### 2016-09-24 doc/paper1-submission/paper3.tex
Update.

\documentclass[fleqn,10pt,lineno,english]{wlpeerj}
\usepackage[T1]{fontenc}
%\usepackage[latin9]{inputenc}
%\usepackage{geometry}
\geometry{verbose,tmargin=2cm,bmargin=2cm,lmargin=2cm,rmargin=2cm}
%\usepackage{xcolor}
%\usepackage{verbatim}
\usepackage{float}
\usepackage{fancybox}
\usepackage{calc}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{natbib}

\makeatletter

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
%% A simple dot to overcome graphicx limitations
\newcommand{\lyxdot}{.}

\floatstyle{ruled}
\newfloat{algorithm}{tbp}{loa}
\providecommand{\algorithmname}{Algorithm}
\floatname{algorithm}{\protect\algorithmname}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
\usepackage{algorithm,algpseudocode}
\usepackage{lmodern}
\definecolor{mylime}{RGB}{128,255,128}

\@ifundefined{showcaptionsetup}{}{%
 \PassOptionsToPackage{caption=false}{subfig}}
\usepackage{subfig}
\makeatother

\usepackage{babel}
\global\long\def\taxonomy{\mbox{\ensuremath{\mathbb{T}}}}

\global\long\def\prunedTaxonomy{\taxonomy_{P}}

\global\long\def\phyloinputs{\mathcal{T}}

\global\long\def\expandedPhylo{\phyloinputs_{E}}

\global\long\def\summaryTree{\mathbb{S}}

\global\long\def\prunedSummary{\summaryTree_{P}}

\global\long\def\collections{\mathcal{C}}

\title{A supertree pipeline for summarizing phylogenetic and taxonomic information
for millions of species}

\author[1,2]{Benjamin D. Redelings}
\author[1]{Mark T. Holder}
\affil[1]{Department of Ecology and Evolutionary Biology, University of Kansas, Lawrence KS, US}
\affil[2]{Department of Biology, Duke University, Durham NC, US}
\corrauthor[1]{Mark T. Holder}{mtholder@ku.edu}

% \keywords{Keyword1, Keyword2, Keyword3}

\begin{abstract}
We present a new supertree method and software implementation that
enables rapid estimation of a summary tree on the scale of millions
of leaves. This software pipeline is called \emph{``propinquity.''}
It uses the dependency-tracking tool \emph{make} to direct the series
of operations required to construct the tree. It relies heavily on
\emph{``otcetera''} - a set of C++ tools to performs most of the
steps of the pipeline. All of the components are free software and
are available on GitHub. This pipeline is currently used by the Open
Tree of Life project to produce all of the versions of project's ``synthetic
tree'' starting at version 5. The propinquity-based pipeline improves
on previous supertree construction methods because it produces no
unsupported branches, and it avoids unnecessary polytomies. In addition
to producing a tree, the pipeline writes an annotations file that
reports which grouping in the input trees support each of the groups
in the supertree.
\end{abstract}

\begin{document}

\flushbottom
\maketitle
\thispagestyle{empty}

\section{Background}

The Open Tree of Life project seeks to build a platform for summarizing
what is known about phylogenetic relationships across all of Life
\citep{HinchliffEtAl2015}. One primary goal of the project is to
build a summary tree from a comprehensive taxonomic tree and a set
of published trees. The summary tree is intended to transparently
and justifiably represent phylogenetic information from these inputs.
The taxonomic tree is derived from the Open Tree Taxonomy (OTT hereafter,
publication in preparation). The phylogenetic inputs are published
trees that have been curated to align the tips to OTT and to identify
the correct rooting (see \citealt{McTavishEtAt2015} for further details
of the curation tools). Unlike OTT, these phylogenetic trees do not
include all leaf taxa. The inputs (taxonomy and phylogenetic trees)
and the output summary supertree are all rooted. Here we describe
the software pipeline (\emph{propinquity}) that summarizes and integrates
these smaller source trees and the taxonomy tree into a single supertree
and the noteworthy tools for manipulating and solving supertrees in
the \emph{otcetera} package.

\subsection{Goals}

Translating the goals of the Open Tree of Life's summary tree into
an explicit set of criteria is not trivial. The summary supertree
should represent the phylogenetic information from source trees in
a transparent and justifiable fashion. We would like to allow users
to correct errors in the supertree by improving the input information
rather than requiring modification to the supertree algorithm. The
pipeline was designed to create a tree which:
\begin{enumerate}
\item displays no unsupported groups,
\item defers to groupings from higher ranked trees in the case of conflict, 
\item contains no unnecessary polytomies, and
\item displays as many groupings from input trees as possible.
\end{enumerate}
These goals are described more fully below. In order to accomplish
transparency and justification, our pipeline also produces annotations
files with information about conflict and support.

\subsubsection{Goal 1: Each grouping is supported by at least one input}

We require that each edge in the supertree be supported by at least
one input tree edge. In addition to aiding interpretability, this
requirement keeps the supertree from arbitrarily representing information
that comes from none of the input trees. Of course, in a supertree
analysis, the full tree will imply some relationships for subsets
of the taxa that are not found in any input tree. So, the meaning
of ``supported by'' needs some clarification.

\paragraph*{Notation, terminology, and the definition of ``supported by''}

Let $\summaryTree$ denote a supertree, and $T_{i}$ denote the $i$th
input tree. The set of taxa that are mapped to the tips of the tree
$T_{i}$ is $\mathcal{L}(i)$. $\summaryTree(i)$ denotes the summary
tree induced by tip nodes that are mapped to taxa in $\mathcal{L}(i)$
and the most recent common ancestor of those leaves, and any other
node that is an ancestor of some but not all of these leaves. We say
that edge $j$ of the supertree is compatible with an input tree,
$T_{i}$ if edge $j$ either is not included in the induced tree $\summaryTree(i)$
or none of the edges in $T_{i}$ are in conflict with edge $j$ in
the induced tree.

We can consider whether or not an node in an input tree is displayed
by $\summaryTree$. For any such node $j$ there is a set of taxa
that are mapped to the tips that descend from the node. This set of
taxa is the ``cluster'' of taxa corresponding to node $j$; it can
be denoted $\mathcal{L}(i,j)$. It will also be referred to as the
``include set'' of the node. The ``exclude set'' of $j$ in $T_{i}$
is the set of taxa in $\mathcal{L}(i)$ but not in $\mathcal{L}(i,j)$.
If the cluster of taxa for any supertree node in $\summaryTree(i)$
is identical to $\mathcal{L}(i,j)$, then we say that $\summaryTree$
displays node $j$ of $T_{i}$. We say that the summary tree \emph{displays}
edge $j$ if the summary tree displays the child node of edge $j$.
Operationally, we can find the most recent common ancestor (MRCA)
node of $\mathcal{L}(i,j)$ in $\summaryTree$; the summary tree displays
$j$ if and only if that MRCA node is not an ancestor of any member
of the exclude set of $j$. We say that edge $k$ of the summary tree
is \emph{supported by} node $j$ in $T_{i}$ if the summary tree displays
node $j$, but if we contracted edge $k$ then the modified summary
tree would no longer display node $j$.

Note that stating that a node in the summary tree is supported by
an input does not imply that every descendant of that node must be
present in the input nor that every taxon that is not a descendant
must be excluded in order to display the node. Consider the problem
shown in figure \ref{fig:toyambig}; panels (\ref{fig:toyambig}a)
and (\ref{fig:toyambig}b) show two input trees. Because taxa A and
E do not occur together in either input, there is some uncertainty
about where to place them. By our terminology, either output shown
in (\ref{fig:toyambig}c) or (\ref{fig:toyambig}d) would be characterized
as a tree that displays all of the input groupings and which has no
unsupported groups. Clearly these criteria are insufficient to specify
a unique solution, and users of the output tree need to be aware that
it may be possible for some taxa to ``float'' to multiple positions.
In figure \ref{fig:toyambig}, taxon E floats to different positions
in (\ref{fig:toyambig}c) and (\ref{fig:toyambig}d), whereas taxon
A does not.

\begin{figure}
\subfloat[Input tree 1]{\includegraphics[width=0.45\textwidth]{Figures__toy_ambig__abcd}

}\hfill{}\subfloat[Input tree 2]{\includegraphics[width=0.45\textwidth]{Figures__toy_ambig__bcde}

}

\subfloat[Summary tree]{\includegraphics[width=0.45\textwidth]{Figures__toy_ambig__soln1}

}\hfill{}\subfloat[Summary tree]{\includegraphics[width=0.45\textwidth]{Figures__toy_ambig__soln2}

}

\caption{An example demonstrating that our definition of ``supported by''
does not imply entire composition of a grouping. (a) and (b) show
2 input trees and (c) and (d) depict trees that each display each
of the groupings in the input trees and which have no unsupported
nodes. The BUILD algorithm (section \ref{fig:subproblem-solutions.})
would choose tree (d) that floats taxon E closer to the root.}

\label{fig:toyambig}
\end{figure}

One of our aims in supertree construction is to minimize the amount
of information in the supertree that does not come from input trees.
We permit information that comes from combinations of input trees,
but not any single input tree. However, we seek to exclude information
that comes from none of the input trees. This motivates the criterion
of not having any unsupported edges, since these edges could be removed
without decreasing the support from any input tree.

\subsubsection{Goal 2: Tree ranking}

An appealing goal for the summarization would be to find the supertree
that displays the largest number of input tree edges. As discussed
in \citet[pages 92 and 131;][]{HusonRS2010} the maximum compatibility
problem is known to be $NP$-hard via a reduction to Max-Clique \citep{Karp1972}.
In addition to being computationally daunting, this formulation of
the supertree problem does not provide biologists who use the summarization
tool with an obvious avenue for fixing perceived problems with the
summary tree. For example, a grouping that a biologist expected may
not be present in the supertree, but it may not conflict with any
of the input groupings which are displayed. This can happen because
displaying both node $a$ from $T_{1}$ and node $b$ from tree $T_{2}$
in a summary tree may only be possible by displaying a grouping that
is present in no input tree. All other factors being equal, if this
implied grouping conflicts with input node $c$ in tree $T_{3}$,
then $c$ will not be displayed in the summary tree, but a biologist
will not necessarily know how to fix this problem. One solution is
to use a ranking of groupings. If an expert were quite confident in
the $c$ grouping, then she could assign that input node a high ranking.
A supertree that used ranks could then recover this grouping even
if its inclusion did not increase the total number of input nodes
that are displayed by the summary tree. Figure (\ref{fig:pairwisecompat})
shows an example of 3 input trees for which there is no pairwise incompatibility,
but no solution displays all of the input groups. Alternative rankings
of inputs can result in one of three summary trees shown in panels
(d-f).

\begin{figure}
\subfloat[Input tree 1]{\includegraphics[scale=0.6]{Figures__toy_pairwise_compat__abc}

}\hfill{}\subfloat[Input tree 2]{\includegraphics[scale=0.6]{Figures__toy_pairwise_compat__acd}

}\hfill{}\subfloat[Input tree 3]{\includegraphics[scale=0.6]{Figures__toy_pairwise_compat__bcd}

}

\subfloat[Summary tree]{\includegraphics[scale=0.6]{Figures__toy_pairwise_compat__s12}

}\hfill{}\subfloat[Summary tree]{\includegraphics[scale=0.6]{Figures__toy_pairwise_compat__s13}

}\hfill{}\subfloat[Summary tree]{\includegraphics[scale=0.6]{Figures__toy_pairwise_compat__s23}

}

\caption{An example of 3 input trees shown in (a), (b), and (c) which do not
conflict in a pairwise manner, but cannot be jointly displayed in
one tree. The 3 solution trees are shown in panels (d-f). Panel (d)
for ranking the tree in (c) lowest. Panel (e) shows the solution if
the tree in (b) has the lowest rank. Panel (f) shows the solution
if the tree in (a) is ranked lowest. Each of the solutions displays
2 of the 3 input groupings. }

\label{fig:pairwisecompat}
\end{figure}

Any approach to supertree construction must deal with the need to
adjudicate between conflicting input trees. We choose to deal with
conflict by ranking the input trees, and preferring to include edges
from higher-ranked trees. The merits of using tree ranking are questionable
because the system does not mediate conflicts based on the relative
amount of evidence for each alternative. However, it is a reasonable
starting point. It has the benefits of making it easy to see why some
groups are included or not (transparency), and it allows simpler and
cleaner algorithms.

Note that if some edge $c$ conflicts with a higher-ranked edge $b$,
then $c$ may still be included in the supertree. This can occur when
the higher ranked edge $b$ conflicts with a yet-higher ranked edge
$a$, and thus $b$ is not included. In that case, it will be possible
for $c$ to be represented in the summary tree. Thus, the fact that
the summary tree displays an input edge does not imply that none of
the higher ranked input trees conflict with that edge.

In order to produce a comprehensive supertree, we also require a rooted
taxonomy tree in addition to the ranked list of rooted input trees.
Unlike other input trees, the taxonomy tree is required to contain
all taxa, and thus has the maximal leaf set. We make the taxonomy
tree the lowest ranked tree. In our current formulation, the taxonomy
tree is also unique in that the taxonomy is the only source of taxonomic
names. Each node in the taxonomy tree corresponds to a named group.
Taxonomic groups may have the same name, but each node in the taxonomy
tree is identified by a unique number (its OTT Id). Taxonomic groups
are identified in the summary supertree by finding a branch (or ``node'')
that has exactly the same include|exclude relationship. The taxonomy
supertree can meaningfully possess degree-two nodes. Although these
nodes can be removed without affecting the relationships of the leaves,
they do represent nested taxonomic groups that contain exactly one
subgroup. The taxonomy is also used to determine which tips are terminal
taxa.

\subsubsection{Goal 3: Contain no unnecessary polytomies}

The supertree should be as resolved as possible - in other words,
it should have no unnecessary polytomies. Thus, for each input edge
that is \emph{not} included, we can point to a reason for non-inclusion
by showing that the input edge conflicts with some edge of the summary
tree. Note, that the requirement to not display unsupported groups
leads to some ``necessary'' polytomies. For example any resolution
of the polytomy shown in figure (\ref{fig:pairwisecompat}e) would
continue to display the same two input groups, but the additional
grouping would be unsupported, because the unresolved tree already
displays the two input groups. Thus, the additional group would be
unsupported and the unresolved tree would be preferred by our criterion.
However, collapsing either internal edge of the tree shown in figure
(\ref{fig:pairwisecompat}d) would result in a tree which displays
only one input grouping. This tree would contain an unnecessary polytomy,
because the polytomy would permit refinement to the depicted tree
which displays more input groupings.

\subsubsection{Goal 4: Display as many input nodes as feasible}

We also seek to construct a supertree that represents as many input
tree nodes as possible. Since non-included input tree nodes must conflict
with the supertree (or they would have been added), this criterion
is the same as minimizing the number of input nodes that conflict
with the supertree.

\subsubsection{Summary of goals}

These optimality criteria help to define what it means for the supertree
to represent the input trees, as well as justifying and explaining
why various features of the supertree exist. The pipeline described
below produces a supertree that satisfies the first three optimality
criteria and is a greedy approximation of a solution to the fourth
goal. It is not guaranteed to display as many input nodes as possible.
Even if the summary tree does accomplish goal 4, it is not necessarily
a \emph{unique} optimum. The pipeline takes a greedy approach to producing
a summary tree by attempting to add groupings from the trees in order
of the trees ranking. This can be viewed as a greedy solution to the
problem of finding the tree with the maximum sum of displayed groups'
weighted scores criterion (described in the appendix A) where the
weights from the trees are so extreme that displaying one group from
a highly ranked tree is preferred to displaying all of the groupings
from lower ranked trees.

\section{Description of the propinquity pipeline}

\subsection{Preprocessing steps}

Propinquity was designed to function as a part of the Open Tree of
Life software architecture, so the first few steps of the pipeline
involve transforming artifacts from that project into a set of rooted
trees and a phylogenetic taxonomy. The phylesystem API \citep{McTavishEtAt2015}
of Open Tree allows users to curate published estimates of trees and
create ranked collections of these trees. Early steps in the propinquity
pipeline manipulate the phylogenetic input trees to improve their
usability and reliability. The first steps of the pipeline (see Figure
\ref{fig:pipeline}) collect a list of trees to include (in the \texttt{phylo\_input}
subdirectory) and store copies of these files (in the \texttt{phylo\_snapshot}
subdirectory) to make it easier to replicate the operation (because
the collection of trees and the tree files change due to curation).
\begin{figure}
\begin{centering}
\includegraphics[height=0.75\paperheight]{paper-pipeline-tools.pdf}
\par\end{centering}
\caption{Organization of the \emph{propinquity} pipeline. Each colored pentagon
labels a program (blue for otcetera-based tools and red for python
scripts in the propinquity or peyotl repository) that performs the
important operations in each step; the number before the tool name
refers to the section in this paper that describes the operation.
The output of each step corresponds to a subdirectory of the propinquity
system which will hold the output artifacts for the step. Ovals are
resources that are required (OTT and Open Tree's phylesystem repository).
White pentagons are user-controlled inputs.}
\label{fig:pipeline}
\end{figure}

\subsubsection{Pruning questionable taxa from the taxonomy}

OTT is a hierarchy of taxonomic names that implies a phylogenetic
taxonomy. An OTT ID has a position in the hierarchy, a taxonomic name,
and set of references to the same name in different taxonomies. In
addition, the ID may also be associated with a set of flags that can
indicate that the taxon may be questionable. These flags can either
encode information taken from an input taxonomy (for example, taxa
the NCBI refers to as ``unplaced'' are assigned an ``unplaced''
flag) or can arise because of some form of conflict during taxonomy
construction (for example, if two taxonomies disagree on the name
for a taxon, then the taxon can be merged and the name will be retained
without any descendants; this name will have an OTT Id, but will be
flagged as ``barren''). Propinquity prunes the OTT down to a more
reliable taxonomy by pruning off parts of the tree that are flagged
with suspicious flags. The set of flags that lead to a subtree of
the taxonomy being pruned is under the control of the user (the set
of flags used by the Open Tree of Life project can be found in the
\texttt{config.opentree.synth} file in the propinquity repository).
For the purpose of the rest of the pipeline, an OTT Id that has been
pruned from the taxonomy will be treated in the same way as invalid
OTT Id. The output of this step is stored in propinquity's \texttt{cleaned\_ott}
subdirectory; this operation only needs to be performed when the OTT
or the pruning flags change.

\subsubsection{Pruning problematically mapped tips from input phylogenetic trees}

Frequently, phylogenetic estimates are rooted using the outgroup criterion,
which is an assumption about the monophyly of the ingroup taxa. Because
the rooting of the branches in the outgroup portion of the tree is
often uncertain, data curators can identify the ingroup node of the
tree; propinquity uses this annotation to prune off the outgroup taxa.

Frequently, not all tips in a phylogenetic input will have been mapped
to a taxon in the current version OTT. Unmapped leaves are pruned
from each phylogenetic input. In some cases, the OTT has changed and
a taxon has been unambiguously mapped to another taxon. This can occur
when multiple species in one version of the taxonomy are ``lumped''
into a single taxon in a subsequent version. OTT maintains a set of
``forwarding'' statements about IDs that have been removed but can
be mapped to an existing taxon; propinquity uses these statements
to update the OTU mapping of input trees.

Finally some leaves are mapped to taxa that occur more than once in
the tree, or taxa that have ancestors represented as tips of the tree.
In these cases, leaves are pruned to assure that tips are mapped to
unique taxa that are not nested. In the case of nested taxa, the tip
mapped to the higher level taxon is pruned, and one of the lower level
tips is retained. In the case of duplicate occurrences of an OTT taxon,
propinquity checks to see if a data curator has selected one of the
taxa to be the exemplar for the taxon. If this selection has not been
made, then the node with the lexicographically lowest ID is chosen
to exemplify the taxon. This choice is arbitrary, but repeatable.
The pruned phylogenetic inputs are stored in a \texttt{cleaned\_phylo}
subdirectory of propinquity.

\subsubsection{Exemplifying tips mapped to higher taxa}

Many input trees have tips that are not terminal taxa, but higher-order
taxonomic groups. It is not clear how to interpret a tip in a phylogenetic
estimate that is labeled with the name of a higher taxon. Several
scenarios can lead to these cases: the data for the tip could have
been created by merging a chimeric set of character scores from constituent
taxa; the species sampled may not have been identified to the lowest
taxonomic rank; or the researcher may simply have used a higher taxonomic
name because he/she assumed that the taxon is monophyletic and the
higher level name would be more recognizable. Rather than allowing
the ambiguity about interpretation of the higher-taxon mapped tips
to propagate throughout the entire pipeline, we transform the input
trees by replacing higher taxa at tips with a set of terminal-taxon
exemplars for each taxon. One approach would be to simply determine
all descendant terminal taxa and attach them as children of the problematic
tip. However, this would create a clade rather than a tip; subsequent
steps in the supertree would interpret the clade as a claim of monophyly
for the taxon. The input tree may not have tested monophyly of the
clade, so this interpretation is unwarranted. We avoid it by attaching
exemplar taxa as child nodes of the higher taxonomic tip but then
collapsing the edge between the node that connected the higher taxon
to its parent. Thus, if $A$ is a non-terminal taxon containing terminal
descendants $a_{1}$ and $a_{2}$ and $B$ is a non-terminal taxon
containing terminal descendants $b_{1}$ and $b_{2}$ we would replace
the subtree $((A,B),c)$ with $((a_{1},a_{2},b_{1},b_{2}),c)$ instead
of the subtree $(((a_{1},a_{2})A,(b_{1},b_{2})B),c)$.

If a taxon is only present in the taxonomy (not in any of the input
trees), then it can be pruned from the taxonomy for the construction
of the supertree and then grafted back on to the summary tree later.
Performing this pruning reduces the size of the supertree problem,
reducing the running time of the pipeline. Similarly, when we expand
a higher taxon in the exemplification step, we can omit members of
the taxon if they do not occur in any of the phylogenetic inputs.
If there are no members of the higher taxon sampled in any other input
tree, then we arbitrarily choose one terminal taxon to represent the
higher taxon. During the exemplification step, a tool from otcetera
(\texttt{otc-nonterminals-to-exemplars}) reads the taxonomy and all
of the ``cleaned'' phylogenetic estimates from the previous step.
Reading all of the inputs is necessary to assure that each higher
taxon is replaced with the same set of exemplars regardless of which
tree the higher taxon occurs in, and that the exemplars for a higher
taxon is the union on the set of descendant terminal taxa that have
been sampled in a phylogenetic input. Figure \ref{fig:exemplify-example}
shows an example of how the trees in figure \ref{fig:cleaned-phylo-example}
would be exemplified.

We prune the taxonomy by removing tips that are not present in any
input tree to produce the pruned taxonomy $\mathbb{T}_{p}$. The tips
pruned in this step will be grafted back onto the skeleton of the
summary tree in a subsequent step. A terminal taxon that is represented
only in the taxonomy can be pruned and then regrafted onto the solution
without affecting which nodes are displayed by the final summary tree.
Thus, this procedure does not impede our ability to find a good summary
tree. Removing these tips produces a smaller input to the rest of
the pipeline, which reduces running times. After producing the set
of ``exemplified'' phylogenetic inputs, this tool exports a pruned
down version of the taxonomy that only contains tips that are present
in at least one phylogenetic input. In version 5.0 of the Open Tree
of Life synthetic tree, this version of the taxonomic tree contained
only 41,226 leaves (while the flag-cleaned version of OTT had 2,424,255
leaves).
\begin{figure}
\subfloat[Tree 1]{\includegraphics[scale=0.4]{Figures__cleaned_phylo__ex_2@tree1\lyxdot tre}

}\hfill{}\subfloat[Tree 2]{\includegraphics[scale=0.4]{Figures__cleaned_phylo__ex_2@tree2\lyxdot tre}

}\hfill{}\subfloat[Taxonomy Tree]{\includegraphics[scale=0.4]{Figures__cleaned_ott__cleaned_ott\lyxdot tre}

}

\caption{Input trees and taxonomy tree}
\label{fig:cleaned-phylo-example}
\end{figure}

\begin{figure}
\subfloat[Tree 1]{\includegraphics[scale=0.4]{Figures__exemplified_phylo__ex_2@tree1\lyxdot tre}

}\hfill{}\subfloat[Tree 2]{\includegraphics[scale=0.4]{Figures__exemplified_phylo__ex_2@tree2\lyxdot tre}

}\hfill{}\subfloat[Pruned taxonomy tree $\protect\prunedTaxonomy$]{\includegraphics[scale=0.4]{Figures__exemplified_phylo_taxonomy_uncontested\lyxdot tre}

}

\caption{Exemplified input trees and taxonomy tree from figure \ref{fig:cleaned-phylo-example}.
E in tree1 is exemplified by E1. Pruned taxa are E2, F2, and D. The
taxa E and F are retained as monotypic taxa in the pruned taxonomy
$\protect\prunedTaxonomy$, but are not shown in panel c. The red
edge in the pruned taxonomy tree is an uncontested higher taxon in
the exemplified taxonomy (as explained in section \ref{subsec:Subproblem-decomposition})}
\label{fig:exemplify-example}
\end{figure}

\subsection{Summary tree construction}

After the preprocessing steps, the inputs have been converted to a
set of rooted phylogenetic estimates in which each leaf is mapped
to a terminal taxon in the exemplified taxonomic tree. The goal of
the remainder of the pipeline is to construct a tree that maximizes
the sum displayed groups' weighted scores criterion. This is accomplished
in four steps: (1) dividing the full problem into subproblems based
on uncontested taxa, (2) constructing a summary solution for each
subproblem by greedily creating a maximally-sized list of groupings
that can all be displayed simultaneously; (3) grafting the subproblem
solutions into a single supertree; and (4) grafting (or ``unpruning'')
the taxonomy-only taxa onto the solution to produce a complete summary
tree.

\subsubsection{Subproblem decomposition\label{subsec:Subproblem-decomposition}}

For the sake of efficiency, propinquity uses a divide-and-conquer
approach to construct the supertree. Subproblems are identified by
searching through the taxonomy tree to find any taxa are not contested
by any single input tree. Here we say that input tree $T_{i}$ contests
taxon $x$ in the pruned taxonomy, if $x$ is not monophyletic in
any resolution of tree $T_{i}$. Thus, polytomies in an input tree
are treated as soft polytomies, and a taxon is not contested merely
because it is not displayed by an input tree.

This operation is performed by the \texttt{otc-uncontested-decompose}
tool in otcetera; see appendix \ref{sec:Decompose-algorithm} for
a description of the algorithm. The output is a series of subproblems,
each of which corresponds to a slice of the taxonomy and corresponding
slices through each relevant input tree. Each uncontested non-terminal
and non-root taxon will show up in two subproblems: it will be the
root of its own subproblem and it will be tip in the subproblem that
covers the next slice deeper in the tree. The red edge in figure \ref{fig:exemplify-example}c
highlights the taxa that are not contested by the input shown in \ref{fig:exemplify-example};
figure \ref{fig:subproblems-example} shows the subproblems that would
be emitted as a result of this set of inputs. The supertree operation
of \citet{HinchliffEtAl2015} also used this otcetera-based decomposition
step.

Note that decomposition into uncontested groups does not necessarily
allow us to find the tree that maximizes the MSDGWS score. For example,
see figure \ref{fig:decompose-worsens}; that example is a variant
of the situation shown in figure \ref{fig:pairwisecompat}. In this
case the groupings from each of the phylogenetic estimates, shown
in panels \ref{fig:decompose-worsens}a and (\ref{fig:decompose-worsens}b,
could be displayed. That solution is shown in panel \ref{fig:decompose-worsens}d,
it displays two of the three input splits, but is optimal because
no solution displays all three input groupings and the depicted solution
displays the two highest ranked groupings. However, neither of the
trees shown in panels \ref{fig:decompose-worsens}a or \ref{fig:decompose-worsens}b
contest the taxon $B$ shown in the taxonomy panel \ref{fig:decompose-worsens}c.
Thus, when using our decomposition, the branches leading to taxa $B1$
and $B2$ in the input phylogenetic trees would be sliced during the
decomposition, and relabeled to refer to taxon $B$. This taxonomically-informed
interpretation of the inputs views the two phylogenetic inputs as
in conflict; so the solution returned by propinquity would defer to
the higher ranked tree. The tree shown in panel \ref{fig:decompose-worsens}e
would be returned. This example arises from the fact that the trees
in \ref{fig:decompose-worsens}a and \ref{fig:decompose-worsens}b
jointly contest taxon $B$, but neither contests taxon $B$ when the
trees are considered in isolation.

Despite the fact that the use of \texttt{otc-uncontested-decompose}
can worsen the final score of the summary tree, we use this approach
in propinquity because it makes the construction of the tree faster
and it is easy for users to correct issues caused by incorrect taxa
being constrained to be monophyletic. By adding a tree (even a low-ranked
tree) that contests a taxon to the corpus of input trees , then the
next synthetic tree will no longer consider the taxon to be uncontested.
Thus the procedure encourages curation of more phylogenetic inputs
as a means of improving the summary tree.

In version 5.0 of the summary tree, this decomposition procedure produced
5,545 subproblems, but only 1,586 of these were non-trivial to solve.
If a subproblem contains only two tips it is trivial; 2,443 subproblems
were trivial in this way. Similarly, if a subproblem contains only
2 trees it is trivial to solve because the solution will simply be
all of the groupings from the first tree combined with all of the
groupings from the second tree that are compatible with the first
tree; 2,838 subproblems were trivial in this way. The subproblem with
the largest number of tips contained 953 tips. The largest subproblem,
in terms of the number of input trees (including the taxonomic tree)
that were relevant, had 18 trees. Without decomposition, the supertree
problem would have had 512 input trees and 41,226 leaves.

\begin{figure}
  \subfloat[Subproblem \emph{ABCD}]{\includegraphics[width=0.3\textwidth]{Figures__subproblems__14__1\lyxdot tre}%
    \hfill{}\includegraphics[width=0.3\textwidth]{Figures__subproblems__14__2\lyxdot tre}%
    \hfill{}\includegraphics[width=0.3\textwidth]{Figures__subproblems__14__3\lyxdot tre}

}

  \subfloat[Subproblem \emph{root}]{\includegraphics[width=0.3\textwidth]{Figures__subproblems__17__1\lyxdot tre}%
    \hfill{}\includegraphics[width=0.3\textwidth]{Figures__subproblems__17__2\lyxdot tre}%
    \hfill{}\includegraphics[width=0.3\textwidth]{Figures__subproblems__17__3\lyxdot tre}

}

\caption{Subproblems generated from the exemplified trees shown in figure \ref{fig:exemplify-example}.
A trivial statemen from the first tree that a taxon labelled $ABCD$
is sister to $E$ has been omitted, because trees with only 2 leaves
do not contain phylogenetic information.}
\label{fig:subproblems-example}
\end{figure}
\begin{figure}
\subfloat[]{\includegraphics[scale=0.6]{Figures__uncontested_worsens_score__ab1c}

}\hfill{}\subfloat[]{\includegraphics[scale=0.6]{Figures__uncontested_worsens_score__acb2}

}\hfill{}\subfloat[]{\includegraphics[scale=0.6]{Figures__uncontested_worsens_score__tax}

}

\hfill{}\subfloat[]{\includegraphics[scale=0.6]{Figures__uncontested_worsens_score__opt}

}\hfill{}\subfloat[]{\includegraphics[scale=0.6]{Figures__uncontested_worsens_score__returned}

}\hfill{}

\caption{An example with three input trees: the highest ranked phylogenetic
input panel (a), the second ranked phylogenetic input (b), and the
taxonomy in panel (c). The summary tree in panel (d) has the highest
possible score, but the summary shown in panel (e) would be returned
from the pipeline that uses uncontested taxon decomposition.}
\label{fig:decompose-worsens}
\end{figure}

\subsubsection{Subproblem solution\label{subsec:Subproblem-solution}}

When solving sub-problems, we sequentially incorporate splits from
trees in order of ranking, retaining splits that are compatible with
the current set of splits (Alg \ref{alg:ConsistentSplitsFromRankedList}).
The order of splits from the same tree is not specified by this approach,
and we incorporate splits using one of the possible post-order traversals
of the tree. We make use of the BUILD algorithm \citep{AhoSSU1981}
to assess compatibility. This strategy avoids unnecessary polytomies,
since splits of later input trees are only rejected from the summary
supertree if they conflict with higher-priority splits. Finally, we
use the BUILD algorithm to construct a supertree displaying all of
the splits in the set of compatible splits. Using the BUILD algorithm
to construct the subproblem summary tree satisfies criterion 3, because
trees from the BUILD algorithm do not contain unsupported branches.
\begin{algorithm}
\begin{algorithmic}
\Require An ordered list of $M$ splits, $\mathcal{R} = [R_1, R_2, R_3, \ldots, R3, \ldots, R_M]$
\State $\mathcal{C} = [R_1]$
\For{each split $i$ in $[2, 3 \ldots M]$}
   \State $\mathcal{T} \leftarrow \mathcal{C} + R_i$ \Comment{where `+' means concatenating 2 lists}
   \If{\textsc{BUILD}$(\mathcal{T})$ does not return null}
      \State $\mathcal{C} \leftarrow \mathcal{T}$
   \EndIf
\EndFor{}
\State\Return $\mathcal{C}$
\end{algorithmic}

\caption{\label{alg:ConsistentSplitsFromRankedList}ConsistentSplitsFromRankedList}
\end{algorithm}

The BUILD algorithm as originally stated by \citet{AhoSSU1981} applies
to a collection of rooted triplets. Instead of decomposing each input
split into a collection of rooted triplets, we instead modify the
BUILD algorithm to apply directly to larger rooted splits. The modified
BUILD algorithm constructs a tree compatible with a collection of
rooted splits, and returns failure if such a tree does not exist.
This modified algorithm recovers the original BUILD algorithm if only
rooted triplets are supplied as input. When larger splits are supplied
as input, the results are the same as if each was was decomposed into
all implied triplets. The modified build algorithm has order $O(N^{2}+N^{2}E+NL)$
where $N$ is the number of splits passed in, $E$ is the average
size of the exclude group, and $L$ is the total number of leaves.
This simplifies to $O(N^{2})$ if all splits are triplets. In this
approach splits are either entirely retained or entirely discarded
- consistent rooted triplets from conflicting splits are not retained.
However, when unpruning taxonomy-only taxa (see below), we make an
attempt to break ties in a way that preserves some partial information
from conflicting splits by attaching taxa from conflicting splits
at their common ancestor. Figure \ref{fig:subproblem-solutions.}
shows the solutions that would be obtained by applying our modified
version of the BUILD algorithm to the subproblems show in figure \ref{fig:subproblems-example}.

\begin{figure}
\hfill{}\subfloat[Solution to subproblem \emph{ABCD}]{\includegraphics[scale=0.45]{Figures__subproblem_solutions__ott14\lyxdot tre}

}\hfill{}\subfloat[Solution to subproblem \emph{root}]{\includegraphics[scale=0.45]{Figures__subproblem_solutions__ott17\lyxdot tre}

}\hfill{}

\caption{Solutions to the subproblems depicted in figure \ref{fig:subproblems-example}}
\label{fig:subproblem-solutions.}
\end{figure}

\subsubsection{Solution grafting}

To produce a tree that spans all of the taxa sampled in the exemplified
set of input trees, we graft the subproblem solutions into a single
tree (stored as the \texttt{grafted\_solution/grafted\_solution.tre}
by propinquity). Recall that each non-root uncontested taxon used
for decomposition occurs as a leaf taxon in one subproblem and as
a root taxon in one other subproblem. Thus, the grafting operation
simply consists of reading all of the subproblem solutions into memory
and then merging the nodes that are labeled with the same OTT ID.

\begin{figure}
\begin{centering}
\includegraphics[scale=0.5]{Figures__grafted_solution__grafted_solution_ottnames\lyxdot tre}
\par\end{centering}
\caption{Grafted solution. The nodes E and F are monotypic here, but will end
up being polytypic after unpruning is performed.}
\end{figure}

\subsubsection{Unpruning unsampled taxa}

As described above, taxa that do not have any descendants in a sampled
phylogenetic input are pruned from the taxonomy for the sake of efficiency.
These taxa are reattached by an ``unpruning step.'' For those taxa
that are compatible with the grafted tree, this step simply amounts
to adding any unsampled taxonomic children to the node that represents
the taxon in the grafted solution tree.

However, a taxon may be incompatible with the grafted solution; we
refer to such taxa as ``broken taxa.'' If a broken taxon contains
some unsampled children, it is not clear where these unsampled children
should be attached to the grafted solution. One approach would be
to mimic the application of apply Algorithm \ref{alg:ConsistentSplitsFromRankedList}
to the full (unpruned) taxonomic tree. This would be equivalent to
collapsing each edge in the taxonomy that attaches a broken taxon
to its parent. The unsampled children of broken taxa would attach
at their least inclusive ancestral taxon which is unbroken. In cases
of several adjacent taxa are broken, this can lead to polytomies of
very high degree deep in the tree. This can make the summary tree
difficult to navigate. Thus, we have adopted an alternative solution.
The \texttt{otc-unprune-solution-and-name-unnamed-nodes} tool from
otcetera attaches the unsampled children of a broken taxa to the grafted
solution as children of the MRCA of the sampled children.

Figure \ref{fig:Two-approaches-to-unpruning} illustrates the two
approaches to unpruning. Taxa G, M, and R (Fig \ref{fig:Two-approaches-to-unpruning}a)
are broken because they conflict with the grafted solution (Fig \ref{fig:Two-approaches-to-unpruning}b);
among these, only taxon R has children that were unsampled in the
grafted solution. Ignoring all broken taxa when unpruning would cause
the unsampled children (R4, R5, and R6) to attached directly at taxon
N (as in the tree shown in Fig \ref{fig:Two-approaches-to-unpruning}c),
because that is the least inclusive unbroken ancestor of R. The tree
illustrated in Fig \ref{fig:Two-approaches-to-unpruning}d shows the
tree that would be produced by propinquity; the children of the broken
taxon R and instead attached at the MRCA of sampled children (R1,
R2, and R3). Their attachment point does not correspond to a taxon
in the taxonomic tree.

\begin{figure}
\begin{centering}
\subfloat[Taxonomy]{\begin{centering}
\begin{minipage}[c][1\totalheight][t]{0.45\columnwidth}%
\begin{center}
\includegraphics[scale=0.4]{example3__cleaned_ott-labels\lyxdot tre}
\par\end{center}%
\end{minipage}
\par\end{centering}
}\hfill{}\subfloat[Grafted solution]{\begin{centering}
\begin{minipage}[c][1\totalheight][t]{0.45\columnwidth}%
\begin{center}
\includegraphics[scale=0.4]{example3__grafted_solution-labels\lyxdot tre}
\par\end{center}%
\end{minipage}
\par\end{centering}
}\hfill{}
\par\end{centering}
\begin{centering}
\subfloat[Unpruned tree - broken taxa removed]{\begin{centering}
\includegraphics[scale=0.4]{example3__unprune-old-labels\lyxdot tre}
\par\end{centering}
}\hfill{}\subfloat[Unpruned tree with MRCA-attachment]{\begin{centering}
\includegraphics[scale=0.4]{example3__unprune-new-labels2\lyxdot tre}
\par\end{centering}
}
\par\end{centering}
\caption{\label{fig:Two-approaches-to-unpruning}Two approaches to unpruning.
Taxa G and R in the taxonomy (a) are broken because they conflict
with the grafted solution (b). Removing these broken taxa from the
taxonomy before unpruning leads to taxa R4, R5, and R6 being attached
directly at taxon N, as in tree (c). In tree (d), the children of
the broken taxon R and instead attached at the MRCA of R1, R2, and
R3.}
\end{figure}

\begin{figure}
\begin{centering}
\includegraphics[scale=0.5]{Figures__labelled_supertree__labelled_supertree_only_taxon_names\lyxdot tre}
\par\end{centering}
\caption{Unpruned tree with internal taxa. Nodes E2, F2, and D have been re-added
to the tree. The nodes E and F are no longer monotypic.}
\end{figure}

\subsubsection{Naming unnamed nodes}

In order to annotate each node in the summary supertree, it is first
necessary that each node have a unique identifier. Nodes whose include
group correspond exactly to the include group of a node in the taxonomy
are given the same identifier as the corresponding taxonomy node.
These identifier are of the form $ottX$ where $X$ is an integer
OTT id. We generate a label of the form $mrcaottX_{1}ottX_{2}$ for
an non-taxonomic node $n$ where $X_{1}$ and $X_{2}$ are the OTT
ids for two leaves, $n$ is the MRCA of these leaves, and $X_{1}$
is the numerically smallest OTT ID that is a descendant of $n$, and
$X_{2}$ is the next the smallest ID that can be chosen to designate
$n$ as the MRCA. Because new taxa added to OTT will be given higher
OTT Ids, the use of the lowest numbered OTT Ids as designators increases
the chance that a node label can be encountered in a subsequent version
of the tree (though the taxonomic content may change). The deterministic
choice of designators also makes the labeling insensitive to branch
rotation of the grafted solution tree.

\subsubsection{Annotation}

\begin{figure}
\subfloat[Input tree $T_{1}$ named ``tree1'']{%
\begin{minipage}[c][1\totalheight][t]{0.25\columnwidth}%
\includegraphics[scale=0.5]{annotation_example__input1\lyxdot tre}%
\end{minipage}

}\hfill{}\subfloat[Summary Tree $\protect\summaryTree$]{%
\begin{minipage}[c][1\totalheight][t]{0.35\columnwidth}%
\includegraphics[scale=0.5]{annotation_example__summary\lyxdot tre}%
\end{minipage}

}\hfill{}\subfloat[Induced Tree $\protect\summaryTree(1)$]{%
\begin{minipage}[c][1\totalheight][t]{0.35\columnwidth}%
\includegraphics[scale=0.5]{annotation_example__induced1_color}%
\end{minipage}

}

\bigskip{}

\bigskip{}

\hfill{}\subfloat[JSON annotations relating edges of (c) to edges of (a)]{%
\shadowbox{\begin{minipage}[t]{0.8\columnwidth}%
\texttt{\{}

\texttt{~~\textquotedbl{}nodes\textquotedbl{}: \{}

\texttt{~~~~\textquotedbl{}mrcaott1ott4\textquotedbl{}: \{ \textquotedbl{}}\texttt{\textcolor{blue}{partial\_path\_of}}\texttt{\textquotedbl{}: \{
\textquotedbl{}tree1\textquotedbl{}: \textquotedbl{}node15\textquotedbl{},
... \} \},}

\texttt{~~~~\textquotedbl{}ott10\textquotedbl{}:~~~~~~~~~\{
\textquotedbl{}}\texttt{\textcolor{blue}{partial\_path\_of}}\texttt{\textquotedbl{}: \{
\textquotedbl{}tree1\textquotedbl{}: \textquotedbl{}node15\textquotedbl{},
... \} \},}

\texttt{~~~~\textquotedbl{}mrcaott5ott7\textquotedbl{}: \{ \textquotedbl{}\textcolor{mylime}{supported\_by}\textquotedbl{}:~~~
\{ \textquotedbl{}tree1\textquotedbl{}: \textquotedbl{}node16\textquotedbl{},
... \} \},}

\texttt{~~~~\textquotedbl{}ott1\textquotedbl{}:~~~~~~~~~~\{
\textquotedbl{}}\texttt{\textcolor{orange}{terminal}}\texttt{\textquotedbl{}: ~~~~~~~\{
\textquotedbl{}tree1\textquotedbl{}: \textquotedbl{}ott1\textquotedbl{},
... ~~\} \},}

\texttt{~~~~\textquotedbl{}ott2\textquotedbl{}:~~~~~~~~~~\{
\textquotedbl{}}\texttt{\textcolor{orange}{terminal}}\texttt{\textquotedbl{}: ~~~~~~~\{
\textquotedbl{}tree1\textquotedbl{}: \textquotedbl{}ott2\textquotedbl{},
... ~~\} \},}

\texttt{~~~~\textquotedbl{}ott5\textquotedbl{}:~~~~~~~~~~\{
\textquotedbl{}}\texttt{\textcolor{orange}{terminal}}\texttt{\textquotedbl{}: ~~~~~~~\{
\textquotedbl{}tree1\textquotedbl{}: \textquotedbl{}ott5\textquotedbl{},
... ~~\} \},}

\texttt{~~~~\textquotedbl{}ott11\textquotedbl{}:~~~~~~~~~\{
\textquotedbl{}}\texttt{\textcolor{orange}{terminal}}\texttt{\textquotedbl{}: ~~~~~~~\{
\textquotedbl{}tree1\textquotedbl{}: \textquotedbl{}ott5\textquotedbl{},
... ~~\} \},}

\texttt{~~~~\textquotedbl{}ott7\textquotedbl{}:~~~~~~~~~~\{
\textquotedbl{}}\texttt{\textcolor{orange}{terminal}}\texttt{\textquotedbl{}: ~~~~~~~\{
\textquotedbl{}tree1\textquotedbl{}: \textquotedbl{}ott7\textquotedbl{},
... ~~\} \}}

\texttt{~~\},}

\texttt{~~...}

\texttt{\} }%
\end{minipage}}

}\hfill{}

\caption{\label{fig:displayed-annotations}The relationship of edges in summary
tree $\protect\summaryTree$ (b) to edges in the input tree $T_{1}$
(a). Only edges of $\protect\summaryTree$ that are present in the
induced tree $\protect\summaryTree(1)$ are represented by JSON annotations
(d). Taxon names are here suppressed in favor of OTT ids, and edges
are referenced via their tipward nodes. Edges in $\protect\summaryTree(1)$
that correspond to terminal edges of $T_{1}$ are orange; edges of
$\protect\summaryTree(1)$ that are supported by edges of $T_{1}$
are blue; where multiple edges of $\protect\summaryTree(1)$ correspond
to the same edge of $T_{1}$ they are green. There is no conflict
in this example. Also, if this were output from propinquity, then
each internal node of $\protect\summaryTree$ would be supported by
other inputs trees that are not shown here. }
\end{figure}

\paragraph{Annotations }

To reveal the connections between the groupings found in the a summary
supertree and the input trees, propinquity uses a few Python scripts
and the \texttt{otc-annotate-synth} tool from otcetera to create an
annotations file describing the pipeline used and the connections
between phylogenetic information in the inputs and the summary. The
JSON file produced by \texttt{otc-annotate-synth} encodes a ``nodes''
property that holds a mapping between a node name for the summary
tree (using the naming convention described in the previous section)
and a node provenance object categorizes the relationship between
the node and the inputs. The node provenance object for node $x$
uses several properties to categorize the relationship between the
node and the inputs; each property in the node provenance object maps
to a structure storing the tree ID and node IDs for the input tree
nodes.

Conceptually, this annotation operation in equivalent to considering
every node $j$ in each input tree $i$ and the summary tree node
$x$. Because the vast majority of nodes in the input studies will
be compatible but not directly relevant to node $x$ we do not list
all of the compatible groupings. If node $x$ is not included in the
induced tree $\mathcal{\mathbb{S}}(i)$, then none of the nodes of
tree $i$ will be referred to in the annotations for node $x$. Even
if $x$ is included in $\mathcal{\mathbb{S}}(i)$, many of the nodes
of $T_{i}$ will be compatible with $x$ while being relevant to other
parts of the summary tree. The only input nodes listed for node $x$
are with rooted taxon bipartitions which conflict with, are displayed
by, or are resolved by the the rooted taxonomic bipartition associated
with node $x$. All input nodes that cannot be displayed by any supertree
that contains $x$ are stored in a ``conflicts\_with'' property
of the node provenance object. If node $j$ of $T_{i}$ is displayed
by the summary tree and $x$ is part of the path of $\mathcal{\mathbb{S}}(i)$
that displays the split between descendants of $j$ and other taxa,
then a reference to the node $j$ will be in the node provenance object.
The exact categorization of this annotation will depend on the configuration
of node $x$ on the induced tree $\mathcal{\mathbb{S}}(i)$:
\begin{itemize}
\item if $x$ is a terminal path in the induced tree then node $j$ will
be listed in the ``terminal'' property;
\item if $x$ is along an internal path that contains some nodes with out-degree
equal to 1, then node $j$ will be listed in the ``partial\_path\_of''
property; and
\item if $x$ is along an internal path without any node of out-degree 1,
then node $j$ will be listed in the ``supported\_by'' property,
because node $j$ supports the existence of grouping $x$ in the sense
that collapsing the edge that separates $x$ from its parent would
cause the summary tree to no longer display node $j$.
\end{itemize}
These three relationships are illustrated in Figure \ref{fig:displayed-annotations}.
Finally, if $T_{i}$ does not display $x$ from $\mathcal{\mathbb{S}}(i)$,
but there exists an unresolved node $j$ in $T_{i}$ which could be
resolved such that the tree would then display $x$, then a reference
to node $j$ will be listed in the ``resolves'' property of node
$x$.

The otcetera annotation tool can also detect cases in which including
information from node $j$ in $T_{i}$ could further resolve a polytomy
$x$ in the summary tree; such a case would be annotated using the
``resolved\_by'' property of $x$. However, because of our goal
of excluding unnecessary polytomies, none of the nodes in propinquity's
summary tree will use this annotation when they are annotated with
the set of input trees.

\subsubsection{Self-documentation}

An optional step in the propinquity pipeline (triggered by the executing
the ``make html'' target) can compose an ``index.html'' file for
each directory created during the pipeline to explain the artifacts
held in that directory and report summary statistics about the summarization
run.

\section{Conclusions}

Here we have described the motivation and methodology used by the
propinquity tool that is currently used by the Open Tree of Life project
to build its summary tree. A modified version of the treemachine software
which built the summary tree described in the \citet{HinchliffEtAl2015}
paper is used by the project to serve the tree produced by propinquity
via Open Tree of Life APIs. The migration of summary tree construction
from treemachine (used for version 4) to propinquity (for all versions
from v5.0 to present) has decreased the computational time required
to construct a supertree from several hours to about 8 minutes (after
some preprocessing steps which only have to be performed when the
input taxonomy changes). The amount of RAM required during tree construction
has also decreased substantially. We ran propinquity on the same input
trees used to construct version Open Tree version 4. Unlike the version
4 tree, the tree from propinquity contained no unsupported nodes or
unnecessary polytomies. The number of displayed input splits increased
from 35,518 (for version 4) to 39,639; an 11.2\% increase. The number
of input splits that do not conflict with the summary tree, but are
not incorporated dropped from 2718 to 0. The number of input splits
that conflict with the phylogeny dropped from 2760 to 1357, a decrease
of 1357, or 49.2\%. %

\section{Acknowledgements}

Thanks to Karen Cranston, Jonathan Rees, Jim Allman, Cody Hinchliff,
Stephen Smith, and Joseph Brown for discussions and feedback. Thanks
to NSF grant 1208393 (DEB), part of the collaborative Open Tree of
Life awards, the University of Kansas, and the Heidelberg Institute
for Theoretical Studies for funding.

\bibliographystyle{upmplainnat}
\bibliography{otcetera}

\appendix

\section{\label{sec:The-maximum-sum}The maximum sum displayed groups' weighted
scores criterion}

Let $\mathcal{W}$ is a weighting function that maps any input tree's
internal node to a non-negative number. If $I(\mathbb{S},i,j)$ is
an indicator function that is 1 if summary tree $\mathbb{S}$ displays
the node $V(i,j)$ and 0 otherwise then: $SDGWS(\mathbb{S})=\sum_{i}\sum_{j}I(\mathbb{S},i,j)\mathcal{W}(i,j)$
is the ``sum of displayed groups' weighted scores'' for a tree where
$i$ indexes all of the input trees and $j$ indexes each non-root
internal node in tree $i$. Preference for this tree is referred to
as the maximum sum displayed groups' weighted scores criterion (MSDGWS
criterion). The summary tree constructed by the propinquity pipeline
is a greedy heuristic for finding a tree that maximizes this score
when the weights for a node are determined by the tree's weight and
the difference in weighting is so large that displaying one node from
a highly ranked tree is preferred to displaying all of the nodes in
the trees with lower rank.

\section{Description of the decomposition algorithm of otc-uncontested-decompose}\label{sec:Decompose-algorithm}

The input is the a ranked list of input trees and comprehensive taxonomy.

\subsection{Creation of multigraph of the taxonomy with embedded trees}

The tool creates a multigraph by starting with a graph is isomorphic
to the taxonomic tree. The nodes and edges created in this step will
be referred to as the ``taxonomic graph.'' Next, we add nodes and
edges to that graph in a procedure that we refer to as ``embedding''
the input trees into the taxonomy. A node is introduced for each node
in an input tree, and these nodes are mapped the MRCA nodes in the
taxonomic graph. In other words, for any node $y$ in an input tree
with a cluster of descendants, $\mathcal{C},$ we find the most tipward
node $z$ in the taxonomic graph that is an ancestor to all of the
taxa in $\mathcal{C}$; let $m(y)=z$ refer to this mapping, and $m^{\prime}(z)=y$
refer to the reverse mapping. Each edge $e_{ij}$ in a source tree
$i$ connects ancestor to its descendant, $a(e_{ij})\rightarrow d(e_{ij})$.
The edges are introduced into the graph. We also introduce new edges
to create a from $m(a(e_{ij}))$ through its descendants to $m(d(e_{ij}))$;
we denote this path $p(e_{ij})$ and refer to the edges in the path
as ``embedding edges for tree $i$''. Note, that this is a path
through the taxonomic nodes, while the edge $e_{ij}$ connects source
tree nodes. The mapping between $e_{ij}$ and $p(e_{ij})$ is stored,
and the edges are labelled with the index $i$ so that it is clear
which tree created them. Because the taxonomic tree is highly unresolved,
it is frequently the case that $m(a(e_{ij}))$ is the same node as
$m(d(e_{ij}))$; in these cases the embedding edge is a loop. This
situation occurs whenever edge $e_{ij}$ can resolve part of the polytomy
represented by a taxon.

We treat the taxonomy as the lowest ranked input tree. The next step
will collapse contested edges in the taxonomic graph. To retain all
of the information from the taxonomy we embed the taxonomy into the
taxonomic graph as if it were another input tree

\subsection{Detection of uncontested higher taxa}

After every input tree and the taxonomy tree have been embedded into
the taxonomic graph, we perform postorder traversal over the taxonomic
graph that underlies the multi-graph. For any internal node (each
of which corresponds to a non-terminal taxon) we determine whether
or not it is contested by examining each input tree. We can determine
tree $i$ contests the taxon represented by taxonomic node $x$ by
looking at the parents of all of the ``exiting'' embedding edges
for tree $i$. These are the set of embedding edges that have $x$
as a daughter and have a parent node that is not $x$ (ergo a parent
node that is taxonomically higher than $x$). If there are more than
one parent nodes in this set of exiting embedding edges, then tree
$i$ contests that taxon. If there is only parent node, then the all
of the constituent taxa belonging to this taxon that are present in
tree $i$ have one parent that is more inclusive; this means that
the input tree does not contest monophyly of the taxon.

If the taxon is uncontested, then it may be the case that some input
trees do not contest the taxon, but contain polytomies that could
be resolved to display the taxon. The cases can be identified by finding
multiple exiting embedding edges that have the same taxon $y$ as
their parent node. In these cases, a pseudo input tree node is created
and becomes the parent node for these edges; this new node is then
connected to $y$ as if it had been an input edge. This operation
is equivalent to resolving an input tree's polytomy in favor of the
monophyly of the uncontested taxon. This is the only way in which
the input trees are modified during the decomposition.

\subsection{Collapsing contested taxa from the taxonomic graph}

If a taxonomic node $x$ fails the ``uncontested'' test described
in the previous section, then the node corresponding to the taxon
is removed from the taxonomic graph and the set of edges (and mappings
between input edges and embedded paths) is updated as if this taxon
had not been present in when the taxonomic graph was created. This
consists of detecting changing any edge in an embedding path that
is adjacent to $x$ by replacing the reference to $x$ with a reference
to its parent $a(x)$. Note that we do not collapse the edge corresponding
to this taxon in the part of the graph that represents the embedding
of the taxonomy into the taxonomic graph. Thus, the taxonomy will
still claim the monophyly of the taxon. This is relevant if the input
grouping that contests taxon $x$ is overruled (in the subproblem
solution step) by a higher ranked split. In other words, the fact
that the a taxon is contested during the decomposition is not a guarantee
that the taxon will not be monophyletic in the final supertree.

\subsection{Emitting subproblems}

Whenever an uncontested taxon is identified, the appropriate slice
of each input trees that intersect with the taxon is written to a
file. Then the multigraph is simplified by slicing any off the taxon.
This slicing is accomplished by examing all of the exiting embedding
edges for the taxon. The descendant taxa of each input tree is relabeled
with the identifier of the contested taxa and all of that node's descendants
are removed. Thus this input node will act as as if it were a leaf
mapped to the uncontested taxon. All descendants of the taxonomic
graph are also pruned off.
\end{document}

### 2016-09-24 doc/paper1-submission/paper3.tex
Update ms.

@@
-\documentclass[fleqn,10pt,lineno,english]{wlpeerj}
+\documentclass[fleqn,12pt,english]{wlpeerj}
 \usepackage[T1]{fontenc}
@@
 %\usepackage{geometry}
-\geometry{verbose,tmargin=2cm,bmargin=2cm,lmargin=2cm,rmargin=2cm}
+\geometry{verbose,tmargin=1in,bmargin=1in,lmargin=1in,rmargin=1in}
 %\usepackage{xcolor}
@@
 
+%\def\fps@figure{ht}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
@@
 \author[1,2]{Benjamin D. Redelings}
-\author[1]{Mark T. Holder}
-\affil[1]{Department of Ecology and Evolutionary Biology, University of Kansas, Lawrence KS, US}
-\affil[2]{Department of Biology, Duke University, Durham NC, US}
-\corrauthor[1]{Mark T. Holder}{mtholder@ku.edu}
+\author[2,3,4]{Mark T. Holder}
+\affil[1]{Department of Biology, Duke University, Durham NC, US}
+\affil[2]{Department of Ecology and Evolutionary Biology, University of Kansas, Lawrence KS, US}
+\affil[3]{Biodiversity Institute, University of Kansas, Lawrence KS, US}
+\affil[4]{Heidelberg Institute for Theoretical Studies, Schloss-Wolfsbrunnenweg 35, 69118 Heidelberg, Germany}
+\corrauthor[2,3,4]{Mark T. Holder}{mtholder@ku.edu}
 
@@
 \begin{figure}
-\subfloat[Input tree 1]{\includegraphics[scale=0.6]{Figures__toy_pairwise_compat__abc}
-
-}\hfill{}\subfloat[Input tree 2]{\includegraphics[scale=0.6]{Figures__toy_pairwise_compat__acd}
-
-}\hfill{}\subfloat[Input tree 3]{\includegraphics[scale=0.6]{Figures__toy_pairwise_compat__bcd}
+\subfloat[Input tree 1]{\includegraphics[scale=0.57]{Figures__toy_pairwise_compat__abc}
+
+}\hfill{}\subfloat[Input tree 2]{\includegraphics[scale=0.57]{Figures__toy_pairwise_compat__acd}
+
+}\hfill{}\subfloat[Input tree 3]{\includegraphics[scale=0.57]{Figures__toy_pairwise_compat__bcd}
 
@@
 
-\subfloat[Summary tree]{\includegraphics[scale=0.6]{Figures__toy_pairwise_compat__s12}
-
-}\hfill{}\subfloat[Summary tree]{\includegraphics[scale=0.6]{Figures__toy_pairwise_compat__s13}
-
-}\hfill{}\subfloat[Summary tree]{\includegraphics[scale=0.6]{Figures__toy_pairwise_compat__s23}
+\subfloat[Summary tree]{\includegraphics[scale=0.57]{Figures__toy_pairwise_compat__s12}
+
+}\hfill{}\subfloat[Summary tree]{\includegraphics[scale=0.57]{Figures__toy_pairwise_compat__s13}
+
+}\hfill{}\subfloat[Summary tree]{\includegraphics[scale=0.57]{Figures__toy_pairwise_compat__s23}
 
@@
 \begin{figure}
-\subfloat[]{\includegraphics[scale=0.6]{Figures__uncontested_worsens_score__ab1c}
-
-}\hfill{}\subfloat[]{\includegraphics[scale=0.6]{Figures__uncontested_worsens_score__acb2}
-
-}\hfill{}\subfloat[]{\includegraphics[scale=0.6]{Figures__uncontested_worsens_score__tax}
+\subfloat[]{\includegraphics[scale=0.57]{Figures__uncontested_worsens_score__ab1c}
+
+}\hfill{}\subfloat[]{\includegraphics[scale=0.57]{Figures__uncontested_worsens_score__acb2}
+
+}\hfill{}\subfloat[]{\includegraphics[scale=0.57]{Figures__uncontested_worsens_score__tax}
 
@@
 
-\hfill{}\subfloat[]{\includegraphics[scale=0.6]{Figures__uncontested_worsens_score__opt}
-
-}\hfill{}\subfloat[]{\includegraphics[scale=0.6]{Figures__uncontested_worsens_score__returned}
+\hfill{}\subfloat[]{\includegraphics[scale=0.57]{Figures__uncontested_worsens_score__opt}
+
+}\hfill{}\subfloat[]{\includegraphics[scale=0.57]{Figures__uncontested_worsens_score__returned}
 
@@
 \begin{figure}
-\hfill{}\subfloat[Solution to subproblem \emph{ABCD}]{\includegraphics[scale=0.45]{Figures__subproblem_solutions__ott14\lyxdot tre}
-
-}\hfill{}\subfloat[Solution to subproblem \emph{root}]{\includegraphics[scale=0.45]{Figures__subproblem_solutions__ott17\lyxdot tre}
+\hfill{}\subfloat[Solution to subproblem \emph{ABCD}]{\includegraphics[scale=0.55]{Figures__subproblem_solutions__ott14\lyxdot tre}
+
+}\hfill{}\subfloat[Solution to subproblem \emph{root}]{\includegraphics[scale=0.55]{Figures__subproblem_solutions__ott17\lyxdot tre}
 
@@
 \begin{center}
-\includegraphics[scale=0.4]{example3__cleaned_ott-labels\lyxdot tre}
+\includegraphics[scale=0.38]{example3__cleaned_ott-labels\lyxdot tre}
 \par\end{center}%
@@
 \begin{center}
-\includegraphics[scale=0.4]{example3__grafted_solution-labels\lyxdot tre}
+\includegraphics[scale=0.38]{example3__grafted_solution-labels\lyxdot tre}
 \par\end{center}%
@@
 \subfloat[Unpruned tree - broken taxa removed]{\begin{centering}
-\includegraphics[scale=0.4]{example3__unprune-old-labels\lyxdot tre}
+\includegraphics[scale=0.38]{example3__unprune-old-labels\lyxdot tre}
 \par\end{centering}
 }\hfill{}\subfloat[Unpruned tree with MRCA-attachment]{\begin{centering}
-\includegraphics[scale=0.4]{example3__unprune-new-labels2\lyxdot tre}
+\includegraphics[scale=0.38]{example3__unprune-new-labels2\lyxdot tre}
 \par\end{centering}
@@
 \hfill{}\subfloat[JSON annotations relating edges of (c) to edges of (a)]{%
-\shadowbox{\begin{minipage}[t]{0.8\columnwidth}%
+\shadowbox{\begin{minipage}[t]{1.0\columnwidth}%
 \texttt{\{}
@@
 \end{figure}
-
-\paragraph{Annotations }
 
@@
 
-\section{\label{sec:The-maximum-sum}The maximum sum displayed groups' weighted
-scores criterion}
+\section{The maximum sum displayed groups' weighted scores criterion}\label{sec:The-maximum-sum}
 

### 2016-09-24 doc/paper1-submission/copy_figures.sh

@@ -4,15 +4,16 @@ FIGURES=$(grep '{.*/' paper.tex | sed 's|.*{\(.*/.*\)}.*|\1|'| sed 's|\\lyxdot |
 for figure in $FIGURES ; do
     filename="../${figure}"
     new_filename=$(echo $figure | sed 's|/|__|g')
-    new_filename=${new_filename%%.pdf}.pdf
     echo "\n\n'$figure' -> ${new_filename}"
     # If the lyx svg-to-pdf command is correctly set to do "--export-area-drawing", then regenerating pdfs is unnecessary.
+    if [ -f "${filename}" ] ; then
+	ln -s "$filename" "$new_filename"
+    fi
     if [ -f "${filename}.svg" ] ; then
-	filename="${filename}.svg"
-	inkscape "$filename" "--export-pdf=${new_filename}" --export-area-drawing
+	ln -s "${filename}.svg" "${new_filename}.svg"
+	inkscape "${new_filename}.svg" "--export-pdf=${new_filename}.pdf" --export-area-drawing
     else
-	filename="${filename%.pdf}.pdf"
-	cp "${filename}" "${new_filename}"
+	ln -s "${filename%.pdf}.pdf" "${new_filename%.pdf}.pdf"
     fi
 done
 

### 2016-09-26 doc/paper1-submission/paper2.tex

--- 2016-09-24 doc/paper1-submission/paper3.tex
@@
-\documentclass[fleqn,10pt,lineno,english]{wlpeerj}
+%% LyX 2.2.0 created this file.  For more info, see http:____www.lyx.org__.
+%% Do not edit unless you really know what you are doing.
+\documentclass[english]{paper}
 \usepackage[T1]{fontenc}
-%\usepackage[latin9]{inputenc}
-%\usepackage{geometry}
+\usepackage[latin9]{inputenc}
+\usepackage{geometry}
 \geometry{verbose,tmargin=2cm,bmargin=2cm,lmargin=2cm,rmargin=2cm}
-%\usepackage{xcolor}
-%\usepackage{verbatim}
+\usepackage{xcolor}
+\usepackage{verbatim}
 \usepackage{float}
@@
 \usepackage{graphicx}
-\usepackage{natbib}
+\usepackage[authoryear]{natbib}
 
@@
 \usepackage{babel}
+\begin{document}
 \global\long\def\taxonomy{\mbox{\ensuremath{\mathbb{T}}}}
@@
 
-\author[1,2]{Benjamin D. Redelings}
-\author[1]{Mark T. Holder}
-\affil[1]{Department of Ecology and Evolutionary Biology, University of Kansas, Lawrence KS, US}
-\affil[2]{Department of Biology, Duke University, Durham NC, US}
-\corrauthor[1]{Mark T. Holder}{mtholder@ku.edu}
-
-% \keywords{Keyword1, Keyword2, Keyword3}
-
+\author{Benjamin D. Redelings$^{1,2}$, Mark T. Holder$^{1,3}$}
+\maketitle
 \begin{abstract}
@@
 \end{abstract}
-
-\begin{document}
-
-\flushbottom
-\maketitle
-\thispagestyle{empty}
 
@@
 of 1357, or 49.2\%. %
+\begin{comment}
+ARE THESE numbers for the same set of inputs? or v4 to v5 (the latter
+had the barnacles tree collection, and perhaps a few other tree differences?
+\end{comment}
 
@@
 
-\section{Description of the decomposition algorithm of otc-uncontested-decompose}\label{sec:Decompose-algorithm}
+\section{\label{sec:Decompose-algorithm}Description of the decomposition
+algorithm of otc-uncontested-decompose}
 

### 2016-10-03 doc/paper1-submission/paper4.tex
Make abstract a bit more explicit.

--- 2016-09-24 doc/paper1-submission/paper3.tex
@@
-\documentclass[fleqn,12pt,english]{wlpeerj}
+\documentclass[fleqn,12pt,lineno,english]{wlpeerj}
 \usepackage[T1]{fontenc}
@@
 \affil[3]{Biodiversity Institute, University of Kansas, Lawrence KS, US}
-\affil[4]{Heidelberg Institute for Theoretical Studies, Schloss-Wolfsbrunnenweg 35, 69118 Heidelberg, Germany}
+\affil[4]{Heidelberg Institute for Theoretical Studies, Heidelberg, Germany}
 \corrauthor[2,3,4]{Mark T. Holder}{mtholder@ku.edu}
@@
 \begin{abstract}
-We present a new supertree method and software implementation that
-enables rapid estimation of a summary tree on the scale of millions
-of leaves. This software pipeline is called \emph{``propinquity.''}
-It uses the dependency-tracking tool \emph{make} to direct the series
-of operations required to construct the tree. It relies heavily on
-\emph{``otcetera''} - a set of C++ tools to performs most of the
-steps of the pipeline. All of the components are free software and
-are available on GitHub. This pipeline is currently used by the Open
-Tree of Life project to produce all of the versions of project's ``synthetic
-tree'' starting at version 5. The propinquity-based pipeline improves
-on previous supertree construction methods because it produces no
-unsupported branches, and it avoids unnecessary polytomies. In addition
-to producing a tree, the pipeline writes an annotations file that
-reports which grouping in the input trees support each of the groups
-in the supertree.
+We present a new supertree method that enables rapid estimation of a
+summary tree on the scale of millions of leaves.  This supertree
+summarizes a collection of input phylogenies and an input taxonomy.
+We introduce formal requirements that such a supertree must satisfy in
+order to transparently and justifiably represent the input trees.   In
+addition to producing a tree, our method writes an annotations file
+that reports which grouping in the input trees support each of the
+groups in the supertree.
+
+We compare the performance of this new method to a previous supertree
+method on the inputs to the Open Tree of Life version 4, and find that
+the number of displayed input splits increases from 35,518 to 39,639 and the
+number of conflicting input splits decreases from 2760 to 1357.  The
+new supertree method also improves on the previous supertree
+construction method because it produces no unsupported branches, and
+it avoids unnecessary polytomies.
+
+This pipeline is currently used by the Open Tree of Life project to
+produce all of the versions of project's ``synthetic tree'' starting
+at version 5.This software pipeline is called
+\emph{``propinquity.''}. It relies heavily on \emph{``otcetera''} - a
+set of C++ tools to perform most of the steps of the pipeline. All of
+the components are free software and are available on GitHub.
 \end{abstract}

### 2016-10-03 doc/paper1-submission/paper4.tex
Update results section slightly.

@@
 \usepackage{natbib}
-
+\sloppy
 \makeatletter
@@
 \newcommand{\lyxdot}{.}
+\newcommand{\otl}{OTLv4}
+\newcommand{\otlprop}{OTLv4/propinquity}
 
@@
 \usepackage{lmodern}
+\usepackage{xcolor}
+\usepackage{hyperref}
+\hypersetup{
+    colorlinks,
+    linkcolor={red!50!black},
+    citecolor={blue!50!black},
+    urlcolor={blue!80!black}
+}
 \definecolor{mylime}{RGB}{128,255,128}
@@
 summary tree on the scale of millions of leaves.  This supertree
-summarizes a collection of input phylogenies and an input taxonomy.
-We introduce formal requirements that such a supertree must satisfy in
-order to transparently and justifiably represent the input trees.   In
-addition to producing a tree, our method writes an annotations file
-that reports which grouping in the input trees support each of the
-groups in the supertree.
-
-We compare the performance of this new method to a previous supertree
-method on the inputs to the Open Tree of Life version 4, and find that
-the number of displayed input splits increases from 35,518 to 39,639 and the
-number of conflicting input splits decreases from 2760 to 1357.  The
-new supertree method also improves on the previous supertree
-construction method because it produces no unsupported branches, and
-it avoids unnecessary polytomies.
+method summarizes a collection of input phylogenies and an input taxonomy.
+We introduce formal goals and criteria for such a supertree to satisfy
+in order to transparently and justifiably represent the input
+trees. In addition to producing a supertree, our method computes
+annotations that describe which grouping in the input trees support
+and conflict with each group in the supertree.
+
+\medskip
+
+We compare our supertree construction method to a previously published
+supertree construction method by assessing their performance on input
+trees used to construct the Open Tree of Life version 4, and find that
+our method increases the number of displayed input splits from 35,518
+to 39,639 and decreases the number of conflicting input splits from
+2760 to 1357.  The new supertree method also improves on the previous
+supertree construction method in that it produces no unsupported
+branches and avoids unnecessary polytomies.
+
+\medskip
 
@@
 
-\section{Description of the propinquity pipeline}
+\section{Description of the supertree method}
 
@@
 down version of the taxonomy that only contains tips that are present
-in at least one phylogenetic input. In version 5.0 of the Open Tree
-of Life synthetic tree, this version of the taxonomic tree contained
-only 41,226 leaves (while the flag-cleaned version of OTT had 2,424,255
-leaves).
+in at least one phylogenetic input.
+
 \begin{figure}
@@
 as a means of improving the summary tree.
-
-In version 5.0 of the summary tree, this decomposition procedure produced
-5,545 subproblems, but only 1,586 of these were non-trivial to solve.
-If a subproblem contains only two tips it is trivial; 2,443 subproblems
-were trivial in this way. Similarly, if a subproblem contains only
-2 trees it is trivial to solve because the solution will simply be
-all of the groupings from the first tree combined with all of the
-groupings from the second tree that are compatible with the first
-tree; 2,838 subproblems were trivial in this way. The subproblem with
-the largest number of tips contained 953 tips. The largest subproblem,
-in terms of the number of input trees (including the taxonomic tree)
-that were relevant, had 18 trees. Without decomposition, the supertree
-problem would have had 512 input trees and 41,226 leaves.
 
@@
 
+\section{Results}
+
+\subsection{Inputs}
+
+We seek to assess the performance of our new supertree method by
+comparing it to the supertree method of \cite{HinchliffEtAl2015}.  The
+method of \cite{HinchliffEtAl2015} was used to construct the Open Tree
+of Life v4 (\otl{}).  Therefore, in order to facilitate comparison, we 
+applied our method to same input trees and taxonomy used by
+\otl{}.  We used the same cleaning flags to trim the taxonomy and
+input trees.  We refer to the resulting supertree as \otlprop{} since it was
+constructed by applying the propinquity pipeline to the same inputs as
+\otl{}.
+The flag-cleaned version of OTT contained 2,424,255 leaves.  The
+\otl{} supertree was constructed from 482 phylogenetic inputs,
+containing a total of 45385 leaves, of which 41029 were unique.
+After flag-cleaning and exemplification by propinquity, these trees contained 40323
+unique tips.
+
+\subsection{Subproblems}
+In the \otlprob{} summary tree, the decomposition procedure produced
+5,406 subproblems, but only 1,422 of these were non-trivial to solve.
+If a subproblem contains only two tips it is trivial; 2,362 subproblems
+were trivial in this way. Similarly, if a subproblem contains only
+2 trees it is trivial to solve because the solution will simply be
+all of the groupings from the first tree combined with all of the
+groupings from the second tree that are compatible with the first
+tree; 3,052 subproblems were trivial in this way. The subproblem with
+the largest number of tips contained 946 tips. The largest subproblem,
+in terms of the number of input trees (including the taxonomic tree)
+that were relevant, had 16 trees. Without decomposition, the supertree
+problem would have had 482 input trees and 41,226 leaves.
+
+\subsection{splits}
+We ran propinquity on the same input
+trees used to construct version Open Tree version 4. Unlike the version
+4 tree, the tree from propinquity contained no unsupported nodes or
+unnecessary polytomies. The number of displayed input splits increased
+from 35,518 (for version 4) to 39,639; an 11.2\% increase. The number
+of input splits that do not conflict with the summary tree, but are
+not incorporated dropped from 2718 to 0. The number of input splits
+that conflict with the phylogeny dropped from 2760 to 1357, a decrease
+of 1357, or 49.2\%. %
+
 \section{Conclusions}
@@
 input taxonomy changes). The amount of RAM required during tree construction
-has also decreased substantially. We ran propinquity on the same input
-trees used to construct version Open Tree version 4. Unlike the version
-4 tree, the tree from propinquity contained no unsupported nodes or
-unnecessary polytomies. The number of displayed input splits increased
-from 35,518 (for version 4) to 39,639; an 11.2\% increase. The number
-of input splits that do not conflict with the summary tree, but are
-not incorporated dropped from 2718 to 0. The number of input splits
-that conflict with the phylogeny dropped from 2760 to 1357, a decrease
-of 1357, or 49.2\%. %
+has also decreased substantially.
 

### 2016-10-03 doc/paper1-submission/paper4.tex
typo.

@@
 \subsection{Subproblems}
-In the \otlprob{} summary tree, the decomposition procedure produced
+In the \otlprop{} summary tree, the decomposition procedure produced
 5,406 subproblems, but only 1,422 of these were non-trivial to solve.

### 2016-10-10 doc/paper1-submission/paper4.tex
Update.

@@
 
-\subsection{Inputs}
-
 We seek to assess the performance of our new supertree method by
@@
 \otl{}.
+
+\subsection{Inputs}
+
 The flag-cleaned version of OTT contained 2,424,255 leaves.  The
@@
 
+% table
+
+% outputs - representation of input trees
+% outputs - annotations
+
+% conflicts in v4: 2760
+% conflicts in v4propr: 1357
+
+% displayed in v4: 35,518
+% displayed in v4propr: 39,639
+
+% # of input splits represented by v4, but not v4prop
+% # of input splits represented by v4prop, but not v4
+% # of input splits represented by v4 and v4prop
+
 \subsection{splits}
+
+\begin{tabular}{lcccccc}
+ & supported\_by & partial\_path\_of & terminal & conflicts\_with & resolves & resolved\_by\tabularnewline
+\hline 
+OTLv4 & 34,595 & 923 & 45,385 & 2760 & 473 & 2718\tabularnewline
+\hline 
+\textendash{} only & 745 & 54 & 0 & 2055 & 0 & 2718\tabularnewline
+\hline 
+OTLv4/propinquity & 38,521 & 1,118 & 45,385 & 1,357 & 515 & 0\tabularnewline
+\hline 
+\textendash{} only & 4,671 & 249 & 0 & 652 & 42 & 0\tabularnewline
+\hline 
+\end{tabular}
+
 We ran propinquity on the same input

### 2016-10-11 doc/paper1-submission/paper4.tex
Use non-numbered pipeline figure.
Refer to out tree as OTLv4' to save space.
Update and add second table for representing taxonomy.

@@
 \newcommand{\otl}{OTLv4}
-\newcommand{\otlprop}{OTLv4/propinquity}
+\newcommand{\otlprop}{OTLv4$^\prime$}
 
@@
 \begin{centering}
-\includegraphics[height=0.75\paperheight]{paper-pipeline-tools.pdf}
+\includegraphics[height=0.75\paperheight]{pipeline-tools.pdf}
 \par\end{centering}
@@
 of Life v4 (\otl{}).  Therefore, in order to facilitate comparison, we 
-applied our method to same input trees and taxonomy used by
-\otl{}.  We used the same cleaning flags to trim the taxonomy and
-input trees.  We refer to the resulting supertree as \otlprop{} since it was
+applied our method to the same input trees and taxonomy used by
+\otl{}.  
+We refer to the resulting supertree as \otlprop{} since it was
 constructed by applying the propinquity pipeline to the same inputs as
@@
 After flag-cleaning and exemplification by propinquity, these trees contained 40323
-unique tips.
+unique tips. We used the same cleaning flags to trim the taxonomy and
+input trees when constructing \otlprop{}, so \otlprop{} contains the
+same number of leaves.
 
@@
 % displayed in v4: 35,518
-% displayed in v4propr: 39,639
+% displayed in v4prop: 39,639
 
@@
 
-\subsection{splits}
-
+\subsection{Representing input splits}
+
+\begin{table}
+
+\caption{Representation of input splits in the \otl{} tree and the \otlprop{} tree.}
+\resizebox{\textwidth}{!}{%
 \begin{tabular}{lcccccc}
@@
 \hline 
-OTLv4 & 34,595 & 923 & 45,385 & 2760 & 473 & 2718\tabularnewline
+\otl{} & 34,595 & 923 & 45,385 & 2760 & 473 & 2718\tabularnewline
 \hline 
-\textendash{} only & 745 & 54 & 0 & 2055 & 0 & 2718\tabularnewline
+\otl{} only & 745 & 54 & 0 & 2055 & 0 & 2718\tabularnewline
 \hline 
-OTLv4/propinquity & 38,521 & 1,118 & 45,385 & 1,357 & 515 & 0\tabularnewline
+\otlprop{} & 38,521 & 1,118 & 45,385 & 1,357 & 515 & 0\tabularnewline
 \hline 
-\textendash{} only & 4,671 & 249 & 0 & 652 & 42 & 0\tabularnewline
+\otlprop{} only & 4,671 & 249 & 0 & 652 & 42 & 0\tabularnewline
 \hline 
-\end{tabular}
+\end{tabular}}
+
+\end{table}
+
+\begin{table}
+
+\caption{Representation of taxonomy splits in the \otl{} tree and the \otlprop{} tree.}
+\resizebox{\textwidth}{!}{%
+\begin{tabular}{lcccccc}
+ & supported\_by & partial\_path\_of & terminal & conflicts\_with & resolves & resolved\_by\tabularnewline
+\hline 
+\otl{} & 125,384 & 0 & 2,424,255 & 1998 & 3676 & 4\tabularnewline
+\hline 
+\otl{} only & 296 & 0 & 0 & 19 & 17 & 4\tabularnewline
+\hline 
+\otlprop{} & 125,107 & 0 & 2,424,255 & 2,279 & 3,883 & 0\tabularnewline
+\hline 
+\otlprop{} only & 19 & 0 & 0 & 300 & 224 & 0\tabularnewline
+\hline 
+\end{tabular}}
+
+\end{table}
+
+We performed an annotation of the \otl{} tree to compare the ability
+of the \otl{} tree and the \otlprop{} to represent splits in the input trees.
 

### 2016-10-11 doc/paper1-submission/paper4.tex
Update.

@@
 \newcommand{\otlprop}{OTLv4$^\prime$}
+\newcommand{\supportedby}{\texttt{supported\_by}}
+\newcommand{\partialpathof}{\texttt{partial\_path\_of}}
+\newcommand{\terminal}{\texttt{terminal}}
+\newcommand{\conflictswith}{\texttt{conflicts\_with}}
+\newcommand{\resolves}{\texttt{resolves}}
+\newcommand{\resolvedby}{\texttt{resolved\_by}}
 
@@
 
-\subsubsection{Annotation}
+\subsubsection{Annotation}\label{sec:Annotation}
 
@@
 tree (using the naming convention described in the previous section)
-and a node provenance object categorizes the relationship between
+and a node provenance object that categorizes the relationship between
 the node and the inputs. The node provenance object for node $x$
@@
 
-Conceptually, this annotation operation in equivalent to considering
+Conceptually, this annotation operation is equivalent to considering
 every node $j$ in each input tree $i$ and the summary tree node
@@
 with node $x$. All input nodes that cannot be displayed by any supertree
-that contains $x$ are stored in a ``conflicts\_with'' property
+that contains $x$ are stored in a ``\conflictswith{}'' property
 of the node provenance object. If node $j$ of $T_{i}$ is displayed
@@
 \begin{itemize}
-\item if $x$ is a terminal path in the induced tree then node $j$ will
-be listed in the ``terminal'' property;
+\item if $x$ is on a terminal path in the induced tree then node $j$ will
+be listed in the ``\terminal{}'' property;
 \item if $x$ is along an internal path that contains some nodes with out-degree
-equal to 1, then node $j$ will be listed in the ``partial\_path\_of''
+equal to 1, then node $j$ will be listed in the ``\partialpathof{}''
 property; and
 \item if $x$ is along an internal path without any node of out-degree 1,
-then node $j$ will be listed in the ``supported\_by'' property,
+then node $j$ will be listed in the ``\supportedby{}'' property,
 because node $j$ supports the existence of grouping $x$ in the sense
@@
 resolved such that the tree would then display $x$, then a reference
-to node $j$ will be listed in the ``resolves'' property of node
+to node $j$ will be listed in the ``\resolves{}'' property of node
 $x$.
@@
 $x$ in the summary tree; such a case would be annotated using the
-``resolved\_by'' property of $x$. However, because of our goal
+``\resolvedby{}'' property of $x$. However, because of our goal
 of excluding unnecessary polytomies, none of the nodes in propinquity's
@@
 
-\subsection{Representing input splits}
-
 \begin{table}
@@
 \begin{tabular}{lcccccc}
- & supported\_by & partial\_path\_of & terminal & conflicts\_with & resolves & resolved\_by\tabularnewline
+ & \supportedby{} & \partialpathof{} & \terminal{} & \conflictswith{} & \resolves{} & \resolvedby{}\tabularnewline
 \hline 
@@
 \hline
+\label{tab:representinputs}
 \end{tabular}}
@@
 \hline 
+\label{tab:representtaxonomy}
 \end{tabular}}
@@
 
-We performed an annotation of the \otl{} tree to compare the ability
-of the \otl{} tree and the \otlprop{} to represent splits in the input trees.
-
-We ran propinquity on the same input
-trees used to construct version Open Tree version 4. Unlike the version
-4 tree, the tree from propinquity contained no unsupported nodes or
-unnecessary polytomies. The number of displayed input splits increased
-from 35,518 (for version 4) to 39,639; an 11.2\% increase. The number
-of input splits that do not conflict with the summary tree, but are
-not incorporated dropped from 2718 to 0. The number of input splits
-that conflict with the phylogeny dropped from 2760 to 1357, a decrease
-of 1357, or 49.2\%. %
+\subsection{Representing input splits}
+
+We performed an annotation of both the \otl{} tree and the \otlprop{}
+tree as described in section \ref{sec:Annotation} to assess the
+ability of our new supertree method to represent splits from input
+phylogenies (Table \ref{tab:representinputs}).  This annotation
+classifies input phylogeny splits according to how they relate to a
+summmary tree, so that each input edge falls in one of \supportedby{},
+\partialpathof{}, \terminal{}, \conflictswith{}, or \resolvedby{}.
+
+The number of displayed input splits (supported\_by +
+partial\_path\_of) increased from 35,518 (for version 4) to 39,639; an
+11.2\% increase. The \otlprop{} tree displays 4920 input splits that
+are not displayed by the \otl{} tree, but does not display 799 input
+splits that are displayed by the \otl{} tree.
+
+Unlike the version 4 tree, the tree from propinquity contained no
+unsupported nodes or unnecessary polytomies. The number of input
+splits that do not conflict with the summary tree, but are not
+incorporated (resolved\_by) dropped from 2718 to 0.
+
+The number of input splits that conflict with the summary
+(conflicts\_with) dropped from 2760 to 1357, a decrease of 1357, or
+49.2\%.
 

### 2016-10-11 doc/paper1-submission/paper4.tex
Update.

@@
 \begin{tabular}{lcccccc}
- & \supportedby{} & \partialpathof{} & \terminal{} & \conflictswith{} & \resolves{} & \resolvedby{}\tabularnewline
+ & \supportedby{} & \partialpathof{} & \terminal{} & \conflictswith{} & \resolvedby{} & \resolves{}\tabularnewline
 \hline 
-\otl{} & 34,595 & 923 & 45,385 & 2760 & 473 & 2718\tabularnewline
+\otl{} & 34,595 & 923 & 45,385 & 2,760 & 2,718 & 473\tabularnewline
 \hline 
-\otl{} only & 745 & 54 & 0 & 2055 & 0 & 2718\tabularnewline
+\otl{} only & 745 & 54 & 0 & 2,055 & 2,718 & 0\tabularnewline
 \hline 
-\otlprop{} & 38,521 & 1,118 & 45,385 & 1,357 & 515 & 0\tabularnewline
+\otlprop{} & 38,521 & 1,118 & 45,385 & 1,357 & 0 & 515\tabularnewline
 \hline 
-\otlprop{} only & 4,671 & 249 & 0 & 652 & 42 & 0\tabularnewline
+\otlprop{} only & 4,671 & 249 & 0 & 652 & 0 & 42\tabularnewline
 \hline
@@
 \begin{tabular}{lcccccc}
- & \supportedby{} & \partialpathof{} & \terminal{} & \conflictswith{} & \resolves{} & \resolvedby{}\tabularnewline
+ & \supportedby{} & \partialpathof{} & \terminal{} & \conflictswith{} & \resolvedby{} & \resolves{}\tabularnewline
 \hline 
-\otl{} & 125,384 & 0 & 2,424,255 & 1998 & 3676 & 4\tabularnewline
+\otl{} & 125,384 & 0 & 2,424,255 & 1998 & 4 & 3676\tabularnewline
 \hline 
-\otl{} only & 296 & 0 & 0 & 19 & 17 & 4\tabularnewline
+\otl{} only & 296 & 0 & 0 & 19 & 4 & 17\tabularnewline
 \hline 
-\otlprop{} & 125,107 & 0 & 2,424,255 & 2,279 & 3,883 & 0\tabularnewline
+\otlprop{} & 125,107 & 0 & 2,424,255 & 2,279 & 0 & 3,883\tabularnewline
 \hline 
-\otlprop{} only & 19 & 0 & 0 & 300 & 224 & 0\tabularnewline
+\otlprop{} only & 19 & 0 & 0 & 300 & 0 & 224\tabularnewline
 \hline 
@@
 ability of our new supertree method to represent splits from input
-phylogenies (Table \ref{tab:representinputs}).  This annotation
-classifies input phylogeny splits according to how they relate to a
-summmary tree, so that each input edge falls in one of \supportedby{},
-\partialpathof{}, \terminal{}, \conflictswith{}, or \resolvedby{}.
-
-The number of displayed input splits (supported\_by +
-partial\_path\_of) increased from 35,518 (for version 4) to 39,639; an
-11.2\% increase. The \otlprop{} tree displays 4920 input splits that
-are not displayed by the \otl{} tree, but does not display 799 input
-splits that are displayed by the \otl{} tree.
-
-Unlike the version 4 tree, the tree from propinquity contained no
-unsupported nodes or unnecessary polytomies. The number of input
+phylogenies.  Table \ref{tab:representinputs} classifies the input
+phylogeny splits according to how they relate to a summmary tree, so
+that each input edge falls in one of \supportedby{}, \partialpathof{},
+\terminal{}, \conflictswith{}, or \resolvedby{}. For example, the
+numbers in the \conflictswith{} column indicate the number of input
+splits $j$ with at least one summary edge $x$ such that the relation
+``$x$ \conflictswith{} $j$'' holds.  The total number of non-terminal
+input phylogeny splits considered thus 40996.
+
+The number of displayed input splits (\supportedby{}
++ \partialpathof{}) increased from 35,518 (for \otl{}) to 39,639 (for \otlprop{});
+an 11.2\% increase. When examining which splits are displayed, we find
+that the \otlprop{} tree displays 4920 input splits that
+are not displayed by the \otl{} tree, whereas the \otl{} tree displays
+only 799 input splits that are not displayed by the \otlprop{} tree.  The number of input
+splits that conflict with the summary (conflicts\_with) dropped from
+2760 to 1357, a decrease of 1357, or 49.2\%.  In accordance with the
+goal of not containing unnecessary polytomies, the number of input
 splits that do not conflict with the summary tree, but are not
-incorporated (resolved\_by) dropped from 2718 to 0.
-
-The number of input splits that conflict with the summary
-(conflicts\_with) dropped from 2760 to 1357, a decrease of 1357, or
-49.2\%.
+incorporated (resolved\_by) dropped from 2718 to 0.  We also find that
+the number of polytomies in input phylogenies that are resolved by the
+summary tree increases from 473 for \otl{} to 515 for \otlprop{}.
+
+We also performed an annotation of the \otl{} tree and the \otlprop{}
+tree to assess the degree to which these trees represent taxonomy
+splits (Table \ref{tab:representtaxonomy}).  The \otlprop{} tree
+conflicts with 281 more taxonomy splits than the \otl{} tree.  Since
+the taxonomy is the lowest ranked input tree, this increased conflict
+with the taxonomy is an expected result of incorporating more splits
+from input phylogenies.
 
@@
 
-Here we have described the motivation and methodology used by the
-propinquity tool that is currently used by the Open Tree of Life project
-to build its summary tree. A modified version of the treemachine software
-which built the summary tree described in the \citet{HinchliffEtAl2015}
-paper is used by the project to serve the tree produced by propinquity
-via Open Tree of Life APIs. The migration of summary tree construction
-from treemachine (used for version 4) to propinquity (for all versions
-from v5.0 to present) has decreased the computational time required
-to construct a supertree from several hours to about 8 minutes (after
-some preprocessing steps which only have to be performed when the
-input taxonomy changes). The amount of RAM required during tree construction
-has also decreased substantially.
+Here we have described the motivation and methodology used by our new
+supertree method that is currently used by the Open Tree of Life
+project to build its summary tree.  Our new method represented 11\%
+more input phylogeny splits with 49\% less conflict compared to the
+Open Tree of Life version 4 summary tree when applied to the same inputs.
+Unless the previous method \citep{HinchliffEtAl2015}, our new method
+is guaranteed to incorporate input splits that do not conflict with
+the summary tree. The method is implemented in the Open Source
+software package propinquity.  A modified version of the treemachine
+software which built the summary tree described in the
+\citet{HinchliffEtAl2015} paper is used by the project to serve the
+tree produced by propinquity via Open Tree of Life APIs.
+
+The migration of summary tree construction from treemachine (used for
+version 4) to propinquity (for all versions from v5.0 to present) has
+increased the pace of synthesis tree releases from the Open Tree of
+Life project.  This is partly because the newly available annotations
+feature has made it possible to identify which input trees are
+responsible for taxa being included or excluded from the summary tree.
+Additionally, the new propinquity software pipeline has decreased the decreased
+the computational time required to construct a supertree from several
+hours to about 8 minutes (after some preprocessing steps which only
+have to be performed when the input taxonomy changes).  The amount of
+RAM required during tree construction has also decreased substantially.
 

### 2016-10-11 doc/paper1-submission/paper4.tex
Update.

@@
 
-The flag-cleaned version of OTT contained 2,424,255 leaves.  The
+The flag-cleaned version of OTT used in the construction of both
+supertrees contained 2,424,255 leaves.  The
 \otl{} supertree was constructed from 482 phylogenetic inputs,
@@
 unique tips. We used the same cleaning flags to trim the taxonomy and
-input trees when constructing \otlprop{}, so \otlprop{} contains the
+input trees when constructing \otlprop{}, so \otl{} and \otlprop{} contain the
 same number of leaves.
@@
 ``$x$ \conflictswith{} $j$'' holds.  The total number of non-terminal
-input phylogeny splits considered thus 40996.
+input phylogeny splits considered was 40996.
 
@@
 + \partialpathof{}) increased from 35,518 (for \otl{}) to 39,639 (for \otlprop{});
-an 11.2\% increase. When examining which splits are displayed, we find
+an 11\% increase. When examining which splits are displayed, we find
 that the \otlprop{} tree displays 4920 input splits that
@@
 only 799 input splits that are not displayed by the \otlprop{} tree.  The number of input
-splits that conflict with the summary (conflicts\_with) dropped from
-2760 to 1357, a decrease of 1357, or 49.2\%.  In accordance with the
+splits that conflict with the summary (\conflictswith{}) dropped from
+2760 to 1357, a decrease of 1403, or 51\%.  In accordance with the
 goal of not containing unnecessary polytomies, the number of input
 splits that do not conflict with the summary tree, but are not
-incorporated (resolved\_by) dropped from 2718 to 0.  We also find that
+incorporated (\resolvedby{}) dropped from 2718 to 0.  We also find that
 the number of polytomies in input phylogenies that are resolved by the
@@
 with the taxonomy is an expected result of incorporating more splits
-from input phylogenies.
+from higher-ranked input phylogenies.
 
@@
 supertree method that is currently used by the Open Tree of Life
-project to build its summary tree.  Our new method represented 11\%
-more input phylogeny splits with 49\% less conflict compared to the
-Open Tree of Life version 4 summary tree when applied to the same inputs.
-Unless the previous method \citep{HinchliffEtAl2015}, our new method
-is guaranteed to incorporate input splits that do not conflict with
-the summary tree. The method is implemented in the Open Source
-software package propinquity.  A modified version of the treemachine
-software which built the summary tree described in the
-\citet{HinchliffEtAl2015} paper is used by the project to serve the
-tree produced by propinquity via Open Tree of Life APIs.
+project to build summary supertrees on the scale of millions of leaves.
+Our new method represented 11\% more input phylogeny splits with 51\%
+less conflict compared to the Open Tree of Life version 4 summary
+tree, when applied to the same inputs. Unlike the previous method
+\citep{HinchliffEtAl2015}, our new method is guaranteed to incorporate
+input splits unless they conflict with the summary tree. The method is
+implemented in the Open Source software package propinquity.  A
+modified version of the treemachine software which built the summary
+tree described in the \citet{HinchliffEtAl2015} paper is used by the
+project to serve the tree produced by propinquity via Open Tree of
+Life APIs.
 

### 2016-10-12 doc/paper1-submission/paper4.tex
Incorporate Emily's comments.

--- 2016-10-11 doc/paper1-submission/paper4.tex
@@
 problem of finding the tree with the maximum sum of displayed groups'
-weighted scores criterion (described in the appendix A) where the
+weighted scores criterion (MSDGWS, described in the appendix A) where the
 weights from the trees are so extreme that displaying one group from
@@
 the remainder of the pipeline is to construct a tree that maximizes
-the sum displayed groups' weighted scores criterion. This is accomplished
+the sum of displayed groups' weighted scores criterion. This is accomplished
 in four steps: (1) dividing the full problem into subproblems based
@@
 approach to construct the supertree. Subproblems are identified by
-searching through the taxonomy tree to find any taxa are not contested
+searching through the taxonomy tree to find any taxa that are not contested
 by any single input tree. Here we say that input tree $T_{i}$ contests
@@
 shows the solutions that would be obtained by applying our modified
-version of the BUILD algorithm to the subproblems show in figure \ref{fig:subproblems-example}.
+version of the BUILD algorithm to the subproblems shown in figure \ref{fig:subproblems-example}.
 
@@
 should be attached to the grafted solution. One approach would be
-to mimic the application of apply Algorithm \ref{alg:ConsistentSplitsFromRankedList}
+to mimic the application of Algorithm \ref{alg:ConsistentSplitsFromRankedList}
 to the full (unpruned) taxonomic tree. This would be equivalent to
@@
 directly at taxon N, as in tree (c). In tree (d), the children of
-the broken taxon R and instead attached at the MRCA of R1, R2, and
-R3.}
+the broken taxon R are instead attached at the MRCA of R1, R2, and
+R3. Our method follows the second approach.}
 \end{figure}
@@
 are given the same identifier as the corresponding taxonomy node.
-These identifier are of the form $ottX$ where $X$ is an integer
+These identifiers are of the form $ottX$ where $X$ is an integer
 OTT id. We generate a label of the form $mrcaottX_{1}ottX_{2}$ for
@@
 
-The tool creates a multigraph by starting with a graph is isomorphic
+The tool creates a multigraph by starting with a graph isomorphic
 to the taxonomic tree. The nodes and edges created in this step will
@@
 one parent nodes in this set of exiting embedding edges, then tree
-$i$ contests that taxon. If there is only parent node, then the all
+$i$ contests that taxon. If there is only one parent node, then the all
 of the constituent taxa belonging to this taxon that are present in

### 2016-10-12 doc/paper1-submission/paper4.tex
final MTH edits

@@
 to 39,639 and decreases the number of conflicting input splits from
-2760 to 1357.  The new supertree method also improves on the previous
+2,760 to 1,357.  The new supertree method also improves on the previous
 supertree construction method in that it produces no unsupported
@@
 
-We can consider whether or not an node in an input tree is displayed
+We can consider whether or not a node in an input tree is displayed
 by $\summaryTree$. For any such node $j$ there is a set of taxa
@@
 $j$ if and only if that MRCA node is not an ancestor of any member
-of the exclude set of $j$. We say that edge $k$ of the summary tree
+of the exclude set of $j$. We say that node $k$ of the summary tree
 is \emph{supported by} node $j$ in $T_{i}$ if the summary tree displays
-node $j$, but if we contracted edge $k$ then the modified summary
+node $j$, but if we contracted edge that separates $k$ from its parent then the modified summary
 tree would no longer display node $j$.
@@
 Taxonomic groups may have the same name, but each node in the taxonomy
-tree is identified by a unique number (its OTT Id). Taxonomic groups
+tree is identified by a unique number (its OTT ID). Taxonomic groups
 are identified in the summary supertree by finding a branch (or ``node'')
@@
 of the polytomy shown in figure (\ref{fig:pairwisecompat}e) would
-continue to display the same two input groups, but the additional
+continue to display the same two input groups. However, the additional
 grouping would be unsupported, because the unresolved tree already
-displays the two input groups. Thus, the additional group would be
-unsupported and the unresolved tree would be preferred by our criterion.
+displays both input groups. Thus, the 
+unresolved tree would be preferred by our criterion.
 However, collapsing either internal edge of the tree shown in figure
@@
 for a taxon, then the taxon can be merged and the name will be retained
-without any descendants; this name will have an OTT Id, but will be
+without any descendants; this name will have an OTT ID, but will be
 flagged as ``barren''). Propinquity prunes the OTT down to a more
@@
 \texttt{config.opentree.synth} file in the propinquity repository).
-For the purpose of the rest of the pipeline, an OTT Id that has been
+For the purpose of the rest of the pipeline, an OTT ID that has been
 pruned from the taxonomy will be treated in the same way as invalid
-OTT Id. The output of this step is stored in propinquity's \texttt{cleaned\_ott}
+OTT ID. The output of this step is stored in propinquity's \texttt{cleaned\_ott}
 subdirectory; this operation only needs to be performed when the OTT
@@
 exemplar taxa as child nodes of the higher taxonomic tip but then
-collapsing the edge between the node that connected the higher taxon
-to its parent. Thus, if $A$ is a non-terminal taxon containing terminal
+collapsing the edge between the former tip node and
+its parent. Thus, if $A$ is a non-terminal taxon containing terminal
 descendants $a_{1}$ and $a_{2}$ and $B$ is a non-terminal taxon
@@
 the remainder of the pipeline is to construct a tree that maximizes
-the sum of displayed groups' weighted scores criterion. This is accomplished
+the sum of displayed groups' weighted scores (MSDGWS) criterion. This is accomplished
 in four steps: (1) dividing the full problem into subproblems based
@@
 case the groupings from each of the phylogenetic estimates, shown
-in panels \ref{fig:decompose-worsens}a and (\ref{fig:decompose-worsens}b,
+in panels \ref{fig:decompose-worsens}a and \ref{fig:decompose-worsens}b,
 could be displayed. That solution is shown in panel \ref{fig:decompose-worsens}d,
@@
 taxon R and instead attached at the MRCA of sampled children (R1,
-R2, and R3). Their attachment point does not correspond to a taxon
+R2, and R3). Their attachment point does not correspond to any taxon
 in the taxonomic tree.
@@
 These identifiers are of the form $ottX$ where $X$ is an integer
-OTT id. We generate a label of the form $mrcaottX_{1}ottX_{2}$ for
+OTT ID. We generate a label of the form $mrcaottX_{1}ottX_{2}$ for
 an non-taxonomic node $n$ where $X_{1}$ and $X_{2}$ are the OTT
-ids for two leaves, $n$ is the MRCA of these leaves, and $X_{1}$
+IDs for two leaves, $n$ is the MRCA of these leaves, and $X_{1}$
 is the numerically smallest OTT ID that is a descendant of $n$, and
@@
 $n$ as the MRCA. Because new taxa added to OTT will be given higher
-OTT Ids, the use of the lowest numbered OTT Ids as designators increases
+OTT IDs, the use of the lowest numbered OTT IDs as designators increases
 the chance that a node label can be encountered in a subsequent version
@@
 induced tree $\protect\summaryTree(1)$ are represented by JSON annotations
-(d). Taxon names are here suppressed in favor of OTT ids, and edges
+(d). Taxon names are here suppressed in favor of OTT IDs, and edges
 are referenced via their tipward nodes. Edges in $\protect\summaryTree(1)$
@@
 \otl{} supertree was constructed from 482 phylogenetic inputs,
-containing a total of 45385 leaves, of which 41029 were unique.
-After flag-cleaning and exemplification by propinquity, these trees contained 40323
+containing a total of 45,385 leaves, of which 41,029 were unique.
+After flag-cleaning and exemplification by propinquity, these trees contained 40,323
 unique tips. We used the same cleaning flags to trim the taxonomy and
@@
 
-% conflicts in v4: 2760
-% conflicts in v4propr: 1357
+% conflicts in v4: 2,760
+% conflicts in v4propr: 1,357
 
@@
 \hline 
-\otl{} & 125,384 & 0 & 2,424,255 & 1998 & 4 & 3676\tabularnewline
+\otl{} & 125,384 & 0 & 2,424,255 & 1,998 & 4 & 3,676\tabularnewline
 \hline 
@@
 ``$x$ \conflictswith{} $j$'' holds.  The total number of non-terminal
-input phylogeny splits considered thus 40996.
+input phylogeny splits considered was 40,996.
 
@@
 an 11.2\% increase. When examining which splits are displayed, we find
-that the \otlprop{} tree displays 4920 input splits that
+that the \otlprop{} tree displays 4,920 input splits that
 are not displayed by the \otl{} tree, whereas the \otl{} tree displays
@@
 splits that conflict with the summary (conflicts\_with) dropped from
-2760 to 1357, a decrease of 1357, or 49.2\%.  In accordance with the
+2,760 to 1,357, a decrease of 1,357, or 49.2\%.  In accordance with the
 goal of not containing unnecessary polytomies, the number of input
 splits that do not conflict with the summary tree, but are not
-incorporated (resolved\_by) dropped from 2718 to 0.  We also find that
+incorporated (resolved\_by) dropped from 2,718 to 0.  We also find that
 the number of polytomies in input phylogenies that are resolved by the
@@
 Open Tree of Life version 4 summary tree when applied to the same inputs.
-Unless the previous method \citep{HinchliffEtAl2015}, our new method
+Unlike the previous method \citep{HinchliffEtAl2015}, our new method
 is guaranteed to incorporate input splits that do not conflict with
@@
 
-Thanks to Karen Cranston, Jonathan Rees, Jim Allman, Cody Hinchliff,
+Thanks to Emily Jane McTavish, Karen Cranston, Jonathan Rees, Jim Allman, Cody Hinchliff,
 Stephen Smith, and Joseph Brown for discussions and feedback. Thanks

### 2016-10-13 doc/paper1-submission/paper4.tex

@@
 + \partialpathof{}) increased from 35,518 (for \otl{}) to 39,639 (for \otlprop{});
-an 11.2\% increase. When examining which splits are displayed, we find
+an 11\% increase. When examining which splits are displayed, we find
 that the \otlprop{} tree displays 4,920 input splits that
@@
 only 799 input splits that are not displayed by the \otlprop{} tree.  The number of input
-splits that conflict with the summary (conflicts\_with) dropped from
-2,760 to 1,357, a decrease of 1,357, or 49.2\%.  In accordance with the
+splits that conflict with the summary (\conflictswith{}) dropped from
+2,760 to 1,357, a decrease of 1,403, or 51\%.  In accordance with the
 goal of not containing unnecessary polytomies, the number of input
 splits that do not conflict with the summary tree, but are not
-incorporated (resolved\_by) dropped from 2,718 to 0.  We also find that
+incorporated (\resolvedby{}) dropped from 2,718 to 0.  We also find that
 the number of polytomies in input phylogenies that are resolved by the
@@
 supertree method that is currently used by the Open Tree of Life
-project to build its summary tree.  Our new method represented 11\%
-more input phylogeny splits with 49\% less conflict compared to the
-Open Tree of Life version 4 summary tree when applied to the same inputs.
-Unlike the previous method \citep{HinchliffEtAl2015}, our new method
-is guaranteed to incorporate input splits that do not conflict with
-the summary tree. The method is implemented in the Open Source
-software package propinquity.  A modified version of the treemachine
-software which built the summary tree described in the
-\citet{HinchliffEtAl2015} paper is used by the project to serve the
-tree produced by propinquity via Open Tree of Life APIs.
+project to build summary supertrees on the scale of millions of leaves.
+Our new method represented 11\% more input phylogeny splits with 51\%
+less conflict compared to the Open Tree of Life version 4 summary
+tree, when applied to the same inputs. Unlike the previous method
+\citep{HinchliffEtAl2015}, our new method is guaranteed to incorporate
+input splits unless they conflict with the summary tree. The method is
+implemented in the Open Source software package propinquity.  A
+modified version of the treemachine software which built the summary
+tree described in the \citet{HinchliffEtAl2015} paper is used by the
+project to serve the tree produced by propinquity via Open Tree of
+Life APIs.
 

### 2016-11-22 doc/paper1-submission/paper.tex

--- 2016-09-26 doc/paper1-submission/paper2.tex
@@
-%% LyX 2.2.0 created this file.  For more info, see http:____www.lyx.org__.
+%% LyX 2.2.0 created this file.  For more info, see http://www.lyx.org/.
 %% Do not edit unless you really know what you are doing.
@@
 \begin{figure}
-\subfloat[Input tree 1]{\includegraphics[width=0.45\textwidth]{Figures__toy_ambig__abcd}
-
-}\hfill{}\subfloat[Input tree 2]{\includegraphics[width=0.45\textwidth]{Figures__toy_ambig__bcde}
+\subfloat[Input tree 1]{\includegraphics[width=0.45\textwidth]{Figures/toy_ambig/abcd}
+
+}\hfill{}\subfloat[Input tree 2]{\includegraphics[width=0.45\textwidth]{Figures/toy_ambig/bcde}
 
@@
 
-\subfloat[Summary tree]{\includegraphics[width=0.45\textwidth]{Figures__toy_ambig__soln1}
-
-}\hfill{}\subfloat[Summary tree]{\includegraphics[width=0.45\textwidth]{Figures__toy_ambig__soln2}
+\subfloat[Summary tree]{\includegraphics[width=0.45\textwidth]{Figures/toy_ambig/soln1}
+
+}\hfill{}\subfloat[Summary tree]{\includegraphics[width=0.45\textwidth]{Figures/toy_ambig/soln2}
 
@@
 \begin{figure}
-\subfloat[Input tree 1]{\includegraphics[scale=0.6]{Figures__toy_pairwise_compat__abc}
-
-}\hfill{}\subfloat[Input tree 2]{\includegraphics[scale=0.6]{Figures__toy_pairwise_compat__acd}
-
-}\hfill{}\subfloat[Input tree 3]{\includegraphics[scale=0.6]{Figures__toy_pairwise_compat__bcd}
+\subfloat[Input tree 1]{\includegraphics[scale=0.6]{Figures/toy_pairwise_compat/abc}
+
+}\hfill{}\subfloat[Input tree 2]{\includegraphics[scale=0.6]{Figures/toy_pairwise_compat/acd}
+
+}\hfill{}\subfloat[Input tree 3]{\includegraphics[scale=0.6]{Figures/toy_pairwise_compat/bcd}
 
@@
 
-\subfloat[Summary tree]{\includegraphics[scale=0.6]{Figures__toy_pairwise_compat__s12}
-
-}\hfill{}\subfloat[Summary tree]{\includegraphics[scale=0.6]{Figures__toy_pairwise_compat__s13}
-
-}\hfill{}\subfloat[Summary tree]{\includegraphics[scale=0.6]{Figures__toy_pairwise_compat__s23}
+\subfloat[Summary tree]{\includegraphics[scale=0.6]{Figures/toy_pairwise_compat/s12}
+
+}\hfill{}\subfloat[Summary tree]{\includegraphics[scale=0.6]{Figures/toy_pairwise_compat/s13}
+
+}\hfill{}\subfloat[Summary tree]{\includegraphics[scale=0.6]{Figures/toy_pairwise_compat/s23}
 
@@
 \begin{centering}
-\includegraphics[height=0.75\paperheight]{paper-pipeline-tools.pdf}
+\includegraphics[height=0.75\paperheight,bb = 0 0 200 100, draft, type=eps]{paper-pipeline-tools.png}
 \par\end{centering}
@@
 \begin{figure}
-\subfloat[Tree 1]{\includegraphics[scale=0.4]{Figures__cleaned_phylo__ex_2@tree1\lyxdot tre}
-
-}\hfill{}\subfloat[Tree 2]{\includegraphics[scale=0.4]{Figures__cleaned_phylo__ex_2@tree2\lyxdot tre}
-
-}\hfill{}\subfloat[Taxonomy Tree]{\includegraphics[scale=0.4]{Figures__cleaned_ott__cleaned_ott\lyxdot tre}
+\subfloat[Tree 1]{\includegraphics[scale=0.4]{Figures/cleaned_phylo/ex_2@tree1\lyxdot tre}
+
+}\hfill{}\subfloat[Tree 2]{\includegraphics[scale=0.4]{Figures/cleaned_phylo/ex_2@tree2\lyxdot tre}
+
+}\hfill{}\subfloat[Taxonomy Tree]{\includegraphics[scale=0.4]{Figures/cleaned_ott/cleaned_ott\lyxdot tre}
 
@@
 \begin{figure}
-\subfloat[Tree 1]{\includegraphics[scale=0.4]{Figures__exemplified_phylo__ex_2@tree1\lyxdot tre}
-
-}\hfill{}\subfloat[Tree 2]{\includegraphics[scale=0.4]{Figures__exemplified_phylo__ex_2@tree2\lyxdot tre}
-
-}\hfill{}\subfloat[Pruned taxonomy tree $\protect\prunedTaxonomy$]{\includegraphics[scale=0.4]{Figures__exemplified_phylo_taxonomy_uncontested\lyxdot tre}
+\subfloat[Tree 1]{\includegraphics[scale=0.4]{Figures/exemplified_phylo/ex_2@tree1\lyxdot tre}
+
+}\hfill{}\subfloat[Tree 2]{\includegraphics[scale=0.4]{Figures/exemplified_phylo/ex_2@tree2\lyxdot tre}
+
+}\hfill{}\subfloat[Pruned taxonomy tree $\protect\prunedTaxonomy$]{\includegraphics[scale=0.4]{Figures/exemplified_phylo_taxonomy_uncontested\lyxdot tre}
 
@@
 \begin{figure}
-  \subfloat[Subproblem \emph{ABCD}]{\includegraphics[width=0.3\textwidth]{Figures__subproblems__14__1\lyxdot tre}%
-    \hfill{}\includegraphics[width=0.3\textwidth]{Figures__subproblems__14__2\lyxdot tre}%
-    \hfill{}\includegraphics[width=0.3\textwidth]{Figures__subproblems__14__3\lyxdot tre}
+  \subfloat[Subproblem \emph{ABCD}]{\includegraphics[width=0.3\textwidth]{Figures/subproblems/14/1\lyxdot tre}%
+    \hfill{}\includegraphics[width=0.3\textwidth]{Figures/subproblems/14/2\lyxdot tre}%
+    \hfill{}\includegraphics[width=0.3\textwidth]{Figures/subproblems/14/3\lyxdot tre}
 
@@
 
-  \subfloat[Subproblem \emph{root}]{\includegraphics[width=0.3\textwidth]{Figures__subproblems__17__1\lyxdot tre}%
-    \hfill{}\includegraphics[width=0.3\textwidth]{Figures__subproblems__17__2\lyxdot tre}%
-    \hfill{}\includegraphics[width=0.3\textwidth]{Figures__subproblems__17__3\lyxdot tre}
+  \subfloat[Subproblem \emph{root}]{\includegraphics[width=0.3\textwidth]{Figures/subproblems/17/1\lyxdot tre}%
+    \hfill{}\includegraphics[width=0.3\textwidth]{Figures/subproblems/17/2\lyxdot tre}%
+    \hfill{}\includegraphics[width=0.3\textwidth]{Figures/subproblems/17/3\lyxdot tre}
 
@@
 \begin{figure}
-\subfloat[]{\includegraphics[scale=0.6]{Figures__uncontested_worsens_score__ab1c}
-
-}\hfill{}\subfloat[]{\includegraphics[scale=0.6]{Figures__uncontested_worsens_score__acb2}
-
-}\hfill{}\subfloat[]{\includegraphics[scale=0.6]{Figures__uncontested_worsens_score__tax}
+\subfloat[]{\includegraphics[scale=0.6]{Figures/uncontested_worsens_score/ab1c}
+
+}\hfill{}\subfloat[]{\includegraphics[scale=0.6]{Figures/uncontested_worsens_score/acb2}
+
+}\hfill{}\subfloat[]{\includegraphics[scale=0.6]{Figures/uncontested_worsens_score/tax}
 
@@
 
-\hfill{}\subfloat[]{\includegraphics[scale=0.6]{Figures__uncontested_worsens_score__opt}
-
-}\hfill{}\subfloat[]{\includegraphics[scale=0.6]{Figures__uncontested_worsens_score__returned}
+\hfill{}\subfloat[]{\includegraphics[scale=0.6]{Figures/uncontested_worsens_score/opt}
+
+}\hfill{}\subfloat[]{\includegraphics[scale=0.6]{Figures/uncontested_worsens_score/returned}
 
@@
 \begin{figure}
-\hfill{}\subfloat[Solution to subproblem \emph{ABCD}]{\includegraphics[scale=0.45]{Figures__subproblem_solutions__ott14\lyxdot tre}
-
-}\hfill{}\subfloat[Solution to subproblem \emph{root}]{\includegraphics[scale=0.45]{Figures__subproblem_solutions__ott17\lyxdot tre}
+\hfill{}\subfloat[Solution to subproblem \emph{ABCD}]{\includegraphics[scale=0.45]{Figures/subproblem_solutions/ott14\lyxdot tre}
+
+}\hfill{}\subfloat[Solution to subproblem \emph{root}]{\includegraphics[scale=0.45]{Figures/subproblem_solutions/ott17\lyxdot tre}
 
@@
 \begin{centering}
-\includegraphics[scale=0.5]{Figures__grafted_solution__grafted_solution_ottnames\lyxdot tre}
+\includegraphics[scale=0.5]{Figures/grafted_solution/grafted_solution_ottnames\lyxdot tre}
 \par\end{centering}
@@
 \begin{center}
-\includegraphics[scale=0.4]{example3__cleaned_ott-labels\lyxdot tre}
+\includegraphics[scale=0.4]{example3/cleaned_ott-labels\lyxdot tre}
 \par\end{center}%
@@
 \begin{center}
-\includegraphics[scale=0.4]{example3__grafted_solution-labels\lyxdot tre}
+\includegraphics[scale=0.4]{example3/grafted_solution-labels\lyxdot tre}
 \par\end{center}%
@@
 \subfloat[Unpruned tree - broken taxa removed]{\begin{centering}
-\includegraphics[scale=0.4]{example3__unprune-old-labels\lyxdot tre}
+\includegraphics[scale=0.4]{example3/unprune-old-labels\lyxdot tre}
 \par\end{centering}
 }\hfill{}\subfloat[Unpruned tree with MRCA-attachment]{\begin{centering}
-\includegraphics[scale=0.4]{example3__unprune-new-labels2\lyxdot tre}
+\includegraphics[scale=0.4]{example3/unprune-new-labels2\lyxdot tre}
 \par\end{centering}
@@
 \begin{centering}
-\includegraphics[scale=0.5]{Figures__labelled_supertree__labelled_supertree_only_taxon_names\lyxdot tre}
+\includegraphics[scale=0.5]{Figures/labelled_supertree/labelled_supertree_only_taxon_names\lyxdot tre}
 \par\end{centering}
@@
 \begin{minipage}[c][1\totalheight][t]{0.25\columnwidth}%
-\includegraphics[scale=0.5]{annotation_example__input1\lyxdot tre}%
+\includegraphics[scale=0.5]{annotation_example/input1\lyxdot tre}%
 \end{minipage}
@@
 \begin{minipage}[c][1\totalheight][t]{0.35\columnwidth}%
-\includegraphics[scale=0.5]{annotation_example__summary\lyxdot tre}%
+\includegraphics[scale=0.5]{annotation_example/summary\lyxdot tre}%
 \end{minipage}
@@
 \begin{minipage}[c][1\totalheight][t]{0.35\columnwidth}%
-\includegraphics[scale=0.5]{annotation_example__induced1_color}%
+\includegraphics[scale=0.5]{annotation_example/induced1_color}%
 \end{minipage}

### 2016-11-23 bin/compare_synthesis_outputs.py
minor documentation changes

@@ -4,8 +4,37 @@
 
 import simplejson as json
 import argparse
-import glob,os,re
+import glob,os,re,csv
 import requests
+from peyotl import ott
+
+def broken_taxa_diffs(run1,run2,verbose,names=False):
+# file of interest is labelled_supertree/broken_taxa.json
+    jsonfile = "{d}/labelled_supertree/broken_taxa.json".format(d=run1)
+    lsbt1 = json.load(open(jsonfile, 'r'))['non_monophyletic_taxa']
+    broken_taxa1 = lsbt1.keys() if lsbt1 else set()
+    jsonfile = "{d}/labelled_supertree/broken_taxa.json".format(d=run2)
+    lsbt2 = json.load(open(jsonfile, 'r'))['non_monophyletic_taxa']
+    broken_taxa2 = lsbt2.keys() if lsbt2 else set()
+    compare_lists("Broken taxa",broken_taxa1,broken_taxa2,verbose)
+    if (names):
+        broken_taxa_report(set(broken_taxa1),set(broken_taxa2))
+
+# writes details of the broken taxa to a file that can be input by
+# report_on_broken_taxa.py
+def broken_taxa_report(s1,s2):
+    # print details of names in 2 but not in 1 (the 'newly broken names')
+    diff = s2.difference(s1)
+    broken_taxa_filename = 'broken_taxa_report.csv'
+    print "printing details of {x} broken taxa to {f}".format(
+        x=len(diff),
+        f=broken_taxa_filename
+        )
+    with open(broken_taxa_filename, 'w') as f:
+        for ottID in diff:
+            (name,rank)=get_taxon_details(ottID)
+            f.write("{n}_{i}\n".format(n=name,i=ottID))
+    f.close()
 
 # generic function to compare two lists: number of items in each,
 # items in first but not second and items in second but not first
@@ -25,13 +54,13 @@ def compare_lists(type,list1,list2,verbose=False):
             print " in common = {l}".format(t=type,l=len(common))
         diff1 = s1.difference(s2)
         if len(diff1) > 0:
-            print " in run1 but not run2 = {l}".format(t=type,l=len(diff1))
+            print " in run1 but not run2 = {l}".format(l=len(diff1))
             if (verbose):
                 print " {t} in run1 but not run2:".format(t=type)
                 print ', '.join(diff1)
         diff2 = s2.difference(s1)
         if len(diff2) > 0:
[…]

### 2016-11-23 bin/report_on_broken_taxa.py

@@ -1,89 +1,124 @@
 #!/usr/bin/env python
+
+# creates an html broken taxon report that can be used for diagnosing
+# which input trees break which taxa
+
+# input file can be created with:
+# grep -E -o  "[)]['A-Za-z][^,)(]+ott[0-9]+" cleaned_ott/cleaned_ott.tre  | sed -E 's/.*ott//'
+
 from cStringIO import StringIO
 import codecs
-import json
-import sys
-import os
-import re
-taxon_pat = re.compile(r'^(.*)([ _]ott)(\d+)\s*$')
-try:
-    id_list_file, out_dir = sys.argv[1:]
-except:
-    sys.exit('Expecting a file with a list of Names + ott### (encoding OTT Ids) as the first argument and a propinquity OUTPUT dir as the second argument.\n')
-btf = os.path.join(out_dir, "labelled_supertree", "broken_taxa.json")
-if not os.path.isfile(btf):
-    sys.exit('The file "{}" was not found\n'.format(btf))
-with codecs.open(btf, 'r', encoding='utf-8') as inp:
-    broken = json.load(inp)['non_monophyletic_taxa']
-ctf = os.path.join(out_dir, "subproblems", "contesting-trees.json")
-if not os.path.isfile(ctf):
-    sys.exit('The file "{}" was not found\n'.format(ctf))
-with codecs.open(ctf, 'r', encoding='utf-8') as inp:
-    contesting = json.load(inp)
+import argparse
+import json, sys, os, re
 
+def read_broken_taxa(out_dir):
+    # check that the broken taxa file exists in the output directory; read
+    # if present
+    btf = os.path.join(out_dir, "labelled_supertree", "broken_taxa.json")
+    if not os.path.isfile(btf):
+        sys.exit('The file "{}" was not found\n'.format(btf))
+    with codecs.open(btf, 'r', encoding='utf-8') as inp:
+        data = json.load(inp)['non_monophyletic_taxa']
+    return data
 
-pref = 'https://devtree.opentreeoflife.org/opentree/argus/opentree7.2'
-hels = []
-with open(id_list_file, 'r') as inp:
-    for line in inp:
-        ls = line.strip()
-        if not ls:
-            continue
-        m = taxon_pat.match(ls)
-        if not m:
-            sys.exit('ID file entry "{}" does not end with ott[0-9]+\n'.format(ls))
-        ott_id = m.group(3)
-        name = m.group(1)
[…]

### 2017-01-10 doc/paper1-submission/paper5.tex
Alter terminology paragraph vis a viz Luay Nakleh.

--- 2016-10-13 doc/paper1-submission/paper4.tex
@@
 
+Cutting any edge \textbf{$j$ }of a rooted tree induces a bi-partition
+$S(j)$ of the leaf taxa into two connected groups $S_{1}(j)$ and
+$S_{2}(j)$. Such a bi-partition is called a rooted split and written
+$S(j)=S_{1}(j)|\bullet S_{2}(j)$. Here $S_{1}(j)$ is called the ``cluster'' of $j$, and contains tip taxa on the side of edge $j$ that does not contain the
+root. $S_{2}(j)$ contains the tip taxa on the side of edge $j$ that
+contains the root. We refer to $S(j)$ as a \emph{rooted} split because
+the right side of the split implicitly contains the root node, as indicated
+by the $\bullet$.
+
+For any two rooted splits $A=A_{1}|\bullet A_{2}$ and $B=B_{1}|\bullet B_{2}$,
+we say that $A$ \emph{displays} $B$ if $B_{1}\subseteq A_{1}$ and
+$B_{2}\subseteq A_{2}$. We also say that $A$ and $B$ \emph{conflict}
+if none of $A_{1}\cap B_{1}$, $A_{1}\cap B_{2}$, and $B_{2}\cap A_{1}$
+are empty\footnote{The intersection of the right sides of the splits is considered to
+be non-empty even if $A_{2}\cap B_{2}$ is empty, because the right
+sides implicitly contain the root.}. If $A$ and $B$ do not conflict, then we say that they are \emph{compatible}\footnote{It is possible to construct a rooted tree that displays both $A$
+and $B$ if and only if $A$ and $B$ are compatible}. If $A$ displays $B$, then $A$ and $B$ must be compatible, since
+$A_{1}\cap B_{2}$ and $B_{1}\cap A_{2}$ must be empty. We adopt
+the shorthand of referring to an edge $k$ when we mean the split
+$S(k)$. So, when we say that edge $j$ conflicts with edge $k$,
+we mean that $S(j)$ conflicts with $S(k)$.
+
 Let $\summaryTree$ denote a supertree, and $T_{i}$ denote the $i$th
-input tree. The set of taxa that are mapped to the tips of the tree
-$T_{i}$ is $\mathcal{L}(i)$. $\summaryTree(i)$ denotes the summary
-tree induced by tip nodes that are mapped to taxa in $\mathcal{L}(i)$
-and the most recent common ancestor of those leaves, and any other
-node that is an ancestor of some but not all of these leaves. We say
-that edge $j$ of the supertree is compatible with an input tree,
-$T_{i}$ if edge $j$ either is not included in the induced tree $\summaryTree(i)$
-or none of the edges in $T_{i}$ are in conflict with edge $j$ in
-the induced tree.
-
-We can consider whether or not a node in an input tree is displayed
-by $\summaryTree$. For any such node $j$ there is a set of taxa
-that are mapped to the tips that descend from the node. This set of
-taxa is the ``cluster'' of taxa corresponding to node $j$; it can
-be denoted $\mathcal{L}(i,j)$. It will also be referred to as the
-``include set'' of the node. The ``exclude set'' of $j$ in $T_{i}$
-is the set of taxa in $\mathcal{L}(i)$ but not in $\mathcal{L}(i,j)$.
-If the cluster of taxa for any supertree node in $\summaryTree(i)$
-is identical to $\mathcal{L}(i,j)$, then we say that $\summaryTree$
-displays node $j$ of $T_{i}$. We say that the summary tree \emph{displays}
-edge $j$ if the summary tree displays the child node of edge $j$.
-Operationally, we can find the most recent common ancestor (MRCA)
-node of $\mathcal{L}(i,j)$ in $\summaryTree$; the summary tree displays
-$j$ if and only if that MRCA node is not an ancestor of any member
-of the exclude set of $j$. We say that node $k$ of the summary tree
-is \emph{supported by} node $j$ in $T_{i}$ if the summary tree displays
-node $j$, but if we contracted edge that separates $k$ from its parent then the modified summary
-tree would no longer display node $j$.
+input tree. We say that $\summaryTree$ displays an edge $j$ of $T_{i}$
+if any edge of $\summaryTree$ displays $j$. We say that $\summaryTree$
+conflicts with a edge $j$ of $T_{i}$ if any edge of $\summaryTree$
+conflicts with $j$; otherwise $\summaryTree$ is compatible with
+$j$. We say that edge $k$ of $\summaryTree$ is \emph{supported
+by} edge $j$ of $T_{i}$ if $\summaryTree$ displays $j$, but $\summaryTree$
+would not display $j$ if we contracted edge $k$. We also adopt the
+shorthand of referring to edges by their tipward endpoint. By combining
+this with the above shorthand, we could say that a node $j$ conflicts
+or does not conflict with another node $k$.
+
+The set of taxa that are mapped to the tips of the tree $T_{i}$ is
+$\mathcal{L}(i)$. $\summaryTree(i)$ denotes the summary tree induced
+by tip nodes that are mapped to taxa in $\mathcal{L}(i)$ and the
+most recent common ancestor of those leaves, and any other node that
+is an ancestor of some but not all of these leaves.
 
@@
 \begin{centering}
-\includegraphics[height=0.75\paperheight]{pipeline-tools.pdf}
+\includegraphics[height=0.70\paperheight]{pipeline-tools.pdf}
 \par\end{centering}

### 2017-01-12 doc/paper1-submission/response.tex
Update.

\documentclass{article}
\usepackage[margin=1in]{geometry}
\usepackage{titlesec}
%\usepackage{a4wide}
%\usepackage{parskip}
\setlength{\parindent}{0pt}
\setlength{\parskip}{0.5em} % 1ex plus 0.5ex minus 0.2ex}
\title{Response to the Reviewer's Comments}
\author{Benjamin Redelings \and Mark Holder}
\titlespacing*{\section}{0pt}{6pt}{0pt}
\titlespacing*{\subsection}{0pt}{3pt}{0pt}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Commands for changes...
\usepackage{color}
%\usepackage{ulem}

\newcommand{\benadd}[1]{#1}
\newcommand{\benremove}[1]{}
\newcommand{\benreplace}[2]{\benadd{#2}}

% \newcommand{\benreplace}[2]{\benremove{#1} \benadd{#2}}

\newenvironment{reply}{$\triangleright$\bf}{$\triangleleft$}
\renewenvironment{quote}
               {\list{}{\rightmargin\leftmargin}%
                \item\relax\normalfont}
               {\endlist}

\begin{document}

\maketitle

Dear Dr. XXX,

We thank the reviewers for their comments.  Here is our response.

Sincerely,

Ben Redelings\\
Mark Holder

\clearpage

\section*{Reviewer 1 (Anonymous)}
\subsection*{Basic reporting}

The basic reporting is sufficient. I have detailed suggestions in my general comments to the authors.

\subsection*{Experimental design}

The analytical design is appropriate for the study.

\subsection*{Validity of the findings}

The findings are valid. One challenge I realized when it comes to reviewing a large-scale supertree effort of this type is the simple fact that the dependence of the method on the quality of the source trees renders empirical evaluation of the results challenging. I provide details in the general comments.

\subsection*{Comments for the Author}
The manuscript ``A supertree pipeline for summarizing phylogenetic and taxonomic information for millions of species'' by Redelings and Holder (hereafter R\&H) describes a novel supertree approach that is already being used to generate the synthetic tree for opentreeoflife.org (hereafter “opentree”). Opentree has the potential to be a very important community resource. Therefore, it seems both important and appropriate to publish a description of the pipeline used to generate the synthetic tree for that project (unless there is a fundamental problem with the proposed approach).

I think there are three issues that should be addressed in a paper like this: 1) does the proposed supertree algorithm have any fundamental flaws?; 2) can the method address the unequal quality of source trees?; and 3) will large-scale synthetic trees generated using this method be useful to the biological community? The third issue is important for a resource like opentree, where a large community might wish to have a single place to retrieve a tree that can be used for comparative methods and/or ancestral state reconstruction. R\&H lay out a relatively convincing case for the first two; they do not address the third.

I do not believe that there is no such problem with the approach proposed by R\&H. They clearly lay out their desiderata for a “good” supertree method. The most important (in my opinion) is described by their first goal: that “each grouping is supported by at least one input [tree]”. This avoids the problem of completely “emergent” groups, which can occur in some supertree methods (Gatesy and Springer 2004). With supermatrix methods it is reasonable (indeed, it is desirable) to find clades that are absent from the estimates of phylogeny for each individual partition. In contrast, the emergence of clades that are not present in any individual source tree is undesirable for supertree methods. The method proposed by R\&H appears to be designed to avoid this problem.

\begin{reply}
1. It seems that the reviewer is satisfied with our criteria for the supertree method.
\end{reply}

The other challenge for supertrees is the low quality of some source trees (this may be a larger problem for older trees). If the tree quality issue is not addressed supertrees are likely to converge on “traditional relationships,” since older trees supporting those relationships may be more numerous and taxonomies often reflect those traditional relationships. This tendency is evident if we focus on a well-studied group (mammals) where an older supertree analyses (e.g., Liu et al. 2001) exhibited this problem (Springer \& de Jong 2001; Gatesy et al. 2002). The problem is even more evident in similar studies of birds, where a supertree effort (Davis and Page 2014) recovered many “traditional” relationships for major groups that were not found in large-scale avian trees (Hackett et al. 2008; Jarvis et al. 2014; Prum et al. 2015). The “newer” molecular relationships were also recovered in a large-scale supermatrix (Burleigh et al. 2015) that provides a large-scale tree with many of the benefits of large-scale supertree studies without the drawbacks

The issue of including data from “bad” trees is especially acute when taxonomies and/or “conglomerated trees” (i.e., trees based on expert evaluation of the literature that largely reflect the consensus at that time; cf. Gatesy et al. 2002). R\&H’s use of weighted trees should address this issue, at least to some degree, since the method downweights specific trees. Hopefully, the downweighted tree will correspond to less reliable sources of phylogenetic information.

I have two concerns regarding the use of tree weights. First, it seems clear that weights could have an extremely large impact in the R\&H method. According to part A of their appendix:

“The summary tree constructed by the propinquity pipeline is a greedy heuristic for finding a tree that maximizes this score when the weights for a node are determined by the tree’s weight and the difference in weighting is so large that displaying one node from a highly ranked tree is preferred to displaying all of the nodes in the trees with lower rank.”

Some elaboration on the sensitivity of their method to tree weights seems warranted. Moreover, the weights are appear to be applied for trees rather than nodes. I recognize that weighting nodes is likely to be impractical, but it could create problems if different source trees have different likelihoods of correctly resolving nodes at various depths in the tree, as seems likely to be the case (e.g., Chen et al. 2015). Perhaps the authors could elaborate on this issue.

Second, I am not sure that weights for trees can be obtained in a truly principled manner. I recognize that one could use any weighting scheme with the R\&H method but it would seem valuable to have an empirical example where a specific weighting scheme is used. Obviously, the weights used in opentree would be the most appropriate to examine. To evaluate those weights I found myself looking to Hinchliff et al. (2015) paper and the current tree on the opentree website. Overall, it would be good for readers to have as much information as possible about the weighting used by Hinchliff et al. (2015); I found the description of the weighting used by Hinchliff et al. (2015) difficult to understand. It would be useful for R\&H to provide as much additional information about Hinchliff et al. (2015) weights as possible, probably in the discussion.

With these concerns out of the way I will add that I am uncertain whether a supertree method that achieves “perfect” (or nearly perfect) weighting can be devised. The only method I can think of that might allow principled estimation of tree weights would be a version of the maximum likelihood supertree method described by Steel and Rodrigo (2008). However, I doubt that such an ML supertree method would meet all of the desiderata for a good supertree method that R\&H articulate (especially their first goal). It would be valuable for the authors to expand on the potential advantages (which appear to be largely computational) and pitfalls of their extreme weighting method.

\begin{reply}
2a. Concern: incorrect input trees could lead to supertree that is not correct.  Response: our goal is to summarize the input trees.  This would work just fine.

2b. Concern: hypotheses that occur in more input trees might not be correct, so a supertree method should not hypothesis just based on frequency.  Response: our method does not support hypotheses because of their frequency.

2c. Concern: bad input trees could compromise the correctness of inferred phylogeny.  Downweighting bad input trees might assuage this.  Response: we are not performing a phylogeny inference.  We don't directly use weights, but instead of community-curated ranks.  While it is possible to construct weights so that we can phrase our rank-based procedure as an optimization problem, at no point do we ever construct weights.

2d. Concern: the weights have a large impact on the method.  Please describe how sensitive the method is to the weights.  Response: we indeed hope that the ranking has a substantial effect on the resulting tree.  If this were not the case, then ranks could not achieve their stated goal of allowing users to use ranks to fix problems in the tree.

Despite this, we ran the pipeline with the ranks reversed.  This drastic alteration of the ranks lead to a supertree that conflicted with 1264 branches of the original supertree.  Compared to this, the opentree 4.0 summary tree that used preceding methodology conflicts with 1387 branches of the 4.0’ tree using current methods.  However, most of the 36533 internal edges of the grafted tree are not altered.

2e. Concern: the weights are applied to trees rather than nodes.  This seems to be the only valid concern.

2f. Concern: the weights cannot be derived in a principled manner.  Results: we are not trying to estimate the weights from the input trees.  We would only need to estimate weights if we were trying to infer the true phylogeny from the input trees, a la the supertree versus supermatrix debate.  But we are not, therefore there is not need to derive weights.

2.  It seems that the reviewer thinks that we are trying to infer the true tree from a datasets consisting of the input trees.  Such an inference would need a principled method of resolving conflicts between input trees using only the information in the set of input trees.  The resulting supertree could then be used to critique the input trees, so that input trees that don't match the supertree would be considered to be incorrect (up to estimation error).

However, our approach is fundamentally different.  We do not seek to infer a tree that is more accurate than any individual input tree.  Instead we seek to summarize and describe the set of input trees.  While the supertree must conflict with some input trees, we do not claim that this constitutes an estimate of which input trees are correct or incorrect.  Instead we seek to represent all input trees, including those that conflict with the phylogeny estimate by linking from summary tree edges to conflicting input tree edges via our annotations file.

Our approach does involve the resolution of conflict.  However, our conflict resolution is not phrased as an estimate of the true tree.  Instead, we supply rankings to allow curators to selection which trees should win conflicts.

We describe how our ranking method could be considered as a weight-based method for the purposes of optimization.  However, such weights should not be interpreted as an attempt to using the (weighted) input trees to decide which alternative phylogenetic hypotheses are correct, and in fact we never say anything like this. Our weights are not computed from the input trees, because we are not performing phylogeny estimation.

The fact that rankings apply to entire trees is indeed an issue with the method.

% Summary: argh.

\end{reply}

The authors do not address my third concern. The degree to which opentree will be used as a source of information for comparative methods is unclear to me. However, regardless of whether it is commonly used for that purpose there is an even more fundamental question: should opentree in its current releases be used by the for comparative methods by the community? I recognize that the R\&H methods could be used outside of opentree, but I think some readers may want an assessment of whether the tree is “ready for prime time.” Other than the broad coverage, is opentree as good as (or better than) than available synthetic trees (e.g., Bininda-Emonds et al. 2007 or Jetz et al. 2012)? I can think of at least one expression of concern regarding the use of taxonomies: the biogeographic study of Wang et al. (2016) that showed a major change in their conclusions when a taxon that appears to be misplaced by taxonomic constraints is included or excluded. The low weight of the taxonomy in opentree might ameliorate this issue. Obviously, addressing assessing this issue might be challenging (e.g., running the opentree analysis without the taxonomy and comparing that tree to a version of opentree pruned to exclude taxa that are only represented in the taxonomy). Although potentially useful I am not certain whether it is warranted.

\begin{reply}
3.

While we believe that our tree is useful, we are do not claim that it is more accurate that the phylo-inputs that are used to construct it.  While we hope to summarize currently available phylogenetic knowledge, we are not trying to produce such knowledge like Bininda-Edmonds.

The tree cannot be used directly for comparative methods, because it does not have branch lengths.
\end{reply}

To summarize, I think the method is extremely valuable. Elaborating a bit more in the discussion, even if that elaboration would be somewhat speculative, would be valuable.

\subsection*{References:}

Bininda-Emonds OR, Cardillo M, Jones KE, et al. (2007). The delayed rise of present-day mammals. Nature, 446, 507-512.

Burleigh JG, Kimball RT \& Braun EL. (2015). Building the avian tree of life using a large-scale, sparse supermatrix. Molecular phylogenetics and evolution, 84, 53-63.

Chen MY, Liang D \& Zhang, P. (2015). Selecting question-specific genes to reduce incongruence in phylogenomics: a case study of jawed vertebrate backbone phylogeny. Systematic biology, 64, 1104-1120.

Davis KE \& Page RDM. (2014). Reweaving the Tapestry: a supertree of birds. PLoS Currents: Tree of Life, doi:10.1371/currents.tol.c1af68dda7c999ed9f1e4b2d2df7a08e. http://currents.plos.org/treeoflife/article/reweaving-the-tapestry-a-supertree-of-birds/

Gatesy J, Matthee C, DeSalle R, \& Hayashi C. (2002). Resolution of a supertree/supermatrix paradox. Systematic Biology, 51, 652-664.

Gatesy J \& Springer MS. (2004). A critique of matrix representation with parsimony supertrees. In Phylogenetic Supertrees (pp. 369-388). Springer Netherlands.

Hackett SJ, Kimball RT, Reddy S, et al. (2008). A phylogenomic study of birds reveals their evolutionary history. Science, 320, 1763-1768.

Hinchliff CE, Smith SA, Allman JF, et al. (2015). Synthesis of phylogeny and taxonomy into a comprehensive tree of life. Proceedings of the National Academy of Sciences, 112:12764–12769.

Jarvis ED, Mirarab S, Aberer AJ, et al. (2014). Whole-genome analyses resolve early branches in the tree of life of modern birds. Science, 346, 1320-1331.

Jetz W, Thomas GH, Joy JB, Hartmann K \& Mooers AO. (2012). The global diversity of birds in space and time. Nature, 491, 444-448.

Liu FGR, Miyamoto MM, Freire NP, Ong PQ, Tennant MR, Young TS \& Gugel KF. (2001). Molecular and morphological supertrees for eutherian (placental) mammals. Science, 291, 1786-1789.

Prum RO, Berv JS, Dornburg A, Field DJ, Townsend JP, Lemmon EM, \& Lemmon AR (2015). A comprehensive phylogeny of birds (Aves) using targeted next-generation DNA sequencing. Nature, 526, 569–573.

Springer MS \& de Jong WW. (2001). Which mammalian supertree to bark up? Science, 291, 1709-1711.

Steel M \& Rodrigo A. (2008). Maximum likelihood supertrees. Systematic biology, 57, 243-250.

Wang N, Kimball RT, Braun EL, Liang B \& Zhang Z. 2016 (early online). Ancestral range reconstruction of Galliformes: the effects of topology and taxon sampling. Journal of Biogeography, doi: 10.1111/jbi.12782. http://onlinelibrary.wiley.com/doi/10.1111/jbi.12782/full

\newpage
\section*{Reviewer 2 (Luay Nakhleh)}
\subsection*{Basic reporting}
The authors report on a new method for constructing supertrees from an input set of trees. Supertree methods are essential if we're to construct trees with millions of taxa, which is what the authors pursue here. The authors describe four criteria that they argue a supertree construction method should satisfy. They then proceed to present a method that satisfies the first three (the method is a greedy algorithm, so it is not guaranteed to satisfy the fourth criterion). They ran the method on a data set of over 2 million taxa (in the resulting supertree) and compared the results to those obtained by the method of Hinchliff et al. (2015).
\subsection*{Experimental design}
The authors test their method on one data set and compare their findings to those obtained by the method of Hinchliff et al.
\subsection*{Validity of the findings}
The results reported on the data set, and comparisons to the method of Hinchliff et al. are reasonable.
\subsection*{Comments for the Author}
I have three general comments:

1. Related work: There is rich literature on methods for supertree methods. I suspect that these existing methods do not handle data sets with millions of taxa, which is why the authors probably chose not to discuss them. But, still, I'd encourage the authors to have a discussion (in the Background) on existing supertree methods and why they chose not to use them. In particular, it would be useful for the community to understand for existing methods which of the four criteria they satisfy.

2. Writing: While the manuscript is well-written in general, I feel the section titled ``Notation, terminology, and the definition of supported by'' needs to be rewritten. The terminology and definitions are confusing. I think it would be much easier for the reader if definitions where made in terms of bipartitions and clusters (induced by edges). If the authors define an edge/node as defining a cluster (in a rooted tree), then compatibility, support, etc., would all be easily defined in terms of set containment and intersection. I feel that I understand what the authors meant to say because I'm familiar with the terminology; that might not be the case for someone not familiar with it.

\begin{reply}
Yes, using standard terminology would make this section clearer.  We have rewritten this section to define ``displays'' and ``conflicts'' in terms of set containment and intersection.
\end{reply}

Also, a minor point with big implications: To prove a problem is NP-hard, a reduction *to* (not *from*) it is necessary. So, please replace "to Max-Clique" by "from Max-Clique" in Section 1.1.2.

\begin{reply}
Thanks - fixed.
\end{reply}

3. Performance: The authors chose one data set. Could they do an analysis of some additional data sets? For example, they can try the "perfect" case in which they take one of the supertrees they got, get out of it many subtrees (so, they know that these subtrees could be put back together into a supertree, and they know what that supertree is), and then run the methods?

\begin{reply}
  We performed this experiment and added the following section describing the experiment and the results:
  \begin{quote}
    To test the reliability of our methods, we divided the OpenTree 4' summary tree into a collection of input trees. The input trees were obtained by splitting the grafted tree at each taxonomy node, so that each input tree has a taxonomy node at the root, and taxonomy nodes as tips, but no taxonomy nodes internally. \emph{is this right?}  This approach ensures that any taxa that were contested in the original analyses remain contested in this new analysis.  We then constructed a summary tree from these input trees using the propinquity pipeline.  The resulting summary tree was identical to the OpenTree 4' summary tree.
  \end{quote}
\end{reply}

\end{document}

### 2017-01-12 doc/paper1-submission/paper5.tex
Update.

@@
 in \citet[pages 92 and 131;][]{HusonRS2010} the maximum compatibility
-problem is known to be $NP$-hard via a reduction to Max-Clique \citep{Karp1972}.
+problem is known to be $NP$-hard via a reduction from Max-Clique \citep{Karp1972}.
 In addition to being computationally daunting, this formulation of
@@
 
+\subsection{Checking reliability of the method}
+    To test the reliability of our methods, we divided the OpenTree 4' summary tree into a collection of input trees.
+The input trees were obtained by splitting the grafted tree at each taxonomy node, so that each input tree has a taxonomy node at the root, and taxonomy nodes as tips, but no taxonomy nodes internally.
+\emph{is this right?}  This approach ensures that any taxa that were contested in the original analyses remain contested in this new analysis.
+We then constructed a summary tree from these input trees using the propinquity pipeline.
+The resulting summary tree was identical to the OpenTree 4' summary tree.
+
 \section{Conclusions}

### 2017-01-12 doc/paper1-submission/response.tex
Fixes.

@@
 \begin{reply}
-2a. Concern: incorrect input trees could lead to supertree that is not correct.  Response: our goal is to summarize the input trees.  This would work just fine.
+2a.
+Concern: incorrect input trees could lead to a supertree that is not correct.
+Response: our goal is to summarize the input trees, not to infer the true tree.
+This would work just fine with incorrect input trees.
 
 2b.
-Concern: hypotheses that occur in more input trees might not be correct, so a supertree method should not hypothesis just based on frequency.
+Concern: hypotheses that occur in more input trees might not be correct, so a supertree method should not favor hypotheses just based on frequency.
 Response: our method does not support hypotheses because of their frequency.
+We use ranks instead.
 
@@
 Downweighting bad input trees might assuage this.
-Response: we are not performing a phylogeny inference.
-We don't directly use weights, but instead of community-curated ranks.
-While it is possible to construct weights so that we can phrase our rank-based procedure as an optimization problem, at no point do we ever construct weights.
+Response: we are not performing a phylogeny inference, but attemptint to summarize input trees, so the correctness of the phylogeny is not an issue.
+While it is possible to describe weights so that we can phrase our rank-based procedure as an optimization problem, at no point do we ever construct weights.
 
@@
 We would only need to estimate weights if we were trying to infer the true phylogeny from the input trees, a la the supertree versus supermatrix debate.
-But we are not, therefore there is not need to derive weights.
+But we are not, therefore there is no need to derive weights.
 
 2.
-It seems that the reviewer thinks that we are trying to infer the true tree from a datasets consisting of the input trees.
+It seems that the reviewer thinks that we are trying to infer the true tree from a dataset consisting of the input trees, like Bininda-Edmonds.
 Such an inference would need a principled method of resolving conflicts between input trees using only the information in the set of input trees.
@@
 While the supertree must conflict with some input trees, we do not claim that this constitutes an estimate of which input trees are correct or incorrect.
-Instead we seek to represent all input trees, including those that conflict with the phylogeny estimate by linking from summary tree edges to conflicting input tree edges via our annotations file.
+Instead we seek to represent all input trees (including those that conflict with the phylogeny estimate) by linking from summary tree edges to conflicting input tree edges via our annotations file.
 
@@
 However, our conflict resolution is not phrased as an estimate of the true tree.
-Instead, we supply rankings to allow curators to selection which trees should win conflicts.
+Instead, we supply rankings to allow curators to select which trees should win conflicts.
 

### 2017-01-17 doc/paper1-submission/response.tex
Update new section 3.4.

@@
 \begin{reply}
-  We performed this experiment and added the following section describing the experiment and the results:
+Thanks for this suggestion.  We performed this experiment and added the following section describing the experiment and the results: \textcolor{red}{\textbf{Mark please check}}
 \begin{quote}
-    To test the reliability of our methods, we divided the OpenTree 4' summary tree into a collection of input trees.
-The input trees were obtained by splitting the grafted tree at each taxonomy node, so that each input tree has a taxonomy node at the root, and taxonomy nodes as tips, but no taxonomy nodes internally.
-\emph{is this right?}  This approach ensures that any taxa that were contested in the original analyses remain contested in this new analysis.
-We then constructed a summary tree from these input trees using the propinquity pipeline.
-The resulting summary tree was identical to the OpenTree 4' summary tree.
+To test the reliability of our methods, we divided the OTLv4' summary
+tree into a collection of input trees. The input trees were obtained
+by splitting the grafted tree at each taxonomy node, so that each
+input tree has a taxonomy node at the root, and taxonomy nodes as
+tips, but no taxonomy nodes internally.  Each leaf node that was an
+internal taxon was then replaced with a leaf taxon. This approach
+ensures that any taxa that were contested in the original analyses
+remain contested in this new analysis.  We then constructed a summary
+tree from these input trees using the propinquity pipeline.  The
+resulting summary tree was identical to the OTLV4' summary tree.  
 \end{quote}
+
+Note that we are not trying to perform phylogeny estimation here like Bininda-Edmonds, so we are not treating the input trees as stand-ins for the data matrices that they may have been constructed from.
+Because of this, we aren't treating the input tree collection as a data set.
+To clarify this to the readers, we add the following text to the Background section:
+\begin{quote}
+Add non-goals section here (to be written).
+\end{quote}
+
 \end{reply}

### 2017-01-17 doc/paper1-submission/paper5.tex
Update new section 3.4.

@@
 \subsection{Checking reliability of the method}
-    To test the reliability of our methods, we divided the OpenTree 4' summary tree into a collection of input trees.
-The input trees were obtained by splitting the grafted tree at each taxonomy node, so that each input tree has a taxonomy node at the root, and taxonomy nodes as tips, but no taxonomy nodes internally.
-\emph{is this right?}  This approach ensures that any taxa that were contested in the original analyses remain contested in this new analysis.
-We then constructed a summary tree from these input trees using the propinquity pipeline.
-The resulting summary tree was identical to the OpenTree 4' summary tree.
-
+To test the reliability of our methods, we divided the OTLv4' summary
+tree into a collection of input trees. The input trees were obtained
+by splitting the grafted tree at each taxonomy node, so that each
+input tree has a taxonomy node at the root, and taxonomy nodes as
+tips, but no taxonomy nodes internally.  Each leaf node that was an
+internal taxon was then replaced with a leaf taxon. This approach
+ensures that any taxa that were contested in the original analyses
+remain contested in this new analysis.  We then constructed a summary
+tree from these input trees using the propinquity pipeline.  The
+resulting summary tree was identical to the OTLV4' summary tree.  
 \section{Conclusions}

### 2017-01-17 doc/paper1-submission/paper5.tex
Add new section on a non-goal of the summary method.

@@
 files with information about conflict and support.
+
+Although other researchers construct supertrees for the purpose of
+inferring the true tree from the input trees \citep{BinindaEmonds2007b,Davis2014},
+we note that phylogeny estimation is an explicit non-goal of our
+supertree method.  We do not treat input trees as data or as surrogates for
+data matrices \citep{gatesy2004critique}, and we do not claim that our
+summary supertree is more accurate than the input trees because more
+data lies behind it. Instead we seek to \emph{summarize and represent}
+the input trees, even when they conflict with each other.  The
+annotations file constructed by our pipeline is essential for this
+task because it includes information about conflict with our supertree
+as well as support for it.
 
@@
 \caption{Subproblems generated from the exemplified trees shown in figure \ref{fig:exemplify-example}.
-A trivial statemen from the first tree that a taxon labelled $ABCD$
+A trivial statement from the first tree that a taxon labelled $ABCD$
 is sister to $E$ has been omitted, because trees with only 2 leaves

### 2017-01-17 doc/paper1-submission/response.tex
Update.

@@
 \usepackage{titlesec}
+\usepackage{natbib}
 %\usepackage{a4wide}
@@
 
+\begin{reply}
+The first reviewer doesn't number their points, so we will divide them into
+three groups (1, 2, and 3) following the reviewer's ``three issues'',
+with sub-points indicated by letters.
+\end{reply}.
+
 I do not believe that there is no such problem with the approach proposed by R\&H.
@@
 \begin{reply}
-1. It seems that the reviewer is satisfied with our criteria for the supertree method.
+1. It seems that the reviewer is satisfied with our criteria for what
+counts as a good supertree.
 \end{reply}
@@
 
+\begin{reply}
+2a.
+The reviewer seems to assume that out supertree algorithm is designed to infer the true tree from the input trees, as supertree algorithms were used in the supertree-versus-supermatrix debate.
+However, we don't say that anywhere, and are not in fact trying to do that.
+We agree with the reviewer that there are a lot of problems with trying to infer phylogenies from input trees, as expressed in Gatesy et al (2004).
+We will address this underlying concern at the end.
+We note addressing individual concerns in the order in which they are written.
+
+We appreciate that attempting to infer the true tree from a collection of input trees could be undermined by the inclusion of multiple trees that all represent ``traditional'' relationships.
+However: (1) our supertree approach is not designed to estimate the true tree, but to represent the input trees even if they are incorrect (2) we handle conflict by allow curators to influence the supertree via ranking as discussed in the text:
+
+\begin{quote}
+If an expert were quite confident in the c grouping, then she could assign that input node a high ranking.
+A supertree that used ranks could then recover this grouping even if its inclusion did not increase the total number of input nodes that are displayed by the summary tree.
+\end{quote}
+
+Note that in our rank-based system ``traditional'' relationships are not favored in conflict resolution, even if they do occur in more input trees.
+\end{reply}
+
 The issue of including data from “bad” trees is especially acute when taxonomies and/or “conglomerated trees” (i.e., trees based on expert evaluation of the literature that largely reflect the consensus at that time; cf.
@@
 Hopefully, the downweighted tree will correspond to less reliable sources of phylogenetic information.
+
+\begin{reply}
+2b.
+We agree that if we were attempting to infer the true tree from the input trees, downweighting bad trees might improve the results somewhat.
+However, we really are trying to represent the input trees, so this is not directly a concern.
+
+However, while it is indeed possible to recast our method as an optimization problem by assigning weights to the input trees, we are not trying to weight the evidence in the input trees, and we do not actually ever calculate these weights in our pipeline.
+\end{reply}
 
@@
 Perhaps the authors could elaborate on this issue.
+
+\begin{reply}
+2c.
+We agree that the ranks have a large impact on the supertree algorithm.
+If this were not the case, then ranks could not achieve their stated goal of allowing users to use ranks to fix problems in the tree.
+If we were performing phylogeny estimation, then these ranks might not give low-ranked trees sufficient weight in conflict resolution.
+However, again, we do not seek to infer the correct tree by treating the input trees as data.
+
+We agree with the reviewer that applying weights to trees rather than nodes is problematic, and have added the following to the discussion:
+\begin{quote}
+ say that applying weights to trees is problematic.
+\end{quote}
+\end{reply}
 
@@
 
+\begin{reply}
+2d. Explaining the Hinchliff (2015) paper is beyond the scope of the current manuscript.
+\end{reply}
+
 With these concerns out of the way I will add that I am uncertain whether a supertree method that achieves “perfect” (or nearly perfect) weighting can be devised.
@@
 \begin{reply}
-2a.
-Concern: incorrect input trees could lead to a supertree that is not correct.
-Response: our goal is to summarize the input trees, not to infer the true tree.
-This would work just fine with incorrect input trees.
-
-2b.
-Concern: hypotheses that occur in more input trees might not be correct, so a supertree method should not favor hypotheses just based on frequency.
-Response: our method does not support hypotheses because of their frequency.
-We use ranks instead.
-
-2c.
-Concern: bad input trees could compromise the correctness of inferred phylogeny.
-Downweighting bad input trees might assuage this.
-Response: we are not performing a phylogeny inference, but attemptint to summarize input trees, so the correctness of the phylogeny is not an issue.
-While it is possible to describe weights so that we can phrase our rank-based procedure as an optimization problem, at no point do we ever construct weights.
-
-2d.
-Concern: the weights have a large impact on the method.
-Please describe how sensitive the method is to the weights.
-Response: we indeed hope that the ranking has a substantial effect on the resulting tree.
-If this were not the case, then ranks could not achieve their stated goal of allowing users to use ranks to fix problems in the tree.
-
-Despite this, we ran the pipeline with the ranks reversed.
-This drastic alteration of the ranks lead to a supertree that conflicted with 1264 branches of the original supertree.
-Compared to this, the opentree 4.0 summary tree that used preceding methodology conflicts with 1387 branches of the 4.0’ tree using current methods.
-However, most of the 36533 internal edges of the grafted tree are not altered.
-
-2e. Concern: the weights are applied to trees rather than nodes.  This seems to be the only valid concern.
-
-2f.
-Concern: the weights cannot be derived in a principled manner.
-Results: we are not trying to estimate the weights from the input trees.
-We would only need to estimate weights if we were trying to infer the true phylogeny from the input trees, a la the supertree versus supermatrix debate.
-But we are not, therefore there is no need to derive weights.
+We agree that trying to derive weights from the input trees alone in a principled manner would be extremely difficult.
+That is why we do not attempt it.
+Again, our goal in constructing the supertree is not to find the correct tree, and we rely use ranks to allow curators to influence conflict resolution.
 
@@
 However, our conflict resolution is not phrased as an estimate of the true tree.
-Instead, we supply rankings to allow curators to select which trees should win conflicts.
+Instead, we supply rankings to allow curators to influence conflict resolution to produce the summary tree that is most useful to them.
 
-We describe how our ranking method could be considered as a weight-based method for the purposes of optimization.
-However, such weights should not be interpreted as an attempt to using the (weighted) input trees to decide which alternative phylogenetic hypotheses are correct, and in fact we never say anything like this.
-Our weights are not computed from the input trees, because we are not performing phylogeny estimation.
+We appreciate the amount of thought the reviewer put into articulating their concerns, and suspect that a number of readers will have the same assumptions.
+To avoid confusion, we therefore added the following to our section on goals:
+\begin{quote}
+Although other researchers construct supertrees for the purpose of
+inferring the true tree from the input trees \citep{BinindaEmonds2007b,Davis2014},
+we note that phylogeny estimation is an explicit non-goal of our
+supertree method.  We do not treat input trees as data or as surrogates for
+data matrices \citep{gatesy2004critique}, and we do not claim that our
+summary supertree is more accurate than the input trees because more
+data lies behind it. Instead we seek to \emph{summarize and represent}
+the input trees, even when they conflict with each other.  The
+annotations file constructed by our pipeline is essential for this
+task because it includes information about conflict with our supertree
+as well as support for it.
+\end{quote}
 
-The fact that rankings apply to entire trees is indeed an issue with the method.
+Furthermore, it would be interesting to know how much effect the ranks had.  We therefore add the following section:
+\begin{quote}
+We ran the pipeline on the OTL4 data with the ranks reversed.
+This drastic alteration of the ranks lead to a summary supertree that conflicted with 1264 branches of the OTLv4'.
+Compared to this, the OTLv4 supertree tree conflicts with 1387 branches of the OTLv4’.
+However, a total of 36533 edges of the grafted tree could be altered by changing the ranks, and most of these edges are unaffected.
+\end{quote}
 
@@
 
-Note that we are not trying to perform phylogeny estimation here like Bininda-Edmonds, so we are not treating the input trees as stand-ins for the data matrices that they may have been constructed from.
-Because of this, we aren't treating the input tree collection as a data set.
-To clarify this to the readers, we add the following text to the Background section:
-\begin{quote}
-Add non-goals section here (to be written).
-\end{quote}
-
 \end{reply}
 
+\bibliographystyle{upmplainnat}
+\bibliography{otcetera}
 \end{document}

### 2017-01-17 doc/paper1-submission/paper5.tex
Update.

@@
 \subsection{Checking reliability of the method}
-To test the reliability of our methods, we divided the OTLv4' summary
+To test the reliability of our methods, we divided the \otlprop{} summary
 tree into a collection of input trees. The input trees were obtained
@@
 tree from these input trees using the propinquity pipeline.  The
-resulting summary tree was identical to the OTLV4' summary tree.  
+resulting summary tree was identical to the \otlprop{} summary tree.
+
+\subsection{Sensitivity of the \otlprop{} supertree to ranks}
+We ran the pipeline on the OTL4 data with the ranks reversed.
+This drastic alteration of the ranks lead to a summary supertree that conflicted with 1264 branches of the \otlprop{}.
+Compared to this, the \otl{} supertree tree conflicts with 1387 branches of the \otlprop{}.
+However, a total of 36533 edges of the grafted tree could be altered by changing the ranks, and most of these edges are unaffected.
+
 \section{Conclusions}

### 2017-01-18 doc/paper1-submission/response.tex
Update.

@@
 
-\begin{reply}
-The first reviewer doesn't number their points, so we will divide them into
-three groups (1, 2, and 3) following the reviewer's ``three issues'',
-with sub-points indicated by letters.
-\end{reply}.
-
+1.
 I do not believe that there is no such problem with the approach proposed by R\&H.
@@
 \begin{reply}
-1. It seems that the reviewer is satisfied with our criteria for what
-counts as a good supertree.
+We are pleased that the reviewer is satisfied with our criteria for what counts as a good supertree.
 \end{reply}
 
+2.
 The other challenge for supertrees is the low quality of some source trees (this may be a larger problem for older trees).
@@
 
-\begin{reply}
-2a.
-The reviewer seems to assume that out supertree algorithm is designed to infer the true tree from the input trees, as supertree algorithms were used in the supertree-versus-supermatrix debate.
-However, we don't say that anywhere, and are not in fact trying to do that.
-We agree with the reviewer that there are a lot of problems with trying to infer phylogenies from input trees, as expressed in Gatesy et al (2004).
-We will address this underlying concern at the end.
-We note addressing individual concerns in the order in which they are written.
-
-We appreciate that attempting to infer the true tree from a collection of input trees could be undermined by the inclusion of multiple trees that all represent ``traditional'' relationships.
-However: (1) our supertree approach is not designed to estimate the true tree, but to represent the input trees even if they are incorrect (2) we handle conflict by allow curators to influence the supertree via ranking as discussed in the text:
-
-\begin{quote}
-If an expert were quite confident in the c grouping, then she could assign that input node a high ranking.
-A supertree that used ranks could then recover this grouping even if its inclusion did not increase the total number of input nodes that are displayed by the summary tree.
-\end{quote}
-
-Note that in our rank-based system ``traditional'' relationships are not favored in conflict resolution, even if they do occur in more input trees.
-\end{reply}
-
 The issue of including data from “bad” trees is especially acute when taxonomies and/or “conglomerated trees” (i.e., trees based on expert evaluation of the literature that largely reflect the consensus at that time; cf.
@@
 Hopefully, the downweighted tree will correspond to less reliable sources of phylogenetic information.
-
-\begin{reply}
-2b.
-We agree that if we were attempting to infer the true tree from the input trees, downweighting bad trees might improve the results somewhat.
-However, we really are trying to represent the input trees, so this is not directly a concern.
-
-However, while it is indeed possible to recast our method as an optimization problem by assigning weights to the input trees, we are not trying to weight the evidence in the input trees, and we do not actually ever calculate these weights in our pipeline.
-\end{reply}
 
@@
 Perhaps the authors could elaborate on this issue.
-
-\begin{reply}
-2c.
-We agree that the ranks have a large impact on the supertree algorithm.
-If this were not the case, then ranks could not achieve their stated goal of allowing users to use ranks to fix problems in the tree.
-If we were performing phylogeny estimation, then these ranks might not give low-ranked trees sufficient weight in conflict resolution.
-However, again, we do not seek to infer the correct tree by treating the input trees as data.
-
-We agree with the reviewer that applying weights to trees rather than nodes is problematic, and have added the following to the discussion:
-\begin{quote}
- say that applying weights to trees is problematic.
-\end{quote}
-\end{reply}
 
@@
 
-\begin{reply}
-2d. Explaining the Hinchliff (2015) paper is beyond the scope of the current manuscript.
-\end{reply}
-
 With these concerns out of the way I will add that I am uncertain whether a supertree method that achieves “perfect” (or nearly perfect) weighting can be devised.
@@
 \begin{reply}
-We agree that trying to derive weights from the input trees alone in a principled manner would be extremely difficult.
-That is why we do not attempt it.
-Again, our goal in constructing the supertree is not to find the correct tree, and we rely use ranks to allow curators to influence conflict resolution.
+We appreciate the thought that the reviewer put into articulating their concerns.
+The reviewer seems to assume that our supertree algorithm is designed to infer the true tree from the input trees like Bininda-Emonds.
+However, we don't say that anywhere, and are not in fact trying to do that.
+We agree with the reviewer that there are a lot of problems with trying to infer phylogenies from input trees, as expressed in the Gatesy et al (2004) critique of supertree methods.
 
-2.
-It seems that the reviewer thinks that we are trying to infer the true tree from a dataset consisting of the input trees, like Bininda-Edmonds.
-Such an inference would need a principled method of resolving conflicts between input trees using only the information in the set of input trees.
-The resulting supertree could then be used to critique the input trees, so that input trees that don't match the supertree would be considered to be incorrect (up to estimation error).
-
-However, our approach is fundamentally different.
-We do not seek to infer a tree that is more accurate than any individual input tree.
-Instead we seek to summarize and describe the set of input trees.
-While the supertree must conflict with some input trees, we do not claim that this constitutes an estimate of which input trees are correct or incorrect.
-Instead we seek to represent all input trees (including those that conflict with the phylogeny estimate) by linking from summary tree edges to conflicting input tree edges via our annotations file.
-
-Our approach does involve the resolution of conflict.
-However, our conflict resolution is not phrased as an estimate of the true tree.
-Instead, we supply rankings to allow curators to influence conflict resolution to produce the summary tree that is most useful to them.
-
-We appreciate the amount of thought the reviewer put into articulating their concerns, and suspect that a number of readers will have the same assumptions.
-To avoid confusion, we therefore added the following to our section on goals:
+To prevent confusion among readers who might assume that we are trying to infer the true tree.  To avoid this interpretation, we add the following text to the Background:
 \begin{quote}
@@
 
-Furthermore, it would be interesting to know how much effect the ranks had.  We therefore add the following section:
+With that clarification behind us, we can respond to a number of specific concerns:
+
+2a.
+The reviewer seems concerned that low-quality source trees will undermine the estimation, and specifically that relationships present in more trees will dominate, leading to an incorrect estimate of the true tree.
+However: (1) our supertree approach is not designed to estimate the true tree, but to represent the input trees even if they are incorrect; (2) we handle conflict by allowing curators to influence the supertree via ranking as discussed in the text.
+Our conflict resolution is not designed to infer the tree.
+
+\begin{quote}
+If an expert were quite confident in the c grouping, then she could assign that input node a high ranking.
+A supertree that used ranks could then recover this grouping even if its inclusion did not increase the total number of input nodes that are displayed by the summary tree.
+\end{quote}
+
+Note that in our rank-based system ``traditional'' relationships are not favored in conflict resolution, even if they do occur in more input trees, because our supertree algorithm does not favor relationships that occur more often.
+
+% We probably do need to state that, unlike MRP and MinCut, our approach doesn't favor relationships that occur more often.
+
+2b.
+The reviewer seems concerned that our tree weights are extreme, and that it is difficult to estimate good tree weights from the input trees.
+While it is possible to recast our method as an optimization problem by assigning weights to the input trees, we are not trying to weight the evidence in the input trees, and we do not actually ever calculate these weights in our pipeline.
+
+We agree that the ranks have a large impact on conflict resolution in the supertree algorithm.
+If this were not the case, then ranks could not achieve their stated goal of allowing curators to fix problems in the tree.
+If we were performing phylogeny estimation, then these ranks might not give low-ranked trees sufficient weight in conflict resolution.
+However, again, we do not seek to infer the correct tree by treating the input trees as data.
+
+%Our approach is somewhat novel in that we replace a kind of pseudo-estimation framework with a curation-based framework in which curators can influence conflict resolution by specifying high ranks for trees containing a desired relationship:
+
+2c. We agree that it would be interesting to know how much effect the ranks had.  We therefore add the following section:
 \begin{quote}
@@
 \end{quote}
+
+2d. We agree with the reviewer that applying ranks to trees rather than nodes is problematic, and have added the following to the discussion:
+\begin{quote}
+ say that applying ranks to trees is problematic.
+\end{quote}
+
+\begin{reply}
+2e. Explaining the Hinchliff (2015) paper is beyond the scope of the current manuscript.
+\end{reply}
+
+%2.
+It seems that the reviewer thinks that we are trying to infer the true tree from a dataset consisting of the input trees, like Bininda-Edmonds.
+Such an inference would need a principled method of resolving conflicts between input trees using only the information in the set of input trees.
+The resulting supertree could then be used to critique the input trees, so that input trees that don't match the supertree would be considered to be incorrect (up to estimation error).
+
+%However, our approach is fundamentally different.
+We do not seek to infer a tree that is more accurate than any individual input tree.
+Instead we seek to summarize and describe the set of input trees.
+While the supertree must conflict with some input trees, we do not claim that this constitutes an estimate of which input trees are correct or incorrect.
+Instead we seek to represent all input trees (including those that conflict with the phylogeny estimate) by linking from summary tree edges to conflicting input tree edges via our annotations file.
 

### 2017-01-18 doc/paper1-submission/response.tex
Update.

@@
 \titlespacing*{\subsection}{0pt}{3pt}{0pt}
+
+\newcommand{\otl}{OTLv4}
+\newcommand{\otlprop}{OTLv4$^\prime$}
 
@@
                {\endlist}
-
+\sloppy
 \begin{document}
@@
 
-We thank the reviewers for their comments.  Here is our response.
+We thank the reviewers for their comments.  In response their suggestions, we performed
+two experiments designed to check the reliability of our reconstruction, and to assess
+the sensitivity of our method to the given ranks.
 
@@
 \begin{reply}
-We are pleased that the reviewer is satisfied with our criteria for what counts as a good supertree.
+Response 1. We are pleased that the reviewer is satisfied with our criteria for what counts as a good supertree.
 \end{reply}
@@
 I have two concerns regarding the use of tree weights. First, it seems clear that weights could have an extremely large impact in the R\&H method. According to part A of their appendix:
-
-“The summary tree constructed by the propinquity pipeline is a greedy heuristic for finding a tree that maximizes this score when the weights for a node are determined by the tree’s weight and the difference in weighting is so large that displaying one node from a highly ranked tree is preferred to displaying all of the nodes in the trees with lower rank.”
-
+\begin{quote}
+The summary tree constructed by the propinquity pipeline is a greedy heuristic for finding a tree that maximizes this score when the weights for a node are determined by the tree’s weight and the difference in weighting is so large that displaying one node from a highly ranked tree is preferred to displaying all of the nodes in the trees with lower rank.
+\end{quote}
 Some elaboration on the sensitivity of their method to tree weights seems warranted.
@@
 \begin{reply}
+Response 2.
 We appreciate the thought that the reviewer put into articulating their concerns.
@@
 However, we don't say that anywhere, and are not in fact trying to do that.
+Instead we are trying to summarize the input trees, including low-quality inputs.
 We agree with the reviewer that there are a lot of problems with trying to infer phylogenies from input trees, as expressed in the Gatesy et al (2004) critique of supertree methods.
 
-To prevent confusion among readers who might assume that we are trying to infer the true tree.  To avoid this interpretation, we add the following text to the Background:
+We add the following text to the Background clarify that we are not trying to infer the true tree from the input trees:
 \begin{quote}
@@
 
-Note that in our rank-based system ``traditional'' relationships are not favored in conflict resolution, even if they do occur in more input trees, because our supertree algorithm does not favor relationships that occur more often.
+Note that in our rank-based method of conflict-resolution, ``traditional'' relationships are not favored in conflict resolution, even if they do occur in more input trees, because our supertree algorithm does not favor relationships that occur more often.
 
@@
 We ran the pipeline on the OTL4 data with the ranks reversed.
-This drastic alteration of the ranks lead to a summary supertree that conflicted with 1264 branches of the OTLv4'.
-Compared to this, the OTLv4 supertree tree conflicts with 1387 branches of the OTLv4’.
-However, a total of 36533 edges of the grafted tree could be altered by changing the ranks, and most of these edges are unaffected.
+This drastic alteration of the ranks lead to a summary supertree that conflicted with 1264 branches of the \otlprop{}.
+Compared to this, the \otl{} supertree tree conflicts with 1387 branches of the \otlprop{} supertree.
+Thus, reversing the ranks makes a smaller different in the \otlprop{} supertree than the new synthesis algorithm.
+We also note that a total of 36533 edges of the grafted tree could be altered by changing the ranks, and most of these edges are unaffected.
 \end{quote}
@@
 
-\begin{reply}
 2e. Explaining the Hinchliff (2015) paper is beyond the scope of the current manuscript.
 \end{reply}
-
 %2.
@@
 % Summary: argh.
-
-\end{reply}
 
@@
 \begin{reply}
-3.
+Response 3.
 
@@
 \end{quote}
-
 \end{reply}

### 2017-01-18 doc/paper1-submission/paper5.tex
Update.

@@
 \thispagestyle{empty}
+
+% TODO: add stuff to paper
+% R1.2 Mention that applying ranks to trees instead of nodes is problematic.
+%      Mention transparency?
+% R2.1: Add paragraph mentioning MRP, MinCut, and how they are different. - 1.1.6
 
@@
 This drastic alteration of the ranks lead to a summary supertree that conflicted with 1264 branches of the \otlprop{}.
-Compared to this, the \otl{} supertree tree conflicts with 1387 branches of the \otlprop{}.
-However, a total of 36533 edges of the grafted tree could be altered by changing the ranks, and most of these edges are unaffected.
+Compared to this, the \otl{} supertree tree conflicts with 1387 branches of the \otlprop{} supertree.
+Thus, reversing the ranks makes a smaller different in the \otlprop{} supertree than the new synthesis algorithm.
+We also note that a total of 36533 edges of the grafted tree could be altered by changing the ranks, and most of these edges are unaffected.
 

### 2017-01-19 doc/paper1-submission/paper5.tex
Update.

@@
 from lower ranked trees.
+
+\subsubsection{Comparison to other supertree methods}
+
+So, we also consider MRP and MinCut.  MRP can display edges not found
+in any input tree, and so fails criterion #1.  MRP is used in many
+studies that attempt to infer the true tree from a lot of input
+trees.  To some extent, MRP tries to treat each tree as data, which
+differs from our approach here.
+
+MinCut (and Modified MinCut) attempts to favor relationships present
+in multiple input trees during conflict resolution.
+
+Both MRP and MinCut attempt to adjudicate conflicts in a way that
+favors relationships present in more input trees.  Thus, they perform
+something more like a weighted average.  However, the reason the
+supertree resolves conflicts in a particular way is not clear and
+transparent, and does not allow a method for curators to adjust the
+supertree.
+
+A number of other supertree methods exist, including Matrix
+Representation with Parsimony (MRP) and MinCut.  MRP
+% http://currents.plos.org/treeoflife/article/reweaving-the-tapestry-a-supertree-of-birds/
+MRP - displays unsupported groups (#1), not rank-based (#2), ?
+SuperFine - computes a strict consensus, then MRP
+Minimum Flip - 
+(Modified) MinCut -
 

### 2017-01-19 doc/paper1-submission/paper5.tex
minor tweaks

@@
 \begin{abstract}
-We present a new supertree method that enables rapid estimation of a
+We present a new supertree method that enables rapid construction of a
 summary tree on the scale of millions of leaves.  This supertree
@@
 We compare our supertree construction method to a previously published
-supertree construction method by assessing their performance on input
+method by assessing their performance on input
 trees used to construct the Open Tree of Life version 4, and find that
@@
 and justifiably represent phylogenetic information from these inputs.
-The taxonomic tree is derived from the Open Tree Taxonomy (OTT hereafter,
+The taxonomic tree is derived from the Open Tree Taxonomy (OTT hereafter;
 publication in preparation). The phylogenetic inputs are published
@@
 if none of $A_{1}\cap B_{1}$, $A_{1}\cap B_{2}$, and $B_{2}\cap A_{1}$
-are empty\footnote{The intersection of the right sides of the splits is considered to
+are empty.
+Note that the intersection of the right sides of the splits is considered to
 be non-empty even if $A_{2}\cap B_{2}$ is empty, because the right
-sides implicitly contain the root.}. If $A$ and $B$ do not conflict, then we say that they are \emph{compatible}\footnote{It is possible to construct a rooted tree that displays both $A$
+sides implicitly contain the root. If $A$ and $B$ do not conflict, then we say that they are \emph{compatible}\footnote{It is possible to construct a rooted tree that displays both $A$
 and $B$ if and only if $A$ and $B$ are compatible}. If $A$ displays $B$, then $A$ and $B$ must be compatible, since
@@
 in \citet[pages 92 and 131;][]{HusonRS2010} the maximum compatibility
-problem is known to be $NP$-hard via a reduction from Max-Clique \citep{Karp1972}.
+problem is known to be $NP$-hard via a reduction to Max-Clique \citep{Karp1972}.
 In addition to being computationally daunting, this formulation of
@@
 So, we also consider MRP and MinCut.  MRP can display edges not found
-in any input tree, and so fails criterion #1.  MRP is used in many
+in any input tree, and so fails criterion \#1.  MRP is used in many
 studies that attempt to infer the true tree from a lot of input
@@
 % http://currents.plos.org/treeoflife/article/reweaving-the-tapestry-a-supertree-of-birds/
-MRP - displays unsupported groups (#1), not rank-based (#2), ?
+MRP - displays unsupported groups (\#1), not rank-based (\#2), ?
 SuperFine - computes a strict consensus, then MRP

### 2017-01-19 doc/paper1-submission/paper5.tex
fig legends and referencing

@@
 from lower ranked trees.
+Note that in this extreme form of weighting, we do not have to calculate
+the score of a tree, we simply traverse the groupings by order of rank
+and add a group if it is compatible with the previously added groups.
 
@@
 
-So, we also consider MRP and MinCut.  MRP can display edges not found
+We also consider matrix representation with parsimony (MRP) and MinCut.
+MRP can display edges not found
 in any input tree, and so fails criterion \#1.  MRP is used in many
@@
 differs from our approach here.
-
 MinCut (and Modified MinCut) attempts to favor relationships present
 in multiple input trees during conflict resolution.
-
 Both MRP and MinCut attempt to adjudicate conflicts in a way that
@@
 tree (stored as the \texttt{grafted\_solution/grafted\_solution.tre}
-by propinquity). Recall that each non-root uncontested taxon used
+by propinquity); see figure \ref{fig:grafted-solution}. Recall that each non-root uncontested taxon used
 for decomposition occurs as a leaf taxon in one subproblem and as
@@
 \par\end{centering}
-\caption{Grafted solution. The nodes E and F are monotypic here, but will end
-up being polytypic after unpruning is performed.}
+\caption{Grafted solution produced from the subproblems from
+figure \ref{fig:subproblem-solutions.} and which is the 
+backbone onto which taxa that are not included in any phylogeny
+will be placed.}
+\label{fig:grafted-solution}
 \end{figure}

### 2017-01-19 doc/paper1-submission/paper5.tex
expanded legends to deal with our confusing omission of the trivial subproblems

@@
 is sister to $E$ has been omitted, because trees with only 2 leaves
-do not contain phylogenetic information.}
+do not contain phylogenetic information.
+Similarly, the trivial problems that list E1 and F1 as the only daughters
+of taxa E and F have been omitted.}
 \label{fig:subproblems-example}
@@
 backbone onto which taxa that are not included in any phylogeny
-will be placed.}
+will be placed.
+In a real analysis, the tips shown as E and F here
+would be monotypic taxa (with children E1 and F1, respectively)
+from the trivial subproblems that were omitted from from figure \ref{fig:subproblems-example}.}
 \label{fig:grafted-solution}
@@
 \par\end{centering}
-\caption{Unpruned tree with internal taxa. Nodes E2, F2, and D have been re-added
-to the tree. The nodes E and F are no longer monotypic.}
+\caption{Unpruned tree with internal taxa. Taxa unsampled in phylogenetic
+statements have been added to the grafted tree shown
+in figure \ref{fig:grafted-solution}.}
 \end{figure}

### 2017-01-23 doc/paper1-submission/paper5.tex
Add section about output tree accuracy.

@@
 
+Our supertree pipeline makes it possible to summarize and merge
+information from across the tree of life.  However, we do not claim
+that our supertree is more accurate than the input trees that is
+summarizes.  We demonstrate that ranks supplied by curators have
+substantial effects on the output of the pipeline.  Currently every
+node in an input tree must recieve the same rank, and bootstrap
+proportions and posterior probabilities are not taken into account.
+Therefore, before using the output of our supertree pipeiline in an
+evolutionary analysis, researchers must first assess the accuracy of
+the output supertree, possibly adjusting the ranks of the input trees
+to produce a more useful outcome.
+
 The migration of summary tree construction from treemachine (used for

### 2017-01-23 doc/paper1-submission/paper5.tex
Move "non-goal" paragraph into subsubsection and clarify.

@@
 files with information about conflict and support.
-
-Although other researchers construct supertrees for the purpose of
-inferring the true tree from the input trees \citep{BinindaEmonds2007b,Davis2014},
-we note that phylogeny estimation is an explicit non-goal of our
-supertree method.  We do not treat input trees as data or as surrogates for
-data matrices \citep{gatesy2004critique}, and we do not claim that our
-summary supertree is more accurate than the input trees because more
-data lies behind it. Instead we seek to \emph{summarize and represent}
-the input trees, even when they conflict with each other.  The
-annotations file constructed by our pipeline is essential for this
-task because it includes information about conflict with our supertree
-as well as support for it.
 
@@
 with the supertree.
+
+\subsubsection{Non-goal: phylogeny estimation}
+Many researchers construct supertrees for the purpose of inferring the true tree from the input trees \citep{BinindaEmonds2007b,Davis2014}.
+Such approaches treat input trees as data or as surrogates for data matrices \citep{gatesy2004critique}, and the supertree is seen as more accurate than the individual input trees because more input trees (and presumably more data) stand behind it.
+For such methods, conflict resolution is a primary aim of the method.
+
+Our supertree method operates in a different paradigm.
+Phylogeny estimation is an explicit non-goal of our supertree algorithm.
+We do not claim that the supertree is more accurate than the input trees it summarizes, and so conflict resolution is not a primary aim of the method.
+Most of our input trees have non-overlapping taxon sets, and so conflict resolution is not the primary problem.
+Instead, seek to merge and aggregate phylogenetic information on the largest possible scale.
+Therefore, the primary aim of our supertree method is to \emph{summarize and represent} the input trees, even when they conflict with each other.
+The annotations file constructed by our pipeline is essential output because it includes information about conflict with our supertree as well as support for it.
+
+Our method must resolve conflicts in order to construct a single supertree.
+However, the rank information used to resolve conflicts is an input to the method, not an output from the method.
+We thus perform curation-based conflict resolution, not inference-based conflict resolution.
 

### 2017-01-23 doc/paper1-submission/paper5.tex
Mention that weights are not weights of evidence.

@@
 from lower ranked trees.
+These large weights help to achieve our goal of allowing curators to fix problems in the tree.
+While it is possible to view our pipeline as an optimization algorithm by assigning weights to the input trees in this way, we are not trying to weight the evidence in the input trees, and we do not actually ever calculate these weights in our pipeline.
 Note that in this extreme form of weighting, we do not have to calculate

### 2017-01-23 doc/paper1-submission/paper5.tex
Typo fix.

@@
 responsible for taxa being included or excluded from the summary tree.
-Additionally, the new propinquity software pipeline has decreased the decreased
+Additionally, the new propinquity software pipeline has decreased
 the computational time required to construct a supertree from several

### 2017-01-23 doc/paper1-submission/paper5.tex
Fix typo.

@@
 Most of our input trees have non-overlapping taxon sets, and so conflict resolution is not the primary problem.
-Instead, seek to merge and aggregate phylogenetic information on the largest possible scale.
+Instead, we seek to merge and aggregate phylogenetic information on the largest possible scale.
 Therefore, the primary aim of our supertree method is to \emph{summarize and represent} the input trees, even when they conflict with each other.
-The annotations file constructed by our pipeline is essential output because it includes information about conflict with our supertree as well as support for it.
+The annotations file constructed by our pipeline is an essential for this task because it includes information about conflict with our supertree as well as support for it.
 

### 2017-01-23 doc/paper1-submission/paper5.tex
Add section on MRP and MinCut.

@@
 \thispagestyle{empty}
-
-% TODO: add stuff to paper
-% R1.2 Mention that applying ranks to trees instead of nodes is problematic.
-%      Mention transparency?
-% R2.1: Add paragraph mentioning MRP, MinCut, and how they are different. - 1.1.6
 
@@
 
-We also consider matrix representation with parsimony (MRP) and MinCut.
-MRP can display edges not found
-in any input tree, and so fails criterion \#1.  MRP is used in many
-studies that attempt to infer the true tree from a lot of input
-trees.  To some extent, MRP tries to treat each tree as data, which
-differs from our approach here.
-MinCut (and Modified MinCut) attempts to favor relationships present
-in multiple input trees during conflict resolution.
-Both MRP and MinCut attempt to adjudicate conflicts in a way that
-favors relationships present in more input trees.  Thus, they perform
-something more like a weighted average.  However, the reason the
-supertree resolves conflicts in a particular way is not clear and
-transparent, and does not allow a method for curators to adjust the
-supertree.
-
-A number of other supertree methods exist, including Matrix
-Representation with Parsimony (MRP) and MinCut.  MRP
-% http://currents.plos.org/treeoflife/article/reweaving-the-tapestry-a-supertree-of-birds/
-MRP - displays unsupported groups (\#1), not rank-based (\#2), ?
-SuperFine - computes a strict consensus, then MRP
-Minimum Flip - 
-(Modified) MinCut -
+Matrix Representation Parsimony (MRP) is one of the
+most widely used supertree approaches \citep{baum1992combining,BinindaEmonds2007b,Davis2014}.
+MRP attempts to encode branches of input trees as characters in a data
+matrix, and then find an optimal tree according to the maximum
+parsimony criterion. Such an approach does not provide a clear and transparent
+explanation for including particular edges in the supertree.
+MRP can also infer edges that exist in none of the input trees, which 
+violates our Goal \#1 \citep{gatesy2004critique}.
+
+Another commonly used method is MinCut \citep{semple2000supertree} or
+Modified MinCut \citep{page2002modified}.  Like our approach, MinCut
+is based on the BUILD algorithm \citep{AhoSSU1981}.  However, MinCut deals
+with conflicts by modifying the BUILD algorithm to resolve
+incompatibilities by discarding edges that are present in the smallest 
+number of input trees.  This approach thus violates our Goal \#2 of
+resolving conflict via ranks that can be altered by a curator to
+influence the output tree.
 

### 2017-01-24 doc/paper1-submission/response.tex
Update.

@@
 
-Dear Dr. XXX,
+Dear Dr. Wilke,
 
-We thank the reviewers for their comments.  In response their suggestions, we performed
-two experiments designed to check the reliability of our reconstruction, and to assess
-the sensitivity of our method to the given ranks.
+We appreciate the suggestions of Dr. Nakleh and the anonymous
+reviewer. In response to their suggestions, we performed two
+additional experiments on our pipeline to (1) ensure that the pipeline
+could accurately recover a tree from its constituent pieces, and (2)
+assessing the effect of input tree ranks on the supertree.
+
+We also rewrote our section on notation and terminology, added a
+discussion of other supertree methods, and clarified how our method
+approaches tree weights differently than MRP and MinCut.
+
+The first reviewer did not break their comments down into discrete
+sub-points and so our reply is also in a somewhat discursive style.
+However, we have responded to specific points as summarized by us.
 
@@
 
-We add the following text to the Background clarify that we are not trying to infer the true tree from the input trees:
+We suspect a lot of readers may have the same question.
+We therefore add the following text to the Background to clarify that we are not trying to infer the true tree from the input trees, and why not:
 \begin{quote}
-Although other researchers construct supertrees for the purpose of
-inferring the true tree from the input trees \citep{BinindaEmonds2007b,Davis2014},
-we note that phylogeny estimation is an explicit non-goal of our
-supertree method.  We do not treat input trees as data or as surrogates for
-data matrices \citep{gatesy2004critique}, and we do not claim that our
-summary supertree is more accurate than the input trees because more
-data lies behind it. Instead we seek to \emph{summarize and represent}
-the input trees, even when they conflict with each other.  The
-annotations file constructed by our pipeline is essential for this
-task because it includes information about conflict with our supertree
-as well as support for it.
+\textbf{Non-goal: phylogeny estimation}
+
+Many researchers construct supertrees for the purpose of inferring the true tree from the input trees \citep{BinindaEmonds2007b,Davis2014}.
+Such approaches treat input trees as data or as surrogates for data matrices \citep{gatesy2004critique}, and the supertree is seen as more accurate than the individual input trees because more input trees (and presumably more data) stand behind it.
+For such methods, conflict resolution is a primary aim of the method.
+
+Our supertree method operates in a different paradigm.
+Phylogeny estimation is an explicit non-goal of our supertree algorithm.
+We do not claim that the supertree is more accurate than the input trees it summarizes, and so conflict resolution is not a primary aim of the method.
+Most of our input trees have non-overlapping taxon sets, and so conflict resolution is not the primary problem.
+Instead, we seek to merge and aggregate phylogenetic information on the largest possible scale.
+Therefore, the primary aim of our supertree method is to \emph{summarize and represent} the input trees, even when they conflict with each other.
+The annotations file constructed by our pipeline is essential for this task because it includes information about conflict with our supertree as well as support for it.
+
+Our method must resolve conflicts in order to construct a single supertree.
+However, the rank information used to resolve conflicts is an input to the method, not an output from the method.
+We thus perform curation-based conflict resolution, not inference-based conflict resolution.
 \end{quote}
@@
 The reviewer seems concerned that low-quality source trees will undermine the estimation, and specifically that relationships present in more trees will dominate, leading to an incorrect estimate of the true tree.
-However: (1) our supertree approach is not designed to estimate the true tree, but to represent the input trees even if they are incorrect; (2) we handle conflict by allowing curators to influence the supertree via ranking as discussed in the text.
-Our conflict resolution is not designed to infer the tree.
+
+However we are not seeking to estimate the true tree, and our rank-based approach to conflict resolution doesn't favor relationships that are more frequent:
 
@@
 
-Note that in our rank-based method of conflict-resolution, ``traditional'' relationships are not favored in conflict resolution, even if they do occur in more input trees, because our supertree algorithm does not favor relationships that occur more often.
-
 % We probably do need to state that, unlike MRP and MinCut, our approach doesn't favor relationships that occur more often.
 
-2b.
-The reviewer seems concerned that our tree weights are extreme, and that it is difficult to estimate good tree weights from the input trees.
-While it is possible to recast our method as an optimization problem by assigning weights to the input trees, we are not trying to weight the evidence in the input trees, and we do not actually ever calculate these weights in our pipeline.
+2b. The reviewer seems concerned that our tree weights are extreme, and that it is difficult to estimate good tree weights from the input trees.
 
-We agree that the ranks have a large impact on conflict resolution in the supertree algorithm.
-If this were not the case, then ranks could not achieve their stated goal of allowing curators to fix problems in the tree.
-If we were performing phylogeny estimation, then these ranks might not give low-ranked trees sufficient weight in conflict resolution.
-However, again, we do not seek to infer the correct tree by treating the input trees as data.
+We appreciate these concerns and add the following text to explain our approach more clearly:
 
-%Our approach is somewhat novel in that we replace a kind of pseudo-estimation framework with a curation-based framework in which curators can influence conflict resolution by specifying high ranks for trees containing a desired relationship:
+\begin{quote}
+These large weights help to achieve our goal of allowing curators to fix problems in the tree.
+While it is possible to view our pipeline as an optimization algorithm by assigning weights to the input trees in this way, we are not trying to weight the evidence in the input trees, and we do not actually ever calculate these weights in our pipeline.
+\end{quote}
+
+We sidestep the problem of estimating good tree weights (which is indeed difficult, if not impossible) by relying on curators to impose rankings instead.
 
@@
 \begin{quote}
- say that applying ranks to trees is problematic.
+However, one downside of our current approach is that every node in an input
+tree must recieve the same rank.  As a result, bootstrap proportions
+and posterior probabilities are not taken into account.
 \end{quote}
 
-2e. Explaining the Hinchliff (2015) paper is beyond the scope of the current manuscript.
+2e. We think that explaining the Hinchliff (2015) paper is beyond the scope of the current manuscript.
 \end{reply}
-%2.
-It seems that the reviewer thinks that we are trying to infer the true tree from a dataset consisting of the input trees, like Bininda-Edmonds.
-Such an inference would need a principled method of resolving conflicts between input trees using only the information in the set of input trees.
-The resulting supertree could then be used to critique the input trees, so that input trees that don't match the supertree would be considered to be incorrect (up to estimation error).
-
-%However, our approach is fundamentally different.
-We do not seek to infer a tree that is more accurate than any individual input tree.
-Instead we seek to summarize and describe the set of input trees.
-While the supertree must conflict with some input trees, we do not claim that this constitutes an estimate of which input trees are correct or incorrect.
-Instead we seek to represent all input trees (including those that conflict with the phylogeny estimate) by linking from summary tree edges to conflicting input tree edges via our annotations file.
-
-% Summary: argh.
 
@@
 \begin{reply}
-Response 3.
+Since we are not trying to construct a phylogeny estimate like Bininda-Edmonds, the question of whether our tree is ``ready for prime time'' may not be the right question.
+However, since this may be the question of many readers, we add the following paragraph to the Conclusions to clarify that we are not claiming superior accuracy for our supertree:
 
-While we believe that our tree is useful, we are do not claim that it is more accurate that the phylo-inputs that are used to construct it.
-While we hope to summarize currently available phylogenetic knowledge, we are not trying to produce such knowledge like Bininda-Edmonds.
-
-The tree cannot be used directly for comparative methods, because it does not have branch lengths.
+\begin{quote}
+Our supertree pipeline makes it possible to summarize and merge
+information from across the Tree of Life.  However, we do not claim
+that our supertree is more accurate than the input trees that it
+summarizes.  Before using the output of our supertree pipeline
+as input to an evolutionary analysis, researchers should first assess
+the accuracy of the output supertree in the clade of interest,
+possibly adjusting the ranks of the input trees to produce a more
+useful outcome.  We have demonstrated that ranks supplied by curators
+have substantial effects on the output of the pipeline.  This
+can be considered to be a benefit of our approach, since it allows
+curators to adjust the output of the pipeline, and it is easy to see
+why the supertree contains the relationships that it does. However,
+one downside of our current approach is that every node in an input
+tree must recieve the same rank.  As a result, bootstrap proportions
+and posterior probabilities are not taken into account.
+\end{quote}
 \end{reply}
@@
 
+\begin{reply}
+We add the following two paragraphs to explain why we do not use MRP or MinCut, two commonly-used supertree methods.
+\begin{quote}
+\textbf{Comparison to other supertree methods}
+
+Matrix Representation Parsimony (MRP) is one of the
+most widely used supertree approaches \citep{baum1992combining,BinindaEmonds2007b,Davis2014}.
+MRP attempts to encode branches of input trees as characters in a data
+matrix, and then find an optimal tree according to the maximum
+parsimony criterion. Such an approach does not provide a clear and transparent
+explanation for including particular edges in the supertree.
+MRP can also infer edges that exist in none of the input trees, which 
+violates our Goal \#1 \citep{gatesy2004critique}.
+
+Another commonly used method is MinCut \citep{semple2000supertree} or
+Modified MinCut \citep{page2002modified}.  Like our approach, MinCut
+is based on the BUILD algorithm \citep{AhoSSU1981}.  However, MinCut deals
+with conflicts by modifying the BUILD algorithm to resolve
+incompatibilities by discarding edges that are present in the smallest 
+number of input trees.  This approach thus violates our Goal \#2 of
+resolving conflict via ranks that can be altered by a curator to
+influence the output tree.
+\end{quote}
+\end{reply}
+
 2.
@@
 \begin{reply}
-Thanks for this suggestion.  We performed this experiment and added the following section describing the experiment and the results: \textcolor{red}{\textbf{Mark please check}}
+Thanks for this suggestion.  We performed this experiment and added the following section describing the experiment and the results:
 \begin{quote}
-To test the reliability of our methods, we divided the OTLv4' summary
+To test the reliability of our methods, we divided the \otlprop{} summary
 tree into a collection of input trees. The input trees were obtained
@@
 tree from these input trees using the propinquity pipeline.  The
-resulting summary tree was identical to the OTLV4' summary tree.  
+resulting summary tree was identical to the \otlprop{} summary tree.
 \end{quote}

### 2017-01-24 doc/paper1-submission/paper5.tex
Update.

@@
 
-Cutting any edge \textbf{$j$ }of a rooted tree induces a bi-partition
+Cutting any edge $j$ of a rooted tree induces a bi-partition
 $S(j)$ of the leaf taxa into two connected groups $S_{1}(j)$ and
@@
 Therefore, the primary aim of our supertree method is to \emph{summarize and represent} the input trees, even when they conflict with each other.
-The annotations file constructed by our pipeline is an essential for this task because it includes information about conflict with our supertree as well as support for it.
+The annotations file constructed by our pipeline is essential for this task because it includes information about conflict with our supertree as well as support for it.
 
@@
 We ran the pipeline on the OTL4 data with the ranks reversed.
-This drastic alteration of the ranks lead to a summary supertree that conflicted with 1264 branches of the \otlprop{}.
+This drastic alteration of the ranks led to a summary supertree that conflicted with 1264 branches of the \otlprop{}.
 Compared to this, the \otl{} supertree tree conflicts with 1387 branches of the \otlprop{} supertree.
@@
 Our supertree pipeline makes it possible to summarize and merge
-information from across the tree of life.  However, we do not claim
-that our supertree is more accurate than the input trees that is
-summarizes.  We demonstrate that ranks supplied by curators have
-substantial effects on the output of the pipeline.  Currently every
-node in an input tree must recieve the same rank, and bootstrap
-proportions and posterior probabilities are not taken into account.
-Therefore, before using the output of our supertree pipeiline in an
-evolutionary analysis, researchers must first assess the accuracy of
-the output supertree, possibly adjusting the ranks of the input trees
-to produce a more useful outcome.
+information from across the Tree of Life.  However, we do not claim
+that our supertree is more accurate than the input trees that it
+summarizes.  Before using the output of our supertree pipeline
+as input to an evolutionary analysis, researchers should first assess
+the accuracy of the output supertree in the clade of interest,
+possibly adjusting the ranks of the input trees to produce a more
+useful outcome.  We have demonstrated that ranks supplied by curators
+have substantial effects on the output of the pipeline.  This
+can be considered to be a benefit of our approach, since it allows
+curators to adjust the output of the pipeline, and it is easy to see
+why the supertree contains the relationships that it does.  However,
+one downside of our current approach is that every node in an input
+tree must recieve the same rank.  As a result, bootstrap proportions
+and posterior probabilities are not taken into account.
 

### 2017-01-25 doc/paper1-submission/paper5.tex
Number figures in order of reference.

@@
 if its inclusion did not increase the total number of input nodes
-that are displayed by the summary tree. Figure (\ref{fig:pairwisecompat})
+that are displayed by the summary tree. Figure \ref{fig:pairwisecompat}
 shows an example of 3 input trees for which there is no pairwise incompatibility,
@@
 taxon is the union on the set of descendant terminal taxa that have
-been sampled in a phylogenetic input. Figure \ref{fig:exemplify-example}
-shows an example of how the trees in figure \ref{fig:cleaned-phylo-example}
-would be exemplified.
+been sampled in a phylogenetic input. For example, the trees in Figure
+\ref{fig:cleaned-phylo-example} would exemplified as shown in Figure
+\ref{fig:exemplify-example}.
 
@@
 to adding any unsampled taxonomic children to the node that represents
-the taxon in the grafted solution tree.
+the taxon in the grafted solution tree.  For example, the tree in
+Figure \ref{fig:grafted-solution} is unpruned to produce the tree in
+Figure \ref{fig:unprune}.
 
@@
 in the taxonomic tree.
+
+\begin{figure}
+
+\begin{centering}
+\includegraphics[scale=0.5]{Figures__labelled_supertree__labelled_supertree_only_taxon_names\lyxdot tre}
+\par\end{centering}
+\caption{\label{fig:unprune}Unpruned tree with internal taxa. Taxa unsampled in phylogenetic
+statements have been added to the grafted tree shown
+in figure \ref{fig:grafted-solution}.}
+\end{figure}
 
@@
 R3. Our method follows the second approach.}
-\end{figure}
-
-\begin{figure}
-\begin{centering}
-\includegraphics[scale=0.5]{Figures__labelled_supertree__labelled_supertree_only_taxon_names\lyxdot tre}
-\par\end{centering}
-\caption{Unpruned tree with internal taxa. Taxa unsampled in phylogenetic
-statements have been added to the grafted tree shown
-in figure \ref{fig:grafted-solution}.}
 \end{figure}

### 2017-01-25 doc/paper1-submission/paper5.tex
Standardize on Fig. #

@@
 must be excluded in order to display the node. Consider the problem
-shown in figure \ref{fig:toyambig}; panels (\ref{fig:toyambig}a)
-and (\ref{fig:toyambig}b) show two input trees. Because taxa A and
+shown in Fig. \ref{fig:toyambig}; panels \ref{fig:toyambig}a
+and \ref{fig:toyambig}b show two input trees. Because taxa A and
 E do not occur together in either input, there is some uncertainty
 about where to place them. By our terminology, either output shown
-in (\ref{fig:toyambig}c) or (\ref{fig:toyambig}d) would be characterized
+in Fig. \ref{fig:toyambig}c or Fig. \ref{fig:toyambig}d would be characterized
 as a tree that displays all of the input groupings and which has no
@@
 it may be possible for some taxa to ``float'' to multiple positions.
-In figure \ref{fig:toyambig}, taxon E floats to different positions
-in (\ref{fig:toyambig}c) and (\ref{fig:toyambig}d), whereas taxon
+In Fig. \ref{fig:toyambig}, taxon E floats to different positions
+in \ref{fig:toyambig}c and \ref{fig:toyambig}d, whereas taxon
 A does not.
@@
 of the groupings in the input trees and which have no unsupported
-nodes. The BUILD algorithm (section \ref{fig:subproblem-solutions.})
+nodes. The BUILD algorithm (section \ref{subsec:Subproblem-solution})
 would choose tree (d) that floats taxon E closer to the root.}
@@
 if its inclusion did not increase the total number of input nodes
-that are displayed by the summary tree. Figure \ref{fig:pairwisecompat}
+that are displayed by the summary tree. Fig. \ref{fig:pairwisecompat}
 shows an example of 3 input trees for which there is no pairwise incompatibility,
@@
 of inputs can result in one of three summary trees shown in panels
-(d-f).
+d-f.
 
@@
 leads to some ``necessary'' polytomies. For example any resolution
-of the polytomy shown in figure (\ref{fig:pairwisecompat}e) would
+of the polytomy shown in Fig. \ref{fig:pairwisecompat}e would
 continue to display the same two input groups. However, the additional
@@
 unresolved tree would be preferred by our criterion.
-However, collapsing either internal edge of the tree shown in figure
-(\ref{fig:pairwisecompat}d) would result in a tree which displays
+However, collapsing either internal edge of the tree shown in Fig.
+\ref{fig:pairwisecompat}d would result in a tree which displays
 only one input grouping. This tree would contain an unnecessary polytomy,
@@
 pipeline manipulate the phylogenetic input trees to improve their
-usability and reliability. The first steps of the pipeline (see Figure
+usability and reliability. The first steps of the pipeline (see Fig.
 \ref{fig:pipeline}) collect a list of trees to include (in the \texttt{phylo\_input}
@@
 taxon is the union on the set of descendant terminal taxa that have
-been sampled in a phylogenetic input. For example, the trees in Figure
-\ref{fig:cleaned-phylo-example} would exemplified as shown in Figure
+been sampled in a phylogenetic input. For example, the trees in Fig.
+\ref{fig:cleaned-phylo-example} would exemplified as shown in Fig.
 \ref{fig:exemplify-example}.
@@
 
-\caption{Exemplified input trees and taxonomy tree from figure \ref{fig:cleaned-phylo-example}.
+\caption{Exemplified input trees and taxonomy tree from Fig. \ref{fig:cleaned-phylo-example}.
 E in tree1 is exemplified by E1. Pruned taxa are E2, F2, and D. The
@@
 root of its own subproblem and it will be tip in the subproblem that
-covers the next slice deeper in the tree. The red edge in figure \ref{fig:exemplify-example}c
-highlights the taxa that are not contested by the input shown in \ref{fig:exemplify-example};
-figure \ref{fig:subproblems-example} shows the subproblems that would
+covers the next slice deeper in the tree. The red edge in Fig. \ref{fig:exemplify-example}c
+highlights the taxa that are not contested by the input shown in Fig. \ref{fig:exemplify-example};
+Fig. \ref{fig:subproblems-example} shows the subproblems that would
 be emitted as a result of this set of inputs. The supertree operation
@@
 allow us to find the tree that maximizes the MSDGWS score. For example,
-see figure \ref{fig:decompose-worsens}; that example is a variant
-of the situation shown in figure \ref{fig:pairwisecompat}. In this
+see Fig. \ref{fig:decompose-worsens}; that example is a variant
+of the situation shown in Fig. \ref{fig:pairwisecompat}. In this
 case the groupings from each of the phylogenetic estimates, shown
@@
 
-\caption{Subproblems generated from the exemplified trees shown in figure \ref{fig:exemplify-example}.
+\caption{Subproblems generated from the exemplified trees shown in Fig. \ref{fig:exemplify-example}.
 A trivial statement from the first tree that a taxon labelled $ABCD$
@@
 from conflicting splits by attaching taxa from conflicting splits
-at their common ancestor. Figure \ref{fig:subproblem-solutions.}
+at their common ancestor. Fig. \ref{fig:subproblem-solutions.}
 shows the solutions that would be obtained by applying our modified
-version of the BUILD algorithm to the subproblems shown in figure \ref{fig:subproblems-example}.
+version of the BUILD algorithm to the subproblems shown in Fig. \ref{fig:subproblems-example}.
 
@@
 
-\caption{Solutions to the subproblems depicted in figure \ref{fig:subproblems-example}}
+\caption{Solutions to the subproblems depicted in Fig. \ref{fig:subproblems-example}}
 \label{fig:subproblem-solutions.}
@@
 tree (stored as the \texttt{grafted\_solution/grafted\_solution.tre}
-by propinquity); see figure \ref{fig:grafted-solution}. Recall that each non-root uncontested taxon used
+by propinquity); see Fig. \ref{fig:grafted-solution}. Recall that each non-root uncontested taxon used
 for decomposition occurs as a leaf taxon in one subproblem and as
@@
 \caption{Grafted solution produced from the subproblems from
-figure \ref{fig:subproblem-solutions.} and which is the 
+Fig. \ref{fig:subproblem-solutions.} and which is the 
 backbone onto which taxa that are not included in any phylogeny
@@
 would be monotypic taxa (with children E1 and F1, respectively)
-from the trivial subproblems that were omitted from from figure \ref{fig:subproblems-example}.}
+from the trivial subproblems that were omitted from from Fig. \ref{fig:subproblems-example}.}
 \label{fig:grafted-solution}
@@
 the taxon in the grafted solution tree.  For example, the tree in
-Figure \ref{fig:grafted-solution} is unpruned to produce the tree in
-Figure \ref{fig:unprune}.
+Fig. \ref{fig:grafted-solution} is unpruned to produce the tree in
+Fig. \ref{fig:unprune}.
 
@@
 
-Figure \ref{fig:Two-approaches-to-unpruning} illustrates the two
+Fig. \ref{fig:Two-approaches-to-unpruning} illustrates the two
 approaches to unpruning. Taxa G, M, and R (Fig \ref{fig:Two-approaches-to-unpruning}a)
@@
 the unsampled children (R4, R5, and R6) to attached directly at taxon
-N (as in the tree shown in Fig \ref{fig:Two-approaches-to-unpruning}c),
+N (as in the tree shown in Fig. \ref{fig:Two-approaches-to-unpruning}c),
 because that is the least inclusive unbroken ancestor of R. The tree
-illustrated in Fig \ref{fig:Two-approaches-to-unpruning}d shows the
+illustrated in Fig. \ref{fig:Two-approaches-to-unpruning}d shows the
 tree that would be produced by propinquity; the children of the broken
@@
 statements have been added to the grafted tree shown
-in figure \ref{fig:grafted-solution}.}
+in Fig. \ref{fig:grafted-solution}.}
 \end{figure}
@@
 \end{itemize}
-These three relationships are illustrated in Figure \ref{fig:displayed-annotations}.
+These three relationships are illustrated in Fig. \ref{fig:displayed-annotations}.
 Finally, if $T_{i}$ does not display $x$ from $\mathcal{\mathbb{S}}(i)$,

### 2017-01-25 doc/paper1-submission/paper5-adjusted.tex
Convert figures 1-7

--- 2017-01-25 doc/paper1-submission/paper5.tex
@@
 \begin{figure}
-\subfloat[Input tree 1]{\includegraphics[width=0.45\textwidth]{Figures__toy_ambig__abcd}
-
-}\hfill{}\subfloat[Input tree 2]{\includegraphics[width=0.45\textwidth]{Figures__toy_ambig__bcde}
+\subfloat[Input tree 1]{\includegraphics[width=0.45\textwidth]{Fig1a}
+
+}\hfill{}\subfloat[Input tree 2]{\includegraphics[width=0.45\textwidth]{Fig1b}
 
@@
 
-\subfloat[Summary tree]{\includegraphics[width=0.45\textwidth]{Figures__toy_ambig__soln1}
-
-}\hfill{}\subfloat[Summary tree]{\includegraphics[width=0.45\textwidth]{Figures__toy_ambig__soln2}
+\subfloat[Summary tree]{\includegraphics[width=0.45\textwidth]{Fig1c}
+
+}\hfill{}\subfloat[Summary tree]{\includegraphics[width=0.45\textwidth]{Fig1d}
 
@@
 \begin{figure}
-\subfloat[Input tree 1]{\includegraphics[scale=0.57]{Figures__toy_pairwise_compat__abc}
-
-}\hfill{}\subfloat[Input tree 2]{\includegraphics[scale=0.57]{Figures__toy_pairwise_compat__acd}
-
-}\hfill{}\subfloat[Input tree 3]{\includegraphics[scale=0.57]{Figures__toy_pairwise_compat__bcd}
+\subfloat[Input tree 1]{\includegraphics[scale=0.57]{Fig2a}
+
+}\hfill{}\subfloat[Input tree 2]{\includegraphics[scale=0.57]{Fig2b}
+
+}\hfill{}\subfloat[Input tree 3]{\includegraphics[scale=0.57]{Fig2c}
 
@@
 
-\subfloat[Summary tree]{\includegraphics[scale=0.57]{Figures__toy_pairwise_compat__s12}
-
-}\hfill{}\subfloat[Summary tree]{\includegraphics[scale=0.57]{Figures__toy_pairwise_compat__s13}
-
-}\hfill{}\subfloat[Summary tree]{\includegraphics[scale=0.57]{Figures__toy_pairwise_compat__s23}
+\subfloat[Summary tree]{\includegraphics[scale=0.57]{Fig2d}
+
+}\hfill{}\subfloat[Summary tree]{\includegraphics[scale=0.57]{Fig2e}
+
+}\hfill{}\subfloat[Summary tree]{\includegraphics[scale=0.57]{Fig2f}
 
@@
 \begin{centering}
-\includegraphics[height=0.70\paperheight]{pipeline-tools.pdf}
+\includegraphics[height=0.70\paperheight]{Fig3.pdf}
 \par\end{centering}
@@
 \begin{figure}
-\subfloat[Tree 1]{\includegraphics[scale=0.4]{Figures__cleaned_phylo__ex_2@tree1\lyxdot tre}
-
-}\hfill{}\subfloat[Tree 2]{\includegraphics[scale=0.4]{Figures__cleaned_phylo__ex_2@tree2\lyxdot tre}
-
-}\hfill{}\subfloat[Taxonomy Tree]{\includegraphics[scale=0.4]{Figures__cleaned_ott__cleaned_ott\lyxdot tre}
+\subfloat[Tree 1]{\includegraphics[scale=0.4]{Fig4a}
+
+}\hfill{}\subfloat[Tree 2]{\includegraphics[scale=0.4]{Fig4b}
+
+}\hfill{}\subfloat[Taxonomy Tree]{\includegraphics[scale=0.4]{Fig4c}
 
@@
 \begin{figure}
-\subfloat[Tree 1]{\includegraphics[scale=0.4]{Figures__exemplified_phylo__ex_2@tree1\lyxdot tre}
-
-}\hfill{}\subfloat[Tree 2]{\includegraphics[scale=0.4]{Figures__exemplified_phylo__ex_2@tree2\lyxdot tre}
-
-}\hfill{}\subfloat[Pruned taxonomy tree $\protect\prunedTaxonomy$]{\includegraphics[scale=0.4]{Figures__exemplified_phylo_taxonomy_uncontested\lyxdot tre}
+\subfloat[Tree 1]{\includegraphics[scale=0.4]{Fig5a}
+
+}\hfill{}\subfloat[Tree 2]{\includegraphics[scale=0.4]{Fig5b}
+
+}\hfill{}\subfloat[Pruned taxonomy tree $\protect\prunedTaxonomy$]{\includegraphics[scale=0.4]{Fig5c}
 
@@
 \begin{figure}
-  \subfloat[Subproblem \emph{ABCD}]{\includegraphics[width=0.3\textwidth]{Figures__subproblems__14__1\lyxdot tre}%
-    \hfill{}\includegraphics[width=0.3\textwidth]{Figures__subproblems__14__2\lyxdot tre}%
-    \hfill{}\includegraphics[width=0.3\textwidth]{Figures__subproblems__14__3\lyxdot tre}
+  \subfloat[Subproblem \emph{ABCD}]{\includegraphics[width=0.3\textwidth]{Fig6a}%
+    \hfill{}\includegraphics[width=0.3\textwidth]{Fig6b}%
+    \hfill{}\includegraphics[width=0.3\textwidth]{Fig6c}
 
@@
 
-  \subfloat[Subproblem \emph{root}]{\includegraphics[width=0.3\textwidth]{Figures__subproblems__17__1\lyxdot tre}%
-    \hfill{}\includegraphics[width=0.3\textwidth]{Figures__subproblems__17__2\lyxdot tre}%
-    \hfill{}\includegraphics[width=0.3\textwidth]{Figures__subproblems__17__3\lyxdot tre}
+  \subfloat[Subproblem \emph{root}]{\includegraphics[width=0.3\textwidth]{Fig6d}%
+    \hfill{}\includegraphics[width=0.3\textwidth]{Fig6e}%
+    \hfill{}\includegraphics[width=0.3\textwidth]{Fig6f}
 
@@
 \begin{figure}
-\subfloat[]{\includegraphics[scale=0.57]{Figures__uncontested_worsens_score__ab1c}
-
-}\hfill{}\subfloat[]{\includegraphics[scale=0.57]{Figures__uncontested_worsens_score__acb2}
-
-}\hfill{}\subfloat[]{\includegraphics[scale=0.57]{Figures__uncontested_worsens_score__tax}
+\subfloat[]{\includegraphics[scale=0.57]{Fig7a}
+
+}\hfill{}\subfloat[]{\includegraphics[scale=0.57]{Fig7b}
+
+}\hfill{}\subfloat[]{\includegraphics[scale=0.57]{Fig7c}
 
@@
 
-\hfill{}\subfloat[]{\includegraphics[scale=0.57]{Figures__uncontested_worsens_score__opt}
-
-}\hfill{}\subfloat[]{\includegraphics[scale=0.57]{Figures__uncontested_worsens_score__returned}
+\hfill{}\subfloat[]{\includegraphics[scale=0.57]{Fig7d}
+
+}\hfill{}\subfloat[]{\includegraphics[scale=0.57]{Fig7e}
 

### 2017-01-25 doc/paper1-submission/paper5-adjusted.tex
Convert Algorithm to Figure.

@@
 trees in order of ranking, retaining splits that are compatible with
-the current set of splits (Alg \ref{alg:ConsistentSplitsFromRankedList}).
+the current set of splits (Fig. \ref{alg:ConsistentSplitsFromRankedList}).
 The order of splits from the same tree is not specified by this approach,
@@
 trees from the BUILD algorithm do not contain unsupported branches.
-\begin{algorithm}
-\begin{algorithmic}
-\Require An ordered list of $M$ splits, $\mathcal{R} = [R_1, R_2, R_3, \ldots, R3, \ldots, R_M]$
-\State $\mathcal{C} = [R_1]$
-\For{each split $i$ in $[2, 3 \ldots M]$}
-   \State $\mathcal{T} \leftarrow \mathcal{C} + R_i$ \Comment{where `+' means concatenating 2 lists}
-   \If{\textsc{BUILD}$(\mathcal{T})$ does not return null}
-      \State $\mathcal{C} \leftarrow \mathcal{T}$
-   \EndIf
-\EndFor{}
-\State\Return $\mathcal{C}$
-\end{algorithmic}
-
-\caption{\label{alg:ConsistentSplitsFromRankedList}ConsistentSplitsFromRankedList}
-\end{algorithm}
+\begin{figure}
+\includegraphics[width=0.9\textwidth]{fig8}
+\caption{\label{alg:ConsistentSplitsFromRankedList}Algorithm \texttt{ConsistentSplitsFromRankedList}}
+\end{figure}
 
@@
 should be attached to the grafted solution. One approach would be
-to mimic the application of Algorithm \ref{alg:ConsistentSplitsFromRankedList}
+to mimic the application of Algorithm $\texttt{ConsistentSplitsFromRankedList}$ (Fig. \ref{alg:ConsistentSplitsFromRankedList})
 to the full (unpruned) taxonomic tree. This would be equivalent to

### 2017-01-25 doc/paper1-submission/paper5-adjusted.tex
Convert Figures 9 and 10.

@@
 \begin{figure}
-\hfill{}\subfloat[Solution to subproblem \emph{ABCD}]{\includegraphics[scale=0.55]{Figures__subproblem_solutions__ott14\lyxdot tre}
-
-}\hfill{}\subfloat[Solution to subproblem \emph{root}]{\includegraphics[scale=0.55]{Figures__subproblem_solutions__ott17\lyxdot tre}
+\hfill{}\subfloat[Solution to subproblem \emph{ABCD}]{\includegraphics[scale=0.55]{Fig9a}
+
+}\hfill{}\subfloat[Solution to subproblem \emph{root}]{\includegraphics[scale=0.55]{Fig9b}
 
@@
 \begin{centering}
-\includegraphics[scale=0.5]{Figures__grafted_solution__grafted_solution_ottnames\lyxdot tre}
+\includegraphics[scale=0.5]{Fig10.pdf}
 \par\end{centering}

### 2017-01-25 doc/paper1-submission/paper5-adjusted.tex
Convert figures 11-13

@@
 \begin{centering}
-\includegraphics[scale=0.5]{Figures__labelled_supertree__labelled_supertree_only_taxon_names\lyxdot tre}
+\includegraphics[scale=0.5]{Fig11}
 \par\end{centering}
@@
 \begin{center}
-\includegraphics[scale=0.38]{example3__cleaned_ott-labels\lyxdot tre}
+\includegraphics[scale=0.38]{Fig12a}
 \par\end{center}%
@@
 \begin{center}
-\includegraphics[scale=0.38]{example3__grafted_solution-labels\lyxdot tre}
+\includegraphics[scale=0.38]{Fig12b}
 \par\end{center}%
@@
 \subfloat[Unpruned tree - broken taxa removed]{\begin{centering}
-\includegraphics[scale=0.38]{example3__unprune-old-labels\lyxdot tre}
+\includegraphics[scale=0.38]{Fig12c}
 \par\end{centering}
 }\hfill{}\subfloat[Unpruned tree with MRCA-attachment]{\begin{centering}
-\includegraphics[scale=0.38]{example3__unprune-new-labels2\lyxdot tre}
+\includegraphics[scale=0.38]{Fig12d}
 \par\end{centering}
@@
 \begin{minipage}[c][1\totalheight][t]{0.25\columnwidth}%
-\includegraphics[scale=0.5]{annotation_example__input1\lyxdot tre}%
+\includegraphics[scale=0.5]{Fig13a}%
 \end{minipage}
@@
 \begin{minipage}[c][1\totalheight][t]{0.35\columnwidth}%
-\includegraphics[scale=0.5]{annotation_example__summary\lyxdot tre}%
+\includegraphics[scale=0.5]{Fig13b}%
 \end{minipage}
@@
 \begin{minipage}[c][1\totalheight][t]{0.35\columnwidth}%
-\includegraphics[scale=0.5]{annotation_example__induced1_color}%
+\includegraphics[scale=0.5]{Fig13c}%
 \end{minipage}
@@
 \hfill{}\subfloat[JSON annotations relating edges of (c) to edges of (a)]{%
-\shadowbox{\begin{minipage}[t]{1.0\columnwidth}%
-\texttt{\{}
-
-\texttt{~~\textquotedbl{}nodes\textquotedbl{}: \{}
-
-\texttt{~~~~\textquotedbl{}mrcaott1ott4\textquotedbl{}: \{ \textquotedbl{}}\texttt{\textcolor{blue}{partial\_path\_of}}\texttt{\textquotedbl{}: \{
-\textquotedbl{}tree1\textquotedbl{}: \textquotedbl{}node15\textquotedbl{},
-... \} \},}
-
-\texttt{~~~~\textquotedbl{}ott10\textquotedbl{}:~~~~~~~~~\{
-\textquotedbl{}}\texttt{\textcolor{blue}{partial\_path\_of}}\texttt{\textquotedbl{}: \{
-\textquotedbl{}tree1\textquotedbl{}: \textquotedbl{}node15\textquotedbl{},
-... \} \},}
-
-\texttt{~~~~\textquotedbl{}mrcaott5ott7\textquotedbl{}: \{ \textquotedbl{}\textcolor{mylime}{supported\_by}\textquotedbl{}:~~~
-\{ \textquotedbl{}tree1\textquotedbl{}: \textquotedbl{}node16\textquotedbl{},
-... \} \},}
-
-\texttt{~~~~\textquotedbl{}ott1\textquotedbl{}:~~~~~~~~~~\{
-\textquotedbl{}}\texttt{\textcolor{orange}{terminal}}\texttt{\textquotedbl{}: ~~~~~~~\{
-\textquotedbl{}tree1\textquotedbl{}: \textquotedbl{}ott1\textquotedbl{},
-... ~~\} \},}
-
-\texttt{~~~~\textquotedbl{}ott2\textquotedbl{}:~~~~~~~~~~\{
-\textquotedbl{}}\texttt{\textcolor{orange}{terminal}}\texttt{\textquotedbl{}: ~~~~~~~\{
-\textquotedbl{}tree1\textquotedbl{}: \textquotedbl{}ott2\textquotedbl{},
-... ~~\} \},}
-
-\texttt{~~~~\textquotedbl{}ott5\textquotedbl{}:~~~~~~~~~~\{
-\textquotedbl{}}\texttt{\textcolor{orange}{terminal}}\texttt{\textquotedbl{}: ~~~~~~~\{
-\textquotedbl{}tree1\textquotedbl{}: \textquotedbl{}ott5\textquotedbl{},
-... ~~\} \},}
-
-\texttt{~~~~\textquotedbl{}ott11\textquotedbl{}:~~~~~~~~~\{
-\textquotedbl{}}\texttt{\textcolor{orange}{terminal}}\texttt{\textquotedbl{}: ~~~~~~~\{
-\textquotedbl{}tree1\textquotedbl{}: \textquotedbl{}ott5\textquotedbl{},
-... ~~\} \},}
-
-\texttt{~~~~\textquotedbl{}ott7\textquotedbl{}:~~~~~~~~~~\{
-\textquotedbl{}}\texttt{\textcolor{orange}{terminal}}\texttt{\textquotedbl{}: ~~~~~~~\{
-\textquotedbl{}tree1\textquotedbl{}: \textquotedbl{}ott7\textquotedbl{},
-... ~~\} \}}
-
-\texttt{~~\},}
-
-\texttt{~~...}
-
-\texttt{\} }%
-\end{minipage}}
-
+\includegraphics{Fig13d}%
 }\hfill{}

### 2017-01-26 doc/paper1-submission/paper5-adjusted.tex
Make references to panels more explicit.

@@
 
-\caption{Input trees and taxonomy tree}
+\caption{Input trees (a-b) and taxonomy tree (c).}
 \label{fig:cleaned-phylo-example}
@@
 
-\caption{Exemplified input trees and taxonomy tree from Fig. \ref{fig:cleaned-phylo-example}.
-E in tree1 is exemplified by E1. Pruned taxa are E2, F2, and D. The
+\caption{Exemplified input trees (a-b) and taxonomy tree (c) from Fig. \ref{fig:cleaned-phylo-example}.
+Taxon E in (a) is exemplified by E1. Pruned taxa are E2, F2, and D. The
 taxa E and F are retained as monotypic taxa in the pruned taxonomy
-$\protect\prunedTaxonomy$, but are not shown in panel c. The red
+$\protect\prunedTaxonomy$, but are not shown in panel (c). The red
 edge in the pruned taxonomy tree is an uncontested higher taxon in
@@
 
-\caption{Subproblems generated from the exemplified trees shown in Fig. \ref{fig:exemplify-example}.
+\caption{Subproblems \emph{ABCD} and \emph{root} generated from the exemplified trees shown in Fig. \ref{fig:exemplify-example}.
 A trivial statement from the first tree that a taxon labelled $ABCD$
@@
 
-\caption{Solutions to the subproblems depicted in Fig. \ref{fig:subproblems-example}}
+\caption{Solutions to subproblem \emph{ABCD} and subproblem \emph{root} depicted in Fig. \ref{fig:subproblems-example}.}
 \label{fig:subproblem-solutions.}

### 2017-01-26 doc/paper1-submission/Fig9.tex
Convert Fig9

--- 2017-01-25 doc/paper1-submission/paper5-adjusted.tex
@@
-\documentclass[fleqn,12pt,lineno,english]{wlpeerj}
+\documentclass[12pt,preview]{standalone}
+\usepackage[labelfont={bf,sf}]{caption}
+\usepackage[verbose,tmargin=1in,bmargin=1in,lmargin=1in,rmargin=1in]{geometry}
 \usepackage[T1]{fontenc}
-%\usepackage[latin9]{inputenc}
-%\usepackage{geometry}
-\geometry{verbose,tmargin=1in,bmargin=1in,lmargin=1in,rmargin=1in}
-%\usepackage{xcolor}
-%\usepackage{verbatim}
 \usepackage{float}
-\usepackage{fancybox}
-\usepackage{calc}
+\usepackage{graphicx}
+\usepackage{lmodern}
+\usepackage{subfig}
 \usepackage{amssymb}
-\usepackage{graphicx}
-\usepackage{natbib}
-\sloppy
-\makeatletter
-
-%\def\fps@figure{ht}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
-%% A simple dot to overcome graphicx limitations
-\newcommand{\lyxdot}{.}
-\newcommand{\otl}{OTLv4}
-\newcommand{\otlprop}{OTLv4$^\prime$}
-\newcommand{\supportedby}{\texttt{supported\_by}}
-\newcommand{\partialpathof}{\texttt{partial\_path\_of}}
-\newcommand{\terminal}{\texttt{terminal}}
-\newcommand{\conflictswith}{\texttt{conflicts\_with}}
-\newcommand{\resolves}{\texttt{resolves}}
-\newcommand{\resolvedby}{\texttt{resolved\_by}}
-
-\floatstyle{ruled}
-\newfloat{algorithm}{tbp}{loa}
-\providecommand{\algorithmname}{Algorithm}
-\floatname{algorithm}{\protect\algorithmname}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
-\usepackage{algorithm,algpseudocode}
-\usepackage{lmodern}
-\usepackage{xcolor}
-\usepackage{hyperref}
-\hypersetup{
-    colorlinks,
-    linkcolor={red!50!black},
-    citecolor={blue!50!black},
-    urlcolor={blue!80!black}
-}
-\definecolor{mylime}{RGB}{128,255,128}
-
-\@ifundefined{showcaptionsetup}{}{%
- \PassOptionsToPackage{caption=false}{subfig}}
-\usepackage{subfig}
-\makeatother
-
-\usepackage{babel}
+%\usepackage{subcaption}
 \global\long\def\taxonomy{\mbox{\ensuremath{\mathbb{T}}}}
@@
 \global\long\def\collections{\mathcal{C}}
-
-\title{A supertree pipeline for summarizing phylogenetic and taxonomic information
-for millions of species}
-
-\author[1,2]{Benjamin D. Redelings}
-\author[2,3,4]{Mark T. Holder}
-\affil[1]{Department of Biology, Duke University, Durham NC, US}
-\affil[2]{Department of Ecology and Evolutionary Biology, University of Kansas, Lawrence KS, US}
-\affil[3]{Biodiversity Institute, University of Kansas, Lawrence KS, US}
-\affil[4]{Heidelberg Institute for Theoretical Studies, Heidelberg, Germany}
-\corrauthor[2,3,4]{Mark T. Holder}{mtholder@ku.edu}
-
-% \keywords{Keyword1, Keyword2, Keyword3}
-
-\begin{abstract}
-We present a new supertree method that enables rapid construction of a
-summary tree on the scale of millions of leaves.  This supertree
-method summarizes a collection of input phylogenies and an input taxonomy.
-We introduce formal goals and criteria for such a supertree to satisfy
-in order to transparently and justifiably represent the input
-trees. In addition to producing a supertree, our method computes
-annotations that describe which grouping in the input trees support
-and conflict with each group in the supertree.
-
-\medskip
-
-We compare our supertree construction method to a previously published
-method by assessing their performance on input
-trees used to construct the Open Tree of Life version 4, and find that
-our method increases the number of displayed input splits from 35,518
-to 39,639 and decreases the number of conflicting input splits from
-2,760 to 1,357.  The new supertree method also improves on the previous
-supertree construction method in that it produces no unsupported
-branches and avoids unnecessary polytomies.
-
-\medskip
-
-This pipeline is currently used by the Open Tree of Life project to
-produce all of the versions of project's ``synthetic tree'' starting
-at version 5.This software pipeline is called
-\emph{``propinquity.''}. It relies heavily on \emph{``otcetera''} - a
-set of C++ tools to perform most of the steps of the pipeline. All of
-the components are free software and are available on GitHub.
-\end{abstract}
-
 \begin{document}
-
-\flushbottom
-\maketitle
-\thispagestyle{empty}
-
-\section{Background}
-
-The Open Tree of Life project seeks to build a platform for summarizing
-what is known about phylogenetic relationships across all of Life
-\citep{HinchliffEtAl2015}. One primary goal of the project is to
-build a summary tree from a comprehensive taxonomic tree and a set
-of published trees. The summary tree is intended to transparently
-and justifiably represent phylogenetic information from these inputs.
-The taxonomic tree is derived from the Open Tree Taxonomy (OTT hereafter;
-publication in preparation). The phylogenetic inputs are published
-trees that have been curated to align the tips to OTT and to identify
-the correct rooting (see \citealt{McTavishEtAt2015} for further details
-of the curation tools). Unlike OTT, these phylogenetic trees do not
-include all leaf taxa. The inputs (taxonomy and phylogenetic trees)
-and the output summary supertree are all rooted. Here we describe
-the software pipeline (\emph{propinquity}) that summarizes and integrates
-these smaller source trees and the taxonomy tree into a single supertree
-and the noteworthy tools for manipulating and solving supertrees in
-the \emph{otcetera} package.
-
-\subsection{Goals}
-
-Translating the goals of the Open Tree of Life's summary tree into
-an explicit set of criteria is not trivial. The summary supertree
-should represent the phylogenetic information from source trees in
-a transparent and justifiable fashion. We would like to allow users
-to correct errors in the supertree by improving the input information
-rather than requiring modification to the supertree algorithm. The
-pipeline was designed to create a tree which:
-\begin{enumerate}
-\item displays no unsupported groups,
-\item defers to groupings from higher ranked trees in the case of conflict, 
-\item contains no unnecessary polytomies, and
-\item displays as many groupings from input trees as possible.
-\end{enumerate}
-These goals are described more fully below. In order to accomplish
-transparency and justification, our pipeline also produces annotations
-files with information about conflict and support.
-
-\subsubsection{Goal 1: Each grouping is supported by at least one input}
-
-We require that each edge in the supertree be supported by at least
-one input tree edge. In addition to aiding interpretability, this
-requirement keeps the supertree from arbitrarily representing information
-that comes from none of the input trees. Of course, in a supertree
-analysis, the full tree will imply some relationships for subsets
-of the taxa that are not found in any input tree. So, the meaning
-of ``supported by'' needs some clarification.
-
-\paragraph*{Notation, terminology, and the definition of ``supported by''}
-
-Cutting any edge $j$ of a rooted tree induces a bi-partition
-$S(j)$ of the leaf taxa into two connected groups $S_{1}(j)$ and
-$S_{2}(j)$. Such a bi-partition is called a rooted split and written
-$S(j)=S_{1}(j)|\bullet S_{2}(j)$. Here $S_{1}(j)$ is called the ``cluster'' of $j$, and contains tip taxa on the side of edge $j$ that does not contain the
-root. $S_{2}(j)$ contains the tip taxa on the side of edge $j$ that
-contains the root. We refer to $S(j)$ as a \emph{rooted} split because
-the right side of the split implicitly contains the root node, as indicated
-by the $\bullet$.
-
-For any two rooted splits $A=A_{1}|\bullet A_{2}$ and $B=B_{1}|\bullet B_{2}$,
-we say that $A$ \emph{displays} $B$ if $B_{1}\subseteq A_{1}$ and
-$B_{2}\subseteq A_{2}$. We also say that $A$ and $B$ \emph{conflict}
-if none of $A_{1}\cap B_{1}$, $A_{1}\cap B_{2}$, and $B_{2}\cap A_{1}$
-are empty.
-Note that the intersection of the right sides of the splits is considered to
-be non-empty even if $A_{2}\cap B_{2}$ is empty, because the right
-sides implicitly contain the root. If $A$ and $B$ do not conflict, then we say that they are \emph{compatible}\footnote{It is possible to construct a rooted tree that displays both $A$
-and $B$ if and only if $A$ and $B$ are compatible}. If $A$ displays $B$, then $A$ and $B$ must be compatible, since
-$A_{1}\cap B_{2}$ and $B_{1}\cap A_{2}$ must be empty. We adopt
-the shorthand of referring to an edge $k$ when we mean the split
-$S(k)$. So, when we say that edge $j$ conflicts with edge $k$,
-we mean that $S(j)$ conflicts with $S(k)$.
-
-Let $\summaryTree$ denote a supertree, and $T_{i}$ denote the $i$th
-input tree. We say that $\summaryTree$ displays an edge $j$ of $T_{i}$
-if any edge of $\summaryTree$ displays $j$. We say that $\summaryTree$
-conflicts with a edge $j$ of $T_{i}$ if any edge of $\summaryTree$
-conflicts with $j$; otherwise $\summaryTree$ is compatible with
-$j$. We say that edge $k$ of $\summaryTree$ is \emph{supported
-by} edge $j$ of $T_{i}$ if $\summaryTree$ displays $j$, but $\summaryTree$
-would not display $j$ if we contracted edge $k$. We also adopt the
-shorthand of referring to edges by their tipward endpoint. By combining
-this with the above shorthand, we could say that a node $j$ conflicts
-or does not conflict with another node $k$.
-
-The set of taxa that are mapped to the tips of the tree $T_{i}$ is
-$\mathcal{L}(i)$. $\summaryTree(i)$ denotes the summary tree induced
-by tip nodes that are mapped to taxa in $\mathcal{L}(i)$ and the
-most recent common ancestor of those leaves, and any other node that
-is an ancestor of some but not all of these leaves.
-
-Note that stating that a node in the summary tree is supported by
-an input does not imply that every descendant of that node must be
-present in the input nor that every taxon that is not a descendant
-must be excluded in order to display the node. Consider the problem
-shown in Fig. \ref{fig:toyambig}; panels \ref{fig:toyambig}a
-and \ref{fig:toyambig}b show two input trees. Because taxa A and
-E do not occur together in either input, there is some uncertainty
-about where to place them. By our terminology, either output shown
-in Fig. \ref{fig:toyambig}c or Fig. \ref{fig:toyambig}d would be characterized
-as a tree that displays all of the input groupings and which has no
-unsupported groups. Clearly these criteria are insufficient to specify
-a unique solution, and users of the output tree need to be aware that
-it may be possible for some taxa to ``float'' to multiple positions.
-In Fig. \ref{fig:toyambig}, taxon E floats to different positions
-in \ref{fig:toyambig}c and \ref{fig:toyambig}d, whereas taxon
-A does not.
-
-\begin{figure}
-\subfloat[Input tree 1]{\includegraphics[width=0.45\textwidth]{Fig1a}
-
-}\hfill{}\subfloat[Input tree 2]{\includegraphics[width=0.45\textwidth]{Fig1b}
-
-}
-
-\subfloat[Summary tree]{\includegraphics[width=0.45\textwidth]{Fig1c}
-
-}\hfill{}\subfloat[Summary tree]{\includegraphics[width=0.45\textwidth]{Fig1d}
-
-}
-
-\caption{An example demonstrating that our definition of ``supported by''
-does not imply entire composition of a grouping. (a) and (b) show
-2 input trees and (c) and (d) depict trees that each display each
-of the groupings in the input trees and which have no unsupported
-nodes. The BUILD algorithm (section \ref{subsec:Subproblem-solution})
-would choose tree (d) that floats taxon E closer to the root.}
-
-\label{fig:toyambig}
-\end{figure}
-
-One of our aims in supertree construction is to minimize the amount
-of information in the supertree that does not come from input trees.
-We permit information that comes from combinations of input trees,
-but not any single input tree. However, we seek to exclude information
-that comes from none of the input trees. This motivates the criterion
-of not having any unsupported edges, since these edges could be removed
-without decreasing the support from any input tree.
-
-\subsubsection{Goal 2: Tree ranking}
-
-An appealing goal for the summarization would be to find the supertree
-that displays the largest number of input tree edges. As discussed
-in \citet[pages 92 and 131;][]{HusonRS2010} the maximum compatibility
-problem is known to be $NP$-hard via a reduction to Max-Clique \citep{Karp1972}.
-In addition to being computationally daunting, this formulation of
-the supertree problem does not provide biologists who use the summarization
-tool with an obvious avenue for fixing perceived problems with the
-summary tree. For example, a grouping that a biologist expected may
-not be present in the supertree, but it may not conflict with any
-of the input groupings which are displayed. This can happen because
-displaying both node $a$ from $T_{1}$ and node $b$ from tree $T_{2}$
-in a summary tree may only be possible by displaying a grouping that
-is present in no input tree. All other factors being equal, if this
-implied grouping conflicts with input node $c$ in tree $T_{3}$,
-then $c$ will not be displayed in the summary tree, but a biologist
-will not necessarily know how to fix this problem. One solution is
-to use a ranking of groupings. If an expert were quite confident in
-the $c$ grouping, then she could assign that input node a high ranking.
-A supertree that used ranks could then recover this grouping even
-if its inclusion did not increase the total number of input nodes
-that are displayed by the summary tree. Fig. \ref{fig:pairwisecompat}
-shows an example of 3 input trees for which there is no pairwise incompatibility,
-but no solution displays all of the input groups. Alternative rankings
-of inputs can result in one of three summary trees shown in panels
-d-f.
-
-\begin{figure}
-\subfloat[Input tree 1]{\includegraphics[scale=0.57]{Fig2a}
-
-}\hfill{}\subfloat[Input tree 2]{\includegraphics[scale=0.57]{Fig2b}
-
-}\hfill{}\subfloat[Input tree 3]{\includegraphics[scale=0.57]{Fig2c}
-
-}
-
-\subfloat[Summary tree]{\includegraphics[scale=0.57]{Fig2d}
-
-}\hfill{}\subfloat[Summary tree]{\includegraphics[scale=0.57]{Fig2e}
-
-}\hfill{}\subfloat[Summary tree]{\includegraphics[scale=0.57]{Fig2f}
-
-}
-
-\caption{An example of 3 input trees shown in (a), (b), and (c) which do not
-conflict in a pairwise manner, but cannot be jointly displayed in
-one tree. The 3 solution trees are shown in panels (d-f). Panel (d)
-for ranking the tree in (c) lowest. Panel (e) shows the solution if
-the tree in (b) has the lowest rank. Panel (f) shows the solution
-if the tree in (a) is ranked lowest. Each of the solutions displays
-2 of the 3 input groupings. }
-
-\label{fig:pairwisecompat}
-\end{figure}
-
-Any approach to supertree construction must deal with the need to
-adjudicate between conflicting input trees. We choose to deal with
-conflict by ranking the input trees, and preferring to include edges
-from higher-ranked trees. The merits of using tree ranking are questionable
-because the system does not mediate conflicts based on the relative
-amount of evidence for each alternative. However, it is a reasonable
-starting point. It has the benefits of making it easy to see why some
-groups are included or not (transparency), and it allows simpler and
-cleaner algorithms.
-
-Note that if some edge $c$ conflicts with a higher-ranked edge $b$,
-then $c$ may still be included in the supertree. This can occur when
-the higher ranked edge $b$ conflicts with a yet-higher ranked edge
-$a$, and thus $b$ is not included. In that case, it will be possible
-for $c$ to be represented in the summary tree. Thus, the fact that
-the summary tree displays an input edge does not imply that none of
-the higher ranked input trees conflict with that edge.
-
-In order to produce a comprehensive supertree, we also require a rooted
-taxonomy tree in addition to the ranked list of rooted input trees.
-Unlike other input trees, the taxonomy tree is required to contain
-all taxa, and thus has the maximal leaf set. We make the taxonomy
-tree the lowest ranked tree. In our current formulation, the taxonomy
-tree is also unique in that the taxonomy is the only source of taxonomic
-names. Each node in the taxonomy tree corresponds to a named group.
-Taxonomic groups may have the same name, but each node in the taxonomy
-tree is identified by a unique number (its OTT ID). Taxonomic groups
-are identified in the summary supertree by finding a branch (or ``node'')
-that has exactly the same include|exclude relationship. The taxonomy
-supertree can meaningfully possess degree-two nodes. Although these
-nodes can be removed without affecting the relationships of the leaves,
-they do represent nested taxonomic groups that contain exactly one
-subgroup. The taxonomy is also used to determine which tips are terminal
-taxa.
-
-\subsubsection{Goal 3: Contain no unnecessary polytomies}
-
-The supertree should be as resolved as possible - in other words,
-it should have no unnecessary polytomies. Thus, for each input edge
-that is \emph{not} included, we can point to a reason for non-inclusion
-by showing that the input edge conflicts with some edge of the summary
-tree. Note, that the requirement to not display unsupported groups
-leads to some ``necessary'' polytomies. For example any resolution
-of the polytomy shown in Fig. \ref{fig:pairwisecompat}e would
-continue to display the same two input groups. However, the additional
-grouping would be unsupported, because the unresolved tree already
-displays both input groups. Thus, the 
-unresolved tree would be preferred by our criterion.
-However, collapsing either internal edge of the tree shown in Fig.
-\ref{fig:pairwisecompat}d would result in a tree which displays
-only one input grouping. This tree would contain an unnecessary polytomy,
-because the polytomy would permit refinement to the depicted tree
-which displays more input groupings.
-
-\subsubsection{Goal 4: Display as many input nodes as feasible}
-
-We also seek to construct a supertree that represents as many input
-tree nodes as possible. Since non-included input tree nodes must conflict
-with the supertree (or they would have been added), this criterion
-is the same as minimizing the number of input nodes that conflict
-with the supertree.
-
-\subsubsection{Non-goal: phylogeny estimation}
-Many researchers construct supertrees for the purpose of inferring the true tree from the input trees \citep{BinindaEmonds2007b,Davis2014}.
-Such approaches treat input trees as data or as surrogates for data matrices \citep{gatesy2004critique}, and the supertree is seen as more accurate than the individual input trees because more input trees (and presumably more data) stand behind it.
-For such methods, conflict resolution is a primary aim of the method.
-
-Our supertree method operates in a different paradigm.
-Phylogeny estimation is an explicit non-goal of our supertree algorithm.
-We do not claim that the supertree is more accurate than the input trees it summarizes, and so conflict resolution is not a primary aim of the method.
-Most of our input trees have non-overlapping taxon sets, and so conflict resolution is not the primary problem.
-Instead, we seek to merge and aggregate phylogenetic information on the largest possible scale.
-Therefore, the primary aim of our supertree method is to \emph{summarize and represent} the input trees, even when they conflict with each other.
-The annotations file constructed by our pipeline is essential for this task because it includes information about conflict with our supertree as well as support for it.
-
-Our method must resolve conflicts in order to construct a single supertree.
-However, the rank information used to resolve conflicts is an input to the method, not an output from the method.
-We thus perform curation-based conflict resolution, not inference-based conflict resolution.
-
-\subsubsection{Summary of goals}
-
-These optimality criteria help to define what it means for the supertree
-to represent the input trees, as well as justifying and explaining
-why various features of the supertree exist. The pipeline described
-below produces a supertree that satisfies the first three optimality
-criteria and is a greedy approximation of a solution to the fourth
-goal. It is not guaranteed to display as many input nodes as possible.
-Even if the summary tree does accomplish goal 4, it is not necessarily
-a \emph{unique} optimum. The pipeline takes a greedy approach to producing
-a summary tree by attempting to add groupings from the trees in order
-of the trees ranking. This can be viewed as a greedy solution to the
-problem of finding the tree with the maximum sum of displayed groups'
-weighted scores criterion (MSDGWS, described in the appendix A) where the
-weights from the trees are so extreme that displaying one group from
-a highly ranked tree is preferred to displaying all of the groupings
-from lower ranked trees.
-These large weights help to achieve our goal of allowing curators to fix problems in the tree.
-While it is possible to view our pipeline as an optimization algorithm by assigning weights to the input trees in this way, we are not trying to weight the evidence in the input trees, and we do not actually ever calculate these weights in our pipeline.
-Note that in this extreme form of weighting, we do not have to calculate
-the score of a tree, we simply traverse the groupings by order of rank
-and add a group if it is compatible with the previously added groups.
-
-\subsubsection{Comparison to other supertree methods}
-
-Matrix Representation Parsimony (MRP) is one of the
-most widely used supertree approaches \citep{baum1992combining,BinindaEmonds2007b,Davis2014}.
-MRP attempts to encode branches of input trees as characters in a data
-matrix, and then find an optimal tree according to the maximum
-parsimony criterion. Such an approach does not provide a clear and transparent
-explanation for including particular edges in the supertree.
-MRP can also infer edges that exist in none of the input trees, which 
-violates our Goal \#1 \citep{gatesy2004critique}.
-
-Another commonly used method is MinCut \citep{semple2000supertree} or
-Modified MinCut \citep{page2002modified}.  Like our approach, MinCut
-is based on the BUILD algorithm \citep{AhoSSU1981}.  However, MinCut deals
-with conflicts by modifying the BUILD algorithm to resolve
-incompatibilities by discarding edges that are present in the smallest 
-number of input trees.  This approach thus violates our Goal \#2 of
-resolving conflict via ranks that can be altered by a curator to
-influence the output tree.
-
-\section{Description of the supertree method}
-
-\subsection{Preprocessing steps}
-
-Propinquity was designed to function as a part of the Open Tree of
-Life software architecture, so the first few steps of the pipeline
-involve transforming artifacts from that project into a set of rooted
-trees and a phylogenetic taxonomy. The phylesystem API \citep{McTavishEtAt2015}
-of Open Tree allows users to curate published estimates of trees and
-create ranked collections of these trees. Early steps in the propinquity
-pipeline manipulate the phylogenetic input trees to improve their
-usability and reliability. The first steps of the pipeline (see Fig.
-\ref{fig:pipeline}) collect a list of trees to include (in the \texttt{phylo\_input}
-subdirectory) and store copies of these files (in the \texttt{phylo\_snapshot}
-subdirectory) to make it easier to replicate the operation (because
-the collection of trees and the tree files change due to curation).
-\begin{figure}
-\begin{centering}
-\includegraphics[height=0.70\paperheight]{Fig3.pdf}
-\par\end{centering}
-\caption{Organization of the \emph{propinquity} pipeline. Each colored pentagon
-labels a program (blue for otcetera-based tools and red for python
-scripts in the propinquity or peyotl repository) that performs the
-important operations in each step; the number before the tool name
-refers to the section in this paper that describes the operation.
-The output of each step corresponds to a subdirectory of the propinquity
-system which will hold the output artifacts for the step. Ovals are
-resources that are required (OTT and Open Tree's phylesystem repository).
-White pentagons are user-controlled inputs.}
-\label{fig:pipeline}
-\end{figure}
-
-\subsubsection{Pruning questionable taxa from the taxonomy}
-
-OTT is a hierarchy of taxonomic names that implies a phylogenetic
-taxonomy. An OTT ID has a position in the hierarchy, a taxonomic name,
-and set of references to the same name in different taxonomies. In
-addition, the ID may also be associated with a set of flags that can
-indicate that the taxon may be questionable. These flags can either
-encode information taken from an input taxonomy (for example, taxa
-the NCBI refers to as ``unplaced'' are assigned an ``unplaced''
-flag) or can arise because of some form of conflict during taxonomy
-construction (for example, if two taxonomies disagree on the name
-for a taxon, then the taxon can be merged and the name will be retained
-without any descendants; this name will have an OTT ID, but will be
-flagged as ``barren''). Propinquity prunes the OTT down to a more
-reliable taxonomy by pruning off parts of the tree that are flagged
-with suspicious flags. The set of flags that lead to a subtree of
-the taxonomy being pruned is under the control of the user (the set
-of flags used by the Open Tree of Life project can be found in the
-\texttt{config.opentree.synth} file in the propinquity repository).
-For the purpose of the rest of the pipeline, an OTT ID that has been
-pruned from the taxonomy will be treated in the same way as invalid
-OTT ID. The output of this step is stored in propinquity's \texttt{cleaned\_ott}
-subdirectory; this operation only needs to be performed when the OTT
-or the pruning flags change.
-
-\subsubsection{Pruning problematically mapped tips from input phylogenetic trees}
-
-Frequently, phylogenetic estimates are rooted using the outgroup criterion,
-which is an assumption about the monophyly of the ingroup taxa. Because
-the rooting of the branches in the outgroup portion of the tree is
-often uncertain, data curators can identify the ingroup node of the
-tree; propinquity uses this annotation to prune off the outgroup taxa.
-
-Frequently, not all tips in a phylogenetic input will have been mapped
-to a taxon in the current version OTT. Unmapped leaves are pruned
-from each phylogenetic input. In some cases, the OTT has changed and
-a taxon has been unambiguously mapped to another taxon. This can occur
-when multiple species in one version of the taxonomy are ``lumped''
-into a single taxon in a subsequent version. OTT maintains a set of
-``forwarding'' statements about IDs that have been removed but can
-be mapped to an existing taxon; propinquity uses these statements
-to update the OTU mapping of input trees.
-
-Finally some leaves are mapped to taxa that occur more than once in
-the tree, or taxa that have ancestors represented as tips of the tree.
-In these cases, leaves are pruned to assure that tips are mapped to
-unique taxa that are not nested. In the case of nested taxa, the tip
-mapped to the higher level taxon is pruned, and one of the lower level
-tips is retained. In the case of duplicate occurrences of an OTT taxon,
-propinquity checks to see if a data curator has selected one of the
-taxa to be the exemplar for the taxon. If this selection has not been
-made, then the node with the lexicographically lowest ID is chosen
-to exemplify the taxon. This choice is arbitrary, but repeatable.
-The pruned phylogenetic inputs are stored in a \texttt{cleaned\_phylo}
-subdirectory of propinquity.
-
-\subsubsection{Exemplifying tips mapped to higher taxa}
-
-Many input trees have tips that are not terminal taxa, but higher-order
-taxonomic groups. It is not clear how to interpret a tip in a phylogenetic
-estimate that is labeled with the name of a higher taxon. Several
-scenarios can lead to these cases: the data for the tip could have
-been created by merging a chimeric set of character scores from constituent
-taxa; the species sampled may not have been identified to the lowest
-taxonomic rank; or the researcher may simply have used a higher taxonomic
-name because he/she assumed that the taxon is monophyletic and the
-higher level name would be more recognizable. Rather than allowing
-the ambiguity about interpretation of the higher-taxon mapped tips
-to propagate throughout the entire pipeline, we transform the input
-trees by replacing higher taxa at tips with a set of terminal-taxon
-exemplars for each taxon. One approach would be to simply determine
-all descendant terminal taxa and attach them as children of the problematic
-tip. However, this would create a clade rather than a tip; subsequent
-steps in the supertree would interpret the clade as a claim of monophyly
-for the taxon. The input tree may not have tested monophyly of the
-clade, so this interpretation is unwarranted. We avoid it by attaching
-exemplar taxa as child nodes of the higher taxonomic tip but then
-collapsing the edge between the former tip node and
-its parent. Thus, if $A$ is a non-terminal taxon containing terminal
-descendants $a_{1}$ and $a_{2}$ and $B$ is a non-terminal taxon
-containing terminal descendants $b_{1}$ and $b_{2}$ we would replace
-the subtree $((A,B),c)$ with $((a_{1},a_{2},b_{1},b_{2}),c)$ instead
-of the subtree $(((a_{1},a_{2})A,(b_{1},b_{2})B),c)$.
-
-If a taxon is only present in the taxonomy (not in any of the input
-trees), then it can be pruned from the taxonomy for the construction
-of the supertree and then grafted back on to the summary tree later.
-Performing this pruning reduces the size of the supertree problem,
-reducing the running time of the pipeline. Similarly, when we expand
-a higher taxon in the exemplification step, we can omit members of
-the taxon if they do not occur in any of the phylogenetic inputs.
-If there are no members of the higher taxon sampled in any other input
-tree, then we arbitrarily choose one terminal taxon to represent the
-higher taxon. During the exemplification step, a tool from otcetera
-(\texttt{otc-nonterminals-to-exemplars}) reads the taxonomy and all
-of the ``cleaned'' phylogenetic estimates from the previous step.
-Reading all of the inputs is necessary to assure that each higher
-taxon is replaced with the same set of exemplars regardless of which
-tree the higher taxon occurs in, and that the exemplars for a higher
-taxon is the union on the set of descendant terminal taxa that have
-been sampled in a phylogenetic input. For example, the trees in Fig.
-\ref{fig:cleaned-phylo-example} would exemplified as shown in Fig.
-\ref{fig:exemplify-example}.
-
-We prune the taxonomy by removing tips that are not present in any
-input tree to produce the pruned taxonomy $\mathbb{T}_{p}$. The tips
-pruned in this step will be grafted back onto the skeleton of the
-summary tree in a subsequent step. A terminal taxon that is represented
-only in the taxonomy can be pruned and then regrafted onto the solution
-without affecting which nodes are displayed by the final summary tree.
-Thus, this procedure does not impede our ability to find a good summary
-tree. Removing these tips produces a smaller input to the rest of
-the pipeline, which reduces running times. After producing the set
-of ``exemplified'' phylogenetic inputs, this tool exports a pruned
-down version of the taxonomy that only contains tips that are present
-in at least one phylogenetic input.
-
-\begin{figure}
-\subfloat[Tree 1]{\includegraphics[scale=0.4]{Fig4a}
-
-}\hfill{}\subfloat[Tree 2]{\includegraphics[scale=0.4]{Fig4b}
-
-}\hfill{}\subfloat[Taxonomy Tree]{\includegraphics[scale=0.4]{Fig4c}
-
-}
-
-\caption{Input trees and taxonomy tree}
-\label{fig:cleaned-phylo-example}
-\end{figure}
-
-\begin{figure}
-\subfloat[Tree 1]{\includegraphics[scale=0.4]{Fig5a}
-
-}\hfill{}\subfloat[Tree 2]{\includegraphics[scale=0.4]{Fig5b}
-
-}\hfill{}\subfloat[Pruned taxonomy tree $\protect\prunedTaxonomy$]{\includegraphics[scale=0.4]{Fig5c}
-
-}
-
-\caption{Exemplified input trees and taxonomy tree from Fig. \ref{fig:cleaned-phylo-example}.
-E in tree1 is exemplified by E1. Pruned taxa are E2, F2, and D. The
-taxa E and F are retained as monotypic taxa in the pruned taxonomy
-$\protect\prunedTaxonomy$, but are not shown in panel c. The red
-edge in the pruned taxonomy tree is an uncontested higher taxon in
-the exemplified taxonomy (as explained in section \ref{subsec:Subproblem-decomposition})}
-\label{fig:exemplify-example}
-\end{figure}
-
-\subsection{Summary tree construction}
-
-After the preprocessing steps, the inputs have been converted to a
-set of rooted phylogenetic estimates in which each leaf is mapped
-to a terminal taxon in the exemplified taxonomic tree. The goal of
-the remainder of the pipeline is to construct a tree that maximizes
-the sum of displayed groups' weighted scores (MSDGWS) criterion. This is accomplished
-in four steps: (1) dividing the full problem into subproblems based
-on uncontested taxa, (2) constructing a summary solution for each
-subproblem by greedily creating a maximally-sized list of groupings
-that can all be displayed simultaneously; (3) grafting the subproblem
-solutions into a single supertree; and (4) grafting (or ``unpruning'')
-the taxonomy-only taxa onto the solution to produce a complete summary
-tree.
-
-\subsubsection{Subproblem decomposition\label{subsec:Subproblem-decomposition}}
-
-For the sake of efficiency, propinquity uses a divide-and-conquer
-approach to construct the supertree. Subproblems are identified by
-searching through the taxonomy tree to find any taxa that are not contested
-by any single input tree. Here we say that input tree $T_{i}$ contests
-taxon $x$ in the pruned taxonomy, if $x$ is not monophyletic in
-any resolution of tree $T_{i}$. Thus, polytomies in an input tree
-are treated as soft polytomies, and a taxon is not contested merely
-because it is not displayed by an input tree.
-
-This operation is performed by the \texttt{otc-uncontested-decompose}
-tool in otcetera; see appendix \ref{sec:Decompose-algorithm} for
-a description of the algorithm. The output is a series of subproblems,
-each of which corresponds to a slice of the taxonomy and corresponding
-slices through each relevant input tree. Each uncontested non-terminal
-and non-root taxon will show up in two subproblems: it will be the
-root of its own subproblem and it will be tip in the subproblem that
-covers the next slice deeper in the tree. The red edge in Fig. \ref{fig:exemplify-example}c
-highlights the taxa that are not contested by the input shown in Fig. \ref{fig:exemplify-example};
-Fig. \ref{fig:subproblems-example} shows the subproblems that would
-be emitted as a result of this set of inputs. The supertree operation
-of \citet{HinchliffEtAl2015} also used this otcetera-based decomposition
-step.
-
-Note that decomposition into uncontested groups does not necessarily
-allow us to find the tree that maximizes the MSDGWS score. For example,
-see Fig. \ref{fig:decompose-worsens}; that example is a variant
-of the situation shown in Fig. \ref{fig:pairwisecompat}. In this
-case the groupings from each of the phylogenetic estimates, shown
-in panels \ref{fig:decompose-worsens}a and \ref{fig:decompose-worsens}b,
-could be displayed. That solution is shown in panel \ref{fig:decompose-worsens}d,
-it displays two of the three input splits, but is optimal because
-no solution displays all three input groupings and the depicted solution
-displays the two highest ranked groupings. However, neither of the
-trees shown in panels \ref{fig:decompose-worsens}a or \ref{fig:decompose-worsens}b
-contest the taxon $B$ shown in the taxonomy panel \ref{fig:decompose-worsens}c.
-Thus, when using our decomposition, the branches leading to taxa $B1$
-and $B2$ in the input phylogenetic trees would be sliced during the
-decomposition, and relabeled to refer to taxon $B$. This taxonomically-informed
-interpretation of the inputs views the two phylogenetic inputs as
-in conflict; so the solution returned by propinquity would defer to
-the higher ranked tree. The tree shown in panel \ref{fig:decompose-worsens}e
-would be returned. This example arises from the fact that the trees
-in \ref{fig:decompose-worsens}a and \ref{fig:decompose-worsens}b
-jointly contest taxon $B$, but neither contests taxon $B$ when the
-trees are considered in isolation.
-
-Despite the fact that the use of \texttt{otc-uncontested-decompose}
-can worsen the final score of the summary tree, we use this approach
-in propinquity because it makes the construction of the tree faster
-and it is easy for users to correct issues caused by incorrect taxa
-being constrained to be monophyletic. By adding a tree (even a low-ranked
-tree) that contests a taxon to the corpus of input trees , then the
-next synthetic tree will no longer consider the taxon to be uncontested.
-Thus the procedure encourages curation of more phylogenetic inputs
-as a means of improving the summary tree.
-
-\begin{figure}
-  \subfloat[Subproblem \emph{ABCD}]{\includegraphics[width=0.3\textwidth]{Fig6a}%
-    \hfill{}\includegraphics[width=0.3\textwidth]{Fig6b}%
-    \hfill{}\includegraphics[width=0.3\textwidth]{Fig6c}
-
-}
-
-  \subfloat[Subproblem \emph{root}]{\includegraphics[width=0.3\textwidth]{Fig6d}%
-    \hfill{}\includegraphics[width=0.3\textwidth]{Fig6e}%
-    \hfill{}\includegraphics[width=0.3\textwidth]{Fig6f}
-
-}
-
-\caption{Subproblems generated from the exemplified trees shown in Fig. \ref{fig:exemplify-example}.
-A trivial statement from the first tree that a taxon labelled $ABCD$
-is sister to $E$ has been omitted, because trees with only 2 leaves
-do not contain phylogenetic information.
-Similarly, the trivial problems that list E1 and F1 as the only daughters
-of taxa E and F have been omitted.}
-\label{fig:subproblems-example}
-\end{figure}
-\begin{figure}
-\subfloat[]{\includegraphics[scale=0.57]{Fig7a}
-
-}\hfill{}\subfloat[]{\includegraphics[scale=0.57]{Fig7b}
-
-}\hfill{}\subfloat[]{\includegraphics[scale=0.57]{Fig7c}
-
-}
-
-\hfill{}\subfloat[]{\includegraphics[scale=0.57]{Fig7d}
-
-}\hfill{}\subfloat[]{\includegraphics[scale=0.57]{Fig7e}
-
-}\hfill{}
-
-\caption{An example with three input trees: the highest ranked phylogenetic
-input panel (a), the second ranked phylogenetic input (b), and the
-taxonomy in panel (c). The summary tree in panel (d) has the highest
-possible score, but the summary shown in panel (e) would be returned
-from the pipeline that uses uncontested taxon decomposition.}
-\label{fig:decompose-worsens}
-\end{figure}
-
-\subsubsection{Subproblem solution\label{subsec:Subproblem-solution}}
-
-When solving sub-problems, we sequentially incorporate splits from
-trees in order of ranking, retaining splits that are compatible with
-the current set of splits (Fig. \ref{alg:ConsistentSplitsFromRankedList}).
-The order of splits from the same tree is not specified by this approach,
-and we incorporate splits using one of the possible post-order traversals
-of the tree. We make use of the BUILD algorithm \citep{AhoSSU1981}
-to assess compatibility. This strategy avoids unnecessary polytomies,
-since splits of later input trees are only rejected from the summary
-supertree if they conflict with higher-priority splits. Finally, we
-use the BUILD algorithm to construct a supertree displaying all of
-the splits in the set of compatible splits. Using the BUILD algorithm
-to construct the subproblem summary tree satisfies criterion 3, because
-trees from the BUILD algorithm do not contain unsupported branches.
-\begin{figure}
-\includegraphics[width=0.9\textwidth]{fig8}
-\caption{\label{alg:ConsistentSplitsFromRankedList}Algorithm \texttt{ConsistentSplitsFromRankedList}}
-\end{figure}
-
-The BUILD algorithm as originally stated by \citet{AhoSSU1981} applies
-to a collection of rooted triplets. Instead of decomposing each input
-split into a collection of rooted triplets, we instead modify the
-BUILD algorithm to apply directly to larger rooted splits. The modified
-BUILD algorithm constructs a tree compatible with a collection of
-rooted splits, and returns failure if such a tree does not exist.
-This modified algorithm recovers the original BUILD algorithm if only
-rooted triplets are supplied as input. When larger splits are supplied
-as input, the results are the same as if each was was decomposed into
-all implied triplets. The modified build algorithm has order $O(N^{2}+N^{2}E+NL)$
-where $N$ is the number of splits passed in, $E$ is the average
-size of the exclude group, and $L$ is the total number of leaves.
-This simplifies to $O(N^{2})$ if all splits are triplets. In this
-approach splits are either entirely retained or entirely discarded
-- consistent rooted triplets from conflicting splits are not retained.
-However, when unpruning taxonomy-only taxa (see below), we make an
-attempt to break ties in a way that preserves some partial information
-from conflicting splits by attaching taxa from conflicting splits
-at their common ancestor. Fig. \ref{fig:subproblem-solutions.}
-shows the solutions that would be obtained by applying our modified
-version of the BUILD algorithm to the subproblems shown in Fig. \ref{fig:subproblems-example}.
-
 \begin{figure}
@@
 
-\caption{Solutions to the subproblems depicted in Fig. \ref{fig:subproblems-example}}
-\label{fig:subproblem-solutions.}
 \end{figure}
-
-\subsubsection{Solution grafting}
-
-To produce a tree that spans all of the taxa sampled in the exemplified
-set of input trees, we graft the subproblem solutions into a single
-tree (stored as the \texttt{grafted\_solution/grafted\_solution.tre}
-by propinquity); see Fig. \ref{fig:grafted-solution}. Recall that each non-root uncontested taxon used
-for decomposition occurs as a leaf taxon in one subproblem and as
-a root taxon in one other subproblem. Thus, the grafting operation
-simply consists of reading all of the subproblem solutions into memory
-and then merging the nodes that are labeled with the same OTT ID.
-
-\begin{figure}
-\begin{centering}
-\includegraphics[scale=0.5]{Fig10.pdf}
-\par\end{centering}
-\caption{Grafted solution produced from the subproblems from
-Fig. \ref{fig:subproblem-solutions.} and which is the 
-backbone onto which taxa that are not included in any phylogeny
-will be placed.
-In a real analysis, the tips shown as E and F here
-would be monotypic taxa (with children E1 and F1, respectively)
-from the trivial subproblems that were omitted from from Fig. \ref{fig:subproblems-example}.}
-\label{fig:grafted-solution}
-\end{figure}
-
-\subsubsection{Unpruning unsampled taxa}
-
-As described above, taxa that do not have any descendants in a sampled
-phylogenetic input are pruned from the taxonomy for the sake of efficiency.
-These taxa are reattached by an ``unpruning step.'' For those taxa
-that are compatible with the grafted tree, this step simply amounts
-to adding any unsampled taxonomic children to the node that represents
-the taxon in the grafted solution tree.  For example, the tree in
-Fig. \ref{fig:grafted-solution} is unpruned to produce the tree in
-Fig. \ref{fig:unprune}.
-
-However, a taxon may be incompatible with the grafted solution; we
-refer to such taxa as ``broken taxa.'' If a broken taxon contains
-some unsampled children, it is not clear where these unsampled children
-should be attached to the grafted solution. One approach would be
-to mimic the application of Algorithm $\texttt{ConsistentSplitsFromRankedList}$ (Fig. \ref{alg:ConsistentSplitsFromRankedList})
-to the full (unpruned) taxonomic tree. This would be equivalent to
-collapsing each edge in the taxonomy that attaches a broken taxon
-to its parent. The unsampled children of broken taxa would attach
-at their least inclusive ancestral taxon which is unbroken. In cases
-of several adjacent taxa are broken, this can lead to polytomies of
-very high degree deep in the tree. This can make the summary tree
-difficult to navigate. Thus, we have adopted an alternative solution.
-The \texttt{otc-unprune-solution-and-name-unnamed-nodes} tool from
-otcetera attaches the unsampled children of a broken taxa to the grafted
-solution as children of the MRCA of the sampled children.
-
-Fig. \ref{fig:Two-approaches-to-unpruning} illustrates the two
-approaches to unpruning. Taxa G, M, and R (Fig \ref{fig:Two-approaches-to-unpruning}a)
-are broken because they conflict with the grafted solution (Fig \ref{fig:Two-approaches-to-unpruning}b);
-among these, only taxon R has children that were unsampled in the
-grafted solution. Ignoring all broken taxa when unpruning would cause
-the unsampled children (R4, R5, and R6) to attached directly at taxon
-N (as in the tree shown in Fig. \ref{fig:Two-approaches-to-unpruning}c),
-because that is the least inclusive unbroken ancestor of R. The tree
-illustrated in Fig. \ref{fig:Two-approaches-to-unpruning}d shows the
-tree that would be produced by propinquity; the children of the broken
-taxon R and instead attached at the MRCA of sampled children (R1,
-R2, and R3). Their attachment point does not correspond to any taxon
-in the taxonomic tree.
-
-\begin{figure}
-
-\begin{centering}
-\includegraphics[scale=0.5]{Fig11}
-\par\end{centering}
-\caption{\label{fig:unprune}Unpruned tree with internal taxa. Taxa unsampled in phylogenetic
-statements have been added to the grafted tree shown
-in Fig. \ref{fig:grafted-solution}.}
-\end{figure}
-
-\begin{figure}
-\begin{centering}
-\subfloat[Taxonomy]{\begin{centering}
-\begin{minipage}[c][1\totalheight][t]{0.45\columnwidth}%
-\begin{center}
-\includegraphics[scale=0.38]{Fig12a}
-\par\end{center}%
-\end{minipage}
-\par\end{centering}
-}\hfill{}\subfloat[Grafted solution]{\begin{centering}
-\begin{minipage}[c][1\totalheight][t]{0.45\columnwidth}%
-\begin{center}
-\includegraphics[scale=0.38]{Fig12b}
-\par\end{center}%
-\end{minipage}
-\par\end{centering}
-}\hfill{}
-\par\end{centering}
-\begin{centering}
-\subfloat[Unpruned tree - broken taxa removed]{\begin{centering}
-\includegraphics[scale=0.38]{Fig12c}
-\par\end{centering}
-}\hfill{}\subfloat[Unpruned tree with MRCA-attachment]{\begin{centering}
-\includegraphics[scale=0.38]{Fig12d}
-\par\end{centering}
-}
-\par\end{centering}
-\caption{\label{fig:Two-approaches-to-unpruning}Two approaches to unpruning.
-Taxa G and R in the taxonomy (a) are broken because they conflict
-with the grafted solution (b). Removing these broken taxa from the
-taxonomy before unpruning leads to taxa R4, R5, and R6 being attached
-directly at taxon N, as in tree (c). In tree (d), the children of
-the broken taxon R are instead attached at the MRCA of R1, R2, and
-R3. Our method follows the second approach.}
-\end{figure}
-
-\subsubsection{Naming unnamed nodes}
-
-In order to annotate each node in the summary supertree, it is first
-necessary that each node have a unique identifier. Nodes whose include
-group correspond exactly to the include group of a node in the taxonomy
-are given the same identifier as the corresponding taxonomy node.
-These identifiers are of the form $ottX$ where $X$ is an integer
-OTT ID. We generate a label of the form $mrcaottX_{1}ottX_{2}$ for
-an non-taxonomic node $n$ where $X_{1}$ and $X_{2}$ are the OTT
-IDs for two leaves, $n$ is the MRCA of these leaves, and $X_{1}$
-is the numerically smallest OTT ID that is a descendant of $n$, and
-$X_{2}$ is the next the smallest ID that can be chosen to designate
-$n$ as the MRCA. Because new taxa added to OTT will be given higher
-OTT IDs, the use of the lowest numbered OTT IDs as designators increases
-the chance that a node label can be encountered in a subsequent version
-of the tree (though the taxonomic content may change). The deterministic
-choice of designators also makes the labeling insensitive to branch
-rotation of the grafted solution tree.
-
-\subsubsection{Annotation}\label{sec:Annotation}
-
-\begin{figure}
-\subfloat[Input tree $T_{1}$ named ``tree1'']{%
-\begin{minipage}[c][1\totalheight][t]{0.25\columnwidth}%
-\includegraphics[scale=0.5]{Fig13a}%
-\end{minipage}
-
-}\hfill{}\subfloat[Summary Tree $\protect\summaryTree$]{%
-\begin{minipage}[c][1\totalheight][t]{0.35\columnwidth}%
-\includegraphics[scale=0.5]{Fig13b}%
-\end{minipage}
-
-}\hfill{}\subfloat[Induced Tree $\protect\summaryTree(1)$]{%
-\begin{minipage}[c][1\totalheight][t]{0.35\columnwidth}%
-\includegraphics[scale=0.5]{Fig13c}%
-\end{minipage}
-
-}
-
-\bigskip{}
-
-\bigskip{}
-
-\hfill{}\subfloat[JSON annotations relating edges of (c) to edges of (a)]{%
-\includegraphics{Fig13d}%
-}\hfill{}
-
-\caption{\label{fig:displayed-annotations}The relationship of edges in summary
-tree $\protect\summaryTree$ (b) to edges in the input tree $T_{1}$
-(a). Only edges of $\protect\summaryTree$ that are present in the
-induced tree $\protect\summaryTree(1)$ are represented by JSON annotations
-(d). Taxon names are here suppressed in favor of OTT IDs, and edges
-are referenced via their tipward nodes. Edges in $\protect\summaryTree(1)$
-that correspond to terminal edges of $T_{1}$ are orange; edges of
-$\protect\summaryTree(1)$ that are supported by edges of $T_{1}$
-are blue; where multiple edges of $\protect\summaryTree(1)$ correspond
-to the same edge of $T_{1}$ they are green. There is no conflict
-in this example. Also, if this were output from propinquity, then
-each internal node of $\protect\summaryTree$ would be supported by
-other inputs trees that are not shown here. }
-\end{figure}
-
-To reveal the connections between the groupings found in the a summary
-supertree and the input trees, propinquity uses a few Python scripts
-and the \texttt{otc-annotate-synth} tool from otcetera to create an
-annotations file describing the pipeline used and the connections
-between phylogenetic information in the inputs and the summary.  The
-JSON file produced by \texttt{otc-annotate-synth} encodes a ``nodes''
-property that holds a mapping between a node name for the summary
-tree (using the naming convention described in the previous section)
-and a node provenance object that categorizes the relationship between
-the node and the inputs. The node provenance object for node $x$
-uses several properties to categorize the relationship between the
-node and the inputs; each property in the node provenance object maps
-to a structure storing the tree ID and node IDs for the input tree
-nodes.
-
-Conceptually, this annotation operation is equivalent to considering
-every node $j$ in each input tree $i$ and the summary tree node
-$x$. Because the vast majority of nodes in the input studies will
-be compatible but not directly relevant to node $x$ we do not list
-all of the compatible groupings. If node $x$ is not included in the
-induced tree $\mathcal{\mathbb{S}}(i)$, then none of the nodes of
-tree $i$ will be referred to in the annotations for node $x$. Even
-if $x$ is included in $\mathcal{\mathbb{S}}(i)$, many of the nodes
-of $T_{i}$ will be compatible with $x$ while being relevant to other
-parts of the summary tree. The only input nodes listed for node $x$
-are with rooted taxon bipartitions which conflict with, are displayed
-by, or are resolved by the the rooted taxonomic bipartition associated
-with node $x$. All input nodes that cannot be displayed by any supertree
-that contains $x$ are stored in a ``\conflictswith{}'' property
-of the node provenance object. If node $j$ of $T_{i}$ is displayed
-by the summary tree and $x$ is part of the path of $\mathcal{\mathbb{S}}(i)$
-that displays the split between descendants of $j$ and other taxa,
-then a reference to the node $j$ will be in the node provenance object.
-The exact categorization of this annotation will depend on the configuration
-of node $x$ on the induced tree $\mathcal{\mathbb{S}}(i)$:
-\begin{itemize}
-\item if $x$ is on a terminal path in the induced tree then node $j$ will
-be listed in the ``\terminal{}'' property;
-\item if $x$ is along an internal path that contains some nodes with out-degree
-equal to 1, then node $j$ will be listed in the ``\partialpathof{}''
-property; and
-\item if $x$ is along an internal path without any node of out-degree 1,
-then node $j$ will be listed in the ``\supportedby{}'' property,
-because node $j$ supports the existence of grouping $x$ in the sense
-that collapsing the edge that separates $x$ from its parent would
-cause the summary tree to no longer display node $j$.
-\end{itemize}
-These three relationships are illustrated in Fig. \ref{fig:displayed-annotations}.
-Finally, if $T_{i}$ does not display $x$ from $\mathcal{\mathbb{S}}(i)$,
-but there exists an unresolved node $j$ in $T_{i}$ which could be
-resolved such that the tree would then display $x$, then a reference
-to node $j$ will be listed in the ``\resolves{}'' property of node
-$x$.
-
-The otcetera annotation tool can also detect cases in which including
-information from node $j$ in $T_{i}$ could further resolve a polytomy
-$x$ in the summary tree; such a case would be annotated using the
-``\resolvedby{}'' property of $x$. However, because of our goal
-of excluding unnecessary polytomies, none of the nodes in propinquity's
-summary tree will use this annotation when they are annotated with
-the set of input trees.
-
-\subsubsection{Self-documentation}
-
-An optional step in the propinquity pipeline (triggered by the executing
-the ``make html'' target) can compose an ``index.html'' file for
-each directory created during the pipeline to explain the artifacts
-held in that directory and report summary statistics about the summarization
-run.
-
-\section{Results}
-
-We seek to assess the performance of our new supertree method by
-comparing it to the supertree method of \cite{HinchliffEtAl2015}.  The
-method of \cite{HinchliffEtAl2015} was used to construct the Open Tree
-of Life v4 (\otl{}).  Therefore, in order to facilitate comparison, we 
-applied our method to the same input trees and taxonomy used by
-\otl{}.  
-We refer to the resulting supertree as \otlprop{} since it was
-constructed by applying the propinquity pipeline to the same inputs as
-\otl{}.
-
-\subsection{Inputs}
-
-The flag-cleaned version of OTT used in the construction of both
-supertrees contained 2,424,255 leaves.  The
-\otl{} supertree was constructed from 482 phylogenetic inputs,
-containing a total of 45,385 leaves, of which 41,029 were unique.
-After flag-cleaning and exemplification by propinquity, these trees contained 40,323
-unique tips. We used the same cleaning flags to trim the taxonomy and
-input trees when constructing \otlprop{}, so \otl{} and \otlprop{} contain the
-same number of leaves.
-
-\subsection{Subproblems}
-In the \otlprop{} summary tree, the decomposition procedure produced
-5,406 subproblems, but only 1,422 of these were non-trivial to solve.
-If a subproblem contains only two tips it is trivial; 2,362 subproblems
-were trivial in this way. Similarly, if a subproblem contains only
-2 trees it is trivial to solve because the solution will simply be
-all of the groupings from the first tree combined with all of the
-groupings from the second tree that are compatible with the first
-tree; 3,052 subproblems were trivial in this way. The subproblem with
-the largest number of tips contained 946 tips. The largest subproblem,
-in terms of the number of input trees (including the taxonomic tree)
-that were relevant, had 16 trees. Without decomposition, the supertree
-problem would have had 482 input trees and 41,226 leaves.
-
-% table
-
-% outputs - representation of input trees
-% outputs - annotations
-
-% conflicts in v4: 2,760
-% conflicts in v4propr: 1,357
-
-% displayed in v4: 35,518
-% displayed in v4prop: 39,639
-
-% # of input splits represented by v4, but not v4prop
-% # of input splits represented by v4prop, but not v4
-% # of input splits represented by v4 and v4prop
-
-\begin{table}
-
-\caption{Representation of input splits in the \otl{} tree and the \otlprop{} tree.}
-\resizebox{\textwidth}{!}{%
-\begin{tabular}{lcccccc}
- & \supportedby{} & \partialpathof{} & \terminal{} & \conflictswith{} & \resolvedby{} & \resolves{}\tabularnewline
-\hline 
-\otl{} & 34,595 & 923 & 45,385 & 2,760 & 2,718 & 473\tabularnewline
-\hline 
-\otl{} only & 745 & 54 & 0 & 2,055 & 2,718 & 0\tabularnewline
-\hline 
-\otlprop{} & 38,521 & 1,118 & 45,385 & 1,357 & 0 & 515\tabularnewline
-\hline 
-\otlprop{} only & 4,671 & 249 & 0 & 652 & 0 & 42\tabularnewline
-\hline
-\label{tab:representinputs}
-\end{tabular}}
-
-\end{table}
-
-\begin{table}
-
-\caption{Representation of taxonomy splits in the \otl{} tree and the \otlprop{} tree.}
-\resizebox{\textwidth}{!}{%
-\begin{tabular}{lcccccc}
- & \supportedby{} & \partialpathof{} & \terminal{} & \conflictswith{} & \resolvedby{} & \resolves{}\tabularnewline
-\hline 
-\otl{} & 125,384 & 0 & 2,424,255 & 1,998 & 4 & 3,676\tabularnewline
-\hline 
-\otl{} only & 296 & 0 & 0 & 19 & 4 & 17\tabularnewline
-\hline 
-\otlprop{} & 125,107 & 0 & 2,424,255 & 2,279 & 0 & 3,883\tabularnewline
-\hline 
-\otlprop{} only & 19 & 0 & 0 & 300 & 0 & 224\tabularnewline
-\hline 
-\label{tab:representtaxonomy}
-\end{tabular}}
-
-\end{table}
-
-\subsection{Representing input splits}
-
-We performed an annotation of both the \otl{} tree and the \otlprop{}
-tree as described in section \ref{sec:Annotation} to assess the
-ability of our new supertree method to represent splits from input
-phylogenies.  Table \ref{tab:representinputs} classifies the input
-phylogeny splits according to how they relate to a summmary tree, so
-that each input edge falls in one of \supportedby{}, \partialpathof{},
-\terminal{}, \conflictswith{}, or \resolvedby{}. For example, the
-numbers in the \conflictswith{} column indicate the number of input
-splits $j$ with at least one summary edge $x$ such that the relation
-``$x$ \conflictswith{} $j$'' holds.  The total number of non-terminal
-input phylogeny splits considered was 40,996.
-
-The number of displayed input splits (\supportedby{}
-+ \partialpathof{}) increased from 35,518 (for \otl{}) to 39,639 (for \otlprop{});
-an 11\% increase. When examining which splits are displayed, we find
-that the \otlprop{} tree displays 4,920 input splits that
-are not displayed by the \otl{} tree, whereas the \otl{} tree displays
-only 799 input splits that are not displayed by the \otlprop{} tree.  The number of input
-splits that conflict with the summary (\conflictswith{}) dropped from
-2,760 to 1,357, a decrease of 1,403, or 51\%.  In accordance with the
-goal of not containing unnecessary polytomies, the number of input
-splits that do not conflict with the summary tree, but are not
-incorporated (\resolvedby{}) dropped from 2,718 to 0.  We also find that
-the number of polytomies in input phylogenies that are resolved by the
-summary tree increases from 473 for \otl{} to 515 for \otlprop{}.
-
-We also performed an annotation of the \otl{} tree and the \otlprop{}
-tree to assess the degree to which these trees represent taxonomy
-splits (Table \ref{tab:representtaxonomy}).  The \otlprop{} tree
-conflicts with 281 more taxonomy splits than the \otl{} tree.  Since
-the taxonomy is the lowest ranked input tree, this increased conflict
-with the taxonomy is an expected result of incorporating more splits
-from higher-ranked input phylogenies.
-
-\subsection{Checking reliability of the method}
-To test the reliability of our methods, we divided the \otlprop{} summary
-tree into a collection of input trees. The input trees were obtained
-by splitting the grafted tree at each taxonomy node, so that each
-input tree has a taxonomy node at the root, and taxonomy nodes as
-tips, but no taxonomy nodes internally.  Each leaf node that was an
-internal taxon was then replaced with a leaf taxon. This approach
-ensures that any taxa that were contested in the original analyses
-remain contested in this new analysis.  We then constructed a summary
-tree from these input trees using the propinquity pipeline.  The
-resulting summary tree was identical to the \otlprop{} summary tree.
-
-\subsection{Sensitivity of the \otlprop{} supertree to ranks}
-We ran the pipeline on the OTL4 data with the ranks reversed.
-This drastic alteration of the ranks led to a summary supertree that conflicted with 1264 branches of the \otlprop{}.
-Compared to this, the \otl{} supertree tree conflicts with 1387 branches of the \otlprop{} supertree.
-Thus, reversing the ranks makes a smaller different in the \otlprop{} supertree than the new synthesis algorithm.
-We also note that a total of 36533 edges of the grafted tree could be altered by changing the ranks, and most of these edges are unaffected.
-
-\section{Conclusions}
-
-Here we have described the motivation and methodology used by our new
-supertree method that is currently used by the Open Tree of Life
-project to build summary supertrees on the scale of millions of leaves.
-Our new method represented 11\% more input phylogeny splits with 51\%
-less conflict compared to the Open Tree of Life version 4 summary
-tree, when applied to the same inputs. Unlike the previous method
-\citep{HinchliffEtAl2015}, our new method is guaranteed to incorporate
-input splits unless they conflict with the summary tree. The method is
-implemented in the Open Source software package propinquity.  A
-modified version of the treemachine software which built the summary
-tree described in the \citet{HinchliffEtAl2015} paper is used by the
-project to serve the tree produced by propinquity via Open Tree of
-Life APIs.
-
-Our supertree pipeline makes it possible to summarize and merge
-information from across the Tree of Life.  However, we do not claim
-that our supertree is more accurate than the input trees that it
-summarizes.  Before using the output of our supertree pipeline
-as input to an evolutionary analysis, researchers should first assess
-the accuracy of the output supertree in the clade of interest,
-possibly adjusting the ranks of the input trees to produce a more
-useful outcome.  We have demonstrated that ranks supplied by curators
-have substantial effects on the output of the pipeline.  This
-can be considered to be a benefit of our approach, since it allows
-curators to adjust the output of the pipeline, and it is easy to see
-why the supertree contains the relationships that it does.  However,
-one downside of our current approach is that every node in an input
-tree must recieve the same rank.  As a result, bootstrap proportions
-and posterior probabilities are not taken into account.
-
-The migration of summary tree construction from treemachine (used for
-version 4) to propinquity (for all versions from v5.0 to present) has
-increased the pace of synthesis tree releases from the Open Tree of
-Life project.  This is partly because the newly available annotations
-feature has made it possible to identify which input trees are
-responsible for taxa being included or excluded from the summary tree.
-Additionally, the new propinquity software pipeline has decreased
-the computational time required to construct a supertree from several
-hours to about 8 minutes (after some preprocessing steps which only
-have to be performed when the input taxonomy changes).  The amount of
-RAM required during tree construction has also decreased substantially.
-
-\section{Acknowledgements}
-
-Thanks to Emily Jane McTavish, Karen Cranston, Jonathan Rees, Jim Allman, Cody Hinchliff,
-Stephen Smith, and Joseph Brown for discussions and feedback. Thanks
-to NSF grant 1208393 (DEB), part of the collaborative Open Tree of
-Life awards, the University of Kansas, and the Heidelberg Institute
-for Theoretical Studies for funding.
-
-\bibliographystyle{upmplainnat}
-\bibliography{otcetera}
-
-\appendix
-
-\section{The maximum sum displayed groups' weighted scores criterion}\label{sec:The-maximum-sum}
-
-Let $\mathcal{W}$ is a weighting function that maps any input tree's
-internal node to a non-negative number. If $I(\mathbb{S},i,j)$ is
-an indicator function that is 1 if summary tree $\mathbb{S}$ displays
-the node $V(i,j)$ and 0 otherwise then: $SDGWS(\mathbb{S})=\sum_{i}\sum_{j}I(\mathbb{S},i,j)\mathcal{W}(i,j)$
-is the ``sum of displayed groups' weighted scores'' for a tree where
-$i$ indexes all of the input trees and $j$ indexes each non-root
-internal node in tree $i$. Preference for this tree is referred to
-as the maximum sum displayed groups' weighted scores criterion (MSDGWS
-criterion). The summary tree constructed by the propinquity pipeline
-is a greedy heuristic for finding a tree that maximizes this score
-when the weights for a node are determined by the tree's weight and
-the difference in weighting is so large that displaying one node from
-a highly ranked tree is preferred to displaying all of the nodes in
-the trees with lower rank.
-
-\section{Description of the decomposition algorithm of otc-uncontested-decompose}\label{sec:Decompose-algorithm}
-
-The input is the a ranked list of input trees and comprehensive taxonomy.
-
-\subsection{Creation of multigraph of the taxonomy with embedded trees}
-
-The tool creates a multigraph by starting with a graph isomorphic
-to the taxonomic tree. The nodes and edges created in this step will
-be referred to as the ``taxonomic graph.'' Next, we add nodes and
-edges to that graph in a procedure that we refer to as ``embedding''
-the input trees into the taxonomy. A node is introduced for each node
-in an input tree, and these nodes are mapped the MRCA nodes in the
-taxonomic graph. In other words, for any node $y$ in an input tree
-with a cluster of descendants, $\mathcal{C},$ we find the most tipward
-node $z$ in the taxonomic graph that is an ancestor to all of the
-taxa in $\mathcal{C}$; let $m(y)=z$ refer to this mapping, and $m^{\prime}(z)=y$
-refer to the reverse mapping. Each edge $e_{ij}$ in a source tree
-$i$ connects ancestor to its descendant, $a(e_{ij})\rightarrow d(e_{ij})$.
-The edges are introduced into the graph. We also introduce new edges
-to create a from $m(a(e_{ij}))$ through its descendants to $m(d(e_{ij}))$;
-we denote this path $p(e_{ij})$ and refer to the edges in the path
-as ``embedding edges for tree $i$''. Note, that this is a path
-through the taxonomic nodes, while the edge $e_{ij}$ connects source
-tree nodes. The mapping between $e_{ij}$ and $p(e_{ij})$ is stored,
-and the edges are labelled with the index $i$ so that it is clear
-which tree created them. Because the taxonomic tree is highly unresolved,
-it is frequently the case that $m(a(e_{ij}))$ is the same node as
-$m(d(e_{ij}))$; in these cases the embedding edge is a loop. This
-situation occurs whenever edge $e_{ij}$ can resolve part of the polytomy
-represented by a taxon.
-
-We treat the taxonomy as the lowest ranked input tree. The next step
-will collapse contested edges in the taxonomic graph. To retain all
-of the information from the taxonomy we embed the taxonomy into the
-taxonomic graph as if it were another input tree
-
-\subsection{Detection of uncontested higher taxa}
-
-After every input tree and the taxonomy tree have been embedded into
-the taxonomic graph, we perform postorder traversal over the taxonomic
-graph that underlies the multi-graph. For any internal node (each
-of which corresponds to a non-terminal taxon) we determine whether
-or not it is contested by examining each input tree. We can determine
-tree $i$ contests the taxon represented by taxonomic node $x$ by
-looking at the parents of all of the ``exiting'' embedding edges
-for tree $i$. These are the set of embedding edges that have $x$
-as a daughter and have a parent node that is not $x$ (ergo a parent
-node that is taxonomically higher than $x$). If there are more than
-one parent nodes in this set of exiting embedding edges, then tree
-$i$ contests that taxon. If there is only one parent node, then the all
-of the constituent taxa belonging to this taxon that are present in
-tree $i$ have one parent that is more inclusive; this means that
-the input tree does not contest monophyly of the taxon.
-
-If the taxon is uncontested, then it may be the case that some input
-trees do not contest the taxon, but contain polytomies that could
-be resolved to display the taxon. The cases can be identified by finding
-multiple exiting embedding edges that have the same taxon $y$ as
-their parent node. In these cases, a pseudo input tree node is created
-and becomes the parent node for these edges; this new node is then
-connected to $y$ as if it had been an input edge. This operation
-is equivalent to resolving an input tree's polytomy in favor of the
-monophyly of the uncontested taxon. This is the only way in which
-the input trees are modified during the decomposition.
-
-\subsection{Collapsing contested taxa from the taxonomic graph}
-
-If a taxonomic node $x$ fails the ``uncontested'' test described
-in the previous section, then the node corresponding to the taxon
-is removed from the taxonomic graph and the set of edges (and mappings
-between input edges and embedded paths) is updated as if this taxon
-had not been present in when the taxonomic graph was created. This
-consists of detecting changing any edge in an embedding path that
-is adjacent to $x$ by replacing the reference to $x$ with a reference
-to its parent $a(x)$. Note that we do not collapse the edge corresponding
-to this taxon in the part of the graph that represents the embedding
-of the taxonomy into the taxonomic graph. Thus, the taxonomy will
-still claim the monophyly of the taxon. This is relevant if the input
-grouping that contests taxon $x$ is overruled (in the subproblem
-solution step) by a higher ranked split. In other words, the fact
-that the a taxon is contested during the decomposition is not a guarantee
-that the taxon will not be monophyletic in the final supertree.
-
-\subsection{Emitting subproblems}
-
-Whenever an uncontested taxon is identified, the appropriate slice
-of each input trees that intersect with the taxon is written to a
-file. Then the multigraph is simplified by slicing any off the taxon.
-This slicing is accomplished by examing all of the exiting embedding
-edges for the taxon. The descendant taxa of each input tree is relabeled
-with the identifier of the contested taxa and all of that node's descendants
-are removed. Thus this input node will act as as if it were a leaf
-mapped to the uncontested taxon. All descendants of the taxonomic
-graph are also pruned off.
 \end{document}

### 2017-01-26 doc/paper1-submission/paper5-adjusted.tex
Unify Figure 1.
Unify Figure 2.
Unify Fig4
Unify Fig5
Unify Figure 6.
Convert Fig7
Convert Fig9
Unify Fig12

@@
 \begin{figure}
-\subfloat[Input tree 1]{\includegraphics[width=0.45\textwidth]{Fig1a}
-
-}\hfill{}\subfloat[Input tree 2]{\includegraphics[width=0.45\textwidth]{Fig1b}
-
-}
-
-\subfloat[Summary tree]{\includegraphics[width=0.45\textwidth]{Fig1c}
-
-}\hfill{}\subfloat[Summary tree]{\includegraphics[width=0.45\textwidth]{Fig1d}
-
-}
+\includegraphics[width=1.0\textwidth]{Fig1}
 
@@
 \begin{figure}
-\subfloat[Input tree 1]{\includegraphics[scale=0.57]{Fig2a}
-
-}\hfill{}\subfloat[Input tree 2]{\includegraphics[scale=0.57]{Fig2b}
-
-}\hfill{}\subfloat[Input tree 3]{\includegraphics[scale=0.57]{Fig2c}
-
-}
-
-\subfloat[Summary tree]{\includegraphics[scale=0.57]{Fig2d}
-
-}\hfill{}\subfloat[Summary tree]{\includegraphics[scale=0.57]{Fig2e}
-
-}\hfill{}\subfloat[Summary tree]{\includegraphics[scale=0.57]{Fig2f}
-
-}
+\includegraphics[width=1.0\textwidth]{Fig2}
 
@@
 \begin{figure}
-\subfloat[Tree 1]{\includegraphics[scale=0.4]{Fig4a}
-
-}\hfill{}\subfloat[Tree 2]{\includegraphics[scale=0.4]{Fig4b}
-
-}\hfill{}\subfloat[Taxonomy Tree]{\includegraphics[scale=0.4]{Fig4c}
-
-}
+\includegraphics[width=1.0\textwidth]{Fig4}
 
@@
 \begin{figure}
-\subfloat[Tree 1]{\includegraphics[scale=0.4]{Fig5a}
-
-}\hfill{}\subfloat[Tree 2]{\includegraphics[scale=0.4]{Fig5b}
-
-}\hfill{}\subfloat[Pruned taxonomy tree $\protect\prunedTaxonomy$]{\includegraphics[scale=0.4]{Fig5c}
-
-}
+\includegraphics[width=1.0\textwidth]{Fig5}
 
@@
 \begin{figure}
-  \subfloat[Subproblem \emph{ABCD}]{\includegraphics[width=0.3\textwidth]{Fig6a}%
-    \hfill{}\includegraphics[width=0.3\textwidth]{Fig6b}%
-    \hfill{}\includegraphics[width=0.3\textwidth]{Fig6c}
-
-}
-
-  \subfloat[Subproblem \emph{root}]{\includegraphics[width=0.3\textwidth]{Fig6d}%
-    \hfill{}\includegraphics[width=0.3\textwidth]{Fig6e}%
-    \hfill{}\includegraphics[width=0.3\textwidth]{Fig6f}
-
-}
+\includegraphics[width=1.0\textwidth]{Fig6}
 
@@
 \begin{figure}
-\subfloat[]{\includegraphics[scale=0.57]{Fig7a}
-
-}\hfill{}\subfloat[]{\includegraphics[scale=0.57]{Fig7b}
-
-}\hfill{}\subfloat[]{\includegraphics[scale=0.57]{Fig7c}
-
-}
-
-\hfill{}\subfloat[]{\includegraphics[scale=0.57]{Fig7d}
-
-}\hfill{}\subfloat[]{\includegraphics[scale=0.57]{Fig7e}
-
-}\hfill{}
+\includegraphics[width=1.0\textwidth]{Fig7}
 
@@
 \begin{figure}
-\hfill{}\subfloat[Solution to subproblem \emph{ABCD}]{\includegraphics[scale=0.55]{Fig9a}
-
-}\hfill{}\subfloat[Solution to subproblem \emph{root}]{\includegraphics[scale=0.55]{Fig9b}
-
-}\hfill{}
+\includegraphics[width=1.0\textwidth]{Fig9}
 
@@
 \begin{figure}
-\begin{centering}
-\subfloat[Taxonomy]{\begin{centering}
-\begin{minipage}[c][1\totalheight][t]{0.45\columnwidth}%
-\begin{center}
-\includegraphics[scale=0.38]{Fig12a}
-\par\end{center}%
-\end{minipage}
-\par\end{centering}
-}\hfill{}\subfloat[Grafted solution]{\begin{centering}
-\begin{minipage}[c][1\totalheight][t]{0.45\columnwidth}%
-\begin{center}
-\includegraphics[scale=0.38]{Fig12b}
-\par\end{center}%
-\end{minipage}
-\par\end{centering}
-}\hfill{}
-\par\end{centering}
-\begin{centering}
-\subfloat[Unpruned tree - broken taxa removed]{\begin{centering}
-\includegraphics[scale=0.38]{Fig12c}
-\par\end{centering}
-}\hfill{}\subfloat[Unpruned tree with MRCA-attachment]{\begin{centering}
-\includegraphics[scale=0.38]{Fig12d}
-\par\end{centering}
-}
-\par\end{centering}
+\includegraphics[width=1.0\textwidth]{Fig12}
+
 \caption{\label{fig:Two-approaches-to-unpruning}Two approaches to unpruning.

### 2017-01-30 doc/paper1-submission/paper5-adjusted.tex
Unify Fig13
Remove funding statements.
Show monotypic taxa in panel 5c.

@@
 taxa E and F are retained as monotypic taxa in the pruned taxonomy
-$\protect\prunedTaxonomy$, but are not shown in panel (c). The red
+$\protect\prunedTaxonomy$. The red
 edge in the pruned taxonomy tree is an uncontested higher taxon in
@@
 \begin{figure}
-\includegraphics[width=1.0\textwidth]{Fig7}
+\subfloat[]{\includegraphics[scale=0.57]{Fig7a}
+
+}\hfill{}\subfloat[]{\includegraphics[scale=0.57]{Fig7b}
+
+}\hfill{}\subfloat[]{\includegraphics[scale=0.57]{Fig7c}
+
+}
+
+\hfill{}\subfloat[]{\includegraphics[scale=0.57]{Fig7d}
+
+}\hfill{}\subfloat[]{\includegraphics[scale=0.57]{Fig7e}
+
+}\hfill{}
 
@@
 \begin{figure}
-\includegraphics[width=1.0\textwidth]{Fig9}
+\hfill{}\subfloat[Solution to subproblem \emph{ABCD}]{\includegraphics[scale=0.55]{Fig9a}
+
+}\hfill{}\subfloat[Solution to subproblem \emph{root}]{\includegraphics[scale=0.55]{Fig9b}
+
+}\hfill{}
 
@@
 \begin{figure}
-\includegraphics[width=1.0\textwidth]{Fig12}
-
+\begin{centering}
+\subfloat[Taxonomy]{\begin{centering}
+\begin{minipage}[c][1\totalheight][t]{0.45\columnwidth}%
+\begin{center}
+\includegraphics[scale=0.38]{Fig12a}
+\par\end{center}%
+\end{minipage}
+\par\end{centering}
+}\hfill{}\subfloat[Grafted solution]{\begin{centering}
+\begin{minipage}[c][1\totalheight][t]{0.45\columnwidth}%
+\begin{center}
+\includegraphics[scale=0.38]{Fig12b}
+\par\end{center}%
+\end{minipage}
+\par\end{centering}
+}\hfill{}
+\par\end{centering}
+\begin{centering}
+\subfloat[Unpruned tree - broken taxa removed]{\begin{centering}
+\includegraphics[scale=0.38]{Fig12c}
+\par\end{centering}
+}\hfill{}\subfloat[Unpruned tree with MRCA-attachment]{\begin{centering}
+\includegraphics[scale=0.38]{Fig12d}
+\par\end{centering}
+}
+\par\end{centering}
 \caption{\label{fig:Two-approaches-to-unpruning}Two approaches to unpruning.

### 2017-01-30 doc/paper1-submission/paper5-adjusted.tex
Remove appendices.

@@
 This operation is performed by the \texttt{otc-uncontested-decompose}
-tool in otcetera; see appendix \ref{sec:Decompose-algorithm} for
+tool in otcetera; see appendix B for
 a description of the algorithm. The output is a series of subproblems,
@@
 
-\appendix
-
-\section{The maximum sum displayed groups' weighted scores criterion}\label{sec:The-maximum-sum}
-
-Let $\mathcal{W}$ is a weighting function that maps any input tree's
-internal node to a non-negative number. If $I(\mathbb{S},i,j)$ is
-an indicator function that is 1 if summary tree $\mathbb{S}$ displays
-the node $V(i,j)$ and 0 otherwise then: $SDGWS(\mathbb{S})=\sum_{i}\sum_{j}I(\mathbb{S},i,j)\mathcal{W}(i,j)$
-is the ``sum of displayed groups' weighted scores'' for a tree where
-$i$ indexes all of the input trees and $j$ indexes each non-root
-internal node in tree $i$. Preference for this tree is referred to
-as the maximum sum displayed groups' weighted scores criterion (MSDGWS
-criterion). The summary tree constructed by the propinquity pipeline
-is a greedy heuristic for finding a tree that maximizes this score
-when the weights for a node are determined by the tree's weight and
-the difference in weighting is so large that displaying one node from
-a highly ranked tree is preferred to displaying all of the nodes in
-the trees with lower rank.
-
-\section{Description of the decomposition algorithm of otc-uncontested-decompose}\label{sec:Decompose-algorithm}
-
-The input is the a ranked list of input trees and comprehensive taxonomy.
-
-\subsection{Creation of multigraph of the taxonomy with embedded trees}
-
-The tool creates a multigraph by starting with a graph isomorphic
-to the taxonomic tree. The nodes and edges created in this step will
-be referred to as the ``taxonomic graph.'' Next, we add nodes and
-edges to that graph in a procedure that we refer to as ``embedding''
-the input trees into the taxonomy. A node is introduced for each node
-in an input tree, and these nodes are mapped the MRCA nodes in the
-taxonomic graph. In other words, for any node $y$ in an input tree
-with a cluster of descendants, $\mathcal{C},$ we find the most tipward
-node $z$ in the taxonomic graph that is an ancestor to all of the
-taxa in $\mathcal{C}$; let $m(y)=z$ refer to this mapping, and $m^{\prime}(z)=y$
-refer to the reverse mapping. Each edge $e_{ij}$ in a source tree
-$i$ connects ancestor to its descendant, $a(e_{ij})\rightarrow d(e_{ij})$.
-The edges are introduced into the graph. We also introduce new edges
-to create a from $m(a(e_{ij}))$ through its descendants to $m(d(e_{ij}))$;
-we denote this path $p(e_{ij})$ and refer to the edges in the path
-as ``embedding edges for tree $i$''. Note, that this is a path
-through the taxonomic nodes, while the edge $e_{ij}$ connects source
-tree nodes. The mapping between $e_{ij}$ and $p(e_{ij})$ is stored,
-and the edges are labelled with the index $i$ so that it is clear
-which tree created them. Because the taxonomic tree is highly unresolved,
-it is frequently the case that $m(a(e_{ij}))$ is the same node as
-$m(d(e_{ij}))$; in these cases the embedding edge is a loop. This
-situation occurs whenever edge $e_{ij}$ can resolve part of the polytomy
-represented by a taxon.
-
-We treat the taxonomy as the lowest ranked input tree. The next step
-will collapse contested edges in the taxonomic graph. To retain all
-of the information from the taxonomy we embed the taxonomy into the
-taxonomic graph as if it were another input tree
-
-\subsection{Detection of uncontested higher taxa}
-
-After every input tree and the taxonomy tree have been embedded into
-the taxonomic graph, we perform postorder traversal over the taxonomic
-graph that underlies the multi-graph. For any internal node (each
-of which corresponds to a non-terminal taxon) we determine whether
-or not it is contested by examining each input tree. We can determine
-tree $i$ contests the taxon represented by taxonomic node $x$ by
-looking at the parents of all of the ``exiting'' embedding edges
-for tree $i$. These are the set of embedding edges that have $x$
-as a daughter and have a parent node that is not $x$ (ergo a parent
-node that is taxonomically higher than $x$). If there are more than
-one parent nodes in this set of exiting embedding edges, then tree
-$i$ contests that taxon. If there is only one parent node, then the all
-of the constituent taxa belonging to this taxon that are present in
-tree $i$ have one parent that is more inclusive; this means that
-the input tree does not contest monophyly of the taxon.
-
-If the taxon is uncontested, then it may be the case that some input
-trees do not contest the taxon, but contain polytomies that could
-be resolved to display the taxon. The cases can be identified by finding
-multiple exiting embedding edges that have the same taxon $y$ as
-their parent node. In these cases, a pseudo input tree node is created
-and becomes the parent node for these edges; this new node is then
-connected to $y$ as if it had been an input edge. This operation
-is equivalent to resolving an input tree's polytomy in favor of the
-monophyly of the uncontested taxon. This is the only way in which
-the input trees are modified during the decomposition.
-
-\subsection{Collapsing contested taxa from the taxonomic graph}
-
-If a taxonomic node $x$ fails the ``uncontested'' test described
-in the previous section, then the node corresponding to the taxon
-is removed from the taxonomic graph and the set of edges (and mappings
-between input edges and embedded paths) is updated as if this taxon
-had not been present in when the taxonomic graph was created. This
-consists of detecting changing any edge in an embedding path that
-is adjacent to $x$ by replacing the reference to $x$ with a reference
-to its parent $a(x)$. Note that we do not collapse the edge corresponding
-to this taxon in the part of the graph that represents the embedding
-of the taxonomy into the taxonomic graph. Thus, the taxonomy will
-still claim the monophyly of the taxon. This is relevant if the input
-grouping that contests taxon $x$ is overruled (in the subproblem
-solution step) by a higher ranked split. In other words, the fact
-that the a taxon is contested during the decomposition is not a guarantee
-that the taxon will not be monophyletic in the final supertree.
-
-\subsection{Emitting subproblems}
-
-Whenever an uncontested taxon is identified, the appropriate slice
-of each input trees that intersect with the taxon is written to a
-file. Then the multigraph is simplified by slicing any off the taxon.
-This slicing is accomplished by examing all of the exiting embedding
-edges for the taxon. The descendant taxa of each input tree is relabeled
-with the identifier of the contested taxa and all of that node's descendants
-are removed. Thus this input node will act as as if it were a leaf
-mapped to the uncontested taxon. All descendants of the taxonomic
-graph are also pruned off.
 \end{document}

### 2017-01-30 doc/paper1-submission/paper5-adjusted.tex

@@
 \begin{figure}
-\subfloat[Input tree $T_{1}$ named ``tree1'']{%
-\begin{minipage}[c][1\totalheight][t]{0.25\columnwidth}%
-\includegraphics[scale=0.5]{Fig13a}%
-\end{minipage}
-
-}\hfill{}\subfloat[Summary Tree $\protect\summaryTree$]{%
-\begin{minipage}[c][1\totalheight][t]{0.35\columnwidth}%
-\includegraphics[scale=0.5]{Fig13b}%
-\end{minipage}
-
-}\hfill{}\subfloat[Induced Tree $\protect\summaryTree(1)$]{%
-\begin{minipage}[c][1\totalheight][t]{0.35\columnwidth}%
-\includegraphics[scale=0.5]{Fig13c}%
-\end{minipage}
-
-}
-
-\bigskip{}
-
-\bigskip{}
-
-\hfill{}\subfloat[JSON annotations relating edges of (c) to edges of (a)]{%
-\includegraphics{Fig13d}%
-}\hfill{}
+\includegraphics[width=1.0\textwidth]{Fig13}
 
@@
 Thanks to Emily Jane McTavish, Karen Cranston, Jonathan Rees, Jim Allman, Cody Hinchliff,
-Stephen Smith, and Joseph Brown for discussions and feedback. Thanks
-to NSF grant 1208393 (DEB), part of the collaborative Open Tree of
-Life awards, the University of Kansas, and the Heidelberg Institute
-for Theoretical Studies for funding.
+Stephen Smith, and Joseph Brown for discussions and feedback.
 

### 2017-03-15 bin/compare_synthesis_outputs.py
more deets on broken taxa

@@ -7,25 +7,42 @@ import simplejson as json
 import argparse
 import glob,os,re,csv
 import requests
-from peyotl import ott
+import peyotl.ott as ott
 
 def broken_taxa_diffs(bt1,bt2,verbose):
     compare_lists("Broken taxa",bt1,bt2,verbose)
 
 # writes details of the broken taxa to a file that can be input by
 # report_on_broken_taxa.py
-def newly_broken_taxa_report(bt1,bt2):
+def newly_broken_taxa_report(run1,run2):
+    # load local copy of OTT
+    taxonomy = ott.OTT()
+    id2names = taxonomy.ott_id_to_names
+    id2ranks = taxonomy.ott_id_to_ranks
     # print details of names in 2 but not in 1 (the 'newly broken names')
+    bt1=set(run1)
+    bt2=set(run2)
     diff = bt2.difference(bt1)
     broken_taxa_filename = 'broken_taxa_report.csv'
     print "printing details of {x} broken taxa to {f}".format(
         x=len(diff),
         f=broken_taxa_filename
         )
+    print "using OTT version {v}".format(v=taxonomy.version)
     with open(broken_taxa_filename, 'w') as f:
         for ottID in diff:
-            (name,rank)=get_taxon_details(ottID)
-            f.write("{n}_{i}\n".format(n=name,i=ottID))
+            # strip off the 'ott' part
+            pattern = re.compile(r'ott')
+            int_id = int(re.sub(pattern,'',ottID))
+            name = "no name"
+            rank = "no rank"
+            if int_id in id2names:
+                name = id2names[int_id]
+                if (isinstance(name,tuple)):
+                    name = name[0]
+                if int_id in id2ranks:
+                    rank = id2ranks[int_id]
+            f.write("{i},{n},{r}\n".format(i=int_id,n=name,r=rank))
     f.close()
 
 # generic function to compare two lists: number of items in each,
@@ -364,7 +381,7 @@ if __name__ == "__main__":
     print "\n# Comparing broken taxa"
     broken_taxa_diffs(run1.broken_taxa,run2.broken_taxa,args.verbose)
     if args.print_broken_taxa:
-        newly_broken_taxa_report(bt1,bt2)
+        newly_broken_taxa_report(run1.broken_taxa,run2.broken_taxa)
 
     print "\n# Synthetic tree summary"
     #synthesis_tree_diffs(args.run1,args.run2)

### 2017-03-15 bin/document_outputs.py
add names + ranks to broken taxa table

@@ -20,6 +20,7 @@ except ImportError:
     import ConfigParser as configparser
 from peyotl import read_as_json
 from peyotl.utility import propinquity_fn_to_study_tree
+import peyotl.ott as ott
 import subprocess
 import peyotl
 import codecs
@@ -274,6 +275,30 @@ def node_label2obj(nl):
         el['node_id'] = node_id
     return el
 
+# add OTT metadata to the monophyletic taxa blob
+def add_taxonomy_metadata(non_monophyletic_taxa):
+    broken_taxa = non_monophyletic_taxa['non_monophyletic_taxa']
+    taxonomy = ott.OTT()
+    id2names = taxonomy.ott_id_to_names
+    id2ranks = taxonomy.ott_id_to_ranks
+
+    for oid in broken_taxa:
+        pattern = re.compile(r'ott')
+        int_id = int(re.sub(pattern,'',oid))
+        name = "no name"
+        rank = "no rank"
+        if int_id in id2names:
+            name = id2names[int_id]
+            if (isinstance(name,tuple)):
+                name = name[0]
+            if int_id in id2ranks:
+                rank = id2ranks[int_id]
+        print(oid,name,rank)
+        broken_taxa[oid]['name'] = name
+        broken_taxa[oid]['rank'] = rank
+    non_monophyletic_taxa['non_monophyletic_taxa']=broken_taxa
+    return non_monophyletic_taxa
+
 class DocGen(object):
     def __init__(self, propinquity_dir, supertree_output_dir, config_filepath):
         self.top_output_dir = supertree_output_dir
@@ -320,6 +345,7 @@ class DocGen(object):
         blob.non_monophyletic_taxa = read_as_json(os.path.join(d, 'broken_taxa.json'))
         if blob.non_monophyletic_taxa['non_monophyletic_taxa'] is None:
             blob.non_monophyletic_taxa['non_monophyletic_taxa'] = {}
+        blob.non_monophyletic_taxa = add_taxonomy_metadata(blob.non_monophyletic_taxa)
         return blob
     def read_subproblem_solutions(self):
         d = os.path.join(self.top_output_dir, 'subproblem_solutions')

### 2018-02-19 doc/incertae_sedis_latex.tex
LaTeX export from LyX

%% LyX 2.2.3 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[english]{article}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{geometry}
\geometry{verbose,tmargin=2cm,bmargin=2cm,lmargin=2cm,rmargin=2cm}
\setlength{\parskip}{6pt}
\setlength{\parindent}{0pt}
\usepackage{babel}
\usepackage{url}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage[authoryear]{natbib}
\usepackage[unicode=true]
 {hyperref}

\makeatletter

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
%% Special footnote code from the package 'stblftnt.sty'
%% Author: Robin Fairbairns -- Last revised Dec 13 1996
\let\SF@@footnote\footnote
\def\footnote{\ifx\protect\@typeset@protect
    \expandafter\SF@@footnote
  \else
    \expandafter\SF@gobble@opt
  \fi
}
\expandafter\def\csname SF@gobble@opt \endcsname{\@ifnextchar[%]
  \SF@gobble@twobracket
  \@gobble
}
\edef\SF@gobble@opt{\noexpand\protect
  \expandafter\noexpand\csname SF@gobble@opt \endcsname}
\def\SF@gobble@twobracket[#1]#2{}
%% A simple dot to overcome graphicx limitations
\newcommand{\lyxdot}{.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
\sloppy
%\setlength{\parskip}{6pt}

\@ifundefined{showcaptionsetup}{}{%
 \PassOptionsToPackage{caption=false}{subfig}}
\usepackage{subfig}
\makeatother

\begin{document}
\global\long\def\taxonomy{\mbox{\ensuremath{\mathbb{T}}}}

\global\long\def\prunedTaxonomy{\taxonomy_{P}}

\global\long\def\phyloinputs{\mathcal{T}}

\global\long\def\expandedPhylo{\phyloinputs_{E}}

\global\long\def\summaryTree{\mathbb{S}}

\global\long\def\prunedSummary{\summaryTree_{P}}

\global\long\def\collections{\mathcal{C}}

\title{\texttt{Handling }\texttt{\emph{incertae sedis}}\texttt{ taxa in
propinquity}}
\maketitle

\section{Introduction}

The phrase \textquotedblleft \emph{incertae sedis}\textquotedblright ,
which is a Latin phrase meaning ``uncertain seat'', is used in classifications
to indicate that a group is of uncertain taxonomic placement. Such
groups arise in practice when they are classified within a parent
group, such as Fungi, but the precise location within the parent group
is uncertain. Such a taxon would then be attached as a child of the
parent group, but labeled as \emph{incertae sedis}''.

Currently \emph{incertae sedis} taxa and their descendants are excluded
from the OpenTree synthetic tree. This is problematic, because the
OpenTree project seeks to build a comprehensive tree covering all
of life. However, including unplaced taxa within the current synthesis
framework would be equally problematic, since placement of taxa within
groups is currently considered to be a conflict with those groups.
This conflict causes both a loss of the taxon name for the conflicting
taxon, and a loss of structure in the synthesis tree.

\subsection{Some examples in OTT\protect\footnote{This paragraph pulled from intro to otcetera/doc/handling-incertae-sedis.pdf}}

The problem is becoming more acute because NCBI is putting a larger
number of taxa into groups that are marked as \textquotedblleft unclassified.\textquotedblright{}
For example, when OTT 2.9 was created NCBI\textquoteright s clasification
of the bird family Sylvidae included a group five genera that were
placed within \textquotedblleft \href{http://www.ncbi.nlm.nih.gov/Taxonomy/Browser/wwwtax.cgi?mode=Undef&id=36270&lvl=3&lin=f&keep=1&srchmode=1&unlock}{unclassified Sylviidae}.\textquotedblright{}
This includes the genus Regulus. Thus in OTT 2.9 Regulus (ott ID =
3599326) is placed inside the family Silvidae (OTT ID = 259942). The
\textquotedblleft unclassified Sylviidae\textquotedblright{} does
not appear in OTT; instead Regulus is flagged as \textquotedblleft unclassified,sibling
higher\textquotedblright{} and all of the species within Regulus are
flagged as \textquotedblleft unclassified inherited.\textquotedblright{}

The taxonomy (as of version 2.9) also contains 685 cases of taxa that
are flagged as both \textquotedblleft unclassified inherited\textquotedblright{}
and \textquotedblleft unclassified\textquotedblright \footnote{based on \texttt{grep unclass.{*}unclassified\textbackslash{}\_inherited
taxonomy.tsv}}.

We would like to stop suppressing (pruning) incertae sedis taxa, so
that groups such as Regulus can appear appear in the synthetic tree.

\subsection{Handling incertae sedis taxa}

To solve this problem, we describe an extension of our previous supertree
synthesis procedure that can use phylogeny information to place \emph{incertae
sedis} taxa into containing taxa without considering the containing
taxa to be broken. This extension requires that we first describe
an extended semantics for taxa in the presences of \emph{incertae
sedis}, which is not a trivial problem. We describe both an updated
semantics for conflict with taxonomy edges, and an updated semantics
for assigning taxonomy names to nodes in the synthetic tree.

The ability to place taxa within the taxonomy without breaking their
containing taxa enables us to include thousands of new taxa that were
previously filtered out to avoid broken taxa. It also enables us to
use phylogenetic trees to update the taxonomy by extending taxa to
include \emph{incertae sedis} taxa placed within them. This makes
substantial progress towards our goal of complete representation of
taxa. It also enables us to include extinct taxa, since many of these
taxa are incertae sedis.

\subsection{Incertae sedis taxa in taxonomies}

\textbf{Flags.} The Open Tree\textquoteright s reference taxonomy
is produced by a tool called smasher that notices hints that a taxon
is incertae sedis and labels that taxon with one of five flags\footnote{see \url{https://github.com/OpenTreeOfLife/reference-taxonomy/wiki/Taxon-flags}.This
document is not concerned with the flagging system \emph{per se},
so \textquotedblleft \emph{incertae sedis}\textquotedblright{} will
be used here to refer to all of the flags that denote taxa with uncertain
placement.}. While a taxonomist may intend to use incertae sedis to indicate
a limited number of possible positions for a taxon to go in the taxonomy,
we do not retain any such details.

\textbf{Container nodes.} Some taxonomies have nodes with names like
``Incertae sedis (Bacteria)''. Such nodes are called containers.
They indicate that each child node is an \emph{incertae sedis} child
of the container node's parent. The OTT retains container nodes with
the \emph{was\_container} flag, but seems to have already moved all
of the container's children to the container's parent.

\textbf{Taxonomy merging.} Not all \emph{incertae sedis} taxa in our
taxonomy are directly labeled \emph{incertae sedis} by taxonomists.
\emph{Incertae sedis} taxa can also result from automatic merging
of taxonomies to create the OpenTree taxonomy. For example, in Figure
3 of \citet{rees2017automated}, cases \#4 and \#6 illustrate examples
where merging of two taxonomies leads to a taxonomy with a taxon of
uncertain placement. The reason is primarily that if taxonomy $\taxonomy_{1}$
contains more levels of hierarchy than taxonomy $\taxonomy_{2}$,
then we must add internal nodes to $\taxonomy_{2}$ to align it to
$\taxonomy_{1}$. However, if taxonomy $\taxonomy_{2}$ contains more
leaves than $\taxonomy_{1}$, then it is unclear if these extra leaves
should be nested inside the additional internal nodes, or not. Thus,
the extra leaves are marked \emph{incertae sedis}.

For example, if $\taxonomy_{1}=((a,b)x,(c,d)y)z$ and $\taxonomy_{2}=(a,b,c,d,e)z$
then $a$ and $b$ in $\taxonomy_{2}$ should be nested within $x$,
but we do not know if $e$ should be nested within $x$ or not. Thus
we obtain $((a,b)x,(c,d)y,?e)z$, where $?$ indicates that taxon
$e$ is marked \emph{incertae sedis}.

\subsection{Formalism}

We consider a ranked collection $\phyloinputs=\left\{ T_{1},\ldots,T_{n}\right\} $
of input trees and a single taxonomy tree $\taxonomy$. We have a
set of labels $\mathcal{L}$ that correspond to taxon names. Therefore
taxonomy node has a label, and every label correspond to a unique
taxonomy node.

In contrast, input phylogenies only have labels on their leaf nodes.
Furthermore, and the set of labels $\mathcal{L}(i)$ on input tree
$T_{i}$ need not include all leaf labels, and in practice always
contains only a relatively small part of $\mathcal{L}$.

We extend this framework by adding a set $\mathcal{I}\subseteq\mathcal{L}$
of labels that have the \emph{incertae sedis }property. The taxonomy
consists of the labels $\mathcal{L}$, the taxonomy tree $\taxonomy$,
and the \emph{incertae sedis} property $\mathcal{I}$ of taxon names.

\section{\label{sec:Semantics-of-incertae}Semantics of \emph{incertae sedis}
taxa}

Incertae sedis taxa affect the synthesis procedure in two ways: they
affect what it means for a taxonomy edge to conflict with the synthesis
tree, and they affect what it means to place a taxon name on a particular
node in the synthesis tree. We examine each of these effects in turn.

\subsection{Splits}

Each edge of a standard tree divides the tip taxa $\mathcal{L}$ into
two groups: the include set $\mathcal{I}(e)$ which does not contain
the root, and the exclude set $\mathcal{E}(e)$ which does contain
the root. Such a split may be written
\[
\mathcal{I}(e)|\bullet\mathcal{E}(e),
\]
with the exclude set always on the right. If no taxa are \emph{incertae
sedis}, then the exclude set for a node is just the total tip set
minus the include set for the node:
\begin{align*}
\mathcal{E}(n) & =\mathcal{L}-\mathcal{I}(n).
\end{align*}

For a node $n$ on the tipward side of an edge $e$, we may also write
$\mathcal{I}(n)$ for $\mathcal{I}(e)$, and $\mathcal{E}(n)$ for
$\mathcal{E}(e)$. We consider the exclude set of the root node to
be empty, and the include set of the root node to contain all tip
taxa:
\begin{align*}
\mathcal{E}(root) & =\{\}\\
\mathcal{I}(root) & =\mathcal{L}.
\end{align*}
In this case the root node stands for an edge that connects the root
to the root's parent.

\subsubsection{Reduced exclude sets for \emph{incertae sedis} taxa}

Marking a taxon as \emph{incertae sedis} changes the meaning of its
sibling taxa. Since the \emph{incertae sedis} taxon can be placed
inside its sibling subtrees, the exclude sets of the sibling subtrees
should not include the \emph{incertae sedis }taxon. Although exclude
sets of sibling taxa are shrunk, the include sets of the sibling taxa
are unaffected.

An \emph{incertae sedis} taxon can be moved into any of the descendants
of its siblings. Therefore, the exclude $\mathcal{E}(n)$ set for
a node $n$ contains the children of all the ancestors of $n$, unless
those children are incertae sedis. We can compute exclude sets recursively
by writing the exclude set of a node $n$ in terms of the exclude
set of its parent. If we use the terminology that the include and
exclude sets for a node $n$ are $\mathcal{I}(n)$ and $\mathcal{E}(n)$,
then we have 
\begin{align}
\mathcal{E}(n) & =\mathcal{E}(parent(n))\cup\left[\mathcal{I}(m)\big|m\in siblings(n),m\text{ not \emph{incertae} \emph{sedis}}\right],\label{eq:exclude-set-formula}
\end{align}
This leads to a pre-order recursion that terminates if the exclude
set for the root note is set to $\{\}$ as a boundary condition.

\subsubsection{Alternative formulation}

Consider an I.S. clade $A$ that is more rootward than an I.S. clade
$B$. With the above formulation, we could not place $A$ as sister
to $B$ and then place $B$ within $A$, because $A$ excludes $B$.
To solve this problem, we could modify formula (\ref{eq:exclude-set-formula})
above to avoid excluding \texttt{incertae\_sedis\_inherited} tips
of a sister taxon $m$. Thus, we could modify formula (\ref{eq:exclude-set-formula})
to refer to $\mathcal{I}^{\prime}(m)$ instead of $\mathcal{I}(m)$,
where 
\begin{align*}
\mathcal{I}^{\prime}(m) & =\begin{cases}
m & \text{if }m\text{ is a leaf}\\
\left[\mathcal{I}^{\prime}(c)\big|c\in children(m),c\text{ not \emph{incertae sedis}}\right] & \text{otherwise.}
\end{cases}
\end{align*}
This would not rule out placing $B$ as sister to $A$ and then placing
$A$ within $B$.

\subsection{Naming }

After solving a supertree (sub) problem, we need to assign taxon names
to the supertree nodes based on the taxonomy tree in the problem.
Each taxon name $n$ corresponds to a split $S(n)=S(n)_{1}|S(n)_{2}$
on the corresponding branch of the taxonomy tree. Without \emph{incertae
sedis}, such splits are always of the form $S(n)_{1}|\mathcal{L}-S(n)_{1}$,
but with \emph{incertae sedis} taxa $S_{2}(n)$ may be smaller than
$\mathcal{L}-S(n)_{1}$.

Without \emph{incertae sedis}, each name applies to at most one node,
and each node can take at most one name, with the exception of monotypic
taxa. Thus, we may simply search the solution tree for a node that
has the same cluster $S(n)_{1}$ and apply the name $n$ to that node.

However, in the \emph{incertae sedis }framework we must raise the
question of whether one name could apply to multiple nodes, or whether
multiple names could apply to one node.

\textbf{BDR:} \emph{It is well known that monotypic taxa are indistinguishable
if you consider only splits on leaf labels, but are distinguishable
if you consider splits on all node labels. It seems that some (all?)
of the problems with assigning multiple names to the same nodes actually
comes from the fact that moving IS taxa can leave the parent as a
degree-2 node. The fact that propinquity handles monotypic taxa indicates
that we actually implicitly consider all taxa to have labels. It is
possible that a trivial extension to the sub-problem solved could
thus handle monotypic taxa and issues with mapping 2 names to one
node with incertae sedis taxa.}

\subsubsection{Multiple nodes that fit one name}

Suppose the taxonomy is (((A1,A2)A,B)AB,C,D{*}) and the solution tree
is (((A1,A2)x,D{*})y,B)AB,C). In this case the name $A$ leads to
the split S(A) = A1 A2 | B C root. This name can apply to both the
node x and the node y.

\textbf{Solution:} In this case, we find the most tipward node and
attach the name to this node.

\subsubsection{Multiple names fit a single node}

\paragraph{Example 1}

Suppose the taxonomy is (((B1,B2)B,C{*})A,Y) and the input tree is
(((B1,C),B2)x,Y) then the names A and B both to the node x. In this
case the names A and B are ordered.

\textbf{Case 1:} If a taxon contains 2 non-IS taxa, then it cannot
be identical with any of its children in the synthesis tree.

\textbf{Case 2:} If a taxon contains 1 non-IS taxon and $\ge1$ IS
taxa, then the taxon could be identical with is non-IS child in the
synthesis tree, if the IS taxa are placed within the child.

\textbf{Case 3:} If a node contains 0 non-IS taxa and 1 IS taxon,
then the IS taxa behaves no differently than a non-IS taxa, since
is has no siblings it could be placed into.

\textbf{Case 4:} If a node contains 0 non-IS taxa and $\ge2$ IS taxa,
then then taxon \emph{could} be identical with a non-IS child in the
synthesis tree, if all but one IS children are placed with in one
of the IS children.

\textbf{Solution:} When we assign multiple names to the same node,
then we expand the node with multiple names to have monotypic parents,
and assign the series of names to the monotypic parents. Another way
of saying this is that when a node has $\le1$ non-IS taxon and $\ge2$
taxa then the node could become monotypic by placement of the IS taxa.

\textbf{Example 2}

If ((A1,A2)A,(B1,B2)B{*},(C1,C2)C{*}); is the taxonomy with asterisks
denoting incertae sedis taxa, then the solution ((A1,A2)A,((B1,C1)mrcaB1C1,(B2,C2)mrcaB2C2)x);
has a node x that could be called B{*} or C{*}.

\textbf{Case:} If a taxa B{*} and C{*} are IS, then their tips can
be intermingled in a new clade x. Both names would then apply.

\textbf{Solution:} ??

In summary, a taxon with split $A_{1}|\bullet B_{1}$ attaches to
the most tipward node $n$ where $A_{1}\subseteq S_{1}(n)$ and $B_{1}\subseteq S_{2}(n)$.
If multiple names end up on the same node, we try to resolve the problem
by creating a monotypic node for a name that is more rootward than
all other names at that node. If this fails to resolve the problem,
we arbitrarily choose one of the names.

\section{Synthesis and conflict resolution with incertae sedis taxa}

\subsection{Placement causes broken taxa}

Synthesis with \emph{incertae sedis} taxa has the potential to resolve
uncertain taxon placements using information from phylogenies. The
propinquity pipeline has always been able to do this kind of resolution.
However, without special consideration given to \emph{naming}, placing
a taxon $A$ within a taxon $B$ results in conflict with taxon $B$
in the taxonomy. In order to avoid a situation where input phylogenies
conflict with a large number of taxa when \emph{incertae sedis} taxa
are placed within them, we have filtered \emph{incertae sedis} taxa
from the taxonomy when constructing all prior synthesis trees.

When the synthesis tree conflicts with a taxonomy node, we way that
the taxon $B$ at that node is a broken taxon. Broken taxa have two
main effects, both of which are negative. First, the name $B$ of
the broken taxon is removed from the synthesis tree. Second, the conflicting
edge for taxon $B$ is not included in the synthesis tree. This means
that any children of $B$ that are taxonomy-only will not be placed
with the children of $B$ that are mentioned in input phylogenies.
Instead the taxonomy-only children of a broken taxon move towards
the root of the synthesis tree and attach at the first higher-ranked
taxon that is an ancestor of $B$ but is not broken.

\subsection{Correctly handling placement}

Thus, we seek a synthesis method that can correctly place taxa within
containing taxa without breaking the containing taxa. We change the
semantics of names to imply, not the exclusion of all non-included
taxa, but only some non-included taxa, as described in section \ref{sec:Semantics-of-incertae}.
As a result of this change in semantics, placing an IS taxon $A$
within a sister taxon $B$ no longer results in conflict with $B$.
This allows us to retain the split for $B$ within the synthesis tree,
so that taxonomy-only children of $B$ are correctly grouped with
their siblings that are referenced by the input trees. We may then
retain the name for the no-longer-broken taxon. Finally, we are then
able to stop filtering \emph{incertae sedis} taxa, so that they appear
in the synthesis tree. Thus, the synthesis tree is able to represent
substantially more species, without suffering the loss of taxa and
the loss of structure.

\subsection{Conflict with incertae sedis taxa}

\subsubsection{Conflicting placement among input trees}

The addition of \emph{incertae sedis} taxa allows new types of conflict
between input trees. For example, different input trees might place
an incertae sedis taxon in conflicting locations. This is illustrated
in Figure \ref{fig:placement-example1}, where the IS taxon (ott7,ott6)ott10
is placed as sister to ott1 by phylogeny $\tau_{1}$ and as sister
ott3 by phylogeny $\tau_{2}$.

When this happens, the placement of the IS taxon is not influenced
by its being marked IS on the taxonomy. Thus, in Example 1, the higher
ranked tree $\tau_{1}$ will be reflected in the synthesis tree, ott10
will be placed as sister to ott1. In contrast, the conflicting placement
in $\tau_{2}$ will not be reflected in the synth tree.

All this would occur in the previous version of propinquity. Where
the updated version differs that (a) the names ott9 and ott8 are retained
instead of being dropped. (b) as a result of not breaking ott8 and
ott9, we do not move ott3 and ott2 up to ott11. {[}BDR: Extend more
figures!{]}

\subsubsection{Example B}

However, incertae sedis taxa are not always tip nodes, but may themselves
contain other taxa. In such cases, it is possible for input trees
to conflict with the incertae sedis taxon itself. For example, consider
the following example
\begin{itemize}
\item $T_{1}=((w_{1},w_{2},z_{1}),y_{1})$
\item $T_{2}=((y_{1},y_{2},z_{2}),w_{1})$
\item Taxonomy tree $((w_{1},w_{2})w,(y_{1},y_{2})y,(z_{1},z_{2},z_{3})z)$
with $z$ marked \emph{incertae sedis}.
\begin{itemize}
\item splits $w_{1}w_{2}|\bullet y_{1}y_{2}$, $y_{1}y_{2}|\bullet w_{1}w_{2}$
and $z_{1}z_{2}|\bullet w_{1}w_{2}y_{1}y_{2}$
\end{itemize}
\end{itemize}
In this case, tree $T_{1}$ places $z_{1}$ within $w$, while $T_{2}$
places $z_{2}$ within $y$. Since different members of $z$ are placed,
neither placement for $z$ is rejected. Instead the taxon $z$ is
broken, the name $z$ disappears, and $z_{3}$ floats to the top level.

Furthermore, since taxon $z$ is a broken \emph{incertae sedis }taxon,
all of its children are effectively incertae sedis independently,
with the difference that they cannot be placed within each other.
Therefore, the names $w$ and $y$ are not lost, since the taxa placed
within them are \emph{incertae sedis} names. Note that $z_{3}$ maybe
therefore be placed in a third location, since it is also \emph{incertae
sedis}.

The situation here would be different if the monophyly of $z$ was
supported by an input phylogeny.

\emph{Note that the synthesis of all input trees before the taxonomy
is unaffected by incertae sedis information.}

\subsubsection{Example C - Nested \emph{incertae sedis} taxa}

In addition to incertae sedis taxa containing other taxa, it is also
possible for incertae sedis taxa to contain nested \emph{incertae
sedis} taxa. \textbf{What issues might this raise?}

\subsection{Placement}

Placement of incertae sedis taxa by input trees is unfortunately not
quite as simple as finding a single location where an I.S. taxon should
attach. For example, when an incertae sedis taxon is broken, its children
need to be ``placed'' separately.

Each input tree can relate to an \emph{incertae sedis} taxon $(A,B,C)D$
in a number of ways
\begin{itemize}
\item it could resolve $A$, $B$, $C$, or $D$.
\item it could place $D$ on a degree-2 (=out-degree-1) node that bisects
a branch
\item it could place a descendant taxon of $D$ 
\item it could place a descendant taxon of $D$ in a \emph{different place}
than another input tree.
\item it could place children of $D$ in multiple places, thus conflicting
with the branch. If the \emph{incertae sedis} taxon $(A,B,C)$ is
broken, then $A$, $B$ and $C$ become \emph{incertae sedis} clades
in their own right, that may attach separately, except that they .
This is because none of $A$, $B$, or $C$ is in the exclude set
of the siblings of $D$.
\end{itemize}

\section{Handling \emph{incertae sedis} taxa in the propinquity pipeline}

In order to handle \emph{incertae sedis} taxa within propinquity,
we must modify some of the stages of the propinquity pipeline. Subproblem
decomposition must place \emph{incertae sedis} taxa in the correct
subproblem. Subproblem files must indicate which taxa are incertae
sedis. The subproblem solver must read this information, account for
\emph{incertae sedis} taxa when solving subproblems, and correctly
name taxa that have been modified by having \emph{incertae sedis}
taxa place inside them. The unpruner must be aware of \emph{incertae
sedis} taxa. Annotations of the tree must be aware of \emph{incertae
sedis} taxa so that it does not consider taxa broken when they have
an incertae sedis taxon placed inside them.

\subsection{Exemplifying taxa}

One current problem is that well-known taxa like Fungi or Mammalia
tend to have a very large number of incertae sedis children, making
browsing in the tree viewer difficult. This can happen when, for example,
fossils or other hard-to-place taxa get classified only to the level
of these well-known nodes and no further. This leads to a situation
where well-known taxa serve as a dumping ground for unplaced taxa.

Our current approach to this problem is to perform a second round
of pruning, or ``cleaning'', during the exemplification step. Incertae
sedis taxa are pruned at this stage if they do not occur in any input
trees. We thus generate a second ``cleaned taxonomy'' that has undergone
this further round of cleaning. This approach improves on the previous
approach in that \emph{incertae sedis} taxa in input trees are no
longer pruned. This approach also removes tons of \emph{incertae sedis}
children from nodes like ``Fungi'', where a lot of unplaced fossils
with few observable characters have been dumped.

However, this approach has the negative effect of pruning some incertae
sedis taxa that need not be pruned. For example, suppose \emph{incertae
sedis} taxon $A$ contains 5 children, of which only 1 child $A_{1}$
occurs in an input tree. If the taxon $A$ is not broken, then it
should be possible to attach the other 4 members of $A$ next to $A_{1}$,
without cluttering up the synthesis tree. Such taxa have been successfully
placed even though they are not in any input tree. This can only be
discovered after synthesis is complete, though.

Additionally, it should also be possible to filter unplaced taxa in
the tree viewer instead of in the synthesis pipeline.

\subsection{Sub-problem decomposition}

The presence of \emph{incertae sedis }taxa poses a problem to sub-problem
decomposition, since taxonomy edges no longer completely separate
subproblems. Instead, \emph{incertae sedis} taxa may attach on either
side of a taxonomy edge. We seek to place \emph{incertae sedis} taxa
into subproblems in such a way that the subproblem solver can perform
the placement inside the subproblem. This approach postpones handling
of conflict in \emph{incertae sedis} taxa to the subproblem solver,
where the problem is well formulated in terms of splits. However,
it does have the effect of creating larger subproblems.

We must also handle conflicting placements of \emph{incertae sedis}
taxa by different input trees. Thus, if one input tree places the
\emph{incertae sedis} taxon $X$ in $((X)B)A$ and another places
$X$ in $((X)C)A$ then we must mark both edges $B$ and $C$ as contested
edges, even if these edges would \emph{not} be contested were taxon
$X$ to be removed. This results in a new way to contest edges that
involves the interaction of two input trees, and not just the interaction
of each input tree with the taxonomy.

We choose to solve these problems by merging any subproblems that
an \emph{incertae sedis} taxon might be placed in. The simplest way
to achieve this is simply to regard any taxon that has an \emph{incertae
sedis }taxon placed within it as contested. This results in marking
both $B$ and $C$ as contested edges in the example above. In fact,
this is the current behaviour of the non-\emph{incertae-sedis} aware
subproblem decomposer. One downside of this approach is that, if we
have $((X)B)A)$ in one input tree, and $X$ is mentioned nowhere
else, then by marking $B$ as contested, we are merging subproblems
unnecessarily. We could instead placed $X$ in $B$ and avoid contesting
the edge $B$. However, this approach is more complex and does not
seem necessary in practice.

\begin{figure}
\subfloat[Taxonomy $\protect\taxonomy$]{\includegraphics[scale=0.5]{Figures/placement1/tax\lyxdot tre}

}\subfloat[$\tau_{1}$]{\includegraphics[scale=0.5]{Figures/placement1/input1\lyxdot tre}

}\subfloat[$\tau_{2}$]{\includegraphics[scale=0.5]{Figures/placement1/input2\lyxdot tre}

}

\caption{\label{fig:placement-example1}Example. An incertae sedis clade (ott6,ott7)
is placed in different subtrees by input trees $\tau_{1}$ and $\tau_{2}$.
In $\tau_{1}$, two nodes that correspond to the taxonomy their ingroup
extended to include (ott6,ott7), and the branches leading to these
nodes have been colored blue. The dashed blue edge leads to a node
that is a newly-introduced degree-2 node which does not correspond
to any taxonomy node. In $\tau_{2}$, only one node that corresponds
to a taxonomy node needs to have its ingroup extended. The placement
of (ott6,ott7) into ott8 toward ott1 by $\tau_{1}$ conflicts with
the placement of (ott6,ott7) into ott9 toward ott3 by $\tau_{2},$}
\end{figure}
 For example, in Figure \ref{fig:placement-example1}, input tree
$\tau_{1}$ contests ott9 and input tree $\tau_{2}$ contests ott9
and ott8. Thus ott1, ott2, ott3, ott6, and ott7 end up in the same
sub-problem.

\subsection{Subproblem solution}

Our sub-problem solver naturally handles \emph{incertae sedis} taxa.
This is because we define the semantics of \emph{incertae sedis} taxa
in terms of partial splits, and our solver natively supports building
trees from partial splits through its use of the BUILD algorithm.
Handling \emph{incertae sedis} taxa thus requires loading incertae
sedis information and computing partial splits for \emph{incertae
sedis} taxa before solving a sub-problem. After solving a sub-problem,
we must apply taxon names from the taxonomy tree to the sub-problem
solution tree. The solution tree is considered to a fixed tree and
not to have any \emph{incertae sedis} nodes, or any other forms of
uncertainty.

\subsubsection{Reading incertae sedis information}

Currently, we read the \emph{incertae sedis} information as a list
of OTT ids for \emph{incertae sedis} taxa. This does not require adding
further annotations to the node names. Only taxonomy nodes can be
\emph{incertae sedis} at the moment, and only the taxonomy tree for
the subproblem contains OTT ids for internal nodes. Therefore we handle
\emph{incertae sedis} information by constructing modified split sets
for the lowest-ranked tree when the list of \emph{incertae sedis}
nodes is not empty.

\subsubsection{Exclude sets modified by \emph{incertae sedis} marks}

Equation (\ref{eq:exclude-set-formula}) leads to the following algorithm
to compute the exclude set for all nodes in a tree.
\begin{enumerate}
\item Set the exclude set of the root node to be empty
\item For each \emph{node} (except the root) in preorder
\begin{itemize}
\item combine the \emph{exclude} set of the parent node with the \emph{include}
set of non-\emph{incertae-sedis} siblings.
\item store this set in a hash, with key \emph{node}
\end{itemize}
\end{enumerate}
This algorithm is currently implemented in \emph{otc-solve-subproblem}.
We store the sets as \emph{std::set}.

\subsubsection{Implementation: finding the node for a name}

To find the node for a name $n$, we find the MRCA of the cluster
$S_{1}(n)$. If the MRCA excludes the entire exclude group $S_{2}(n)$
then the name applies to the MRCA; otherwise the taxon does not exist
on the tree.

\subsubsection{Implementation: handling name clashes}

When multiple names $N=\{n_{1},\ldots n_{N}\}$ map to the same solution
node $x$, then these names must satisfy some tree structure on the
taxonomy, such that $n_{1}<n_{2}$ if $n_{1}$ is a descendant of
$n_{2}$ in the taxonomy. If it is possible to find a name $n_{max}$
that is the unique maximal element of $N$, then it is permissible
to 
\begin{enumerate}
\item create a monotypic parent $p(x)$ of $x$, and assign $n_{max}$ to
$p(x)$
\item continue handling name clashes at $x$ with the set of possible names
reduced to $N-n_{max}$.
\end{enumerate}
However, its certainly possible that there might not be any such $N_{max}$,
in which case we could just choose a name for $x$ from $N$ (perhaps
not an \emph{incertae sedis} name) and then record all the other names
as equivalents somewhere.

\textbf{BDR:}\textbf{\emph{ }}\emph{we might get this behavior in
a nice an automatic way if we create a single fake leaf for each monotypic
taxonomy node that holds the node's leaf label.}

\subsubsection{Caveats}

When multiple I.S. taxa have been moved to the root node of a subproblem,
they may be I.S. over the entire subproblem, and some may be I.S.
over others in an asymmetric manner. Therefore, we might need to specify
additional information about the original attachment location of the
I.S. taxa, such as their depth. This only affects problems that have
been decomposed.

\textbf{BDR:} \emph{currently we don't actually move taxa to get them
into a subproblem. So, is this even an issue?}

\subsection{Grafted supertree}

\emph{Question:} Does the synthesis tree contain any \emph{incertae
sedis} groups?\\
\emph{Answer:} The grafted supertree will not contain any \emph{incertae
sedis} groups. However, when we attach pruned nodes to a parent in
the grafted supertree, we could mark such nodes \emph{incertae sedis}
if we want.

\subsection{Unpruning}

Currently the unpruner \emph{does not} require that the OTT ids are
named in the grafted solution before unpruning starts. According to
Mark's document, he wasn't sure if such names were generated for nodes
that had an IS taxon placed inside of them, so otc-unprune-solution-and-name-unnamed
nodes throws away all the names and generates them itself.

\textbf{BDR}: See document \texttt{otcetera/doc/unprune-solution-and-name-unnamed-nodes.pdf}

The unpruner should record when unpruned nodes are \emph{incertae
sedis}. Such nodes are unaffected by phylogenies, and so \emph{incertae
sedis} annotations for them make good sense.

\subsection{Annotation}

Annotation primarily involves running a conflict analysis between
the synthesis tree and each input tree. Since neither tree has any
\emph{incertae sedis} taxa, the conflict algorithm does not need to
change. Furthermore, if we allow \emph{incertae sedis} taxa that are
taxonomy-only to be annotated as \emph{incertae sedis} on the synth
tree, then such groups will not affect conflict with the input trees.
We would also like to allow running a conflict analysis between the
synthesis tree and the taxonomy tree. However, naming the nodes \emph{is}
a (almost) run of conflict analysis on the taxonomy tree, and this
has already been done in a prior step. So, the current annotation
procedure actually works as-is.

It would be nice to allow running conflict against the cleaned taxonomy,
though. One way to do this would be to generated a ``placed taxonomy'',
with groups extended to include \emph{incertae sedis} taxa that have
been placed within them. This would not require any updating to the
conflict-analysis code in the annotation step.

\subsection{Conflict service}

The current conflict service considers a group $A$ to conflict with
the taxonomy if group $A$ has an incertae sedis group $B$ placed
within it. This doesn't affect the annotations, since taxon names
are added by the unpruner. But it could make perfectly fine input
trees incorrectly look like they are the cause of broken taxa, if
they contain IS taxa. Thus, it would be nice to have a modified conflict
algorithm.

\subsubsection{Current conflict algorithm}

The current conflict algorithm is pretty fast, but it works by classifying
tips into either (i) the include group or (ii) the exclude group.
To avoid counting the exclude group for every split, we instead count
the total number of children for each node, and assume that any children
not in the include set are in the exclude set. This is no longer true
when we have incertae sedis taxa. I suspect that if we want to handle
incertae sedies, we'd need a third category (iii) for ``neither include
group nor exclude group''.
\begin{enumerate}
\item Get induced trees on intersection of leaf sets
\item Compute depth for each node (nd->depth)
\item Compute number of tips at or below each node (nd->n\_tips)
\item for each input tree node -> $nd$
\begin{enumerate}
\item skip the root
\item skip monotypic
\item if its a tip then find corresponding (``terminal'') edges in synth
tree and continue
\item leaves1 <- get the list of leaves in the include group of $nd$ (in
input)
\item L2 <- find the total number of tips (L2 = sum {[}nd->n\_tips| nd <-
leaves1{]})
\item leaves2 <- get list of corresponding synth leaf nodes (in synth)
\item nodes <- find all nodes between leaves2 and the MRCA (in synth)
\item MRCA <- mrca of leaves2 (in synth. this uses the nd->depth annotation)
\item Compute number of tips in the include set (nd->include\_tips) below
each node in $nodes$ (in synth)
\item if n\_include\_tips(MRCA) == n\_tips(MRCA) then the MRCA displays
$nd$
\item if n\_include\_tips(MRCA) < n\_tips(MRCA) then
\begin{itemize}
\item foreach node in nodes
\begin{itemize}
\item if (n\_include\_tipes(nd) < n\_tips(nd) and n\_include\_tips(nd) <
l2)
\begin{itemize}
\item this is a conflict!
\end{itemize}
\end{itemize}
\item if there are no conflicts, then this is a resolved\_by.
\end{itemize}
\end{enumerate}
\end{enumerate}

\subsubsection{Modified conflict algorithm?}

This probably is outside the scope of the paper, but if we could come
up with a modified conflict algorithm, that would be nice/useful,
and probably novel. It would also probably be slower...

\section{Results}

Should we do this? We could say: 
\begin{itemize}
\item we placed $x$ incertae sedis taxa.
\item we avoiding breaking $y$ taxa that had IS taxa placed inside them.
\item we allowed $z_{1}$ new taxa into the synthesis tree that were incertae
sedis.
\item we allowed $z_{2}$ new taxa into the synthesis tree that are marked
as extinct.
\item some nodes have as many as $w$\emph{ incertae sedis} children, making
them unbrowseable when incertae sedis children are not excluded.
\item $v_{1}$ input trees w were previously excluded \emph{entirely} because
they are nested within in an incertae sedis taxon.
\item $v_{2}$ input trees w were previously excluded \emph{partially} because
they are nested within in an incertae sedis taxon.
\end{itemize}
Currently the numbers $z_{1}$ and $z_{2}$.

\section{Discussion}

One thing we could do (perhaps) that we are not currently doing, is
to have nodes marked as incertae sedis on the synth tree. This would
be easy enough if such nodes are not affected in any way by the input
trees. Thus, when unpruning nodes we could mark any nodes \emph{incertae
sedis} if they were marked \emph{incertae sedis} on the taxonomy.

Secondly, I think we need to distinguish \emph{incertae sedis} taxa
that are ``unplaced'' from \emph{incertae sedis} taxa that do not
occur in any input tree. I think that if $A$ contains child $A_{1}$
that is an input tree, and the taxon $A$ is not broken, then $A$
will be placed, and thus any other children $A_{2},A_{3},\ldots,A_{n}$
will also be placed, since they will be added as children of the (placed)
node $A$ by the unpruner. This could be considered when deciding
which nodes to suppress in the tree viewer.

\bibliographystyle{upmplainnat}
\bibliography{otcetera}

\end{document}

### 2018-02-21 doc/incertae_sedis_latex2.tex
Export to TEX.

--- 2018-02-19 doc/incertae_sedis_latex.tex
@@
 \usepackage{babel}
+\usepackage{verbatim}
 \usepackage{url}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Textclass specific LaTeX commands.
+\newenvironment{lyxcode}
+{\par\begin{list}{}{
+\setlength{\rightmargin}{\leftmargin}
+\setlength{\listparindent}{0pt}% needed for AMS classes
+\raggedright
+\setlength{\itemsep}{0pt}
+\setlength{\parsep}{0pt}
+\normalfont\ttfamily}%
+ \item[]}
+{\end{list}}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
@@
 
-\title{\texttt{Handling }\texttt{\emph{incertae sedis}}\texttt{ taxa in
-propinquity}}
+\title{\texttt{\emph{Incertae sedis}}\texttt{ taxa in supertree construction}}
 \maketitle
@@
 
-Currently \emph{incertae sedis} taxa and their descendants are excluded
-from the OpenTree synthetic tree. This is problematic, because the
-OpenTree project seeks to build a comprehensive tree covering all
-of life. However, including unplaced taxa within the current synthesis
-framework would be equally problematic, since placement of taxa within
-groups is currently considered to be a conflict with those groups.
-This conflict causes both a loss of the taxon name for the conflicting
-taxon, and a loss of structure in the synthesis tree.
+The OpenTree project seeks to build a comprehensive supertree tree
+covering all of life. The approach is to combine information in published
+phylogenies with a comprehensive taxonomy that supplies taxon names.
+We describe an extension to our supertree method so that we can resolve
+taxonomic uncertainty by using published phylogenies to place \emph{incertae
+sedis} taxa. We describe an updated semantics for conflict between
+phylogeny edges and taxonomy edges and an updated semantics for assigning
+taxonomy names to nodes in the synthetic tree.
+
+The ability to place taxa within the taxonomy without breaking their
+containing taxa enables us to include thousands of new taxa that were
+previously filtered out to avoid broken taxa. It also enables us to
+use phylogenetic trees to update the taxonomy by extending taxa to
+include \emph{incertae sedis} taxa placed within them. This makes
+substantial progress towards our goal of complete representation of
+taxa. It also enables us to include extinct taxa, since many of these
+taxa are incertae sedis.
+
+\emph{TODO: add cases}
+
+\paragraph{Taxonomy merging}
+
+Not all \emph{incertae sedis} taxa in our taxonomy are directly labeled
+\emph{incertae sedis} by taxonomists. \emph{Incertae sedis} taxa can
+also result from automatic merging of taxonomies to create the OpenTree
+taxonomy. For example, in Figure 3 of \citet{rees2017automated},
+cases \#4 and \#6 illustrate examples where merging of two taxonomies
+leads to a taxonomy with a taxon of uncertain placement. The reason
+is primarily that if taxonomy $\taxonomy_{1}$ contains more levels
+of hierarchy than taxonomy $\taxonomy_{2}$, then we must add internal
+nodes to $\taxonomy_{2}$ to align it to $\taxonomy_{1}$. However,
+if taxonomy $\taxonomy_{2}$ contains more leaves than $\taxonomy_{1}$,
+then it is unclear if these extra leaves should be nested inside the
+additional internal nodes, or not. Thus, the extra leaves are marked
+\emph{incertae sedis}.
+
+For example, if $\taxonomy_{1}=((a,b)x,(c,d)y)z$ and $\taxonomy_{2}=(a,b,c,d,e)z$
+then $a$ and $b$ in $\taxonomy_{2}$ should be nested within $x$,
+but we do not know if $e$ should be nested within $x$ or not. Thus
+we obtain $((a,b)x,(c,d)y,?e)z$, where $?$ indicates that taxon
+$e$ is marked \emph{incertae sedis}.
+
+\paragraph{Formalism}
+
+Formalism. We consider a ranked collection $\phyloinputs=\left\{ T_{1},\ldots,T_{n}\right\} $
+of input trees and a single taxonomy tree $\taxonomy$. We have a
+set of labels $\mathcal{L}$ that correspond to taxon names. Therefore
+taxonomy node has a label, and every label correspond to a unique
+taxonomy node.
+
+In contrast, input phylogenies only have labels on their leaf nodes.
+Furthermore, and the set of labels $\mathcal{L}(i)$ on input tree
+$T_{i}$ need not include all leaf labels, and in practice always
+contains only a relatively small part of $\mathcal{L}$.
+
+We extend this framework by adding a set $\mathcal{I}\subseteq\mathcal{L}$
+of labels that have the \emph{incertae sedis }property. The taxonomy
+consists of the labels $\mathcal{L}$, the taxonomy tree $\taxonomy$,
+and the \emph{incertae sedis} property $\mathcal{I}$ of taxon names.
+
+\section{Details \& Methods}
 
@@
 
-\subsection{Handling incertae sedis taxa}
-
-To solve this problem, we describe an extension of our previous supertree
-synthesis procedure that can use phylogeny information to place \emph{incertae
-sedis} taxa into containing taxa without considering the containing
-taxa to be broken. This extension requires that we first describe
-an extended semantics for taxa in the presences of \emph{incertae
-sedis}, which is not a trivial problem. We describe both an updated
-semantics for conflict with taxonomy edges, and an updated semantics
-for assigning taxonomy names to nodes in the synthetic tree.
-
-The ability to place taxa within the taxonomy without breaking their
-containing taxa enables us to include thousands of new taxa that were
-previously filtered out to avoid broken taxa. It also enables us to
-use phylogenetic trees to update the taxonomy by extending taxa to
-include \emph{incertae sedis} taxa placed within them. This makes
-substantial progress towards our goal of complete representation of
-taxa. It also enables us to include extinct taxa, since many of these
-taxa are incertae sedis.
-
-\subsection{Incertae sedis taxa in taxonomies}
+\subsection{Taxonomy details}
 
@@
 
-\textbf{Taxonomy merging.} Not all \emph{incertae sedis} taxa in our
-taxonomy are directly labeled \emph{incertae sedis} by taxonomists.
-\emph{Incertae sedis} taxa can also result from automatic merging
-of taxonomies to create the OpenTree taxonomy. For example, in Figure
-3 of \citet{rees2017automated}, cases \#4 and \#6 illustrate examples
-where merging of two taxonomies leads to a taxonomy with a taxon of
-uncertain placement. The reason is primarily that if taxonomy $\taxonomy_{1}$
-contains more levels of hierarchy than taxonomy $\taxonomy_{2}$,
-then we must add internal nodes to $\taxonomy_{2}$ to align it to
-$\taxonomy_{1}$. However, if taxonomy $\taxonomy_{2}$ contains more
-leaves than $\taxonomy_{1}$, then it is unclear if these extra leaves
-should be nested inside the additional internal nodes, or not. Thus,
-the extra leaves are marked \emph{incertae sedis}.
-
-For example, if $\taxonomy_{1}=((a,b)x,(c,d)y)z$ and $\taxonomy_{2}=(a,b,c,d,e)z$
-then $a$ and $b$ in $\taxonomy_{2}$ should be nested within $x$,
-but we do not know if $e$ should be nested within $x$ or not. Thus
-we obtain $((a,b)x,(c,d)y,?e)z$, where $?$ indicates that taxon
-$e$ is marked \emph{incertae sedis}.
-
-\subsection{Formalism}
-
-We consider a ranked collection $\phyloinputs=\left\{ T_{1},\ldots,T_{n}\right\} $
-of input trees and a single taxonomy tree $\taxonomy$. We have a
-set of labels $\mathcal{L}$ that correspond to taxon names. Therefore
-taxonomy node has a label, and every label correspond to a unique
-taxonomy node.
-
-In contrast, input phylogenies only have labels on their leaf nodes.
-Furthermore, and the set of labels $\mathcal{L}(i)$ on input tree
-$T_{i}$ need not include all leaf labels, and in practice always
-contains only a relatively small part of $\mathcal{L}$.
-
-We extend this framework by adding a set $\mathcal{I}\subseteq\mathcal{L}$
-of labels that have the \emph{incertae sedis }property. The taxonomy
-consists of the labels $\mathcal{L}$, the taxonomy tree $\taxonomy$,
-and the \emph{incertae sedis} property $\mathcal{I}$ of taxon names.
-
 \section{\label{sec:Semantics-of-incertae}Semantics of \emph{incertae sedis}
@@
 
-Incertae sedis taxa affect the synthesis procedure in two ways: they
-affect what it means for a taxonomy edge to conflict with the synthesis
-tree, and they affect what it means to place a taxon name on a particular
-node in the synthesis tree. We examine each of these effects in turn.
+In order to incorporate \emph{incertae sedis }taxa into a supertree
+analyses, we first tackle the issue of the semantics of \emph{incertae
+sedis} taxa. We discuss various different things the term can mean
+within the literature, and settle on a specific usage here. We discuss
+how to move from an operational ``you can move this guy here'' semantics
+to a splits-based semantics. This affects both what it means for a
+taxonomy edge to conflict with the synthesis tree, and what it means
+to place a taxon name on a particular node in the synthesis tree.
+Finally, after a supertree has been constructed, we consider how to
+interpret the supertree as (perhaps) placing an \emph{incertae sedis}
+taxon $A$ within its sister taxon $B$.
+
+\subsection{Usage in the literature}
+
+The term \emph{incertae sedis} can be used in a number of different
+ways. For example,\emph{ }one author might describe \emph{incertae
+sedis} genera that are asserted to be monophyletic, whereas another
+author might describe \emph{incertae sedis} genera that may be interdigitated.
 
@@
 $A$ within $B$.
+
+\subsubsection{Comparison to operational definition}
+
+An definition in terms of moving node $A$ to position $x$ could
+cause trouble when moving multiple nodes, since the order of the movement
+would matter.
+
+Also, the splits-based semantics might not completely recover the
+edit-operation semantics. The corner cases are ... .
 
@@
 we arbitrarily choose one of the names.
+
+\subsection{Placement}
+
+After a synthesis tree is constructed, we attach taxon names for to
+the synthesis tree for each non-conflicting taxon. These names attach
+at the MRCA of the include group. Then, for each taxon $A$ on the
+synth tree, we associate with the first taxon $B$ found in a rootward
+walk on the synthesis tree. If $A$ is not a descendant of $B$ on
+the taxonomy, then we say that $A$ is \emph{placed} within $B$ by
+the synthesis tree. While this may occur if $A$ is \emph{incertae
+sedis}, more complex scenarios are possible. We note that if method
+does not handle \emph{incertae sedis} taxa, then this scenario could
+not occur because any taxa $B$ that $A$ is placed within would be
+considered incompatible with the synthesis tree and so their names
+would not be applied.
+
+In order to fully describe a placement of $A$ within $B$ we consider
+the sequence of taxa encountered on a walk from $A$ or $B$ up to
+their MRCA $M$ on the taxonomy tree. Let us denote the sequence of
+taxa encountered from $A$ as $A_{1},\ldots,A_{n},M$ where $A_{1}=A$.
+Let $B_{1},\ldots,B_{m},M$ be the sequence of taxa encountered from
+$B$ where $B_{1}=B$. Then $A_{n}$ must be \emph{incertae sedis},
+and $A_{2},\ldots,A_{n}$ must be incompatible with the synthesis
+tree (i.e. broken taxa). We then have that $A$ is placed successively
+within $B_{m},B_{m-1},\ldots,B_{2},B_{1}$, and $m$ is the depth
+of the placement of $A$.
+
+One question is whether this approach is able to handle nested incertae
+sedis taxa.
+
+Placement of incertae sedis taxa by input trees is unfortunately not
+quite as simple as finding a single location where an I.S. taxon should
+attach. For example, when an incertae sedis taxon is broken, its children
+need to be ``placed'' separately.
+
+Each input tree can relate to an \emph{incertae sedis} taxon $(A,B,C)D$
+in a number of ways
+\begin{itemize}
+\item it could resolve $A$, $B$, $C$, or $D$.
+\item it could place $D$ on a degree-2 (=out-degree-1) node that bisects
+a branch
+\item it could place a descendant taxon of $D$ 
+\item it could place a descendant taxon of $D$ in a \emph{different place}
+than another input tree.
+\item it could place children of $D$ in multiple places, thus conflicting
+with the branch. If the \emph{incertae sedis} taxon $(A,B,C)$ is
+broken, then $A$, $B$ and $C$ become \emph{incertae sedis} clades
+in their own right, that may attach separately, except that they .
+This is because none of $A$, $B$, or $C$ is in the exclude set
+of the siblings of $D$.
+\end{itemize}
 
@@
 sedis} taxa. \textbf{What issues might this raise?}
-
-\subsection{Placement}
-
-Placement of incertae sedis taxa by input trees is unfortunately not
-quite as simple as finding a single location where an I.S. taxon should
-attach. For example, when an incertae sedis taxon is broken, its children
-need to be ``placed'' separately.
-
-Each input tree can relate to an \emph{incertae sedis} taxon $(A,B,C)D$
-in a number of ways
-\begin{itemize}
-\item it could resolve $A$, $B$, $C$, or $D$.
-\item it could place $D$ on a degree-2 (=out-degree-1) node that bisects
-a branch
-\item it could place a descendant taxon of $D$ 
-\item it could place a descendant taxon of $D$ in a \emph{different place}
-than another input tree.
-\item it could place children of $D$ in multiple places, thus conflicting
-with the branch. If the \emph{incertae sedis} taxon $(A,B,C)$ is
-broken, then $A$, $B$ and $C$ become \emph{incertae sedis} clades
-in their own right, that may attach separately, except that they .
-This is because none of $A$, $B$, or $C$ is in the exclude set
-of the siblings of $D$.
-\end{itemize}
 
@@
 \section{Results}
+\begin{lyxcode}
+\begin{comment}
+\begin{lyxcode}
+?ott6145835~{[}6145835{]}~<-~ott698424~{[}698424{]}~placed~under~ott4942423~{[}4942423{]}~<-~ott698424~{[}698424{]}
 
+?ott6145873~{[}6145873{]}~<-~ott698424~{[}698424{]}~placed~under~ott4942423~{[}4942423{]}~<-~ott698424~{[}698424{]}~
+
+?ott18090~{[}18090{]}~<-~ott672501~{[}672501{]}~placed~under~ott275501~{[}275501{]}~<-~ott672501~{[}672501{]}
+
+?ott190269~{[}190269{]}~<-~ott672501~{[}672501{]}~placed~under~ott275501~{[}275501{]}~<-~ott672501~{[}672501{]}~
+
+?ott349638~{[}349638{]}~<-~ott672501~{[}672501{]}~placed~under~ott275501~{[}275501{]}~<-~ott672501~{[}672501{]}~
+
+?ott1009313~{[}1009313{]}~<-~ott672501~{[}672501{]}~placed~under~ott275501~{[}275501{]}~<-~ott672501~{[}672501{]}~
+
+?ott564110~{[}564110{]}~<-~ott939130~{[}939130{]}~placed~under~ott757344~{[}757344{]}~<-~ott939130~{[}939130{]}~
+
+?ott679485~{[}679485{]}~<-~ott1074933~{[}1074933{]}~placed~under~ott534796~{[}534796{]}~<-~ott1074933~{[}1074933{]}~
+
+?ott729184~{[}729184{]}~<-~ott1074933~{[}1074933{]}~placed~under~ott534796~{[}534796{]}~<-~ott1074933~{[}1074933{]}~
+
+?ott5144555~{[}5144555{]}~<-~ott23373~{[}23373{]}~placed~under~ott447802~{[}447802{]}~<-~ott23373~{[}23373{]}
+
+?ott441413~{[}441413{]}~<-~ott1014022~{[}1014022{]}~placed~under~ott571842~{[}571842{]}~<-~ott997381~{[}997381{]}~<-~ott1014022~{[}1014022{]}~
+
+?ott31033~{[}31033{]}~<-~ott627035~{[}627035{]}~placed~under~ott6019314~{[}6019314{]}~<-~ott627035~{[}627035{]}~
+
+?ott4018918~{[}4018918{]}~<-~ott5248084~{[}5248084{]}~placed~under~ott167767~{[}167767{]}~<-~ott5248084~{[}5248084{]}~
+
+?ott4792759~{[}4792759{]}~<-~ott5248084~{[}5248084{]}~placed~under~ott167767~{[}167767{]}~<-~ott5248084~{[}5248084{]}~
+\end{lyxcode}
+\end{comment}
+\end{lyxcode}
 Should we do this? We could say: 
 \begin{itemize}
-\item we placed $x$ incertae sedis taxa.
-\item we avoiding breaking $y$ taxa that had IS taxa placed inside them.
+\item we placed 14 incertae sedis taxa.
+\item we avoiding breaking 8 taxa that had IS taxa placed inside them.
 \item we allowed $z_{1}$ new taxa into the synthesis tree that were incertae
@@
 them unbrowseable when incertae sedis children are not excluded.
-\item $v_{1}$ input trees w were previously excluded \emph{entirely} because
+\item $v_{1}$ input trees were previously excluded \emph{entirely} because
 they are nested within in an incertae sedis taxon.
-\item $v_{2}$ input trees w were previously excluded \emph{partially} because
+\item $v_{2}$ input trees were previously excluded \emph{partially} because
 they are nested within in an incertae sedis taxon.

### 2018-03-01 doc/incertae_sedis_latex2.tex
Update.

@@
 
-The OpenTree project seeks to build a comprehensive supertree tree
-covering all of life. The approach is to combine information in published
-phylogenies with a comprehensive taxonomy that supplies taxon names.
+The OpenTree project seeks to build a comprehensive supertree covering
+all of life. The approach is to combine information in published phylogenies
+with a comprehensive taxonomy that supplies taxon names \citep{redelings2017supertree}.
 We describe an extension to our supertree method so that we can resolve
@@
 
-\paragraph{Taxonomy merging}
-
-Not all \emph{incertae sedis} taxa in our taxonomy are directly labeled
-\emph{incertae sedis} by taxonomists. \emph{Incertae sedis} taxa can
-also result from automatic merging of taxonomies to create the OpenTree
-taxonomy. For example, in Figure 3 of \citet{rees2017automated},
-cases \#4 and \#6 illustrate examples where merging of two taxonomies
-leads to a taxonomy with a taxon of uncertain placement. The reason
-is primarily that if taxonomy $\taxonomy_{1}$ contains more levels
-of hierarchy than taxonomy $\taxonomy_{2}$, then we must add internal
-nodes to $\taxonomy_{2}$ to align it to $\taxonomy_{1}$. However,
-if taxonomy $\taxonomy_{2}$ contains more leaves than $\taxonomy_{1}$,
-then it is unclear if these extra leaves should be nested inside the
-additional internal nodes, or not. Thus, the extra leaves are marked
-\emph{incertae sedis}.
-
-For example, if $\taxonomy_{1}=((a,b)x,(c,d)y)z$ and $\taxonomy_{2}=(a,b,c,d,e)z$
-then $a$ and $b$ in $\taxonomy_{2}$ should be nested within $x$,
-but we do not know if $e$ should be nested within $x$ or not. Thus
-we obtain $((a,b)x,(c,d)y,?e)z$, where $?$ indicates that taxon
-$e$ is marked \emph{incertae sedis}.
-
 \paragraph{Formalism}
@@
 In contrast, input phylogenies only have labels on their leaf nodes.
-Furthermore, and the set of labels $\mathcal{L}(i)$ on input tree
-$T_{i}$ need not include all leaf labels, and in practice always
-contains only a relatively small part of $\mathcal{L}$.
+Furthermore, the set of labels $\mathcal{L}(i)$ on input tree $T_{i}$
+need not include all leaf labels, and in practice always contains
+only a relatively small part of $\mathcal{L}$.
 
@@
 that groups such as Regulus can appear appear in the synthetic tree.
+
+\begin{comment}
+While we get Regulus back, we do not end up placing it in Sylviidae
+because Sylviidae is broken. It conflicts with about 20 different
+trees.
+\end{comment}
 
@@
 \section{Results}
+
+\subsection{Case 1}
+
+taxonomy = (((a1,a2)A,(b1,b2)B)AB,(c1,c2)?C,D)root;
+
+\subsubsection{If we place $c$ within $A$}
+
+tree1 = (((a1,c1),b1),d1);
+
+synth-with-is: ((((a1,(c1,c2)C),a2)A,(b1,b2)B)AB),(d1)D)
+
+placement: C within A <- AB
+
+synth-no-is: ((a1,(c1,c2)C),a1,(b1,b2)B,(d1)D)
+
+\subsubsection{If we place $c$ within $A$ but break $C$}
+
+tree1 = ((a1,c1),(a2,c1),b1)
+
+synth-with-is: ((((a1,c1),(a2,c2))A,(b1,b2)B)AB, (d1)D)
+
+synth-no-is: ??
+
+placement: 
+\begin{itemize}
+\item c1 <- (broken) C within A <- AB.
+\item c2<- (broken) C within A <- AB
+\end{itemize}
+result: we lose $C$, but keep $A$ and $AB$.
+
+\subsection{Case 2: nested is}
+
+taxonomy = (((a1,a2)A,(b1,b2)B),( ( (c11,c12)C1,(c21,c22)C2,(ci1,ci2)?CI
+)?C) ,(d1)D)
+
+tree1 = ((a1,(c11,(ci1,c12))),b1)
+
+synth-with-is: places CI wihin C1, and C within A.
+
+\subsection{Case 3: interleaving taxa}
+
+taxonomy = ((a1,a2)A,(b1,b2)?B,(c1,c2)?c)
+
+input tree: ((b1,c1),(b2,c2))
+
+synth-with-is: ((a1,a2)A,((b1,c1),(b2,c2))B) or ((a1,a2)A,((b1,c1),(b2,c2))C)
+
+Result: we name the interleaved taxon $B$ or $C$, but cannot name
+it both names.
+
+\subsection{Case 4: creation of monotypic taxa}
+
+see test cases.
 \begin{lyxcode}
@@
 \begin{lyxcode}
-?ott6145835~{[}6145835{]}~<-~ott698424~{[}698424{]}~placed~under~ott4942423~{[}4942423{]}~<-~ott698424~{[}698424{]}
+\textbf{Whale:}~?ott6145835~{[}6145835~Brandtocetus/genus{]}~<-~ott698424~{[}698424~Cetacea/order{]}~placed~under~ott4942423~{[}4942423~Cetotherium~/~genus{]}~<-~ott698424~{[}698424~Cetacea/order{]}
 
-?ott6145873~{[}6145873{]}~<-~ott698424~{[}698424{]}~placed~under~ott4942423~{[}4942423{]}~<-~ott698424~{[}698424{]}~
+\textbf{Whale:}~?ott6145873~{[}6145873{]}~<-~ott698424~{[}698424~Cetacea~/~order{]}~placed~under~ott4942423~{[}4942423{]}~<-~ott698424~{[}698424{]}~
 
-?ott18090~{[}18090{]}~<-~ott672501~{[}672501{]}~placed~under~ott275501~{[}275501{]}~<-~ott672501~{[}672501{]}
+-~Cetacea~broken.
 
-?ott190269~{[}190269{]}~<-~ott672501~{[}672501{]}~placed~under~ott275501~{[}275501{]}~<-~ott672501~{[}672501{]}~
+\textbf{Insect:}~?ott18090~{[}18090{]}~<-~ott672501~{[}672501{]}~placed~under~ott275501~{[}275501{]}~<-~ott672501~{[}672501{]}
 
-?ott349638~{[}349638{]}~<-~ott672501~{[}672501{]}~placed~under~ott275501~{[}275501{]}~<-~ott672501~{[}672501{]}~
+\textbf{Insect:}~?ott190269~{[}190269{]}~<-~ott672501~{[}672501{]}~placed~under~ott275501~{[}275501{]}~<-~ott672501~{[}672501{]}~
 
-?ott1009313~{[}1009313{]}~<-~ott672501~{[}672501{]}~placed~under~ott275501~{[}275501{]}~<-~ott672501~{[}672501{]}~
+\textbf{Insect:}~?ott349638~{[}349638{]}~<-~ott672501~{[}672501{]}~placed~under~ott275501~{[}275501{]}~<-~ott672501~{[}672501{]}~
 
-?ott564110~{[}564110{]}~<-~ott939130~{[}939130{]}~placed~under~ott757344~{[}757344{]}~<-~ott939130~{[}939130{]}~
+\textbf{Insect:}~?ott1009313~{[}1009313{]}~<-~ott672501~{[}672501~Reduviidae~/~family{]}~placed~under~ott275501~{[}275501{]}~<-~ott672501~{[}672501{]}~
 
-?ott679485~{[}679485{]}~<-~ott1074933~{[}1074933{]}~placed~under~ott534796~{[}534796{]}~<-~ott1074933~{[}1074933{]}~
+-~Reduviidae~broken.
 
-?ott729184~{[}729184{]}~<-~ott1074933~{[}1074933{]}~placed~under~ott534796~{[}534796{]}~<-~ott1074933~{[}1074933{]}~
+\textbf{Insect:~}?ott564110~{[}564110{]}~<-~ott939130~{[}939130~Thyrididae~/~family{]}~placed~under~ott757344~{[}757344{]}~<-~ott939130~{[}939130{]}~
 
-?ott5144555~{[}5144555{]}~<-~ott23373~{[}23373{]}~placed~under~ott447802~{[}447802{]}~<-~ott23373~{[}23373{]}
+\textbf{Plant:}~?ott679485~{[}679485{]}~<-~ott1074933~{[}1074933~Lamioidea~/~subfamily{]}~placed~under~ott534796~{[}534796{]}~<-~
 
-?ott441413~{[}441413{]}~<-~ott1014022~{[}1014022{]}~placed~under~ott571842~{[}571842{]}~<-~ott997381~{[}997381{]}~<-~ott1014022~{[}1014022{]}~
+\textbf{Plant:}~?ott729184~{[}729184{]}~<-~ott1074933~{[}1074933~Lamioidea~/~subfamily~{]}~placed~under~ott534796~{[}534796{]}~<-~ott1074933~{[}1074933{]}~
 
-?ott31033~{[}31033{]}~<-~ott627035~{[}627035{]}~placed~under~ott6019314~{[}6019314{]}~<-~ott627035~{[}627035{]}~
+ott1074933~{[}1074933{]}~
 
-?ott4018918~{[}4018918{]}~<-~ott5248084~{[}5248084{]}~placed~under~ott167767~{[}167767{]}~<-~ott5248084~{[}5248084{]}~
+-~Lamioidea~\emph{not}~broken
 
-?ott4792759~{[}4792759{]}~<-~ott5248084~{[}5248084{]}~placed~under~ott167767~{[}167767{]}~<-~ott5248084~{[}5248084{]}~
+\textbf{Plant:}~?ott5144555~{[}5144555{]}~<-~ott23373~{[}23373{]}~placed~under~ott447802~{[}447802{]}~<-~ott23373~{[}23373{]}
+
+-~Orobanchaceae~broken
+
+\textbf{Fly:}~?ott441413~{[}441413~Rhodesiella~lungliensis/species{]}~<-~ott1014022~{[}1014022~Rubiaceae{]}~placed~under~ott571842~{[}571842~Pavetteae~/~tribe{]}~<-~ott997381~{[}997381~Ixoroidea/subfamily{]}~<-~ott1014022~{[}1014022~Rubiaceae/family{]}
+
+\textbf{Plant:}~?ott31033~{[}31033{]}~<-~ott627035~{[}627035{]}~placed~under~ott6019314~{[}6019314{]}~<-~ott627035~{[}627035{]}~
+
+\textbf{Bacteria:}~?ott4018918~{[}4018918{]}~<-~ott5248084~{[}5248084{]}~placed~under~ott167767~{[}167767{]}~<-~ott5248084~{[}5248084{]}~
+
+\textbf{Bacteria:}?ott4792759~{[}4792759{]}~<-~ott5248084~{[}5248084{]}~placed~under~ott167767~{[}167767{]}~<-~ott5248084~{[}5248084{]}~
+
+-~Erysipelotrichaceae~broken
 \end{lyxcode}
@@
 \begin{itemize}
-\item we placed 14 incertae sedis taxa.
+\item there are 331 incertae sedis taxa that are mentioned in input trees.
+\begin{itemize}
+\item what is the kingom / phylum / class / order / genus / species
+\end{itemize}
+\item we placed 14 incertae sedis taxa inside a sister taxon.
+\item we confirmed $n$ incertae sedis taxa as being separate from all sister
+(sampled) taxa??
 \item we avoiding breaking 8 taxa that had IS taxa placed inside them.
@@
 
+\paragraph{Taxonomy merging}
+
+Not all \emph{incertae sedis} taxa in our taxonomy are directly labeled
+\emph{incertae sedis} by taxonomists. \emph{Incertae sedis} taxa can
+also result from automatic merging of taxonomies to create the OpenTree
+taxonomy. For example, in Figure 3 of \citet{rees2017automated},
+cases \#4 and \#6 illustrate examples where merging of two taxonomies
+leads to a taxonomy with a taxon of uncertain placement. The reason
+is primarily that if taxonomy $\taxonomy_{1}$ contains more levels
+of hierarchy than taxonomy $\taxonomy_{2}$, then we must add internal
+nodes to $\taxonomy_{2}$ to align it to $\taxonomy_{1}$. However,
+if taxonomy $\taxonomy_{2}$ contains more leaves than $\taxonomy_{1}$,
+then it is unclear if these extra leaves should be nested inside the
+additional internal nodes, or not. Thus, the extra leaves are marked
+\emph{incertae sedis}.
+
+For example, if $\taxonomy_{1}=((a,b)x,(c,d)y)z$ and $\taxonomy_{2}=(a,b,c,d,e)z$
+then $a$ and $b$ in $\taxonomy_{2}$ should be nested within $x$,
+but we do not know if $e$ should be nested within $x$ or not. Thus
+we obtain $((a,b)x,(c,d)y,?e)z$, where $?$ indicates that taxon
+$e$ is marked \emph{incertae sedis}.
+
 \bibliographystyle{upmplainnat}

### 2018-03-05 doc/strip_ott_labels.py
Add some figures for incertae sedis clade case.

@@ -1,0 +1,9 @@
+#!/usr/bin/python
+
+import sys
+import re
+
+for line in sys.stdin:
+    line = re.sub(r'_ott\d+','',line)
+    print(line)
+    

### 2018-03-05 doc/incertae_sedis_latex2.tex
Update document for revised definition of i.s. exclude sets.

@@
 
-\subsubsection{Reduced exclude sets for \emph{incertae sedis} taxa}
+\subsubsection{Partially reduced exclude sets for \emph{incertae sedis} taxa}
 
@@
 \begin{align}
-\mathcal{E}(n) & =\mathcal{E}(parent(n))\cup\left[\mathcal{I}(m)\big|m\in siblings(n),m\text{ not \emph{incertae} \emph{sedis}}\right],\label{eq:exclude-set-formula}
+\mathcal{E}(n) & =\mathcal{E}(parent(n))\cup\left[\mathcal{I}(m)\big|m\in siblings(n),m\text{ not \emph{incertae} \emph{sedis}}\right],\label{eq:exclude-set-formula-1}
 \end{align}
@@
 
-\subsubsection{Alternative formulation}
+\subsubsection{Extended formulation}
 
-Consider an I.S. clade $A$ that is more rootward than an I.S. clade
-$B$. With the above formulation, we could not place $A$ as sister
-to $B$ and then place $B$ within $A$, because $A$ excludes $B$.
-To solve this problem, we could modify formula (\ref{eq:exclude-set-formula})
-above to avoid excluding \texttt{incertae\_sedis\_inherited} tips
-of a sister taxon $m$. Thus, we could modify formula (\ref{eq:exclude-set-formula})
-to refer to $\mathcal{I}^{\prime}(m)$ instead of $\mathcal{I}(m)$,
-where 
+If an \emph{incertae sedis} taxon $A$ is placed within a sibling
+clade, then it may gain a new \emph{incertae sedis} sibling $B$.
+We could then possibly place $B$ within $A$. However, the above
+definition does not allow this possibility. We therefore adjust the
+above definition to exclude only descendants of sibling taxon only
+the path to the descendant does not pass through a node labeled as
+\emph{incertae sedis}. Such descendants of a sibling $m$ are given
+by:
 \begin{align*}
@@
 \end{align*}
-This would not rule out placing $B$ as sister to $A$ and then placing
-$A$ within $B$.
+We then modify the formula above to exclude $\mathcal{I}^{\prime}(m)$
+instead of $\mathcal{I}(m)$ for each sibling $m$:
+\begin{align}
+\mathcal{E}^{\prime}(n) & =\mathcal{E}^{\prime}(parent(n))\cup\left[\mathcal{I}^{\prime}(m)\big|m\in siblings(n),m\text{ not \emph{incertae} \emph{sedis}}\right],\label{eq:exclude-set-formula-2}
+\end{align}
+
+This less stringent formula is the one we will apply throughout this
+paper.
 
@@
 
-Equation (\ref{eq:exclude-set-formula}) leads to the following algorithm
-to compute the exclude set for all nodes in a tree.
+Equation (\ref{eq:exclude-set-formula-1}) leads to the following
+algorithm to compute the exclude set for all nodes in a tree.
 \begin{enumerate}
@@
 \end{itemize}
-\item we placed 14 incertae sedis taxa inside a sister taxon.
+\item we placed 14 incertae sedis taxa inside a sister taxon. (``incertae
+sedis'', ``versus ``unplaced''?)
+\item we placed $o$ incertae sedis taxa \emph{outside} all sister taxa.
 \item we confirmed $n$ incertae sedis taxa as being separate from all sister

### 2018-03-07 doc/Figures/is-within-is/draw.py
Update figures and doc.

@@ -1,0 +1,9 @@
+#!/usr/bin/python
+
+import sys
+from subprocess import call
+
+filename = sys.argv[1]
+
+call(["../../strip_ott_labels.py"], stdin=open(filename), stdout=open("temp","w"))
+call(["../../drawtree","--input=temp","--output={}.svg".format(filename),"--format=svg"])

### 2018-03-07 doc/incertae_sedis_latex2.tex
Update tex and pdf.

@@
 \usepackage{geometry}
-\geometry{verbose,tmargin=2cm,bmargin=2cm,lmargin=2cm,rmargin=2cm}
+\geometry{verbose,tmargin=3cm,bmargin=3cm,lmargin=3cm,rmargin=3cm}
 \setlength{\parskip}{6pt}
@@
 sedis} taxa. \textbf{What issues might this raise?}
+
+\section{Cases}
+
+When only tip nodes on the taxonomy are \emph{incertae sedis}, we
+have a very simple form of the \emph{incertae sedis} supertree problem.
+However, in practice entire clades may be \emph{incertae sedis}, and
+so more complex issue arise. In this section we consider a number
+of cases that must be handled when attempting to place \emph{incertae
+sedis} taxa.
+
+\subsection{Case 1: An \emph{incertae sedis} clade}
+
+When a clade is marked as \emph{incertae sedis}, we need to consider
+two cases. In the first case, the clade may be placed within a sister
+clade intact (Figure \ref{fig:Handling-incertae-sedis}). However,
+if the clade is not monophyletic in the synthesis tree, then we allow
+the members of the clade to be placed separately (Figure \ref{fig:Broken-incertae-sedis}).
+
+\begin{figure}
+\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-clade/tree1-shaded}
+
+}\hfill{}\subfloat[Taxonomy]{\includegraphics[scale=0.5]{Figures/is-clade/tax-shaded}
+
+}
+
+\subfloat[Synthesis Tree \#1]{\includegraphics[scale=0.5]{Figures/is-clade/synth1-no-is-shaded}
+
+}\hfill{}\subfloat[Synthesis Tree \#1]{\includegraphics[scale=0.5]{Figures/is-clade/synth1-shaded}
+
+}
+
+\caption{\label{fig:Handling-incertae-sedis}Handling \emph{incertae sedis
+}taxa recovers additional edges and taxon names. Supertree construction
+on input tree (a) and taxonomy tree (b) with \emph{incertae sedis}
+clade $C$ leads to synthesis tree (c). However, supertree construction
+with \emph{incertae sedis} handling constructs tree (d), which recovers
+taxon name $AB$ and $A$, as well as the edge to clade $AB$. Taxon
+names and edges that are conditional on handling \emph{incertae sedis}
+are shaded grey. }
+
+\end{figure}
+
+\begin{figure}
+
+\hfill{}\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-clade/tree2-shaded}
+
+}\hfill{}\subfloat[Taxonomy]{\includegraphics[scale=0.5]{Figures/is-clade/tax-shaded}
+
+}\hfill{}
+
+\hfill{}\subfloat[Synthesis Tree \#1]{\includegraphics[scale=0.5]{Figures/is-clade/synth2-shaded}
+
+}\hfill{}
+
+\caption{Broken incertae sedis clade\label{fig:Broken-incertae-sedis}. Taxonomy
+tree (b) contains \emph{incertae sedis }taxon $C$ that is broken
+by input tree (a). Supertree construction results in a synthesis tree
+(c) that places $C1$ and $C2$ separately within $A$.}
+
+\end{figure}
+
+\subsection{Case 2: A nested \emph{incertae sedis} clade}
+
+When \emph{incertae sedis} clades are allowed, it is possible for
+an \emph{incertae sedis} taxon to be nested within another \emph{incertae
+sedis} taxon. In such a case we must allow the higher-ranked taxon
+to be placed within a sibling, while preserving the right of the lower
+ranked taxon to be placed within a lower-level sibling (Figure \#)
+
+\subsection{Case 3: Placement of one incertae sedis taxon within another.}
+
+In our semantics of incertae sedis taxa, it is possible for to place
+one \emph{incertae sedis} sibling inside another. A special case of
+this is when a higher-rank incertae sedis taxon $A$ is placed as
+sister to an incertae sedis taxon $B$, and then $B$ is placed within
+$A$ (Figure \#).
 

### 2018-03-15 doc/incertae_sedis_latex2.tex
Update doc.

@@
-%% LyX 2.2.3 created this file.  For more info, see http://www.lyx.org/.
+%% LyX 2.3.0 created this file.  For more info, see http://www.lyx.org/.
 %% Do not edit unless you really know what you are doing.
@@
 \setlength{\parindent}{0pt}
-\usepackage{babel}
 \usepackage{verbatim}
-\usepackage{url}
 \usepackage{amsmath}
@@
 \usepackage[authoryear]{natbib}
-\usepackage[unicode=true]
- {hyperref}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
-%% Special footnote code from the package 'stblftnt.sty'
-%% Author: Robin Fairbairns -- Last revised Dec 13 1996
-\let\SF@@footnote\footnote
-\def\footnote{\ifx\protect\@typeset@protect
-    \expandafter\SF@@footnote
-  \else
-    \expandafter\SF@gobble@opt
-  \fi
-}
-\expandafter\def\csname SF@gobble@opt \endcsname{\@ifnextchar[%]
-  \SF@gobble@twobracket
-  \@gobble
-}
-\edef\SF@gobble@opt{\noexpand\protect
-  \expandafter\noexpand\csname SF@gobble@opt \endcsname}
-\def\SF@gobble@twobracket[#1]#2{}
 %% A simple dot to overcome graphicx limitations
@@
 
+\usepackage{babel}
 \begin{document}
@@
 
-\title{\texttt{\emph{Incertae sedis}}\texttt{ taxa in supertree construction}}
+\title{\texttt{Supertree construction with}\texttt{\emph{ Incertae sedis}}\texttt{
+taxa}}
 \maketitle
@@
 The phrase \textquotedblleft \emph{incertae sedis}\textquotedblright ,
-which is a Latin phrase meaning ``uncertain seat'', is used in classifications
-to indicate that a group is of uncertain taxonomic placement. Such
-groups arise in practice when they are classified within a parent
-group, such as Fungi, but the precise location within the parent group
-is uncertain. Such a taxon would then be attached as a child of the
-parent group, but labeled as \emph{incertae sedis}''.
+which is a Latin phrase meaning ``uncertain seat'', indicates that
+a taxon is of uncertain placement. Such groups arise in practice when
+they are classified within a parent group, such as Fungi, but the
+precise location within the parent group is uncertain. Such a taxon
+would then be attached as a child of the parent group, but labeled
+as \emph{incertae sedis}''.
 
@@
 with a comprehensive taxonomy that supplies taxon names \citep{redelings2017supertree}.
-We describe an extension to our supertree method so that we can resolve
+We describe an extension to this supertree method so that we can resolve
 taxonomic uncertainty by using published phylogenies to place \emph{incertae
@@
 
-The ability to place taxa within the taxonomy without breaking their
-containing taxa enables us to include thousands of new taxa that were
-previously filtered out to avoid broken taxa. It also enables us to
-use phylogenetic trees to update the taxonomy by extending taxa to
-include \emph{incertae sedis} taxa placed within them. This makes
-substantial progress towards our goal of complete representation of
-taxa. It also enables us to include extinct taxa, since many of these
-taxa are incertae sedis.
-
-\emph{TODO: add cases}
-
-\paragraph{Formalism}
-
-Formalism. We consider a ranked collection $\phyloinputs=\left\{ T_{1},\ldots,T_{n}\right\} $
-of input trees and a single taxonomy tree $\taxonomy$. We have a
-set of labels $\mathcal{L}$ that correspond to taxon names. Therefore
-taxonomy node has a label, and every label correspond to a unique
-taxonomy node.
-
-In contrast, input phylogenies only have labels on their leaf nodes.
-Furthermore, the set of labels $\mathcal{L}(i)$ on input tree $T_{i}$
-need not include all leaf labels, and in practice always contains
-only a relatively small part of $\mathcal{L}$.
-
-We extend this framework by adding a set $\mathcal{I}\subseteq\mathcal{L}$
-of labels that have the \emph{incertae sedis }property. The taxonomy
-consists of the labels $\mathcal{L}$, the taxonomy tree $\taxonomy$,
-and the \emph{incertae sedis} property $\mathcal{I}$ of taxon names.
-
-\section{Details \& Methods}
-
-\subsection{Some examples in OTT\protect\footnote{This paragraph pulled from intro to otcetera/doc/handling-incertae-sedis.pdf}}
-
-The problem is becoming more acute because NCBI is putting a larger
-number of taxa into groups that are marked as \textquotedblleft unclassified.\textquotedblright{}
-For example, when OTT 2.9 was created NCBI\textquoteright s clasification
-of the bird family Sylvidae included a group five genera that were
-placed within \textquotedblleft \href{http://www.ncbi.nlm.nih.gov/Taxonomy/Browser/wwwtax.cgi?mode=Undef&id=36270&lvl=3&lin=f&keep=1&srchmode=1&unlock}{unclassified Sylviidae}.\textquotedblright{}
-This includes the genus Regulus. Thus in OTT 2.9 Regulus (ott ID =
-3599326) is placed inside the family Silvidae (OTT ID = 259942). The
-\textquotedblleft unclassified Sylviidae\textquotedblright{} does
-not appear in OTT; instead Regulus is flagged as \textquotedblleft unclassified,sibling
-higher\textquotedblright{} and all of the species within Regulus are
-flagged as \textquotedblleft unclassified inherited.\textquotedblright{}
-
-The taxonomy (as of version 2.9) also contains 685 cases of taxa that
-are flagged as both \textquotedblleft unclassified inherited\textquotedblright{}
-and \textquotedblleft unclassified\textquotedblright \footnote{based on \texttt{grep unclass.{*}unclassified\textbackslash{}\_inherited
-taxonomy.tsv}}.
-
-We would like to stop suppressing (pruning) incertae sedis taxa, so
-that groups such as Regulus can appear appear in the synthetic tree.
-
-\begin{comment}
-While we get Regulus back, we do not end up placing it in Sylviidae
-because Sylviidae is broken. It conflicts with about 20 different
-trees.
-\end{comment}
-
-\subsection{Taxonomy details}
-
-\textbf{Flags.} The Open Tree\textquoteright s reference taxonomy
-is produced by a tool called smasher that notices hints that a taxon
-is incertae sedis and labels that taxon with one of five flags\footnote{see \url{https://github.com/OpenTreeOfLife/reference-taxonomy/wiki/Taxon-flags}.This
-document is not concerned with the flagging system \emph{per se},
-so \textquotedblleft \emph{incertae sedis}\textquotedblright{} will
-be used here to refer to all of the flags that denote taxa with uncertain
-placement.}. While a taxonomist may intend to use incertae sedis to indicate
-a limited number of possible positions for a taxon to go in the taxonomy,
-we do not retain any such details.
-
-\textbf{Container nodes.} Some taxonomies have nodes with names like
-``Incertae sedis (Bacteria)''. Such nodes are called containers.
-They indicate that each child node is an \emph{incertae sedis} child
-of the container node's parent. The OTT retains container nodes with
-the \emph{was\_container} flag, but seems to have already moved all
-of the container's children to the container's parent.
+The ability to place taxa within the taxonomy without breaking containing
+taxa enables us to include thousands of new taxa that were previously
+filtered out to avoid broken taxa. It also enables us to use published
+phylogenies to update the taxonomy by extending taxa to include \emph{incertae
+sedis} taxa placed within them. This makes substantial progress towards
+our goal of complete representation of taxa. It also enables us to
+include extinct taxa, since many of these taxa are incertae sedis.
+
+\paragraph{Background}
+
+We briefly describe the supertree algorithm proposed in \citet{redelings2017supertree}.
+This supertree algorithm takes as input a ranked list $\phyloinputs=\left\{ T_{1},\ldots,T_{n}\right\} $
+of rooted input trees and a single rooted taxonomy tree $\taxonomy$.
+The taxonomy tree is considered by ranked below any of the input trees.
+Each node $i$ in the taxonomy tree is associated a taxon name $n(i)$
+in the set of taxon names $\mathcal{N}$, and each taxon name is corresponds
+to only one node of the taxonomy tree. Thus the taxonomy is comprehensive
+and contains all taxa. In contrast, input phylogenies only have labels
+on their leaf nodes, and generally include a small fraction of the
+total number of taxa. Furthermore, taxa in input trees need not correspond
+to leaf taxa on the taxonomy tree, and multiple tips may be labeled
+with the same taxon name. \citet{redelings2017supertree} described
+a method called 'exemplification' that reduces this problem to a simpler
+form where all taxon names of input trees (i) correspond to leaf taxa
+and (ii) occur only once in an input tree. For the purposes of this
+paper, we therefore assume that this is the case.
+
+Given the exemplified input trees and taxonomy, the supertree algorithm
+first divides the supertree problem into independent pieces, solves
+the problem on these pieces, and then glues the pieces back together.
+This is made possible by the assumption that any taxonomy edges that
+are not contested by any of the input trees can be assumed to occur
+in the supertree. We may then solve separately for what happens on
+each side of such an uncontested edge. When the uncontested edge is
+attached to the solution below the edge and the solution above the
+edge, the complete solution is then obtained.
+
+Supertrees for each subproblem are obtained by greedily constructing
+a set of rooted splits $\mathcal{C}$ that are jointly consistent
+and correspond to branches of the input trees. We construct an ordered
+list of splits $S(T_{i})$ by walking the branches of tree $T_{i}$
+in post-order and appending the corresponding splits to the list.
+These lists combined to form the list $\Sigma=S(T_{1})+S(T_{2})+\ldots+S(T_{n})+S(\mathbb{T})$,
+where '$+$' denotes concatenation. The set $C$ is initialized to
+the empty set, and we then iteratively consider each split in the
+list $\Sigma$ and insert it into $C$ if the resulting set remains
+jointly compatible. The BUILD algorithm is used both to check compatibility,
+and to construct a compatible tree from the final set $C$.
 
@@
 taxon $A$ within its sister taxon $B$.
+
+\subsection{Notation}
+
+We extend this framework by adding a set $\mathcal{I}\subseteq\mathcal{N}$
+of labels that have the \emph{incertae sedis }property. The taxonomy
+consists of the labels $\mathcal{N}$, the taxonomy tree $\taxonomy$,
+and the \emph{incertae sedis} property $\mathcal{I}$ of taxon names.
 

### 2018-03-15 doc/incertae_sedis_latex2.tex
Update TEX and PDF.

@@
 of rooted input trees and a single rooted taxonomy tree $\taxonomy$.
-The taxonomy tree is considered by ranked below any of the input trees.
-Each node $i$ in the taxonomy tree is associated a taxon name $n(i)$
-in the set of taxon names $\mathcal{N}$, and each taxon name is corresponds
+The taxonomy tree is ranked below any of the input trees. Each node
+$i$ in the taxonomy tree is associated a taxon name $n(i)$ in the
+set of taxon names $\mathcal{N}$, and each taxon name is corresponds
 to only one node of the taxonomy tree. Thus the taxonomy is comprehensive
@@
 on their leaf nodes, and generally include a small fraction of the
-total number of taxa. Furthermore, taxa in input trees need not correspond
-to leaf taxa on the taxonomy tree, and multiple tips may be labeled
+total number of taxa. Furthermore, taxa in input trees may correspond
+to higher taxa on the taxonomy tree, and multiple tips may be labeled
 with the same taxon name. \citet{redelings2017supertree} described
@@
 
-\subsection{Splits}
+\subsection{A splits-based semantics for \emph{incertae sedis} taxa}
+
+\subsubsection{Splits}
 
@@
 \[
-\mathcal{I}(e)|\bullet\mathcal{E}(e),
+\mathcal{I}(e)|\bullet\mathcal{E}(e).
 \]
-with the exclude set always on the right. If no taxa are \emph{incertae
+where the $\bullet$ indicates the root. If no taxa are \emph{incertae
 sedis}, then the exclude set for a node is just the total tip set
@@
 
-\subsubsection{Partially reduced exclude sets for \emph{incertae sedis} taxa}
+\subsubsection{Reduced exclude sets allow \emph{incertae sedis} to be placed into
+sibling taxa}
 
 Marking a taxon as \emph{incertae sedis} changes the meaning of its
-sibling taxa. Since the \emph{incertae sedis} taxon can be placed
-inside its sibling subtrees, the exclude sets of the sibling subtrees
-should not include the \emph{incertae sedis }taxon. Although exclude
-sets of sibling taxa are shrunk, the include sets of the sibling taxa
-are unaffected.
+sibling taxa. We seek to represent this by constructing modified splits
+for each branch of the taxonomy tree. The include sets of these splits
+remain unchanged, but we construct reduced exclude sets to indicate
+cases where \emph{incertae sedis} taxa are allowed to intrude into
+a taxon.
 
 An \emph{incertae sedis} taxon can be moved into any of the descendants
-of its siblings. Therefore, the exclude $\mathcal{E}(n)$ set for
-a node $n$ contains the children of all the ancestors of $n$, unless
-those children are incertae sedis. We can compute exclude sets recursively
-by writing the exclude set of a node $n$ in terms of the exclude
-set of its parent. If we use the terminology that the include and
-exclude sets for a node $n$ are $\mathcal{I}(n)$ and $\mathcal{E}(n)$,
-then we have 
+of its siblings. Therefore, the exclude set $\mathcal{E}(n)$ for
+node $n$ should not include any of its siblings marked \emph{incertae
+}sedis\emph{,} but should exclude the descendants of its other siblings\emph{.}
+Additionally, the exclude $\mathcal{E}(n)$ should contain the children
+of all the ancestors of $n$, unless those children are incertae sedis.
+
+We may also consider what happens if an \emph{incertae sedis} taxon
+$A$ is placed within a sister group. Suppose $A$ is placed next
+to a sibling $B$. If $B$ is \emph{incertae sedis}, then it should
+be possible to place $B$ within $A$. As a result, $A$ should not
+exclude descendants of $B$ than can only be reached from $A$ by
+passing through a node marked \emph{incertae sedis. }
+
+We can compute exclude sets recursively by writing the exclude set
+of a node $n$ in terms of the exclude set of its parent:
 \begin{align}
-\mathcal{E}(n) & =\mathcal{E}(parent(n))\cup\left[\mathcal{I}(m)\big|m\in siblings(n),m\text{ not \emph{incertae} \emph{sedis}}\right],\label{eq:exclude-set-formula-1}
+\mathcal{E}(n) & =\mathcal{E}(parent(n))\cup\left[\mathcal{I^{\prime}}(m)\big|m\in siblings(n)\right],\label{eq:exclude-set-formula-1}
 \end{align}
-This leads to a pre-order recursion that terminates if the exclude
-set for the root note is set to $\{\}$ as a boundary condition.
-
-\subsubsection{Extended formulation}
-
-If an \emph{incertae sedis} taxon $A$ is placed within a sibling
-clade, then it may gain a new \emph{incertae sedis} sibling $B$.
-We could then possibly place $B$ within $A$. However, the above
-definition does not allow this possibility. We therefore adjust the
-above definition to exclude only descendants of sibling taxon only
-the path to the descendant does not pass through a node labeled as
-\emph{incertae sedis}. Such descendants of a sibling $m$ are given
-by:
+where $\mathcal{I}^{\prime}(n)$ indicates the descendants of $n$
+that are reachable from $n$ without passing through an \emph{incertae
+sedis} node:
 \begin{align*}
-\mathcal{I}^{\prime}(m) & =\begin{cases}
-m & \text{if }m\text{ is a leaf}\\
-\left[\mathcal{I}^{\prime}(c)\big|c\in children(m),c\text{ not \emph{incertae sedis}}\right] & \text{otherwise.}
+\mathcal{I}^{\prime}(n) & =\begin{cases}
+\emptyset & \text{if }n\text{ is \emph{incertae sedis}}\\
+\{n\} & \text{if }n\text{ is a leaf node}\\
+\left[\mathcal{I^{\prime}}(m)\big|m\in children(n)\right] & \text{if }n\text{ is an internal node}
 \end{cases}
 \end{align*}
-We then modify the formula above to exclude $\mathcal{I}^{\prime}(m)$
-instead of $\mathcal{I}(m)$ for each sibling $m$:
-\begin{align}
-\mathcal{E}^{\prime}(n) & =\mathcal{E}^{\prime}(parent(n))\cup\left[\mathcal{I}^{\prime}(m)\big|m\in siblings(n),m\text{ not \emph{incertae} \emph{sedis}}\right],\label{eq:exclude-set-formula-2}
-\end{align}
-
-This less stringent formula is the one we will apply throughout this
-paper.
-
-\subsubsection{Comparison to operational definition}
-
-An definition in terms of moving node $A$ to position $x$ could
-cause trouble when moving multiple nodes, since the order of the movement
-would matter.
-
-Also, the splits-based semantics might not completely recover the
-edit-operation semantics. The corner cases are ... .
+This recursion terminates if the exclude set for the root node is
+set to $\emptyset$ as a boundary condition.
 
@@
 to be placed within a sibling, while preserving the right of the lower
-ranked taxon to be placed within a lower-level sibling (Figure \#)
+ranked taxon to be placed within a lower-level sibling (Figure \ref{fig:Nested-incertae-sedis}).
+
+\begin{figure}
+\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-nested/tree3-shaded}
+
+}\hfill{}\subfloat[Taxonomy tree]{\includegraphics[scale=0.5]{Figures/is-nested/tax-shaded}
+
+}
+
+\subfloat[Synthesis tree (\emph{incertae sedis} ignored)]{\includegraphics[scale=0.5]{Figures/is-nested/synth3-no-is-shaded}
+
+}\hfill{}\subfloat[Synthesis tree (\emph{incertae sedis} considered)]{\includegraphics[scale=0.5]{Figures/is-nested/synth3-is-shaded}
+
+}
+
+\caption{\label{fig:Nested-incertae-sedis}Nested incertae sedis taxa. (a)
+Input tree \#1 places CI1 within CG, and C within $A$. (b) In the
+taxonomy, CI is \emph{incertae sedis }within C, while C is \emph{incertae
+sedis} under the root. (c) The synthesis tree when \emph{incertae
+sedis} taxa are not considered. The names AB, A, and CG are lost.
+(d) The synthesis tree when \emph{incertae sedis} taxa are considered.
+The names AB, A, and CG are regained, as well as the edges leading
+to AB and CG.}
+
+\end{figure}
 
@@
 $A$ (Figure \#).
+
+\emph{This case makes placement more complicated. Use a sequence of
+ordered placements, preorder?}
+\begin{figure}
+
+\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-within-is/tree1\lyxdot tre}
+
+}\subfloat[Taxonomy]{\includegraphics[scale=0.5]{Figures/is-within-is/tax-shaded}
+
+}\subfloat[Synthesis tree]{\includegraphics[scale=0.5]{Figures/is-within-is/synth1-shaded}
+
+}
+
+\caption{Placing one incertae sedis group within another. Here I is placed
+within A, and then AI is placed within A.}
+
+\end{figure}
 
@@
 
+\subsubsection{Comparison to operational definition}
+
+The splits-based semantics for incertae sedis taxa has many desirable
+properties. In allows the use of the BUILD algorithm to assess compatibility
+of split sets. However, it also has some properties that may not be
+expected. For example, two incertae sedis siblings can be freely interdigitated,
+which might not be the expected outcome. The current definition, also
+ignore ranks. Thus if we had two incertae sedis siblings that were
+genera, but other siblings were families, then one might expect that
+the genus-level incertae sedis taxa could be placed within families,
+but not intermixed. Since the current approach ignores ranks, it cannot
+do that.
+
+The above definition of incertae sedis taxa does allow two incertae
+sedis sister taxa to be interdigitated. It also
+
+Also, the splits-based semantics might not completely recover the
+edit-operation semantics. The corner cases are ... .
+
 \bibliographystyle{upmplainnat}

### 2018-03-20 doc/incertae_sedis_latex2.tex
Update ms.

@@
 
+\section{New Introduction}
+
+Supertree methods combine a collection of input trees with different
+taxon sets into a single tree on the combined taxon set. While such
+methods are usually applied to leaf-labeled phylogenies that have
+been estimated from DNA sequence data, we additionally seek to include
+taxonomic information. We thus require that one of the input trees
+is the taxonomy tree. The taxonomy has a number of features not found
+in the input trees, such as having labels for internal nodes, containing
+monotypic nodes, and containing \emph{incertae sedis} nodes. We refer
+to labels of taxonomy nodes as ``taxa'', and require that labels
+and taxonomy nodes have a one-to-one correspondence. We additionally
+require that the taxonomy be comprehensive and contain all labels.
+
+One of the problems that arises when attempting to include taxonomic
+information is the problem of \emph{incertae sedis }taxa. \emph{Incertae
+sedis} taxa are taxa that have unknown taxonomic placement. The common
+interpretation of such taxa is that they may not be place further
+root-ward, but may be placed within their sibling taxa. This frequently
+occurs when a specimen is identified to a taxonomic level such as
+the family level but not further. It also occurs when it is unclear
+whether a genus forms its own family, or is part of an existing family.
+(Mark rewrite?)
+
 \section{Introduction}
@@
 would then be attached as a child of the parent group, but labeled
-as \emph{incertae sedis}''.
-
-The OpenTree project seeks to build a comprehensive supertree covering
-all of life. The approach is to combine information in published phylogenies
-with a comprehensive taxonomy that supplies taxon names \citep{redelings2017supertree}.
-We describe an extension to this supertree method so that we can resolve
-taxonomic uncertainty by using published phylogenies to place \emph{incertae
-sedis} taxa. We describe an updated semantics for conflict between
-phylogeny edges and taxonomy edges and an updated semantics for assigning
-taxonomy names to nodes in the synthetic tree.
-
-The ability to place taxa within the taxonomy without breaking containing
-taxa enables us to include thousands of new taxa that were previously
-filtered out to avoid broken taxa. It also enables us to use published
-phylogenies to update the taxonomy by extending taxa to include \emph{incertae
-sedis} taxa placed within them. This makes substantial progress towards
-our goal of complete representation of taxa. It also enables us to
-include extinct taxa, since many of these taxa are incertae sedis.
+as \emph{incertae sedis}.
+
+The Open Tree of Life project seeks to build a comprehensive supertree
+covering all of life. The approach is to combine information in published
+phylogenies with a comprehensive taxonomy that supplies taxon names
+\citep{redelings2017supertree}. We describe an extension to this
+supertree method so that we can resolve taxonomic uncertainty by using
+published phylogenies to place \emph{incertae sedis} taxa. We describe
+an updated semantics for conflict between phylogeny edges and taxonomy
+edges and an updated semantics for assigning taxonomy names to nodes
+in the synthetic tree.
+
+Our approach to \emph{incertae sedis} taxa enables us to use published
+phylogenies to update the taxonomy by extending taxonomic groups to
+include \emph{incertae sedis} taxa placed within them. The ability
+to update the taxonomy is important, since the placement of \emph{incertae
+sedis} taxa creates artificial conflicts with a non-updated taxonomy.
+Thus, the ability to correctly assess compatibility with the taxonomy
+when the taxonomy contains \emph{incertae sedis taxa} allows us to
+include thousands of new taxa that were previously filtered out to
+avoid conflict. It also enables us to include extinct taxa, since
+many of these taxa are \emph{incertae sedis}. This makes substantial
+progress towards our goal of completely representation of taxa.
 
@@
 This supertree algorithm takes as input a ranked list $\phyloinputs=\left\{ T_{1},\ldots,T_{n}\right\} $
-of rooted input trees and a single rooted taxonomy tree $\taxonomy$.
-The taxonomy tree is ranked below any of the input trees. Each node
-$i$ in the taxonomy tree is associated a taxon name $n(i)$ in the
-set of taxon names $\mathcal{N}$, and each taxon name is corresponds
-to only one node of the taxonomy tree. Thus the taxonomy is comprehensive
-and contains all taxa. In contrast, input phylogenies only have labels
-on their leaf nodes, and generally include a small fraction of the
-total number of taxa. Furthermore, taxa in input trees may correspond
-to higher taxa on the taxonomy tree, and multiple tips may be labeled
-with the same taxon name. \citet{redelings2017supertree} described
-a method called 'exemplification' that reduces this problem to a simpler
-form where all taxon names of input trees (i) correspond to leaf taxa
-and (ii) occur only once in an input tree. For the purposes of this
-paper, we therefore assume that this is the case.
+of rooted input phylogenies and a single rooted taxonomy tree $\taxonomy$.
+Each node in the taxonomy tree is associated with a taxon name in
+the set of taxon names $\mathcal{N}$, and each taxon name is corresponds
+to only one node of the taxonomy tree. We use \emph{$\mathcal{L}$
+}to indicate taxon names corresponding to leaf nodes of the taxonomy
+tree. Correspondence between the input trees and the taxonomy is established
+by annotating leaf nodes of each input tree with taxon names. For
+purposes of conflict resolution, the taxonomy tree is ranked below
+all of the input trees.
+
+\paragraph{Exemplification}
+
+While the taxonomy is comprehensive and contains all taxa, input phylogenies
+only have labels on their leaf nodes, and generally include a small
+fraction of the total number of taxa. Furthermore, taxa in input trees
+may correspond to higher taxa on the taxonomy tree, and multiple tips
+may be labeled with the same taxon name. \citet{redelings2017supertree}
+described a method called 'exemplification' that reduces this problem
+to a simpler form where all taxon names of input trees (i) correspond
+to leaf taxa and (ii) occur only once in an input tree. For the purposes
+of this paper, we therefore assume that this is the case.
+
+\paragraph{Problem decomposition}
 
@@
 
+\paragraph{Splits-based synthesis method}
+
 Supertrees for each subproblem are obtained by greedily constructing
@@
 and correspond to branches of the input trees. We construct an ordered
-list of splits $S(T_{i})$ by walking the branches of tree $T_{i}$
+list of splits $S(T_{i})$ by walking the branches of each tree $T_{i}$
 in post-order and appending the corresponding splits to the list.
-These lists combined to form the list $\Sigma=S(T_{1})+S(T_{2})+\ldots+S(T_{n})+S(\mathbb{T})$,
+These lists are combined to form the list $\Sigma=S(T_{1})+S(T_{2})+\ldots+S(T_{n})+S(\mathbb{T})$,
 where '$+$' denotes concatenation. The set $C$ is initialized to
@@
 
-In order to incorporate \emph{incertae sedis }taxa into a supertree
-analyses, we first tackle the issue of the semantics of \emph{incertae
-sedis} taxa. We discuss various different things the term can mean
-within the literature, and settle on a specific usage here. We discuss
-how to move from an operational ``you can move this guy here'' semantics
-to a splits-based semantics. This affects both what it means for a
-taxonomy edge to conflict with the synthesis tree, and what it means
-to place a taxon name on a particular node in the synthesis tree.
-Finally, after a supertree has been constructed, we consider how to
-interpret the supertree as (perhaps) placing an \emph{incertae sedis}
-taxon $A$ within its sister taxon $B$.
-
-\subsection{Notation}
-
-We extend this framework by adding a set $\mathcal{I}\subseteq\mathcal{N}$
-of labels that have the \emph{incertae sedis }property. The taxonomy
-consists of the labels $\mathcal{N}$, the taxonomy tree $\taxonomy$,
-and the \emph{incertae sedis} property $\mathcal{I}$ of taxon names.
-
-\subsection{Usage in the literature}
-
-The term \emph{incertae sedis} can be used in a number of different
-ways. For example,\emph{ }one author might describe \emph{incertae
-sedis} genera that are asserted to be monophyletic, whereas another
-author might describe \emph{incertae sedis} genera that may be interdigitated.
+We extend this framework by adding a set $\mathcal{U}\subseteq\mathcal{N}$
+of labels that have the \emph{incertae sedis }property. The extended
+taxonomy then consists of the taxonomy tree $\taxonomy$, the labels
+$\mathcal{N}$, and set $\mathcal{U}$ of \emph{incertae sedis }taxa.
 
@@
 
-\subsubsection{Splits}
-
-Each edge of a standard tree divides the tip taxa $\mathcal{L}$ into
-two groups: the include set $\mathcal{I}(e)$ which does not contain
-the root, and the exclude set $\mathcal{E}(e)$ which does contain
-the root. Such a split may be written
+In order to apply supertree methods to rooted trees with \emph{incertae
+sedis} taxa, we first propose a semantics of \emph{incertae sedis
+}in terms of rooted splits.
+
+We define rooted splits by noting that each edge of a tree divides
+the tip taxa $\mathcal{L}$ into two groups: the include set $\mathcal{I}(e)$
+which does not contain the root, and the exclude set $\mathcal{E}(e)$
+which does contain the root. Such a split may be written
 \[
@@
 
-For a node $n$ on the tipward side of an edge $e$, we may also write
-$\mathcal{I}(n)$ for $\mathcal{I}(e)$, and $\mathcal{E}(n)$ for
-$\mathcal{E}(e)$. We consider the exclude set of the root node to
-be empty, and the include set of the root node to contain all tip
-taxa:
-\begin{align*}
-\mathcal{E}(root) & =\{\}\\
-\mathcal{I}(root) & =\mathcal{L}.
-\end{align*}
-In this case the root node stands for an edge that connects the root
-to the root's parent.
+For a node $n$ on the tip-ward side of an edge $e$, we may also
+write $\mathcal{I}(n)$ for $\mathcal{I}(e)$, and $\mathcal{E}(n)$
+for $\mathcal{E}(e)$.
 
@@
 Additionally, the exclude $\mathcal{E}(n)$ should contain the children
-of all the ancestors of $n$, unless those children are incertae sedis.
+of all the ancestors of $n$, unless those children are \emph{incertae
+sedis}.
 
@@
 $A$ is placed within a sister group. Suppose $A$ is placed next
-to a sibling $B$. If $B$ is \emph{incertae sedis}, then it should
-be possible to place $B$ within $A$. As a result, $A$ should not
-exclude descendants of $B$ than can only be reached from $A$ by
-passing through a node marked \emph{incertae sedis. }
+to a sibling $B$. If $B$ is also \emph{incertae sedis}, then it
+should be possible to place $B$ within $A$. As a result, $A$ should
+not exclude descendants of $B$ than can only be reached from $A$
+by passing through a node marked \emph{incertae sedis. }
 
@@
 
-However, in the \emph{incertae sedis }framework we must raise the
-question of whether one name could apply to multiple nodes, or whether
-multiple names could apply to one node.
-
-\textbf{BDR:} \emph{It is well known that monotypic taxa are indistinguishable
-if you consider only splits on leaf labels, but are distinguishable
-if you consider splits on all node labels. It seems that some (all?)
-of the problems with assigning multiple names to the same nodes actually
-comes from the fact that moving IS taxa can leave the parent as a
-degree-2 node. The fact that propinquity handles monotypic taxa indicates
-that we actually implicitly consider all taxa to have labels. It is
-possible that a trivial extension to the sub-problem solved could
-thus handle monotypic taxa and issues with mapping 2 names to one
-node with incertae sedis taxa.}
-
-\subsubsection{Multiple nodes that fit one name}
-
-Suppose the taxonomy is (((A1,A2)A,B)AB,C,D{*}) and the solution tree
-is (((A1,A2)x,D{*})y,B)AB,C). In this case the name $A$ leads to
-the split S(A) = A1 A2 | B C root. This name can apply to both the
-node x and the node y.
-
-\textbf{Solution:} In this case, we find the most tipward node and
-attach the name to this node.
+However, in the \emph{incertae sedis} framework, it is possible for
+one name to apply to multiple nodes. For example, in Figure \ref{fig:One-name-can},
+the name $A$ can apply to multiple nodes. Here the name $A$ corresponds
+to the split $A1\,A2\,|\bullet B\,C$, leaving out $D$ since it is
+\emph{incertae sedis}. The two nodes $x$ and $y$ imply the splits
+$A1\,A2\,|\bullet B\,C\,D$ and $A1\,A2\,D|\bullet B\,C$ respectively,
+and both of these splits imply the split $A1\,A2\,|\bullet B\,C$,
+so the name $A$ can apply to both $x$ and \footnote{Define ``implies'', as in the propinquity paper: $A_{1}|\bullet A_{2}\implies B_{1}|\bullet B_{2}$
+means $B_{1}\subseteq A_{1}$ and $B_{2}\subseteq A_{2}$.}. This cannot happen without\emph{ incertae sedis} taxa except at
+monotypic nodes. When faced with a choice about where to place a name,
+we find the most tip-ward node where the name can apply and attach
+the name to this node. This corresponds to a conservative choice about
+whether \emph{incertae sedis} taxa should be added to an existing
+taxon.
+\begin{figure}
+\hfill{}\subfloat[Taxonomy tree]{\includegraphics[scale=0.5]{Figures/name1nodes2/tax-shaded}
+
+}\hfill{}\subfloat[Solution tree]{\includegraphics[scale=0.5]{Figures/name1nodes2/solution-shaded}
+
+}\hfill{}
+
+\caption{\label{fig:One-name-can}One name can be consistent with multiple
+nodes.}
+\end{figure}
+
+It is also possible for multiple names to apply to a single node.
+
+<--------------------- resume editting here ------------->
+
+\emph{}
 
@@
 
+\subsubsection{Usage in the literature}
+
+The term \emph{incertae sedis} can be used in a number of different
+ways. For example,\emph{ }one author might describe \emph{incertae
+sedis} genera that are asserted to be monophyletic, whereas another
+author might describe \emph{incertae sedis} genera that may be interdigitated.
+
 \bibliographystyle{upmplainnat}

### 2018-03-27 doc/incertae_sedis_latex2.tex
Update ms.

@@
 to the supertree nodes based on the taxonomy tree in the problem.
-Each taxon name $n$ corresponds to a split $S(n)=S(n)_{1}|S(n)_{2}$
+Each taxon name $n$ corresponds to a split $S(n)=S(n)_{1}|\bullet S(n)_{2}$
 on the corresponding branch of the taxonomy tree. Without \emph{incertae
-sedis}, such splits are always of the form $S(n)_{1}|\mathcal{L}-S(n)_{1}$,
+sedis}, such splits are always of the form $S(n)_{1}|\bullet\mathcal{L}-S(n)_{1}$,
 but with \emph{incertae sedis} taxa $S_{2}(n)$ may be smaller than
@@
 However, in the \emph{incertae sedis} framework, it is possible for
-one name to apply to multiple nodes. For example, in Figure \ref{fig:One-name-can},
-the name $A$ can apply to multiple nodes. Here the name $A$ corresponds
-to the split $A1\,A2\,|\bullet B\,C$, leaving out $D$ since it is
-\emph{incertae sedis}. The two nodes $x$ and $y$ imply the splits
-$A1\,A2\,|\bullet B\,C\,D$ and $A1\,A2\,D|\bullet B\,C$ respectively,
-and both of these splits imply the split $A1\,A2\,|\bullet B\,C$,
-so the name $A$ can apply to both $x$ and \footnote{Define ``implies'', as in the propinquity paper: $A_{1}|\bullet A_{2}\implies B_{1}|\bullet B_{2}$
+one name to apply to multiple nodes. For example, in Figure \ref{fig:One-name-can}b,
+the name $A$ can apply to nodes $x$ and $y$. Here the name $A$
+corresponds to the split $A1\,A2\,|\bullet B\,C$, leaving out $D$
+since it is \emph{incertae sedis}. The two nodes $x$ and $y$ imply
+the splits $A1\,A2\,|\bullet B\,C\,D$ and $A1\,A2\,D|\bullet B\,C$
+respectively, and both of these splits imply the split $A1\,A2\,|\bullet B\,C$,
+so the name $A$ can apply to both $x$ and $y$\footnote{Define ``implies'', as in the propinquity paper: $A_{1}|\bullet A_{2}\implies B_{1}|\bullet B_{2}$
 means $B_{1}\subseteq A_{1}$ and $B_{2}\subseteq A_{2}$.}. This cannot happen without\emph{ incertae sedis} taxa except at
 monotypic nodes. When faced with a choice about where to place a name,
-we find the most tip-ward node where the name can apply and attach
-the name to this node. This corresponds to a conservative choice about
-whether \emph{incertae sedis} taxa should be added to an existing
-taxon.
+our solution is to find the most tip-ward node where the name can
+apply and attach the name to this node. This corresponds to a conservative
+choice about whether \emph{incertae sedis} taxa should be added to
+an existing taxon.
 \begin{figure}
@@
 It is also possible for multiple names to apply to a single node.
+For example, in Figure \ref{fig:Multiple-names-can}a, the taxon $A$
+corresponds to the split $B1\,B2\,C\,|\bullet\,Y$, and is child taxon
+$B$ corresponds to the split $B1\,B2\,|\bullet\,Y$. The edge leading
+to $A$ is consistent with the split for $B$, but the name $B$ is
+applied to the node with the smallest include group. However, in the
+solution tree (Fig. \ref{fig:Multiple-names-can}b), there is is only
+node for both names $A$ and $B$ to apply to. In this case, we solve
+this problem by introducing a monotypic parent, and applying $A$
+to the newly created parent node (Fig. \ref{fig:Multiple-names-can}c).
+\begin{figure}
+\hfill{}\subfloat[Taxonomy tree]{\includegraphics[scale=0.5]{Figures/names2nodes1/tax-shaded}
+
+}\hfill{}\subfloat[Solution tree]{\includegraphics[scale=0.5]{Figures/names2nodes1/synth-shaded}
+
+}\hfill{}\subfloat[Solution tree]{\includegraphics[scale=0.5]{Figures/names2nodes1/solution2-shaded}
+
+}\hfill{}
+
+\caption{\label{fig:Multiple-names-can}Multiple names can apply to a single
+node.}
+
+\end{figure}
+
+Not all cases where multiple names apply to a single node can be resolved
+in this way.
 

### 2018-04-02 doc/incertae_sedis_latex2.tex
Update.

@@
 sedis} taxa, we first propose a semantics of \emph{incertae sedis
-}in terms of rooted splits.
-
-We define rooted splits by noting that each edge of a tree divides
-the tip taxa $\mathcal{L}$ into two groups: the include set $\mathcal{I}(e)$
-which does not contain the root, and the exclude set $\mathcal{E}(e)$
-which does contain the root. Such a split may be written
+}in terms of rooted splits. We define rooted splits by noting that
+each edge of a tree divides the tip taxa $\mathcal{L}$ into two groups:
+the include set $\mathcal{I}(e)$ which does not contain the root,
+and the exclude set $\mathcal{E}(e)$ which does contain the root.
+Such a split may be written
 \[
@@
 
-Marking a taxon as \emph{incertae sedis} changes the meaning of its
-sibling taxa. We seek to represent this by constructing modified splits
-for each branch of the taxonomy tree. The include sets of these splits
-remain unchanged, but we construct reduced exclude sets to indicate
-cases where \emph{incertae sedis} taxa are allowed to intrude into
-a taxon.
-
 An \emph{incertae sedis} taxon can be moved into any of the descendants
-of its siblings. Therefore, the exclude set $\mathcal{E}(n)$ for
-node $n$ should not include any of its siblings marked \emph{incertae
-}sedis\emph{,} but should exclude the descendants of its other siblings\emph{.}
-Additionally, the exclude $\mathcal{E}(n)$ should contain the children
-of all the ancestors of $n$, unless those children are \emph{incertae
-sedis}.
-
-We may also consider what happens if an \emph{incertae sedis} taxon
-$A$ is placed within a sister group. Suppose $A$ is placed next
-to a sibling $B$. If $B$ is also \emph{incertae sedis}, then it
-should be possible to place $B$ within $A$. As a result, $A$ should
-not exclude descendants of $B$ than can only be reached from $A$
-by passing through a node marked \emph{incertae sedis. }
-
-We can compute exclude sets recursively by writing the exclude set
-of a node $n$ in terms of the exclude set of its parent:
-\begin{align}
-\mathcal{E}(n) & =\mathcal{E}(parent(n))\cup\left[\mathcal{I^{\prime}}(m)\big|m\in siblings(n)\right],\label{eq:exclude-set-formula-1}
-\end{align}
-where $\mathcal{I}^{\prime}(n)$ indicates the descendants of $n$
-that are reachable from $n$ without passing through an \emph{incertae
-sedis} node:
+of its siblings. Therefore marking a taxon as \emph{incertae sedis}
+changes the meaning of its sibling taxa. We seek to represent this
+by constructing modified splits for each branch of the taxonomy tree.
+The include sets of these splits remain unchanged, but we construct
+reduced exclude sets in order to allow \emph{incertae sedis} taxa
+to intrude into their sibling taxa.
+
+The reduced exclude set $\mathcal{E}(n)$ for node $n$ should therefore
+not contain the descendants of its siblings marked \emph{incertae
+}sedis\emph{,} but should contain the descendants of its other siblings\emph{.
+}We additionally stipulate that it is not possible for \emph{incertae
+sedis} taxa to intrude into their \emph{incertae sedis} siblings.
+(This is not the only way to construct a split-based semantics for
+\emph{incertae sedis}, but it avoids certain complications, as noted
+in the discussion.) We thus define $\mathcal{I}(m,n)$ to indicate
+descendants of $m$ that are excluded from $n$. Thus:
 \begin{align*}
-\mathcal{I}^{\prime}(n) & =\begin{cases}
-\emptyset & \text{if }n\text{ is \emph{incertae sedis}}\\
-\{n\} & \text{if }n\text{ is a leaf node}\\
-\left[\mathcal{I^{\prime}}(m)\big|m\in children(n)\right] & \text{if }n\text{ is an internal node}
+\mathcal{I}(m,n) & =\begin{cases}
+\emptyset & \text{if }n\text{ is \emph{incertae sedis}\,and }m\text{ is not}\\
+\mathcal{I}(m) & \text{otherwise}
 \end{cases}
 \end{align*}
-This recursion terminates if the exclude set for the root node is
-set to $\emptyset$ as a boundary condition.
+
+Additionally, the exclude set $\mathcal{E}(n)$ should contain the
+children of all the ancestors of $n$, unless those children are \emph{incertae
+sedis}. To avoid traversing all ancestors separately for each node
+$n$, we note that the excluded children of any ancestors will be
+in the exclude set of the parent of $n$ already. Therefore, we can
+write the exclude set of a node $n$ in terms of the exclude set of
+its parent:
+
+\begin{align}
+\mathcal{E}(n) & =\mathcal{E}(parent(n))\cup\left[\mathcal{I}(m,n)\big|m\in siblings(n)\right].\label{eq:exclude-set-formula-1}
+\end{align}
+This formula allows us to compute exclude sets via a pre-order recursion
+on the taxonomy tree. This recursion terminates if the exclude set
+for the root node is set to $\emptyset$ as a boundary condition.
+
+\textbf{Problem:} how about the goal that we should be able to get
+the same outcome by moving a taxon up one level and marking it as
+incertae sedis? This might require that moving A next to B, and then
+moving B into A is allowed.
+
+\textbf{Problem:} should we be able to move incertae sedis taxa down
+into incertae sedis taxa? If not, then we must change the recursion.
 
@@
 
-After solving a supertree (sub) problem, we need to assign taxon names
+After constructing a supertree, we still need to assign taxon names
 to the supertree nodes based on the taxonomy tree in the problem.
@@
 For example, in Figure \ref{fig:Multiple-names-can}a, the taxon $A$
-corresponds to the split $B1\,B2\,C\,|\bullet\,Y$, and is child taxon
-$B$ corresponds to the split $B1\,B2\,|\bullet\,Y$. The edge leading
-to $A$ is consistent with the split for $B$, but the name $B$ is
-applied to the node with the smallest include group. However, in the
-solution tree (Fig. \ref{fig:Multiple-names-can}b), there is is only
-node for both names $A$ and $B$ to apply to. In this case, we solve
-this problem by introducing a monotypic parent, and applying $A$
-to the newly created parent node (Fig. \ref{fig:Multiple-names-can}c).
+corresponds to the split $B1\,B2\,C\,|\bullet\,Y$, and its child
+taxon $B$ corresponds to the split $B1\,B2\,|\bullet\,Y$. The edge
+leading to $A$ is consistent with the split for $B$, but the name
+$B$ is applied to the node with the smallest include group. However,
+in the solution tree (Fig. \ref{fig:Multiple-names-can}b), there
+is is only node for both names $A$ and $B$ to apply to. In this
+case, we solve this problem by introducing a monotypic parent, and
+applying $A$ to the newly created parent node (Fig. \ref{fig:Multiple-names-can}c).
 \begin{figure}
@@
 
-Not all cases where multiple names apply to a single node can be resolved
-in this way.
-
-<--------------------- resume editting here ------------->
-
 \emph{}
-
-\subsubsection{Multiple names fit a single node}
-
-\paragraph{Example 1}
-
-Suppose the taxonomy is (((B1,B2)B,C{*})A,Y) and the input tree is
-(((B1,C),B2)x,Y) then the names A and B both to the node x. In this
-case the names A and B are ordered.
-
-\textbf{Case 1:} If a taxon contains 2 non-IS taxa, then it cannot
-be identical with any of its children in the synthesis tree.
-
-\textbf{Case 2:} If a taxon contains 1 non-IS taxon and $\ge1$ IS
-taxa, then the taxon could be identical with is non-IS child in the
-synthesis tree, if the IS taxa are placed within the child.
-
-\textbf{Case 3:} If a node contains 0 non-IS taxa and 1 IS taxon,
-then the IS taxa behaves no differently than a non-IS taxa, since
-is has no siblings it could be placed into.
-
-\textbf{Case 4:} If a node contains 0 non-IS taxa and $\ge2$ IS taxa,
-then then taxon \emph{could} be identical with a non-IS child in the
-synthesis tree, if all but one IS children are placed with in one
-of the IS children.
-
-\textbf{Solution:} When we assign multiple names to the same node,
-then we expand the node with multiple names to have monotypic parents,
-and assign the series of names to the monotypic parents. Another way
-of saying this is that when a node has $\le1$ non-IS taxon and $\ge2$
-taxa then the node could become monotypic by placement of the IS taxa.
-
-\textbf{Example 2}
-
-If ((A1,A2)A,(B1,B2)B{*},(C1,C2)C{*}); is the taxonomy with asterisks
-denoting incertae sedis taxa, then the solution ((A1,A2)A,((B1,C1)mrcaB1C1,(B2,C2)mrcaB2C2)x);
-has a node x that could be called B{*} or C{*}.
-
-\textbf{Case:} If a taxa B{*} and C{*} are IS, then their tips can
-be intermingled in a new clade x. Both names would then apply.
-
-\textbf{Solution:} ??
-
-In summary, a taxon with split $A_{1}|\bullet B_{1}$ attaches to
-the most tipward node $n$ where $A_{1}\subseteq S_{1}(n)$ and $B_{1}\subseteq S_{2}(n)$.
-If multiple names end up on the same node, we try to resolve the problem
-by creating a monotypic node for a name that is more rootward than
-all other names at that node. If this fails to resolve the problem,
-we arbitrarily choose one of the names.
 
@@
 
+If incertae sedis taxa are cannot be placed within other incertae
+sedis taxa, then such an approach does not work if $y$ is already
+incertae sedis. However, we take the approach that \emph{incertae
+sedis} notations are not a general solution to expression of uncertain
+placement.
+
 \subsubsection{Comparison to operational definition}
@@
 
+\subsubsection{Intrusion into incertae sedis taxa}
+
+In theory it could be possible for an \emph{incertae sedis} taxon
+$A$ to be placed within an \emph{incertae sedis} sibling $B$. We
+initially implemented this semantics. It requires updating the recursion
+to define $I(m,n)$ as the descendants of $m$ that can be access
+from $n$ without traversing an \emph{incertae sedis} node.
+
+However, if we allow this behavior by reducing the exclude sets for
+$A$ and $B$, then this has the unfortunate consequence that the
+members of $A$ and $B$ may be inter-digitated. In many cases, \emph{incertae
+sedis} genera have siblings that are families, and the expected semantics
+is that genera may be placed within their sibling families, or may
+remain outside existing families. In such cases, inter-digitating
+the genera would not be expected behavior. Thus, we choose the semantics
+that \emph{incertae sedis} taxa exclude their \emph{incertae sedis}
+siblings. This is not because the other semantics is impossible. However,
+we note that use of the interdigitating semantics allows a situation
+where the names for $A$ and $B$ both apply to the same node.
+
+This semantics also complicates placement, since it is possible to
+place $A$ as a sibling of $B$, and then place $B$ within $A$.
+This violates the current invariant that taxa are only ever placed
+within more tipward taxa. This makes the description of placement
+more complicated, since we can no longer simply record all cases where
+a parent node is not ancestor on the taxonomy.
+
 \subsubsection{Usage in the literature}

### 2018-04-03 doc/incertae_sedis_latex2.tex
Update.

@@
 
-\title{\texttt{Supertree construction with}\texttt{\emph{ Incertae sedis}}\texttt{
-taxa}}
+\title{\texttt{Taxonomic supertree construction with}\texttt{\emph{ Incertae
+sedis}}\texttt{ taxa}}
 \maketitle
 
-\section{New Introduction}
+\section{Introduction}
 
@@
 methods are usually applied to leaf-labeled phylogenies that have
-been estimated from DNA sequence data, we additionally seek to include
-taxonomic information. We thus require that one of the input trees
-is the taxonomy tree. The taxonomy has a number of features not found
-in the input trees, such as having labels for internal nodes, containing
-monotypic nodes, and containing \emph{incertae sedis} nodes. We refer
-to labels of taxonomy nodes as ``taxa'', and require that labels
-and taxonomy nodes have a one-to-one correspondence. We additionally
-require that the taxonomy be comprehensive and contain all labels.
-
-One of the problems that arises when attempting to include taxonomic
-information is the problem of \emph{incertae sedis }taxa. \emph{Incertae
-sedis} taxa are taxa that have unknown taxonomic placement. The common
-interpretation of such taxa is that they may not be place further
-root-ward, but may be placed within their sibling taxa. This frequently
-occurs when a specimen is identified to a taxonomic level such as
-the family level but not further. It also occurs when it is unclear
-whether a genus forms its own family, or is part of an existing family.
-(Mark rewrite?)
-
-\section{Introduction}
-
-The phrase \textquotedblleft \emph{incertae sedis}\textquotedblright ,
-which is a Latin phrase meaning ``uncertain seat'', indicates that
-a taxon is of uncertain placement. Such groups arise in practice when
-they are classified within a parent group, such as Fungi, but the
-precise location within the parent group is uncertain. Such a taxon
-would then be attached as a child of the parent group, but labeled
-as \emph{incertae sedis}.
+been estimated from DNA sequence data, we seek to include taxonomic
+information also. Taxonomies can provide a comprehensive coverage
+of known taxa that inferred phylogenies cannot. Additionally, taxonomies
+can provide taxonomic names, which are needed to align input phylogenies
+to each other, and to allow references to existing groups.
+
+We thus require that one of the input trees is a taxonomy tree. The
+taxonomy has a number of features not found in the input phylogenies.
+While the input trees need not cover the entire taxon set, we require
+the taxonomy to be comprehensive. The taxonomy also has labels for
+internal nodes, whereas input phylogenies only have labels for leaf
+nodes. We refer to labels of taxonomy nodes as ``taxon names'',
+and require that taxon names and taxonomy nodes have a one-to-one
+correspondence. The taxonomy may also contain out-degree 1 nodes.
+These nodes correspond to monotypic taxa which contain a single child
+of lower taxonomic rank.
+
+In practice, taxonomies may also contain nodes with uncertain placement.
+These nodes are often labelled ``\emph{incertae sedis}'', which
+means ``uncertain seat'' in Latin. The common interpretation of
+such taxa is that they may not be moved further towards the root,
+but may be moved into their sibling taxa. For example, a genus with
+a sibling that is a family may be annotated as``\emph{incertae familia}'',
+indicating that it is unclear which family the genus should be placed
+in. \emph{Incertae sedis} taxa frequently occur when a specimen is
+identified up to a given taxonomic level, but no further. Extinct
+taxa are often \emph{incertae sedis}.
 
@@
 
-\paragraph{Background}
+\subsection{Background}
+
+\paragraph{Terminology}
 
@@
 
+In this manuscript we extend this framework by adding a set $\mathcal{U}\subseteq\mathcal{N}$
+of labels that have the \emph{incertae sedis }property. The extended
+taxonomy then consists of the taxonomy tree $\taxonomy$, the labels
+$\mathcal{N}$, and set $\mathcal{U}$ of \emph{incertae sedis }taxa.
+
 \paragraph{Exemplification}
@@
 
-We extend this framework by adding a set $\mathcal{U}\subseteq\mathcal{N}$
-of labels that have the \emph{incertae sedis }property. The extended
-taxonomy then consists of the taxonomy tree $\taxonomy$, the labels
-$\mathcal{N}$, and set $\mathcal{U}$ of \emph{incertae sedis }taxa.
-
-\subsection{A splits-based semantics for \emph{incertae sedis} taxa}
+ We seek a semantics for supertrees with \emph{incertae sedis }taxa
+that satisfy the following properties:
+\begin{enumerate}
+\item An incertae sedis node may intrude into its siblings and their descendants.
+\item Two incertae sedis siblings may not be interdigitated.
+\item Any placement $T\to T^{*}$ of a tree $T$ with incertae sedis taxa
+is also accessible from a tree $T^{\prime}$ that is created from
+$T$ by taking a taxon, attaching it to an ancestor, and marking it
+\emph{incertae sedis}.
+\item The semantics of an \emph{incertae sedis} node does not depend on
+assigned ranks, but only on the tree.
+\item Semantics is based on deriving a split for each branch of the tree.
+\end{enumerate}
+Here we focus on a split-based semantics instead of a semantics based
+on apply of a series of edit operations to the tree. The split-based
+semantics allows us to seek trees that are simultaneously consistent
+with a set of split-constraints. The use of a split-based semantics
+does impose some limitations. These come primarily because in the
+edit-based semantics, after we place an \emph{incertae sedis} taxon,
+it is no loner \emph{incertae sedis}, whereas with the split-based
+semantics taxa are either always \emph{incertae sedis} or never \emph{incertae
+sedis}.
+
+This has an effect, for example, when we consider two incertae sedis
+siblings $A$ and $B$. If we can place $A$ within $B$, and $B$
+within $A$, then a split-based semantics allows us to freely inter-digitate
+$A$ and $B$. If we do not want to allow $A$ and $B$ to be inter-digitated,
+then we cannot both $A$ within $B$ and $B$ within $A$. In addition,
+we can imagine that a family $A$ is incertae sedis within a kingdom.
+In an edit-based semantics, we could place $A$ as a sibling to a
+genus $B$, mark $A$ as non-\emph{incertae-sedis}, and then place
+$B$ within $A$. In a split-based semantics, we must disallow either
+$B$ within $A$ or $A$ within $B$ if we want to disallow interdigitation.
+
+As a result of this, we cannot satisfy all five of these criteria.
+Thus, we describe two alternative split-based semantics. The non-interdigitation
+semantics gets \# (1,2,4,5), and the interdigitation semantics gets
+(1,3,4,5). This makes the point that each semantics results from different
+choices about which criteria to satisfy.
+
+\subsection{Split-based semantics for \emph{incertae sedis} taxa}
 
@@
 
-\subsubsection{Reduced exclude sets allow \emph{incertae sedis} to be placed into
-sibling taxa}
+\subsubsection{A semantics that disallows inter-digitating \emph{incertae sedis}
+siblings}
 
 An \emph{incertae sedis} taxon can be moved into any of the descendants
-of its siblings. Therefore marking a taxon as \emph{incertae sedis}
-changes the meaning of its sibling taxa. We seek to represent this
-by constructing modified splits for each branch of the taxonomy tree.
-The include sets of these splits remain unchanged, but we construct
-reduced exclude sets in order to allow \emph{incertae sedis} taxa
-to intrude into their sibling taxa.
+of its siblings. We seek to represent this by constructing modified
+splits for each branch of the taxonomy tree. The include sets of these
+splits remain unchanged, but we construct reduced exclude sets in
+order to allow \emph{incertae sedis} taxa to intrude into their sibling
+taxa.
 
@@
 into incertae sedis taxa? If not, then we must change the recursion.
+
+\subsubsection{An alternative semantics that allows inter-digitating incertae sedis
+siblings}
+
+If we want to allow placing two incertae sedis siblings within each
+other, then incertae sedis siblings are never excluded.
+
+Furthermore, consider placing $A$ as a sibling of $B$, and then
+placing $B$ within $A$. In order for this to work, $A$ must not
+exclude the incertae sedis children of $B$.
+
+In order to accomplish both of these goals, we define the descendants
+$\mathcal{I}^{\prime}(n)$ that are accessible without passing through
+an incertae sedis node.
+
+\begin{align*}
+\mathcal{I^{\prime}}(n) & =\begin{cases}
+\emptyset & \text{if }n\in\mathcal{U}\\
+\{n\} & \text{if }n\text{ is a leaf}\\{}
+[\mathcal{I}^{\prime}(c)|c\in children(n)] & \text{otherwise.}
+\end{cases}
+\end{align*}
+We may then define the taxa excluded from $n$ as:
+
+\begin{align*}
+\mathcal{E}(n) & =\mathcal{E}(parent(n))\cup[\mathcal{I}^{\prime}(s)|s\in siblings(n)].
+\end{align*}
+
+This semantics satisfies criteria 1,3,4, and 5 above.
+
+\subsubsection{A third semantics}
+
+OK, the idea here is that we want to avoid inter-digitation. \emph{Incertae
+sedis} taxa thus exclude\emph{ incertae sedis} siblings. However,
+where the first semantics allows \emph{incertae sedis }taxa to intrude
+into\emph{ incertae sedis }descendants of siblings, but not allow\emph{
+incertae sedis} descendants of siblings to intrude into \emph{incertae
+sedis} taxa, here we do the reverse.
+
+OK, so the idea here is that an\emph{ incertae sedis} group excludes
+\begin{itemize}
+\item \emph{incertae sedis} siblings
+\item \emph{incertae sedis }descendants of ancestors
+\item but not\emph{ incertae sedis} descendants of non-incertae-sedis siblings
+\end{itemize}
+It is also possible to allow placing $A$ as a sibling of $B$ and
+then placing $B$ within $A$ if we disallow placing $A$ into incertae
+sedis descendants of its siblings. We may then allow placing those
+descendants within $A$.
+
+\begin{align*}
+\mathcal{I^{\prime}}(n) & =\begin{cases}
+\emptyset & \text{if }n\in\mathcal{U}\\
+\{n\} & \text{if }n\text{ is a leaf}\\{}
+[\mathcal{I}^{\prime}(c)|c\in children(n)] & \text{otherwise.}
+\end{cases}
+\end{align*}
+
+We use $\mathcal{I}(m,n)$ to indicate the descendants of $m$ that
+are excluded from $n$.
+\begin{align*}
+\mathcal{I}(m,n) & =\begin{cases}
+\emptyset & \text{if }n\text{ is \emph{incertae sedis}\,and }m\text{ is not}\\
+\mathcal{I}(m) & \text{if }n\text{ and }m\text{ are both }incertae\,sedis\\
+\mathcal{I}^{\prime}(m) & \text{otherwise.}
+\end{cases}
+\end{align*}
+
+We may then define the taxa excluded from $n$ as:
+
+\begin{align*}
+\mathcal{E}(n) & =\begin{cases}
+\left[\mathcal{L}-\mathcal{I}(parent(n))\right]\cup[\mathcal{I}(s,n)|s\in siblings(n)] & \text{if }n\in\mathcal{U}\\
+\mathcal{E}(parent(n))\cup[\mathcal{I}(s,n)|s\in siblings(n)] & \text{otherwise.}
+\end{cases}
+\end{align*}
+I think this works... but need to double-check.
+
+Also: can we genericize $\mathcal{E}(parent(n))$?
 
@@
 
-After a synthesis tree is constructed, we attach taxon names for to
-the synthesis tree for each non-conflicting taxon. These names attach
-at the MRCA of the include group. Then, for each taxon $A$ on the
-synth tree, we associate with the first taxon $B$ found in a rootward
-walk on the synthesis tree. If $A$ is not a descendant of $B$ on
-the taxonomy, then we say that $A$ is \emph{placed} within $B$ by
-the synthesis tree. While this may occur if $A$ is \emph{incertae
+After a synthesis tree is constructed, we attach taxon names to the
+synthesis tree for each non-conflicting taxon. These names attach
+at the MRCA of the include group. Then, for each taxon $A$ that is
+found on the synth tree, we associate with the first taxon $B$ found
+in a rootward walk on the synthesis tree. If $A$ is not a descendant
+of $B$ on the taxonomy, then we say that $A$ is \emph{placed} within
+$B$ by the synthesis tree. While this may occur if $A$ is \emph{incertae
 sedis}, more complex scenarios are possible. We note that if method
@@
 within $B_{m},B_{m-1},\ldots,B_{2},B_{1}$, and $m$ is the depth
-of the placement of $A$.
-
+of the placement of $A$.%
+\begin{comment}
 One question is whether this approach is able to handle nested incertae
@@
 \end{itemize}
+\end{comment}
 
@@
 
+With a splits-based semantics, we cannot allow both (i) placing $A$
+in $B$ and (ii) placing $B$ in $A$ and also forbid interdigitating
+$A$ and $B$.
+
 However, if we allow this behavior by reducing the exclude sets for

### 2018-04-17 doc/incertae_sedis_latex2.tex
Start reorganizing sections 3 + 4.

@@
 \usepackage[T1]{fontenc}
-\usepackage[latin9]{inputenc}
+\usepackage[latin9]{luainputenc}
 \usepackage{geometry}
@@
 sedis}}\texttt{ taxa}}
+
+\author{Benjamin D. Redelings and Mark T. Holder}
 \maketitle
@@
 
-We thus require that one of the input trees is a taxonomy tree. The
-taxonomy has a number of features not found in the input phylogenies.
-While the input trees need not cover the entire taxon set, we require
-the taxonomy to be comprehensive. The taxonomy also has labels for
-internal nodes, whereas input phylogenies only have labels for leaf
-nodes. We refer to labels of taxonomy nodes as ``taxon names'',
-and require that taxon names and taxonomy nodes have a one-to-one
-correspondence. The taxonomy may also contain out-degree 1 nodes.
-These nodes correspond to monotypic taxa which contain a single child
-of lower taxonomic rank.
+Here we describe a supertree method that takes a taxonomy as one of
+its inputs. The taxonomy has a number of features not found in the
+input phylogenies. While the input trees need not cover the entire
+taxon set, our method requires that the taxonomy to be comprehensive.
+The taxonomy also has labels for internal nodes, whereas input phylogenies
+only have labels for leaf nodes. We refer to labels of taxonomy nodes
+as ``taxon names'', and require that taxon names and taxonomy nodes
+have a one-to-one correspondence. The taxonomy may also contain out-degree
+1 nodes. These nodes correspond to monotypic taxa which contain a
+single child of lower taxonomic rank.
 
@@
 phylogenies with a comprehensive taxonomy that supplies taxon names
-\citep{redelings2017supertree}. We describe an extension to this
-supertree method so that we can resolve taxonomic uncertainty by using
-published phylogenies to place \emph{incertae sedis} taxa. We describe
-an updated semantics for conflict between phylogeny edges and taxonomy
-edges and an updated semantics for assigning taxonomy names to nodes
-in the synthetic tree.
+\citep{redelings2017supertree}. In practice, the method has been
+used with the Open Tree of Life taxonomy \citep{rees2017automated}.We
+describe an extension to this supertree method so that we can resolve
+taxonomic uncertainty by using published phylogenies to place \emph{incertae
+sedis} taxa. We describe an updated semantics for conflict between
+phylogeny edges and taxonomy edges and an updated semantics for assigning
+taxonomy names to nodes in the synthetic tree.
 
@@
 Thus, the ability to correctly assess compatibility with the taxonomy
-when the taxonomy contains \emph{incertae sedis taxa} allows us to
+when the taxonomy contains \emph{incertae sedis }taxa allows us to
 include thousands of new taxa that were previously filtered out to
@@
 
-After a synthesis tree is constructed, we attach taxon names to the
-synthesis tree for each non-conflicting taxon. These names attach
-at the MRCA of the include group. Then, for each taxon $A$ that is
-found on the synth tree, we associate with the first taxon $B$ found
-in a rootward walk on the synthesis tree. If $A$ is not a descendant
-of $B$ on the taxonomy, then we say that $A$ is \emph{placed} within
-$B$ by the synthesis tree. While this may occur if $A$ is \emph{incertae
-sedis}, more complex scenarios are possible. We note that if method
-does not handle \emph{incertae sedis} taxa, then this scenario could
-not occur because any taxa $B$ that $A$ is placed within would be
-considered incompatible with the synthesis tree and so their names
-would not be applied.
-
-In order to fully describe a placement of $A$ within $B$ we consider
-the sequence of taxa encountered on a walk from $A$ or $B$ up to
-their MRCA $M$ on the taxonomy tree. Let us denote the sequence of
-taxa encountered from $A$ as $A_{1},\ldots,A_{n},M$ where $A_{1}=A$.
-Let $B_{1},\ldots,B_{m},M$ be the sequence of taxa encountered from
-$B$ where $B_{1}=B$. Then $A_{n}$ must be \emph{incertae sedis},
-and $A_{2},\ldots,A_{n}$ must be incompatible with the synthesis
-tree (i.e. broken taxa). We then have that $A$ is placed successively
-within $B_{m},B_{m-1},\ldots,B_{2},B_{1}$, and $m$ is the depth
-of the placement of $A$.%
+After we have attached taxon names to the synthesis tree, we would
+like to interpret the position of these names in terms of placing
+\emph{incertae sedis }taxa in some spot, possibly a new spot. This
+would allow us to interpret the synthesis tree as saying that phylogenetic
+information has (for example) placed genus A within family B, or perhaps
+in a monotypic, unnamed family that contains only $A$. The simplest
+approach to placement involves noting whenever a taxon $B$ is a descendant
+of a taxon $A$ on the named synthesis tree but not the taxonomy tree.
+For example, in Figure \ref{fig:Handling-incertae-sedis}, taxon $C$
+is a descendant of taxon $A$ in the synthetic tree (Fig. \ref{fig:Handling-incertae-sedis}d),
+but not the taxonomy (Fig. \ref{fig:Handling-incertae-sedis}b). Thus,
+we could say that the synthetic tree places $C$ within $A$. Furthermore,
+the synthetic tree additionally places $C$ within $AB$.
+
+A slightly more general case is shown in Figure \ref{fig:Broken-incertae-sedis}.
+Here taxon $C1$ is an descendant of $A$ in the synthesis tree, but
+not the taxonomy. Thus $C1$ is placed within $A$. We note that the
+most recent common ancestor (MRCA) of $A$ and $C1$ in the taxonomy
+is the root node, and the path to the MRCA from $A$ is $A\to AB\to root$,
+while the path to the MRCA from $C1$ is $C1\to C\to root$. Here
+the additional node $AB$ on the path to the MRCA from $A$ indicates
+that $C1$ is placed within $AB$ as well as $A$. In contrast, the
+additional node $C$ on the path to the MRCA indicates that $C1$
+is a broken \emph{incertae sedis} taxon. Thus we may consider an algorithm
+that, for each taxon $B$ on the synthesis tree, finds the closest
+ancestral taxon $A$, and checks if $A$ is an ancestor of $B$ on
+the taxonomy. If not, we compute the paths $A\to A_{1}\to\ldots\to MRCA(A,B)$
+and $B\to B_{1}\to\ldots\to MRCA(A,B)$.
+
+However, this assumes that \emph{incertae sedis} taxa are always placed
+within their siblings, or descendants of their siblings. As such,
+it does not handle cases such as Figure \ref{fig:Placing-one-incertae},
+in which $I$ is first placed as sibling to $AI$, and then $AI$
+is placed within $I$. Note that $I$ is not a sibling of $AI$ or
+a descendant of a sibling. Furthermore, the MRCA of $I$ and $AI$
+is the root, and the path from $AI$ to the root includes $A$, but
+$A$ is not broken. We suggest that this conundrum may be resolved
+by first processing the placement of $I$ within $A$ by modifying
+the taxonomy to reflect this placement\footnote{Does this actually work? Some placements break taxa.}.
+When we consider the placement of $AI$ within $I$ on this modified
+taxonomy, the MRCA of $I$ and $AI$ is now $A$ instead of the root,
+and the correct interpretation results. We suggest that by walking
+the tree in a pre-order fashion and processing shallowing placements
+first, the previous approach of finding the MRCA and labelling intermediate
+nodes as either additional placements or broken incertae sedis (or
+incertae\_sedis\_inherited) taxa holds. Such an approach should also
+handle cases containing nested incertae sedis taxa, as in Figure \ref{fig:Placing-one-incertae}.%
 \begin{comment}
@@
 Synthesis with \emph{incertae sedis} taxa has the potential to resolve
-uncertain taxon placements using information from phylogenies. The
-propinquity pipeline has always been able to do this kind of resolution.
-However, without special consideration given to \emph{naming}, placing
-a taxon $A$ within a taxon $B$ results in conflict with taxon $B$
-in the taxonomy. In order to avoid a situation where input phylogenies
-conflict with a large number of taxa when \emph{incertae sedis} taxa
-are placed within them, we have filtered \emph{incertae sedis} taxa
+uncertain taxon placements using information from phylogenies. Since
+phylogenies are ranked higher than taxonomy, they may place members
+of incertae sedis clades as they wish even when we do not take into
+account the semantics of incertae sedis taxa. However, without the
+incertae sedis semantics described above, placing a taxon $A$ within
+a taxon $B$ results in conflict with taxon $B$ in the taxonomy.
+This has three main effects:
+\begin{enumerate}
+\item The edge leading to clade $B$ is lost.
+\item Any taxonomy-only taxa are placed at the MRCA of the phylogenetically
+informed taxa, instead of at node $B$.
+\item The name for taxon $B$ is lost.
+\item We do not record $A$ as being placed within $B$.
+\end{enumerate}
+In order to avoid a situation where input phylogenies conflict with
+a large number of taxa when \emph{incertae sedis} taxa are placed
+within them, we have previously filtered \emph{incertae sedis} taxa
 from the taxonomy when constructing all prior synthesis trees.
@@
 
-\subsection{Correctly handling placement}
-
-Thus, we seek a synthesis method that can correctly place taxa within
-containing taxa without breaking the containing taxa. We change the
-semantics of names to imply, not the exclusion of all non-included
-taxa, but only some non-included taxa, as described in section \ref{sec:Semantics-of-incertae}.
-As a result of this change in semantics, placing an IS taxon $A$
-within a sister taxon $B$ no longer results in conflict with $B$.
-This allows us to retain the split for $B$ within the synthesis tree,
-so that taxonomy-only children of $B$ are correctly grouped with
-their siblings that are referenced by the input trees. We may then
-retain the name for the no-longer-broken taxon. Finally, we are then
-able to stop filtering \emph{incertae sedis} taxa, so that they appear
-in the synthesis tree. Thus, the synthesis tree is able to represent
-substantially more species, without suffering the loss of taxa and
-the loss of structure.
+We therefore seek a synthesis method that can correctly introduce
+\emph{incertae sedis} taxa into containing taxa without breaking the
+containing taxa. We change the semantics of names to imply, not the
+exclusion of all non-included taxa, but only some non-included taxa,
+as described in section \ref{sec:Semantics-of-incertae}. As a result
+of this change in semantics, placing an IS taxon $A$ within a sister
+taxon $B$ no longer results in conflict with $B$. This allows us
+to retain the split for $B$ within the synthesis tree, so that taxonomy-only
+children of $B$ are correctly grouped with their siblings that are
+referenced by the input trees. We may then retain the name for the
+no-longer-broken taxon. Finally, we are then able to stop filtering
+\emph{incertae sedis} taxa, so that they appear in the synthesis tree.
+Thus, the synthesis tree is able to represent substantially more species,
+without suffering the loss of taxa and the loss of structure.
 
@@
 
-\subsubsection{Conflicting placement among input trees}
+When only tip nodes on the taxonomy are \emph{incertae sedis}, we
+have a very simple form of the \emph{incertae sedis} supertree problem.
+However, in practice entire clades may be \emph{incertae sedis}, and
+so more complex issue arise. In this section we consider a number
+of cases that must be handled when attempting to place \emph{incertae
+sedis} taxa.
+
+\emph{Note that the synthesis of all input trees before the taxonomy
+is unaffected by incertae sedis information.}
+
+\subsection{Conflicting placement among input trees}
+
+{[}\textbf{BDR:} \emph{maybe we should merge this with Figure. \ref{fig:Broken-incertae-sedis}.
+Although the point of figure \ref{fig:placement-example1} is to talk
+about sub-problem decomposition as well as what the result looks like.}{]}
 
@@
 figures!{]}
-
-\subsubsection{Example B}
-
-However, incertae sedis taxa are not always tip nodes, but may themselves
-contain other taxa. In such cases, it is possible for input trees
-to conflict with the incertae sedis taxon itself. For example, consider
-the following example
-\begin{itemize}
-\item $T_{1}=((w_{1},w_{2},z_{1}),y_{1})$
-\item $T_{2}=((y_{1},y_{2},z_{2}),w_{1})$
-\item Taxonomy tree $((w_{1},w_{2})w,(y_{1},y_{2})y,(z_{1},z_{2},z_{3})z)$
-with $z$ marked \emph{incertae sedis}.
-\begin{itemize}
-\item splits $w_{1}w_{2}|\bullet y_{1}y_{2}$, $y_{1}y_{2}|\bullet w_{1}w_{2}$
-and $z_{1}z_{2}|\bullet w_{1}w_{2}y_{1}y_{2}$
-\end{itemize}
-\end{itemize}
-In this case, tree $T_{1}$ places $z_{1}$ within $w$, while $T_{2}$
-places $z_{2}$ within $y$. Since different members of $z$ are placed,
-neither placement for $z$ is rejected. Instead the taxon $z$ is
-broken, the name $z$ disappears, and $z_{3}$ floats to the top level.
-
-Furthermore, since taxon $z$ is a broken \emph{incertae sedis }taxon,
-all of its children are effectively incertae sedis independently,
-with the difference that they cannot be placed within each other.
-Therefore, the names $w$ and $y$ are not lost, since the taxa placed
-within them are \emph{incertae sedis} names. Note that $z_{3}$ maybe
-therefore be placed in a third location, since it is also \emph{incertae
-sedis}.
-
-The situation here would be different if the monophyly of $z$ was
-supported by an input phylogeny.
-
-\emph{Note that the synthesis of all input trees before the taxonomy
-is unaffected by incertae sedis information.}
-
-\subsubsection{Example C - Nested \emph{incertae sedis} taxa}
-
-In addition to incertae sedis taxa containing other taxa, it is also
-possible for incertae sedis taxa to contain nested \emph{incertae
-sedis} taxa. \textbf{What issues might this raise?}
-
-\section{Cases}
-
-When only tip nodes on the taxonomy are \emph{incertae sedis}, we
-have a very simple form of the \emph{incertae sedis} supertree problem.
-However, in practice entire clades may be \emph{incertae sedis}, and
-so more complex issue arise. In this section we consider a number
-of cases that must be handled when attempting to place \emph{incertae
-sedis} taxa.
 
@@
 \begin{figure}
-
-\hfill{}\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-clade/tree2-shaded}
-
-}\hfill{}\subfloat[Taxonomy]{\includegraphics[scale=0.5]{Figures/is-clade/tax-shaded}
+\hfill{}\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-clade/tree1-shaded}
+
+}\hfill{}\subfloat[Input tree \#2]{\includegraphics[scale=0.5]{Figures/is-clade/tree3-shaded}
 
@@
 
-\hfill{}\subfloat[Synthesis Tree \#1]{\includegraphics[scale=0.5]{Figures/is-clade/synth2-shaded}
+\hfill{}\subfloat[Taxonomy]{\includegraphics[scale=0.5]{Figures/is-clade/tax-shaded}
+
+}\hfill{}\subfloat[Synthesis Tree \#1]{\includegraphics[scale=0.5]{Figures/is-clade/synth2-shaded}
 
@@
 
-\caption{Broken incertae sedis clade\label{fig:Broken-incertae-sedis}. Taxonomy
-tree (b) contains \emph{incertae sedis }taxon $C$ that is broken
-by input tree (a). Supertree construction results in a synthesis tree
-(c) that places $C1$ and $C2$ separately within $A$.}
+\caption{\label{fig:Broken-incertae-sedis} \emph{Incertae sedis} clade $C$
+broken by conflicting placement. Input trees (a) and (b) conflict
+in the placement of taxa in $C$. The synthesis supertree (d) places
+places $C1$ and $C2$ separately within $A$, while unplaced taxa
+float upwards. (\textbf{Note:} This is what the sub-problem solver
+does. The entire pipeline would attach $C3$ at $A$ actually, since
+it is the MRCA of $A1$ and $A2$. Should we change this?) }
 
 \end{figure}
+
+When the input phylogenies conflict on where members of an incertae
+sedis clade should be placed, neither placement is rejected. Instead,
+the taxon $C$ is broken, the name $C$ disappears, and $C3$ floats
+to the top level. Algorithmic, alternatively placements of children
+of an incertae sedis clade lead to contesting of all edges along with
+taxa might be placed, creating a large, merged sub-problem, as in
+Figure \ref{fig:placement-example1}.
+
+In Figure \ref{fig:Broken-incertae-sedis}, tree $T_{1}$ places $C1$
+next to $A1$, while $T_{2}$ places $C2$ next to $A2$. Since different
+members of $C$ are placed, neither placement for $C$ is rejected.
+Instead the taxon $C$ is broken, the name $C$ disappears, and $C3$
+floats to the top level. Furthermore, since taxon $C$ is a broken
+\emph{incertae sedis }taxon, all of its children are effectively incertae
+sedis independently, with the exception that they cannot be placed
+within each other. Therefore, the names $A$ and $AB$ are not lost,
+since the taxa placed within them are \emph{incertae sedis} names.
+
+The situation here would be different if the monophyly of $C$ was
+supported by an high-ranked input phylogeny. In that case, instead
+of breaking $C$, we would choose the placement in the highest-ranked
+tree.
 
@@
 
-\caption{Placing one incertae sedis group within another. Here I is placed
-within A, and then AI is placed within A.}
+\caption{\label{fig:Placing-one-incertae}Placing one incertae sedis group
+within another. Here I is placed within A, and then AI is placed within
+A.}
 
@@
 been placed within them. This would not require any updating to the
-conflict-analysis code in the annotation step.
+conflict-analysis code in the annotation step.%
+\begin{comment}
 
@@
 and probably novel. It would also probably be slower...
+\end{comment}
 

### 2018-04-19 doc/incertae_sedis.tex
lost the notes

--- 2018-04-17 doc/incertae_sedis_latex2.tex
@@
 \usepackage[T1]{fontenc}
-\usepackage[latin9]{luainputenc}
+\usepackage[latin9]{inputenc}
 \usepackage{geometry}
@@
 Supertree methods combine a collection of input trees with different
-taxon sets into a single tree on the combined taxon set. While such
-methods are usually applied to leaf-labeled phylogenies that have
-been estimated from DNA sequence data, we seek to include taxonomic
-information also. Taxonomies can provide a comprehensive coverage
-of known taxa that inferred phylogenies cannot. Additionally, taxonomies
-can provide taxonomic names, which are needed to align input phylogenies
-to each other, and to allow references to existing groups.
-
-Here we describe a supertree method that takes a taxonomy as one of
-its inputs. The taxonomy has a number of features not found in the
-input phylogenies. While the input trees need not cover the entire
-taxon set, our method requires that the taxonomy to be comprehensive.
-The taxonomy also has labels for internal nodes, whereas input phylogenies
-only have labels for leaf nodes. We refer to labels of taxonomy nodes
-as ``taxon names'', and require that taxon names and taxonomy nodes
-have a one-to-one correspondence. The taxonomy may also contain out-degree
-1 nodes. These nodes correspond to monotypic taxa which contain a
-single child of lower taxonomic rank.
-
-In practice, taxonomies may also contain nodes with uncertain placement.
-These nodes are often labelled ``\emph{incertae sedis}'', which
-means ``uncertain seat'' in Latin. The common interpretation of
-such taxa is that they may not be moved further towards the root,
-but may be moved into their sibling taxa. For example, a genus with
-a sibling that is a family may be annotated as``\emph{incertae familia}'',
-indicating that it is unclear which family the genus should be placed
-in. \emph{Incertae sedis} taxa frequently occur when a specimen is
-identified up to a given taxonomic level, but no further. Extinct
-taxa are often \emph{incertae sedis}.
-
-The Open Tree of Life project seeks to build a comprehensive supertree
+taxon sets into a single tree on the combined taxon set. The methods
+are usually applied to leaf-labeled phylogenetic estimates. Here we
+describe a supertree approach that overcomes issues associated with
+using a taxonomy as one of the inputs to a supertree method.
+
+The Open Tree of Life project \citep{HinchliffEtAl2015} seeks to
+build a comprehensive supertree \citep[see][]{redelings2017supertree}
 covering all of life. The approach is to combine information in published
-phylogenies with a comprehensive taxonomy that supplies taxon names
-\citep{redelings2017supertree}. In practice, the method has been
-used with the Open Tree of Life taxonomy \citep{rees2017automated}.We
-describe an extension to this supertree method so that we can resolve
-taxonomic uncertainty by using published phylogenies to place \emph{incertae
-sedis} taxa. We describe an updated semantics for conflict between
-phylogeny edges and taxonomy edges and an updated semantics for assigning
-taxonomy names to nodes in the synthetic tree.
-
-Our approach to \emph{incertae sedis} taxa enables us to use published
-phylogenies to update the taxonomy by extending taxonomic groups to
-include \emph{incertae sedis} taxa placed within them. The ability
-to update the taxonomy is important, since the placement of \emph{incertae
-sedis} taxa creates artificial conflicts with a non-updated taxonomy.
-Thus, the ability to correctly assess compatibility with the taxonomy
-when the taxonomy contains \emph{incertae sedis }taxa allows us to
-include thousands of new taxa that were previously filtered out to
-avoid conflict. It also enables us to include extinct taxa, since
-many of these taxa are \emph{incertae sedis}. This makes substantial
-progress towards our goal of completely representation of taxa.
+phylogenies with a comprehensive taxonomy that supplies taxon names.
+The supertree method has been used with the Open Tree of Taxonomy
+\citep[OTT hereafter, see][]{rees2017automated}, but in principle
+is general. The taxonomy is a crucial input for several reasons: it
+covers a wider range of species, its list of names allows for alignment
+of different phylogenetic estimates to each other, and it provides
+names for clades. Only a small proportion of known species have been
+included in a phylogenetic analysis, thus a taxonomy is crucial for
+achieving comprehensive coverage of known taxa. Phylogenetic estimates
+collected from the published literature often use different names
+for the same species. Lists of synonyms in OTT (and other taxonomies)
+allow data curators to ``align'' input phylogenies to the taxonomy
+\citep[see][ for discussion of the curation process in the Open Tree of Life project]{McTavishEtAt2015}.
+This alignment is important for recognizing when two different estimates
+refer to the same taxon. Biologists are familiar with names for ``higher''
+taxa (taxa above the species rank). Thus, naming clades in the supertree
+makes the tree easier to navigate and use.
+
+While some taxonomies may indicate hybrid taxa, the supertree approach
+described here does not consider any hybridization. Thus the taxonomic
+hierarchy can be converted to a tree. While not all taxonomists believe
+that named taxa should correspond to clades \citep[see, for example,][]{HorandlS2010},
+the priniciples of phylogenetic classification are used by such a
+large majority of practicing taxonomists, that we have chosen to treat
+the taxonomic tree that mirrors the hierarchy of the taxonomy as an
+estimate of phylogenetic relationships. We refer to labels of taxonomy
+nodes as ``taxon names'', and assume that taxon names and taxonomy
+nodes have a one-to-one correspondence. Note that the taxonomy may
+also contain out-degree 1 nodes. These nodes correspond to monotypic
+taxa which contain a single child of lower taxonomic rank.
+
+The supertree method described by \citet{redelings2017supertree}
+is able to use a set of phylogentic estimates and taxonomy to produce
+a supertree with clades names according to the taxonomy. However,
+the method was not able to accommodate the fact that taxonomies frequently
+contain nodes with uncertain placement. These nodes are often labelled
+``\emph{incertae sedis}'', which means ``uncertain seat'' in Latin.
+The common interpretation of such taxa is that they may not be moved
+further towards the root, but may be moved into their sibling taxa.
+For example, a genus with a sibling that is a family may be annotated
+as ``\emph{incertae familia}'', indicating that it is unclear which
+family the genus should be placed in. \emph{Incertae sedis} taxa frequently
+occur when a specimen is identified down to a given taxonomic level,
+but no further. Extinct taxa are often \emph{incertae sedis}. OTT
+is constructed from several source taxonomies, and these sources include
+various ways of indicating that the position of a taxon within the
+hierarchy is uncertain. OTT uses a series of ``flags'' to annotate
+these taxa. The supertree of by \citet{redelings2017supertree} simply
+pruned these taxa from the taxonomy and input trees. Thus, the final
+supertree lacked taxa which were \emph{incertae sedis}.
+
+Here describe a supertree method so that we can resolve taxonomic
+uncertainty by using published phylogenies to place \emph{incertae
+sedis} taxa. Developing this method required deciding on a set operational
+semantics to be used when interpreting the \emph{incertae sedis} annotation.
+These semantics affect the interpretation of the phylogenetic statements
+being made by the input taxonomy and the rules for applying taxonomic
+names to the final supertree. In addition to discussing how to interpret
+in the \emph{incertae sedis} annotation, we describe the algorithmic
+changes to the supertree pipeline that were required to adequately
+represent the taxonomic uncertainty. The approach described here allows
+us to include thousands of new taxa in our supertree analysis that
+were previously filtered out. It will also enable the Open Tree of
+Life project to include extinct taxa, since many of these taxa are
+\emph{incertae sedis}. This makes substantial progress towards our
+goal of comprehensive inclusion of known taxa into the Open Tree of
+Life summary tree.
 
@@
 
-\paragraph{Terminology}
-
-We briefly describe the supertree algorithm proposed in \citet{redelings2017supertree}.
-This supertree algorithm takes as input a ranked list $\phyloinputs=\left\{ T_{1},\ldots,T_{n}\right\} $
-of rooted input phylogenies and a single rooted taxonomy tree $\taxonomy$.
-Each node in the taxonomy tree is associated with a taxon name in
-the set of taxon names $\mathcal{N}$, and each taxon name is corresponds
-to only one node of the taxonomy tree. We use \emph{$\mathcal{L}$
-}to indicate taxon names corresponding to leaf nodes of the taxonomy
-tree. Correspondence between the input trees and the taxonomy is established
-by annotating leaf nodes of each input tree with taxon names. For
-purposes of conflict resolution, the taxonomy tree is ranked below
-all of the input trees.
-
-In this manuscript we extend this framework by adding a set $\mathcal{U}\subseteq\mathcal{N}$
-of labels that have the \emph{incertae sedis }property. The extended
-taxonomy then consists of the taxonomy tree $\taxonomy$, the labels
-$\mathcal{N}$, and set $\mathcal{U}$ of \emph{incertae sedis }taxa.
-
-\paragraph{Exemplification}
-
-While the taxonomy is comprehensive and contains all taxa, input phylogenies
-only have labels on their leaf nodes, and generally include a small
-fraction of the total number of taxa. Furthermore, taxa in input trees
-may correspond to higher taxa on the taxonomy tree, and multiple tips
-may be labeled with the same taxon name. \citet{redelings2017supertree}
-described a method called 'exemplification' that reduces this problem
-to a simpler form where all taxon names of input trees (i) correspond
-to leaf taxa and (ii) occur only once in an input tree. For the purposes
-of this paper, we therefore assume that this is the case.
-
-\paragraph{Problem decomposition}
-
-Given the exemplified input trees and taxonomy, the supertree algorithm
-first divides the supertree problem into independent pieces, solves
-the problem on these pieces, and then glues the pieces back together.
-This is made possible by the assumption that any taxonomy edges that
-are not contested by any of the input trees can be assumed to occur
-in the supertree. We may then solve separately for what happens on
-each side of such an uncontested edge. When the uncontested edge is
-attached to the solution below the edge and the solution above the
-edge, the complete solution is then obtained.
-
-\paragraph{Splits-based synthesis method}
-
-Supertrees for each subproblem are obtained by greedily constructing
-a set of rooted splits $\mathcal{C}$ that are jointly consistent
-and correspond to branches of the input trees. We construct an ordered
-list of splits $S(T_{i})$ by walking the branches of each tree $T_{i}$
-in post-order and appending the corresponding splits to the list.
-These lists are combined to form the list $\Sigma=S(T_{1})+S(T_{2})+\ldots+S(T_{n})+S(\mathbb{T})$,
-where '$+$' denotes concatenation. The set $C$ is initialized to
-the empty set, and we then iteratively consider each split in the
-list $\Sigma$ and insert it into $C$ if the resulting set remains
-jointly compatible. The BUILD algorithm is used both to check compatibility,
-and to construct a compatible tree from the final set $C$.
+\paragraph{Previous algorithm}
+
+The supertree algorithm of \citet{redelings2017supertree} takes as
+input a ranked list $\phyloinputs=\left\{ T_{1},\ldots,T_{n}\right\} $
+of leaf-labeled rooted input phylogenies and a single rooted taxonomy
+tree $\taxonomy_{p}$ that is ranked below all of the input trees.
+As discussed in \citet{redelings2017supertree}, the preferred tree
+would display as many of the highly ranked input splits as possible.
+If computational barriers were not an issue, the method would act
+by first producing a ranked list of rooted splits: $\mathcal{S}=S(T_{1})+S(T_{2})+\ldots+S(T_{n})+S(\mathbb{T})$
+where $S(T_{i})$ denotes a list of non-trivial rooted splits created
+from a post-order traversal of tree $i$ and '$+$' denotes concatenation.
+Then the supertree would be produced in a greedy manner by iterating
+over the list $S$, and adding each rooted split to a growing rooted
+supertree if that split is compatible with the supertree. In other
+words, the set $C$ is initialized to the empty set, and we then iteratively
+consider each split in the list $\mathcal{S}$ and insert it into
+$C$ if the resulting set remains jointly compatible. The BUILD algorithm\citep{AhoSSU1981}
+is used both to check compatibility, and to construct a compatible
+tree from the final set $C$.
+
+Because of the large number (over 2.6 million) of tips in the full
+tree, the supertree method of \citet{redelings2017supertree} uses
+an approximation approach that relies on decomposition of the full
+problem into subproblems. The decomposition partitioning the list
+of taxonomy splits $S(\taxonomy_{p})$ into two lists: $\mathcal{Z}(\taxonomy_{p}),$the
+list of splits from the taxonomy tree which are compatible with every
+rooted split in the phylogenetic inputs, and $\mathcal{Y}(\taxonomy_{p})$,
+the list of taxonomic splits that conflict with at least one phylogenetic
+input split. Conceptually the approximate algorithm works by greedily
+adding all compatible splits in altered ranked list of splits: $\mathcal{S}^{\prime}=\mathcal{Z}(\taxonomy_{p})+S(T_{1})+S(T_{2})+\ldots+S(T_{n})+\mathcal{Y}(\taxonomy_{p})$.
+
+In practice, the decomposition allows the compatibility of splits
+to be performed on smaller, relabeled trees. This can be done much
+more efficiently than if compatibility was checked using the BUILD
+algorithm of \citet{AhoSSU1981} on the full leaf set. Thus, the supertree
+algorithm first divides the supertree problem into independent pieces
+by bisecting trees at the uncontested taxa which correspond to $Z$,
+then solves the supertree problem on these pieces, and finally glues
+the pieces back together to produce the full tree.
+
+The taxonomy tree, $\taxonomy_{p}$, is derived from OTT. As mentioned
+above, OTT not only contains a taxonomic hierarchy, but also contains
+taxonomic labels and annotations (referred to as ``flags'') on the
+taxa to carry extra information. The entire hiearchy encoded by OTT
+is a complete taxonomic tree, $\taxonomy^{\ast}.$ Nodes in $\taxonomy^{\ast}$
+have a one-to-one correspondence with a set of labels $\mathcal{N}$
+that are called taxon names. We will decompose $\mathcal{N}$ into
+the taxon names for leaf taxa \emph{$\mathcal{L}$ }and higher taxa
+$\mathcal{H}$. We will use $\mathcal{U}$ to refer to the subset
+($\mathcal{U}\subseteq\mathcal{N}$) of taxa that are annotated with
+the \emph{incertae sedis }property. Other annotations identify dubious
+taxa and taxonomic names that are unwanted artifacts of the process
+of building OTT. Thus, the taxonomic tree which is used as an input
+to the supertree algorithm is pruned version of $\taxonomy^{\ast}$.
+For our purposes, the relevant information from the taxonomy consists
+of the taxonomy tree $\taxonomy^{\ast}$, the labels $\mathcal{N}$,
+and set $\mathcal{U}$ of \emph{incertae sedis }taxa. In previous
+work, taxa in $\mathcal{U}$ were all pruned from the taxonomic hierarchy
+when $\taxonomy_{p}$ is extracted from OTT. In the present work,
+we consider supertree approaches that operate on a taxonomic tree
+$\taxonomy,$which is produced from $\taxonomy^{\ast}$but does not
+prune \emph{incertae sedis} taxa.
+
+Throughout we assume that leaf labels on input phylogenetic trees
+are only taken from the leaf set of the taxonomic tree that is being
+used and that each label occurs at most once in each input tree. \citet{redelings2017supertree}
+describe a method called ``exemplification'' that ensures this is
+the case. Thus pruning of the \emph{incertae sedis} taxa from the
+taxonomy was also accompanied by pruning those taxa from the input
+phylogenetic trees.
 
@@
 taxa}
+
+\paragraph{Equivalence of a taxonomy tree and a list of rooted splits}
+
+Note that, for the purposes of the supertree method, all of the information
+from the taxonomic tree is contained in the set of rooted splits.
+Thus, one could imagine replacing that taxonomy, $\taxonomy_{p}$,
+with a set of trees $\mathcal{R}(\taxonomy_{p})$ each of which has
+a single internal edge such that each edge in $\taxonomy_{p}$is included
+in one tree in $\mathcal{R}(\taxonomy_{p})$. If the order of trees
+is set to be (with some abuse of notation): $\mathcal{R}(\taxonomy_{p})=\mathcal{Z}(\taxonomy_{p})+\mathcal{Y}(\taxonomy_{p})$,
+then using $\mathcal{R}(\taxonomy_{p})$ in place of $\taxonomy_{p}$in
+the supertree algorithm would produce the same output.
+
+We define rooted splits for each edge by noting that removing the
+edge would produce two trees. The leaf label sets of these trees the
+partition the full leaf set $\mathcal{L}$ into two groups. The include
+set $\mathcal{I}(e)$ which does not contain the root, and the exclude
+set $\mathcal{E}(e)$ which does contain the root. Such a split may
+be written
+\[
+\mathcal{I}(e)|\bullet\mathcal{E}(e).
+\]
+where the $\bullet$ indicates the root. The include set is the set
+of leaves that are descendants of the daughter node of the edge. The
+exclude set is the set of leaves that are not descendants of that
+node. If no taxa are \emph{incertae sedis}, then the exclude set for
+a node is just the total tip set minus the include set for the node:
+\begin{align*}
+\mathcal{E}(e) & =\mathcal{L}-\mathcal{I}(e).
+\end{align*}
+
+For a node $n$ on the tip-ward side of an edge $e$, we may also
+write $\mathcal{I}(n)$ for $\mathcal{I}(e)$, and $\mathcal{E}(n)$
+for $\mathcal{E}(e)$. For any two rooted splits $A=A_{1}|\bullet A_{2}$
+and $B=B_{1}|\bullet B_{2}$ , we say that $A$ \emph{displays }$B$
+if $B_{1}\subseteq A_{1}$ and $B_{2}\subseteq A_{2}$.
+
+Note that when a taxon rooted at node $n$ is not marked as \emph{incertae
+sedis}, then the leaf labels that are descendants of $n$ will be
+in the include set for $n$ and all its ancestors and will be in the
+exclude set for all other nodes/edges.
+
+\paragraph{Information content of the \emph{incertae sedis }annotation}
+
+In order to incorporate \emph{incertae sedis }taxa into a supertree
+analyses, we must arrive at an operational definition of the meaning
+of the \emph{incertae sedis} label. The core meaning of the phrase
+is that the annotated taxon may actually be a member of one of the
+taxa that appear as siblings to it in the heirarchical representation
+of the taxonomy. In the most expansive interpretation, the author
+of the taxonomy is stating that the \emph{incertae sedis} taxon could
+be placed anywhere inside one of the sibling taxa or their descendants.
+However, frequently only a subset of the possible placement points
+would be viewed as plausible. Ideally, the \emph{incertae sedis} annotation
+some taxon \emph{x} would be accompanied by a specific statement about
+which other taxa might include \emph{x}. Detailed readings of the
+character argumentation used by taxonomists could provide this sort
+of specific information in many cases, but we lack such information
+on the scale of the entire OTT. Here we consider a few operational
+approaches for handling the \emph{incertae sedis} label.
+
+In terms of rooted split interpretation, an \emph{incertae sedis}
+taxon differs from a normal taxon because its leaves should not automatically
+be mentioned in the exclude set of all of the edges that does not
+descend from. If a taxon rooted at node $n$ is marked as \emph{incertae
+sedis}, then the leaf labels that are descendants of $n$ will still
+be in the include set for $n$ and all its ancestors, but these labels
+will not necessarily be the exclude set for all other nodes/edges.
+They may simply be omitted in some rooted split representations, resulting
+in partial splits.
+
+\paragraph{Naive interpretation: ignoring the annotation}
+
+The most straightforward approach for increasing the taxonomic completeness
+of the supertree is simply to no longer prune taxa from $\taxonomy$
+because they are flagged as \emph{incertae sedis}, and then run the
+rest of the pipeline of \citet{redelings2017supertree} unaltered.
+However this approach results in a less resolved tree because taxonomic
+vagueness (in the form of the \emph{incertae sedis} annotation) is
+interpreted as an unwarranted statement that the \emph{incertae sedis}
+taxa are disallowed from being members of their sibling taxa.
+
+DO WE HAVE SOME NUMBERS ON THIS (PRESUMABLY FROM JUST RUNNING THE
+ONLY PROPINQUITY WITH A DIFFERENT SET OF PRUNE FLAGS)?
+
+\paragraph{Rich taxon-specific attachment points}
+
+As mentioned above, the incertae sedis may mean that a taxon is unplaced,
+but only a few placements within the taxonomy are possible. For example,
+a genus may not be characterized as incertae sedis at the level of
+the subfamily. Thus, it could be placed inside any of the hypothesized
+sub-families, or within its own subfamily, but it would be unexpected
+for it to be placed inside one of the other genera within a family.
+
+In principle this information is easy to incorporate. If $\mathcal{P}$is
+the set of taxa that are potential ancestors for incertae sedis taxon
+$x$, then $\mathcal{L}(x)$ will occur in neither the include nor
+the exclude set of any taxon in $\mathcal{P}$. Unfortunately, utilizing
+taxon-specific interpretations of the incertae sedis flag relies on
+character argumentation (which is lacking in OTT) or rank-based information
+(which is frequently untrustworthy in OTT). Thus, here we pursue an
+approach of ascribing a meaning to the \emph{incertae sedis} label
+which attempts to capture the core of the idea that it articulates,
+but which can be applied automatically across the taxonomy without
+additional information.
+
+\paragraph{Splits-based synthesis method}
 
@@
 \begin{enumerate}
-\item An incertae sedis node may intrude into its siblings and their descendants.
-\item Two incertae sedis siblings may not be interdigitated.
-\item Any placement $T\to T^{*}$ of a tree $T$ with incertae sedis taxa
-is also accessible from a tree $T^{\prime}$ that is created from
-$T$ by taking a taxon, attaching it to an ancestor, and marking it
-\emph{incertae sedis}.
+\item An \emph{incertae sedis} node may intrude into its siblings and their
+descendants.
+\item Two \emph{incertae sedis} siblings may not be interdigitated.
+\item Any placement $T\to T^{*}$ of a tree $T$ with \emph{incertae sedis}
+taxa is also accessible from a tree $T^{\prime}$ that is created
+from $T$ by taking a taxon, attaching it to an ancestor, and marking
+it \emph{incertae sedis}.%
+\begin{comment}
+MTH: I don't understand this
+\end{comment}
 \item The semantics of an \emph{incertae sedis} node does not depend on
@@
 
+For any two rooted splits $A=A_{1}|\bullet A_{2}$ and $B=B_{1}|\bullet B_{2}$
+, we say that $A$ \emph{displays }$B$ if $B_{1}\subseteq A_{1}$
+and $B_{2}\subseteq A_{2}$.
+
 \subsubsection{A semantics that disallows inter-digitating \emph{incertae sedis}
@@
 The reduced exclude set $\mathcal{E}(n)$ for node $n$ should therefore
-not contain the descendants of its siblings marked \emph{incertae
+not contain the descendants of its siblings which are marked \emph{incertae
 }sedis\emph{,} but should contain the descendants of its other siblings\emph{.
@@
 \end{align}
-This formula allows us to compute exclude sets via a pre-order recursion
-on the taxonomy tree. This recursion terminates if the exclude set
-for the root node is set to $\emptyset$ as a boundary condition.
+This formula allows us to compute exclude sets via a pre-order traversal
+on the taxonomy tree with exclude set for the root node defined to
+be $\emptyset$.
 
@@
 where the first semantics allows \emph{incertae sedis }taxa to intrude
-into\emph{ incertae sedis }descendants of siblings, but not allow\emph{
+into \emph{incertae sedis} descendants of siblings, but not allow\emph{
 incertae sedis} descendants of siblings to intrude into \emph{incertae
@@
 
-After constructing a supertree, we still need to assign taxon names
-to the supertree nodes based on the taxonomy tree in the problem.
-Each taxon name $n$ corresponds to a split $S(n)=S(n)_{1}|\bullet S(n)_{2}$
-on the corresponding branch of the taxonomy tree. Without \emph{incertae
-sedis}, such splits are always of the form $S(n)_{1}|\bullet\mathcal{L}-S(n)_{1}$,
-but with \emph{incertae sedis} taxa $S_{2}(n)$ may be smaller than
-$\mathcal{L}-S(n)_{1}$.
+After constructing a supertree, tip nodes already have names in \emph{$\mathcal{L}$}.
+However, we still need to assign higher taxon names to internal supertree
+nodes based on the taxonomy tree in the problem. Each taxon name $n$
+corresponds to a split $S(n)=S(n)_{1}|\bullet S(n)_{2}$ on the corresponding
+branch of the taxonomy tree. Without \emph{incertae sedis}, such splits
+are always of the form $S(n)_{1}|\bullet\mathcal{L}-S(n)_{1}$, but
+with \emph{incertae sedis} taxa $S_{2}(n)$ may be smaller than $\mathcal{L}-S(n)_{1}$.
 
@@
 corresponds to the split $A1\,A2\,|\bullet B\,C$, leaving out $D$
-since it is \emph{incertae sedis}. The two nodes $x$ and $y$ imply
+since it is \emph{incertae sedis}. The two nodes $x$ and $y$ display
 the splits $A1\,A2\,|\bullet B\,C\,D$ and $A1\,A2\,D|\bullet B\,C$
-respectively, and both of these splits imply the split $A1\,A2\,|\bullet B\,C$,
-so the name $A$ can apply to both $x$ and $y$\footnote{Define ``implies'', as in the propinquity paper: $A_{1}|\bullet A_{2}\implies B_{1}|\bullet B_{2}$
-means $B_{1}\subseteq A_{1}$ and $B_{2}\subseteq A_{2}$.}. This cannot happen without\emph{ incertae sedis} taxa except at
-monotypic nodes. When faced with a choice about where to place a name,
-our solution is to find the most tip-ward node where the name can
-apply and attach the name to this node. This corresponds to a conservative
-choice about whether \emph{incertae sedis} taxa should be added to
-an existing taxon.
+respectively, and both of these splits display the split $A1\,A2\,|\bullet B\,C$,
+so the name $A$ can apply to both $x$ and $y$. This cannot happen
+without\emph{ incertae sedis} taxa except at monotypic nodes. When
+faced with a choice about where to place a name, our solution is to
+find the most tip-ward node where the name can apply and attach the
+name to this node. This corresponds to a conservative choice about
+whether \emph{incertae sedis} taxa should be added to an existing
+taxon.
 \begin{figure}
@@
 
-\subsection{Placement}
+\subsection{\label{subsec:Placement}Placement}
 
@@
 like to interpret the position of these names in terms of placing
-\emph{incertae sedis }taxa in some spot, possibly a new spot. This
-would allow us to interpret the synthesis tree as saying that phylogenetic
-information has (for example) placed genus A within family B, or perhaps
-in a monotypic, unnamed family that contains only $A$. The simplest
-approach to placement involves noting whenever a taxon $B$ is a descendant
-of a taxon $A$ on the named synthesis tree but not the taxonomy tree.
-For example, in Figure \ref{fig:Handling-incertae-sedis}, taxon $C$
-is a descendant of taxon $A$ in the synthetic tree (Fig. \ref{fig:Handling-incertae-sedis}d),
+\emph{incertae sedis }taxa in a revised taxonomy. This would allow
+us to interpret the synthesis tree as saying that phylogenetic information
+has (for example) placed genus A within family B, or perhaps outside
+of all named families. The simplest approach to placement involves
+noting whenever a taxon $B$ is a descendant of a taxon $A$ on the
+named synthesis tree but not the taxonomy tree. For example, in Figure
+\ref{fig:Handling-incertae-sedis}, taxon $C$ is a descendant of
+taxon $A$ in the synthetic tree (Fig. \ref{fig:Handling-incertae-sedis}d),
 but not the taxonomy (Fig. \ref{fig:Handling-incertae-sedis}b). Thus,
@@
 \item The edge leading to clade $B$ is lost.
-\item Any taxonomy-only taxa are placed at the MRCA of the phylogenetically
+\item Any taxonomy-only children of $B$ are placed at the MRCA of the phylogenetically
 informed taxa, instead of at node $B$.
@@
 
-When the synthesis tree conflicts with a taxonomy node, we way that
+When the synthesis tree conflicts with a taxonomy node, we say that
 the taxon $B$ at that node is a broken taxon. Broken taxa have two
@@
 
-\subfloat[Synthesis Tree \#1]{\includegraphics[scale=0.5]{Figures/is-clade/synth1-no-is-shaded}
-
-}\hfill{}\subfloat[Synthesis Tree \#1]{\includegraphics[scale=0.5]{Figures/is-clade/synth1-shaded}
+\subfloat[Synthesis Tree (no \emph{incertae sedis})]{\includegraphics[scale=0.5]{Figures/is-clade/synth1-no-is-shaded}
+
+}\hfill{}\subfloat[Synthesis Tree (\emph{incertae sedis} aware)]{\includegraphics[scale=0.5]{Figures/is-clade/synth1-shaded}
 
@@
 
-\emph{This case makes placement more complicated. Use a sequence of
-ordered placements, preorder?}
+\emph{}
 \begin{figure}
-
 \subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-within-is/tree1\lyxdot tre}
@@
 A.}
-
 \end{figure}
@@
 
-Additionally, it should also be possible to filter unplaced taxa in
-the tree viewer instead of in the synthesis pipeline.
+We should be able to to figure out which incertae sedis taxa are placed
+(see section \ref{subsec:Placement}), and then mark the other ones
+incertae sedis in the \emph{synthesis tree}. We could then filter
+out the taxa that \emph{remain }incertae sedis (or incertae\_sedis
+inherited) in the tree viewer instead of the synthesis pipeline.
 
@@
 have $((X)B)A)$ in one input tree, and $X$ is mentioned nowhere
-else, then by marking $B$ as contested, we are merging subproblems
-unnecessarily. We could instead placed $X$ in $B$ and avoid contesting
-the edge $B$. However, this approach is more complex and does not
-seem necessary in practice.
+else, then by marking $B$ as contested, we are creating unnecessarily
+large subproblems. We could instead placed $X$ in $B$ and avoid
+contesting the edge $B$. However, this approach is more complex and
+does not seem necessary in practice.
 

### 2018-04-19 doc/incertae_sedis.tex
note 1

@@
 \global\long\def\collections{\mathcal{C}}
+
+\newcommand{\note}[1]{}
 
@@
 phylogenetic trees.
+
+\note{
+    1 Details & Methods
+
+1.1 Some examples in OTT This paragraph pulled from intro to otcetera/doc/handling-incertae-sedis.pdf
+
+The problem is becoming more acute because NCBI is putting a larger number of taxa into groups that are marked as “unclassified.” For example, when OTT 2.9 was created NCBI’s clasification of the bird family Sylvidae included a group five genera that were placed within “[http://www.ncbi.nlm.nih.gov/Taxonomy/Browser/wwwtax.cgi?mode=Undef&id=36270&lvl=3&lin=f&keep=1&srchmode=1&unlock||unclassified Sylviidae].” This includes the genus Regulus.
+Thus in OTT 2.9 Regulus (ott ID = 3599326) is placed inside the family Silvidae (OTT ID = 259942).
+The “unclassified Sylviidae” does not appear in OTT; instead Regulus is flagged as “unclassified,sibling higher” and all of the species within Regulus are flagged as “unclassified inherited.”
+
+The taxonomy (as of version 2.9) also contains 685 cases of taxa that are flagged as both “unclassified inherited” and “unclassified”based on grep unclass.*unclassified\_inherited taxonomy.tsv.
+
+We would like to stop suppressing (pruning) incertae sedis taxa, so that groups such as Regulus can appear appear in the synthetic tree.
+
+While we get Regulus back, we do not end up placing it in Sylviidae because Sylviidae is broken. It conflicts with about 20 different trees.
+
+1.2 Taxonomy details
+
+Flags.
+The Open Tree’s reference taxonomy is produced by a tool called smasher that notices hints that a taxon is incertae sedis and labels that taxon with one of five flagssee https://github.com/OpenTreeOfLife/reference-taxonomy/wiki/Taxon-flags.This document is not concerned with the flagging system per se, so “incertae sedis” will be used here to refer to all of the flags that denote taxa with uncertain placement..
+While a taxonomist may intend to use incertae sedis to indicate a limited number of possible positions for a taxon to go in the taxonomy, we do not retain any such details.
+
+Container nodes.
+Some taxonomies have nodes with names like “Incertae sedis (Bacteria)”.
+Such nodes are called containers.
+They indicate that each child node is an incertae sedis child of the container node's parent.
+The OTT retains container nodes with the was_container flag, but seems to have already moved all of the container's children to the container's parent.
+}
 

### 2018-04-19 doc/incertae_sedis.tex
line wrapping

@@
-%% LyX 2.3.0 created this file.  For more info, see http://www.lyx.org/.
-%% Do not edit unless you really know what you are doing.
 \documentclass[english]{article}
@@
 \usepackage{babel}
+\usepackage{xspace}
+\usepackage{color}
 \begin{document}
+
 \global\long\def\taxonomy{\mbox{\ensuremath{\mathbb{T}}}}
@@
 
-\title{\texttt{Taxonomic supertree construction with}\texttt{\emph{ Incertae
-sedis}}\texttt{ taxa}}
-
-\author{Benjamin D. Redelings and Mark T. Holder}
-\maketitle
+\newcommand{\mthcomment}[1]{{\color{red} #1}\xspace}
+
+\title{\texttt{Taxonomic supertree construction with}\texttt{\emph{
+Incertae sedis}}\texttt{ taxa}}
+
+\author{Benjamin D. Redelings and Mark T. Holder} maketitle
 
@@
 
-The Open Tree of Life project \citep{HinchliffEtAl2015} seeks to
-build a comprehensive supertree \citep[see][]{redelings2017supertree}
-covering all of life. The approach is to combine information in published
-phylogenies with a comprehensive taxonomy that supplies taxon names.
-The supertree method has been used with the Open Tree of Taxonomy
-\citep[OTT hereafter, see][]{rees2017automated}, but in principle
-is general. The taxonomy is a crucial input for several reasons: it
-covers a wider range of species, its list of names allows for alignment
-of different phylogenetic estimates to each other, and it provides
-names for clades. Only a small proportion of known species have been
-included in a phylogenetic analysis, thus a taxonomy is crucial for
-achieving comprehensive coverage of known taxa. Phylogenetic estimates
-collected from the published literature often use different names
-for the same species. Lists of synonyms in OTT (and other taxonomies)
-allow data curators to ``align'' input phylogenies to the taxonomy
-\citep[see][ for discussion of the curation process in the Open Tree of Life project]{McTavishEtAt2015}.
-This alignment is important for recognizing when two different estimates
-refer to the same taxon. Biologists are familiar with names for ``higher''
-taxa (taxa above the species rank). Thus, naming clades in the supertree
-makes the tree easier to navigate and use.
+The Open Tree of Life project \citep{HinchliffEtAl2015} seeks to build
+a comprehensive supertree \citep[see][]{redelings2017supertree}
+covering all of life. The approach is to combine information in
+published phylogenies with a comprehensive taxonomy that supplies
+taxon names. The supertree method has been used with the Open Tree of
+Taxonomy \citep[OTT hereafter, see][]{rees2017automated}, but in
+principle is general. The taxonomy is a crucial input for several
+reasons: it covers a wider range of species, its list of names allows
+for alignment of different phylogenetic estimates to each other, and
+it provides names for clades. Only a small proportion of known species
+have been included in a phylogenetic analysis, thus a taxonomy is
+crucial for achieving comprehensive coverage of known taxa.
+Phylogenetic estimates collected from the published literature often
+use different names for the same species. Lists of synonyms in OTT
+(and other taxonomies) allow data curators to ``align'' input
+phylogenies to the taxonomy \citep[see][ for discussion of the
+curation process in the Open Tree of Life project]{McTavishEtAt2015}.
+This alignment is important for recognizing when two different
+estimates refer to the same taxon. Biologists are familiar with names
+for ``higher'' taxa (taxa above the species rank). Thus, naming clades
+in the supertree makes the tree easier to navigate and use.
 
@@
 described here does not consider any hybridization. Thus the taxonomic
-hierarchy can be converted to a tree. While not all taxonomists believe
-that named taxa should correspond to clades \citep[see, for example,][]{HorandlS2010},
-the priniciples of phylogenetic classification are used by such a
-large majority of practicing taxonomists, that we have chosen to treat
-the taxonomic tree that mirrors the hierarchy of the taxonomy as an
-estimate of phylogenetic relationships. We refer to labels of taxonomy
-nodes as ``taxon names'', and assume that taxon names and taxonomy
-nodes have a one-to-one correspondence. Note that the taxonomy may
-also contain out-degree 1 nodes. These nodes correspond to monotypic
-taxa which contain a single child of lower taxonomic rank.
-
-The supertree method described by \citet{redelings2017supertree}
-is able to use a set of phylogentic estimates and taxonomy to produce
-a supertree with clades names according to the taxonomy. However,
-the method was not able to accommodate the fact that taxonomies frequently
+hierarchy can be converted to a tree. While not all taxonomists
+believe that named taxa should correspond to clades \citep[see, for
+example,][]{HorandlS2010}, the priniciples of phylogenetic
+classification are used by such a large majority of practicing
+taxonomists, that we have chosen to treat the taxonomic tree that
+mirrors the hierarchy of the taxonomy as an estimate of phylogenetic
+relationships. We refer to labels of taxonomy nodes as ``taxon
+names'', and assume that taxon names and taxonomy nodes have a one-to-
+one correspondence. Note that the taxonomy may also contain out-degree
+1 nodes. These nodes correspond to monotypic taxa which contain a
+single child of lower taxonomic rank.
+
+The supertree method described by \citet{redelings2017supertree} is
+able to use a set of phylogentic estimates and taxonomy to produce a
+supertree with clades names according to the taxonomy. However, the
+method was not able to accommodate the fact that taxonomies frequently
 contain nodes with uncertain placement. These nodes are often labelled
@@
 as ``\emph{incertae familia}'', indicating that it is unclear which
-family the genus should be placed in. \emph{Incertae sedis} taxa frequently
-occur when a specimen is identified down to a given taxonomic level,
-but no further. Extinct taxa are often \emph{incertae sedis}. OTT
-is constructed from several source taxonomies, and these sources include
-various ways of indicating that the position of a taxon within the
-hierarchy is uncertain. OTT uses a series of ``flags'' to annotate
-these taxa. The supertree of by \citet{redelings2017supertree} simply
-pruned these taxa from the taxonomy and input trees. Thus, the final
-supertree lacked taxa which were \emph{incertae sedis}.
+family the genus should be placed in. \emph{Incertae sedis} taxa
+frequently occur when a specimen is identified down to a given
+taxonomic level, but no further. Extinct taxa are often \emph{incertae
+sedis}. OTT is constructed from several source taxonomies, and these
+sources include various ways of indicating that the position of a
+taxon within the hierarchy is uncertain. OTT uses a series of
+``flags'' to annotate these taxa. The supertree of by
+\citet{redelings2017supertree} simply pruned these taxa from the
+taxonomy and input trees. Thus, the final supertree lacked taxa which
+were \emph{incertae sedis}.
 
@@
 uncertainty by using published phylogenies to place \emph{incertae
-sedis} taxa. Developing this method required deciding on a set operational
-semantics to be used when interpreting the \emph{incertae sedis} annotation.
-These semantics affect the interpretation of the phylogenetic statements
-being made by the input taxonomy and the rules for applying taxonomic
-names to the final supertree. In addition to discussing how to interpret
-in the \emph{incertae sedis} annotation, we describe the algorithmic
-changes to the supertree pipeline that were required to adequately
-represent the taxonomic uncertainty. The approach described here allows
-us to include thousands of new taxa in our supertree analysis that
-were previously filtered out. It will also enable the Open Tree of
-Life project to include extinct taxa, since many of these taxa are
-\emph{incertae sedis}. This makes substantial progress towards our
-goal of comprehensive inclusion of known taxa into the Open Tree of
-Life summary tree.
+sedis} taxa. Developing this method required deciding on a set
+operational semantics to be used when interpreting the \emph{incertae
+sedis} annotation. These semantics affect the interpretation of the
+phylogenetic statements being made by the input taxonomy and the rules
+for applying taxonomic names to the final supertree. In addition to
+discussing how to interpret in the \emph{incertae sedis} annotation,
+we describe the algorithmic changes to the supertree pipeline that
+were required to adequately represent the taxonomic uncertainty. The
+approach described here allows us to include thousands of new taxa in
+our supertree analysis that were previously filtered out. It will also
+enable the Open Tree of Life project to include extinct taxa, since
+many of these taxa are \emph{incertae sedis}. This makes substantial
+progress towards our goal of comprehensive inclusion of known taxa
+into the Open Tree of Life summary tree.
 
@@
 of leaf-labeled rooted input phylogenies and a single rooted taxonomy
-tree $\taxonomy_{p}$ that is ranked below all of the input trees.
-As discussed in \citet{redelings2017supertree}, the preferred tree
-would display as many of the highly ranked input splits as possible.
-If computational barriers were not an issue, the method would act
-by first producing a ranked list of rooted splits: $\mathcal{S}=S(T_{1})+S(T_{2})+\ldots+S(T_{n})+S(\mathbb{T})$
-where $S(T_{i})$ denotes a list of non-trivial rooted splits created
-from a post-order traversal of tree $i$ and '$+$' denotes concatenation.
-Then the supertree would be produced in a greedy manner by iterating
-over the list $S$, and adding each rooted split to a growing rooted
+tree $\taxonomy_{p}$ that is ranked below all of the input trees. As
+discussed in \citet{redelings2017supertree}, the preferred tree would
+display as many of the highly ranked input splits as possible. If
+computational barriers were not an issue, the method would act by
+first producing a ranked list of rooted splits:
+$\mathcal{S}=S(T_{1})+S(T_{2})+\ldots+S(T_{n})+S(\mathbb{T})$ where
+$S(T_{i})$ denotes a list of non-trivial rooted splits created from a
+post-order traversal of tree $i$ and '$+$' denotes concatenation. Then
+the supertree would be produced in a greedy manner by iterating over
+the list $S$, and adding each rooted split to a growing rooted
 supertree if that split is compatible with the supertree. In other
-words, the set $C$ is initialized to the empty set, and we then iteratively
-consider each split in the list $\mathcal{S}$ and insert it into
-$C$ if the resulting set remains jointly compatible. The BUILD algorithm\citep{AhoSSU1981}
-is used both to check compatibility, and to construct a compatible
-tree from the final set $C$.
+words, the set $C$ is initialized to the empty set, and we then
+iteratively consider each split in the list $\mathcal{S}$ and insert
+it into $C$ if the resulting set remains jointly compatible. The BUILD
+algorithm\citep{AhoSSU1981} is used both to check compatibility, and
+to construct a compatible tree from the final set $C$.
 
 Because of the large number (over 2.6 million) of tips in the full
-tree, the supertree method of \citet{redelings2017supertree} uses
-an approximation approach that relies on decomposition of the full
-problem into subproblems. The decomposition partitioning the list
-of taxonomy splits $S(\taxonomy_{p})$ into two lists: $\mathcal{Z}(\taxonomy_{p}),$the
-list of splits from the taxonomy tree which are compatible with every
-rooted split in the phylogenetic inputs, and $\mathcal{Y}(\taxonomy_{p})$,
-the list of taxonomic splits that conflict with at least one phylogenetic
-input split. Conceptually the approximate algorithm works by greedily
-adding all compatible splits in altered ranked list of splits: $\mathcal{S}^{\prime}=\mathcal{Z}(\taxonomy_{p})+S(T_{1})+S(T_{2})+\ldots+S(T_{n})+\mathcal{Y}(\taxonomy_{p})$.
-
-In practice, the decomposition allows the compatibility of splits
-to be performed on smaller, relabeled trees. This can be done much
-more efficiently than if compatibility was checked using the BUILD
-algorithm of \citet{AhoSSU1981} on the full leaf set. Thus, the supertree
-algorithm first divides the supertree problem into independent pieces
-by bisecting trees at the uncontested taxa which correspond to $Z$,
-then solves the supertree problem on these pieces, and finally glues
-the pieces back together to produce the full tree.
+tree, the supertree method of \citet{redelings2017supertree} uses an
+approximation approach that relies on decomposition of the full
+problem into subproblems. The decomposition partitioning the list of
+taxonomy splits $S(\taxonomy_{p})$ into two lists:
+$\mathcal{Z}(\taxonomy_{p}),$the list of splits from the taxonomy tree
+which are compatible with every rooted split in the phylogenetic
+inputs, and $\mathcal{Y}(\taxonomy_{p})$, the list of taxonomic splits
+that conflict with at least one phylogenetic input split. Conceptually
+the approximate algorithm works by greedily adding all compatible
+splits in altered ranked list of splits:
+$\mathcal{S}^{\prime}=\mathcal{Z}(\taxonomy_{p})+S(T_{1})+S(T_{2})+\l
+dots+S(T_{n})+\mathcal{Y}(\taxonomy_{p})$.
+
+In practice, the decomposition allows the compatibility of splits to
+be performed on smaller, relabeled trees. This can be done much more
+efficiently than if compatibility was checked using the BUILD
+algorithm of \citet{AhoSSU1981} on the full leaf set. Thus, the
+supertree algorithm first divides the supertree problem into
+independent pieces by bisecting trees at the uncontested taxa which
+correspond to $Z$, then solves the supertree problem on these pieces,
+and finally glues the pieces back together to produce the full tree.
 
@@
 taxonomic labels and annotations (referred to as ``flags'') on the
-taxa to carry extra information. The entire hiearchy encoded by OTT
-is a complete taxonomic tree, $\taxonomy^{\ast}.$ Nodes in $\taxonomy^{\ast}$
-have a one-to-one correspondence with a set of labels $\mathcal{N}$
-that are called taxon names. We will decompose $\mathcal{N}$ into
-the taxon names for leaf taxa \emph{$\mathcal{L}$ }and higher taxa
-$\mathcal{H}$. We will use $\mathcal{U}$ to refer to the subset
-($\mathcal{U}\subseteq\mathcal{N}$) of taxa that are annotated with
-the \emph{incertae sedis }property. Other annotations identify dubious
-taxa and taxonomic names that are unwanted artifacts of the process
-of building OTT. Thus, the taxonomic tree which is used as an input
-to the supertree algorithm is pruned version of $\taxonomy^{\ast}$.
-For our purposes, the relevant information from the taxonomy consists
-of the taxonomy tree $\taxonomy^{\ast}$, the labels $\mathcal{N}$,
-and set $\mathcal{U}$ of \emph{incertae sedis }taxa. In previous
-work, taxa in $\mathcal{U}$ were all pruned from the taxonomic hierarchy
-when $\taxonomy_{p}$ is extracted from OTT. In the present work,
-we consider supertree approaches that operate on a taxonomic tree
-$\taxonomy,$which is produced from $\taxonomy^{\ast}$but does not
-prune \emph{incertae sedis} taxa.
-
-Throughout we assume that leaf labels on input phylogenetic trees
-are only taken from the leaf set of the taxonomic tree that is being
-used and that each label occurs at most once in each input tree. \citet{redelings2017supertree}
-describe a method called ``exemplification'' that ensures this is
-the case. Thus pruning of the \emph{incertae sedis} taxa from the
-taxonomy was also accompanied by pruning those taxa from the input
-phylogenetic trees.
-
-\note{
-    1 Details & Methods
-
-1.1 Some examples in OTT This paragraph pulled from intro to otcetera/doc/handling-incertae-sedis.pdf
-
-The problem is becoming more acute because NCBI is putting a larger number of taxa into groups that are marked as “unclassified.” For example, when OTT 2.9 was created NCBI’s clasification of the bird family Sylvidae included a group five genera that were placed within “[http://www.ncbi.nlm.nih.gov/Taxonomy/Browser/wwwtax.cgi?mode=Undef&id=36270&lvl=3&lin=f&keep=1&srchmode=1&unlock||unclassified Sylviidae].” This includes the genus Regulus.
-Thus in OTT 2.9 Regulus (ott ID = 3599326) is placed inside the family Silvidae (OTT ID = 259942).
-The “unclassified Sylviidae” does not appear in OTT; instead Regulus is flagged as “unclassified,sibling higher” and all of the species within Regulus are flagged as “unclassified inherited.”
-
-The taxonomy (as of version 2.9) also contains 685 cases of taxa that are flagged as both “unclassified inherited” and “unclassified”based on grep unclass.*unclassified\_inherited taxonomy.tsv.
-
-We would like to stop suppressing (pruning) incertae sedis taxa, so that groups such as Regulus can appear appear in the synthetic tree.
-
-While we get Regulus back, we do not end up placing it in Sylviidae because Sylviidae is broken. It conflicts with about 20 different trees.
+taxa to carry extra information. The entire hiearchy encoded by OTT is
+a complete taxonomic tree, $\taxonomy^{\ast}.$ Nodes in
+$\taxonomy^{\ast}$ have a one-to-one correspondence with a set of
+labels $\mathcal{N}$ that are called taxon names. We will decompose
+$\mathcal{N}$ into the taxon names for leaf taxa \emph{$\mathcal{L}$
+}and higher taxa $\mathcal{H}$. We will use $\mathcal{U}$ to refer to
+the subset ($\mathcal{U}\subseteq\mathcal{N}$) of taxa that are
+annotated with the \emph{incertae sedis }property. Other annotations
+identify dubious taxa and taxonomic names that are unwanted artifacts
+of the process of building OTT. Thus, the taxonomic tree which is used
+as an input to the supertree algorithm is pruned version of
+$\taxonomy^{\ast}$. For our purposes, the relevant information from
+the taxonomy consists of the taxonomy tree $\taxonomy^{\ast}$, the
+labels $\mathcal{N}$, and set $\mathcal{U}$ of \emph{incertae sedis
+}taxa. In previous work, taxa in $\mathcal{U}$ were all pruned from
+the taxonomic hierarchy when $\taxonomy_{p}$ is extracted from OTT. In
+the present work, we consider supertree approaches that operate on a
+taxonomic tree $\taxonomy,$which is produced from
+$\taxonomy^{\ast}$but does not prune \emph{incertae sedis} taxa.
+
+Throughout we assume that leaf labels on input phylogenetic trees are
+only taken from the leaf set of the taxonomic tree that is being used
+and that each label occurs at most once in each input tree.
+\citet{redelings2017supertree} describe a method called
+``exemplification'' that ensures this is the case. Thus pruning of the
+\emph{incertae sedis} taxa from the taxonomy was also accompanied by
+pruning those taxa from the input phylogenetic trees.
+
+\note{     1 Details & Methods
+
+1.1 Some examples in OTT This paragraph pulled from intro to
+otcetera/doc /handling-incertae-sedis.pdf
+
+The problem is becoming more acute because NCBI is putting a larger
+number of taxa into groups that are marked as “unclassified.” For
+example, when OTT 2.9 was created NCBI’s clasification of the bird
+family Sylvidae included a group five genera that were placed within
+“[http://www.ncbi.nlm.nih.gov/Taxonomy/Brows er/wwwtax.cgi?mode=Undef&
+id=36270&lvl=3&lin=f&keep=1&srchmode=1&unlock||unclassi fied
+Sylviidae].” This includes the genus Regulus. Thus in OTT 2.9 Regulus
+(ott ID = 3599326) is placed inside the family Silvidae (OTT ID =
+259942). The “unclassified Sylviidae” does not appear in OTT; instead
+Regulus is flagged as “unclassified,sibling higher” and all of the
+species within Regulus are flagged as “unclassified inherited.”
+
+The taxonomy (as of version 2.9) also contains 685 cases of taxa that
+are flagged as both “unclassified inherited” and “unclassified”based
+on grep unclass.*unclassified\_inherited taxonomy.tsv.
+
+We would like to stop suppressing (pruning) incertae sedis taxa, so
+that groups such as Regulus can appear appear in the synthetic tree.
+
+While we get Regulus back, we do not end up placing it in Sylviidae
+because Sylviidae is broken. It conflicts with about 20 different
+trees.
 
@@
 
-Flags.
-The Open Tree’s reference taxonomy is produced by a tool called smasher that notices hints that a taxon is incertae sedis and labels that taxon with one of five flagssee https://github.com/OpenTreeOfLife/reference-taxonomy/wiki/Taxon-flags.This document is not concerned with the flagging system per se, so “incertae sedis” will be used here to refer to all of the flags that denote taxa with uncertain placement..
-While a taxonomist may intend to use incertae sedis to indicate a limited number of possible positions for a taxon to go in the taxonomy, we do not retain any such details.
-
-Container nodes.
-Some taxonomies have nodes with names like “Incertae sedis (Bacteria)”.
-Such nodes are called containers.
-They indicate that each child node is an incertae sedis child of the container node's parent.
-The OTT retains container nodes with the was_container flag, but seems to have already moved all of the container's children to the container's parent.
-}
-
-\section{\label{sec:Semantics-of-incertae}Semantics of \emph{incertae sedis}
-taxa}
+Flags. The Open Tree’s reference taxonomy is produced by a tool called
+smasher that notices hints that a taxon is incertae sedis and labels
+that taxon with one of five flagssee https://github.com/OpenTreeOfLife
+/reference-taxonomy/wiki /Taxon-flags.This document is not concerned
+with the flagging system per se, so “incertae sedis” will be used here
+to refer to all of the flags that denote taxa with uncertain
+placement.. While a taxonomist may intend to use incertae sedis to
+indicate a limited number of possible positions for a taxon to go in
+the taxonomy, we do not retain any such details.
+
+Container nodes. Some taxonomies have nodes with names like “Incertae
+sedis (Bacteria)”. Such nodes are called containers. They indicate
+that each child node is an incertae sedis child of the container
+node's parent. The OTT retains container nodes with the was_container
+flag, but seems to have already moved all of the container's children
+to the container's parent. }
+
+\section{\label{sec:Semantics-of-incertae}Semantics of \emph{incertae
+sedis} taxa}
 
@@
 
-Note that, for the purposes of the supertree method, all of the information
-from the taxonomic tree is contained in the set of rooted splits.
-Thus, one could imagine replacing that taxonomy, $\taxonomy_{p}$,
-with a set of trees $\mathcal{R}(\taxonomy_{p})$ each of which has
-a single internal edge such that each edge in $\taxonomy_{p}$is included
-in one tree in $\mathcal{R}(\taxonomy_{p})$. If the order of trees
-is set to be (with some abuse of notation): $\mathcal{R}(\taxonomy_{p})=\mathcal{Z}(\taxonomy_{p})+\mathcal{Y}(\taxonomy_{p})$,
-then using $\mathcal{R}(\taxonomy_{p})$ in place of $\taxonomy_{p}$in
-the supertree algorithm would produce the same output.
-
-We define rooted splits for each edge by noting that removing the
-edge would produce two trees. The leaf label sets of these trees the
+Note that, for the purposes of the supertree method, all of the
+information from the taxonomic tree is contained in the set of rooted
+splits. Thus, one could imagine replacing that taxonomy,
+$\taxonomy_{p}$, with a set of trees $\mathcal{R}(\taxonomy_{p})$ each
+of which has a single internal edge such that each edge in
+$\taxonomy_{p}$is included in one tree in
+$\mathcal{R}(\taxonomy_{p})$. If the order of trees is set to be (with
+some abuse of notation):
+$\mathcal{R}(\taxonomy_{p})=\mathcal{Z}(\taxonomy_{p}) + 
+\mathcal{Y}(\taxonomy_{p})$, then using $\mathcal{R}(\taxonomy_{p})$ in
+place of $\taxonomy_{p}$in the supertree algorithm would produce the
+same output.
+
+We define rooted splits for each edge by noting that removing the edge
+would produce two trees. The leaf label sets of these trees the
 partition the full leaf set $\mathcal{L}$ into two groups. The include
 set $\mathcal{I}(e)$ which does not contain the root, and the exclude
-set $\mathcal{E}(e)$ which does contain the root. Such a split may
-be written
-\[
-\mathcal{I}(e)|\bullet\mathcal{E}(e).
-\]
-where the $\bullet$ indicates the root. The include set is the set
-of leaves that are descendants of the daughter node of the edge. The
-exclude set is the set of leaves that are not descendants of that
-node. If no taxa are \emph{incertae sedis}, then the exclude set for
-a node is just the total tip set minus the include set for the node:
-\begin{align*}
-\mathcal{E}(e) & =\mathcal{L}-\mathcal{I}(e).
-\end{align*}
-
-For a node $n$ on the tip-ward side of an edge $e$, we may also
-write $\mathcal{I}(n)$ for $\mathcal{I}(e)$, and $\mathcal{E}(n)$
-for $\mathcal{E}(e)$. For any two rooted splits $A=A_{1}|\bullet A_{2}$
-and $B=B_{1}|\bullet B_{2}$ , we say that $A$ \emph{displays }$B$
-if $B_{1}\subseteq A_{1}$ and $B_{2}\subseteq A_{2}$.
-
-Note that when a taxon rooted at node $n$ is not marked as \emph{incertae
-sedis}, then the leaf labels that are descendants of $n$ will be
-in the include set for $n$ and all its ancestors and will be in the
-exclude set for all other nodes/edges.
-
-\paragraph{Information content of the \emph{incertae sedis }annotation}
+set $\mathcal{E}(e)$ which does contain the root. Such a split may be
+written \[ \mathcal{I}(e)|\bullet\mathcal{E}(e). \] where the
+$\bullet$ indicates the root. The include set is the set of leaves
+that are descendants of the daughter node of the edge. The exclude set
+is the set of leaves that are not descendants of that node. If no taxa
+are \emph{incertae sedis}, then the exclude set for a node is just the
+total tip set minus the include set for the node: \begin{align*}
+\mathcal{E}(e) & =\mathcal{L}-\mathcal{I}(e). \end{align*}
+
+For a node $n$ on the tip-ward side of an edge $e$, we may also write
+$\mathcal{I}(n)$ for $\mathcal{I}(e)$, and $\mathcal{E}(n)$ for
+$\mathcal{E}(e)$. For any two rooted splits $A=A_{1}|\bullet A_{2}$
+and $B=B_{1}|\bullet B_{2}$ , we say that $A$ \emph{displays }$B$ if
+$B_{1}\subseteq A_{1}$ and $B_{2}\subseteq A_{2}$.
+
+Note that when a taxon rooted at node $n$ is not marked as
+\emph{incertae sedis}, then the leaf labels that are descendants of
+$n$ will be in the include set for $n$ and all its ancestors and will
+be in the exclude set for all other nodes/edges.
+
+\paragraph{Information content of the emph{incertae sedis }annotation}
 
@@
 analyses, we must arrive at an operational definition of the meaning
-of the \emph{incertae sedis} label. The core meaning of the phrase
-is that the annotated taxon may actually be a member of one of the
-taxa that appear as siblings to it in the heirarchical representation
-of the taxonomy. In the most expansive interpretation, the author
-of the taxonomy is stating that the \emph{incertae sedis} taxon could
-be placed anywhere inside one of the sibling taxa or their descendants.
+of the \emph{incertae sedis} label. The core meaning of the phrase is
+that the annotated taxon may actually be a member of one of the taxa
+that appear as siblings to it in the heirarchical representation of
+the taxonomy. In the most expansive interpretation, the author of the
+taxonomy is stating that the \emph{incertae sedis} taxon could be
+placed anywhere inside one of the sibling taxa or their descendants.
 However, frequently only a subset of the possible placement points
-would be viewed as plausible. Ideally, the \emph{incertae sedis} annotation
-some taxon \emph{x} would be accompanied by a specific statement about
-which other taxa might include \emph{x}. Detailed readings of the
-character argumentation used by taxonomists could provide this sort
-of specific information in many cases, but we lack such information
-on the scale of the entire OTT. Here we consider a few operational
-approaches for handling the \emph{incertae sedis} label.
+would be viewed as plausible. Ideally, the \emph{incertae sedis}
+annotation some taxon \emph{x} would be accompanied by a specific
+statement about which other taxa might include \emph{x}. Detailed
+readings of the character argumentation used by taxonomists could
+provide this sort of specific information in many cases, but we lack
+such information on the scale of the entire OTT. Here we consider a
+few operational approaches for handling the \emph{incertae sedis}
+label.
 
 In terms of rooted split interpretation, an \emph{incertae sedis}
-taxon differs from a normal taxon because its leaves should not automatically
-be mentioned in the exclude set of all of the edges that does not
-descend from. If a taxon rooted at node $n$ is marked as \emph{incertae
-sedis}, then the leaf labels that are descendants of $n$ will still
-be in the include set for $n$ and all its ancestors, but these labels
-will not necessarily be the exclude set for all other nodes/edges.
-They may simply be omitted in some rooted split representations, resulting
-in partial splits.
+taxon differs from a normal taxon because its leaves should not
+automatically be mentioned in the exclude set of all of the edges that
+does not descend from. If a taxon rooted at node $n$ is marked as
+\emph{incertae sedis}, then the leaf labels that are descendants of
+$n$ will still be in the include set for $n$ and all its ancestors,
+but these labels will not necessarily be the exclude set for all other
+nodes/edges. They may simply be omitted in some rooted split
+representations, resulting in partial splits.
 
@@
 
-The most straightforward approach for increasing the taxonomic completeness
-of the supertree is simply to no longer prune taxa from $\taxonomy$
-because they are flagged as \emph{incertae sedis}, and then run the
-rest of the pipeline of \citet{redelings2017supertree} unaltered.
-However this approach results in a less resolved tree because taxonomic
-vagueness (in the form of the \emph{incertae sedis} annotation) is
-interpreted as an unwarranted statement that the \emph{incertae sedis}
-taxa are disallowed from being members of their sibling taxa.
-
-DO WE HAVE SOME NUMBERS ON THIS (PRESUMABLY FROM JUST RUNNING THE
-ONLY PROPINQUITY WITH A DIFFERENT SET OF PRUNE FLAGS)?
+The most straightforward approach for increasing the taxonomic
+completeness of the supertree is simply to no longer prune taxa from
+$\taxonomy$ because they are flagged as \emph{incertae sedis}, and
+then run the rest of the pipeline of \citet{redelings2017supertree}
+unaltered. However this approach results in a less resolved tree
+because taxonomic vagueness (in the form of the \emph{incertae sedis}
+annotation) is interpreted as an unwarranted statement that the
+\emph{incertae sedis} taxa are disallowed from being members of their
+sibling taxa.
+
+\mthcomment{ Do we have some numbers on this (presumably from just
+running the only propinquity with a different set of prune flags)? }
 
@@
 
-As mentioned above, the incertae sedis may mean that a taxon is unplaced,
-but only a few placements within the taxonomy are possible. For example,
-a genus may not be characterized as incertae sedis at the level of
-the subfamily. Thus, it could be placed inside any of the hypothesized
-sub-families, or within its own subfamily, but it would be unexpected
-for it to be placed inside one of the other genera within a family.
-
-In principle this information is easy to incorporate. If $\mathcal{P}$is
-the set of taxa that are potential ancestors for incertae sedis taxon
-$x$, then $\mathcal{L}(x)$ will occur in neither the include nor
-the exclude set of any taxon in $\mathcal{P}$. Unfortunately, utilizing
-taxon-specific interpretations of the incertae sedis flag relies on
-character argumentation (which is lacking in OTT) or rank-based information
-(which is frequently untrustworthy in OTT). Thus, here we pursue an
-approach of ascribing a meaning to the \emph{incertae sedis} label
-which attempts to capture the core of the idea that it articulates,
-but which can be applied automatically across the taxonomy without
-additional information.
+As mentioned above, the incertae sedis may mean that a taxon is
+unplaced, but only a few placements within the taxonomy are possible.
+For example, a genus may not be characterized as incertae sedis at the
+level of the subfamily. Thus, it could be placed inside any of the
+hypothesized sub-families, or within its own subfamily, but it would
+be unexpected for it to be placed inside one of the other genera
+within a family.
+
+In principle this information is easy to incorporate. If
+$\mathcal{P}$is the set of taxa that are potential ancestors for
+incertae sedis taxon $x$, then $\mathcal{L}(x)$ will occur in neither
+the include nor the exclude set of any taxon in $\mathcal{P}$.
+Unfortunately, utilizing taxon-specific interpretations of the
+incertae sedis flag relies on character argumentation (which is
+lacking in OTT) or rank-based information (which is frequently
+untrustworthy in OTT). Thus, here we pursue an approach of ascribing a
+meaning to the \emph{incertae sedis} label which attempts to capture
+the core of the idea that it articulates, but which can be applied
+automatically across the taxonomy without additional information.
 
@@
 \begin{enumerate}
-\item An \emph{incertae sedis} node may intrude into its siblings and their
+    \item An
+\emph{incertae sedis} node may intrude into its siblings and their
 descendants.
-\item Two \emph{incertae sedis} siblings may not be interdigitated.
-\item Any placement $T\to T^{*}$ of a tree $T$ with \emph{incertae sedis}
-taxa is also accessible from a tree $T^{\prime}$ that is created
-from $T$ by taking a taxon, attaching it to an ancestor, and marking
-it \emph{incertae sedis}.%
-\begin{comment}
-MTH: I don't understand this
-\end{comment}
-\item The semantics of an \emph{incertae sedis} node does not depend on
-assigned ranks, but only on the tree.
-\item Semantics is based on deriving a split for each branch of the tree.
+\item Two \emph{incertae sedis} siblings may not be
+interdigitated.
+\item Any placement $T\to T^{*}$ of a tree $T$ with
+\emph{incertae sedis} taxa is also accessible from a tree $T^{\prime}$
+that is created from $T$ by taking a taxon, attaching it to an
+ancestor, and marking it \emph{incertae sedis}.
+\begin{comment} MTH:
+I don't understand this \end{comment} \item The semantics of an
+\emph{incertae sedis} node does not depend on assigned ranks, but only
+on the tree.
+\item Semantics is based on deriving a split for each
+branch of the tree.
 \end{enumerate}
-Here we focus on a split-based semantics instead of a semantics based
-on apply of a series of edit operations to the tree. The split-based
-semantics allows us to seek trees that are simultaneously consistent
-with a set of split-constraints. The use of a split-based semantics
-does impose some limitations. These come primarily because in the
-edit-based semantics, after we place an \emph{incertae sedis} taxon,
-it is no loner \emph{incertae sedis}, whereas with the split-based
-semantics taxa are either always \emph{incertae sedis} or never \emph{incertae
-sedis}.
+Here we focus on a split-based
+semantics instead of a semantics based on apply of a series of edit
+operations to the tree. The split-based semantics allows us to seek
+trees that are simultaneously consistent with a set of split-
+constraints. The use of a split-based semantics does impose some
+limitations. These come primarily because in the edit-based semantics,
+after we place an \emph{incertae sedis} taxon, it is no loner
+\emph{incertae sedis}, whereas with the split-based semantics taxa are
+either always \emph{incertae sedis} or never \emph{incertae sedis}.
 
 This has an effect, for example, when we consider two incertae sedis
-siblings $A$ and $B$. If we can place $A$ within $B$, and $B$
-within $A$, then a split-based semantics allows us to freely inter-digitate
-$A$ and $B$. If we do not want to allow $A$ and $B$ to be inter-digitated,
-then we cannot both $A$ within $B$ and $B$ within $A$. In addition,
-we can imagine that a family $A$ is incertae sedis within a kingdom.
-In an edit-based semantics, we could place $A$ as a sibling to a
-genus $B$, mark $A$ as non-\emph{incertae-sedis}, and then place
+siblings $A$ and $B$. If we can place $A$ within $B$, and $B$ within
+$A$, then a split- based semantics allows us to freely inter-digitate
+$A$ and $B$. If we do not want to allow $A$ and $B$ to be inter-
+digitated, then we cannot both $A$ within $B$ and $B$ within $A$. In
+addition, we can imagine that a family $A$ is incertae sedis within a
+kingdom. In an edit-based semantics, we could place $A$ as a sibling
+to a genus $B$, mark $A$ as non-\emph{incertae-sedis}, and then place
 $B$ within $A$. In a split-based semantics, we must disallow either
-$B$ within $A$ or $A$ within $B$ if we want to disallow interdigitation.
+$B$ within $A$ or $A$ within $B$ if we want to disallow
+interdigitation.
 
 As a result of this, we cannot satisfy all five of these criteria.
-Thus, we describe two alternative split-based semantics. The non-interdigitation
-semantics gets \# (1,2,4,5), and the interdigitation semantics gets
-(1,3,4,5). This makes the point that each semantics results from different
-choices about which criteria to satisfy.
-
-\subsection{Split-based semantics for \emph{incertae sedis} taxa}
-
-In order to apply supertree methods to rooted trees with \emph{incertae
-sedis} taxa, we first propose a semantics of \emph{incertae sedis
-}in terms of rooted splits. We define rooted splits by noting that
-each edge of a tree divides the tip taxa $\mathcal{L}$ into two groups:
-the include set $\mathcal{I}(e)$ which does not contain the root,
-and the exclude set $\mathcal{E}(e)$ which does contain the root.
-Such a split may be written
-\[
-\mathcal{I}(e)|\bullet\mathcal{E}(e).
-\]
-where the $\bullet$ indicates the root. If no taxa are \emph{incertae
-sedis}, then the exclude set for a node is just the total tip set
-minus the include set for the node:
-\begin{align*}
-\mathcal{E}(n) & =\mathcal{L}-\mathcal{I}(n).
+Thus, we describe two alternative split-based semantics. The non-
+interdigitation semantics gets \# (1,2,4,5), and the interdigitation
+semantics gets (1,3,4,5). This makes the point that each semantics
+results from different choices about which criteria to satisfy.
+
+\subsection{Split-based semantics for emph{incertae sedis} taxa}
+
+In order to apply supertree methods to rooted trees with
+\emph{incertae sedis} taxa, we first propose a semantics of
+\emph{incertae sedis }in terms of rooted splits. We define rooted
+splits by noting that each edge of a tree divides the tip taxa
+$\mathcal{L}$ into two groups: the include set $\mathcal{I}(e)$ which
+does not contain the root, and the exclude set $\mathcal{E}(e)$ which
+does contain the root. Such a split may be written \[
+\mathcal{I}(e)|\bullet\mathcal{E}(e). \] where the $\bullet$ indicates
+the root. If no taxa are \emph{incertae sedis}, then the exclude set
+for a node is just the total tip set minus the include set for the
+node: \begin{align*} \mathcal{E}(n) & =\mathcal{L}-\mathcal{I}(n).
 \end{align*}
 
-For a node $n$ on the tip-ward side of an edge $e$, we may also
-write $\mathcal{I}(n)$ for $\mathcal{I}(e)$, and $\mathcal{E}(n)$
-for $\mathcal{E}(e)$.
-
-For any two rooted splits $A=A_{1}|\bullet A_{2}$ and $B=B_{1}|\bullet B_{2}$
-, we say that $A$ \emph{displays }$B$ if $B_{1}\subseteq A_{1}$
+For a node $n$ on the tip-ward side of an edge $e$, we may also write
+$\mathcal{I}(n)$ for $\mathcal{I}(e)$, and $\mathcal{E}(n)$ for
+$\mathcal{E}(e)$.
+
+For any two rooted splits $A=A_{1}|\bullet A_{2}$ and $B=B_{1}|\bullet
+B_{2}$ , we say that $A$ \emph{displays }$B$ if $B_{1}\subseteq A_{1}$
 and $B_{2}\subseteq A_{2}$.
 
-\subsubsection{A semantics that disallows inter-digitating \emph{incertae sedis}
-siblings}
-
-An \emph{incertae sedis} taxon can be moved into any of the descendants
-of its siblings. We seek to represent this by constructing modified
-splits for each branch of the taxonomy tree. The include sets of these
-splits remain unchanged, but we construct reduced exclude sets in
-order to allow \emph{incertae sedis} taxa to intrude into their sibling
-taxa.
+\subsubsection{A semantics that disallows inter-digitating
+\emph{incertae sedis} siblings}
+
+An \emph{incertae sedis} taxon can be moved into any of the
+descendants of its siblings. We seek to represent this by constructing
+modified splits for each branch of the taxonomy tree. The include sets
+of these splits remain unchanged, but we construct reduced exclude
+sets in order to allow \emph{incertae sedis} taxa to intrude into
+their sibling taxa.
 
 The reduced exclude set $\mathcal{E}(n)$ for node $n$ should therefore
-not contain the descendants of its siblings which are marked \emph{incertae
-}sedis\emph{,} but should contain the descendants of its other siblings\emph{.
-}We additionally stipulate that it is not possible for \emph{incertae
-sedis} taxa to intrude into their \emph{incertae sedis} siblings.
-(This is not the only way to construct a split-based semantics for
-\emph{incertae sedis}, but it avoids certain complications, as noted
-in the discussion.) We thus define $\mathcal{I}(m,n)$ to indicate
-descendants of $m$ that are excluded from $n$. Thus:
-\begin{align*}
-\mathcal{I}(m,n) & =\begin{cases}
-\emptyset & \text{if }n\text{ is \emph{incertae sedis}\,and }m\text{ is not}\\
-\mathcal{I}(m) & \text{otherwise}
-\end{cases}
-\end{align*}
+not contain the descendants of its siblings which are marked
+\emph{incertae }sedis\emph{,} but should contain the descendants of
+its other siblings\emph{. }We additionally stipulate that it is not
+possible for \emph{incertae sedis} taxa to intrude into their
+\emph{incertae sedis} siblings. (This is not the only way to construct
+a split-based semantics for \emph{incertae sedis}, but it avoids
+certain complications, as noted in the discussion.) We thus define
+$\mathcal{I}(m,n)$ to indicate descendants of $m$ that are excluded
+from $n$. Thus: \begin{align*} \mathcal{I}(m,n) & =\begin{cases}
+\emptyset & \text{if }n\text{ is \emph{incertae sedis}\,and }m\text{
+is not}\\ \mathcal{I}(m) & \text{otherwise} \end{cases} \end{align*}
 
 Additionally, the exclude set $\mathcal{E}(n)$ should contain the
-children of all the ancestors of $n$, unless those children are \emph{incertae
-sedis}. To avoid traversing all ancestors separately for each node
-$n$, we note that the excluded children of any ancestors will be
-in the exclude set of the parent of $n$ already. Therefore, we can
-write the exclude set of a node $n$ in terms of the exclude set of
+children of all the ancestors of $n$, unless those children are
+\emph{incertae sedis}. To avoid traversing all ancestors separately
+for each node $n$, we note that the excluded children of any ancestors
+will be in the exclude set of the parent of $n$ already. Therefore, we
+can write the exclude set of a node $n$ in terms of the exclude set of
 its parent:
 
-\begin{align}
-\mathcal{E}(n) & =\mathcal{E}(parent(n))\cup\left[\mathcal{I}(m,n)\big|m\in siblings(n)\right].\label{eq:exclude-set-formula-1}
-\end{align}
-This formula allows us to compute exclude sets via a pre-order traversal
-on the taxonomy tree with exclude set for the root node defined to
-be $\emptyset$.
-
-\textbf{Problem:} how about the goal that we should be able to get
-the same outcome by moving a taxon up one level and marking it as
-incertae sedis? This might require that moving A next to B, and then
-moving B into A is allowed.
+\begin{align} \mathcal{E}(n) &
+=\mathcal{E}(parent(n))\cup\left[\mathcal{I}(m,n)\big|m\in
+siblings(n)\right].\label{eq:exclude-set-formula-1} \end{align} This
+formula allows us to compute exclude sets via a pre-order traversal on
+the taxonomy tree with exclude set for the root node defined to be
+$\emptyset$.
+
+\textbf{Problem:} how about the goal that we should be able to get the
+same outcome by moving a taxon up one level and marking it as incertae
+sedis? This might require that moving A next to B, and then moving B
+into A is allowed.
 
@@
 
-\subsubsection{An alternative semantics that allows inter-digitating incertae sedis
-siblings}
+\subsubsection{An alternative semantics that allows inter-digitating
+incertae sedis siblings}
 
@@
 
+\begin{align*} \mathcal{I^{\prime}}(n) & =\begin{cases} \emptyset &
+\text{if }n\in\mathcal{U}\\ \{n\} & \text{if }n\text{ is a leaf}\\{}
+[\mathcal{I}^{\prime}(c)|c\in children(n)] & \text{otherwise.}
+\end{cases} \end{align*} We may then define the taxa excluded from $n$
+as:
+
 \begin{align*} 
-\mathcal{I^{\prime}}(n) & =\begin{cases}
-\emptyset & \text{if }n\in\mathcal{U}\\
-\{n\} & \text{if }n\text{ is a leaf}\\{}
+\mathcal{E}(n) & =\mathcal{E}(parent(n))\cup\left[\mathcal{I}(m,n)\big|m\in siblings(n)\right].\label{eq:exclude-set-formula-1}
+\end{align*}
+
+This semantics satisfies criteria 1,3,4, and 5 above.
+
+\subsubsection{A third semantics}
+
+OK, the idea here is that we want to avoid inter-digitation.
+\emph{Incertae sedis} taxa thus exclude\emph{ incertae sedis}
+siblings. However, where the first semantics allows \emph{incertae
+sedis }taxa to intrude into \emph{incertae sedis} descendants of
+siblings, but not allow\emph{ incertae sedis} descendants of siblings
+to intrude into \emph{incertae sedis} taxa, here we do the reverse.
+
+OK, so the idea here is that an\emph{ incertae sedis} group excludes
+\begin{itemize} \item \emph{incertae sedis} siblings \item
+\emph{incertae sedis }descendants of ancestors \item but not\emph{
+incertae sedis} descendants of non-incertae-sedis siblings
+\end{itemize} It is also possible to allow placing $A$ as a sibling of
+$B$ and then placing $B$ within $A$ if we disallow placing $A$ into
+incertae sedis descendants of its siblings. We may then allow placing
+those descendants within $A$.
+
+\begin{align*} \mathcal{I^{\prime}}(n) & =\begin{cases} \emptyset &
+\text{if }n\in\mathcal{U}\\ \{n\} & \text{if }n\text{ is a leaf}\\{}
 [\mathcal{I}^{\prime}(c)|c\in children(n)] & \text{otherwise.}
-\end{cases}
-\end{align*}
+\end{cases} \end{align*}
+
+We use $\mathcal{I}(m,n)$ to indicate the descendants of $m$ that are
+excluded from $n$. \begin{align*} \mathcal{I}(m,n) & =\begin{cases}
+\emptyset & \text{if }n\text{ is \emph{incertae sedis}\,and }m\text{
+is not}\\ \mathcal{I}(m) & \text{if }n\text{ and }m\text{ are both
+}incertae\,sedis\\ \mathcal{I}^{\prime}(m) & \text{otherwise.}
+\end{cases} \end{align*}
+
 We may then define the taxa excluded from $n$ as:
-
-\begin{align*}
-\mathcal{E}(n) & =\mathcal{E}(parent(n))\cup[\mathcal{I}^{\prime}(s)|s\in siblings(n)].
-\end{align*}
-
-This semantics satisfies criteria 1,3,4, and 5 above.
-
-\subsubsection{A third semantics}
-
-OK, the idea here is that we want to avoid inter-digitation. \emph{Incertae
-sedis} taxa thus exclude\emph{ incertae sedis} siblings. However,
-where the first semantics allows \emph{incertae sedis }taxa to intrude
-into \emph{incertae sedis} descendants of siblings, but not allow\emph{
-incertae sedis} descendants of siblings to intrude into \emph{incertae
-sedis} taxa, here we do the reverse.
-
-OK, so the idea here is that an\emph{ incertae sedis} group excludes
-\begin{itemize}
-\item \emph{incertae sedis} siblings
-\item \emph{incertae sedis }descendants of ancestors
-\item but not\emph{ incertae sedis} descendants of non-incertae-sedis siblings
-\end{itemize}
-It is also possible to allow placing $A$ as a sibling of $B$ and
-then placing $B$ within $A$ if we disallow placing $A$ into incertae
-sedis descendants of its siblings. We may then allow placing those
-descendants within $A$.
-
-\begin{align*}
-\mathcal{I^{\prime}}(n) & =\begin{cases}
-\emptyset & \text{if }n\in\mathcal{U}\\
-\{n\} & \text{if }n\text{ is a leaf}\\{}
-[\mathcal{I}^{\prime}(c)|c\in children(n)] & \text{otherwise.}
-\end{cases}
-\end{align*}
-
-We use $\mathcal{I}(m,n)$ to indicate the descendants of $m$ that
-are excluded from $n$.
-\begin{align*}
-\mathcal{I}(m,n) & =\begin{cases}
-\emptyset & \text{if }n\text{ is \emph{incertae sedis}\,and }m\text{ is not}\\
-\mathcal{I}(m) & \text{if }n\text{ and }m\text{ are both }incertae\,sedis\\
-\mathcal{I}^{\prime}(m) & \text{otherwise.}
-\end{cases}
-\end{align*}
-
-We may then define the taxa excluded from $n$ as:
-
 \begin{align*}
@@
 \end{align*}
-I think this works... but need to double-check.
+I think this works... but
+need to double-check.
 
@@
 
-After constructing a supertree, tip nodes already have names in \emph{$\mathcal{L}$}.
-However, we still need to assign higher taxon names to internal supertree
-nodes based on the taxonomy tree in the problem. Each taxon name $n$
-corresponds to a split $S(n)=S(n)_{1}|\bullet S(n)_{2}$ on the corresponding
-branch of the taxonomy tree. Without \emph{incertae sedis}, such splits
-are always of the form $S(n)_{1}|\bullet\mathcal{L}-S(n)_{1}$, but
-with \emph{incertae sedis} taxa $S_{2}(n)$ may be smaller than $\mathcal{L}-S(n)_{1}$.
+After constructing a supertree, tip nodes already have names in
+\emph{$\mathcal{L}$}. However, we still need to assign higher taxon
+names to internal supertree nodes based on the taxonomy tree in the
+problem. Each taxon name $n$ corresponds to a split
+$S(n)=S(n)_{1}|\bullet S(n)_{2}$ on the corresponding branch of the
+taxonomy tree. Without \emph{incertae sedis}, such splits are always
+of the form $S(n)_{1}|\bullet\mathcal{L}-S(n)_{1}$, but with
+\emph{incertae sedis} taxa $S_{2}(n)$ may be smaller than
+$\mathcal{L}-S(n)_{1}$.
 
 Without \emph{incertae sedis}, each name applies to at most one node,
-and each node can take at most one name, with the exception of monotypic
-taxa. Thus, we may simply search the solution tree for a node that
-has the same cluster $S(n)_{1}$ and apply the name $n$ to that node.
+and each node can take at most one name, with the exception of
+monotypic taxa. Thus, we may simply search the solution tree for a
+node that has the same cluster $S(n)_{1}$ and apply the name $n$ to
+that node.
 
 However, in the \emph{incertae sedis} framework, it is possible for
-one name to apply to multiple nodes. For example, in Figure \ref{fig:One-name-can}b,
-the name $A$ can apply to nodes $x$ and $y$. Here the name $A$
-corresponds to the split $A1\,A2\,|\bullet B\,C$, leaving out $D$
-since it is \emph{incertae sedis}. The two nodes $x$ and $y$ display
-the splits $A1\,A2\,|\bullet B\,C\,D$ and $A1\,A2\,D|\bullet B\,C$
-respectively, and both of these splits display the split $A1\,A2\,|\bullet B\,C$,
-so the name $A$ can apply to both $x$ and $y$. This cannot happen
-without\emph{ incertae sedis} taxa except at monotypic nodes. When
-faced with a choice about where to place a name, our solution is to
-find the most tip-ward node where the name can apply and attach the
-name to this node. This corresponds to a conservative choice about
-whether \emph{incertae sedis} taxa should be added to an existing
-taxon.
+one name to apply to multiple nodes. For example, in Figure \ref{fig
+:One-name-can}b, the name $A$ can apply to nodes $x$ and $y$. Here the
+name $A$ corresponds to the split $A1\,A2\,|\bullet B\,C$, leaving out
+$D$ since it is \emph{incertae sedis}. The two nodes $x$ and $y$
+display the splits $A1\,A2\,|\bullet B\,C\,D$ and $A1\,A2\,D|\bullet
+B\,C$ respectively, and both of these splits display the split
+$A1\,A2\,|\bullet B\,C$, so the name $A$ can apply to both $x$ and
+$y$. This cannot happen without\emph{ incertae sedis} taxa except at
+monotypic nodes. When faced with a choice about where to place a name,
+our solution is to find the most tip-ward node where the name can
+apply and attach the name to this node. This corresponds to a
+conservative choice about whether \emph{incertae sedis} taxa should be
+added to an existing taxon.
 \begin{figure}
-\hfill{}\subfloat[Taxonomy tree]{\includegraphics[scale=0.5]{Figures/name1nodes2/tax-shaded}
-
-}\hfill{}\subfloat[Solution tree]{\includegraphics[scale=0.5]{Figures/name1nodes2/solution-shaded}
-
-}\hfill{}
-
-\caption{\label{fig:One-name-can}One name can be consistent with multiple
-nodes.}
+\hfill{}\subfloat[Taxonomy tree]{\includegraphics[scale=0.5]{Figures/name1nodes2/tax-shaded}}
+\hfill{}\subfloat[Solution tree]{\includegraphics[scale=0.5]{Figures/name1nodes2/solution-shaded}}
+\hfill{}
+\caption{\label{fig:One-name-can}One name can be consistent with multiple nodes.}
 \end{figure}
 
-It is also possible for multiple names to apply to a single node.
-For example, in Figure \ref{fig:Multiple-names-can}a, the taxon $A$
-corresponds to the split $B1\,B2\,C\,|\bullet\,Y$, and its child
-taxon $B$ corresponds to the split $B1\,B2\,|\bullet\,Y$. The edge
-leading to $A$ is consistent with the split for $B$, but the name
-$B$ is applied to the node with the smallest include group. However,
-in the solution tree (Fig. \ref{fig:Multiple-names-can}b), there
-is is only node for both names $A$ and $B$ to apply to. In this
-case, we solve this problem by introducing a monotypic parent, and
-applying $A$ to the newly created parent node (Fig. \ref{fig:Multiple-names-can}c).
+It is also possible for multiple names to apply to a single node. For
+example, in Figure \ref{fig:Multiple-names-can}a, the taxon $A$
+corresponds to the split $B1\,B2\,C\,|\bullet\,Y$, and its child taxon
+$B$ corresponds to the split $B1\,B2\,|\bullet\,Y$. The edge leading
+to $A$ is consistent with the split for $B$, but the name $B$ is
+applied to the node with the smallest include group. However, in the
+solution tree (Fig. \ref{fig:Multiple-names-can}b), there is is only
+node for both names $A$ and $B$ to apply to. In this case, we solve
+this problem by introducing a monotypic parent, and applying $A$ to
+the newly created parent node (Fig. \ref{fig:Multiple-names-can}c).
 \begin{figure}
-\hfill{}\subfloat[Taxonomy tree]{\includegraphics[scale=0.5]{Figures/names2nodes1/tax-shaded}
-
-}\hfill{}\subfloat[Solution tree]{\includegraphics[scale=0.5]{Figures/names2nodes1/synth-shaded}
-
-}\hfill{}\subfloat[Solution tree]{\includegraphics[scale=0.5]{Figures/names2nodes1/solution2-shaded}
-
-}\hfill{}
-
-\caption{\label{fig:Multiple-names-can}Multiple names can apply to a single
-node.}
-
+\hfill{}\subfloat[Taxonomy tree]{\includegraphics[scale=0.5]{Figures/names2nodes1/tax-shaded}}
+\hfill{}subfloat[Solution tree]{\includegraphics[scale=0.5]{Figures/names2nodes1/synth-shaded}}
+\hfill{}subfloat[Solution tree]{\includegraphics[scale=0.5]{Figures/names2nodes1/solution2-shaded}}
+\hfill{}
+\caption{\label{fig:Multiple-names-can}Multiple names can apply to a single node.}
 \end{figure}
@@
 
-\subsection{\label{subsec:Placement}Placement}
+\subsection{label{subsec:Placement}Placement}
 
@@
 like to interpret the position of these names in terms of placing
-\emph{incertae sedis }taxa in a revised taxonomy. This would allow
-us to interpret the synthesis tree as saying that phylogenetic information
-has (for example) placed genus A within family B, or perhaps outside
-of all named families. The simplest approach to placement involves
-noting whenever a taxon $B$ is a descendant of a taxon $A$ on the
-named synthesis tree but not the taxonomy tree. For example, in Figure
-\ref{fig:Handling-incertae-sedis}, taxon $C$ is a descendant of
-taxon $A$ in the synthetic tree (Fig. \ref{fig:Handling-incertae-sedis}d),
-but not the taxonomy (Fig. \ref{fig:Handling-incertae-sedis}b). Thus,
-we could say that the synthetic tree places $C$ within $A$. Furthermore,
-the synthetic tree additionally places $C$ within $AB$.
-
-A slightly more general case is shown in Figure \ref{fig:Broken-incertae-sedis}.
-Here taxon $C1$ is an descendant of $A$ in the synthesis tree, but
-not the taxonomy. Thus $C1$ is placed within $A$. We note that the
-most recent common ancestor (MRCA) of $A$ and $C1$ in the taxonomy
-is the root node, and the path to the MRCA from $A$ is $A\to AB\to root$,
-while the path to the MRCA from $C1$ is $C1\to C\to root$. Here
-the additional node $AB$ on the path to the MRCA from $A$ indicates
-that $C1$ is placed within $AB$ as well as $A$. In contrast, the
-additional node $C$ on the path to the MRCA indicates that $C1$
-is a broken \emph{incertae sedis} taxon. Thus we may consider an algorithm
-that, for each taxon $B$ on the synthesis tree, finds the closest
-ancestral taxon $A$, and checks if $A$ is an ancestor of $B$ on
-the taxonomy. If not, we compute the paths $A\to A_{1}\to\ldots\to MRCA(A,B)$
-and $B\to B_{1}\to\ldots\to MRCA(A,B)$.
-
-However, this assumes that \emph{incertae sedis} taxa are always placed
-within their siblings, or descendants of their siblings. As such,
-it does not handle cases such as Figure \ref{fig:Placing-one-incertae},
-in which $I$ is first placed as sibling to $AI$, and then $AI$
-is placed within $I$. Note that $I$ is not a sibling of $AI$ or
-a descendant of a sibling. Furthermore, the MRCA of $I$ and $AI$
-is the root, and the path from $AI$ to the root includes $A$, but
-$A$ is not broken. We suggest that this conundrum may be resolved
-by first processing the placement of $I$ within $A$ by modifying
-the taxonomy to reflect this placement\footnote{Does this actually work? Some placements break taxa.}.
-When we consider the placement of $AI$ within $I$ on this modified
-taxonomy, the MRCA of $I$ and $AI$ is now $A$ instead of the root,
-and the correct interpretation results. We suggest that by walking
-the tree in a pre-order fashion and processing shallowing placements
-first, the previous approach of finding the MRCA and labelling intermediate
-nodes as either additional placements or broken incertae sedis (or
-incertae\_sedis\_inherited) taxa holds. Such an approach should also
-handle cases containing nested incertae sedis taxa, as in Figure \ref{fig:Placing-one-incertae}.%
+\emph{incertae sedis }taxa in a revised taxonomy. This would allow us
+to interpret the synthesis tree as saying that phylogenetic
+information has (for example) placed genus A within family B, or
+perhaps outside of all named families. The simplest approach to
+placement involves noting whenever a taxon $B$ is a descendant of a
+taxon $A$ on the named synthesis tree but not the taxonomy tree. For
+example, in Figure \ref{fig:Handling-incertae-sedis}, taxon $C$ is a
+descendant of taxon $A$ in the synthetic tree (Fig. \ref{fig:Handling-
+incertae-sedis}d), but not the taxonomy (Fig. \ref{fig:Handling-
+incertae-sedis}b). Thus, we could say that the synthetic tree places
+$C$ within $A$. Furthermore, the synthetic tree additionally places
+$C$ within $AB$.
+
+A slightly more general case is shown in Figure \ref{fig:Broken-
+incertae-sedis}. Here taxon $C1$ is an descendant of $A$ in the
+synthesis tree, but not the taxonomy. Thus $C1$ is placed within $A$.
+We note that the most recent common ancestor (MRCA) of $A$ and $C1$ in
+the taxonomy is the root node, and the path to the MRCA from $A$ is
+$A\to AB\to root$, while the path to the MRCA from $C1$ is $C1\to C\to
+root$. Here the additional node $AB$ on the path to the MRCA from $A$
+indicates that $C1$ is placed within $AB$ as well as $A$. In contrast,
+the additional node $C$ on the path to the MRCA indicates that $C1$ is
+a broken \emph{incertae sedis} taxon. Thus we may consider an
+algorithm that, for each taxon $B$ on the synthesis tree, finds the
+closest ancestral taxon $A$, and checks if $A$ is an ancestor of $B$
+on the taxonomy. If not, we compute the paths $A\to A_{1}\to\ldots\to
+MRCA(A,B)$ and $B\to B_{1}\to\ldots\to MRCA(A,B)$.
+
+However, this assumes that \emph{incertae sedis} taxa are always
+placed within their siblings, or descendants of their siblings. As
+such, it does not handle cases such as Figure \ref{fig:Placing-one-
+incertae}, in which $I$ is first placed as sibling to $AI$, and then
+$AI$ is placed within $I$. Note that $I$ is not a sibling of $AI$ or a
+descendant of a sibling. Furthermore, the MRCA of $I$ and $AI$ is the
+root, and the path from $AI$ to the root includes $A$, but $A$ is not
+broken. We suggest that this conundrum may be resolved by first
+processing the placement of $I$ within $A$ by modifying the taxonomy
+to reflect this placement\footnote{Does this actually work? Some
+placements break taxa.}. When we consider the placement of $AI$ within
+$I$ on this modified taxonomy, the MRCA of $I$ and $AI$ is now $A$
+instead of the root, and the correct interpretation results. We
+suggest that by walking the tree in a pre-order fashion and processing
+shallowing placements first, the previous approach of finding the MRCA
+and labelling intermediate nodes as either additional placements or
+broken incertae sedis (or incertae\_sedis\_inherited) taxa holds. Such
+an approach should also handle cases containing nested incertae sedis
+taxa, as in Figure \ref{fig:Placing-one-incertae}.
+
 \begin{comment}
-One question is whether this approach is able to handle nested incertae
-sedis taxa.
+One question is whether this approach is able to handle nested
+incertae sedis taxa.
 
 Placement of incertae sedis taxa by input trees is unfortunately not
-quite as simple as finding a single location where an I.S. taxon should
-attach. For example, when an incertae sedis taxon is broken, its children
-need to be ``placed'' separately.
-
-Each input tree can relate to an \emph{incertae sedis} taxon $(A,B,C)D$
-in a number of ways
+quite as simple as finding a single location where an I.S. taxon
+should attach. For example, when an incertae sedis taxon is broken,
+its children need to be ``placed'' separately.
+
+Each input tree can relate to an \emph{incertae sedis} taxon
+$(A,B,C)D$ in a number of ways
 \begin{itemize}
-\item it could resolve $A$, $B$, $C$, or $D$.
-\item it could place $D$ on a degree-2 (=out-degree-1) node that bisects
-a branch
-\item it could place a descendant taxon of $D$ 
-\item it could place a descendant taxon of $D$ in a \emph{different place}
-than another input tree.
-\item it could place children of $D$ in multiple places, thus conflicting
-with the branch. If the \emph{incertae sedis} taxon $(A,B,C)$ is
-broken, then $A$, $B$ and $C$ become \emph{incertae sedis} clades
-in their own right, that may attach separately, except that they .
-This is because none of $A$, $B$, or $C$ is in the exclude set
-of the siblings of $D$.
+    \item it could resolve
+$A$, $B$, $C$, or $D$.
+    \item it could place $D$ on a degree-2 (=out-
+degree-1) node that bisects a branch 
+    \item it could place a descendant
+taxon of $D$ 
+    \item it could place a descendant taxon of $D$ in a
+\emph{different place} than another input tree.
+    \item it could place
+children of $D$ in multiple places, thus conflicting with the branch.
+If the \emph{incertae sedis} taxon $(A,B,C)$ is broken, then $A$, $B$
+and $C$ become \emph{incertae sedis} clades in their own right, that
+may attach separately, except that they . This is because none of $A$,
+$B$, or $C$ is in the exclude set of the siblings of $D$.
 \end{itemize}
@@
 uncertain taxon placements using information from phylogenies. Since
-phylogenies are ranked higher than taxonomy, they may place members
-of incertae sedis clades as they wish even when we do not take into
+phylogenies are ranked higher than taxonomy, they may place members of
+incertae sedis clades as they wish even when we do not take into
 account the semantics of incertae sedis taxa. However, without the
-incertae sedis semantics described above, placing a taxon $A$ within
-a taxon $B$ results in conflict with taxon $B$ in the taxonomy.
-This has three main effects:
-\begin{enumerate}
-\item The edge leading to clade $B$ is lost.
-\item Any taxonomy-only children of $B$ are placed at the MRCA of the phylogenetically
-informed taxa, instead of at node $B$.
-\item The name for taxon $B$ is lost.
-\item We do not record $A$ as being placed within $B$.
-\end{enumerate}
-In order to avoid a situation where input phylogenies conflict with
-a large number of taxa when \emph{incertae sedis} taxa are placed
-within them, we have previously filtered \emph{incertae sedis} taxa
-from the taxonomy when constructing all prior synthesis trees.
+incertae sedis semantics described above, placing a taxon $A$ within a
+taxon $B$ results in conflict with taxon $B$ in the taxonomy. This has
+three main effects: \begin{enumerate} \item The edge leading to clade
+$B$ is lost. \item Any taxonomy-only children of $B$ are placed at the
+MRCA of the phylogenetically informed taxa, instead of at node $B$.
+\item The name for taxon $B$ is lost. \item We do not record $A$ as
+being placed within $B$. \end{enumerate} In order to avoid a situation
+where input phylogenies conflict with a large number of taxa when
+\emph{incertae sedis} taxa are placed within them, we have previously
+filtered \emph{incertae sedis} taxa from the taxonomy when
+constructing all prior synthesis trees.
 
@@
 the taxon $B$ at that node is a broken taxon. Broken taxa have two
-main effects, both of which are negative. First, the name $B$ of
-the broken taxon is removed from the synthesis tree. Second, the conflicting
-edge for taxon $B$ is not included in the synthesis tree. This means
-that any children of $B$ that are taxonomy-only will not be placed
-with the children of $B$ that are mentioned in input phylogenies.
-Instead the taxonomy-only children of a broken taxon move towards
-the root of the synthesis tree and attach at the first higher-ranked
-taxon that is an ancestor of $B$ but is not broken.
+main effects, both of which are negative. First, the name $B$ of the
+broken taxon is removed from the synthesis tree. Second, the
+conflicting edge for taxon $B$ is not included in the synthesis tree.
+This means that any children of $B$ that are taxonomy-only will not be
+placed with the children of $B$ that are mentioned in input
+phylogenies. Instead the taxonomy-only children of a broken taxon move
+towards the root of the synthesis tree and attach at the first higher-
+ranked taxon that is an ancestor of $B$ but is not broken.
 
@@
 exclusion of all non-included taxa, but only some non-included taxa,
-as described in section \ref{sec:Semantics-of-incertae}. As a result
+as described in section \ref{sec:Semantics-of- incertae}. As a result
 of this change in semantics, placing an IS taxon $A$ within a sister
-taxon $B$ no longer results in conflict with $B$. This allows us
-to retain the split for $B$ within the synthesis tree, so that taxonomy-only
-children of $B$ are correctly grouped with their siblings that are
-referenced by the input trees. We may then retain the name for the
+taxon $B$ no longer results in conflict with $B$. This allows us to
+retain the split for $B$ within the synthesis tree, so that taxonomy-
+only children of $B$ are correctly grouped with their siblings that
+are referenced by the input trees. We may then retain the name for the
 no-longer-broken taxon. Finally, we are then able to stop filtering
 \emph{incertae sedis} taxa, so that they appear in the synthesis tree.
-Thus, the synthesis tree is able to represent substantially more species,
-without suffering the loss of taxa and the loss of structure.
+Thus, the synthesis tree is able to represent substantially more
+species, without suffering the loss of taxa and the loss of structure.
 
@@
 
-When only tip nodes on the taxonomy are \emph{incertae sedis}, we
-have a very simple form of the \emph{incertae sedis} supertree problem.
+When only tip nodes on the taxonomy are \emph{incertae sedis}, we have
+a very simple form of the \emph{incertae sedis} supertree problem.
 However, in practice entire clades may be \emph{incertae sedis}, and
-so more complex issue arise. In this section we consider a number
-of cases that must be handled when attempting to place \emph{incertae
+so more complex issue arise. In this section we consider a number of
+cases that must be handled when attempting to place \emph{incertae
 sedis} taxa.
@@
 
-{[}\textbf{BDR:} \emph{maybe we should merge this with Figure. \ref{fig:Broken-incertae-sedis}.
-Although the point of figure \ref{fig:placement-example1} is to talk
-about sub-problem decomposition as well as what the result looks like.}{]}
-
-The addition of \emph{incertae sedis} taxa allows new types of conflict
-between input trees. For example, different input trees might place
-an incertae sedis taxon in conflicting locations. This is illustrated
-in Figure \ref{fig:placement-example1}, where the IS taxon (ott7,ott6)ott10
-is placed as sister to ott1 by phylogeny $\tau_{1}$ and as sister
-ott3 by phylogeny $\tau_{2}$.
-
-When this happens, the placement of the IS taxon is not influenced
-by its being marked IS on the taxonomy. Thus, in Example 1, the higher
+{[}\textbf{BDR:} \emph{maybe we should merge this with Figure.
+\ref{fig:Broken- incertae-sedis}. Although the point of figure
+\ref{fig:placement-example1} is to talk about sub-problem
+decomposition as well as what the result looks like.}{]}
+
+The addition of \emph{incertae sedis} taxa allows new types of
+conflict between input trees. For example, different input trees might
+place an incertae sedis taxon in conflicting locations. This is
+illustrated in Figure \ref{fig :placement-example1}, where the IS
+taxon (ott7,ott6)ott10 is placed as sister to ott1 by phylogeny
+$\tau_{1}$ and as sister ott3 by phylogeny $\tau_{2}$.
+
+When this happens, the placement of the IS taxon is not influenced by
+its being marked IS on the taxonomy. Thus, in Example 1, the higher
 ranked tree $\tau_{1}$ will be reflected in the synthesis tree, ott10
-will be placed as sister to ott1. In contrast, the conflicting placement
-in $\tau_{2}$ will not be reflected in the synth tree.
-
-All this would occur in the previous version of propinquity. Where
-the updated version differs that (a) the names ott9 and ott8 are retained
+will be placed as sister to ott1. In contrast, the conflicting
+placement in $\tau_{2}$ will not be reflected in the synth tree.
+
+All this would occur in the previous version of propinquity. Where the
+updated version differs that (a) the names ott9 and ott8 are retained
 instead of being dropped. (b) as a result of not breaking ott8 and
@@
 
-\subsection{Case 1: An \emph{incertae sedis} clade}
+\subsection{Case 1: An emph{incertae sedis} clade}
 
@@
 two cases. In the first case, the clade may be placed within a sister
-clade intact (Figure \ref{fig:Handling-incertae-sedis}). However,
-if the clade is not monophyletic in the synthesis tree, then we allow
-the members of the clade to be placed separately (Figure \ref{fig:Broken-incertae-sedis}).
+clade intact (Figure \ref{fig:Handling-incertae-sedis}). However, if
+the clade is not monophyletic in the synthesis tree, then we allow the
+members of the clade to be placed separately (Figure \ref{fig:Broken-
+incertae-sedis}).
 
@@
 \end{figure}
-
 \begin{figure}
@@
 sedis clade should be placed, neither placement is rejected. Instead,
-the taxon $C$ is broken, the name $C$ disappears, and $C3$ floats
-to the top level. Algorithmic, alternatively placements of children
-of an incertae sedis clade lead to contesting of all edges along with
-taxa might be placed, creating a large, merged sub-problem, as in
-Figure \ref{fig:placement-example1}.
+the taxon $C$ is broken, the name $C$ disappears, and $C3$ floats to
+the top level. Algorithmic, alternatively placements of children of an
+incertae sedis clade lead to contesting of all edges along with taxa
+might be placed, creating a large, merged sub-problem, as in Figure
+\ref{fig:placement-example1}.
 
@@
 floats to the top level. Furthermore, since taxon $C$ is a broken
-\emph{incertae sedis }taxon, all of its children are effectively incertae
-sedis independently, with the exception that they cannot be placed
-within each other. Therefore, the names $A$ and $AB$ are not lost,
-since the taxa placed within them are \emph{incertae sedis} names.
+\emph{incertae sedis }taxon, all of its children are effectively
+incertae sedis independently, with the exception that they cannot be
+placed within each other. Therefore, the names $A$ and $AB$ are not
+lost, since the taxa placed within them are \emph{incertae sedis}
+names.
 
 The situation here would be different if the monophyly of $C$ was
-supported by an high-ranked input phylogeny. In that case, instead
-of breaking $C$, we would choose the placement in the highest-ranked
+supported by an high-ranked input phylogeny. In that case, instead of
+breaking $C$, we would choose the placement in the highest-ranked
 tree.
 
-\subsection{Case 2: A nested \emph{incertae sedis} clade}
-
-When \emph{incertae sedis} clades are allowed, it is possible for
-an \emph{incertae sedis} taxon to be nested within another \emph{incertae
-sedis} taxon. In such a case we must allow the higher-ranked taxon
-to be placed within a sibling, while preserving the right of the lower
-ranked taxon to be placed within a lower-level sibling (Figure \ref{fig:Nested-incertae-sedis}).
+\subsection{Case 2: A nested emph{incertae sedis} clade}
+
+When \emph{incertae sedis} clades are allowed, it is possible for an
+\emph{incertae sedis} taxon to be nested within another \emph{incertae
+sedis} taxon. In such a case we must allow the higher-ranked taxon to
+be placed within a sibling, while preserving the right of the lower
+ranked taxon to be placed within a lower-level sibling (Figure
+\ref{fig:Nested-incertae-sedis}).
 
@@
 
-\subsection{Case 3: Placement of one incertae sedis taxon within another.}
+\subsection{Case 3: Placement of one incertae sedis taxon within
+another.}
 
@@
 
-\emph{}
 \begin{figure}
@@
 
-\section{Handling \emph{incertae sedis} taxa in the propinquity pipeline}
-
-In order to handle \emph{incertae sedis} taxa within propinquity,
-we must modify some of the stages of the propinquity pipeline. Subproblem
+\section{Handling emph{incertae sedis} taxa in the propinquity
+pipeline}
+
+In order to handle \emph{incertae sedis} taxa within propinquity, we
+must modify some of the stages of the propinquity pipeline. Subproblem
 decomposition must place \emph{incertae sedis} taxa in the correct
@@
 \emph{incertae sedis} taxa when solving subproblems, and correctly
-name taxa that have been modified by having \emph{incertae sedis}
-taxa place inside them. The unpruner must be aware of \emph{incertae
-sedis} taxa. Annotations of the tree must be aware of \emph{incertae
-sedis} taxa so that it does not consider taxa broken when they have
-an incertae sedis taxon placed inside them.
+name taxa that have been modified by having \emph{incertae sedis} taxa
+place inside them. The unpruner must be aware of \emph{incertae sedis}
+taxa. Annotations of the tree must be aware of \emph{incertae sedis}
+taxa so that it does not consider taxa broken when they have an
+incertae sedis taxon placed inside them.
 
@@
 tend to have a very large number of incertae sedis children, making
-browsing in the tree viewer difficult. This can happen when, for example,
-fossils or other hard-to-place taxa get classified only to the level
-of these well-known nodes and no further. This leads to a situation
-where well-known taxa serve as a dumping ground for unplaced taxa.
-
-Our current approach to this problem is to perform a second round
-of pruning, or ``cleaning'', during the exemplification step. Incertae
+browsing in the tree viewer difficult. This can happen when, for
+example, fossils or other hard-to- place taxa get classified only to
+the level of these well-known nodes and no further. This leads to a
+situation where well-known taxa serve as a dumping ground for unplaced
+taxa.
+
+Our current approach to this problem is to perform a second round of
+pruning, or ``cleaning'', during the exemplification step. Incertae
 sedis taxa are pruned at this stage if they do not occur in any input
-trees. We thus generate a second ``cleaned taxonomy'' that has undergone
-this further round of cleaning. This approach improves on the previous
-approach in that \emph{incertae sedis} taxa in input trees are no
-longer pruned. This approach also removes tons of \emph{incertae sedis}
-children from nodes like ``Fungi'', where a lot of unplaced fossils
-with few observable characters have been dumped.
-
-However, this approach has the negative effect of pruning some incertae
-sedis taxa that need not be pruned. For example, suppose \emph{incertae
-sedis} taxon $A$ contains 5 children, of which only 1 child $A_{1}$
-occurs in an input tree. If the taxon $A$ is not broken, then it
-should be possible to attach the other 4 members of $A$ next to $A_{1}$,
-without cluttering up the synthesis tree. Such taxa have been successfully
-placed even though they are not in any input tree. This can only be
-discovered after synthesis is complete, though.
-
-We should be able to to figure out which incertae sedis taxa are placed
-(see section \ref{subsec:Placement}), and then mark the other ones
-incertae sedis in the \emph{synthesis tree}. We could then filter
+trees. We thus generate a second ``cleaned taxonomy'' that has
+undergone this further round of cleaning. This approach improves on
+the previous approach in that \emph{incertae sedis} taxa in input
+trees are no longer pruned. This approach also removes tons of
+\emph{incertae sedis} children from nodes like ``Fungi'', where a lot
+of unplaced fossils with few observable characters have been dumped.
+
+However, this approach has the negative effect of pruning some
+incertae sedis taxa that need not be pruned. For example, suppose
+\emph{incertae sedis} taxon $A$ contains 5 children, of which only 1
+child $A_{1}$ occurs in an input tree. If the taxon $A$ is not broken,
+then it should be possible to attach the other 4 members of $A$ next
+to $A_{1}$, without cluttering up the synthesis tree. Such taxa have
+been successfully placed even though they are not in any input tree.
+This can only be discovered after synthesis is complete, though.
+
+We should be able to to figure out which incertae sedis taxa are
+placed (see section \ref{subsec:Placement}), and then mark the other
+ones incertae sedis in the \emph{synthesis tree}. We could then filter
 out the taxa that \emph{remain }incertae sedis (or incertae\_sedis
@@
 
-The presence of \emph{incertae sedis }taxa poses a problem to sub-problem
-decomposition, since taxonomy edges no longer completely separate
-subproblems. Instead, \emph{incertae sedis} taxa may attach on either
-side of a taxonomy edge. We seek to place \emph{incertae sedis} taxa
-into subproblems in such a way that the subproblem solver can perform
-the placement inside the subproblem. This approach postpones handling
-of conflict in \emph{incertae sedis} taxa to the subproblem solver,
-where the problem is well formulated in terms of splits. However,
-it does have the effect of creating larger subproblems.
+The presence of \emph{incertae sedis }taxa poses a problem to sub-
+problem decomposition, since taxonomy edges no longer completely
+separate subproblems. Instead, \emph{incertae sedis} taxa may attach
+on either side of a taxonomy edge. We seek to place \emph{incertae
+sedis} taxa into subproblems in such a way that the subproblem solver
+can perform the placement inside the subproblem. This approach
+postpones handling of conflict in \emph{incertae sedis} taxa to the
+subproblem solver, where the problem is well formulated in terms of
+splits. However, it does have the effect of creating larger
+subproblems.
 
@@
 taxa by different input trees. Thus, if one input tree places the
-\emph{incertae sedis} taxon $X$ in $((X)B)A$ and another places
-$X$ in $((X)C)A$ then we must mark both edges $B$ and $C$ as contested
-edges, even if these edges would \emph{not} be contested were taxon
-$X$ to be removed. This results in a new way to contest edges that
-involves the interaction of two input trees, and not just the interaction
-of each input tree with the taxonomy.
-
-We choose to solve these problems by merging any subproblems that
-an \emph{incertae sedis} taxon might be placed in. The simplest way
-to achieve this is simply to regard any taxon that has an \emph{incertae
+\emph{incertae sedis} taxon $X$ in $((X)B)A$ and another places $X$ in
+$((X)C)A$ then we must mark both edges $B$ and $C$ as contested edges,
+even if these edges would \emph{not} be contested were taxon $X$ to be
+removed. This results in a new way to contest edges that involves the
+interaction of two input trees, and not just the interaction of each
+input tree with the taxonomy.
+
+We choose to solve these problems by merging any subproblems that an
+\emph{incertae sedis} taxon might be placed in. The simplest way to
+achieve this is simply to regard any taxon that has an \emph{incertae
 sedis }taxon placed within it as contested. This results in marking
@@
 subproblem decomposer. One downside of this approach is that, if we
-have $((X)B)A)$ in one input tree, and $X$ is mentioned nowhere
-else, then by marking $B$ as contested, we are creating unnecessarily
-large subproblems. We could instead placed $X$ in $B$ and avoid
-contesting the edge $B$. However, this approach is more complex and
-does not seem necessary in practice.
+have $((X)B)A)$ in one input tree, and $X$ is mentioned nowhere else,
+then by marking $B$ as contested, we are creating unnecessarily large
+subproblems. We could instead placed $X$ in $B$ and avoid contesting
+the edge $B$. However, this approach is more complex and does not seem
+necessary in practice.
 
@@
 \end{figure}
- For example, in Figure \ref{fig:placement-example1}, input tree
-$\tau_{1}$ contests ott9 and input tree $\tau_{2}$ contests ott9
-and ott8. Thus ott1, ott2, ott3, ott6, and ott7 end up in the same
-sub-problem.
+ For example, in Figure
+\ref{fig:placement-example1}, input tree $\tau_{1}$ contests ott9 and
+input tree $\tau_{2}$ contests ott9 and ott8. Thus ott1, ott2, ott3,
+ott6, and ott7 end up in the same sub-problem.
 
@@
 we must apply taxon names from the taxonomy tree to the sub-problem
-solution tree. The solution tree is considered to a fixed tree and
-not to have any \emph{incertae sedis} nodes, or any other forms of
+solution tree. The solution tree is considered to a fixed tree and not
+to have any \emph{incertae sedis} nodes, or any other forms of
 uncertainty.
@@
 
-Currently, we read the \emph{incertae sedis} information as a list
-of OTT ids for \emph{incertae sedis} taxa. This does not require adding
+Currently, we read the \emph{incertae sedis} information as a list of
+OTT ids for \emph{incertae sedis} taxa. This does not require adding
 further annotations to the node names. Only taxonomy nodes can be
 \emph{incertae sedis} at the moment, and only the taxonomy tree for
-the subproblem contains OTT ids for internal nodes. Therefore we handle
-\emph{incertae sedis} information by constructing modified split sets
-for the lowest-ranked tree when the list of \emph{incertae sedis}
-nodes is not empty.
-
-\subsubsection{Exclude sets modified by \emph{incertae sedis} marks}
+the subproblem contains OTT ids for internal nodes. Therefore we
+handle \emph{incertae sedis} information by constructing modified
+split sets for the lowest-ranked tree when the list of \emph{incertae
+sedis} nodes is not empty.
+
+\subsubsection{Exclude sets modified by emph{incertae sedis} marks}
 
@@
 algorithm to compute the exclude set for all nodes in a tree.
-\begin{enumerate}
-\item Set the exclude set of the root node to be empty
-\item For each \emph{node} (except the root) in preorder
-\begin{itemize}
-\item combine the \emph{exclude} set of the parent node with the \emph{include}
-set of non-\emph{incertae-sedis} siblings.
-\item store this set in a hash, with key \emph{node}
-\end{itemize}
-\end{enumerate}
-This algorithm is currently implemented in \emph{otc-solve-subproblem}.
-We store the sets as \emph{std::set}.
+\begin{enumerate} \item Set the exclude set of the root node to be
+empty \item For each \emph{node} (except the root) in preorder
+\begin{itemize} \item combine the \emph{exclude} set of the parent
+node with the \emph{include} set of non-\emph{incertae-sedis}
+siblings. \item store this set in a hash, with key \emph{node}
+\end{itemize} \end{enumerate} This algorithm is currently implemented
+in \emph{otc-solve- subproblem}. We store the sets as \emph{std::set}.
 
@@
 
-When multiple names $N=\{n_{1},\ldots n_{N}\}$ map to the same solution
-node $x$, then these names must satisfy some tree structure on the
-taxonomy, such that $n_{1}<n_{2}$ if $n_{1}$ is a descendant of
+When multiple names $N=\{n_{1},\ldots n_{N}\}$ map to the same
+solution node $x$, then these names must satisfy some tree structure
+on the taxonomy, such that $n_{1}<n_{2}$ if $n_{1}$ is a descendant of
 $n_{2}$ in the taxonomy. If it is possible to find a name $n_{max}$
-that is the unique maximal element of $N$, then it is permissible
-to 
-\begin{enumerate}
-\item create a monotypic parent $p(x)$ of $x$, and assign $n_{max}$ to
-$p(x)$
-\item continue handling name clashes at $x$ with the set of possible names
-reduced to $N-n_{max}$.
-\end{enumerate}
-However, its certainly possible that there might not be any such $N_{max}$,
-in which case we could just choose a name for $x$ from $N$ (perhaps
-not an \emph{incertae sedis} name) and then record all the other names
-as equivalents somewhere.
-
-\textbf{BDR:}\textbf{\emph{ }}\emph{we might get this behavior in
-a nice an automatic way if we create a single fake leaf for each monotypic
-taxonomy node that holds the node's leaf label.}
+that is the unique maximal element of $N$, then it is permissible to
+\begin{enumerate} \item create a monotypic parent $p(x)$ of $x$, and
+assign $n_{max}$ to $p(x)$ \item continue handling name clashes at $x$
+with the set of possible names reduced to $N-n_{max}$. \end{enumerate}
+However, its certainly possible that there might not be any such
+$N_{max}$, in which case we could just choose a name for $x$ from $N$
+(perhaps not an \emph{incertae sedis} name) and then record all the
+other names as equivalents somewhere.
+
+\textbf{BDR:}\textbf{\emph{ }}\emph{we might get this behavior in a
+nice an automatic way if we create a single fake leaf for each
+monotypic taxonomy node that holds the node's leaf label.}
 
@@
 
-When multiple I.S. taxa have been moved to the root node of a subproblem,
-they may be I.S. over the entire subproblem, and some may be I.S.
-over others in an asymmetric manner. Therefore, we might need to specify
-additional information about the original attachment location of the
-I.S. taxa, such as their depth. This only affects problems that have
-been decomposed.
+When multiple I.S. taxa have been moved to the root node of a
+subproblem, they may be I.S. over the entire subproblem, and some may
+be I.S. over others in an asymmetric manner. Therefore, we might need
+to specify additional information about the original attachment
+location of the I.S. taxa, such as their depth. This only affects
+problems that have been decomposed.
 
@@
 \emph{Question:} Does the synthesis tree contain any \emph{incertae
-sedis} groups?\\
-\emph{Answer:} The grafted supertree will not contain any \emph{incertae
-sedis} groups. However, when we attach pruned nodes to a parent in
-the grafted supertree, we could mark such nodes \emph{incertae sedis}
-if we want.
+sedis} groups?\\ \emph{Answer:} The grafted supertree will not contain
+any \emph{incertae sedis} groups. However, when we attach pruned nodes
+to a parent in the grafted supertree, we could mark such nodes
+\emph{incertae sedis} if we want.
 
@@
 Mark's document, he wasn't sure if such names were generated for nodes
-that had an IS taxon placed inside of them, so otc-unprune-solution-and-name-unnamed
-nodes throws away all the names and generates them itself.
-
-\textbf{BDR}: See document \texttt{otcetera/doc/unprune-solution-and-name-unnamed-nodes.pdf}
+that had an IS taxon placed inside of them, so otc-unprune-solution-
+and-name-unnamed nodes throws away all the names and generates them
+itself.
+
+\textbf{BDR}: See document texttt{otcetera/doc/unprune-solution-and-name-unnamed-nodes.pdf}
 
 The unpruner should record when unpruned nodes are \emph{incertae
-sedis}. Such nodes are unaffected by phylogenies, and so \emph{incertae
-sedis} annotations for them make good sense.
+sedis}. Such nodes are unaffected by phylogenies, and so
+\emph{incertae sedis} annotations for them make good sense.
 
@@
 
-Annotation primarily involves running a conflict analysis between
-the synthesis tree and each input tree. Since neither tree has any
+Annotation primarily involves running a conflict analysis between the
+synthesis tree and each input tree. Since neither tree has any
 \emph{incertae sedis} taxa, the conflict algorithm does not need to
@@
 We would also like to allow running a conflict analysis between the
-synthesis tree and the taxonomy tree. However, naming the nodes \emph{is}
-a (almost) run of conflict analysis on the taxonomy tree, and this
-has already been done in a prior step. So, the current annotation
-procedure actually works as-is.
-
-It would be nice to allow running conflict against the cleaned taxonomy,
-though. One way to do this would be to generated a ``placed taxonomy'',
-with groups extended to include \emph{incertae sedis} taxa that have
-been placed within them. This would not require any updating to the
-conflict-analysis code in the annotation step.%
+synthesis tree and the taxonomy tree. However, naming the nodes
+\emph{is} a (almost) run of conflict analysis on the taxonomy tree,
+and this has already been done in a prior step. So, the current
+annotation procedure actually works as-is.
+
+It would be nice to allow running conflict against the cleaned
+taxonomy, though. One way to do this would be to generated a ``placed
+taxonomy'', with groups extended to include \emph{incertae sedis} taxa
+that have been placed within them. This would not require any updating
+to the conflict-analysis code in the annotation step.
 \begin{comment}
@@
 the taxonomy if group $A$ has an incertae sedis group $B$ placed
-within it. This doesn't affect the annotations, since taxon names
-are added by the unpruner. But it could make perfectly fine input
-trees incorrectly look like they are the cause of broken taxa, if
-they contain IS taxa. Thus, it would be nice to have a modified conflict
+within it. This doesn't affect the annotations, since taxon names are
+added by the unpruner. But it could make perfectly fine input trees
+incorrectly look like they are the cause of broken taxa, if they
+contain IS taxa. Thus, it would be nice to have a modified conflict
 algorithm.
@@
 
-The current conflict algorithm is pretty fast, but it works by classifying
-tips into either (i) the include group or (ii) the exclude group.
-To avoid counting the exclude group for every split, we instead count
-the total number of children for each node, and assume that any children
-not in the include set are in the exclude set. This is no longer true
-when we have incertae sedis taxa. I suspect that if we want to handle
-incertae sedies, we'd need a third category (iii) for ``neither include
-group nor exclude group''.
+The current conflict algorithm is pretty fast, but it works by
+classifying tips into either (i) the include group or (ii) the exclude
+group. To avoid counting the exclude group for every split, we instead
+count the total number of children for each node, and assume that any
+children not in the include set are in the exclude set. This is no
+longer true when we have incertae sedis taxa. I suspect that if we
+want to handle incertae sedies, we'd need a third category (iii) for
+``neither include group nor exclude group''.
 \begin{enumerate}
-\item Get induced trees on intersection of leaf sets
-\item Compute depth for each node (nd->depth)
-\item Compute number of tips at or below each node (nd->n\_tips)
+ \item
+Get induced trees on intersection of leaf sets \item Compute depth for
+each node (nd->depth)
+ \item Compute number of tips at or below each
+node (nd->n\_tips)
  \item for each input tree node -> $nd$
  \begin{enumerate}
-\item skip the root
-\item skip monotypic
-\item if its a tip then find corresponding (``terminal'') edges in synth
-tree and continue
-\item leaves1 <- get the list of leaves in the include group of $nd$ (in
-input)
-\item L2 <- find the total number of tips (L2 = sum {[}nd->n\_tips| nd <-
-leaves1{]})
-\item leaves2 <- get list of corresponding synth leaf nodes (in synth)
-\item nodes <- find all nodes between leaves2 and the MRCA (in synth)
-\item MRCA <- mrca of leaves2 (in synth. this uses the nd->depth annotation)
-\item Compute number of tips in the include set (nd->include\_tips) below
-each node in $nodes$ (in synth)
-\item if n\_include\_tips(MRCA) == n\_tips(MRCA) then the MRCA displays
-$nd$
-\item if n\_include\_tips(MRCA) < n\_tips(MRCA) then
+   \item skip the root \item skip monotypic \item if
+its a tip then find corresponding (``terminal'') edges in synth tree
+and continue \item leaves1 <- get the list of leaves in the include
+group of $nd$ (in input) \item L2 <- find the total number of tips (L2
+= sum {[}nd->n\_tips| nd <- leaves1{]})
+   \item leaves2 <- get list of
+corresponding synth leaf nodes (in synth)
+   \item nodes <- find all
+nodes between leaves2 and the MRCA (in synth)
+   \item MRCA <- mrca of
+leaves2 (in synth. this uses the nd->depth annotation)
+   \item Compute
+number of tips in the include set (nd->include\_tips) below each node
+in $nodes$ (in synth)
+   \item if n\_include\_tips(MRCA) == n\_tips(MRCA)
+then the MRCA displays $nd$
+   \item if n\_include\_tips(MRCA) <
+n\_tips(MRCA) then 
    \begin{itemize} 
@@
       \begin{itemize}
-\item if (n\_include\_tipes(nd) < n\_tips(nd) and n\_include\_tips(nd) <
-l2)
+         \item if (n\_include\_tips(nd) < n\_tips(nd) and
+n\_include\_tips(nd) < l2)
           \begin{itemize}
@@
            \end{itemize}
-\item if there are no conflicts, then this is a resolved\_by.
+         \item if there are no conflicts, then this
+is a resolved_by. 
       \end{itemize}
@@
 This probably is outside the scope of the paper, but if we could come
-up with a modified conflict algorithm, that would be nice/useful,
-and probably novel. It would also probably be slower...
+up with a modified conflict algorithm, that would be nice/useful, and
+probably novel. It would also probably be slower... 
 \end{comment}
@@
 
-placement: 
-\begin{itemize}
-\item c1 <- (broken) C within A <- AB.
-\item c2<- (broken) C within A <- AB
-\end{itemize}
-result: we lose $C$, but keep $A$ and $AB$.
+placement:  \begin{itemize} \item c1 <- (broken) C within A <- AB.
+\item c2<- (broken) C within A <- AB \end{itemize} result: we lose
+$C$, but keep $A$ and $AB$.
 
@@
 
-taxonomy = (((a1,a2)A,(b1,b2)B),( ( (c11,c12)C1,(c21,c22)C2,(ci1,ci2)?CI
-)?C) ,(d1)D)
+taxonomy = (((a1,a2)A,(b1,b2)B),( (
+(c11,c12)C1,(c21,c22)C2,(ci1,ci2)?CI )?C) ,(d1)D)
 
@@
 
-synth-with-is: ((a1,a2)A,((b1,c1),(b2,c2))B) or ((a1,a2)A,((b1,c1),(b2,c2))C)
-
-Result: we name the interleaved taxon $B$ or $C$, but cannot name
-it both names.
+synth-with-is: ((a1,a2)A,((b1,c1),(b2,c2))B) or
+((a1,a2)A,((b1,c1),(b2,c2))C)
+
+Result: we name the interleaved taxon $B$ or $C$, but cannot name it
+both names.
 
@@
 see test cases.
-\begin{lyxcode}
-\begin{comment}
-\begin{lyxcode}
-\textbf{Whale:}~?ott6145835~{[}6145835~Brandtocetus/genus{]}~<-~ott698424~{[}698424~Cetacea/order{]}~placed~under~ott4942423~{[}4942423~Cetotherium~/~genus{]}~<-~ott698424~{[}698424~Cetacea/order{]}
-
-\textbf{Whale:}~?ott6145873~{[}6145873{]}~<-~ott698424~{[}698424~Cetacea~/~order{]}~placed~under~ott4942423~{[}4942423{]}~<-~ott698424~{[}698424{]}~
+\begin{comment} \begin{lyxcode}
+\textbf{Whale:}~ ?ott6145835~{[}6145835~Brandtocetus/genus{]}~<-~ott69
+8424~{[}698424~Cetacea/orde r{]}~placed~under~ott4942423~{[}4942423~Ce
+totherium~/~genus{]}~<-~ott698424~{[}6 98424~Cetacea/order{]}
+
+\textbf{Whale:}~?ott6145873~{[}6145873{]}~<-~ott698424~{[}698424~Cetac
+\ea~/~order
+\{]}~placed~under~ott4942423~{[}4942423{]}~<-~ott698424~{[}698424{]}~
 
@@
 
-\textbf{Insect:}~?ott18090~{[}18090{]}~<-~ott672501~{[}672501{]}~placed~under~ott275501~{[}275501{]}~<-~ott672501~{[}672501{]}
-
-\textbf{Insect:}~?ott190269~{[}190269{]}~<-~ott672501~{[}672501{]}~placed~under~ott275501~{[}275501{]}~<-~ott672501~{[}672501{]}~
-
-\textbf{Insect:}~?ott349638~{[}349638{]}~<-~ott672501~{[}672501{]}~placed~under~ott275501~{[}275501{]}~<-~ott672501~{[}672501{]}~
-
-\textbf{Insect:}~?ott1009313~{[}1009313{]}~<-~ott672501~{[}672501~Reduviidae~/~family{]}~placed~under~ott275501~{[}275501{]}~<-~ott672501~{[}672501{]}~
+\textbf{Insect:}~?ott18090~{[}18090{]}~<-~ott672501~{[}672501{]}~place
+\d~under~ot t275501~{[}275501{]}~<-~ott672501~{[}672501{]}
+
+\textbf{Insect:}~?ott190269~{[}190269{]}~<-~ott672501~{[}672501{]}~pla
+\ced~under~ ott275501~{[}275501{]}~<-~ott672501~{[}672501{]}~
+
+\textbf{Insect:}~?ott349638~{[}349638{]}~<-~ott672501~{[}672501{]}~pla
+\ced~under~ ott275501~{[}275501{]}~<-~ott672501~{[}672501{]}~
+
+\textbf{Insect:}~?ott1009313~{[}1009313{]}~<-~ott672501~{[}672501~Redu
+\viidae~/~f amily{]}~placed~under~ott275501~{[}275501{]}~<-~ott672501~
+\{[}672501{]}~
 
@@
 
-\textbf{Insect:~}?ott564110~{[}564110{]}~<-~ott939130~{[}939130~Thyrididae~/~family{]}~placed~under~ott757344~{[}757344{]}~<-~ott939130~{[}939130{]}~
-
-\textbf{Plant:}~?ott679485~{[}679485{]}~<-~ott1074933~{[}1074933~Lamioidea~/~subfamily{]}~placed~under~ott534796~{[}534796{]}~<-~
-
-\textbf{Plant:}~?ott729184~{[}729184{]}~<-~ott1074933~{[}1074933~Lamioidea~/~subfamily~{]}~placed~under~ott534796~{[}534796{]}~<-~ott1074933~{[}1074933{]}~
+\textbf{Insect:~}?ott564110~{[}564110{]}~<-~ott939130~{[}939130~Thyrid
+\idae~/~fam
+\ily{]}~placed~under~ott757344~{[}757344{]}~<-~ott939130~{[}939130{]}~
+
+\textbf{Plant:}~?ott679485~{[}679485{]}~<-~ott1074933~{[}1074933~Lamio
+\idea~/~sub family{]}~placed~under~ott534796~{[}534796{]}~<-~
+
+\textbf{Plant:}~?ott729184~{[}729184{]}~<-~ott1074933~{[}1074933~Lamio
+\idea~/~sub family~{]}~placed~under~ott534796~{[}534796{]}~<-~ott10749
+\33~{[}1074933{]}~
 
@@
 
-\textbf{Plant:}~?ott5144555~{[}5144555{]}~<-~ott23373~{[}23373{]}~placed~under~ott447802~{[}447802{]}~<-~ott23373~{[}23373{]}
+\textbf{Plant:}~?ott5144555~{[}5144555{]}~<-~ott23373~{[}23373{]}~plac
+\ed~under~o tt447802~{[}447802{]}~<-~ott23373~{[}23373{]}
 
@@
 
-\textbf{Fly:}~?ott441413~{[}441413~Rhodesiella~lungliensis/species{]}~<-~ott1014022~{[}1014022~Rubiaceae{]}~placed~under~ott571842~{[}571842~Pavetteae~/~tribe{]}~<-~ott997381~{[}997381~Ixoroidea/subfamily{]}~<-~ott1014022~{[}1014022~Rubiaceae/family{]}
-
-\textbf{Plant:}~?ott31033~{[}31033{]}~<-~ott627035~{[}627035{]}~placed~under~ott6019314~{[}6019314{]}~<-~ott627035~{[}627035{]}~
-
-\textbf{Bacteria:}~?ott4018918~{[}4018918{]}~<-~ott5248084~{[}5248084{]}~placed~under~ott167767~{[}167767{]}~<-~ott5248084~{[}5248084{]}~
-
-\textbf{Bacteria:}?ott4792759~{[}4792759{]}~<-~ott5248084~{[}5248084{]}~placed~under~ott167767~{[}167767{]}~<-~ott5248084~{[}5248084{]}~
+\textbf{Fly:}~?ott441413~{[}441413~Rhodesiella~lungliensis/species{]}~
+\<-~ott1014 022~{[}1014022~Rubiaceae{]}~placed~under~ott571842~{[}5718
+\42~Pavetteae~/~tribe{ ]}~<-~ott997381~{[}997381~Ixoroidea/subfamily{]
+\}~<-~ott1014022~{[}1014022~Rubia ceae/family{]}
+
+\textbf{Plant:}~?ott31033~{[}31033{]}~<-~ott627035~{[}627035{]}~placed
+\~under~ott 6019314~{[}6019314{]}~<-~ott627035~{[}627035{]}~
+
+\textbf{Bacteria:}~?ott4018918~{[}4018918{]}~<-~ott5248084~{[}5248084{
+\]}~placed~ under~ott167767~{[}167767{]}~<-~ott5248084~{[}5248084{]}~
+
+\textbf{Bacteria:}?ott4792759~{[}4792759{]}~<-~ott5248084~{[}5248084{]
+\}~placed~u nder~ott167767~{[}167767{]}~<-~ott5248084~{[}5248084{]}~
 
 -~Erysipelotrichaceae~broken 
-\end{lyxcode}
-\end{comment}
-\end{lyxcode}
-Should we do this? We could say: 
-\begin{itemize}
-\item there are 331 incertae sedis taxa that are mentioned in input trees.
-\begin{itemize}
+\end{lyxcode} \end{comment}
+Should we do this? We could say:  \begin{itemize} \item there are 331
+incertae sedis taxa that are mentioned in input trees. \begin{itemize}
 \item what is the kingom / phylum / class / order / genus / species
-\end{itemize}
-\item we placed 14 incertae sedis taxa inside a sister taxon. (``incertae
-sedis'', ``versus ``unplaced''?)
-\item we placed $o$ incertae sedis taxa \emph{outside} all sister taxa.
-\item we confirmed $n$ incertae sedis taxa as being separate from all sister
-(sampled) taxa??
-\item we avoiding breaking 8 taxa that had IS taxa placed inside them.
-\item we allowed $z_{1}$ new taxa into the synthesis tree that were incertae
-sedis.
-\item we allowed $z_{2}$ new taxa into the synthesis tree that are marked
-as extinct.
-\item some nodes have as many as $w$\emph{ incertae sedis} children, making
-them unbrowseable when incertae sedis children are not excluded.
-\item $v_{1}$ input trees were previously excluded \emph{entirely} because
-they are nested within in an incertae sedis taxon.
-\item $v_{2}$ input trees were previously excluded \emph{partially} because
-they are nested within in an incertae sedis taxon.
-\end{itemize}
-Currently the numbers $z_{1}$ and $z_{2}$.
+\end{itemize} \item we placed 14 incertae sedis taxa inside a sister
+taxon. (``incertae sedis'', ``versus ``unplaced''?) \item we placed
+$o$ incertae sedis taxa \emph{outside} all sister taxa. \item we
+confirmed $n$ incertae sedis taxa as being separate from all sister
+(sampled) taxa?? \item we avoiding breaking 8 taxa that had IS taxa
+placed inside them. \item we allowed $z_{1}$ new taxa into the
+synthesis tree that were incertae sedis. \item we allowed $z_{2}$ new
+taxa into the synthesis tree that are marked as extinct. \item some
+nodes have as many as $w$\emph{ incertae sedis} children, making them
+unbrowseable when incertae sedis children are not excluded. \item
+$v_{1}$ input trees were previously excluded \emph{entirely} because
+they are nested within in an incertae sedis taxon. \item $v_{2}$ input
+trees were previously excluded \emph{partially} because they are
+nested within in an incertae sedis taxon. \end{itemize} Currently the
+numbers $z_{1}$ and $z_{2}$.
 
@@
 
-One thing we could do (perhaps) that we are not currently doing, is
-to have nodes marked as incertae sedis on the synth tree. This would
-be easy enough if such nodes are not affected in any way by the input
-trees. Thus, when unpruning nodes we could mark any nodes \emph{incertae
-sedis} if they were marked \emph{incertae sedis} on the taxonomy.
+One thing we could do (perhaps) that we are not currently doing, is to
+have nodes marked as incertae sedis on the synth tree. This would be
+easy enough if such nodes are not affected in any way by the input
+trees. Thus, when unpruning nodes we could mark any nodes
+\emph{incertae sedis} if they were marked \emph{incertae sedis} on the
+taxonomy.
 
@@
 occur in any input tree. I think that if $A$ contains child $A_{1}$
-that is an input tree, and the taxon $A$ is not broken, then $A$
-will be placed, and thus any other children $A_{2},A_{3},\ldots,A_{n}$
-will also be placed, since they will be added as children of the (placed)
-node $A$ by the unpruner. This could be considered when deciding
-which nodes to suppress in the tree viewer.
+that is an input tree, and the taxon $A$ is not broken, then $A$ will
+be placed, and thus any other children $A_{2},A_{3},\ldots,A_{n}$ will
+also be placed, since they will be added as children of the (placed)
+node $A$ by the unpruner. This could be considered when deciding which
+nodes to suppress in the tree viewer.
 
@@
 
-Not all \emph{incertae sedis} taxa in our taxonomy are directly labeled
-\emph{incertae sedis} by taxonomists. \emph{Incertae sedis} taxa can
-also result from automatic merging of taxonomies to create the OpenTree
-taxonomy. For example, in Figure 3 of \citet{rees2017automated},
-cases \#4 and \#6 illustrate examples where merging of two taxonomies
-leads to a taxonomy with a taxon of uncertain placement. The reason
-is primarily that if taxonomy $\taxonomy_{1}$ contains more levels
-of hierarchy than taxonomy $\taxonomy_{2}$, then we must add internal
-nodes to $\taxonomy_{2}$ to align it to $\taxonomy_{1}$. However,
-if taxonomy $\taxonomy_{2}$ contains more leaves than $\taxonomy_{1}$,
-then it is unclear if these extra leaves should be nested inside the
-additional internal nodes, or not. Thus, the extra leaves are marked
-\emph{incertae sedis}.
-
-For example, if $\taxonomy_{1}=((a,b)x,(c,d)y)z$ and $\taxonomy_{2}=(a,b,c,d,e)z$
-then $a$ and $b$ in $\taxonomy_{2}$ should be nested within $x$,
-but we do not know if $e$ should be nested within $x$ or not. Thus
-we obtain $((a,b)x,(c,d)y,?e)z$, where $?$ indicates that taxon
-$e$ is marked \emph{incertae sedis}.
+Not all \emph{incertae sedis} taxa in our taxonomy are directly
+labeled \emph{incertae sedis} by taxonomists. \emph{Incertae sedis}
+taxa can also result from automatic merging of taxonomies to create
+the OpenTree taxonomy. For example, in Figure 3 of
+\citet{rees2017automated}, cases \#4 and \#6 illustrate examples where
+merging of two taxonomies leads to a taxonomy with a taxon of
+uncertain placement. The reason is primarily that if taxonomy
+$\taxonomy_{1}$ contains more levels of hierarchy than taxonomy
+$\taxonomy_{2}$, then we must add internal nodes to $\taxonomy_{2}$ to
+align it to $\taxonomy_{1}$. However, if taxonomy $\taxonomy_{2}$
+contains more leaves than $\taxonomy_{1}$, then it is unclear if these
+extra leaves should be nested inside the additional internal nodes, or
+not. Thus, the extra leaves are marked \emph{incertae sedis}.
+
+For example, if $\taxonomy_{1}=((a,b)x,(c,d)y)z$ and
+$\taxonomy_{2}=(a,b,c,d,e)z$ then $a$ and $b$ in $\taxonomy_{2}$
+should be nested within $x$, but we do not know if $e$ should be
+nested within $x$ or not. Thus we obtain $((a,b)x,(c,d)y,?e)z$, where
+$?$ indicates that taxon $e$ is marked \emph{incertae sedis}.
 
@@
 The splits-based semantics for incertae sedis taxa has many desirable
-properties. In allows the use of the BUILD algorithm to assess compatibility
-of split sets. However, it also has some properties that may not be
-expected. For example, two incertae sedis siblings can be freely interdigitated,
-which might not be the expected outcome. The current definition, also
-ignore ranks. Thus if we had two incertae sedis siblings that were
-genera, but other siblings were families, then one might expect that
-the genus-level incertae sedis taxa could be placed within families,
-but not intermixed. Since the current approach ignores ranks, it cannot
-do that.
+properties. In allows the use of the BUILD algorithm to assess
+compatibility of split sets. However, it also has some properties that
+may not be expected. For example, two incertae sedis siblings can be
+freely interdigitated, which might not be the expected outcome. The
+current definition, also ignore ranks. Thus if we had two incertae
+sedis siblings that were genera, but other siblings were families,
+then one might expect that the genus-level incertae sedis taxa could
+be placed within families, but not intermixed. Since the current
+approach ignores ranks, it cannot do that.
 
@@
 
-In theory it could be possible for an \emph{incertae sedis} taxon
-$A$ to be placed within an \emph{incertae sedis} sibling $B$. We
-initially implemented this semantics. It requires updating the recursion
-to define $I(m,n)$ as the descendants of $m$ that can be access
-from $n$ without traversing an \emph{incertae sedis} node.
-
-With a splits-based semantics, we cannot allow both (i) placing $A$
-in $B$ and (ii) placing $B$ in $A$ and also forbid interdigitating
-$A$ and $B$.
+In theory it could be possible for an \emph{incertae sedis} taxon $A$
+to be placed within an \emph{incertae sedis} sibling $B$. We initially
+implemented this semantics. It requires updating the recursion to
+define $I(m,n)$ as the descendants of $m$ that can be access from $n$
+without traversing an \emph{incertae sedis} node.
+
+With a splits-based semantics, we cannot allow both (i) placing $A$ in
+$B$ and (ii) placing $B$ in $A$ and also forbid interdigitating $A$
+and $B$.
 
@@
 $A$ and $B$, then this has the unfortunate consequence that the
-members of $A$ and $B$ may be inter-digitated. In many cases, \emph{incertae
-sedis} genera have siblings that are families, and the expected semantics
-is that genera may be placed within their sibling families, or may
-remain outside existing families. In such cases, inter-digitating
-the genera would not be expected behavior. Thus, we choose the semantics
-that \emph{incertae sedis} taxa exclude their \emph{incertae sedis}
-siblings. This is not because the other semantics is impossible. However,
-we note that use of the interdigitating semantics allows a situation
-where the names for $A$ and $B$ both apply to the same node.
+members of $A$ and $B$ may be inter-digitated. In many cases,
+\emph{incertae sedis} genera have siblings that are families, and the
+expected semantics is that genera may be placed within their sibling
+families, or may remain outside existing families. In such cases,
+inter-digitating the genera would not be expected behavior. Thus, we
+choose the semantics that \emph{incertae sedis} taxa exclude their
+\emph{incertae sedis} siblings. This is not because the other
+semantics is impossible. However, we note that use of the
+interdigitating semantics allows a situation where the names for $A$
+and $B$ both apply to the same node.
 
 This semantics also complicates placement, since it is possible to
-place $A$ as a sibling of $B$, and then place $B$ within $A$.
-This violates the current invariant that taxa are only ever placed
-within more tipward taxa. This makes the description of placement
-more complicated, since we can no longer simply record all cases where
-a parent node is not ancestor on the taxonomy.
+place $A$ as a sibling of $B$, and then place $B$ within $A$. This
+violates the current invariant that taxa are only ever placed within
+more tipward taxa. This makes the description of placement more
+complicated, since we can no longer simply record all cases where a
+parent node is not ancestor on the taxonomy.
 
@@
 sedis} genera that are asserted to be monophyletic, whereas another
-author might describe \emph{incertae sedis} genera that may be interdigitated.
-
-\bibliographystyle{upmplainnat}
-\bibliography{otcetera}
+author might describe \emph{incertae sedis} genera that may be
+interdigitated.
+
+\bibliographystyle{upmplainnat} bibliography{otcetera}
 

### 2018-04-19 doc/incertae_sedis.tex
minor

@@
 \usepackage{graphicx}
+\usepackage{setspace}
+\linespread{1.5}
 \usepackage[authoryear]{natbib}
@@
 \usepackage{color}
+\usepackage[colorlinks,linkcolor=blue,citecolor=blue]{hyperref}
 \begin{document}
@@
 splits in altered ranked list of splits:
-$\mathcal{S}^{\prime}=\mathcal{Z}(\taxonomy_{p})+S(T_{1})+S(T_{2})+\l
-dots+S(T_{n})+\mathcal{Y}(\taxonomy_{p})$.
+$\mathcal{S}^{\prime}=\mathcal{Z}(\taxonomy_{p})+S(T_{1})+S(T_{2})+\ldots+S(T_{n})+\mathcal{Y}(\taxonomy_{p})$.
 
@@
 are \emph{incertae sedis}, then the exclude set for a node is just the
-total tip set minus the include set for the node: \begin{align*}
-\mathcal{E}(e) & =\mathcal{L}-\mathcal{I}(e). \end{align*}
+total tip set minus the include set for the node:
+$\mathcal{E}(e)  =\mathcal{L}-\mathcal{I}(e)$.
 
@@
 \begin{enumerate}
-    \item An
-\emph{incertae sedis} node may intrude into its siblings and their
+    \item An \emph{incertae sedis} node may intrude into its siblings and their
 descendants.
@@
 that is created from $T$ by taking a taxon, attaching it to an
-ancestor, and marking it \emph{incertae sedis}.
-\begin{comment} MTH:
-I don't understand this \end{comment} \item The semantics of an
+ancestor, and marking it \emph{incertae sedis}. \mthcomment{MTH: I don't understand this.}
+    \item The semantics of an
 \emph{incertae sedis} node does not depend on assigned ranks, but only
@@
 for a node is just the total tip set minus the include set for the
-node: \begin{align*} \mathcal{E}(n) & =\mathcal{L}-\mathcal{I}(n).
-\end{align*}
+node: $ \mathcal{E}(n) =\mathcal{L}-\mathcal{I}(n)$.
 
@@
 $\mathcal{I}(m,n)$ to indicate descendants of $m$ that are excluded
-from $n$. Thus: \begin{align*} \mathcal{I}(m,n) & =\begin{cases}
+from $n$. Thus:
+\begin{align*}
+\mathcal{I}(m,n) & =
+    \begin{cases}
         \emptyset & \text{if }n\text{ is \emph{incertae sedis}\,and }m\text{
-is not}\\ \mathcal{I}(m) & \text{otherwise} \end{cases} \end{align*}
+is not}\\
+        \mathcal{I}(m) & \text{otherwise}
+    \end{cases}
+\end{align*}
 
@@
 its parent:
-
-\begin{align} \mathcal{E}(n) &
-=\mathcal{E}(parent(n))\cup\left[\mathcal{I}(m,n)\big|m\in
-siblings(n)\right].\label{eq:exclude-set-formula-1} \end{align} This
+\begin{align}
+    \mathcal{E}(n) & = \mathcal{E}(parent(n))\cup\left[\mathcal{I}(m,n)\big|m\in
+           siblings(n)\right].\label{eq:exclude-set-formula-1}
+\end{align}
+This
 formula allows us to compute exclude sets via a pre-order traversal on
@@
 
-\begin{align*} \mathcal{I^{\prime}}(n) & =\begin{cases} \emptyset &
-\text{if }n\in\mathcal{U}\\ \{n\} & \text{if }n\text{ is a leaf}\\{}
+\begin{align*}
+    \mathcal{I^{\prime}}(n) & =
+    \begin{cases}
+        \emptyset & \text{if }n\in\mathcal{U}\\
+        \{n\} & \text{if }n\text{ is a leaf}\\
         [\mathcal{I}^{\prime}(c)|c\in children(n)] & \text{otherwise.}
-\end{cases} \end{align*} We may then define the taxa excluded from $n$
+\end{cases}
+\end{align*} We may then define the taxa excluded from $n$
 as:
@@
 that is the unique maximal element of $N$, then it is permissible to
-\begin{enumerate} \item create a monotypic parent $p(x)$ of $x$, and
-assign $n_{max}$ to $p(x)$ \item continue handling name clashes at $x$
-with the set of possible names reduced to $N-n_{max}$. \end{enumerate}
+\begin{enumerate}
+    \item create a monotypic parent $p(x)$ of $x$, and
+assign $n_{max}$ to $p(x)$ 
+    \item continue handling name clashes at $x$
+with the set of possible names reduced to $N-n_{max}$. 
+\end{enumerate}
 However, its certainly possible that there might not be any such
@@
 
-\bibliographystyle{upmplainnat} bibliography{otcetera}
+\bibliographystyle{upmplainnat}
+\bibliography{otcetera}
 

### 2018-04-19 doc/incertae_sedis.tex
more minor

@@
 
-\subsection{label{subsec:Placement}Placement}
+\subsection{Placement}\label{subsec:Placement}
 
@@
 
-\subsection{Case 2: A nested emph{incertae sedis} clade}
+\subsection{Case 2: A nested \emph{incertae sedis} clade}
 
@@
 
-\subsection{Case 3: Placement of one incertae sedis taxon within
+\subsection{Case 3: Placement of one \emph{incertae sedis} taxon within
 another.}
@@
 
-\section{Handling emph{incertae sedis} taxa in the propinquity
+\section{Handling \emph{incertae sedis} taxa in the propinquity
 pipeline}

### 2018-04-20 doc/incertae_sedis.tex
Respond to a question.

@@
 \newcommand{\mthcomment}[1]{{\color{red} #1}\xspace}
+
+\newcommand{\bdrcomment}[1]{{\color{blue} #1}\xspace}
 
@@
 ancestor, and marking it \emph{incertae sedis}. \mthcomment{MTH: I don't understand this.}
+\bdrcomment{BDR: Figure \ref{fig:Placing-one-incertae} is an illustration of this.
+Suppose you have a genus that is incertae sedis and a family that is incertae sedis and attached at the kingdom level.
+We could in theory place the family (.e.g $I$ in the figure) at the right level.
+It then sits next to the incertae sedis genus (e.g.
+$AI$ in the figure).
+We could then place the genus within the family.
+
+The goal above is trying to say (and failing to be clear at saying) that if we have a family that is sibling to an incertae sedis genus, then the placement of the genus within the family is still a possibility if we reattach the family as a child of a kingdom node and mark it incertae sedis.
+}
     \item The semantics of an

### 2018-04-20 doc/incertae_sedis.tex
Minor edits and compile fixes.

@@
 
-\author{Benjamin D. Redelings and Mark T. Holder} maketitle
+\author{Benjamin D. Redelings and Mark T. Holder}
+
+\maketitle
 
@@
 The supertree method described by \citet{redelings2017supertree} is
-able to use a set of phylogentic estimates and taxonomy to produce a
+able to use a set of phylogenetic estimates and taxonomy to produce a
 supertree with clades names according to the taxonomy. However, the
@@
 taxonomy and input trees. Thus, the final supertree lacked taxa which
-were \emph{incertae sedis}.
+were \emph{incertae sedis} or the descendants of such taxa.
 
@@
 for applying taxonomic names to the final supertree. In addition to
-discussing how to interpret in the \emph{incertae sedis} annotation,
+discussing how to interpret the \emph{incertae sedis} annotation,
 we describe the algorithmic changes to the supertree pipeline that
@@
 approximation approach that relies on decomposition of the full
-problem into subproblems. The decomposition partitioning the list of
+problem into subproblems. The decomposition partitions the list of
 taxonomy splits $S(\taxonomy_{p})$ into two lists:
-$\mathcal{Z}(\taxonomy_{p}),$the list of splits from the taxonomy tree
+$\mathcal{Z}(\taxonomy_{p}),$ the list of splits from the taxonomy tree
 which are compatible with every rooted split in the phylogenetic
@@
 In practice, the decomposition allows the compatibility of splits to
-be performed on smaller, relabeled trees. This can be done much more
+be checked on smaller, relabeled trees. This can be done much more
 efficiently than if compatibility was checked using the BUILD
@@
 of the process of building OTT. Thus, the taxonomic tree which is used
-as an input to the supertree algorithm is pruned version of
+as an input to the supertree algorithm is a pruned version of
 $\taxonomy^{\ast}$. For our purposes, the relevant information from
@@
 labels $\mathcal{N}$, and set $\mathcal{U}$ of \emph{incertae sedis
-}taxa. In previous work, taxa in $\mathcal{U}$ were all pruned from
+}taxa. In previous work, taxa in $\mathcal{U}$ and all of their descendants were all pruned from
 the taxonomic hierarchy when $\taxonomy_{p}$ is extracted from OTT. In
 the present work, we consider supertree approaches that operate on a
-taxonomic tree $\taxonomy,$which is produced from
+taxonomic tree $\taxonomy$, which is produced from
 $\taxonomy^{\ast}$but does not prune \emph{incertae sedis} taxa.
@@
 would be viewed as plausible. Ideally, the \emph{incertae sedis}
-annotation some taxon \emph{x} would be accompanied by a specific
+annotation for some taxon \emph{x} would be accompanied by a specific
 statement about which other taxa might include \emph{x}. Detailed
@@
 
-As mentioned above, the incertae sedis may mean that a taxon is
+As mentioned above, the \emph{incertae sedis} annotation may mean that a taxon is
 unplaced, but only a few placements within the taxonomy are possible.
@@
 In principle this information is easy to incorporate. If
-$\mathcal{P}$is the set of taxa that are potential ancestors for
+$\mathcal{P}$ is the set of taxa that are potential ancestors for
 incertae sedis taxon $x$, then $\mathcal{L}(x)$ will occur in neither
@@
 However, in the \emph{incertae sedis} framework, it is possible for
-one name to apply to multiple nodes. For example, in Figure \ref{fig
-:One-name-can}b, the name $A$ can apply to nodes $x$ and $y$. Here the
+one name to apply to multiple nodes. For example, in Figure \ref{fig:One-name-can}b,
+the name $A$ can apply to nodes $x$ and $y$. Here the
 name $A$ corresponds to the split $A1\,A2\,|\bullet B\,C$, leaving out
@@
 \hfill{}\subfloat[Taxonomy tree]{\includegraphics[scale=0.5]{Figures/names2nodes1/tax-shaded}}
-\hfill{}subfloat[Solution tree]{\includegraphics[scale=0.5]{Figures/names2nodes1/synth-shaded}}
-\hfill{}subfloat[Solution tree]{\includegraphics[scale=0.5]{Figures/names2nodes1/solution2-shaded}}
+\hfill{}\subfloat[Solution tree]{\includegraphics[scale=0.5]{Figures/names2nodes1/synth-shaded}}
+\hfill{}\subfloat[Solution tree]{\includegraphics[scale=0.5]{Figures/names2nodes1/solution2-shaded}}
 \hfill{}
@@
 example, in Figure \ref{fig:Handling-incertae-sedis}, taxon $C$ is a
-descendant of taxon $A$ in the synthetic tree (Fig. \ref{fig:Handling-
-incertae-sedis}d), but not the taxonomy (Fig. \ref{fig:Handling-
-incertae-sedis}b). Thus, we could say that the synthetic tree places
+descendant of taxon $A$ in the synthetic tree (Fig. \ref{fig:Handling-incertae-sedis}d),
+but not the taxonomy (Fig. \ref{fig:Handling-incertae-sedis}b).
+Thus, we could say that the synthetic tree places
 $C$ within $A$. Furthermore, the synthetic tree additionally places
@@
 
-A slightly more general case is shown in Figure \ref{fig:Broken-
-incertae-sedis}. Here taxon $C1$ is an descendant of $A$ in the
+A slightly more general case is shown in Figure \ref{fig:Broken-incertae-sedis}.
+Here taxon $C1$ is an descendant of $A$ in the
 synthesis tree, but not the taxonomy. Thus $C1$ is placed within $A$.
@@
 placed within their siblings, or descendants of their siblings. As
-such, it does not handle cases such as Figure \ref{fig:Placing-one-
-incertae}, in which $I$ is first placed as sibling to $AI$, and then
+such, it does not handle cases such as Figure \ref{fig:Placing-one-incertae},
+in which $I$ is first placed as sibling to $AI$, and then
 $AI$ is placed within $I$. Note that $I$ is not a sibling of $AI$ or a
@@
 exclusion of all non-included taxa, but only some non-included taxa,
-as described in section \ref{sec:Semantics-of- incertae}. As a result
+as described in section \ref{sec:Semantics-of-incertae}. As a result
 of this change in semantics, placing an IS taxon $A$ within a sister
@@
 {[}\textbf{BDR:} \emph{maybe we should merge this with Figure.
-\ref{fig:Broken- incertae-sedis}. Although the point of figure
+\ref{fig:Broken-incertae-sedis}. Although the point of figure
 \ref{fig:placement-example1} is to talk about sub-problem
@@
 place an incertae sedis taxon in conflicting locations. This is
-illustrated in Figure \ref{fig :placement-example1}, where the IS
+illustrated in Figure \ref{fig:placement-example1}, where the IS
 taxon (ott7,ott6)ott10 is placed as sister to ott1 by phylogeny
@@
 
-\subsection{Case 1: An emph{incertae sedis} clade}
+\subsection{Case 1: An \emph{incertae sedis} clade}
 
@@
 the clade is not monophyletic in the synthesis tree, then we allow the
-members of the clade to be placed separately (Figure \ref{fig:Broken-
-incertae-sedis}).
+members of the clade to be placed separately (Figure \ref{fig:Broken-incertae-sedis}).
 

### 2018-04-20 doc/incertae_sedis.tex
line breaks after .

--- 2018-04-19 doc/incertae_sedis.tex
@@
 
+% \raggedright
+
 \global\long\def\taxonomy{\mbox{\ensuremath{\mathbb{T}}}}
@@
 
-\author{Benjamin D. Redelings and Mark T. Holder} maketitle
+\author{Benjamin D.~Redelings and Mark T.~Holder} 
+\maketitle
 
@@
 Supertree methods combine a collection of input trees with different
-taxon sets into a single tree on the combined taxon set. The methods
-are usually applied to leaf-labeled phylogenetic estimates. Here we
-describe a supertree approach that overcomes issues associated with
+taxon sets into a single tree on the combined taxon set.
+The methods are usually applied to leaf-labeled phylogenetic estimates. 
+Here we describe a supertree approach that overcomes issues associated with
 using a taxonomy as one of the inputs to a supertree method.
@@
 a comprehensive supertree \citep[see][]{redelings2017supertree}
-covering all of life. The approach is to combine information in
+covering all of life.
+The approach is to combine information in
 published phylogenies with a comprehensive taxonomy that supplies
-taxon names. The supertree method has been used with the Open Tree of
+taxon names.
+The supertree method has been used with the Open Tree of
 Taxonomy \citep[OTT hereafter, see][]{rees2017automated}, but in
-principle is general. The taxonomy is a crucial input for several
+principle is general.
+The taxonomy is a crucial input for several
 reasons: it covers a wider range of species, its list of names allows
 for alignment of different phylogenetic estimates to each other, and
-it provides names for clades. Only a small proportion of known species
+it provides names for clades.
+Only a small proportion of known species
 have been included in a phylogenetic analysis, thus a taxonomy is
-crucial for achieving comprehensive coverage of known taxa.
+important for achieving comprehensive coverage of known taxa.
 Phylogenetic estimates collected from the published literature often
-use different names for the same species. Lists of synonyms in OTT
+use different names for the same species.
+Lists of synonyms in OTT
 (and other taxonomies) allow data curators to ``align'' input
@@
 This alignment is important for recognizing when two different
-estimates refer to the same taxon. Biologists are familiar with names
-for ``higher'' taxa (taxa above the species rank). Thus, naming clades
+estimates refer to the same taxon.
+Biologists are familiar with names
+for ``higher'' taxa (taxa above the species rank).
+Thus, naming clades
 in the supertree makes the tree easier to navigate and use.
@@
 While some taxonomies may indicate hybrid taxa, the supertree approach
-described here does not consider any hybridization. Thus the taxonomic
-hierarchy can be converted to a tree. While not all taxonomists
+described here does not consider any hybridization.
+Thus the taxonomic
+hierarchy can be converted to a tree.
+While not all taxonomists
 believe that named taxa should correspond to clades \citep[see, for
@@
 mirrors the hierarchy of the taxonomy as an estimate of phylogenetic
-relationships. We refer to labels of taxonomy nodes as ``taxon
+relationships.
+We refer to labels of taxonomy nodes as ``taxon
 names'', and assume that taxon names and taxonomy nodes have a one-to-
-one correspondence. Note that the taxonomy may also contain out-degree
-1 nodes. These nodes correspond to monotypic taxa which contain a
+one correspondence.
+Note that the taxonomy may also contain out-degree
+1 nodes.
+These nodes correspond to monotypic taxa which contain a
 single child of lower taxonomic rank.
@@
 able to use a set of phylogentic estimates and taxonomy to produce a
-supertree with clades names according to the taxonomy. However, the
+supertree with clades names according to the taxonomy.
+However, the
 method was not able to accommodate the fact that taxonomies frequently
-contain nodes with uncertain placement. These nodes are often labelled
+contain nodes with uncertain placement.
+These nodes are often labelled
 ``\emph{incertae sedis}'', which means ``uncertain seat'' in Latin.
@@
 as ``\emph{incertae familia}'', indicating that it is unclear which
-family the genus should be placed in. \emph{Incertae sedis} taxa
+family the genus should be placed in.
+\emph{Incertae sedis} taxa
 frequently occur when a specimen is identified down to a given
-taxonomic level, but no further. Extinct taxa are often \emph{incertae
-sedis}. OTT is constructed from several source taxonomies, and these
+taxonomic level, but no further.
+Extinct taxa are often \emph{incertae
+sedis}.
+OTT is constructed from several source taxonomies, and these
 sources include various ways of indicating that the position of a
-taxon within the hierarchy is uncertain. OTT uses a series of
-``flags'' to annotate these taxa. The supertree of by
+taxon within the hierarchy is uncertain.
+OTT uses a series of
+``flags'' to annotate these taxa.
+The supertree of by
 \citet{redelings2017supertree} simply pruned these taxa from the
-taxonomy and input trees. Thus, the final supertree lacked taxa which
+taxonomy and input trees.
+Thus, the final supertree lacked taxa which
 were \emph{incertae sedis}.
@@
 uncertainty by using published phylogenies to place \emph{incertae
-sedis} taxa. Developing this method required deciding on a set
+sedis} taxa.
+Developing this method required deciding on a set
 operational semantics to be used when interpreting the \emph{incertae
-sedis} annotation. These semantics affect the interpretation of the
+sedis} annotation.
+These semantics affect the interpretation of the
 phylogenetic statements being made by the input taxonomy and the rules
-for applying taxonomic names to the final supertree. In addition to
+for applying taxonomic names to the final supertree.
+In addition to
 discussing how to interpret in the \emph{incertae sedis} annotation,
 we describe the algorithmic changes to the supertree pipeline that
-were required to adequately represent the taxonomic uncertainty. The
+were required to adequately represent the taxonomic uncertainty.
+The
 approach described here allows us to include thousands of new taxa in
-our supertree analysis that were previously filtered out. It will also
+our supertree analysis that were previously filtered out.
+It will also
 enable the Open Tree of Life project to include extinct taxa, since
-many of these taxa are \emph{incertae sedis}. This makes substantial
+many of these taxa are \emph{incertae sedis}.
+This makes substantial
 progress towards our goal of comprehensive inclusion of known taxa
@@
 of leaf-labeled rooted input phylogenies and a single rooted taxonomy
-tree $\taxonomy_{p}$ that is ranked below all of the input trees. As
+tree $\taxonomy_{p}$ that is ranked below all of the input trees.
+As
 discussed in \citet{redelings2017supertree}, the preferred tree would
-display as many of the highly ranked input splits as possible. If
+display as many of the highly ranked input splits as possible.
+If
 computational barriers were not an issue, the method would act by
@@
 $S(T_{i})$ denotes a list of non-trivial rooted splits created from a
-post-order traversal of tree $i$ and '$+$' denotes concatenation. Then
+post-order traversal of tree $i$ and '$+$' denotes concatenation.
+Then
 the supertree would be produced in a greedy manner by iterating over
 the list $S$, and adding each rooted split to a growing rooted
-supertree if that split is compatible with the supertree. In other
+supertree if that split is compatible with the supertree.
+In other
 words, the set $C$ is initialized to the empty set, and we then
 iteratively consider each split in the list $\mathcal{S}$ and insert
-it into $C$ if the resulting set remains jointly compatible. The BUILD
+it into $C$ if the resulting set remains jointly compatible.
+The BUILD
 algorithm\citep{AhoSSU1981} is used both to check compatibility, and
@@
 approximation approach that relies on decomposition of the full
-problem into subproblems. The decomposition partitioning the list of
+problem into subproblems.
+The decomposition partitioning the list of
 taxonomy splits $S(\taxonomy_{p})$ into two lists:
@@
 inputs, and $\mathcal{Y}(\taxonomy_{p})$, the list of taxonomic splits
-that conflict with at least one phylogenetic input split. Conceptually
+that conflict with at least one phylogenetic input split.
+Conceptually
 the approximate algorithm works by greedily adding all compatible
@@
 In practice, the decomposition allows the compatibility of splits to
-be performed on smaller, relabeled trees. This can be done much more
+be performed on smaller, relabeled trees.
+This can be done much more
 efficiently than if compatibility was checked using the BUILD
-algorithm of \citet{AhoSSU1981} on the full leaf set. Thus, the
+algorithm of \citet{AhoSSU1981} on the full leaf set.
+Thus, the
 supertree algorithm first divides the supertree problem into
@@
 
-The taxonomy tree, $\taxonomy_{p}$, is derived from OTT. As mentioned
+The taxonomy tree, $\taxonomy_{p}$, is derived from OTT.
+As mentioned
 above, OTT not only contains a taxonomic hierarchy, but also contains
 taxonomic labels and annotations (referred to as ``flags'') on the
-taxa to carry extra information. The entire hiearchy encoded by OTT is
+taxa to carry extra information.
+The entire hiearchy encoded by OTT is
 a complete taxonomic tree, $\taxonomy^{\ast}.$ Nodes in
 $\taxonomy^{\ast}$ have a one-to-one correspondence with a set of
-labels $\mathcal{N}$ that are called taxon names. We will decompose
+labels $\mathcal{N}$ that are called taxon names.
+We will decompose
 $\mathcal{N}$ into the taxon names for leaf taxa \emph{$\mathcal{L}$
-}and higher taxa $\mathcal{H}$. We will use $\mathcal{U}$ to refer to
+}and higher taxa $\mathcal{H}$.
+We will use $\mathcal{U}$ to refer to
 the subset ($\mathcal{U}\subseteq\mathcal{N}$) of taxa that are
-annotated with the \emph{incertae sedis }property. Other annotations
+annotated with the \emph{incertae sedis }property.
+Other annotations
 identify dubious taxa and taxonomic names that are unwanted artifacts
-of the process of building OTT. Thus, the taxonomic tree which is used
+of the process of building OTT.
+Thus, the taxonomic tree which is used
 as an input to the supertree algorithm is pruned version of
-$\taxonomy^{\ast}$. For our purposes, the relevant information from
+$\taxonomy^{\ast}$.
+For our purposes, the relevant information from
 the taxonomy consists of the taxonomy tree $\taxonomy^{\ast}$, the
 labels $\mathcal{N}$, and set $\mathcal{U}$ of \emph{incertae sedis
-}taxa. In previous work, taxa in $\mathcal{U}$ were all pruned from
-the taxonomic hierarchy when $\taxonomy_{p}$ is extracted from OTT. In
+}taxa.
+In previous work, taxa in $\mathcal{U}$ were all pruned from
+the taxonomic hierarchy when $\taxonomy_{p}$ is extracted from OTT.
+In
 the present work, we consider supertree approaches that operate on a
@@
 \citet{redelings2017supertree} describe a method called
-``exemplification'' that ensures this is the case. Thus pruning of the
+``exemplification'' that ensures this is the case.
+Thus pruning of the
 \emph{incertae sedis} taxa from the taxonomy was also accompanied by
@@
 id=36270&lvl=3&lin=f&keep=1&srchmode=1&unlock||unclassi fied
-Sylviidae].” This includes the genus Regulus. Thus in OTT 2.9 Regulus
+Sylviidae].” This includes the genus Regulus.
+Thus in OTT 2.9 Regulus
 (ott ID = 3599326) is placed inside the family Silvidae (OTT ID =
-259942). The “unclassified Sylviidae” does not appear in OTT; instead
+259942).
+The “unclassified Sylviidae” does not appear in OTT; instead
 Regulus is flagged as “unclassified,sibling higher” and all of the
@@
 While we get Regulus back, we do not end up placing it in Sylviidae
-because Sylviidae is broken. It conflicts with about 20 different
+because Sylviidae is broken.
+It conflicts with about 20 different
 trees.
@@
 
-Flags. The Open Tree’s reference taxonomy is produced by a tool called
+Flags.
+The Open Tree’s reference taxonomy is produced by a tool called
 smasher that notices hints that a taxon is incertae sedis and labels
@@
 to refer to all of the flags that denote taxa with uncertain
-placement.. While a taxonomist may intend to use incertae sedis to
+placement..
+While a taxonomist may intend to use incertae sedis to
 indicate a limited number of possible positions for a taxon to go in
 the taxonomy, we do not retain any such details.
-
-Container nodes. Some taxonomies have nodes with names like “Incertae
-sedis (Bacteria)”. Such nodes are called containers. They indicate
+Container nodes.
+Some taxonomies have nodes with names like “Incertae
+sedis (Bacteria)”.
+Such nodes are called containers.
+They indicate
 that each child node is an incertae sedis child of the container
-node's parent. The OTT retains container nodes with the was_container
+node's parent.
+The OTT retains container nodes with the was_container
 flag, but seems to have already moved all of the container's children
-to the container's parent. }
+to the container's parent.
+}
 
@@
 information from the taxonomic tree is contained in the set of rooted
-splits. Thus, one could imagine replacing that taxonomy,
+splits.
+Thus, one could imagine replacing that taxonomy,
 $\taxonomy_{p}$, with a set of trees $\mathcal{R}(\taxonomy_{p})$ each
@@
 $\taxonomy_{p}$is included in one tree in
-$\mathcal{R}(\taxonomy_{p})$. If the order of trees is set to be (with
+$\mathcal{R}(\taxonomy_{p})$.
+If the order of trees is set to be (with
 some abuse of notation):
@@
 We define rooted splits for each edge by noting that removing the edge
-would produce two trees. The leaf label sets of these trees the
-partition the full leaf set $\mathcal{L}$ into two groups. The include
+would produce two trees.
+The leaf label sets of these trees the
+partition the full leaf set $\mathcal{L}$ into two groups.
+The include
 set $\mathcal{I}(e)$ which does not contain the root, and the exclude
-set $\mathcal{E}(e)$ which does contain the root. Such a split may be
-written \[ \mathcal{I}(e)|\bullet\mathcal{E}(e). \] where the
-$\bullet$ indicates the root. The include set is the set of leaves
-that are descendants of the daughter node of the edge. The exclude set
-is the set of leaves that are not descendants of that node. If no taxa
+set $\mathcal{E}(e)$ which does contain the root.
+Such a split may be
+written \[ \mathcal{I}(e)|\bullet\mathcal{E}(e).\] where the
+$\bullet$ indicates the root.
+The include set is the set of leaves
+that are descendants of the daughter node of the edge.
+The exclude set
+is the set of leaves that are not descendants of that node.
+If no taxa
 are \emph{incertae sedis}, then the exclude set for a node is just the
@@
 $\mathcal{I}(n)$ for $\mathcal{I}(e)$, and $\mathcal{E}(n)$ for
-$\mathcal{E}(e)$. For any two rooted splits $A=A_{1}|\bullet A_{2}$
+$\mathcal{E}(e)$.
+For any two rooted splits $A=A_{1}|\bullet A_{2}$
 and $B=B_{1}|\bullet B_{2}$ , we say that $A$ \emph{displays }$B$ if
@@
 analyses, we must arrive at an operational definition of the meaning
-of the \emph{incertae sedis} label. The core meaning of the phrase is
+of the \emph{incertae sedis} label.
+The core meaning of the phrase is
 that the annotated taxon may actually be a member of one of the taxa
 that appear as siblings to it in the heirarchical representation of
-the taxonomy. In the most expansive interpretation, the author of the
+the taxonomy.
+In the most expansive interpretation, the author of the
 taxonomy is stating that the \emph{incertae sedis} taxon could be
@@
 However, frequently only a subset of the possible placement points
-would be viewed as plausible. Ideally, the \emph{incertae sedis}
+would be viewed as plausible.
+Ideally, the \emph{incertae sedis}
 annotation some taxon \emph{x} would be accompanied by a specific
-statement about which other taxa might include \emph{x}. Detailed
+statement about which other taxa might include \emph{x}.
+Detailed
 readings of the character argumentation used by taxonomists could
 provide this sort of specific information in many cases, but we lack
-such information on the scale of the entire OTT. Here we consider a
+such information on the scale of the entire OTT.
+Here we consider a
 few operational approaches for handling the \emph{incertae sedis}
@@
 automatically be mentioned in the exclude set of all of the edges that
-does not descend from. If a taxon rooted at node $n$ is marked as
+does not descend from.
+If a taxon rooted at node $n$ is marked as
 \emph{incertae sedis}, then the leaf labels that are descendants of
@@
 but these labels will not necessarily be the exclude set for all other
-nodes/edges. They may simply be omitted in some rooted split
+nodes/edges.
+They may simply be omitted in some rooted split
 representations, resulting in partial splits.
@@
 then run the rest of the pipeline of \citet{redelings2017supertree}
-unaltered. However this approach results in a less resolved tree
+unaltered.
+However this approach results in a less resolved tree
 because taxonomic vagueness (in the form of the \emph{incertae sedis}
@@
 For example, a genus may not be characterized as incertae sedis at the
-level of the subfamily. Thus, it could be placed inside any of the
+level of the subfamily.
+Thus, it could be placed inside any of the
 hypothesized sub-families, or within its own subfamily, but it would
@@
 
-In principle this information is easy to incorporate. If
+In principle this information is easy to incorporate.
+If
 $\mathcal{P}$is the set of taxa that are potential ancestors for
@@
 lacking in OTT) or rank-based information (which is frequently
-untrustworthy in OTT). Thus, here we pursue an approach of ascribing a
+untrustworthy in OTT).
+Thus, here we pursue an approach of ascribing a
 meaning to the \emph{incertae sedis} label which attempts to capture
@@
 that is created from $T$ by taking a taxon, attaching it to an
-ancestor, and marking it \emph{incertae sedis}. \mthcomment{MTH: I don't understand this.}
+ancestor, and marking it \emph{incertae sedis}.
+\mthcomment{MTH: I don't understand this.}
     \item The semantics of an
@@
 semantics instead of a semantics based on apply of a series of edit
-operations to the tree. The split-based semantics allows us to seek
+operations to the tree.
+The split-based semantics allows us to seek
 trees that are simultaneously consistent with a set of split-
-constraints. The use of a split-based semantics does impose some
-limitations. These come primarily because in the edit-based semantics,
+constraints.
+The use of a split-based semantics does impose some
+limitations.
+These come primarily because in the edit-based semantics,
 after we place an \emph{incertae sedis} taxon, it is no loner
@@
 This has an effect, for example, when we consider two incertae sedis
-siblings $A$ and $B$. If we can place $A$ within $B$, and $B$ within
+siblings $A$ and $B$.
+If we can place $A$ within $B$, and $B$ within
 $A$, then a split- based semantics allows us to freely inter-digitate
-$A$ and $B$. If we do not want to allow $A$ and $B$ to be inter-
-digitated, then we cannot both $A$ within $B$ and $B$ within $A$. In
+$A$ and $B$.
+If we do not want to allow $A$ and $B$ to be inter-
+digitated, then we cannot both $A$ within $B$ and $B$ within $A$.
+In
 addition, we can imagine that a family $A$ is incertae sedis within a
-kingdom. In an edit-based semantics, we could place $A$ as a sibling
+kingdom.
+In an edit-based semantics, we could place $A$ as a sibling
 to a genus $B$, mark $A$ as non-\emph{incertae-sedis}, and then place
-$B$ within $A$. In a split-based semantics, we must disallow either
+$B$ within $A$.
+In a split-based semantics, we must disallow either
 $B$ within $A$ or $A$ within $B$ if we want to disallow
@@
 As a result of this, we cannot satisfy all five of these criteria.
-Thus, we describe two alternative split-based semantics. The non-
+Thus, we describe two alternative split-based semantics.
+The non-
 interdigitation semantics gets \# (1,2,4,5), and the interdigitation
-semantics gets (1,3,4,5). This makes the point that each semantics
+semantics gets (1,3,4,5).
+This makes the point that each semantics
 results from different choices about which criteria to satisfy.
@@
 \emph{incertae sedis} taxa, we first propose a semantics of
-\emph{incertae sedis }in terms of rooted splits. We define rooted
+\emph{incertae sedis }in terms of rooted splits.
+We define rooted
 splits by noting that each edge of a tree divides the tip taxa
@@
 does not contain the root, and the exclude set $\mathcal{E}(e)$ which
-does contain the root. Such a split may be written \[
-\mathcal{I}(e)|\bullet\mathcal{E}(e). \] where the $\bullet$ indicates
-the root. If no taxa are \emph{incertae sedis}, then the exclude set
+does contain the root.
+Such a split may be written \[
+\mathcal{I}(e)|\bullet\mathcal{E}(e).
+\] where the $\bullet$ indicates
+the root.
+If no taxa are \emph{incertae sedis}, then the exclude set
 for a node is just the total tip set minus the include set for the
@@
 An \emph{incertae sedis} taxon can be moved into any of the
-descendants of its siblings. We seek to represent this by constructing
-modified splits for each branch of the taxonomy tree. The include sets
+descendants of its siblings.
+We seek to represent this by constructing
+modified splits for each branch of the taxonomy tree.
+The include sets
 of these splits remain unchanged, but we construct reduced exclude
@@
 \emph{incertae }sedis\emph{,} but should contain the descendants of
-its other siblings\emph{. }We additionally stipulate that it is not
+its other siblings\emph{.
+}We additionally stipulate that it is not
 possible for \emph{incertae sedis} taxa to intrude into their
-\emph{incertae sedis} siblings. (This is not the only way to construct
+\emph{incertae sedis} siblings.
+(This is not the only way to construct
 a split-based semantics for \emph{incertae sedis}, but it avoids
@@
 $\mathcal{I}(m,n)$ to indicate descendants of $m$ that are excluded
-from $n$. Thus:
+from $n$.
+Thus:
 \begin{align*}
@@
 children of all the ancestors of $n$, unless those children are
-\emph{incertae sedis}. To avoid traversing all ancestors separately
+\emph{incertae sedis}.
+To avoid traversing all ancestors separately
 for each node $n$, we note that the excluded children of any ancestors
-will be in the exclude set of the parent of $n$ already. Therefore, we
+will be in the exclude set of the parent of $n$ already.
+Therefore, we
 can write the exclude set of a node $n$ in terms of the exclude set of
@@
 Furthermore, consider placing $A$ as a sibling of $B$, and then
-placing $B$ within $A$. In order for this to work, $A$ must not
+placing $B$ within $A$.
+In order for this to work, $A$ must not
 exclude the incertae sedis children of $B$.
@@
 \emph{Incertae sedis} taxa thus exclude\emph{ incertae sedis}
-siblings. However, where the first semantics allows \emph{incertae
+siblings.
+However, where the first semantics allows \emph{incertae
 sedis }taxa to intrude into \emph{incertae sedis} descendants of
@@
 $B$ and then placing $B$ within $A$ if we disallow placing $A$ into
-incertae sedis descendants of its siblings. We may then allow placing
+incertae sedis descendants of its siblings.
+We may then allow placing
 those descendants within $A$.
@@
 We use $\mathcal{I}(m,n)$ to indicate the descendants of $m$ that are
-excluded from $n$. \begin{align*} \mathcal{I}(m,n) & =\begin{cases}
+excluded from $n$.
+\begin{align*} \mathcal{I}(m,n) & =\begin{cases}
 \emptyset & \text{if }n\text{ is \emph{incertae sedis}\,and }m\text{
@@
 \end{align*}
-I think this works... but
+I think this works...
+but
 need to double-check.
@@
 After constructing a supertree, tip nodes already have names in
-\emph{$\mathcal{L}$}. However, we still need to assign higher taxon
+\emph{$\mathcal{L}$}.
+However, we still need to assign higher taxon
 names to internal supertree nodes based on the taxonomy tree in the
-problem. Each taxon name $n$ corresponds to a split
+problem.
+Each taxon name $n$ corresponds to a split
 $S(n)=S(n)_{1}|\bullet S(n)_{2}$ on the corresponding branch of the
-taxonomy tree. Without \emph{incertae sedis}, such splits are always
+taxonomy tree.
+Without \emph{incertae sedis}, such splits are always
 of the form $S(n)_{1}|\bullet\mathcal{L}-S(n)_{1}$, but with
@@
 and each node can take at most one name, with the exception of
-monotypic taxa. Thus, we may simply search the solution tree for a
+monotypic taxa.
+Thus, we may simply search the solution tree for a
 node that has the same cluster $S(n)_{1}$ and apply the name $n$ to
@@
 However, in the \emph{incertae sedis} framework, it is possible for
-one name to apply to multiple nodes. For example, in Figure \ref{fig
-:One-name-can}b, the name $A$ can apply to nodes $x$ and $y$. Here the
+one name to apply to multiple nodes.
+For example, in Figure \ref{fig:One-name-can}b, 
+the name $A$ can apply to nodes $x$ and $y$.
+Here the
 name $A$ corresponds to the split $A1\,A2\,|\bullet B\,C$, leaving out
-$D$ since it is \emph{incertae sedis}. The two nodes $x$ and $y$
+$D$ since it is \emph{incertae sedis}.
+The two nodes $x$ and $y$
 display the splits $A1\,A2\,|\bullet B\,C\,D$ and $A1\,A2\,D|\bullet
@@
 $A1\,A2\,|\bullet B\,C$, so the name $A$ can apply to both $x$ and
-$y$. This cannot happen without\emph{ incertae sedis} taxa except at
-monotypic nodes. When faced with a choice about where to place a name,
+$y$.
+This cannot happen without\emph{ incertae sedis} taxa except at
+monotypic nodes.
+When faced with a choice about where to place a name,
 our solution is to find the most tip-ward node where the name can
-apply and attach the name to this node. This corresponds to a
+apply and attach the name to this node.
+This corresponds to a
 conservative choice about whether \emph{incertae sedis} taxa should be
@@
 
-It is also possible for multiple names to apply to a single node. For
+It is also possible for multiple names to apply to a single node.
+For
 example, in Figure \ref{fig:Multiple-names-can}a, the taxon $A$
 corresponds to the split $B1\,B2\,C\,|\bullet\,Y$, and its child taxon
-$B$ corresponds to the split $B1\,B2\,|\bullet\,Y$. The edge leading
+$B$ corresponds to the split $B1\,B2\,|\bullet\,Y$.
+The edge leading
 to $A$ is consistent with the split for $B$, but the name $B$ is
-applied to the node with the smallest include group. However, in the
-solution tree (Fig. \ref{fig:Multiple-names-can}b), there is is only
-node for both names $A$ and $B$ to apply to. In this case, we solve
+applied to the node with the smallest include group.
+However, in the
+solution tree (Fig.
+\ref{fig:Multiple-names-can}b), there is is only
+node for both names $A$ and $B$ to apply to.
+In this case, we solve
 this problem by introducing a monotypic parent, and applying $A$ to
-the newly created parent node (Fig. \ref{fig:Multiple-names-can}c).
+the newly created parent node (Fig.
+\ref{fig:Multiple-names-can}c).
 \begin{figure}
@@
 like to interpret the position of these names in terms of placing
-\emph{incertae sedis }taxa in a revised taxonomy. This would allow us
+\emph{incertae sedis }taxa in a revised taxonomy.
+This would allow us
 to interpret the synthesis tree as saying that phylogenetic
 information has (for example) placed genus A within family B, or
-perhaps outside of all named families. The simplest approach to
+perhaps outside of all named families.
+The simplest approach to
 placement involves noting whenever a taxon $B$ is a descendant of a
-taxon $A$ on the named synthesis tree but not the taxonomy tree. For
+taxon $A$ on the named synthesis tree but not the taxonomy tree.
+For
 example, in Figure \ref{fig:Handling-incertae-sedis}, taxon $C$ is a
-descendant of taxon $A$ in the synthetic tree (Fig. \ref{fig:Handling-
-incertae-sedis}d), but not the taxonomy (Fig. \ref{fig:Handling-
-incertae-sedis}b). Thus, we could say that the synthetic tree places
+descendant of taxon $A$ in the synthetic tree 
+(Fig.\ref{fig:Handling-incertae-sedis}d), but not the taxonomy 
+(Fig.\ref{fig:Handling-incertae-sedis}b).
+Thus, we could say that the synthetic tree places
 $C$ within $A$. Furthermore, the synthetic tree additionally places
@@
 
-A slightly more general case is shown in Figure \ref{fig:Broken-
-incertae-sedis}. Here taxon $C1$ is an descendant of $A$ in the
-synthesis tree, but not the taxonomy. Thus $C1$ is placed within $A$.
+A slightly more general case is shown in 
+Figure \ref{fig:Broken-incertae-sedis}.
+Here taxon $C1$ is an descendant of $A$ in the
+synthesis tree, but not the taxonomy.
+Thus $C1$ is placed within $A$.
 We note that the most recent common ancestor (MRCA) of $A$ and $C1$ in
@@
 $A\to AB\to root$, while the path to the MRCA from $C1$ is $C1\to C\to
-root$. Here the additional node $AB$ on the path to the MRCA from $A$
-indicates that $C1$ is placed within $AB$ as well as $A$. In contrast,
+root$.
+Here the additional node $AB$ on the path to the MRCA from $A$
+indicates that $C1$ is placed within $AB$ as well as $A$.
+In contrast,
 the additional node $C$ on the path to the MRCA indicates that $C1$ is
-a broken \emph{incertae sedis} taxon. Thus we may consider an
+a broken \emph{incertae sedis} taxon.
+Thus we may consider an
 algorithm that, for each taxon $B$ on the synthesis tree, finds the
 closest ancestral taxon $A$, and checks if $A$ is an ancestor of $B$
-on the taxonomy. If not, we compute the paths $A\to A_{1}\to\ldots\to
+on the taxonomy.
+If not, we compute the paths $A\to A_{1}\to\ldots\to
 MRCA(A,B)$ and $B\to B_{1}\to\ldots\to MRCA(A,B)$.
@@
 However, this assumes that \emph{incertae sedis} taxa are always
-placed within their siblings, or descendants of their siblings. As
+placed within their siblings, or descendants of their siblings.
+As
 such, it does not handle cases such as Figure \ref{fig:Placing-one-
 incertae}, in which $I$ is first placed as sibling to $AI$, and then
-$AI$ is placed within $I$. Note that $I$ is not a sibling of $AI$ or a
-descendant of a sibling. Furthermore, the MRCA of $I$ and $AI$ is the
+$AI$ is placed within $I$.
+Note that $I$ is not a sibling of $AI$ or a
+descendant of a sibling.
+Furthermore, the MRCA of $I$ and $AI$ is the
 root, and the path from $AI$ to the root includes $A$, but $A$ is not
-broken. We suggest that this conundrum may be resolved by first
+broken.
+We suggest that this conundrum may be resolved by first
 processing the placement of $I$ within $A$ by modifying the taxonomy
 to reflect this placement\footnote{Does this actually work? Some
-placements break taxa.}. When we consider the placement of $AI$ within
+placements break taxa.}.
+When we consider the placement of $AI$ within
 $I$ on this modified taxonomy, the MRCA of $I$ and $AI$ is now $A$
-instead of the root, and the correct interpretation results. We
+instead of the root, and the correct interpretation results.
+We
 suggest that by walking the tree in a pre-order fashion and processing
@@
 and labelling intermediate nodes as either additional placements or
-broken incertae sedis (or incertae\_sedis\_inherited) taxa holds. Such
+broken incertae sedis (or incertae\_sedis\_inherited) taxa holds.
+Such
 an approach should also handle cases containing nested incertae sedis
@@
 Placement of incertae sedis taxa by input trees is unfortunately not
-quite as simple as finding a single location where an I.S. taxon
-should attach. For example, when an incertae sedis taxon is broken,
+quite as simple as finding a single location where an I.S.
+taxon
+should attach.
+For example, when an incertae sedis taxon is broken,
 its children need to be ``placed'' separately.
@@
 and $C$ become \emph{incertae sedis} clades in their own right, that
-may attach separately, except that they . This is because none of $A$,
+may attach separately, except that they .
+This is because none of $A$,
 $B$, or $C$ is in the exclude set of the siblings of $D$.
@@
 Synthesis with \emph{incertae sedis} taxa has the potential to resolve
-uncertain taxon placements using information from phylogenies. Since
+uncertain taxon placements using information from phylogenies.
+Since
 phylogenies are ranked higher than taxonomy, they may place members of
 incertae sedis clades as they wish even when we do not take into
-account the semantics of incertae sedis taxa. However, without the
+account the semantics of incertae sedis taxa.
+However, without the
 incertae sedis semantics described above, placing a taxon $A$ within a
-taxon $B$ results in conflict with taxon $B$ in the taxonomy. This has
-three main effects: \begin{enumerate} \item The edge leading to clade
-$B$ is lost. \item Any taxonomy-only children of $B$ are placed at the
+taxon $B$ results in conflict with taxon $B$ in the taxonomy.
+This has
+three main effects: \begin{enumerate}
+    \item The edge leading to clade $B$ is lost.
+    \item Any taxonomy-only children of $B$ are placed at the
         MRCA of the phylogenetically informed taxa, instead of at node $B$.
-\item The name for taxon $B$ is lost. \item We do not record $A$ as
-being placed within $B$. \end{enumerate} In order to avoid a situation
+    \item The name for taxon $B$ is lost.
+    \item We do not record $A$ as being placed within $B$.
+\end{enumerate} In order to avoid a situation
 where input phylogenies conflict with a large number of taxa when
@@
 When the synthesis tree conflicts with a taxonomy node, we say that
-the taxon $B$ at that node is a broken taxon. Broken taxa have two
-main effects, both of which are negative. First, the name $B$ of the
-broken taxon is removed from the synthesis tree. Second, the
+the taxon $B$ at that node is a broken taxon.
+Broken taxa have two
+main effects, both of which are negative.
+First, the name $B$ of the
+broken taxon is removed from the synthesis tree.
+Second, the
 conflicting edge for taxon $B$ is not included in the synthesis tree.
@@
 placed with the children of $B$ that are mentioned in input
-phylogenies. Instead the taxonomy-only children of a broken taxon move
+phylogenies.
+Instead the taxonomy-only children of a broken taxon move
 towards the root of the synthesis tree and attach at the first higher-
@@
 \emph{incertae sedis} taxa into containing taxa without breaking the
-containing taxa. We change the semantics of names to imply, not the
+containing taxa.
+We change the semantics of names to imply, not the
 exclusion of all non-included taxa, but only some non-included taxa,
-as described in section \ref{sec:Semantics-of- incertae}. As a result
+as described in section \ref{sec:Semantics-of- incertae}.
+As a result
 of this change in semantics, placing an IS taxon $A$ within a sister
-taxon $B$ no longer results in conflict with $B$. This allows us to
+taxon $B$ no longer results in conflict with $B$.
+This allows us to
 retain the split for $B$ within the synthesis tree, so that taxonomy-
 only children of $B$ are correctly grouped with their siblings that
-are referenced by the input trees. We may then retain the name for the
-no-longer-broken taxon. Finally, we are then able to stop filtering
+are referenced by the input trees.
+We may then retain the name for the
+no-longer-broken taxon.
+Finally, we are then able to stop filtering
 \emph{incertae sedis} taxa, so that they appear in the synthesis tree.
@@
 However, in practice entire clades may be \emph{incertae sedis}, and
-so more complex issue arise. In this section we consider a number of
+so more complex issue arise.
+In this section we consider a number of
 cases that must be handled when attempting to place \emph{incertae
@@
 {[}\textbf{BDR:} \emph{maybe we should merge this with Figure.
-\ref{fig:Broken- incertae-sedis}. Although the point of figure
+\ref{fig:Broken- incertae-sedis}.
+Although the point of figure
 \ref{fig:placement-example1} is to talk about sub-problem
@@
 The addition of \emph{incertae sedis} taxa allows new types of
-conflict between input trees. For example, different input trees might
-place an incertae sedis taxon in conflicting locations. This is
+conflict between input trees.
+For example, different input trees might
+place an incertae sedis taxon in conflicting locations.
+This is
 illustrated in Figure \ref{fig :placement-example1}, where the IS
@@
 When this happens, the placement of the IS taxon is not influenced by
-its being marked IS on the taxonomy. Thus, in Example 1, the higher
+its being marked IS on the taxonomy.
+Thus, in Example 1, the higher
 ranked tree $\tau_{1}$ will be reflected in the synthesis tree, ott10
-will be placed as sister to ott1. In contrast, the conflicting
+will be placed as sister to ott1.
+In contrast, the conflicting
 placement in $\tau_{2}$ will not be reflected in the synth tree.
 
-All this would occur in the previous version of propinquity. Where the
+All this would occur in the previous version of propinquity.
+Where the
 updated version differs that (a) the names ott9 and ott8 are retained
-instead of being dropped. (b) as a result of not breaking ott8 and
-ott9, we do not move ott3 and ott2 up to ott11. {[}BDR: Extend more
+instead of being dropped.
+(b) as a result of not breaking ott8 and
+ott9, we do not move ott3 and ott2 up to ott11.
+{[}BDR: Extend more
 figures!{]}
@@
 When a clade is marked as \emph{incertae sedis}, we need to consider
-two cases. In the first case, the clade may be placed within a sister
-clade intact (Figure \ref{fig:Handling-incertae-sedis}). However, if
+two cases.
+In the first case, the clade may be placed within a sister
+clade intact (Figure \ref{fig:Handling-incertae-sedis}).
+However, if
 the clade is not monophyletic in the synthesis tree, then we allow the
@@
 \caption{\label{fig:Handling-incertae-sedis}Handling \emph{incertae sedis
-}taxa recovers additional edges and taxon names. Supertree construction
+}taxa recovers additional edges and taxon names.
+Supertree construction
 on input tree (a) and taxonomy tree (b) with \emph{incertae sedis}
-clade $C$ leads to synthesis tree (c). However, supertree construction
+clade $C$ leads to synthesis tree (c).
+However, supertree construction
 with \emph{incertae sedis} handling constructs tree (d), which recovers
-taxon name $AB$ and $A$, as well as the edge to clade $AB$. Taxon
+taxon name $AB$ and $A$, as well as the edge to clade $AB$.
+Taxon
 names and edges that are conditional on handling \emph{incertae sedis}
-are shaded grey. }
+are shaded grey.
+}
 
@@
 \caption{\label{fig:Broken-incertae-sedis} \emph{Incertae sedis} clade $C$
-broken by conflicting placement. Input trees (a) and (b) conflict
-in the placement of taxa in $C$. The synthesis supertree (d) places
+broken by conflicting placement.
+Input trees (a) and (b) conflict
+in the placement of taxa in $C$.
+The synthesis supertree (d) places
 places $C1$ and $C2$ separately within $A$, while unplaced taxa
-float upwards. (\textbf{Note:} This is what the sub-problem solver
-does. The entire pipeline would attach $C3$ at $A$ actually, since
-it is the MRCA of $A1$ and $A2$. Should we change this?) }
+float upwards.
+(\textbf{Note:} This is what the sub-problem solver
+does.
+The entire pipeline would attach $C3$ at $A$ actually, since
+it is the MRCA of $A1$ and $A2$.
+Should we change this?) }
 
@@
 When the input phylogenies conflict on where members of an incertae
-sedis clade should be placed, neither placement is rejected. Instead,
+sedis clade should be placed, neither placement is rejected.
+Instead,
 the taxon $C$ is broken, the name $C$ disappears, and $C3$ floats to
-the top level. Algorithmic, alternatively placements of children of an
+the top level.
+Algorithmic, alternatively placements of children of an
 incertae sedis clade lead to contesting of all edges along with taxa
@@
 In Figure \ref{fig:Broken-incertae-sedis}, tree $T_{1}$ places $C1$
-next to $A1$, while $T_{2}$ places $C2$ next to $A2$. Since different
+next to $A1$, while $T_{2}$ places $C2$ next to $A2$.
+Since different
 members of $C$ are placed, neither placement for $C$ is rejected.
 Instead the taxon $C$ is broken, the name $C$ disappears, and $C3$
-floats to the top level. Furthermore, since taxon $C$ is a broken
+floats to the top level.
+Furthermore, since taxon $C$ is a broken
 \emph{incertae sedis }taxon, all of its children are effectively
 incertae sedis independently, with the exception that they cannot be
-placed within each other. Therefore, the names $A$ and $AB$ are not
+placed within each other.
+Therefore, the names $A$ and $AB$ are not
 lost, since the taxa placed within them are \emph{incertae sedis}
@@
 The situation here would be different if the monophyly of $C$ was
-supported by an high-ranked input phylogeny. In that case, instead of
+supported by an high-ranked input phylogeny.
+In that case, instead of
 breaking $C$, we would choose the placement in the highest-ranked
@@
 \emph{incertae sedis} taxon to be nested within another \emph{incertae
-sedis} taxon. In such a case we must allow the higher-ranked taxon to
+sedis} taxon.
+In such a case we must allow the higher-ranked taxon to
 be placed within a sibling, while preserving the right of the lower
@@
 
-\caption{\label{fig:Nested-incertae-sedis}Nested incertae sedis taxa. (a)
-Input tree \#1 places CI1 within CG, and C within $A$. (b) In the
+\caption{\label{fig:Nested-incertae-sedis}Nested incertae sedis taxa.
+(a)
+Input tree \#1 places CI1 within CG, and C within $A$.
+(b) In the
 taxonomy, CI is \emph{incertae sedis }within C, while C is \emph{incertae
-sedis} under the root. (c) The synthesis tree when \emph{incertae
-sedis} taxa are not considered. The names AB, A, and CG are lost.
+sedis} under the root.
+(c) The synthesis tree when \emph{incertae
+sedis} taxa are not considered.
+The names AB, A, and CG are lost.
 (d) The synthesis tree when \emph{incertae sedis} taxa are considered.
@@
 In our semantics of incertae sedis taxa, it is possible for to place
-one \emph{incertae sedis} sibling inside another. A special case of
+one \emph{incertae sedis} sibling inside another.
+A special case of
 this is when a higher-rank incertae sedis taxon $A$ is placed as
@@
 \caption{\label{fig:Placing-one-incertae}Placing one incertae sedis group
-within another. Here I is placed within A, and then AI is placed within
+within another.
+Here I is placed within A, and then AI is placed within
 A.}
@@
 In order to handle \emph{incertae sedis} taxa within propinquity, we
-must modify some of the stages of the propinquity pipeline. Subproblem
+must modify some of the stages of the propinquity pipeline.
+Subproblem
 decomposition must place \emph{incertae sedis} taxa in the correct
-subproblem. Subproblem files must indicate which taxa are incertae
-sedis. The subproblem solver must read this information, account for
+subproblem.
+Subproblem files must indicate which taxa are incertae
+sedis.
+The subproblem solver must read this information, account for
 \emph{incertae sedis} taxa when solving subproblems, and correctly
 name taxa that have been modified by having \emph{incertae sedis} taxa
-place inside them. The unpruner must be aware of \emph{incertae sedis}
-taxa. Annotations of the tree must be aware of \emph{incertae sedis}
+place inside them.
+The unpruner must be aware of \emph{incertae sedis}
+taxa.
+Annotations of the tree must be aware of \emph{incertae sedis}
 taxa so that it does not consider taxa broken when they have an
@@
 tend to have a very large number of incertae sedis children, making
-browsing in the tree viewer difficult. This can happen when, for
+browsing in the tree viewer difficult.
+This can happen when, for
 example, fossils or other hard-to- place taxa get classified only to
-the level of these well-known nodes and no further. This leads to a
+the level of these well-known nodes and no further.
+This leads to a
 situation where well-known taxa serve as a dumping ground for unplaced
@@
 Our current approach to this problem is to perform a second round of
-pruning, or ``cleaning'', during the exemplification step. Incertae
+pruning, or ``cleaning'', during the exemplification step.
+Incertae
 sedis taxa are pruned at this stage if they do not occur in any input
-trees. We thus generate a second ``cleaned taxonomy'' that has
-undergone this further round of cleaning. This approach improves on
+trees.
+We thus generate a second ``cleaned taxonomy'' that has
+undergone this further round of cleaning.
+This approach improves on
 the previous approach in that \emph{incertae sedis} taxa in input
-trees are no longer pruned. This approach also removes tons of
+trees are no longer pruned.
+This approach also removes tons of
 \emph{incertae sedis} children from nodes like ``Fungi'', where a lot
@@
 However, this approach has the negative effect of pruning some
-incertae sedis taxa that need not be pruned. For example, suppose
+incertae sedis taxa that need not be pruned.
+For example, suppose
 \emph{incertae sedis} taxon $A$ contains 5 children, of which only 1
-child $A_{1}$ occurs in an input tree. If the taxon $A$ is not broken,
+child $A_{1}$ occurs in an input tree.
+If the taxon $A$ is not broken,
 then it should be possible to attach the other 4 members of $A$ next
-to $A_{1}$, without cluttering up the synthesis tree. Such taxa have
+to $A_{1}$, without cluttering up the synthesis tree.
+Such taxa have
 been successfully placed even though they are not in any input tree.
@@
 placed (see section \ref{subsec:Placement}), and then mark the other
-ones incertae sedis in the \emph{synthesis tree}. We could then filter
+ones incertae sedis in the \emph{synthesis tree}.
+We could then filter
 out the taxa that \emph{remain }incertae sedis (or incertae\_sedis
@@
 problem decomposition, since taxonomy edges no longer completely
-separate subproblems. Instead, \emph{incertae sedis} taxa may attach
-on either side of a taxonomy edge. We seek to place \emph{incertae
+separate subproblems.
+Instead, \emph{incertae sedis} taxa may attach
+on either side of a taxonomy edge.
+We seek to place \emph{incertae
 sedis} taxa into subproblems in such a way that the subproblem solver
-can perform the placement inside the subproblem. This approach
+can perform the placement inside the subproblem.
+This approach
 postpones handling of conflict in \emph{incertae sedis} taxa to the
 subproblem solver, where the problem is well formulated in terms of
-splits. However, it does have the effect of creating larger
+splits.
+However, it does have the effect of creating larger
 subproblems.
@@
 We must also handle conflicting placements of \emph{incertae sedis}
-taxa by different input trees. Thus, if one input tree places the
+taxa by different input trees.
+Thus, if one input tree places the
 \emph{incertae sedis} taxon $X$ in $((X)B)A$ and another places $X$ in
@@
 even if these edges would \emph{not} be contested were taxon $X$ to be
-removed. This results in a new way to contest edges that involves the
+removed.
+This results in a new way to contest edges that involves the
 interaction of two input trees, and not just the interaction of each
@@
 We choose to solve these problems by merging any subproblems that an
-\emph{incertae sedis} taxon might be placed in. The simplest way to
+\emph{incertae sedis} taxon might be placed in.
+The simplest way to
 achieve this is simply to regard any taxon that has an \emph{incertae
-sedis }taxon placed within it as contested. This results in marking
-both $B$ and $C$ as contested edges in the example above. In fact,
+sedis }taxon placed within it as contested.
+This results in marking
+both $B$ and $C$ as contested edges in the example above.
+In fact,
 this is the current behaviour of the non-\emph{incertae-sedis} aware
-subproblem decomposer. One downside of this approach is that, if we
+subproblem decomposer.
+One downside of this approach is that, if we
 have $((X)B)A)$ in one input tree, and $X$ is mentioned nowhere else,
 then by marking $B$ as contested, we are creating unnecessarily large
-subproblems. We could instead placed $X$ in $B$ and avoid contesting
-the edge $B$. However, this approach is more complex and does not seem
+subproblems.
+We could instead placed $X$ in $B$ and avoid contesting
+the edge $B$.
+However, this approach is more complex and does not seem
 necessary in practice.
@@
 
-\caption{\label{fig:placement-example1}Example. An incertae sedis clade (ott6,ott7)
+\caption{\label{fig:placement-example1}Example.
+An incertae sedis clade (ott6,ott7)
 is placed in different subtrees by input trees $\tau_{1}$ and $\tau_{2}$.
@@
 extended to include (ott6,ott7), and the branches leading to these
-nodes have been colored blue. The dashed blue edge leads to a node
+nodes have been colored blue.
+The dashed blue edge leads to a node
 that is a newly-introduced degree-2 node which does not correspond
-to any taxonomy node. In $\tau_{2}$, only one node that corresponds
-to a taxonomy node needs to have its ingroup extended. The placement
+to any taxonomy node.
+In $\tau_{2}$, only one node that corresponds
+to a taxonomy node needs to have its ingroup extended.
+The placement
 of (ott6,ott7) into ott8 toward ott1 by $\tau_{1}$ conflicts with
@@
 \ref{fig:placement-example1}, input tree $\tau_{1}$ contests ott9 and
-input tree $\tau_{2}$ contests ott9 and ott8. Thus ott1, ott2, ott3,
+input tree $\tau_{2}$ contests ott9 and ott8.
+Thus ott1, ott2, ott3,
 ott6, and ott7 end up in the same sub-problem.
@@
 sedis information and computing partial splits for \emph{incertae
-sedis} taxa before solving a sub-problem. After solving a sub-problem,
+sedis} taxa before solving a sub-problem.
+After solving a sub-problem,
 we must apply taxon names from the taxonomy tree to the sub-problem
-solution tree. The solution tree is considered to a fixed tree and not
+solution tree.
+The solution tree is considered to a fixed tree and not
 to have any \emph{incertae sedis} nodes, or any other forms of
@@
 Currently, we read the \emph{incertae sedis} information as a list of
-OTT ids for \emph{incertae sedis} taxa. This does not require adding
-further annotations to the node names. Only taxonomy nodes can be
+OTT ids for \emph{incertae sedis} taxa.
+This does not require adding
+further annotations to the node names.
+Only taxonomy nodes can be
 \emph{incertae sedis} at the moment, and only the taxonomy tree for
-the subproblem contains OTT ids for internal nodes. Therefore we
+the subproblem contains OTT ids for internal nodes.
+Therefore we
 handle \emph{incertae sedis} information by constructing modified
@@
 node with the \emph{include} set of non-\emph{incertae-sedis}
-siblings. \item store this set in a hash, with key \emph{node}
+siblings.
+\item store this set in a hash, with key \emph{node}
 \end{itemize} \end{enumerate} This algorithm is currently implemented
-in \emph{otc-solve- subproblem}. We store the sets as \emph{std::set}.
+in \emph{otc-solve- subproblem}.
+We store the sets as \emph{std::set}.
 
@@
 To find the node for a name $n$, we find the MRCA of the cluster
-$S_{1}(n)$. If the MRCA excludes the entire exclude group $S_{2}(n)$
+$S_{1}(n)$.
+If the MRCA excludes the entire exclude group $S_{2}(n)$
 then the name applies to the MRCA; otherwise the taxon does not exist
@@
 on the taxonomy, such that $n_{1}<n_{2}$ if $n_{1}$ is a descendant of
-$n_{2}$ in the taxonomy. If it is possible to find a name $n_{max}$
+$n_{2}$ in the taxonomy.
+If it is possible to find a name $n_{max}$
 that is the unique maximal element of $N$, then it is permissible to
@@
 with the set of possible names reduced to $N-n_{max}$.
+
 \end{enumerate}
@@
 
-When multiple I.S. taxa have been moved to the root node of a
-subproblem, they may be I.S. over the entire subproblem, and some may
-be I.S. over others in an asymmetric manner. Therefore, we might need
+When multiple I.S.
+taxa have been moved to the root node of a
+subproblem, they may be I.S.
+over the entire subproblem, and some may
+be I.S.
+over others in an asymmetric manner.
+Therefore, we might need
 to specify additional information about the original attachment
-location of the I.S. taxa, such as their depth. This only affects
+location of the I.S.
+taxa, such as their depth.
+This only affects
 problems that have been decomposed.
@@
 \textbf{BDR:} \emph{currently we don't actually move taxa to get them
-into a subproblem. So, is this even an issue?}
+into a subproblem.
+So, is this even an issue?}
 
@@
 sedis} groups?\\ \emph{Answer:} The grafted supertree will not contain
-any \emph{incertae sedis} groups. However, when we attach pruned nodes
+any \emph{incertae sedis} groups.
+However, when we attach pruned nodes
 to a parent in the grafted supertree, we could mark such nodes
@@
 Currently the unpruner \emph{does not} require that the OTT ids are
-named in the grafted solution before unpruning starts. According to
+named in the grafted solution before unpruning starts.
+According to
 Mark's document, he wasn't sure if such names were generated for nodes
@@
 The unpruner should record when unpruned nodes are \emph{incertae
-sedis}. Such nodes are unaffected by phylogenies, and so
+sedis}.
+Such nodes are unaffected by phylogenies, and so
 \emph{incertae sedis} annotations for them make good sense.
@@
 Annotation primarily involves running a conflict analysis between the
-synthesis tree and each input tree. Since neither tree has any
+synthesis tree and each input tree.
+Since neither tree has any
 \emph{incertae sedis} taxa, the conflict algorithm does not need to
-change. Furthermore, if we allow \emph{incertae sedis} taxa that are
+change.
+Furthermore, if we allow \emph{incertae sedis} taxa that are
 taxonomy-only to be annotated as \emph{incertae sedis} on the synth
@@
 We would also like to allow running a conflict analysis between the
-synthesis tree and the taxonomy tree. However, naming the nodes
+synthesis tree and the taxonomy tree.
+However, naming the nodes
 \emph{is} a (almost) run of conflict analysis on the taxonomy tree,
-and this has already been done in a prior step. So, the current
+and this has already been done in a prior step.
+So, the current
 annotation procedure actually works as-is.
@@
 It would be nice to allow running conflict against the cleaned
-taxonomy, though. One way to do this would be to generated a ``placed
+taxonomy, though.
+One way to do this would be to generated a ``placed
 taxonomy'', with groups extended to include \emph{incertae sedis} taxa
-that have been placed within them. This would not require any updating
+that have been placed within them.
+This would not require any updating
 to the conflict-analysis code in the annotation step.
@@
 the taxonomy if group $A$ has an incertae sedis group $B$ placed
-within it. This doesn't affect the annotations, since taxon names are
-added by the unpruner. But it could make perfectly fine input trees
+within it.
+This doesn't affect the annotations, since taxon names are
+added by the unpruner.
+But it could make perfectly fine input trees
 incorrectly look like they are the cause of broken taxa, if they
-contain IS taxa. Thus, it would be nice to have a modified conflict
+contain IS taxa.
+Thus, it would be nice to have a modified conflict
 algorithm.
@@
 classifying tips into either (i) the include group or (ii) the exclude
-group. To avoid counting the exclude group for every split, we instead
+group.
+To avoid counting the exclude group for every split, we instead
 count the total number of children for each node, and assume that any
-children not in the include set are in the exclude set. This is no
-longer true when we have incertae sedis taxa. I suspect that if we
+children not in the include set are in the exclude set.
+This is no
+longer true when we have incertae sedis taxa.
+I suspect that if we
 want to handle incertae sedies, we'd need a third category (iii) for
@@
    \item MRCA <- mrca of
-leaves2 (in synth. this uses the nd->depth annotation)
+leaves2 (in synth.
+this uses the nd->depth annotation)
    \item Compute
@@
 is a resolved_by.
+
       \end{itemize}
@@
 up with a modified conflict algorithm, that would be nice/useful, and
-probably novel. It would also probably be slower... 
+probably novel.
+It would also probably be slower...
+
 \end{comment}
@@
 Should we do this? We could say:  \begin{itemize} \item there are 331
-incertae sedis taxa that are mentioned in input trees. \begin{itemize}
+incertae sedis taxa that are mentioned in input trees.
+\begin{itemize}
 \item what is the kingom / phylum / class / order / genus / species
 \end{itemize} \item we placed 14 incertae sedis taxa inside a sister
-taxon. (``incertae sedis'', ``versus ``unplaced''?) \item we placed
-$o$ incertae sedis taxa \emph{outside} all sister taxa. \item we
+taxon.
+(``incertae sedis'', ``versus ``unplaced''?) \item we placed
+$o$ incertae sedis taxa \emph{outside} all sister taxa.
+\item we
 confirmed $n$ incertae sedis taxa as being separate from all sister
 (sampled) taxa?? \item we avoiding breaking 8 taxa that had IS taxa
-placed inside them. \item we allowed $z_{1}$ new taxa into the
-synthesis tree that were incertae sedis. \item we allowed $z_{2}$ new
-taxa into the synthesis tree that are marked as extinct. \item some
+placed inside them.
+\item we allowed $z_{1}$ new taxa into the
+synthesis tree that were incertae sedis.
+\item we allowed $z_{2}$ new
+taxa into the synthesis tree that are marked as extinct.
+\item some
 nodes have as many as $w$\emph{ incertae sedis} children, making them
-unbrowseable when incertae sedis children are not excluded. \item
+unbrowseable when incertae sedis children are not excluded.
+\item
 $v_{1}$ input trees were previously excluded \emph{entirely} because
-they are nested within in an incertae sedis taxon. \item $v_{2}$ input
+they are nested within in an incertae sedis taxon.
+\item $v_{2}$ input
 trees were previously excluded \emph{partially} because they are
-nested within in an incertae sedis taxon. \end{itemize} Currently the
+nested within in an incertae sedis taxon.
+\end{itemize} Currently the
 numbers $z_{1}$ and $z_{2}$.
@@
 One thing we could do (perhaps) that we are not currently doing, is to
-have nodes marked as incertae sedis on the synth tree. This would be
+have nodes marked as incertae sedis on the synth tree.
+This would be
 easy enough if such nodes are not affected in any way by the input
-trees. Thus, when unpruning nodes we could mark any nodes
+trees.
+Thus, when unpruning nodes we could mark any nodes
 \emph{incertae sedis} if they were marked \emph{incertae sedis} on the
@@
 that are ``unplaced'' from \emph{incertae sedis} taxa that do not
-occur in any input tree. I think that if $A$ contains child $A_{1}$
+occur in any input tree.
+I think that if $A$ contains child $A_{1}$
 that is an input tree, and the taxon $A$ is not broken, then $A$ will
@@
 also be placed, since they will be added as children of the (placed)
-node $A$ by the unpruner. This could be considered when deciding which
+node $A$ by the unpruner.
+This could be considered when deciding which
 nodes to suppress in the tree viewer.
@@
 Not all \emph{incertae sedis} taxa in our taxonomy are directly
-labeled \emph{incertae sedis} by taxonomists. \emph{Incertae sedis}
+labeled \emph{incertae sedis} by taxonomists.
+\emph{Incertae sedis}
 taxa can also result from automatic merging of taxonomies to create
-the OpenTree taxonomy. For example, in Figure 3 of
+the OpenTree taxonomy.
+For example, in Figure 3 of
 \citet{rees2017automated}, cases \#4 and \#6 illustrate examples where
 merging of two taxonomies leads to a taxonomy with a taxon of
-uncertain placement. The reason is primarily that if taxonomy
+uncertain placement.
+The reason is primarily that if taxonomy
 $\taxonomy_{1}$ contains more levels of hierarchy than taxonomy
 $\taxonomy_{2}$, then we must add internal nodes to $\taxonomy_{2}$ to
-align it to $\taxonomy_{1}$. However, if taxonomy $\taxonomy_{2}$
+align it to $\taxonomy_{1}$.
+However, if taxonomy $\taxonomy_{2}$
 contains more leaves than $\taxonomy_{1}$, then it is unclear if these
 extra leaves should be nested inside the additional internal nodes, or
-not. Thus, the extra leaves are marked \emph{incertae sedis}.
+not.
+Thus, the extra leaves are marked \emph{incertae sedis}.
 
@@
 should be nested within $x$, but we do not know if $e$ should be
-nested within $x$ or not. Thus we obtain $((a,b)x,(c,d)y,?e)z$, where
+nested within $x$ or not.
+Thus we obtain $((a,b)x,(c,d)y,?e)z$, where
 $?$ indicates that taxon $e$ is marked \emph{incertae sedis}.
@@
 sedis taxa, then such an approach does not work if $y$ is already
-incertae sedis. However, we take the approach that \emph{incertae
+incertae sedis.
+However, we take the approach that \emph{incertae
 sedis} notations are not a general solution to expression of uncertain
@@
 The splits-based semantics for incertae sedis taxa has many desirable
-properties. In allows the use of the BUILD algorithm to assess
-compatibility of split sets. However, it also has some properties that
-may not be expected. For example, two incertae sedis siblings can be
-freely interdigitated, which might not be the expected outcome. The
-current definition, also ignore ranks. Thus if we had two incertae
+properties.
+In allows the use of the BUILD algorithm to assess
+compatibility of split sets.
+However, it also has some properties that
+may not be expected.
+For example, two incertae sedis siblings can be
+freely interdigitated, which might not be the expected outcome.
+The
+current definition, also ignore ranks.
+Thus if we had two incertae
 sedis siblings that were genera, but other siblings were families,
 then one might expect that the genus-level incertae sedis taxa could
-be placed within families, but not intermixed. Since the current
+be placed within families, but not intermixed.
+Since the current
 approach ignores ranks, it cannot do that.
@@
 The above definition of incertae sedis taxa does allow two incertae
-sedis sister taxa to be interdigitated. It also
+sedis sister taxa to be interdigitated.
+It also
 
 Also, the splits-based semantics might not completely recover the
-edit-operation semantics. The corner cases are ... .
+edit-operation semantics.
+The corner cases are ...
+.
 
@@
 In theory it could be possible for an \emph{incertae sedis} taxon $A$
-to be placed within an \emph{incertae sedis} sibling $B$. We initially
-implemented this semantics. It requires updating the recursion to
+to be placed within an \emph{incertae sedis} sibling $B$.
+We initially
+implemented this semantics.
+It requires updating the recursion to
 define $I(m,n)$ as the descendants of $m$ that can be access from $n$
@@
 $A$ and $B$, then this has the unfortunate consequence that the
-members of $A$ and $B$ may be inter-digitated. In many cases,
+members of $A$ and $B$ may be inter-digitated.
+In many cases,
 \emph{incertae sedis} genera have siblings that are families, and the
 expected semantics is that genera may be placed within their sibling
-families, or may remain outside existing families. In such cases,
-inter-digitating the genera would not be expected behavior. Thus, we
+families, or may remain outside existing families.
+In such cases,
+inter-digitating the genera would not be expected behavior.
+Thus, we
 choose the semantics that \emph{incertae sedis} taxa exclude their
-\emph{incertae sedis} siblings. This is not because the other
-semantics is impossible. However, we note that use of the
+\emph{incertae sedis} siblings.
+This is not because the other
+semantics is impossible.
+However, we note that use of the
 interdigitating semantics allows a situation where the names for $A$
@@
 This semantics also complicates placement, since it is possible to
-place $A$ as a sibling of $B$, and then place $B$ within $A$. This
+place $A$ as a sibling of $B$, and then place $B$ within $A$.
+This
 violates the current invariant that taxa are only ever placed within
-more tipward taxa. This makes the description of placement more
+more tipward taxa.
+This makes the description of placement more
 complicated, since we can no longer simply record all cases where a
@@
 The term \emph{incertae sedis} can be used in a number of different
-ways. For example,\emph{ }one author might describe \emph{incertae
+ways.
+For example,\emph{ }one author might describe \emph{incertae
 sedis} genera that are asserted to be monophyletic, whereas another

### 2018-04-20 doc/incertae_sedis.tex

--- 2018-04-20 doc/incertae_sedis.tex
@@
 
-\author{Benjamin D. Redelings and Mark T. Holder}
-
+\author{Benjamin D.~Redelings and Mark T.~Holder} 
 \maketitle
@@
 The supertree method described by \citet{redelings2017supertree} is
+
 able to use a set of phylogenetic estimates and taxonomy to produce a
-supertree with clades names according to the taxonomy. However, the
+supertree with clades names according to the taxonomy.
+However, the
 method was not able to accommodate the fact that taxonomies frequently
-contain nodes with uncertain placement. These nodes are often labelled
+contain nodes with uncertain placement.
+These nodes are often labelled
 ``\emph{incertae sedis}'', which means ``uncertain seat'' in Latin.
@@
 as ``\emph{incertae familia}'', indicating that it is unclear which
-family the genus should be placed in. \emph{Incertae sedis} taxa
+family the genus should be placed in.
+\emph{Incertae sedis} taxa
 frequently occur when a specimen is identified down to a given
-taxonomic level, but no further. Extinct taxa are often \emph{incertae
-sedis}. OTT is constructed from several source taxonomies, and these
+taxonomic level, but no further.
+Extinct taxa are often \emph{incertae
+sedis}.
+OTT is constructed from several source taxonomies, and these
 sources include various ways of indicating that the position of a
-taxon within the hierarchy is uncertain. OTT uses a series of
-``flags'' to annotate these taxa. The supertree of by
+taxon within the hierarchy is uncertain.
+OTT uses a series of
+``flags'' to annotate these taxa.
+The supertree of by
 \citet{redelings2017supertree} simply pruned these taxa from the
-taxonomy and input trees. Thus, the final supertree lacked taxa which
+taxonomy and input trees.
+Thus, the final supertree lacked taxa which
 were \emph{incertae sedis} or the descendants of such taxa.
@@
 uncertainty by using published phylogenies to place \emph{incertae
-sedis} taxa. Developing this method required deciding on a set
+sedis} taxa.
+Developing this method required deciding on a set
 operational semantics to be used when interpreting the \emph{incertae
-sedis} annotation. These semantics affect the interpretation of the
+sedis} annotation.
+These semantics affect the interpretation of the
 phylogenetic statements being made by the input taxonomy and the rules
-for applying taxonomic names to the final supertree. In addition to
+for applying taxonomic names to the final supertree.
+In addition to
 discussing how to interpret the \emph{incertae sedis} annotation,
 we describe the algorithmic changes to the supertree pipeline that
-were required to adequately represent the taxonomic uncertainty. The
+were required to adequately represent the taxonomic uncertainty.
+The
 approach described here allows us to include thousands of new taxa in
-our supertree analysis that were previously filtered out. It will also
+our supertree analysis that were previously filtered out.
+It will also
 enable the Open Tree of Life project to include extinct taxa, since
-many of these taxa are \emph{incertae sedis}. This makes substantial
+many of these taxa are \emph{incertae sedis}.
+This makes substantial
 progress towards our goal of comprehensive inclusion of known taxa
@@
 approximation approach that relies on decomposition of the full
-problem into subproblems. The decomposition partitions the list of
+problem into subproblems.
+The decomposition partitions the list of
 taxonomy splits $S(\taxonomy_{p})$ into two lists:
@@
 In practice, the decomposition allows the compatibility of splits to
-be checked on smaller, relabeled trees. This can be done much more
+be checked on smaller, relabeled trees.
+This can be done much more
 efficiently than if compatibility was checked using the BUILD
-algorithm of \citet{AhoSSU1981} on the full leaf set. Thus, the
+algorithm of \citet{AhoSSU1981} on the full leaf set.
+Thus, the
 supertree algorithm first divides the supertree problem into
@@
 
-The taxonomy tree, $\taxonomy_{p}$, is derived from OTT. As mentioned
+The taxonomy tree, $\taxonomy_{p}$, is derived from OTT.
+As mentioned
 above, OTT not only contains a taxonomic hierarchy, but also contains
 taxonomic labels and annotations (referred to as ``flags'') on the
-taxa to carry extra information. The entire hiearchy encoded by OTT is
+taxa to carry extra information.
+The entire hiearchy encoded by OTT is
 a complete taxonomic tree, $\taxonomy^{\ast}.$ Nodes in
 $\taxonomy^{\ast}$ have a one-to-one correspondence with a set of
-labels $\mathcal{N}$ that are called taxon names. We will decompose
+labels $\mathcal{N}$ that are called taxon names.
+We will decompose
 $\mathcal{N}$ into the taxon names for leaf taxa \emph{$\mathcal{L}$
-}and higher taxa $\mathcal{H}$. We will use $\mathcal{U}$ to refer to
+}and higher taxa $\mathcal{H}$.
+We will use $\mathcal{U}$ to refer to
 the subset ($\mathcal{U}\subseteq\mathcal{N}$) of taxa that are
-annotated with the \emph{incertae sedis }property. Other annotations
+annotated with the \emph{incertae sedis }property.
+Other annotations
 identify dubious taxa and taxonomic names that are unwanted artifacts
-of the process of building OTT. Thus, the taxonomic tree which is used
-as an input to the supertree algorithm is a pruned version of
-$\taxonomy^{\ast}$. For our purposes, the relevant information from
+of the process of building OTT.
+Thus, the taxonomic tree which is used
+as an input to the supertree algorithm is pruned version of
+$\taxonomy^{\ast}$.
+For our purposes, the relevant information from
 the taxonomy consists of the taxonomy tree $\taxonomy^{\ast}$, the
-labels $\mathcal{N}$, and set $\mathcal{U}$ of \emph{incertae sedis
-}taxa. In previous work, taxa in $\mathcal{U}$ and all of their descendants were all pruned from
-the taxonomic hierarchy when $\taxonomy_{p}$ is extracted from OTT. In
+labels $\mathcal{N}$, and set $\mathcal{U}$ of \emph{incertae sedis} taxa.
+In previous work, taxa in $\mathcal{U}$ and all of their descendants 
+were all pruned from
+the taxonomic hierarchy when $\taxonomy_{p}$ is extracted from OTT.
+In
 the present work, we consider supertree approaches that operate on a
@@
 However, frequently only a subset of the possible placement points
-would be viewed as plausible. Ideally, the \emph{incertae sedis}
-annotation for some taxon \emph{x} would be accompanied by a specific
-statement about which other taxa might include \emph{x}. Detailed
+would be viewed as plausible.
+Ideally, the \emph{incertae sedis}
+annotation some taxon $x$ would be accompanied by a specific
+statement about which other taxa might include $x$.
+Detailed
 readings of the character argumentation used by taxonomists could
 provide this sort of specific information in many cases, but we lack
-such information on the scale of the entire OTT. Here we consider a
+such information on the scale of the entire OTT.
+Here we consider a
 few operational approaches for handling the \emph{incertae sedis}
@@
 
-In principle this information is easy to incorporate. If
+In principle this information is easy to incorporate.
+If
 $\mathcal{P}$ is the set of taxa that are potential ancestors for
@@
 However, in the \emph{incertae sedis} framework, it is possible for
-one name to apply to multiple nodes. For example, in Figure \ref{fig:One-name-can}b,
-the name $A$ can apply to nodes $x$ and $y$. Here the
+one name to apply to multiple nodes.
+For example, in Figure \ref{fig:One-name-can}b, 
+the name $A$ can apply to nodes $x$ and $y$.
+Here the
 name $A$ corresponds to the split $A1\,A2\,|\bullet B\,C$, leaving out
-$D$ since it is \emph{incertae sedis}. The two nodes $x$ and $y$
+$D$ since it is \emph{incertae sedis}.
+The two nodes $x$ and $y$
 display the splits $A1\,A2\,|\bullet B\,C\,D$ and $A1\,A2\,D|\bullet
@@
 like to interpret the position of these names in terms of placing
-\emph{incertae sedis }taxa in a revised taxonomy. This would allow us
+\emph{incertae sedis }taxa in a revised taxonomy.
+This would allow us
 to interpret the synthesis tree as saying that phylogenetic
 information has (for example) placed genus A within family B, or
-perhaps outside of all named families. The simplest approach to
+perhaps outside of all named families.
+The simplest approach to
 placement involves noting whenever a taxon $B$ is a descendant of a
-taxon $A$ on the named synthesis tree but not the taxonomy tree. For
+taxon $A$ on the named synthesis tree but not the taxonomy tree.
+For
 example, in Figure \ref{fig:Handling-incertae-sedis}, taxon $C$ is a
-descendant of taxon $A$ in the synthetic tree (Fig. \ref{fig:Handling-incertae-sedis}d),
-but not the taxonomy (Fig. \ref{fig:Handling-incertae-sedis}b).
+descendant of taxon $A$ in the synthetic tree 
+(Fig.\ref{fig:Handling-incertae-sedis}d), but not the taxonomy 
+(Fig.\ref{fig:Handling-incertae-sedis}b).
 Thus, we could say that the synthetic tree places
@@
 
-A slightly more general case is shown in Figure \ref{fig:Broken-incertae-sedis}.
+A slightly more general case is shown in 
+Figure \ref{fig:Broken-incertae-sedis}.
 Here taxon $C1$ is an descendant of $A$ in the
-synthesis tree, but not the taxonomy. Thus $C1$ is placed within $A$.
+synthesis tree, but not the taxonomy.
+Thus $C1$ is placed within $A$.
 We note that the most recent common ancestor (MRCA) of $A$ and $C1$ in
@@
 However, this assumes that \emph{incertae sedis} taxa are always
-placed within their siblings, or descendants of their siblings. As
+placed within their siblings, or descendants of their siblings.
+As
 such, it does not handle cases such as Figure \ref{fig:Placing-one-incertae}, 
 in which $I$ is first placed as sibling to $AI$, and then
-$AI$ is placed within $I$. Note that $I$ is not a sibling of $AI$ or a
-descendant of a sibling. Furthermore, the MRCA of $I$ and $AI$ is the
+$AI$ is placed within $I$.
+Note that $I$ is not a sibling of $AI$ or a
+descendant of a sibling.
+Furthermore, the MRCA of $I$ and $AI$ is the
 root, and the path from $AI$ to the root includes $A$, but $A$ is not
-broken. We suggest that this conundrum may be resolved by first
+broken.
+We suggest that this conundrum may be resolved by first
 processing the placement of $I$ within $A$ by modifying the taxonomy
 to reflect this placement\footnote{Does this actually work? Some
-placements break taxa.}. When we consider the placement of $AI$ within
+placements break taxa.}.
+When we consider the placement of $AI$ within
 $I$ on this modified taxonomy, the MRCA of $I$ and $AI$ is now $A$
-instead of the root, and the correct interpretation results. We
+instead of the root, and the correct interpretation results.
+We
 suggest that by walking the tree in a pre-order fashion and processing
@@
 \emph{incertae sedis} taxa into containing taxa without breaking the
-containing taxa. We change the semantics of names to imply, not the
+containing taxa.
+We change the semantics of names to imply, not the
 exclusion of all non-included taxa, but only some non-included taxa,
-as described in section \ref{sec:Semantics-of-incertae}. As a result
+as described in section \ref{sec:Semantics-of-incertae}.
+As a result
 of this change in semantics, placing an IS taxon $A$ within a sister
-taxon $B$ no longer results in conflict with $B$. This allows us to
+taxon $B$ no longer results in conflict with $B$.
+This allows us to
 retain the split for $B$ within the synthesis tree, so that taxonomy-
 only children of $B$ are correctly grouped with their siblings that
-are referenced by the input trees. We may then retain the name for the
-no-longer-broken taxon. Finally, we are then able to stop filtering
+are referenced by the input trees.
+We may then retain the name for the
+no-longer-broken taxon.
+Finally, we are then able to stop filtering
 \emph{incertae sedis} taxa, so that they appear in the synthesis tree.
@@
 {[}\textbf{BDR:} \emph{maybe we should merge this with Figure.
-\ref{fig:Broken-incertae-sedis}. Although the point of figure
+\ref{fig:Broken-incertae-sedis}.
+Although the point of figure
 \ref{fig:placement-example1} is to talk about sub-problem
@@
 The addition of \emph{incertae sedis} taxa allows new types of
-conflict between input trees. For example, different input trees might
-place an incertae sedis taxon in conflicting locations. This is
-illustrated in Figure \ref{fig:placement-example1}, where the IS
+conflict between input trees.
+For example, different input trees might
+place an incertae sedis taxon in conflicting locations.
+This is
+illustrated in Figure \ref{fig :placement-example1}, where the IS
 taxon (ott7,ott6)ott10 is placed as sister to ott1 by phylogeny

### 2018-04-20 doc/incertae_sedis.tex
minor

@@
 Thus, one could imagine replacing that taxonomy,
-$\taxonomy_{p}$, with a set of trees $\mathcal{R}(\taxonomy_{p})$ each
-of which has a single internal edge such that each edge in
-$\taxonomy_{p}$is included in one tree in
+$\taxonomy_{p}$, with a set of rooted splits $\mathcal{R}(\taxonomy_{p})$ 
+such that each edge in
+$\taxonomy_{p}$ is encoded in one split in
 $\mathcal{R}(\taxonomy_{p})$.
-If the order of trees is set to be (with
-some abuse of notation):
+If the order of trees is set to be:
 $\mathcal{R}(\taxonomy_{p})=\mathcal{Z}(\taxonomy_{p}) + 
 \mathcal{Y}(\taxonomy_{p})$, then using $\mathcal{R}(\taxonomy_{p})$ in
-place of $\taxonomy_{p}$in the supertree algorithm would produce the
+place of $\taxonomy_{p}$ in the supertree algorithm would produce the
 same output.
 
-We define rooted splits for each edge by noting that removing the edge
+As is typically done, we define rooted splits for each edge by
+noting that removing the edge
 would produce two trees.
@@
 $\mathcal{E}(e)  =\mathcal{L}-\mathcal{I}(e)$.
-
 For a node $n$ on the tip-ward side of an edge $e$, we may also write
@@
 
-\paragraph{Information content of the emph{incertae sedis }annotation}
+\paragraph{Information content of the \emph{incertae sedis }annotation}
 
@@
 would be viewed as plausible.
+
 Ideally, the \emph{incertae sedis}
@@
 They may simply be omitted in some rooted split
-representations, resulting in partial splits.
+representations, resulting in ``partial splits'' -- those which
+do not necessarily mention all of the taxa.
+Thus, when interpreted correctly the \emph{incertae sedis} annotations
+are reducing the amount of information that the taxonomy brings to
+the supertree problem.
 
@@
 sibling taxa.
+In essence, this interpretation would overinterpret the taxonomic
+    hierarchy by treating all of the groupings as statements about
+    full rooted splits, when some should be viewed as partial splits.
 

### 2018-04-20 doc/incertae_sedis.tex
context about naming

@@
 
+The interpretation of the \emph{incertae sedis} annotation affects the supertree
+    pipeline at three points:
+\begin{enumerate}
+    \item The subproblem decomposition step relies on identifying uncontested nodes.
+    Thus, varying interpretations of whether or not an \emph{incertae sedis}
+    taxon can be placed within another taxon without contesting that taxon
+    will affect the subproblem decomposition.
+    \item When solving each subproblem, how taxon is represented as a rooted
+    split can affect the solution found.
+    As described below, the \emph{incertae sedis} annotation affects the 
+        exclude groups of rooted split encodings.
+    \item The step of assigning names to the supertree is affected by the 
+        operational phylogenetic definition of each name.
+\end{enumerate}
+
 \paragraph{Equivalence of a taxonomy tree and a list of rooted splits}
@@
 
-\paragraph{Information content of the \emph{incertae sedis }annotation}
+In terms of phylogenetic nomenclature the supertree pipeline acts as
+    the taxonomy defines a branch-based name \citep[see][]{deQueiroz2013}
+    for each higher taxon of the form.
+So for name labeling node $n$ in the taxonomic tree, 
+    the operational definition would be:
+    ``the clade defined by all taxa
+    that are more closely related to $\mathcal{I}(n)$ than to
+    $\mathcal{E}(n)$, if such a clade exists on the tree.''
+
+\paragraph{Information content of the \emph{incertae sedis} annotation}
 

### 2018-04-20 doc/incertae_sedis.tex
consolidating similar points

@@
 
-Ideally, the \emph{incertae sedis}
-annotation some taxon $x$ would be accompanied by a specific
-statement about which other taxa might include $x$.
-Detailed
-readings of the character argumentation used by taxonomists could
-provide this sort of specific information in many cases, but we lack
-such information on the scale of the entire OTT.
-Here we consider a
-few operational approaches for handling the \emph{incertae sedis}
-label.
-
 In terms of rooted split interpretation, an \emph{incertae sedis}
@@
 In principle this information is easy to incorporate.
-If
-$\mathcal{P}$ is the set of taxa that are potential ancestors for
+Ideally, the \emph{incertae sedis}
+annotation some taxon $x$ would be accompanied by a specific
+statement about which other taxa might include $x$.
+If $\mathcal{P}$ is the set of taxa that are potential ancestors for
 incertae sedis taxon $x$, then $\mathcal{L}(x)$ will occur in neither
@@
 Unfortunately, utilizing taxon-specific interpretations of the
-incertae sedis flag relies on character argumentation (which is
-lacking in OTT) or rank-based information (which is frequently
+incertae sedis flag relies on detailed readings of character argumentation
+(which is lacking in OTT) or rank-based information (which is frequently
 untrustworthy in OTT).

### 2018-04-20 doc/incertae_sedis.tex
correction to rank-based i.s.

@@
 unplaced, but only a few placements within the taxonomy are possible.
-For example, a genus may not be characterized as incertae sedis at the
-level of the subfamily.
+For example, a genus might be characterized as \emph{incertae sedis}
+at the level of the subfamily.
 Thus, it could be placed inside any of the
 hypothesized sub-families, or within its own subfamily, but it would
-be unexpected for it to be placed inside one of the other genera
-within a family.
+be unexpected for it to be placed inside one of the other genera.
 

### 2018-04-20 doc/incertae_sedis.tex
restoring \
Fix last undefined reference.

@@
 
-\subsection{Split-based semantics for emph{incertae sedis} taxa}
+\subsection{Split-based semantics for \emph{incertae sedis} taxa}
 
@@
 This is
-illustrated in Figure \ref{fig :placement-example1}, where the IS
+illustrated in Figure \ref{fig:placement-example1}, where the IS
 taxon (ott7,ott6)ott10 is placed as sister to ott1 by phylogeny
@@
 
-\subsubsection{Exclude sets modified by emph{incertae sedis} marks}
+\subsubsection{Exclude sets modified by \emph{incertae sedis} marks}
 

### 2018-04-21 doc/incertae_sedis.tex
Added clarifying comments on the alternative semantics.

@@
 
-\subsubsection{A semantics that disallows inter-digitating
+\subsubsection{semantics \#1a: The original, simplest semantics}
+\begin{align}
+    \mathcal{E}(n) & = \mathcal{E}(parent(n))\cup\left[\mathcal{I}(s)\big|s\in
+           siblings(n), s \text{ not \emph{incertae sedis}}\right].\label{eq:exclude-set-formula-1a}
+\end{align}
+This semantics \emph{does} allow interdigitation.
+
+\subsubsection{semantics \#1b: modified to disallow inter-digitating
 \emph{incertae sedis} siblings}
@@
 
-\subsubsection{An alternative semantics that allows inter-digitating
-incertae sedis siblings}
+\subsubsection{An alternative semantics that allows placing an incertae sedis group within an incertae-sedis great-uncle}
+
+\bdrcomment{This semantics is actually designed to allow the situation in Figure \ref{fig:Placing-one-incertae}.
+
+  The problem is (might be) as follows: if we have the tree ``((?A1,A2)A,((D)C)B)MRCA;'', where A1 is incertae sedis, then we can place A1 within D.
+This semantics does not exclude A1 from B, C, or D.
+So, if A is broken, then nothing prohibits A1 from intruding into B, C, D, etc.
+
+  The purpose of this was to allow placing B as a siblign to A1 and then placing A1 within B -- \emph{if} B were incertae sedis.  But B is not incertae sedis.
+
+  We could try to fix up this situation by adding cases to the exclude rules to say that B excludes A1 if B is not incertae sedis.
+But adding cases makes things messier.
+I think it also does not actually solve the problem: suppose we mark B incertae sedis, but do NOT place it as sibling to A1.
+Then again if A is broken, A1 can intrude into B.
+We really want to allow A1 into B only if B is placed as sibling to A1, and I'm not sure how to do that with the partial-split-based semantics.
+If you see a way, then great!
+}
 
@@
 \subsubsection{A third semantics}
+
+\bdrcomment{I didn't really finish working this one out.}
 

### 2018-04-24 doc/incertae_sedis.tex
Tiny edits.

@@
 taxon sets into a single tree on the combined taxon set.
-The methods are usually applied to leaf-labeled phylogenetic estimates. 
+These methods are usually applied to leaf-labeled phylogenetic estimates. 
 Here we describe a supertree approach that overcomes issues associated with
@@
 The supertree method described by \citet{redelings2017supertree} is
-
 able to use a set of phylogenetic estimates and taxonomy to produce a
@@
 ``flags'' to annotate these taxa.
-The supertree of by
-\citet{redelings2017supertree} simply pruned these taxa from the
+The supertree of \citet{redelings2017supertree} simply pruned these taxa from the
 taxonomy and input trees.
@@
 sedis} taxa.
-Developing this method required deciding on a set
+Developing this method required deciding on a set of
 operational semantics to be used when interpreting the \emph{incertae
@@
 The BUILD
-algorithm\citep{AhoSSU1981} is used both to check compatibility, and
+algorithm \citep{AhoSSU1981} is used both to check compatibility, and
 to construct a compatible tree from the final set $C$.
@@
 taxa to carry extra information.
-The entire hiearchy encoded by OTT is
+The entire hierarchy encoded by OTT is
 a complete taxonomic tree, $\taxonomy^{\ast}.$ Nodes in
@@
 \emph{incertae sedis} taxa from the taxonomy was also accompanied by
-pruning those taxa from the input phylogenetic trees.
+pruning those taxa and all of their descendants from the input phylogenetic
+trees.
 
@@
 In terms of phylogenetic nomenclature the supertree pipeline acts as
-    the taxonomy defines a branch-based name \citep[see][]{deQueiroz2013}
+if the taxonomy defines a branch-based name \citep[see][]{deQueiroz2013}
     for each higher taxon of the form.
-So for name labeling node $n$ in the taxonomic tree, 
+So for the name labeling node $n$ in the taxonomic tree, 
     the operational definition would be:

### 2018-04-24 doc/incertae_sedis.tex
Reorder and reformat.

@@
 
-The interpretation of the \emph{incertae sedis} annotation affects the supertree
-    pipeline at three points:
-\begin{enumerate}
-    \item The subproblem decomposition step relies on identifying uncontested nodes.
-    Thus, varying interpretations of whether or not an \emph{incertae sedis}
-    taxon can be placed within another taxon without contesting that taxon
-    will affect the subproblem decomposition.
-    \item When solving each subproblem, how taxon is represented as a rooted
-    split can affect the solution found.
-    As described below, the \emph{incertae sedis} annotation affects the 
-        exclude groups of rooted split encodings.
-    \item The step of assigning names to the supertree is affected by the 
-        operational phylogenetic definition of each name.
-\end{enumerate}
-
-\paragraph{Equivalence of a taxonomy tree and a list of rooted splits}
-
-Note that, for the purposes of the supertree method, all of the
-information from the taxonomic tree is contained in the set of rooted
-splits.
-Thus, one could imagine replacing that taxonomy,
-$\taxonomy_{p}$, with a set of rooted splits $\mathcal{R}(\taxonomy_{p})$ 
-such that each edge in
-$\taxonomy_{p}$ is encoded in one split in
-$\mathcal{R}(\taxonomy_{p})$.
-If the order of trees is set to be:
-$\mathcal{R}(\taxonomy_{p})=\mathcal{Z}(\taxonomy_{p}) + 
-\mathcal{Y}(\taxonomy_{p})$, then using $\mathcal{R}(\taxonomy_{p})$ in
-place of $\taxonomy_{p}$ in the supertree algorithm would produce the
-same output.
-
-As is typically done, we define rooted splits for each edge by
-noting that removing the edge
-would produce two trees.
-The leaf label sets of these trees the
-partition the full leaf set $\mathcal{L}$ into two groups.
-The include
-set $\mathcal{I}(e)$ which does not contain the root, and the exclude
-set $\mathcal{E}(e)$ which does contain the root.
-Such a split may be
-written \[ \mathcal{I}(e)|\bullet\mathcal{E}(e).\] where the
-$\bullet$ indicates the root.
-The include set is the set of leaves
-that are descendants of the daughter node of the edge.
-The exclude set
-is the set of leaves that are not descendants of that node.
-If no taxa
-are \emph{incertae sedis}, then the exclude set for a node is just the
-total tip set minus the include set for the node:
-$\mathcal{E}(e)  =\mathcal{L}-\mathcal{I}(e)$.
-For a node $n$ on the tip-ward side of an edge $e$, we may also write
-$\mathcal{I}(n)$ for $\mathcal{I}(e)$, and $\mathcal{E}(n)$ for
-$\mathcal{E}(e)$.
-For any two rooted splits $A=A_{1}|\bullet A_{2}$
-and $B=B_{1}|\bullet B_{2}$ , we say that $A$ \emph{displays }$B$ if
-$B_{1}\subseteq A_{1}$ and $B_{2}\subseteq A_{2}$.
-
-Note that when a taxon rooted at node $n$ is not marked as
-\emph{incertae sedis}, then the leaf labels that are descendants of
-$n$ will be in the include set for $n$ and all its ancestors and will
-be in the exclude set for all other nodes/edges.
-
-In terms of phylogenetic nomenclature the supertree pipeline acts as
-if the taxonomy defines a branch-based name \citep[see][]{deQueiroz2013}
-    for each higher taxon of the form.
-So for the name labeling node $n$ in the taxonomic tree, 
-    the operational definition would be:
-    ``the clade defined by all taxa
-    that are more closely related to $\mathcal{I}(n)$ than to
-    $\mathcal{E}(n)$, if such a clade exists on the tree.''
-
-\paragraph{Information content of the \emph{incertae sedis} annotation}
-
 In order to incorporate \emph{incertae sedis }taxa into a supertree
@@
 would be viewed as plausible.
+
+We seek a semantics for supertrees with \emph{incertae sedis }taxa
+that satisfy the following properties:
+\begin{enumerate}
+    \item An \emph{incertae sedis} node may intrude into its siblings and their
+descendants.
+    \item The semantics of an \emph{incertae sedis} node does not depend on assigned ranks, but only on the tree.
+    \item Semantics is based on deriving a rooted (partial) split for each branch of the taxonomy.
+      % These could be goals, but maybe discuss them as side-effects:
+      % Side effect #1. Don't interdigitate
+      % Side effect #2. Allow placing A next to B, and then placing B within A. (See Fig. 6)
+\end{enumerate}
+
+The rooted split derived from a node $n$ serves to determine both when $n$ is in conflict
+with an input tree, and to where to place the taxon name for $n$ on the resulting synthesis tree.
+% Removed mention of incertae sedis possibly affecting subproblem decomposition.  It
+% actually doesn't but that is surprising, and therefore kind of a result.
+
+\paragraph{Equivalence of a taxonomy tree and a list of rooted splits}
+
+Note that, for the purposes of the supertree method, all of the information from
+the taxonomic tree is contained in the set of rooted splits. Thus, one could
+imagine replacing that taxonomy, $\taxonomy_{p}$, with a set of rooted splits $\mathcal{R}(\taxonomy_{p})$ such that
+each edge in $\taxonomy_{p}$ is encoded in one split in $\mathcal{R}(\taxonomy_{p})$. If the order of trees is set
+to be: $\mathcal{R}(\taxonomy_{p})=\mathcal{Z}(\taxonomy_{p}) + 
+\mathcal{Y}(\taxonomy_{p})$, then using $\mathcal{R}(\taxonomy_{p})$ in place of $\taxonomy_{p}$ in the supertree
+algorithm would produce the same output.
+
+As is typically done, we define rooted splits for each edge by noting that
+removing the edge would produce two trees. The leaf label sets of these trees
+the partition the full leaf set $\mathcal{L}$ into two groups. The include set $\mathcal{I}(e)$ which
+does not contain the root, and the exclude set $\mathcal{E}(e)$ which does contain the root.
+Such a split may be written
+\[ \mathcal{I}(e)|\bullet\mathcal{E}(e).\]
+where the $\bullet$ indicates the root. The include set is the set of leaves that are
+descendants of the daughter node of the edge. The exclude set is the set of
+leaves that are not descendants of that node. If no taxa are \emph{incertae sedis},
+then the exclude set for a node is just the total tip set minus the include set
+for the node: $\mathcal{E}(e)  =\mathcal{L}-\mathcal{I}(e)$. For a node $n$ on the tip-ward side of an edge $e$, we
+may also write $\mathcal{I}(n)$ for $\mathcal{I}(e)$, and $\mathcal{E}(n)$ for $\mathcal{E}(e)$. For any two rooted splits $A=A_{1}|\bullet A_{2}$
+and $B=B_{1}|\bullet B_{2}$ , we say that $A$ \emph{displays }$B$ if $B_{1}\subseteq A_{1}$ and $B_{2}\subseteq A_{2}$.
+
+Note that when a taxon rooted at node $n$ is not marked as \emph{incertae sedis},
+then the leaf labels that are descendants of $n$ will be in the include set for
+$n$ and all its ancestors and will be in the exclude set for all other nodes/edges.
+
+In terms of phylogenetic nomenclature the supertree pipeline acts as if the
+taxonomy defines a branch-based name \citep[see][]{deQueiroz2013} for each
+higher taxon of the form. So for the name labeling node $n$ in the taxonomic
+tree, the operational definition would be: ``the clade defined by all taxa that
+are more closely related to $\mathcal{I}(n)$ than to $\mathcal{E}(n)$, if such a clade exists on the tree.''
+
+\paragraph{Information content of the \emph{incertae sedis} annotation}
 

### 2018-04-24 doc/incertae_sedis.tex
Reformating to be readable with preview-latex.

@@
 
-In terms of rooted split interpretation, an \emph{incertae sedis}
-taxon differs from a normal taxon because its leaves should not
-automatically be mentioned in the exclude set of all of the edges that
-does not descend from.
-If a taxon rooted at node $n$ is marked as
-\emph{incertae sedis}, then the leaf labels that are descendants of
-$n$ will still be in the include set for $n$ and all its ancestors,
+In terms of rooted split interpretation, an \emph{incertae sedis} taxon differs
+from a normal taxon because its leaves should not automatically be mentioned in
+the exclude set of all of the edges that does not descend from. If a taxon
+rooted at node $n$ is marked as \emph{incertae sedis}, then the leaf labels that
+are descendants of $n$ will still be in the include set for $n$ and all its ancestors,
 but these labels will not necessarily be the exclude set for all other
-nodes/edges.
-They may simply be omitted in some rooted split
-representations, resulting in ``partial splits'' -- those which
-do not necessarily mention all of the taxa.
-Thus, when interpreted correctly the \emph{incertae sedis} annotations
-are reducing the amount of information that the taxonomy brings to
-the supertree problem.
+nodes/edges. They may simply be omitted in some rooted split representations,
+resulting in ``partial splits'' -- those which do not necessarily mention all of
+the taxa. Thus, when interpreted correctly the \emph{incertae sedis} annotations
+are reducing the amount of information that the taxonomy brings to the supertree
+problem.
 
@@
 
-We seek a semantics for supertrees with \emph{incertae sedis }taxa
-that satisfy the following properties:
-\begin{enumerate}
-    \item An \emph{incertae sedis} node may intrude into its siblings and their
-descendants.
-    \item Two \emph{incertae sedis} siblings may not be
-interdigitated.
-    \item Any placement $T\to T^{*}$ of a tree $T$ with
-\emph{incertae sedis} taxa is also accessible from a tree $T^{\prime}$
-that is created from $T$ by taking a taxon, attaching it to an
-ancestor, and marking it \emph{incertae sedis}. \mthcomment{MTH: I don't understand this.}
-\bdrcomment{BDR: Figure \ref{fig:Placing-one-incertae} is an illustration of this.
-Suppose you have a genus that is incertae sedis and a family that is incertae sedis and attached at the kingdom level.
-We could in theory place the family (.e.g $I$ in the figure) at the right level.
-It then sits next to the incertae sedis genus (e.g.
-$AI$ in the figure).
-We could then place the genus within the family.
-
-The goal above is trying to say (and failing to be clear at saying) that if we have a family that is sibling to an incertae sedis genus, then the placement of the genus within the family is still a possibility if we reattach the family as a child of a kingdom node and mark it incertae sedis.
-}
-    \item The semantics of an
-\emph{incertae sedis} node does not depend on assigned ranks, but only
-on the tree.
-    \item Semantics is based on deriving a split for each
-branch of the tree.
-\end{enumerate}
-Here we focus on a split-based
-semantics instead of a semantics based on apply of a series of edit
-operations to the tree.
-The split-based semantics allows us to seek
-trees that are simultaneously consistent with a set of split-
-constraints.
-The use of a split-based semantics does impose some
-limitations.
-These come primarily because in the edit-based semantics,
-after we place an \emph{incertae sedis} taxon, it is no loner
-\emph{incertae sedis}, whereas with the split-based semantics taxa are
-either always \emph{incertae sedis} or never \emph{incertae sedis}.
-
-This has an effect, for example, when we consider two incertae sedis
-siblings $A$ and $B$.
-If we can place $A$ within $B$, and $B$ within
-$A$, then a split- based semantics allows us to freely inter-digitate
-$A$ and $B$.
-If we do not want to allow $A$ and $B$ to be inter-
-digitated, then we cannot both $A$ within $B$ and $B$ within $A$.
-In
-addition, we can imagine that a family $A$ is incertae sedis within a
-kingdom.
-In an edit-based semantics, we could place $A$ as a sibling
-to a genus $B$, mark $A$ as non-\emph{incertae-sedis}, and then place
-$B$ within $A$.
-In a split-based semantics, we must disallow either
-$B$ within $A$ or $A$ within $B$ if we want to disallow
+Here we focus on a split-based semantics instead of a semantics based on apply
+of a series of edit operations to the tree. The split-based semantics allows us
+to seek trees that are simultaneously consistent with a set of split-constraints.
+The use of a split-based semantics does impose some limitations. These come
+primarily because in the edit-based semantics, after we place an
+\emph{incertae sedis} taxon, it is no loner \emph{incertae sedis}, whereas with
+the split-based semantics taxa are either always \emph{incertae sedis} or never
+\emph{incertae sedis}.
+
+This has an effect, for example, when we consider two incertae sedis siblings $A$
+and $B$. If we can place $A$ within $B$, and $B$ within $A$, then a split- based
+semantics allows us to freely inter-digitate $A$ and $B$. If we do not want to
+allow $A$ and $B$ to be inter-digitated, then we cannot both $A$ within $B$ and $B$
+within $A$. In addition, we can imagine that a family $A$ is incertae sedis within a
+kingdom. In an edit-based semantics, we could place $A$ as a sibling to a genus $B$,
+mark $A$ as non-\emph{incertae-sedis}, and then place $B$ within $A$. In a split-based
+semantics, we must disallow either $B$ within $A$ or $A$ within $B$ if we want to disallow
 interdigitation.
 
-As a result of this, we cannot satisfy all five of these criteria.
-Thus, we describe two alternative split-based semantics.
-The non-
-interdigitation semantics gets \# (1,2,4,5), and the interdigitation
-semantics gets (1,3,4,5).
-This makes the point that each semantics
-results from different choices about which criteria to satisfy.
+As a result of this, we cannot satisfy all five of these criteria. Thus, we
+describe two alternative split-based semantics. The non-interdigitation
+semantics gets \# (1,2,4,5), and the interdigitation semantics gets (1,3,4,5).
+This makes the point that each semantics results from different choices about
+which criteria to satisfy.
 
@@
 
-In order to apply supertree methods to rooted trees with
-\emph{incertae sedis} taxa, we first propose a semantics of
-\emph{incertae sedis }in terms of rooted splits.
-We define rooted
-splits by noting that each edge of a tree divides the tip taxa
-$\mathcal{L}$ into two groups: the include set $\mathcal{I}(e)$ which
-does not contain the root, and the exclude set $\mathcal{E}(e)$ which
-does contain the root.
-Such a split may be written \[
+In order to apply supertree methods to rooted trees with \emph{incertae sedis}
+taxa, we first propose a semantics of \emph{incertae sedis }in terms of rooted
+splits. We define rooted splits by noting that each edge of a tree divides the
+tip taxa $\mathcal{L}$ into two groups: the include set $\mathcal{I}(e)$ which does not contain the root,
+and the exclude set $\mathcal{E}(e)$ which does contain the root. Such a split may be written
+\[
 \mathcal{I}(e)|\bullet\mathcal{E}(e).
-\] where the $\bullet$ indicates
-the root.
-If no taxa are \emph{incertae sedis}, then the exclude set
-for a node is just the total tip set minus the include set for the
+\]
+where the $\bullet$ indicates the root. If no taxa are \emph{incertae sedis}, then the
+exclude set for a node is just the total tip set minus the include set for the
 node: $ \mathcal{E}(n) =\mathcal{L}-\mathcal{I}(n)$.
 
-For a node $n$ on the tip-ward side of an edge $e$, we may also write
-$\mathcal{I}(n)$ for $\mathcal{I}(e)$, and $\mathcal{E}(n)$ for
-$\mathcal{E}(e)$.
+For a node $n$ on the tip-ward side of an edge $e$, we may also write $\mathcal{I}(n)$ for $\mathcal{I}(e)$,
+and $\mathcal{E}(n)$ for $\mathcal{E}(e)$.
 
@@
 \end{align*}
-I think this works...
-but
-need to double-check.
+I think this works... but need to double-check.
 
@@
 
-After constructing a supertree, tip nodes already have names in
-\emph{$\mathcal{L}$}.
-However, we still need to assign higher taxon
-names to internal supertree nodes based on the taxonomy tree in the
-problem.
-Each taxon name $n$ corresponds to a split
-$S(n)=S(n)_{1}|\bullet S(n)_{2}$ on the corresponding branch of the
-taxonomy tree.
-Without \emph{incertae sedis}, such splits are always
-of the form $S(n)_{1}|\bullet\mathcal{L}-S(n)_{1}$, but with
-\emph{incertae sedis} taxa $S_{2}(n)$ may be smaller than
-$\mathcal{L}-S(n)_{1}$.
-
-Without \emph{incertae sedis}, each name applies to at most one node,
-and each node can take at most one name, with the exception of
-monotypic taxa.
-Thus, we may simply search the solution tree for a
-node that has the same cluster $S(n)_{1}$ and apply the name $n$ to
-that node.
-
-However, in the \emph{incertae sedis} framework, it is possible for
-one name to apply to multiple nodes.
-For example, in Figure \ref{fig:One-name-can}b, 
-the name $A$ can apply to nodes $x$ and $y$.
-Here the
-name $A$ corresponds to the split $A1\,A2\,|\bullet B\,C$, leaving out
-$D$ since it is \emph{incertae sedis}.
-The two nodes $x$ and $y$
-display the splits $A1\,A2\,|\bullet B\,C\,D$ and $A1\,A2\,D|\bullet
-B\,C$ respectively, and both of these splits display the split
-$A1\,A2\,|\bullet B\,C$, so the name $A$ can apply to both $x$ and
-$y$.
-This cannot happen without\emph{ incertae sedis} taxa except at
-monotypic nodes.
-When faced with a choice about where to place a name,
-our solution is to find the most tip-ward node where the name can
-apply and attach the name to this node.
-This corresponds to a
-conservative choice about whether \emph{incertae sedis} taxa should be
-added to an existing taxon.
+After constructing a supertree, tip nodes already have names in \emph{$\mathcal{L}$}.
+However, we still need to assign higher taxon names to internal supertree nodes
+based on the taxonomy tree in the problem. Each taxon name $n$ corresponds to a
+split $S(n)=S(n)_{1}|\bullet S(n)_{2}$ on the corresponding branch of the taxonomy tree. Without
+\emph{incertae sedis}, such splits are always of the form $S(n)_{1}|\bullet\mathcal{L}-S(n)_{1}$, but
+with \emph{incertae sedis} taxa $S_{2}(n)$ may be smaller than $\mathcal{L}-S(n)_{1}$.
+
+Without \emph{incertae sedis}, each name applies to at most one node, and each
+node can take at most one name, with the exception of monotypic taxa. Thus, we
+may simply search the solution tree for a node that has the same cluster $S(n)_{1}$
+and apply the name $n$ to that node.
+
+However, in the \emph{incertae sedis} framework, it is possible for one name to
+apply to multiple nodes. For example, in Figure \ref{fig:One-name-can}b, the
+name $A$ can apply to nodes $x$ and $y$. Here the name $A$ corresponds to the split
+$A1\,A2\,|\bullet B\,C$, leaving out $D$ since it is \emph{incertae sedis}. The two nodes $x$
+and $y$ display the splits $A1\,A2\,|\bullet B\,C\,D$ and $A1\,A2\,D|\bullet
+B\,C$ respectively, and both
+of these splits display the split $A1\,A2\,|\bullet B\,C$, so the name $A$ can apply to both
+$x$ and $y$. This cannot happen without\emph{ incertae sedis} taxa except at
+monotypic nodes. When faced with a choice about where to place a name, our
+solution is to find the most tip-ward node where the name can apply and attach
+the name to this node. This corresponds to a conservative choice about whether
+\emph{incertae sedis} taxa should be added to an existing taxon.
 \begin{figure}
@@
 
-It is also possible for multiple names to apply to a single node.
-For
-example, in Figure \ref{fig:Multiple-names-can}a, the taxon $A$
-corresponds to the split $B1\,B2\,C\,|\bullet\,Y$, and its child taxon
-$B$ corresponds to the split $B1\,B2\,|\bullet\,Y$.
-The edge leading
-to $A$ is consistent with the split for $B$, but the name $B$ is
-applied to the node with the smallest include group.
-However, in the
-solution tree (Fig.
-\ref{fig:Multiple-names-can}b), there is is only
-node for both names $A$ and $B$ to apply to.
-In this case, we solve
-this problem by introducing a monotypic parent, and applying $A$ to
-the newly created parent node (Fig.
-\ref{fig:Multiple-names-can}c).
+It is also possible for multiple names to apply to a single node. For example,
+in Figure \ref{fig:Multiple-names-can}a, the taxon $A$ corresponds to the split
+$B1\,B2\,C\,|\bullet\,Y$, and its child taxon $B$ corresponds to the split $B1\,B2\,|\bullet\,Y$. The
+edge leading to $A$ is consistent with the split for $B$, but the name $B$ is
+applied to the node with the smallest include group. However, in the
+solution tree (Fig. \ref{fig:Multiple-names-can}b), there is is only node for
+both names $A$ and $B$ to apply to. In this case, we solve this problem by
+introducing a monotypic parent, and applying $A$ to the newly created parent node
+(Fig. \ref{fig:Multiple-names-can}c).
 \begin{figure}
@@
 \end{figure}
-
-\emph{}
 

### 2018-04-24 doc/incertae_sedis.tex
Remove duplicate definition of rooted splits.

@@
 taxa, we first propose a semantics of \emph{incertae sedis }in terms of rooted
-splits. We define rooted splits by noting that each edge of a tree divides the
-tip taxa $\mathcal{L}$ into two groups: the include set $\mathcal{I}(e)$ which does not contain the root,
-and the exclude set $\mathcal{E}(e)$ which does contain the root. Such a split may be written
-\[
-\mathcal{I}(e)|\bullet\mathcal{E}(e).
-\]
-where the $\bullet$ indicates the root. If no taxa are \emph{incertae sedis}, then the
-exclude set for a node is just the total tip set minus the include set for the
-node: $ \mathcal{E}(n) =\mathcal{L}-\mathcal{I}(n)$.
-
-For a node $n$ on the tip-ward side of an edge $e$, we may also write $\mathcal{I}(n)$ for $\mathcal{I}(e)$,
-and $\mathcal{E}(n)$ for $\mathcal{E}(e)$.
-
-For any two rooted splits $A=A_{1}|\bullet A_{2}$ and $B=B_{1}|\bullet
-B_{2}$ , we say that $A$ \emph{displays }$B$ if $B_{1}\subseteq A_{1}$
-and $B_{2}\subseteq A_{2}$.
+splits.
 

### 2018-04-24 doc/incertae_sedis.tex
Some rearranging, updating and more reformatting.

@@
 
-The rooted split derived from a node $n$ serves to determine both when $n$ is in conflict
-with an input tree, and to where to place the taxon name for $n$ on the resulting synthesis tree.
+The rooted split associated with taxonomy node $n$ serves to determine both when $n$ is in conflict
+with an input tree, and to where to place the taxon name for clade $n$ on the synthesis tree.
 % Removed mention of incertae sedis possibly affecting subproblem decomposition.  It
@@
 
-\paragraph{Equivalence of a taxonomy tree and a list of rooted splits}
-
-Note that, for the purposes of the supertree method, all of the information from
-the taxonomic tree is contained in the set of rooted splits. Thus, one could
-imagine replacing that taxonomy, $\taxonomy_{p}$, with a set of rooted splits $\mathcal{R}(\taxonomy_{p})$ such that
-each edge in $\taxonomy_{p}$ is encoded in one split in $\mathcal{R}(\taxonomy_{p})$. If the order of trees is set
-to be: $\mathcal{R}(\taxonomy_{p})=\mathcal{Z}(\taxonomy_{p}) + 
-\mathcal{Y}(\taxonomy_{p})$, then using $\mathcal{R}(\taxonomy_{p})$ in place of $\taxonomy_{p}$ in the supertree
-algorithm would produce the same output.
-
-As is typically done, we define rooted splits for each edge by noting that
-removing the edge would produce two trees. The leaf label sets of these trees
-the partition the full leaf set $\mathcal{L}$ into two groups. The include set $\mathcal{I}(e)$ which
-does not contain the root, and the exclude set $\mathcal{E}(e)$ which does contain the root.
-Such a split may be written
-\[ \mathcal{I}(e)|\bullet\mathcal{E}(e).\]
-where the $\bullet$ indicates the root. The include set is the set of leaves that are
-descendants of the daughter node of the edge. The exclude set is the set of
-leaves that are not descendants of that node. If no taxa are \emph{incertae sedis},
-then the exclude set for a node is just the total tip set minus the include set
-for the node: $\mathcal{E}(e)  =\mathcal{L}-\mathcal{I}(e)$. For a node $n$ on the tip-ward side of an edge $e$, we
-may also write $\mathcal{I}(n)$ for $\mathcal{I}(e)$, and $\mathcal{E}(n)$ for $\mathcal{E}(e)$. For any two rooted splits $A=A_{1}|\bullet A_{2}$
-and $B=B_{1}|\bullet B_{2}$ , we say that $A$ \emph{displays }$B$ if $B_{1}\subseteq A_{1}$ and $B_{2}\subseteq A_{2}$.
-
-Note that when a taxon rooted at node $n$ is not marked as \emph{incertae sedis},
-then the leaf labels that are descendants of $n$ will be in the include set for
-$n$ and all its ancestors and will be in the exclude set for all other nodes/edges.
-
-In terms of phylogenetic nomenclature the supertree pipeline acts as if the
-taxonomy defines a branch-based name \citep[see][]{deQueiroz2013} for each
-higher taxon of the form. So for the name labeling node $n$ in the taxonomic
-tree, the operational definition would be: ``the clade defined by all taxa that
-are more closely related to $\mathcal{I}(n)$ than to $\mathcal{E}(n)$, if such a clade exists on the tree.''
-
-\paragraph{Information content of the \emph{incertae sedis} annotation}
-
-In terms of rooted split interpretation, an \emph{incertae sedis} taxon differs
-from a normal taxon because its leaves should not automatically be mentioned in
-the exclude set of all of the edges that does not descend from. If a taxon
-rooted at node $n$ is marked as \emph{incertae sedis}, then the leaf labels that
-are descendants of $n$ will still be in the include set for $n$ and all its ancestors,
-but these labels will not necessarily be the exclude set for all other
-nodes/edges. They may simply be omitted in some rooted split representations,
-resulting in ``partial splits'' -- those which do not necessarily mention all of
-the taxa. Thus, when interpreted correctly the \emph{incertae sedis} annotations
-are reducing the amount of information that the taxonomy brings to the supertree
-problem.
-
-\paragraph{Naive interpretation: ignoring the annotation}
-
-The most straightforward approach for increasing the taxonomic
-completeness of the supertree is simply to no longer prune taxa from
-$\taxonomy$ because they are flagged as \emph{incertae sedis}, and
-then run the rest of the pipeline of \citet{redelings2017supertree}
-unaltered.
-However this approach results in a less resolved tree
-because taxonomic vagueness (in the form of the \emph{incertae sedis}
-annotation) is interpreted as an unwarranted statement that the
-\emph{incertae sedis} taxa are disallowed from being members of their
-sibling taxa.
-In essence, this interpretation would overinterpret the taxonomic
-    hierarchy by treating all of the groupings as statements about
-    full rooted splits, when some should be viewed as partial splits.
-
-\mthcomment{ Do we have some numbers on this (presumably from just
-running the only propinquity with a different set of prune flags)? }
-
-\paragraph{Rich taxon-specific attachment points}
-
+\paragraph{(Not using rank information) Rich taxon-specific attachment points}
+% Question: why not use rank information?  Answer: well, we could, but lets start simple.
 As mentioned above, the \emph{incertae sedis} annotation may mean that a taxon is
@@
 automatically across the taxonomy without additional information.
+
+\paragraph{(Split-based semantics) Equivalence of a taxonomy tree and a list of rooted splits}
+
+Note that, for the purposes of the supertree method, all of the information from
+the taxonomic tree is contained in the set of rooted splits. Thus, one could
+imagine replacing that taxonomy, $\taxonomy_{p}$, with a set of rooted splits $\mathcal{R}(\taxonomy_{p})$ such that
+each edge in $\taxonomy_{p}$ is encoded in one split in $\mathcal{R}(\taxonomy_{p})$. If the order of trees is set
+to be: $\mathcal{R}(\taxonomy_{p})=\mathcal{Z}(\taxonomy_{p}) + 
+\mathcal{Y}(\taxonomy_{p})$, then using $\mathcal{R}(\taxonomy_{p})$ in place of $\taxonomy_{p}$ in the supertree
+algorithm would produce the same output.
+
+As is typically done, we define rooted splits for each edge by noting that
+removing the edge would produce two trees. The leaf label sets of these trees
+the partition the full leaf set $\mathcal{L}$ into two groups. The include set $\mathcal{I}(e)$ which
+does not contain the root, and the exclude set $\mathcal{E}(e)$ which does contain the root.
+Such a split may be written
+\[ \mathcal{I}(e)|\bullet\mathcal{E}(e).\]
+where the $\bullet$ indicates the root. The include set is the set of leaves that are
+descendants of the daughter node of the edge. The exclude set is the set of
+leaves that are not descendants of that node. If no taxa are \emph{incertae sedis},
+then the exclude set for a node is just the total tip set minus the include set
+for the node: $\mathcal{E}(e)  =\mathcal{L}-\mathcal{I}(e)$. For a node $n$ on the tip-ward side of an edge $e$, we
+may also write $\mathcal{I}(n)$ for $\mathcal{I}(e)$, and $\mathcal{E}(n)$ for $\mathcal{E}(e)$. For any two rooted splits $A=A_{1}|\bullet A_{2}$
+and $B=B_{1}|\bullet B_{2}$ , we say that $A$ \emph{displays }$B$ if $B_{1}\subseteq A_{1}$ and $B_{2}\subseteq A_{2}$.
+
+Note that when a taxon rooted at node $n$ is not marked as \emph{incertae sedis},
+then the leaf labels that are descendants of $n$ will be in the include set for
+$n$ and all its ancestors and will be in the exclude set for all other nodes/edges.
+
+In terms of phylogenetic nomenclature the supertree pipeline acts as if the
+taxonomy defines a branch-based name \citep[see][]{deQueiroz2013} for each
+higher taxon of the form. So for the name labeling node $n$ in the taxonomic
+tree, the operational definition would be: ``the clade defined by all taxa that
+are more closely related to $\mathcal{I}(n)$ than to $\mathcal{E}(n)$, if such a clade exists on the tree.''
+
+\paragraph{(A first semantics) Naive interpretation: ignoring the annotation}
+
+The most straightforward approach for increasing the taxonomic
+completeness of the supertree is simply to no longer prune taxa from
+$\taxonomy$ because they are flagged as \emph{incertae sedis}, and
+then run the rest of the pipeline of \citet{redelings2017supertree}
+unaltered.
+However this approach results in a less resolved tree
+because taxonomic vagueness (in the form of the \emph{incertae sedis}
+annotation) is interpreted as an unwarranted statement that the
+\emph{incertae sedis} taxa are disallowed from being members of their
+sibling taxa.
+In essence, this interpretation would overinterpret the taxonomic
+    hierarchy by treating all of the groupings as statements about
+    full rooted splits, when some should be viewed as partial splits.
+
+% \mthcomment{ Do we have some numbers on this (presumably from just
+%running the only propinquity with a different set of prune flags)? }
+
+% \bdrcomment{ I think we would currently lose about 10 taxa }
+
+\bdrcomment{The reason why we don't do the naive thing is currently in section \ref{conflict}}
+\paragraph{(Split-based semantics) Information content of the \emph{incertae sedis} annotation}
+
+In terms of rooted split interpretation, an \emph{incertae sedis} taxon differs
+from a normal taxon because its leaves should not automatically be mentioned in
+the exclude set of all of the edges that does not descend from. If a taxon
+rooted at node $n$ is marked as \emph{incertae sedis}, then the leaf labels that
+are descendants of $n$ will still be in the include set for $n$ and all its ancestors,
+but these labels will not necessarily be the exclude set for all other
+nodes/edges. They may simply be omitted in some rooted split representations,
+resulting in ``partial splits'' -- those which do not necessarily mention all of
+the taxa. Thus, when interpreted correctly the \emph{incertae sedis} annotations
+are reducing the amount of information that the taxonomy brings to the supertree
+problem.
 
@@
 
-Additionally, the exclude set $\mathcal{E}(n)$ should contain the
-children of all the ancestors of $n$, unless those children are
-\emph{incertae sedis}.
-To avoid traversing all ancestors separately
-for each node $n$, we note that the excluded children of any ancestors
-will be in the exclude set of the parent of $n$ already.
-Therefore, we
-can write the exclude set of a node $n$ in terms of the exclude set of
-its parent:
+Additionally, the exclude set $\mathcal{E}(n)$ should contain the children of
+all the ancestors of $n$, unless those children are \emph{incertae sedis}.
+To avoid traversing all ancestors separately for each node $n$, we note that the
+excluded children of any ancestors will be in the exclude set of the parent of
+$n$ already. Therefore, we can write the exclude set of a node $n$ in terms of
+the exclude set of its parent:
 \begin{align}
@@
 \end{align}
-This
-formula allows us to compute exclude sets via a pre-order traversal on
-the taxonomy tree with exclude set for the root node defined to be
-$\emptyset$.
+This formula allows us to compute exclude sets via a pre-order traversal on the
+taxonomy tree with exclude set for the root node defined to be $\emptyset$.
 
@@
 Also: can we genericize $\mathcal{E}(parent(n))$?
+
+\subsection{Surprising results of the split-based semantics}
+
+1. Inter-digitation
+
+2. Placing A next to B, then B within A.
 
@@
 \subsection{Placement causes broken taxa}
-
+\label{conflict}
 Synthesis with \emph{incertae sedis} taxa has the potential to resolve

### 2018-04-24 doc/incertae_sedis.tex
Update.

@@
 
-\newcommand{\bdrcomment}[1]{{\color{blue} #1}\xspace}
+\newcommand{\bdrcomment}[1]{{\color{cyan} #1}\xspace}
 
@@
 }
+
+\subsection{Placement causes broken taxa}
+\bdrcomment{I duplicated section \ref{conflict} to here, since we might want
+  to discuss Figure \ref{fig:Handling-incertae-sedis} in the intro to answer
+  why anybody would care.  Or not.  But note that the section on the ``naive approach''
+  asks basically this question.}
+
+Synthesis with \emph{incertae sedis} taxa has the potential to resolve
+uncertain taxon placements using information from phylogenies.
+Since
+phylogenies are ranked higher than taxonomy, they may place members of
+incertae sedis clades as they wish even when we do not take into
+account the semantics of incertae sedis taxa.
+However, without the
+incertae sedis semantics described above, placing a taxon $A$ within a
+taxon $B$ results in conflict with taxon $B$ in the taxonomy.
+This has
+three main effects: \begin{enumerate}
+    \item The edge leading to clade $B$ is lost.
+    \item Any taxonomy-only children of $B$ are placed at the
+        MRCA of the phylogenetically informed taxa, instead of at node $B$.
+    \item The name for taxon $B$ is lost.
+    \item We do not record $A$ as being placed within $B$.
+\end{enumerate} In order to avoid a situation
+where input phylogenies conflict with a large number of taxa when
+\emph{incertae sedis} taxa are placed within them, we have previously
+filtered \emph{incertae sedis} taxa from the taxonomy when
+constructing all prior synthesis trees.
+
+When the synthesis tree conflicts with a taxonomy node, we say that
+the taxon $B$ at that node is a broken taxon.
+Broken taxa have two
+main effects, both of which are negative.
+First, the name $B$ of the
+broken taxon is removed from the synthesis tree.
+Second, the
+conflicting edge for taxon $B$ is not included in the synthesis tree.
+This means that any children of $B$ that are taxonomy-only will not be
+placed with the children of $B$ that are mentioned in input
+phylogenies.
+Instead the taxonomy-only children of a broken taxon move
+towards the root of the synthesis tree and attach at the first higher-
+ranked taxon that is an ancestor of $B$ but is not broken.
+
+We therefore seek a synthesis method that can correctly introduce
+\emph{incertae sedis} taxa into containing taxa without breaking the
+containing taxa.
+We change the semantics of names to imply, not the
+exclusion of all non-included taxa, but only some non-included taxa,
+as described in section \ref{sec:Semantics-of-incertae}.
+As a result
+of this change in semantics, placing an IS taxon $A$ within a sister
+taxon $B$ no longer results in conflict with $B$.
+This allows us to
+retain the split for $B$ within the synthesis tree, so that taxonomy-
+only children of $B$ are correctly grouped with their siblings that
+are referenced by the input trees.
+We may then retain the name for the
+no-longer-broken taxon.
+Finally, we are then able to stop filtering
+\emph{incertae sedis} taxa, so that they appear in the synthesis tree.
+Thus, the synthesis tree is able to represent substantially more
+species, without suffering the loss of taxa and the loss of structure.
 

### 2018-04-25 doc/incertae_sedis.tex
minor

@@
 The supertree method has been used with the Open Tree of
-Taxonomy \citep[OTT hereafter, see][]{rees2017automated}, but in
-principle is general.
+Taxonomy \citep[OTT hereafter, see][]{rees2017automated}, but 
+could be used in more general contexts.
 The taxonomy is a crucial input for several
-reasons: it covers a wider range of species, its list of names allows
+reasons: it covers a wide range of species, its list of names allows
 for alignment of different phylogenetic estimates to each other, and
@@
 phylogenies to the taxonomy \citep[see][ for discussion of the
-curation process in the Open Tree of Life project]{McTavishEtAt2015}.
+curation process that the Open Tree of Life project uses to align phylogenetic
+estimates to OTT]{McTavishEtAt2015}.
 This alignment is important for recognizing when two different
@@
 
-While some taxonomies may indicate hybrid taxa, the supertree approach
+Some taxonomies indicate which taxa are thought to be the result of hybridization, 
+but the supertree approach
 described here does not consider any hybridization.
@@
 
-Here describe a supertree method so that we can resolve taxonomic
+Here we describe a supertree method so that we can resolve taxonomic
 uncertainty by using published phylogenies to place \emph{incertae
@@
 The BUILD
-algorithm \citep{AhoSSU1981} is used both to check compatibility, and
-to construct a compatible tree from the final set $C$.
+algorithm \citep{AhoSSU1981} is used both to check compatibility.
+ and to construct many aspects of a compatible tree from the final set $C$.
+Refer to \citet{redelings2017supertree} for details of the procedure, including
+a description of an optimization from
+  pruning taxa that do not occur in any phylogenetic estimates in pre-processing
+  and then reattaching them to a backbone tree.
+This optimization is not crucial to the current work, but it does
+  cause some groups to attach more tipward than they would if the entire
+  tree were constructed using BUILD.
 

### 2018-04-25 doc/incertae_sedis.tex
Reformat and tiny edits.

@@
 
-The supertree algorithm of \citet{redelings2017supertree} takes as
-input a ranked list $\phyloinputs=\left\{ T_{1},\ldots,T_{n}\right\} $
-of leaf-labeled rooted input phylogenies and a single rooted taxonomy
-tree $\taxonomy_{p}$ that is ranked below all of the input trees.
-As
-discussed in \citet{redelings2017supertree}, the preferred tree would
-display as many of the highly ranked input splits as possible.
-If
-computational barriers were not an issue, the method would act by
-first producing a ranked list of rooted splits:
-$\mathcal{S}=S(T_{1})+S(T_{2})+\ldots+S(T_{n})+S(\mathbb{T})$ where
-$S(T_{i})$ denotes a list of non-trivial rooted splits created from a
-post-order traversal of tree $i$ and '$+$' denotes concatenation.
-Then
-the supertree would be produced in a greedy manner by iterating over
-the list $S$, and adding each rooted split to a growing rooted
-supertree if that split is compatible with the supertree.
-In other
-words, the set $C$ is initialized to the empty set, and we then
-iteratively consider each split in the list $\mathcal{S}$ and insert
-it into $C$ if the resulting set remains jointly compatible.
-The BUILD
-algorithm \citep{AhoSSU1981} is used both to check compatibility.
- and to construct many aspects of a compatible tree from the final set $C$.
-Refer to \citet{redelings2017supertree} for details of the procedure, including
-a description of an optimization from
-  pruning taxa that do not occur in any phylogenetic estimates in pre-processing
-  and then reattaching them to a backbone tree.
-This optimization is not crucial to the current work, but it does
-  cause some groups to attach more tipward than they would if the entire
-  tree were constructed using BUILD.
-
-Because of the large number (over 2.6 million) of tips in the full
-tree, the supertree method of \citet{redelings2017supertree} uses an
-approximation approach that relies on decomposition of the full
-problem into subproblems.
-The decomposition partitions the list of
-taxonomy splits $S(\taxonomy_{p})$ into two lists:
-$\mathcal{Z}(\taxonomy_{p}),$ the list of splits from the taxonomy tree
-which are compatible with every rooted split in the phylogenetic
-inputs, and $\mathcal{Y}(\taxonomy_{p})$, the list of taxonomic splits
-that conflict with at least one phylogenetic input split.
-Conceptually
-the approximate algorithm works by greedily adding all compatible
-splits in altered ranked list of splits:
-$\mathcal{S}^{\prime}=\mathcal{Z}(\taxonomy_{p})+S(T_{1})+S(T_{2})+\ldots+S(T_{n})+\mathcal{Y}(\taxonomy_{p})$.
-
-In practice, the decomposition allows the compatibility of splits to
-be checked on smaller, relabeled trees.
-This can be done much more
-efficiently than if compatibility was checked using the BUILD
-algorithm of \citet{AhoSSU1981} on the full leaf set.
-Thus, the
-supertree algorithm first divides the supertree problem into
-independent pieces by bisecting trees at the uncontested taxa which
-correspond to $Z$, then solves the supertree problem on these pieces,
-and finally glues the pieces back together to produce the full tree.
-
-The taxonomy tree, $\taxonomy_{p}$, is derived from OTT.
-As mentioned
-above, OTT not only contains a taxonomic hierarchy, but also contains
-taxonomic labels and annotations (referred to as ``flags'') on the
-taxa to carry extra information.
-The entire hierarchy encoded by OTT is
-a complete taxonomic tree, $\taxonomy^{\ast}.$ Nodes in
-$\taxonomy^{\ast}$ have a one-to-one correspondence with a set of
-labels $\mathcal{N}$ that are called taxon names.
-We will decompose
-$\mathcal{N}$ into the taxon names for leaf taxa \emph{$\mathcal{L}$
-}and higher taxa $\mathcal{H}$.
-We will use $\mathcal{U}$ to refer to
-the subset ($\mathcal{U}\subseteq\mathcal{N}$) of taxa that are
-annotated with the \emph{incertae sedis }property.
-Other annotations
-identify dubious taxa and taxonomic names that are unwanted artifacts
-of the process of building OTT.
-Thus, the taxonomic tree which is used
-as an input to the supertree algorithm is pruned version of
-$\taxonomy^{\ast}$.
-For our purposes, the relevant information from
-the taxonomy consists of the taxonomy tree $\taxonomy^{\ast}$, the
-labels $\mathcal{N}$, and set $\mathcal{U}$ of \emph{incertae sedis} taxa.
-In previous work, taxa in $\mathcal{U}$ and all of their descendants 
-were all pruned from
-the taxonomic hierarchy when $\taxonomy_{p}$ is extracted from OTT.
-In
-the present work, we consider supertree approaches that operate on a
-taxonomic tree $\taxonomy$, which is produced from
-$\taxonomy^{\ast}$but does not prune \emph{incertae sedis} taxa.
-
-Throughout we assume that leaf labels on input phylogenetic trees are
-only taken from the leaf set of the taxonomic tree that is being used
-and that each label occurs at most once in each input tree.
-\citet{redelings2017supertree} describe a method called
-``exemplification'' that ensures this is the case.
-Thus pruning of the
-\emph{incertae sedis} taxa from the taxonomy was also accompanied by
+The supertree algorithm of \citet{redelings2017supertree} takes as input a
+ranked list $\phyloinputs=\left\{ T_{1},\ldots,T_{n}\right\} $ of leaf-labeled rooted input phylogenies and a
+single rooted taxonomy tree $\taxonomy_{p}$ that is ranked below all of the input trees.
+As discussed in \citet{redelings2017supertree}, the preferred tree would
+display as many of the highly ranked input splits as possible. If computational
+barriers were not an issue, the method would act by first producing a ranked
+list of rooted splits: $\mathcal{S}=S(T_{1})+S(T_{2})+\ldots+S(T_{n})+S(\mathbb{T})$ where $S(T_{i})$ denotes a list
+of non-trivial rooted splits created from a post-order traversal of tree $i$ and '$+$'
+denotes concatenation. Then the supertree would be produced in a greedy manner
+by iterating over the list $S$, and adding each rooted split to a growing rooted
+supertree if that split is compatible with the supertree. In other words, the set
+$C$ is initialized to the empty set, and we then iteratively consider each split in
+the list $\mathcal{S}$ and insert it into $C$ if the resulting set remains jointly compatible.
+The BUILD algorithm \citep{AhoSSU1981} is used both to check compatibility,  and
+to construct many aspects of a compatible tree from the final set $C$.  Refer to
+\citet{redelings2017supertree} for details of the procedure, including a
+description of an optimization for pruning taxa that do not occur in any
+phylogenetic estimates during pre-processing and then reattaching them to a
+backbone tree.  This optimization is not crucial to the current work, but it
+does cause some groups to attach more tipward than they would if the entire tree
+were constructed using BUILD.
+
+Because of the large number (over 2.6 million) of tips in the full tree, the
+supertree method of \citet{redelings2017supertree} uses an approximation
+approach that relies on decomposition of the full problem into subproblems.
+The decomposition partitions the list of taxonomy splits $S(\taxonomy_{p})$ into two lists:
+$\mathcal{Z}(\taxonomy_{p}),$ the list of splits from the taxonomy tree which are compatible with every
+rooted split in the phylogenetic inputs, and $\mathcal{Y}(\taxonomy_{p})$, the list of taxonomic splits
+that conflict with at least one phylogenetic input split. Conceptually the
+approximate algorithm works by greedily adding all compatible splits in
+altered ranked list of splits: $\mathcal{S}^{\prime}=\mathcal{Z}(\taxonomy_{p})+S(T_{1})+S(T_{2})+\ldots+S(T_{n})+\mathcal{Y}(\taxonomy_{p})$.
+
+In practice, the decomposition allows the compatibility of splits to be checked
+on smaller, relabeled trees. This can be done much more efficiently than if
+compatibility was checked using the BUILD algorithm of \citet{AhoSSU1981} on the
+full leaf set. Thus, the supertree algorithm first divides the supertree problem
+into independent pieces by bisecting trees at the uncontested taxa which
+correspond to $Z$, then solves the supertree problem on these pieces, and finally
+glues the pieces back together to produce the full tree.
+
+The taxonomy tree, $\taxonomy_{p}$, is derived from OTT. As mentioned above, OTT not only
+contains a taxonomic hierarchy, but also contains taxonomic labels and
+annotations (referred to as ``flags'') on the taxa to carry extra information.
+The entire hierarchy encoded by OTT is a complete taxonomic tree, $\taxonomy^{\ast}.$ Nodes in
+$\taxonomy^{\ast}$ have a one-to-one correspondence with a set of labels $\mathcal{N}$ that are called
+taxon names. We will decompose $\mathcal{N}$ into the taxon names for leaf taxa $\mathcal{L}$ and
+higher taxa $\mathcal{H}$. We will use $\mathcal{U}$ to refer to the subset ($\mathcal{U}\subseteq\mathcal{N}$) of taxa that are
+annotated with the \emph{incertae sedis }property. Other annotations identify
+dubious taxa and taxonomic names that are unwanted artifacts of the process of
+building OTT. Thus, the taxonomic tree which is used as an input to the
+supertree algorithm is pruned version of $\taxonomy^{\ast}$. For our purposes, the relevant
+information from the taxonomy consists of the taxonomy tree $\taxonomy^{\ast}$, the labels $\mathcal{N}$,
+and set $\mathcal{U}$ of \emph{incertae sedis} taxa. In previous work, taxa in $\mathcal{U}$ and all
+of their descendants were all pruned from the taxonomic hierarchy when $\taxonomy_{p}$ is
+extracted from OTT. In the present work, we consider supertree approaches that
+operate on a taxonomic tree $\taxonomy$, which is produced from $\taxonomy^{\ast}$but does not prune
+\emph{incertae sedis} taxa.
+
+Throughout we assume that leaf labels on input phylogenetic trees are only taken
+from the leaf set of the taxonomic tree that is being used and that each label
+occurs at most once in each input tree. \citet{redelings2017supertree} describe
+a method called ``exemplification'' that ensures this is the case. Thus pruning
+of the \emph{incertae sedis} taxa from the taxonomy was also accompanied by
 pruning those taxa and all of their descendants from the input phylogenetic

### 2018-04-25 doc/incertae_sedis.tex
Tweak wording in 1.2 and move figure.

@@
 The problem is becoming more acute because NCBI is putting a larger
-number of taxa into groups that are marked as “unclassified.” For
-example, when OTT 2.9 was created NCBI’s clasification of the bird
+number of taxa into groups that are marked as ``classified''. or
+example, when OTT 2.9 was created NCBI's clasification of the bird
 family Sylvidae included a group five genera that were placed within
-“[http://www.ncbi.nlm.nih.gov/Taxonomy/Brows er/wwwtax.cgi?mode=Undef&
+``[http://www.ncbi.nlm.nih.gov/Taxonomy/Brows er/wwwtax.cgi?mode=Undef&
 id=36270&lvl=3&lin=f&keep=1&srchmode=1&unlock||unclassi fied
-Sylviidae].” This includes the genus Regulus.
+Sylviidae].'' This includes the genus Regulus.
 Thus in OTT 2.9 Regulus
@@
 259942).
-The “unclassified Sylviidae” does not appear in OTT; instead
-Regulus is flagged as “unclassified,sibling higher” and all of the
-species within Regulus are flagged as “unclassified inherited.”
+The ``unclassified Sylviidae'' does not appear in OTT; instead
+Regulus is flagged as ``unclassified,sibling higher'' and all of the
+species within Regulus are flagged as ``unclassified inherited.''
 
 The taxonomy (as of version 2.9) also contains 685 cases of taxa that
-are flagged as both “unclassified inherited” and “unclassified”based
+are flagged as both ``unclassified inherited'' and ``unclassified''based
 on grep unclass.*unclassified\_inherited taxonomy.tsv.
@@
 Flags.
-The Open Tree’s reference taxonomy is produced by a tool called
+The Open Tree's reference taxonomy is produced by a tool called
 smasher that notices hints that a taxon is incertae sedis and labels
@@
 /reference-taxonomy/wiki /Taxon-flags.This document is not concerned
-with the flagging system per se, so “incertae sedis” will be used here
+with the flagging system per se, so ``incertae sedis'' will be used here
 to refer to all of the flags that denote taxa with uncertain
@@
 Container nodes.
-Some taxonomies have nodes with names like “Incertae
-sedis (Bacteria)”.
+Some taxonomies have nodes with names like ``Incertae
+sedis (Bacteria)''.
 Such nodes are called containers.
@@
 }
-
-\subsection{Placement causes broken taxa}
-\bdrcomment{I duplicated section \ref{conflict} to here, since we might want
-  to discuss Figure \ref{fig:Handling-incertae-sedis} in the intro to answer
-  why anybody would care.  Or not.  But note that the section on the ``naive approach''
-  asks basically this question.}
-
-Synthesis with \emph{incertae sedis} taxa has the potential to resolve
-uncertain taxon placements using information from phylogenies.
-Since
-phylogenies are ranked higher than taxonomy, they may place members of
-incertae sedis clades as they wish even when we do not take into
-account the semantics of incertae sedis taxa.
-However, without the
-incertae sedis semantics described above, placing a taxon $A$ within a
-taxon $B$ results in conflict with taxon $B$ in the taxonomy.
-This has
-three main effects: \begin{enumerate}
-    \item The edge leading to clade $B$ is lost.
-    \item Any taxonomy-only children of $B$ are placed at the
-        MRCA of the phylogenetically informed taxa, instead of at node $B$.
-    \item The name for taxon $B$ is lost.
-    \item We do not record $A$ as being placed within $B$.
-\end{enumerate} In order to avoid a situation
-where input phylogenies conflict with a large number of taxa when
-\emph{incertae sedis} taxa are placed within them, we have previously
-filtered \emph{incertae sedis} taxa from the taxonomy when
-constructing all prior synthesis trees.
-
-When the synthesis tree conflicts with a taxonomy node, we say that
-the taxon $B$ at that node is a broken taxon.
-Broken taxa have two
-main effects, both of which are negative.
-First, the name $B$ of the
-broken taxon is removed from the synthesis tree.
-Second, the
-conflicting edge for taxon $B$ is not included in the synthesis tree.
-This means that any children of $B$ that are taxonomy-only will not be
-placed with the children of $B$ that are mentioned in input
-phylogenies.
-Instead the taxonomy-only children of a broken taxon move
-towards the root of the synthesis tree and attach at the first higher-
-ranked taxon that is an ancestor of $B$ but is not broken.
-
-We therefore seek a synthesis method that can correctly introduce
-\emph{incertae sedis} taxa into containing taxa without breaking the
-containing taxa.
-We change the semantics of names to imply, not the
-exclusion of all non-included taxa, but only some non-included taxa,
-as described in section \ref{sec:Semantics-of-incertae}.
-As a result
-of this change in semantics, placing an IS taxon $A$ within a sister
-taxon $B$ no longer results in conflict with $B$.
-This allows us to
-retain the split for $B$ within the synthesis tree, so that taxonomy-
-only children of $B$ are correctly grouped with their siblings that
-are referenced by the input trees.
-We may then retain the name for the
-no-longer-broken taxon.
-Finally, we are then able to stop filtering
-\emph{incertae sedis} taxa, so that they appear in the synthesis tree.
-Thus, the synthesis tree is able to represent substantially more
-species, without suffering the loss of taxa and the loss of structure.
+\begin{figure}
+\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-clade/tree1-shaded}
+
+}\hfill{}\subfloat[Taxonomy]{\includegraphics[scale=0.5]{Figures/is-clade/tax-shaded}
+
+}
+
+\subfloat[Synthesis Tree (no \emph{incertae sedis})]{\includegraphics[scale=0.5]{Figures/is-clade/synth1-no-is-shaded}
+
+}\hfill{}\subfloat[Synthesis Tree (\emph{incertae sedis} aware)]{\includegraphics[scale=0.5]{Figures/is-clade/synth1-shaded}
+
+}
+
+\caption{\label{fig:Handling-incertae-sedis}Handling \emph{incertae sedis
+}taxa recovers additional edges and taxon names.
+Supertree construction on input tree (a) and taxonomy tree (b) with \emph{incertae sedis}
+clade $C$ leads to synthesis tree (c) if \emph{incertae sedis} annotations are
+ignored, but to synthesis tree (d) if \emph{incertae sedis} annotations are respected.
+The tree in (d) recovers taxon names $AB$ and $A$, as well as the edge to clade $AB$
+that groups $A$ and $B$. Taxon names and edges that are conditional on handling
+\emph{incertae sedis}
+are shaded grey.
+}
+
+\end{figure}
+
+\subsection{A naive semantics: consequences of ignoring the annotation}
+
+The most straightforward approach for increasing the taxonomic completeness of
+the supertree is simply to no longer prune taxa from $\taxonomy$ because they are flagged
+as \emph{incertae sedis}, and then run the rest of the pipeline of
+\citet{redelings2017supertree} unaltered.  Before we propose a semantics of
+\emph{incertae sedis}, it is worth pointing out what this naive semantics is lacking.
+
+First, this approach results in a less-resolved tree because taxonomic
+vagueness (in the form of the \emph{incertae sedis} annotation) is interpreted as
+an unwarranted statement that the \emph{incertae sedis} taxa are disallowed from
+being members of their sibling taxa.  When an incertae sedis taxon is placed within
+a sibling taxon by input phylogenies, this is taken as a rejection of the monophyly
+of the sibling taxon, which is then removed from the synthesis tree.  For
+example, in the synthesis tree in Fig. \ref{fig:Handling-incertae-sedis}c, the
+placement of taxon $C$ in group $AB$ is interpreted as a rejection of the
+monophyly of $AB$. The edge leading to $AB$ is therefore not incorporated into
+the synthesis tree.  In contrast, the synthesis tree in Fig.
+\ref{fig:Handling-incertae-sedis}d that does not reject the monophyly of $AB$
+retains the grouping of $A$ and $B$.
+
+Second, this approach results in the removal of clade names for all clades that
+have an \emph{incertae sedis} taxon place within them.  For example, in Fig.
+\ref{fig:Handling-incertae-sedis}c the names for the clades $A$ and $AB$ are lost.
+
+Third, this approach cannot use information from phylogenies to resolve the
+location of uncertain taxa.  For example, the synthesis tree in
+\ref{fig:Handling-incertae-sedis}c has lost the label for $A$, and so does not
+allow us to place $C$ within $A$.  In contrast, the synthesis tree in Fig.
+\ref{fig:Handling-incertae-sedis}d that takes account of \emph{incertae sedis}
+annotations and retains the label for $A$.  This allows us to infer that
+\emph{incertae sedis} taxon $C$ has been placed within $A$, resolving
+the uncertainty.
 
@@
 
-\paragraph{(A first semantics) Naive interpretation: ignoring the annotation}
-
-The most straightforward approach for increasing the taxonomic
-completeness of the supertree is simply to no longer prune taxa from
-$\taxonomy$ because they are flagged as \emph{incertae sedis}, and
-then run the rest of the pipeline of \citet{redelings2017supertree}
-unaltered.
-However this approach results in a less resolved tree
-because taxonomic vagueness (in the form of the \emph{incertae sedis}
-annotation) is interpreted as an unwarranted statement that the
-\emph{incertae sedis} taxa are disallowed from being members of their
-sibling taxa.
-In essence, this interpretation would overinterpret the taxonomic
-    hierarchy by treating all of the groupings as statements about
-    full rooted splits, when some should be viewed as partial splits.
-
-% \mthcomment{ Do we have some numbers on this (presumably from just
-%running the only propinquity with a different set of prune flags)? }
-
-% \bdrcomment{ I think we would currently lose about 10 taxa }
-
-\bdrcomment{The reason why we don't do the naive thing is currently in section \ref{conflict}}
 \paragraph{(Split-based semantics) Information content of the \emph{incertae sedis} annotation}
@@
 members of the clade to be placed separately (Figure \ref{fig:Broken-incertae-sedis}).
-
-\begin{figure}
-\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-clade/tree1-shaded}
-
-}\hfill{}\subfloat[Taxonomy]{\includegraphics[scale=0.5]{Figures/is-clade/tax-shaded}
-
-}
-
-\subfloat[Synthesis Tree (no \emph{incertae sedis})]{\includegraphics[scale=0.5]{Figures/is-clade/synth1-no-is-shaded}
-
-}\hfill{}\subfloat[Synthesis Tree (\emph{incertae sedis} aware)]{\includegraphics[scale=0.5]{Figures/is-clade/synth1-shaded}
-
-}
-
-\caption{\label{fig:Handling-incertae-sedis}Handling \emph{incertae sedis
-}taxa recovers additional edges and taxon names.
-Supertree construction
-on input tree (a) and taxonomy tree (b) with \emph{incertae sedis}
-clade $C$ leads to synthesis tree (c).
-However, supertree construction
-with \emph{incertae sedis} handling constructs tree (d), which recovers
-taxon name $AB$ and $A$, as well as the edge to clade $AB$.
-Taxon
-names and edges that are conditional on handling \emph{incertae sedis}
-are shaded grey.
-}
-
-\end{figure}
 \begin{figure}

### 2018-04-25 doc/incertae_sedis.tex
Try another rewording.

@@
 
-First, this approach results in a less-resolved tree because taxonomic
-vagueness (in the form of the \emph{incertae sedis} annotation) is interpreted as
-an unwarranted statement that the \emph{incertae sedis} taxa are disallowed from
-being members of their sibling taxa.  When an incertae sedis taxon is placed within
-a sibling taxon by input phylogenies, this is taken as a rejection of the monophyly
-of the sibling taxon, which is then removed from the synthesis tree.  For
-example, in the synthesis tree in Fig. \ref{fig:Handling-incertae-sedis}c, the
-placement of taxon $C$ in group $AB$ is interpreted as a rejection of the
-monophyly of $AB$. The edge leading to $AB$ is therefore not incorporated into
-the synthesis tree.  In contrast, the synthesis tree in Fig.
-\ref{fig:Handling-incertae-sedis}d that does not reject the monophyly of $AB$
-retains the grouping of $A$ and $B$.
+This approach incorrectly treats the intrusion of an \emph{incertae sedis} taxon
+into a sibling taxon as a conflicting with monophyly of the sibling taxon. For
+example, in Fig. \ref{fig:Handling-incertae-sedis}, the placement of the
+\emph{incertae sedis} taxon $C$ within $A$ and $AB$, is considered
+to contradict the monophyly of $A$ and $AB$ and thus to conflict with those taxa.
+This has three main effects.  These effects are illustrated in the difference
+between the synthesis tree in Fig. \ref{fig:Handling-incertae-sedis}c, which
+ignores \emph{incertae sedis} annotations, and the synthesis tree in Fig.
+\ref{fig:Handling-incertae-sedis}d, which respects them.
+
+First, this approach results in a less-resolved tree.  While input trees are
+free to place incertae sedis taxa within a sibling, the naive approach
+interprets this placement as conflict with the sibling.  Since the taxonomy
+is ranked below all input trees, conflict is always resolved in favor of the
+input trees, and so the sibling in the taxonomy is removed from the synthesis
+tree.  For example, in the synthesis tree in Fig.
+\ref{fig:Handling-incertae-sedis}c, the placement of taxon $C$ in group $AB$ is
+interpreted as a rejection of the monophyly of $AB$. The edge leading to $AB$ is
+therefore not incorporated into the synthesis tree.  In contrast, the synthesis
+tree in Fig. \ref{fig:Handling-incertae-sedis}d that does not reject the
+monophyly of $AB$ retains the grouping of $A$ and $B$.
 
 Second, this approach results in the removal of clade names for all clades that
-have an \emph{incertae sedis} taxon place within them.  For example, in Fig.
+have an \emph{incertae sedis} taxon placed within them.  For example, in Fig.
 \ref{fig:Handling-incertae-sedis}c the names for the clades $A$ and $AB$ are lost.
@@
 \ref{fig:Handling-incertae-sedis}c has lost the label for $A$, and so does not
-allow us to place $C$ within $A$.  In contrast, the synthesis tree in Fig.
-\ref{fig:Handling-incertae-sedis}d that takes account of \emph{incertae sedis}
-annotations and retains the label for $A$.  This allows us to infer that
-\emph{incertae sedis} taxon $C$ has been placed within $A$, resolving
-the uncertainty.
+allow us to note that $C$ has been placed within $A$.  In contrast, the
+synthesis tree in Fig. \ref{fig:Handling-incertae-sedis}d retains the label for
+$A$, which allows us to resolve taxonomic uncertainty by infering that
+\emph{incertae sedis} taxon $C$ has been placed within $A$.
 

### 2018-04-26 doc/incertae_sedis.tex
Update.

@@
 
-Supertree methods combine a collection of input trees with different
-taxon sets into a single tree on the combined taxon set.
-These methods are usually applied to leaf-labeled phylogenetic estimates. 
-Here we describe a supertree approach that overcomes issues associated with
-using a taxonomy as one of the inputs to a supertree method.
-
-The Open Tree of Life project \citep{HinchliffEtAl2015} seeks to build
-a comprehensive supertree \citep[see][]{redelings2017supertree}
-covering all of life.
-The approach is to combine information in
-published phylogenies with a comprehensive taxonomy that supplies
-taxon names.
-The supertree method has been used with the Open Tree of
-Taxonomy \citep[OTT hereafter, see][]{rees2017automated}, but 
-could be used in more general contexts.
-The taxonomy is a crucial input for several
-reasons: it covers a wide range of species, its list of names allows
-for alignment of different phylogenetic estimates to each other, and
-it provides names for clades.
-Only a small proportion of known species
-have been included in a phylogenetic analysis, thus a taxonomy is
-important for achieving comprehensive coverage of known taxa.
-Phylogenetic estimates collected from the published literature often
-use different names for the same species.
-Lists of synonyms in OTT
-(and other taxonomies) allow data curators to ``align'' input
-phylogenies to the taxonomy \citep[see][ for discussion of the
-curation process that the Open Tree of Life project uses to align phylogenetic
-estimates to OTT]{McTavishEtAt2015}.
-This alignment is important for recognizing when two different
-estimates refer to the same taxon.
-Biologists are familiar with names
-for ``higher'' taxa (taxa above the species rank).
-Thus, naming clades
-in the supertree makes the tree easier to navigate and use.
-
-Some taxonomies indicate which taxa are thought to be the result of hybridization, 
-but the supertree approach
-described here does not consider any hybridization.
-Thus the taxonomic
-hierarchy can be converted to a tree.
-While not all taxonomists
-believe that named taxa should correspond to clades \citep[see, for
-example,][]{HorandlS2010}, the priniciples of phylogenetic
-classification are used by such a large majority of practicing
-taxonomists, that we have chosen to treat the taxonomic tree that
-mirrors the hierarchy of the taxonomy as an estimate of phylogenetic
-relationships.
-We refer to labels of taxonomy nodes as ``taxon
-names'', and assume that taxon names and taxonomy nodes have a one-to-
-one correspondence.
-Note that the taxonomy may also contain out-degree
-1 nodes.
-These nodes correspond to monotypic taxa which contain a
-single child of lower taxonomic rank.
-
-The supertree method described by \citet{redelings2017supertree} is
-able to use a set of phylogenetic estimates and taxonomy to produce a
-supertree with clades names according to the taxonomy.
-However, the
-method was not able to accommodate the fact that taxonomies frequently
-contain nodes with uncertain placement.
-These nodes are often labelled
-``\emph{incertae sedis}'', which means ``uncertain seat'' in Latin.
-The common interpretation of such taxa is that they may not be moved
-further towards the root, but may be moved into their sibling taxa.
-For example, a genus with a sibling that is a family may be annotated
-as ``\emph{incertae familia}'', indicating that it is unclear which
-family the genus should be placed in.
-\emph{Incertae sedis} taxa
-frequently occur when a specimen is identified down to a given
-taxonomic level, but no further.
-Extinct taxa are often \emph{incertae
-sedis}.
-OTT is constructed from several source taxonomies, and these
-sources include various ways of indicating that the position of a
-taxon within the hierarchy is uncertain.
-OTT uses a series of
-``flags'' to annotate these taxa.
-The supertree of \citet{redelings2017supertree} simply pruned these taxa from the
-taxonomy and input trees.
-Thus, the final supertree lacked taxa which
-were \emph{incertae sedis} or the descendants of such taxa.
-
-Here we describe a supertree method so that we can resolve taxonomic
-uncertainty by using published phylogenies to place \emph{incertae
-sedis} taxa.
-Developing this method required deciding on a set of
-operational semantics to be used when interpreting the \emph{incertae
-sedis} annotation.
-These semantics affect the interpretation of the
-phylogenetic statements being made by the input taxonomy and the rules
-for applying taxonomic names to the final supertree.
-In addition to
-discussing how to interpret the \emph{incertae sedis} annotation,
-we describe the algorithmic changes to the supertree pipeline that
-were required to adequately represent the taxonomic uncertainty.
-The
-approach described here allows us to include thousands of new taxa in
-our supertree analysis that were previously filtered out.
-It will also
-enable the Open Tree of Life project to include extinct taxa, since
-many of these taxa are \emph{incertae sedis}.
-This makes substantial
-progress towards our goal of comprehensive inclusion of known taxa
-into the Open Tree of Life summary tree.
+Supertree methods combine a collection of input trees with different taxon sets
+into a single tree on the combined taxon set. These methods are usually applied
+to leaf-labeled phylogenetic estimates.  Here we describe a supertree approach
+that overcomes issues associated with using a taxonomy as one of the inputs to
+a supertree method.
+
+The Open Tree of Life project \citep{HinchliffEtAl2015} seeks to build a
+comprehensive supertree \citep[see][]{redelings2017supertree} covering all of
+life. The approach is to combine information in published phylogenies with a
+comprehensive taxonomy that supplies taxon names. The supertree method has
+been used with the Open Tree Taxonomy \citep[OTT hereafter, see][]{rees2017automated},
+but could be used in more general contexts. The taxonomy is a crucial input for
+several reasons: it covers a wide range of species, its list of names allows
+for alignment of different phylogenetic estimates to each other, and it provides
+names for clades. Only a small proportion of known species have been included in
+a phylogenetic analysis, thus a taxonomy is important for achieving
+comprehensive coverage of known taxa. Phylogenetic estimates collected from the
+published literature often use different names for the same species. Lists of
+synonyms in OTT (and other taxonomies) allow data curators to ``align'' input
+phylogenies to the taxonomy \citep[see][ for discussion of the curation process
+that the Open Tree of Life project uses to align phylogenetic estimates to OTT]{McTavishEtAt2015}.
+This alignment is important for recognizing when two different estimates refer
+to the same taxon. Biologists are familiar with names for ``higher'' taxa (taxa
+above the species rank). Thus, naming clades in the supertree makes the tree
+easier to navigate and use.
+
+Some taxonomies indicate which taxa are thought to be the result of
+hybridization, but the supertree approach described here does not consider any
+hybridization. Thus the taxonomic hierarchy can be converted to a tree. While
+not all taxonomists believe that named taxa should correspond to clades \citep[see,
+for example,][]{HorandlS2010}, the priniciples of phylogenetic classification
+are used by such a large majority of practicing taxonomists, that we have chosen
+to treat the taxonomic tree that mirrors the hierarchy of the taxonomy as an
+estimate of phylogenetic relationships. We refer to labels of taxonomy nodes as
+``taxon names'', and assume that taxon names and taxonomy nodes have a
+one-to-one correspondence. Note that the taxonomy may also contain out-degree
+1 nodes. These nodes correspond to monotypic taxa which contain a single child
+of lower taxonomic rank.
+
+The supertree method described by \citet{redelings2017supertree} is able to use
+a set of phylogenetic estimates and taxonomy to produce a supertree with clades
+named according to the taxonomy. However, the method was not able to accommodate
+the fact that taxonomies frequently contain nodes with uncertain placement.
+These nodes are often labelled ``\emph{incertae sedis}'', which means
+``uncertain seat'' in Latin. The common interpretation of such taxa is that
+they may not be moved further towards the root, but may be moved into their
+sibling taxa. For example, a genus with a sibling that is a family may be
+annotated as ``\emph{incertae familia}'', indicating that it is unclear which
+family the genus should be placed in. \emph{Incertae sedis} taxa frequently
+occur when a specimen is identified down to a given taxonomic level, but no
+further. Extinct taxa are often \emph{incertae sedis}. OTT is constructed from
+several source taxonomies, and these sources include various ways of indicating
+that the position of a taxon within the hierarchy is uncertain. OTT uses a
+series of ``flags'' to annotate these taxa. The supertree of
+\citet{redelings2017supertree} simply pruned these taxa from the taxonomy and
+input trees. Thus, the final supertree lacked taxa which were \emph{incertae
+sedis} or the descendants of such taxa.
+
+Here we describe a supertree method so that we can resolve taxonomic uncertainty
+by using published phylogenies to place \emph{incertae sedis} taxa. Developing
+this method required deciding on a set of operational semantics to be used when
+interpreting the \emph{incertae sedis} annotation. These semantics affect the
+interpretation of the phylogenetic statements being made by the input taxonomy
+and the rules for applying taxonomic names to the final supertree. In addition
+to discussing how to interpret the \emph{incertae sedis} annotation, we describe
+the algorithmic changes to the supertree pipeline that were required to
+adequately represent the taxonomic uncertainty. The approach described here
+allows us to include thousands of new taxa in our supertree analysis that were
+previously filtered out. It will also enable the Open Tree of Life project to
+include extinct taxa, since many of these taxa are \emph{incertae sedis}. This
+makes substantial progress towards our goal of comprehensive inclusion of known
+taxa into the Open Tree of Life summary tree.
 
@@
 example, in Fig. \ref{fig:Handling-incertae-sedis}, the placement of the
-\emph{incertae sedis} taxon $C$ within $A$ and $AB$, is considered
+\emph{incertae sedis} taxon $C$ within $A$ and $AB$ is considered
 to contradict the monophyly of $A$ and $AB$ and thus to conflict with those taxa.
@@
 
-First, this approach results in a less-resolved tree.  While input trees are
+First, the naive approach results in a less-resolved tree.  While input trees are
 free to place incertae sedis taxa within a sibling, the naive approach
@@
 
-Second, this approach results in the removal of clade names for all clades that
+Second, the naive approach results in the removal of clade names for all clades that
 have an \emph{incertae sedis} taxon placed within them.  For example, in Fig.
@@
 
-Third, this approach cannot use information from phylogenies to resolve the
+Third, the naive approach cannot use information from phylogenies to resolve the
 location of uncertain taxa.  For example, the synthesis tree in
 \ref{fig:Handling-incertae-sedis}c has lost the label for $A$, and so does not
-allow us to note that $C$ has been placed within $A$.  In contrast, the
-synthesis tree in Fig. \ref{fig:Handling-incertae-sedis}d retains the label for
-$A$, which allows us to resolve taxonomic uncertainty by infering that
-\emph{incertae sedis} taxon $C$ has been placed within $A$.
+allow us to note that $C$ has been placed within $A$.  In contrast, the synthesis
+tree in Fig. \ref{fig:Handling-incertae-sedis}d retains the label for $A$, which
+allows us to resolve taxonomic uncertainty by inferring that $C$ has been placed
+within $A$.
 
@@
 sedis} taxa}
+
+\subsection{Goals for an \emph{incertae sedis} semantics}
 
@@
 \begin{enumerate}
-    \item An \emph{incertae sedis} node may intrude into its siblings and their
-descendants.
-    \item The semantics of an \emph{incertae sedis} node does not depend on assigned ranks, but only on the tree.
+    \item An \emph{incertae sedis} node may intrude into its siblings and their descendants without conflict.
+    \item An \emph{incertae sedis} node has an unrestricted range within the descendants of its sibling taxa.
+    \item The semantics of an \emph{incertae sedis} node does not depend on side-information such as taxon ranks, but only on the taxonomy tree.
     \item Semantics is based on deriving a rooted (partial) split for each branch of the taxonomy.
@@
 
-\paragraph{(Not using rank information) Rich taxon-specific attachment points}
-% Question: why not use rank information?  Answer: well, we could, but lets start simple.
-As mentioned above, the \emph{incertae sedis} annotation may mean that a taxon is
-unplaced, but only a few placements within the taxonomy are possible.
-For example, a genus might be characterized as \emph{incertae sedis}
-at the level of the subfamily.
-Thus, it could be placed inside any of the
-hypothesized sub-families, or within its own subfamily, but it would
-be unexpected for it to be placed inside one of the other genera.
-
-In principle this information is easy to incorporate.
-Ideally, the \emph{incertae sedis}
-annotation some taxon $x$ would be accompanied by a specific
-statement about which other taxa might include $x$.
-If $\mathcal{P}$ is the set of taxa that are potential ancestors for
-incertae sedis taxon $x$, then $\mathcal{L}(x)$ will occur in neither
-the include nor the exclude set of any taxon in $\mathcal{P}$.
-Unfortunately, utilizing taxon-specific interpretations of the
-incertae sedis flag relies on detailed readings of character argumentation
-(which is lacking in OTT) or rank-based information (which is frequently
-untrustworthy in OTT).
-Thus, here we pursue an approach of ascribing a
-meaning to the \emph{incertae sedis} label which attempts to capture
-the core of the idea that it articulates, but which can be applied
-automatically across the taxonomy without additional information.
-
-\paragraph{(Split-based semantics) Equivalence of a taxonomy tree and a list of rooted splits}
+\subsubsection{Unrestricted range and ignoring side information}
+The choice to allow each \emph{incertae sedis} taxon to attach anywhere within
+the descendants of its siblings requires some explanation, for two reasons.
+First, more precise information about where a taxon can attach is easy to
+incorporate. If $\mathcal{P}$ is the set of taxa that are potential ancestors for \emph{incertae
+sedis} taxon $x$, then we may simply remove the descendants of $x$ from the exclude
+set for each $p\in\mathcal{P}$.  Such an approach would require that the path from $x$ to $p$
+go through only other potential ancestors.
+
+Second, as mentioned above, it may be expected that an \emph{incertae sedis} taxon
+attach only a specific points within the taxonomy. For example, a genus
+might be characterized as \emph{incertae sedis} at the level of the subfamily.
+Thus, it could be placed inside any of the hypothesized sub-families, or within
+its own subfamily, but it would be unexpected for it to be placed inside one of
+the other genera.
+
+Unfortunately, reliable information about the range of valid attachment points
+for each \emph{incertae sedis} taxon is lacking.  Ideally, the \emph{incertae
+  sedis} annotation for some taxon $x$ would be accompanied by a specific statement
+about which other taxa might include $x$.  However, OTT does not provide this
+directly.  Constructing a range for each incertae sedis taxon relies on detailed
+readings of character argumentation (which is lacking in OTT) or rank-based
+information (which is frequently untrustworthy in OTT). Thus, here we pursue
+an approach of ascribing a meaning to the \emph{incertae sedis} label which
+attempts to capture the core of the idea that it articulates, but which can be
+applied automatically across the taxonomy without additional information.
+
+\subsubsection{Split-based semantics}
+\paragraph{Equivalence of a taxonomy tree and a list of rooted splits}
 
@@
 
+\paragraph{Definition of rooted splits}
 As is typically done, we define rooted splits for each edge by noting that
@@
 
-\paragraph{(Split-based semantics) Information content of the \emph{incertae sedis} annotation}
-
-In terms of rooted split interpretation, an \emph{incertae sedis} taxon differs
-from a normal taxon because its leaves should not automatically be mentioned in
-the exclude set of all of the edges that does not descend from. If a taxon
-rooted at node $n$ is marked as \emph{incertae sedis}, then the leaf labels that
-are descendants of $n$ will still be in the include set for $n$ and all its ancestors,
-but these labels will not necessarily be the exclude set for all other
-nodes/edges. They may simply be omitted in some rooted split representations,
-resulting in ``partial splits'' -- those which do not necessarily mention all of
-the taxa. Thus, when interpreted correctly the \emph{incertae sedis} annotations
-are reducing the amount of information that the taxonomy brings to the supertree
-problem.
-
-\paragraph{Splits-based synthesis method}
-
-Here we focus on a split-based semantics instead of a semantics based on apply
-of a series of edit operations to the tree. The split-based semantics allows us
-to seek trees that are simultaneously consistent with a set of split-constraints.
-The use of a split-based semantics does impose some limitations. These come
-primarily because in the edit-based semantics, after we place an
-\emph{incertae sedis} taxon, it is no loner \emph{incertae sedis}, whereas with
-the split-based semantics taxa are either always \emph{incertae sedis} or never
-\emph{incertae sedis}.
-
-This has an effect, for example, when we consider two incertae sedis siblings $A$
-and $B$. If we can place $A$ within $B$, and $B$ within $A$, then a split- based
-semantics allows us to freely inter-digitate $A$ and $B$. If we do not want to
-allow $A$ and $B$ to be inter-digitated, then we cannot both $A$ within $B$ and $B$
-within $A$. In addition, we can imagine that a family $A$ is incertae sedis within a
-kingdom. In an edit-based semantics, we could place $A$ as a sibling to a genus $B$,
-mark $A$ as non-\emph{incertae-sedis}, and then place $B$ within $A$. In a split-based
-semantics, we must disallow either $B$ within $A$ or $A$ within $B$ if we want to disallow
-interdigitation.
-
-As a result of this, we cannot satisfy all five of these criteria. Thus, we
-describe two alternative split-based semantics. The non-interdigitation
-semantics gets \# (1,2,4,5), and the interdigitation semantics gets (1,3,4,5).
-This makes the point that each semantics results from different choices about
-which criteria to satisfy.
-
 \subsection{Split-based semantics for \emph{incertae sedis} taxa}
 
-In order to apply supertree methods to rooted trees with \emph{incertae sedis}
-taxa, we first propose a semantics of \emph{incertae sedis }in terms of rooted
-splits.
-
-\subsubsection{semantics \#1a: The original, simplest semantics}
-\begin{align}
-    \mathcal{E}(n) & = \mathcal{E}(parent(n))\cup\left[\mathcal{I}(s)\big|s\in
-           siblings(n), s \text{ not \emph{incertae sedis}}\right].\label{eq:exclude-set-formula-1a}
-\end{align}
-This semantics \emph{does} allow interdigitation.
-
-\subsubsection{semantics \#1b: modified to disallow inter-digitating
-\emph{incertae sedis} siblings}
-
-An \emph{incertae sedis} taxon can be moved into any of the
-descendants of its siblings.
-We seek to represent this by constructing
-modified splits for each branch of the taxonomy tree.
-The include sets
-of these splits remain unchanged, but we construct reduced exclude
-sets in order to allow \emph{incertae sedis} taxa to intrude into
-their sibling taxa.
-
-The reduced exclude set $\mathcal{E}(n)$ for node $n$ should therefore
-not contain the descendants of its siblings which are marked
-\emph{incertae }sedis\emph{,} but should contain the descendants of
-its other siblings\emph{.
-}We additionally stipulate that it is not
-possible for \emph{incertae sedis} taxa to intrude into their
-\emph{incertae sedis} siblings.
-(This is not the only way to construct
-a split-based semantics for \emph{incertae sedis}, but it avoids
-certain complications, as noted in the discussion.) We thus define
-$\mathcal{I}(m,n)$ to indicate descendants of $m$ that are excluded
-from $n$.
-Thus:
+An \emph{incertae sedis} taxon $n$ differs from a normal taxon because it can be
+moved into the descendants of any of its siblings.  Thus, annotating $n$ as
+\emph{incertae sedis} really modifies the meanings of its siblings so that they
+no longer exclude $n$.  In terms of rooted splits, we can represent this lack of
+exclusion by omitting taxa from the exclude set for a node.  Such a split does
+not mention the complete set of tip labels $\mathcal{L}$ and is therefore called a
+`partial split'', as opposed to a ``full split'' that contains the entirely of $\mathcal{L}$.
+Thus, when interpreted correctly the \emph{incertae sedis} 
+annotations are reducing the amount of information that the taxonomy brings to
+the supertree problem.
+
+We seek to define modified splits for each branch of the taxonomy tree. The
+include sets of these splits remain unchanged, but the exclude sets may be reduced
+\emph{incertae sedis} taxa to intrude into their sibling taxa.  Therefore, a
+semantics can be given simply in terms of the exclude set for each taxonomy node.
+
+In order to determine how the exclude sets should be modified to account for
+\emph{incertae sedis} taxa, we first attempt to express the unmodified exclude
+sets in a helpful way.  We note that a node $n$ ordinarily excludes the
+descendants of $\mathcal{I}(s)$ for all of its siblings, as well as the siblings
+of all of its ancestors.  This leads to the following equation:
 \begin{align*}
-\mathcal{I}(m,n) & =
-    \begin{cases}
-        \emptyset & \text{if }n\text{ is \emph{incertae sedis}\,and }m\text{
-is not}\\
-        \mathcal{I}(m) & \text{otherwise}
-    \end{cases}
+\mathcal{E}(n) & =\left[\mathcal{I}(s)\big|a\in n+ancestors(n),s\in siblings(a)\right]\\
+ & =\left[\mathcal{I}(s)\big|s\in siblings(n)\right]\cup\left[\mathcal{I}(s)\big|a\in ancestors(n),s\in siblings(a)\right]\\
+ & =\left[\mathcal{I}(s)\big|s\in siblings(n)\right]\cup\mathcal{E}(parent(n)) \label{eq:exclude-set-formula-traditional}
 \end{align*}
 
-Additionally, the exclude set $\mathcal{E}(n)$ should contain the children of
-all the ancestors of $n$, unless those children are \emph{incertae sedis}.
-To avoid traversing all ancestors separately for each node $n$, we note that the
-excluded children of any ancestors will be in the exclude set of the parent of
-$n$ already. Therefore, we can write the exclude set of a node $n$ in terms of
-the exclude set of its parent:
-\begin{align}
-    \mathcal{E}(n) & = \mathcal{E}(parent(n))\cup\left[\mathcal{I}(m,n)\big|m\in
-           siblings(n)\right].\label{eq:exclude-set-formula-1}
-\end{align}
+When modifying the exclude set for a node $n$, we seek to find all the \emph{incertae
+sedis} taxa that may intrude into it. It turns out that the incertae sedis taxa that
+might intrude are just the \emph{incertae sedis} siblings of $n$ and all of its ancestors.
+\begin{align*}
+\mathcal{E}(n) & =\left[\mathcal{I}(s)\big|a\in n+ancestors(n),s\in siblings(a),s\text{ not \emph{incertae sedis}}\right]\\
+ & =\left[\mathcal{I}(s)\big|s\in siblings(n),s\text{ not \emph{incertae sedis}}\right]\cup\\
+ & \,\,\,\,\,\,\,\,\,\,\,\,\left[\mathcal{I}(s)\big|a\in ancestors(n),s\in siblings(a),s\text{ not \emph{incertae sedis}}\right]\\
+ & =\left[\mathcal{I}(s)\big|s\in siblings(n),s\text{ not \emph{incertae sedis}}\right]\cup\mathcal{E}(parent(n)).\label{eq:exclude-set-formula-1a}
+\end{align*}
 This formula allows us to compute exclude sets via a pre-order traversal on the
 taxonomy tree with exclude set for the root node defined to be $\emptyset$.
-
-\textbf{Problem:} how about the goal that we should be able to get the
-same outcome by moving a taxon up one level and marking it as incertae
-sedis? This might require that moving A next to B, and then moving B
-into A is allowed.
-
-\textbf{Problem:} should we be able to move incertae sedis taxa down
-into incertae sedis taxa? If not, then we must change the recursion.
-
-\subsubsection{An alternative semantics that allows placing an incertae sedis group within an incertae-sedis great-uncle}
-
-\bdrcomment{This semantics is actually designed to allow the situation in Figure \ref{fig:Placing-one-incertae}.
-
-  The problem is (might be) as follows: if we have the tree ``((?A1,A2)A,((D)C)B)MRCA;'', where A1 is incertae sedis, then we can place A1 within D.
-This semantics does not exclude A1 from B, C, or D.
-So, if A is broken, then nothing prohibits A1 from intruding into B, C, D, etc.
-
-  The purpose of this was to allow placing B as a siblign to A1 and then placing A1 within B -- \emph{if} B were incertae sedis.  But B is not incertae sedis.
-
-  We could try to fix up this situation by adding cases to the exclude rules to say that B excludes A1 if B is not incertae sedis.
-But adding cases makes things messier.
-I think it also does not actually solve the problem: suppose we mark B incertae sedis, but do NOT place it as sibling to A1.
-Then again if A is broken, A1 can intrude into B.
-We really want to allow A1 into B only if B is placed as sibling to A1, and I'm not sure how to do that with the partial-split-based semantics.
-If you see a way, then great!
-}
-
-If we want to allow placing two incertae sedis siblings within each
-other, then incertae sedis siblings are never excluded.
-
-Furthermore, consider placing $A$ as a sibling of $B$, and then
-placing $B$ within $A$.
-In order for this to work, $A$ must not
-exclude the incertae sedis children of $B$.
-
-In order to accomplish both of these goals, we define the descendants
-$\mathcal{I}^{\prime}(n)$ that are accessible without passing through
-an incertae sedis node.
-
-\begin{align*}
-    \mathcal{I^{\prime}}(n) & =
-    \begin{cases}
-        \emptyset & \text{if }n\in\mathcal{U}\\
-        \{n\} & \text{if }n\text{ is a leaf}\\
-        [\mathcal{I}^{\prime}(c)|c\in children(n)] & \text{otherwise.}
-\end{cases}
-\end{align*} We may then define the taxa excluded from $n$
-as:
-
-\begin{align*} 
-\mathcal{E}(n) & =\mathcal{E}(parent(n))\cup\left[\mathcal{I}(m,n)\big|m\in siblings(n)\right].\label{eq:exclude-set-formula-1}
-\end{align*}
-
-This semantics satisfies criteria 1,3,4, and 5 above.
-
-\subsubsection{A third semantics}
-
-\bdrcomment{I didn't really finish working this one out.}
-
-OK, the idea here is that we want to avoid inter-digitation.
-\emph{Incertae sedis} taxa thus exclude\emph{ incertae sedis}
-siblings.
-However, where the first semantics allows \emph{incertae
-sedis }taxa to intrude into \emph{incertae sedis} descendants of
-siblings, but not allow\emph{ incertae sedis} descendants of siblings
-to intrude into \emph{incertae sedis} taxa, here we do the reverse.
-
-OK, so the idea here is that an\emph{ incertae sedis} group excludes
-\begin{itemize} \item \emph{incertae sedis} siblings \item
-\emph{incertae sedis }descendants of ancestors \item but not\emph{
-incertae sedis} descendants of non-incertae-sedis siblings
-\end{itemize} It is also possible to allow placing $A$ as a sibling of
-$B$ and then placing $B$ within $A$ if we disallow placing $A$ into
-incertae sedis descendants of its siblings.
-We may then allow placing
-those descendants within $A$.
-
-\begin{align*} \mathcal{I^{\prime}}(n) & =\begin{cases} \emptyset &
-\text{if }n\in\mathcal{U}\\ \{n\} & \text{if }n\text{ is a leaf}\\{}
-[\mathcal{I}^{\prime}(c)|c\in children(n)] & \text{otherwise.}
-\end{cases} \end{align*}
-
-We use $\mathcal{I}(m,n)$ to indicate the descendants of $m$ that are
-excluded from $n$.
-\begin{align*} \mathcal{I}(m,n) & =\begin{cases}
-\emptyset & \text{if }n\text{ is \emph{incertae sedis}\,and }m\text{
-is not}\\ \mathcal{I}(m) & \text{if }n\text{ and }m\text{ are both
-}incertae\,sedis\\ \mathcal{I}^{\prime}(m) & \text{otherwise.}
-\end{cases} \end{align*}
-
-We may then define the taxa excluded from $n$ as:
-\begin{align*}
-\mathcal{E}(n) & =\begin{cases}
-\left[\mathcal{L}-\mathcal{I}(parent(n))\right]\cup[\mathcal{I}(s,n)|s\in siblings(n)] & \text{if }n\in\mathcal{U}\\
-\mathcal{E}(parent(n))\cup[\mathcal{I}(s,n)|s\in siblings(n)] & \text{otherwise.}
-\end{cases}
-\end{align*}
-I think this works... but need to double-check.
-
-Also: can we genericize $\mathcal{E}(parent(n))$?
-
-\subsection{Surprising results of the split-based semantics}
-
-1. Inter-digitation
-
-2. Placing A next to B, then B within A.
 
@@
 
+\subsection{Cases that this semantics can handle}
+
+\subsection{Surprising results of the split-based semantics}
+
+\subsubsection{Inter-digitation}
+1. Inter-digitation
+
+\bdrcomment{Ben: Add figure}
+
+\subsubsection{Placing one \emph{incertae sedis} group within another}
+It would be nice if our semantics of incertae sedis taxa would allow us to place
+one \emph{incertae sedis} sibling inside another.
+A special case of
+this is when a higher-rank incertae sedis taxon $A$ is placed as
+sister to an incertae sedis taxon $B$, and then $B$ is placed within
+$A$ (Figure \#).
+
+\begin{figure}
+\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-within-is/tree1\lyxdot tre}
+
+}\subfloat[Taxonomy]{\includegraphics[scale=0.5]{Figures/is-within-is/tax-shaded}
+
+}\subfloat[Synthesis tree]{\includegraphics[scale=0.5]{Figures/is-within-is/synth1-shaded}
+
+}
+
+\caption{\label{fig:Placing-one-incertae}Placing one incertae sedis group
+within another.
+Here I is placed within A, and then AI is placed within
+A.}
+\end{figure}
+
+2. Placing A next to B, then B within A.
+
+\bdrcomment{Ben: Write discussion of Figure \ref{fig:Placing-one-incertae}}
+
+\paragraph{Splits-based synthesis method}
+
+Here we focus on a split-based semantics instead of a semantics based on apply
+of a series of edit operations to the tree. The split-based semantics allows us
+to seek trees that are simultaneously consistent with a set of split-constraints.
+The use of a split-based semantics does impose some limitations. These come
+primarily because in the edit-based semantics, after we place an
+\emph{incertae sedis} taxon, it is no loner \emph{incertae sedis}, whereas with
+the split-based semantics taxa are either always \emph{incertae sedis} or never
+\emph{incertae sedis}.
+
+This has an effect, for example, when we consider two incertae sedis siblings $A$
+and $B$. If we can place $A$ within $B$, and $B$ within $A$, then a split- based
+semantics allows us to freely inter-digitate $A$ and $B$. If we do not want to
+allow $A$ and $B$ to be inter-digitated, then we cannot both $A$ within $B$ and $B$
+within $A$. In addition, we can imagine that a family $A$ is incertae sedis within a
+kingdom. In an edit-based semantics, we could place $A$ as a sibling to a genus $B$,
+mark $A$ as non-\emph{incertae-sedis}, and then place $B$ within $A$. In a split-based
+semantics, we must disallow either $B$ within $A$ or $A$ within $B$ if we want to disallow
+interdigitation.
+
+As a result of this, we cannot satisfy all five of these criteria. Thus, we
+describe two alternative split-based semantics. The non-interdigitation
+semantics gets \# (1,2,4,5), and the interdigitation semantics gets (1,3,4,5).
+This makes the point that each semantics results from different choices about
+which criteria to satisfy.
+
 \section{Synthesis and conflict resolution with incertae sedis taxa}
@@
 another.}
-
-In our semantics of incertae sedis taxa, it is possible for to place
-one \emph{incertae sedis} sibling inside another.
-A special case of
-this is when a higher-rank incertae sedis taxon $A$ is placed as
-sister to an incertae sedis taxon $B$, and then $B$ is placed within
-$A$ (Figure \#).
-
-\begin{figure}
-\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-within-is/tree1\lyxdot tre}
-
-}\subfloat[Taxonomy]{\includegraphics[scale=0.5]{Figures/is-within-is/tax-shaded}
-
-}\subfloat[Synthesis tree]{\includegraphics[scale=0.5]{Figures/is-within-is/synth1-shaded}
-
-}
-
-\caption{\label{fig:Placing-one-incertae}Placing one incertae sedis group
-within another.
-Here I is placed within A, and then AI is placed within
-A.}
-\end{figure}
 
@@
 
+\appendix
+\section{Alternative semantics}
+\subsubsection{semantics \#1b: modified to disallow inter-digitating
+\emph{incertae sedis} siblings}
+
+We additionally stipulate that it is not possible for \emph{incertae sedis} taxa
+to intrude into their \emph{incertae sedis} siblings.  We thus define $\mathcal{I}(m,n)$ to
+indicate descendants of $m$ that are excluded from $n$. Thus:
+\begin{align*}
+\mathcal{I}(m,n) & =
+    \begin{cases}
+        \emptyset & \text{if }n\text{ is \emph{incertae sedis}\,and }m\text{
+is not}\\
+        \mathcal{I}(m) & \text{otherwise}
+    \end{cases}
+\end{align*}
+
+We can write the exclude set of a node $n$ in terms of
+the exclude set of its parent:
+\begin{align}
+    \mathcal{E}(n) & = \mathcal{E}(parent(n))\cup\left[\mathcal{I}(m,n)\big|m\in
+           siblings(n)\right].\label{eq:exclude-set-formula-1}
+\end{align}
+\subsubsection{An alternative semantics that allows placing an incertae sedis group within an incertae-sedis great-uncle}
+
+\bdrcomment{This semantics is actually designed to allow the situation in Figure \ref{fig:Placing-one-incertae}.
+
+  The problem is (might be) as follows: if we have the tree ``((?A1,A2)A,((D)C)B)MRCA;'', where A1 is incertae sedis, then we can place A1 within D.
+This semantics does not exclude A1 from B, C, or D.
+So, if A is broken, then nothing prohibits A1 from intruding into B, C, D, etc.
+
+  The purpose of this was to allow placing B as a siblign to A1 and then placing A1 within B -- \emph{if} B were incertae sedis.  But B is not incertae sedis.
+
+  We could try to fix up this situation by adding cases to the exclude rules to say that B excludes A1 if B is not incertae sedis.
+But adding cases makes things messier.
+I think it also does not actually solve the problem: suppose we mark B incertae sedis, but do NOT place it as sibling to A1.
+Then again if A is broken, A1 can intrude into B.
+We really want to allow A1 into B only if B is placed as sibling to A1, and I'm not sure how to do that with the partial-split-based semantics.
+If you see a way, then great!
+}
+
+If we want to allow placing two incertae sedis siblings within each
+other, then incertae sedis siblings are never excluded.
+
+Furthermore, consider placing $A$ as a sibling of $B$, and then
+placing $B$ within $A$.
+In order for this to work, $A$ must not
+exclude the incertae sedis children of $B$.
+
+In order to accomplish both of these goals, we define the descendants
+$\mathcal{I}^{\prime}(n)$ that are accessible without passing through
+an incertae sedis node.
+
+\begin{align*}
+    \mathcal{I^{\prime}}(n) & =
+    \begin{cases}
+        \emptyset & \text{if }n\in\mathcal{U}\\
+        \{n\} & \text{if }n\text{ is a leaf}\\
+        [\mathcal{I}^{\prime}(c)|c\in children(n)] & \text{otherwise.}
+\end{cases}
+\end{align*} We may then define the taxa excluded from $n$
+as:
+
+\begin{align*} 
+\mathcal{E}(n) & =\mathcal{E}(parent(n))\cup\left[\mathcal{I}(m,n)\big|m\in siblings(n)\right].\label{eq:exclude-set-formula-1}
+\end{align*}
+
+This semantics satisfies criteria 1,3,4, and 5 above.
+
+\subsubsection{A third semantics}
+
+\bdrcomment{I didn't really finish working this one out.}
+
+OK, the idea here is that we want to avoid inter-digitation.
+\emph{Incertae sedis} taxa thus exclude\emph{ incertae sedis}
+siblings.
+However, where the first semantics allows \emph{incertae
+sedis }taxa to intrude into \emph{incertae sedis} descendants of
+siblings, but not allow\emph{ incertae sedis} descendants of siblings
+to intrude into \emph{incertae sedis} taxa, here we do the reverse.
+
+OK, so the idea here is that an\emph{ incertae sedis} group excludes
+\begin{itemize} \item \emph{incertae sedis} siblings \item
+\emph{incertae sedis }descendants of ancestors \item but not\emph{
+incertae sedis} descendants of non-incertae-sedis siblings
+\end{itemize} It is also possible to allow placing $A$ as a sibling of
+$B$ and then placing $B$ within $A$ if we disallow placing $A$ into
+incertae sedis descendants of its siblings.
+We may then allow placing
+those descendants within $A$.
+
+\begin{align*} \mathcal{I^{\prime}}(n) & =\begin{cases} \emptyset &
+\text{if }n\in\mathcal{U}\\ \{n\} & \text{if }n\text{ is a leaf}\\{}
+[\mathcal{I}^{\prime}(c)|c\in children(n)] & \text{otherwise.}
+\end{cases} \end{align*}
+
+We use $\mathcal{I}(m,n)$ to indicate the descendants of $m$ that are
+excluded from $n$.
+\begin{align*} \mathcal{I}(m,n) & =\begin{cases}
+\emptyset & \text{if }n\text{ is \emph{incertae sedis}\,and }m\text{
+is not}\\ \mathcal{I}(m) & \text{if }n\text{ and }m\text{ are both
+}incertae\,sedis\\ \mathcal{I}^{\prime}(m) & \text{otherwise.}
+\end{cases} \end{align*}
+
+We may then define the taxa excluded from $n$ as:
+\begin{align*}
+\mathcal{E}(n) & =\begin{cases}
+\left[\mathcal{L}-\mathcal{I}(parent(n))\right]\cup[\mathcal{I}(s,n)|s\in siblings(n)] & \text{if }n\in\mathcal{U}\\
+\mathcal{E}(parent(n))\cup[\mathcal{I}(s,n)|s\in siblings(n)] & \text{otherwise.}
+\end{cases}
+\end{align*}
+I think this works... but need to double-check.
+
+Also: can we genericize $\mathcal{E}(parent(n))$?
+
 \bibliographystyle{upmplainnat}

### 2018-04-26 doc/incertae_sedis.tex
Move some things around.

@@
 \begin{align*}
-\mathcal{E}(n) & =\left[\mathcal{I}(s)\big|a\in n+ancestors(n),s\in siblings(a)\right]\\
+  \mathcal{E}(n) & = \mathcal{L} - \mathcal{I}(n)\\
+ & =\left[\mathcal{I}(s)\big|a\in n+ancestors(n),s\in siblings(a)\right]\\
  & =\left[\mathcal{I}(s)\big|s\in siblings(n)\right]\cup\left[\mathcal{I}(s)\big|a\in ancestors(n),s\in siblings(a)\right]\\
@@
 This formula allows us to compute exclude sets via a pre-order traversal on the
-taxonomy tree with exclude set for the root node defined to be $\emptyset$.
+taxonomy tree with the exclude set for the root node defined to be $\emptyset$.
 
@@
 \end{figure}
+
+\bdrcomment{Ben: discuss naming + interdigitation}
 
@@
 
-\subsection{Cases that this semantics can handle}
-
-\subsection{Surprising results of the split-based semantics}
-
-\subsubsection{Inter-digitation}
-1. Inter-digitation
-
-\bdrcomment{Ben: Add figure}
-
-\subsubsection{Placing one \emph{incertae sedis} group within another}
-It would be nice if our semantics of incertae sedis taxa would allow us to place
-one \emph{incertae sedis} sibling inside another.
-A special case of
-this is when a higher-rank incertae sedis taxon $A$ is placed as
-sister to an incertae sedis taxon $B$, and then $B$ is placed within
-$A$ (Figure \#).
-
-\begin{figure}
-\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-within-is/tree1\lyxdot tre}
-
-}\subfloat[Taxonomy]{\includegraphics[scale=0.5]{Figures/is-within-is/tax-shaded}
-
-}\subfloat[Synthesis tree]{\includegraphics[scale=0.5]{Figures/is-within-is/synth1-shaded}
-
-}
-
-\caption{\label{fig:Placing-one-incertae}Placing one incertae sedis group
-within another.
-Here I is placed within A, and then AI is placed within
-A.}
-\end{figure}
-
-2. Placing A next to B, then B within A.
-
-\bdrcomment{Ben: Write discussion of Figure \ref{fig:Placing-one-incertae}}
-
-\paragraph{Splits-based synthesis method}
-
-Here we focus on a split-based semantics instead of a semantics based on apply
-of a series of edit operations to the tree. The split-based semantics allows us
-to seek trees that are simultaneously consistent with a set of split-constraints.
-The use of a split-based semantics does impose some limitations. These come
-primarily because in the edit-based semantics, after we place an
-\emph{incertae sedis} taxon, it is no loner \emph{incertae sedis}, whereas with
-the split-based semantics taxa are either always \emph{incertae sedis} or never
-\emph{incertae sedis}.
-
-This has an effect, for example, when we consider two incertae sedis siblings $A$
-and $B$. If we can place $A$ within $B$, and $B$ within $A$, then a split- based
-semantics allows us to freely inter-digitate $A$ and $B$. If we do not want to
-allow $A$ and $B$ to be inter-digitated, then we cannot both $A$ within $B$ and $B$
-within $A$. In addition, we can imagine that a family $A$ is incertae sedis within a
-kingdom. In an edit-based semantics, we could place $A$ as a sibling to a genus $B$,
-mark $A$ as non-\emph{incertae-sedis}, and then place $B$ within $A$. In a split-based
-semantics, we must disallow either $B$ within $A$ or $A$ within $B$ if we want to disallow
-interdigitation.
-
-As a result of this, we cannot satisfy all five of these criteria. Thus, we
-describe two alternative split-based semantics. The non-interdigitation
-semantics gets \# (1,2,4,5), and the interdigitation semantics gets (1,3,4,5).
-This makes the point that each semantics results from different choices about
-which criteria to satisfy.
-
 \section{Synthesis and conflict resolution with incertae sedis taxa}
-
-\subsection{Placement causes broken taxa}
-\label{conflict}
-Synthesis with \emph{incertae sedis} taxa has the potential to resolve
-uncertain taxon placements using information from phylogenies.
-Since
-phylogenies are ranked higher than taxonomy, they may place members of
-incertae sedis clades as they wish even when we do not take into
-account the semantics of incertae sedis taxa.
-However, without the
-incertae sedis semantics described above, placing a taxon $A$ within a
-taxon $B$ results in conflict with taxon $B$ in the taxonomy.
-This has
-three main effects: \begin{enumerate}
-    \item The edge leading to clade $B$ is lost.
-    \item Any taxonomy-only children of $B$ are placed at the
-        MRCA of the phylogenetically informed taxa, instead of at node $B$.
-    \item The name for taxon $B$ is lost.
-    \item We do not record $A$ as being placed within $B$.
-\end{enumerate} In order to avoid a situation
-where input phylogenies conflict with a large number of taxa when
-\emph{incertae sedis} taxa are placed within them, we have previously
-filtered \emph{incertae sedis} taxa from the taxonomy when
-constructing all prior synthesis trees.
-
-When the synthesis tree conflicts with a taxonomy node, we say that
-the taxon $B$ at that node is a broken taxon.
-Broken taxa have two
-main effects, both of which are negative.
-First, the name $B$ of the
-broken taxon is removed from the synthesis tree.
-Second, the
-conflicting edge for taxon $B$ is not included in the synthesis tree.
-This means that any children of $B$ that are taxonomy-only will not be
-placed with the children of $B$ that are mentioned in input
-phylogenies.
-Instead the taxonomy-only children of a broken taxon move
-towards the root of the synthesis tree and attach at the first higher-
-ranked taxon that is an ancestor of $B$ but is not broken.
-
-We therefore seek a synthesis method that can correctly introduce
-\emph{incertae sedis} taxa into containing taxa without breaking the
-containing taxa.
-We change the semantics of names to imply, not the
-exclusion of all non-included taxa, but only some non-included taxa,
-as described in section \ref{sec:Semantics-of-incertae}.
-As a result
-of this change in semantics, placing an IS taxon $A$ within a sister
-taxon $B$ no longer results in conflict with $B$.
-This allows us to
-retain the split for $B$ within the synthesis tree, so that taxonomy-
-only children of $B$ are correctly grouped with their siblings that
-are referenced by the input trees.
-We may then retain the name for the
-no-longer-broken taxon.
-Finally, we are then able to stop filtering
-\emph{incertae sedis} taxa, so that they appear in the synthesis tree.
-Thus, the synthesis tree is able to represent substantially more
-species, without suffering the loss of taxa and the loss of structure.
 
@@
 ott9, we do not move ott3 and ott2 up to ott11.
-{[}BDR: Extend more
-figures!{]}
+{[}BDR: Extend more figures!{]}
 
@@
 another.}
+\subsection{Surprising results of the split-based semantics}
+
+\subsubsection{Inter-digitation}
+1. Inter-digitation
+
+\bdrcomment{Ben: Add figure}
+
+\subsubsection{Placing one \emph{incertae sedis} group within another}
+It would be nice if our semantics of incertae sedis taxa would allow us to place
+one \emph{incertae sedis} sibling inside another.
+A special case of
+this is when a higher-rank incertae sedis taxon $A$ is placed as
+sister to an incertae sedis taxon $B$, and then $B$ is placed within
+$A$ (Figure \#).
+
+\begin{figure}
+\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-within-is/tree1\lyxdot tre}
+
+}\subfloat[Taxonomy]{\includegraphics[scale=0.5]{Figures/is-within-is/tax-shaded}
+
+}\subfloat[Synthesis tree]{\includegraphics[scale=0.5]{Figures/is-within-is/synth1-shaded}
+
+}
+
+\caption{\label{fig:Placing-one-incertae}Placing one incertae sedis group
+within another.
+Here I is placed within A, and then AI is placed within
+A.}
+\end{figure}
+
+2. Placing A next to B, then B within A.
+
+\bdrcomment{Ben: Write discussion of Figure \ref{fig:Placing-one-incertae}}
+
+\paragraph{Splits-based synthesis method}
+
+Here we focus on a split-based semantics instead of a semantics based on apply
+of a series of edit operations to the tree. The split-based semantics allows us
+to seek trees that are simultaneously consistent with a set of split-constraints.
+The use of a split-based semantics does impose some limitations. These come
+primarily because in the edit-based semantics, after we place an
+\emph{incertae sedis} taxon, it is no loner \emph{incertae sedis}, whereas with
+the split-based semantics taxa are either always \emph{incertae sedis} or never
+\emph{incertae sedis}.
+
+This has an effect, for example, when we consider two incertae sedis siblings $A$
+and $B$. If we can place $A$ within $B$, and $B$ within $A$, then a split- based
+semantics allows us to freely inter-digitate $A$ and $B$. If we do not want to
+allow $A$ and $B$ to be inter-digitated, then we cannot both $A$ within $B$ and $B$
+within $A$. In addition, we can imagine that a family $A$ is incertae sedis within a
+kingdom. In an edit-based semantics, we could place $A$ as a sibling to a genus $B$,
+mark $A$ as non-\emph{incertae-sedis}, and then place $B$ within $A$. In a split-based
+semantics, we must disallow either $B$ within $A$ or $A$ within $B$ if we want to disallow
+interdigitation.
+
+As a result of this, we cannot satisfy all five of these criteria. Thus, we
+describe two alternative split-based semantics. The non-interdigitation
+semantics gets \# (1,2,4,5), and the interdigitation semantics gets (1,3,4,5).
+This makes the point that each semantics results from different choices about
+which criteria to satisfy.
 

### 2018-04-26 doc/incertae_sedis.tex
odd label issues fixed

@@
 of all of its ancestors.  This leads to the following equation:
-\begin{align*}
+\begin{eqnarray*}
   \mathcal{E}(n) & = \mathcal{L} - \mathcal{I}(n)\\
@@
  & =\left[\mathcal{I}(s)\big|s\in siblings(n)\right]\cup\left[\mathcal{I}(s)\big|a\in ancestors(n),s\in siblings(a)\right]\\
- & =\left[\mathcal{I}(s)\big|s\in siblings(n)\right]\cup\mathcal{E}(parent(n)) \label{eq:exclude-set-formula-traditional}
-\end{align*}
+ & =\left[\mathcal{I}(s)\big|s\in siblings(n)\right]\cup\mathcal{E}(parent(n)) \label{eq:exsetformtraditional}
+\end{eqnarray*}
 
@@
 \begin{align*}
-\mathcal{E}(n) & =\left[\mathcal{I}(s)\big|a\in n+ancestors(n),s\in siblings(a),s\text{ not \emph{incertae sedis}}\right]\\
+\mathcal{E}(n) & =
+    \left[\mathcal{I}(s)\big|a\in n+ancestors(n),s\in siblings(a),s\text{ not \emph{incertae sedis}}\right]\\
 & =\left[\mathcal{I}(s)\big|s\in siblings(n),s\text{ not \emph{incertae sedis}}\right]\cup\\
 & \,\,\,\,\,\,\,\,\,\,\,\,\left[\mathcal{I}(s)\big|a\in ancestors(n),s\in siblings(a),s\text{ not \emph{incertae sedis}}\right]\\
- & =\left[\mathcal{I}(s)\big|s\in siblings(n),s\text{ not \emph{incertae sedis}}\right]\cup\mathcal{E}(parent(n)).\label{eq:exclude-set-formula-1a}
+& =\left[\mathcal{I}(s)\big|s\in siblings(n),s\text{ not \emph{incertae sedis}}\right]\cup\mathcal{E}(parent(n)).
+\label{eq:exsetformonea}
 \end{align*}
@@
 
-Equation (\ref{eq:exclude-set-formula-1}) leads to the following
+Equation (\ref{eq:exsetformone}) leads to the following
 algorithm to compute the exclude set for all nodes in a tree.
@@
 the exclude set of its parent:
-\begin{align}
+\begin{eqnarray*}
     \mathcal{E}(n) & = \mathcal{E}(parent(n))\cup\left[\mathcal{I}(m,n)\big|m\in
-           siblings(n)\right].\label{eq:exclude-set-formula-1}
-\end{align}
+           siblings(n)\right].
+           \label{eq:exsetformone}
+\end{eqnarray*}
 \subsubsection{An alternative semantics that allows placing an incertae sedis group within an incertae-sedis great-uncle}
@@
 
-\begin{align*} 
-\mathcal{E}(n) & =\mathcal{E}(parent(n))\cup\left[\mathcal{I}(m,n)\big|m\in siblings(n)\right].\label{eq:exclude-set-formula-1}
-\end{align*}
+\begin{eqnarray*} 
+\mathcal{E}(n) & =\mathcal{E}(parent(n))\cup\left[\mathcal{I}(m,n)\big|m\in siblings(n)\right].\label{eq:exsetformoneb}
+\end{eqnarray*}
 

### 2018-04-26 doc/incertae_sedis.tex
Fix some issues with equations and labels.

@@
 of all of its ancestors.  This leads to the following equation:
-\begin{eqnarray*}
-  \mathcal{E}(n) & = \mathcal{L} - \mathcal{I}(n)\\
- & =\left[\mathcal{I}(s)\big|a\in n+ancestors(n),s\in siblings(a)\right]\\
- & =\left[\mathcal{I}(s)\big|s\in siblings(n)\right]\cup\left[\mathcal{I}(s)\big|a\in ancestors(n),s\in siblings(a)\right]\\
- & =\left[\mathcal{I}(s)\big|s\in siblings(n)\right]\cup\mathcal{E}(parent(n)) \label{eq:exsetformtraditional}
-\end{eqnarray*}
+\begin{align}
+\mathcal{E}(n) & = \mathcal{L} - \mathcal{I}(n)\nonumber\\
+ & =\left[\mathcal{I}(s)\big|a\in n+ancestors(n),s\in siblings(a)\right]\nonumber\\
+ & =\left[\mathcal{I}(s)\big|s\in siblings(n)\right]\cup\left[\mathcal{I}(s)\big|a\in ancestors(n),s\in siblings(a)\right]\nonumber\\
+ & =\left[\mathcal{I}(s)\big|s\in siblings(n)\right]\cup\mathcal{E}(parent(n)).\label{eq:exsetformonea-traditional}
+\end{align}
 
@@
 might intrude are just the \emph{incertae sedis} siblings of $n$ and all of its ancestors.
-\begin{align*}
+\begin{align}
 \mathcal{E}(n) & =
-    \left[\mathcal{I}(s)\big|a\in n+ancestors(n),s\in siblings(a),s\text{ not \emph{incertae sedis}}\right]\\
-& =\left[\mathcal{I}(s)\big|s\in siblings(n),s\text{ not \emph{incertae sedis}}\right]\cup\\
-& \,\,\,\,\,\,\,\,\,\,\,\,\left[\mathcal{I}(s)\big|a\in ancestors(n),s\in siblings(a),s\text{ not \emph{incertae sedis}}\right]\\
-& =\left[\mathcal{I}(s)\big|s\in siblings(n),s\text{ not \emph{incertae sedis}}\right]\cup\mathcal{E}(parent(n)).
-\label{eq:exsetformonea}
-\end{align*}
+    \left[\mathcal{I}(s)\big|a\in n+ancestors(n),s\in siblings(a),s\text{ not \emph{incertae sedis}}\right]\nonumber\\
+& =\left[\mathcal{I}(s)\big|s\in siblings(n),s\text{ not \emph{incertae sedis}}\right]\cup\nonumber\\
+& \,\,\,\,\,\,\,\,\,\,\,\,\left[\mathcal{I}(s)\big|a\in ancestors(n),s\in siblings(a),s\text{ not \emph{incertae sedis}}\right]\nonumber\\
+& =\left[\mathcal{I}(s)\big|s\in siblings(n),s\text{ not \emph{incertae sedis}}\right]\cup\mathcal{E}(parent(n)).\label{eq:exsetformonea}
+\end{align}
 This formula allows us to compute exclude sets via a pre-order traversal on the

### 2018-04-26 doc/incertae_sedis.tex
Rename subsection.
Add interdigitation figure and discuss in naming section.

@@
 
-\subsubsection{Split-based semantics}
+\subsubsection{Terminology for rooted splits}
 \paragraph{Equivalence of a taxonomy tree and a list of rooted splits}
@@
 
+\begin{figure}
+\hfill{}\subfloat[Taxonomy tree]{\includegraphics[scale=0.5]{Figures/name1nodes2/tax-shaded}}
+\hfill{}\subfloat[Solution tree]{\includegraphics[scale=0.5]{Figures/name1nodes2/solution-shaded}}
+\hfill{}
+\caption{\label{fig:One-name-can}One name can be consistent with multiple nodes.}
+\end{figure}
+
 After constructing a supertree, tip nodes already have names in \emph{$\mathcal{L}$}.
@@
 \emph{incertae sedis} taxa should be added to an existing taxon.
-\begin{figure}
-\hfill{}\subfloat[Taxonomy tree]{\includegraphics[scale=0.5]{Figures/name1nodes2/tax-shaded}}
-\hfill{}\subfloat[Solution tree]{\includegraphics[scale=0.5]{Figures/name1nodes2/solution-shaded}}
-\hfill{}
-\caption{\label{fig:One-name-can}One name can be consistent with multiple nodes.}
-\end{figure}
-
-It is also possible for multiple names to apply to a single node. For example,
-in Figure \ref{fig:Multiple-names-can}a, the taxon $A$ corresponds to the split
-$B1\,B2\,C\,|\bullet\,Y$, and its child taxon $B$ corresponds to the split $B1\,B2\,|\bullet\,Y$. The
-edge leading to $A$ is consistent with the split for $B$, but the name $B$ is
-applied to the node with the smallest include group. However, in the
-solution tree (Fig. \ref{fig:Multiple-names-can}b), there is is only node for
-both names $A$ and $B$ to apply to. In this case, we solve this problem by
-introducing a monotypic parent, and applying $A$ to the newly created parent node
-(Fig. \ref{fig:Multiple-names-can}c).
+
 \begin{figure}
@@
 
-\bdrcomment{Ben: discuss naming + interdigitation}
+It is also possible for multiple names to apply to a single node. For example,
+in Figure \ref{fig:Multiple-names-can}a, the taxon $A$ corresponds to the split
+$B1\,B2\,C\,|\bullet\,Y$, and its child taxon $B$ corresponds to the split $B1\,B2\,|\bullet\,Y$. The
+edge leading to $A$ is consistent with the split for $B$, but the name $B$ is
+applied to the node with the smallest include group. However, in the
+solution tree (Fig. \ref{fig:Multiple-names-can}b), there is is only node for
+both names $A$ and $B$ to apply to.  This kind of situation arises when a taxon
+becomes monotypic after its \emph{incertae sedis} children are placed elsewhere.
+For such cases, we can solve this problem by (i) selecting a taxon at the
+node that is ancestral to all other taxa and (ii) moving this taxon to a
+newly-introduced out-degree-1 parent.  For example, $A$ is moved to an
+out-degree-1 parent in Fig. \ref{fig:Multiple-names-can}c.  We repeatedly perform
+this procedure until no taxon at the node is the parent of all taxa.
+
+However, there are cases where thus procedure does not result in one name per
+node.  This case can occur, for example, when two incertae sedis siblings are 
+``inter-digitated'', or  combined into a single clade in such a way that their
+members cannot be separated.  When the members of the two clades share the same
+MRCA, as in Fig. \ref{fig:interdigitation}, then the names for both clades
+apply to the same node.  However, in this case, neither name is ancestral to
+the other, and the previous solution does not apply.  In such cases, we choose
+one of the names arbitrarily.
+
+\begin{figure}
+\hfill{}\subfloat[Taxonomy tree]{\includegraphics[scale=0.5]{Figures/is-interdigitation/tax-shaded}}
+\hfill{}\subfloat[Solution tree]{\includegraphics[scale=0.5]{Figures/is-interdigitation/synth-shaded}}
+\hfill{}
+\caption{\label{fig:interdigitation}Interdigitation of \emph{incertae sedis} siblings.
+Here the names $A$ and $B$ apply to the same node on the solution tree, but neither is ancestral to the other on the taxonomy.}
+\end{figure}
 
@@
 
-\bdrcomment{Ben: Add figure}
+\bdrcomment{Ben: discuss figure \ref{fig:interdigitation}.}
 

### 2018-04-26 doc/incertae_sedis.tex
Update.

@@
 \subsection{Background}
-
+\label{sec:background}
 \paragraph{Previous algorithm}
@@
 
-\note{     1 Details & Methods
-
-1.1 Some examples in OTT This paragraph pulled from intro to
-otcetera/doc /handling-incertae-sedis.pdf
-
-The problem is becoming more acute because NCBI is putting a larger
-number of taxa into groups that are marked as ``classified''. or
-example, when OTT 2.9 was created NCBI's clasification of the bird
-family Sylvidae included a group five genera that were placed within
-``[http://www.ncbi.nlm.nih.gov/Taxonomy/Brows er/wwwtax.cgi?mode=Undef&
-id=36270&lvl=3&lin=f&keep=1&srchmode=1&unlock||unclassi fied
-Sylviidae].'' This includes the genus Regulus.
-Thus in OTT 2.9 Regulus
-(ott ID = 3599326) is placed inside the family Silvidae (OTT ID =
-259942).
-The ``unclassified Sylviidae'' does not appear in OTT; instead
-Regulus is flagged as ``unclassified,sibling higher'' and all of the
-species within Regulus are flagged as ``unclassified inherited.''
-
-The taxonomy (as of version 2.9) also contains 685 cases of taxa that
-are flagged as both ``unclassified inherited'' and ``unclassified''based
-on grep unclass.*unclassified\_inherited taxonomy.tsv.
-
-We would like to stop suppressing (pruning) incertae sedis taxa, so
-that groups such as Regulus can appear appear in the synthetic tree.
-
-While we get Regulus back, we do not end up placing it in Sylviidae
-because Sylviidae is broken.
-It conflicts with about 20 different
-trees.
-
-1.2 Taxonomy details
-
-Flags.
-The Open Tree's reference taxonomy is produced by a tool called
-smasher that notices hints that a taxon is incertae sedis and labels
-that taxon with one of five flagssee https://github.com/OpenTreeOfLife
-/reference-taxonomy/wiki /Taxon-flags.This document is not concerned
-with the flagging system per se, so ``incertae sedis'' will be used here
-to refer to all of the flags that denote taxa with uncertain
-placement..
-While a taxonomist may intend to use incertae sedis to
-indicate a limited number of possible positions for a taxon to go in
-the taxonomy, we do not retain any such details.
-Container nodes.
-Some taxonomies have nodes with names like ``Incertae
-sedis (Bacteria)''.
-Such nodes are called containers.
-They indicate
-that each child node is an incertae sedis child of the container
-node's parent.
-The OTT retains container nodes with the was_container
-flag, but seems to have already moved all of the container's children
-to the container's parent.
-}
 \begin{figure}
@@
 
-\section{Synthesis and conflict resolution with incertae sedis taxa}
+\section{The rooted-split semantics in practice}
+
+In order to examine the results of the rooted splits-semantics above, we explore
+the outcome on a series of complex cases.  These cases are intended to
+illustrate some of the issues that arise when developing a semantics of
+\emph{incertae sedis}, as well as the outcomes for the rooted-split semantics.
+We think that they constitute one of the main contributions of the paper.  In
+each case, the synthesis trees in this section are constructed using the
+BUILD-based algorithm described in section \ref{sec:background}, without
+decomposing them into smaller subproblems or pruning and then unpruning
+taxonomy-only taxa.  Synthesis trees that are described as \emph{incertae
+sedis}-aware apply the BUILD algorithm to splits constructed using equation
+\ref{eq:exsetformonea}, whereas the synthesis trees that are not \emph{incertae
+sedis}-aware are constructed from splits using equation
+\ref{eq:exsetformonea-traditional}.
+
+Much of the complexity that we explore here comes from the fact that it is not
+only tip taxa that can be \emph{incertae sedis}.  It is also possible to have
+higher ranked  \emph{incertae sedis} taxa, such as genera that are marked
+\emph{incertae familia}.
+
+\subsection{The \emph{incertae sedis} semantics restores taxonomic groups and names}
+
+With the naive semantics, intrusion of an incertae sedis taxon into a sibling is
+interpreted as conflict with that sibling.  This results in the loss of the edge
+corresponding to the sibling taxon as well as the taxon name.
+
+We see from Fig. \ref{fig:Handling-incertae-sedis}d that use of the incertae sedis
+semantics results the taxonomy edge $AB$ grouping $A$ and $B$.  It also restores the
+names $AB$ and $A$.  We may thus determine that $C$ has been placed within $A$ (and
+its ancestor $AB$).
 
@@
 
+\section{Details \& Methods}
+
+\subsection{Cases of incertae sedis taxa}
+1.1 Some examples in OTT This paragraph pulled from intro to
+otcetera/doc /handling-incertae-sedis.pdf
+
+The problem is becoming more acute because NCBI is putting a larger
+number of taxa into groups that are marked as ``classified''. For
+example, when OTT 2.9 was created NCBI's clasification of the bird
+family Sylvidae included a group five genera that were placed within
+\url{http://www.ncbi.nlm.nih.gov/Taxonomy/Browser/wwwtax.cgi?mode=Undef&id=36270&lvl=3&lin=f&keep=1&srchmode=1&unlock||unclassified Sylviidae}.
+This includes the genus Regulus. Thus in OTT 2.9 Regulus (ott ID = 3599326) is placed inside the family Silvidae (OTT ID =
+259942).
+The ``unclassified Sylviidae'' does not appear in OTT; instead
+Regulus is flagged as ``unclassified,sibling higher'' and all of the
+species within Regulus are flagged as ``unclassified inherited.''
+
+The taxonomy (as of version 2.9) also contains 685 cases of taxa that
+are flagged as both ``unclassified inherited'' and ``unclassified'' based
+on {\tt grep unclass.*unclassified\_inherited taxonomy.tsv}.
+
+We would like to stop suppressing (pruning) incertae sedis taxa, so
+that groups such as Regulus can appear appear in the synthetic tree.
+
+\bdrcomment{We get Regulus back, bu we do not end up placing it in Sylviidae
+because Sylviidae is broken. It conflicts with about 20 different trees.}
+
+\subsection{Taxonomy details}
+
+\paragraph{Flags.}
+The Open Tree's reference taxonomy is produced by a tool called
+smasher that notices hints that a taxon is incertae sedis and labels
+that taxon with one of five flags (see https://github.com/OpenTreeOfLife
+/reference-taxonomy/wiki /Taxon-flags). This document is not concerned
+with the flagging system per se, so ``incertae sedis'' will be used here
+to refer to all of the flags that denote taxa with uncertain
+placement.
+
+\paragraph{Limited attachment range}
+While a taxonomist may intend to use incertae sedis to
+indicate a limited number of possible positions for a taxon to go in
+the taxonomy, we do not retain any such details.
+
+\paragraph{Container nodes.}
+Some taxonomies have nodes with names like ``Incertae sedis (Bacteria)''.
+Such nodes are called containers. They indicate that each child node is an
+incertae sedis child of the container node's parent. The OTT retains container
+nodes with the was\_container flag, but seems to have already moved all of the
+container's children to the container's parent.
+
 \bibliographystyle{upmplainnat}

### 2018-04-26 doc/incertae_sedis.tex
Add a word.
Update.

@@
 the outcome on a series of complex cases.  These cases are intended to
-illustrate some of the issues that arise when developing a semantics of
+illustrate some of the issues that arise when developing an operational semantics of
 \emph{incertae sedis}, as well as the outcomes for the rooted-split semantics.
@@
 
-When \emph{incertae sedis} clades are allowed, it is possible for an
-\emph{incertae sedis} taxon to be nested within another \emph{incertae
-sedis} taxon.
-In such a case we must allow the higher-ranked taxon to
-be placed within a sibling, while preserving the right of the lower
-ranked taxon to be placed within a lower-level sibling (Figure
-\ref{fig:Nested-incertae-sedis}).
-
 \begin{figure}
@@
 \end{figure}
+When \emph{incertae sedis} clades are allowed, it is possible for an
+\emph{incertae sedis} taxon to be nested within another \emph{incertae
+sedis} taxon.
+In such a case we must allow the higher-ranked taxon to
+be placed within a sibling, while preserving the right of the lower
+ranked taxon to be placed within a lower-level sibling (Figure
+\ref{fig:Nested-incertae-sedis}).
 
@@
 
-\paragraph{Splits-based synthesis method}
-
-Here we focus on a split-based semantics instead of a semantics based on apply
+\subsection{Splits-based semantics versus edit-based semantics}
+
+Here we focus on a split-based semantics instead of a semantics based on applying
 of a series of edit operations to the tree. The split-based semantics allows us
@@
 
-\subsection{Case 1}
-
-taxonomy = (((a1,a2)A,(b1,b2)B)AB,(c1,c2)?C,D)root;
-
-\subsubsection{If we place $c$ within $A$}
-
-tree1 = (((a1,c1),b1),d1);
-
-synth-with-is: ((((a1,(c1,c2)C),a2)A,(b1,b2)B)AB),(d1)D)
-
-placement: C within A <- AB
-
-synth-no-is: ((a1,(c1,c2)C),a1,(b1,b2)B,(d1)D)
-
-\subsubsection{If we place $c$ within $A$ but break $C$}
-
-tree1 = ((a1,c1),(a2,c1),b1)
-
-synth-with-is: ((((a1,c1),(a2,c2))A,(b1,b2)B)AB, (d1)D)
-
-synth-no-is: ??
-
-placement:  \begin{itemize} \item c1 <- (broken) C within A <- AB.
-\item c2<- (broken) C within A <- AB \end{itemize} result: we lose
-$C$, but keep $A$ and $AB$.
-
-\subsection{Case 2: nested is}
-
-taxonomy = (((a1,a2)A,(b1,b2)B),( (
-(c11,c12)C1,(c21,c22)C2,(ci1,ci2)?CI )?C) ,(d1)D)
-
-tree1 = ((a1,(c11,(ci1,c12))),b1)
-
-synth-with-is: places CI wihin C1, and C within A.
-
-\subsection{Case 3: interleaving taxa}
-
-taxonomy = ((a1,a2)A,(b1,b2)?B,(c1,c2)?c)
-
-input tree: ((b1,c1),(b2,c2))
-
-synth-with-is: ((a1,a2)A,((b1,c1),(b2,c2))B) or
-((a1,a2)A,((b1,c1),(b2,c2))C)
-
-Result: we name the interleaved taxon $B$ or $C$, but cannot name it
-both names.
-
-\subsection{Case 4: creation of monotypic taxa}
-
-see test cases.
-\begin{comment} \begin{lyxcode}
-\textbf{Whale:}~ ?ott6145835~{[}6145835~Brandtocetus/genus{]}~<-~ott69
-8424~{[}698424~Cetacea/orde r{]}~placed~under~ott4942423~{[}4942423~Ce
-totherium~/~genus{]}~<-~ott698424~{[}6 98424~Cetacea/order{]}
-
-\textbf{Whale:}~?ott6145873~{[}6145873{]}~<-~ott698424~{[}698424~Cetac
-\ea~/~order
-\{]}~placed~under~ott4942423~{[}4942423{]}~<-~ott698424~{[}698424{]}~
-
--~Cetacea~broken.
-
-\textbf{Insect:}~?ott18090~{[}18090{]}~<-~ott672501~{[}672501{]}~place
-\d~under~ot t275501~{[}275501{]}~<-~ott672501~{[}672501{]}
-
-\textbf{Insect:}~?ott190269~{[}190269{]}~<-~ott672501~{[}672501{]}~pla
-\ced~under~ ott275501~{[}275501{]}~<-~ott672501~{[}672501{]}~
-
-\textbf{Insect:}~?ott349638~{[}349638{]}~<-~ott672501~{[}672501{]}~pla
-\ced~under~ ott275501~{[}275501{]}~<-~ott672501~{[}672501{]}~
-
-\textbf{Insect:}~?ott1009313~{[}1009313{]}~<-~ott672501~{[}672501~Redu
-\viidae~/~f amily{]}~placed~under~ott275501~{[}275501{]}~<-~ott672501~
-\{[}672501{]}~
-
--~Reduviidae~broken.
-
-\textbf{Insect:~}?ott564110~{[}564110{]}~<-~ott939130~{[}939130~Thyrid
-\idae~/~fam
-\ily{]}~placed~under~ott757344~{[}757344{]}~<-~ott939130~{[}939130{]}~
-
-\textbf{Plant:}~?ott679485~{[}679485{]}~<-~ott1074933~{[}1074933~Lamio
-\idea~/~sub family{]}~placed~under~ott534796~{[}534796{]}~<-~
-
-\textbf{Plant:}~?ott729184~{[}729184{]}~<-~ott1074933~{[}1074933~Lamio
-\idea~/~sub family~{]}~placed~under~ott534796~{[}534796{]}~<-~ott10749
-\33~{[}1074933{]}~
-
-ott1074933~{[}1074933{]}~
-
--~Lamioidea~\emph{not}~broken
-
-\textbf{Plant:}~?ott5144555~{[}5144555{]}~<-~ott23373~{[}23373{]}~plac
-\ed~under~o tt447802~{[}447802{]}~<-~ott23373~{[}23373{]}
-
--~Orobanchaceae~broken
-
-\textbf{Fly:}~?ott441413~{[}441413~Rhodesiella~lungliensis/species{]}~
-\<-~ott1014 022~{[}1014022~Rubiaceae{]}~placed~under~ott571842~{[}5718
-\42~Pavetteae~/~tribe{ ]}~<-~ott997381~{[}997381~Ixoroidea/subfamily{]
-\}~<-~ott1014022~{[}1014022~Rubia ceae/family{]}
-
-\textbf{Plant:}~?ott31033~{[}31033{]}~<-~ott627035~{[}627035{]}~placed
-\~under~ott 6019314~{[}6019314{]}~<-~ott627035~{[}627035{]}~
-
-\textbf{Bacteria:}~?ott4018918~{[}4018918{]}~<-~ott5248084~{[}5248084{
-\]}~placed~ under~ott167767~{[}167767{]}~<-~ott5248084~{[}5248084{]}~
-
-\textbf{Bacteria:}?ott4792759~{[}4792759{]}~<-~ott5248084~{[}5248084{]
-\}~placed~u nder~ott167767~{[}167767{]}~<-~ott5248084~{[}5248084{]}~
-
--~Erysipelotrichaceae~broken 
-\end{lyxcode} \end{comment}
 Should we do this? We could say:  \begin{itemize} \item there are 331
@@
 
+\section{Placements}
+\textbf{Whale:}~ ?ott6145835~{[}6145835~Brandtocetus/genus{]}~<-~ott69
+8424~{[}698424~Cetacea/orde r{]}~placed~under~ott4942423~{[}4942423~Ce
+totherium~/~genus{]}~<-~ott698424~{[}6 98424~Cetacea/order{]}
+
+\textbf{Whale:}~?ott6145873~{[}6145873{]}~<-~ott698424~{[}698424~Cetac
+\ea~/~order
+\{]}~placed~under~ott4942423~{[}4942423{]}~<-~ott698424~{[}698424{]}~
+
+-~Cetacea~broken.
+
+\textbf{Insect:}~?ott18090~{[}18090{]}~<-~ott672501~{[}672501{]}~place
+\d~under~ot t275501~{[}275501{]}~<-~ott672501~{[}672501{]}
+
+\textbf{Insect:}~?ott190269~{[}190269{]}~<-~ott672501~{[}672501{]}~pla
+\ced~under~ ott275501~{[}275501{]}~<-~ott672501~{[}672501{]}~
+
+\textbf{Insect:}~?ott349638~{[}349638{]}~<-~ott672501~{[}672501{]}~pla
+\ced~under~ ott275501~{[}275501{]}~<-~ott672501~{[}672501{]}~
+
+\textbf{Insect:}~?ott1009313~{[}1009313{]}~<-~ott672501~{[}672501~Redu
+\viidae~/~f amily{]}~placed~under~ott275501~{[}275501{]}~<-~ott672501~
+\{[}672501{]}~
+
+-~Reduviidae~broken.
+
+\textbf{Insect:~}?ott564110~{[}564110{]}~<-~ott939130~{[}939130~Thyrid
+\idae~/~fam
+\ily{]}~placed~under~ott757344~{[}757344{]}~<-~ott939130~{[}939130{]}~
+
+\textbf{Plant:}~?ott679485~{[}679485{]}~<-~ott1074933~{[}1074933~Lamio
+\idea~/~sub family{]}~placed~under~ott534796~{[}534796{]}~<-~
+
+\textbf{Plant:}~?ott729184~{[}729184{]}~<-~ott1074933~{[}1074933~Lamio
+\idea~/~sub family~{]}~placed~under~ott534796~{[}534796{]}~<-~ott10749
+\33~{[}1074933{]}~
+
+ott1074933~{[}1074933{]}~
+
+-~Lamioidea~\emph{not}~broken
+
+\textbf{Plant:}~?ott5144555~{[}5144555{]}~<-~ott23373~{[}23373{]}~plac
+\ed~under~o tt447802~{[}447802{]}~<-~ott23373~{[}23373{]}
+
+-~Orobanchaceae~broken
+
+\textbf{Fly:}~?ott441413~{[}441413~Rhodesiella~lungliensis/species{]}~
+\<-~ott1014 022~{[}1014022~Rubiaceae{]}~placed~under~ott571842~{[}5718
+\42~Pavetteae~/~tribe{ ]}~<-~ott997381~{[}997381~Ixoroidea/subfamily{]
+\}~<-~ott1014022~{[}1014022~Rubia ceae/family{]}
+
+\textbf{Plant:}~?ott31033~{[}31033{]}~<-~ott627035~{[}627035{]}~placed
+\~under~ott 6019314~{[}6019314{]}~<-~ott627035~{[}627035{]}~
+
+\textbf{Bacteria:}~?ott4018918~{[}4018918{]}~<-~ott5248084~{[}5248084{
+\]}~placed~ under~ott167767~{[}167767{]}~<-~ott5248084~{[}5248084{]}~
+
+\textbf{Bacteria:}?ott4792759~{[}4792759{]}~<-~ott5248084~{[}5248084{]
+\}~placed~u nder~ott167767~{[}167767{]}~<-~ott5248084~{[}5248084{]}~
+
+-~Erysipelotrichaceae~broken
+
 \bibliographystyle{upmplainnat}

### 2018-05-03 doc/incertae_sedis.tex
compiles

@@
 \section{Placements}
-\textbf{Whale:}~ ?ott6145835~{[}6145835~Brandtocetus/genus{]}~<-~ott69
-8424~{[}698424~Cetacea/orde r{]}~placed~under~ott4942423~{[}4942423~Ce
-totherium~/~genus{]}~<-~ott698424~{[}6 98424~Cetacea/order{]}
-
-\textbf{Whale:}~?ott6145873~{[}6145873{]}~<-~ott698424~{[}698424~Cetac
-\ea~/~order
-\{]}~placed~under~ott4942423~{[}4942423{]}~<-~ott698424~{[}698424{]}~
+\textbf{Whale:}~ ?ott6145835~{[}6145835~Brandtocetus/genus{]}~<-~ott698424~{[}698424~Cetacea/orde r{]}~placed~under~ott4942423~{[}4942423~Cetotherium~/~genus{]}~<-~ott698424~{[}6 98424~Cetacea/order{]}
+
+\textbf{Whale:}~?ott6145873~{[}6145873{]}~<-~ott698424~{[}698424~Cetacea~/~order{]}~placed~under~ott4942423~{[}4942423{]}~<-~ott698424~{[}698424{]}~
 
@@
 
-\textbf{Insect:}~?ott18090~{[}18090{]}~<-~ott672501~{[}672501{]}~place
-\d~under~ot t275501~{[}275501{]}~<-~ott672501~{[}672501{]}
-
-\textbf{Insect:}~?ott190269~{[}190269{]}~<-~ott672501~{[}672501{]}~pla
-\ced~under~ ott275501~{[}275501{]}~<-~ott672501~{[}672501{]}~
-
-\textbf{Insect:}~?ott349638~{[}349638{]}~<-~ott672501~{[}672501{]}~pla
-\ced~under~ ott275501~{[}275501{]}~<-~ott672501~{[}672501{]}~
-
-\textbf{Insect:}~?ott1009313~{[}1009313{]}~<-~ott672501~{[}672501~Redu
-\viidae~/~f amily{]}~placed~under~ott275501~{[}275501{]}~<-~ott672501~
-\{[}672501{]}~
+\textbf{Insect:}~?ott18090~{[}18090{]}~<-~ott672501~{[}672501{]}~placed~under~ot t275501~{[}275501{]}~<-~ott672501~{[}672501{]}
+
+\textbf{Insect:}~?ott190269~{[}190269{]}~<-~ott672501~{[}672501{]}~placed~under~ ott275501~{[}275501{]}~<-~ott672501~{[}672501{]}~
+
+\textbf{Insect:}~?ott349638~{[}349638{]}~<-~ott672501~{[}672501{]}~placed~under~ ott275501~{[}275501{]}~<-~ott672501~{[}672501{]}~
+
+\textbf{Insect:}~?ott1009313~{[}1009313{]}~<-~ott672501~{[}672501~Reduviidae~/~f amily{]}~placed~under~ott275501~{[}275501{]}~<-~ott672501~{[}672501{]}~
 
@@
 
-\textbf{Insect:~}?ott564110~{[}564110{]}~<-~ott939130~{[}939130~Thyrid
-\idae~/~fam
-\ily{]}~placed~under~ott757344~{[}757344{]}~<-~ott939130~{[}939130{]}~
-
-\textbf{Plant:}~?ott679485~{[}679485{]}~<-~ott1074933~{[}1074933~Lamio
-\idea~/~sub family{]}~placed~under~ott534796~{[}534796{]}~<-~
-
-\textbf{Plant:}~?ott729184~{[}729184{]}~<-~ott1074933~{[}1074933~Lamio
-\idea~/~sub family~{]}~placed~under~ott534796~{[}534796{]}~<-~ott10749
-\33~{[}1074933{]}~
+\textbf{Insect:~}?ott564110~{[}564110{]}~<-~ott939130~{[}939130~Thyrididae~/~family{]}~placed~under~ott757344~{[}757344{]}~<-~ott939130~{[}939130{]}~
+
+\textbf{Plant:}~?ott679485~{[}679485{]}~<-~ott1074933~{[}1074933~Lamioidea~/~sub family{]}~placed~under~ott534796~{[}534796{]}~<-~
+
+\textbf{Plant:}~?ott729184~{[}729184{]}~<-~ott1074933~{[}1074933~Lamioidea~/~sub family~{]}~placed~under~ott534796~{[}534796{]}~<-~ott1074933~{[}1074933{]}~
 
@@
 
-\textbf{Plant:}~?ott5144555~{[}5144555{]}~<-~ott23373~{[}23373{]}~plac
-\ed~under~o tt447802~{[}447802{]}~<-~ott23373~{[}23373{]}
+\textbf{Plant:}~?ott5144555~{[}5144555{]}~<-~ott23373~{[}23373{]}~placed~under~o tt447802~{[}447802{]}~<-~ott23373~{[}23373{]}
 
@@
 
-\textbf{Fly:}~?ott441413~{[}441413~Rhodesiella~lungliensis/species{]}~
-\<-~ott1014 022~{[}1014022~Rubiaceae{]}~placed~under~ott571842~{[}5718
-\42~Pavetteae~/~tribe{ ]}~<-~ott997381~{[}997381~Ixoroidea/subfamily{]
-\}~<-~ott1014022~{[}1014022~Rubia ceae/family{]}
-
-\textbf{Plant:}~?ott31033~{[}31033{]}~<-~ott627035~{[}627035{]}~placed
-\~under~ott 6019314~{[}6019314{]}~<-~ott627035~{[}627035{]}~
-
-\textbf{Bacteria:}~?ott4018918~{[}4018918{]}~<-~ott5248084~{[}5248084{
-\]}~placed~ under~ott167767~{[}167767{]}~<-~ott5248084~{[}5248084{]}~
-
-\textbf{Bacteria:}?ott4792759~{[}4792759{]}~<-~ott5248084~{[}5248084{]
-\}~placed~u nder~ott167767~{[}167767{]}~<-~ott5248084~{[}5248084{]}~
+\textbf{Fly:}~?ott441413~{[}441413~Rhodesiella~lungliensis/species{]}~<-~ott1014 022~{[}1014022~Rubiaceae{]}~placed~under~ott571842~{[}571842~Pavetteae~/~tribe{ ]}~<-~ott997381~{[}997381~Ixoroidea/subfamily{]}~<-~ott1014022~{[}1014022~Rubia ceae/family{]}
+
+\textbf{Plant:}~?ott31033~{[}31033{]}~<-~ott627035~{[}627035{]}~placed~under~ott 6019314~{[}6019314{]}~<-~ott627035~{[}627035{]}~
+
+\textbf{Bacteria:}~?ott4018918~{[}4018918{]}~<-~ott5248084~{[}5248084{]}~placed~ under~ott167767~{[}167767{]}~<-~ott5248084~{[}5248084{]}~
+
+\textbf{Bacteria:}?ott4792759~{[}4792759{]}~<-~ott5248084~{[}5248084{]}~placed~u nder~ott167767~{[}167767{]}~<-~ott5248084~{[}5248084{]}~
 

### 2018-05-04 doc/incertae_sedis.tex
emph only

@@
 First, the naive approach results in a less-resolved tree.  While input trees are
-free to place incertae sedis taxa within a sibling, the naive approach
+free to place \emph{incertae sedis} taxa within a sibling, the naive approach
 interprets this placement as conflict with the sibling.  Since the taxonomy

### 2018-05-07 doc/incertae_sedis.tex
minor

@@
     \item An \emph{incertae sedis} node has an unrestricted range within the descendants of its sibling taxa.
-    \item The semantics of an \emph{incertae sedis} node does not depend on side-information such as taxon ranks, but only on the taxonomy tree.
-    \item Semantics is based on deriving a rooted (partial) split for each branch of the taxonomy.
+    \item The semantics of an \emph{incertae sedis} node does not depend on additional information such as taxon ranks, but only on the taxonomy tree.
+    \item The semantics is based on deriving a rooted (partial) split for each branch of the taxonomy.
       % These could be goals, but maybe discuss them as side-effects:
@@
 
+%\subsubsection{Terminology for rooted splits}
+
 The rooted split associated with taxonomy node $n$ serves to determine both when $n$ is in conflict
@@
 % actually doesn't but that is surprising, and therefore kind of a result.
-
-\subsubsection{Unrestricted range and ignoring side information}
+% \paragraph{Definition of rooted splits}
+As is typically done, we define rooted splits for each node by noting that
+removing the edge connecting the node to its parent would produce two trees.
+The leaf label sets of these trees
+the partition the full leaf set $\mathcal{L}$ into two groups.
+The include set $\mathcal{I}(n)$ which
+does not contain the root, and the exclude set $\mathcal{E}(n)$ which does contain the root.
+Such a split may be written
+\[ \mathcal{I}(n)|\bullet\mathcal{E}(n).\]
+where the $\bullet$ indicates the root. 
+The include set is the set of labels for leaves that are
+descendants of the node.
+The exclude set is the set of
+labels for leaves that are not descendants of that node. 
+If no taxa are \emph{incertae sedis},
+then the exclude set for a node is just the total tip set minus the include set
+for the node: $\mathcal{E}(n)  =\mathcal{L}-\mathcal{I}(n)$.
+%For a node $n$ on the tip-ward side of an edge $e$, we
+%may also write $\mathcal{I}(n)$ for $\mathcal{I}(e)$, and $\mathcal{E}(n)$ for $\mathcal{E}(e)$.
+For any two rooted splits $A=A_{1}|\bullet A_{2}$
+and $B=B_{1}|\bullet B_{2}$ , we say that $A$ \emph{displays }$B$ if $B_{1}\subseteq A_{1}$ and $B_{2}\subseteq A_{2}$.
+
+Note that when a taxon rooted at node $n$ is not marked as \emph{incertae sedis},
+then the leaf labels that are descendants of $n$ will be in the include set for
+$n$ and all its ancestors and will be in the exclude set for all other nodes.
+
+In terms of phylogenetic nomenclature the supertree pipeline acts as if the
+taxonomy defines a conditional node-based name \citep[see][]{deQueiroz2013} for each
+higher taxon.
+So for the name labeling node $n$ in the taxonomic
+tree, the operational definition would be: ``the clade defined by the most recent
+common ancestor (MRCA) of $\mathcal{I}(n)$ as long as that MRCA is not an ancestor of
+$\mathcal{E}(n)$.''
+OTT is not based on a set of explicitly defined phylogenetic name definitions.
+In many contexts, a branch-based definition for $n$ (``the clade by all of the taxa more
+closely related to all fo them members of $\mathcal{I}(n)$ than to any member of
+$\mathcal{E}(n)$, if such a clade exists'') would result in similar behavior
+during the tree construction. 
+An exception to this (discussed below) arises in the step of assigning names to the 
+    supertree.
+
+\paragraph{Equivalence of a taxonomy tree and a list of rooted splits}
+
+Note that, for the purposes of the supertree method, all of the information from
+the taxonomic tree is contained in the set of rooted splits. Thus, one could
+imagine replacing that taxonomy, $\taxonomy_{p}$, with a set of rooted splits $\mathcal{R}(\taxonomy_{p})$ such that
+each edge in $\taxonomy_{p}$ is encoded in one split in $\mathcal{R}(\taxonomy_{p})$. If the order of trees is set
+to be: $\mathcal{R}(\taxonomy_{p})=\mathcal{Z}(\taxonomy_{p}) + 
+\mathcal{Y}(\taxonomy_{p})$, then using $\mathcal{R}(\taxonomy_{p})$ in place of $\taxonomy_{p}$ in the supertree
+algorithm would produce the same output.
+
+\subsubsection{Unrestricted range and ignoring additional information}
 The choice to allow each \emph{incertae sedis} taxon to attach anywhere within
@@
 First, more precise information about where a taxon can attach is easy to
-incorporate. If $\mathcal{P}$ is the set of taxa that are potential ancestors for \emph{incertae
+incorporate. 
+If $\mathcal{P}$ is the set of taxa that are potential ancestors for \emph{incertae
 sedis} taxon $x$, then we may simply remove the descendants of $x$ from the exclude
@@
 Unfortunately, reliable information about the range of valid attachment points
-for each \emph{incertae sedis} taxon is lacking.  Ideally, the \emph{incertae
+for each \emph{incertae sedis} taxon is lacking in OTT and most other large-scale
+taxonomies.
+Ideally, the \emph{incertae
   sedis} annotation for some taxon $x$ would be accompanied by a specific statement
@@
 
-\subsubsection{Terminology for rooted splits}
-\paragraph{Equivalence of a taxonomy tree and a list of rooted splits}
-
-Note that, for the purposes of the supertree method, all of the information from
-the taxonomic tree is contained in the set of rooted splits. Thus, one could
-imagine replacing that taxonomy, $\taxonomy_{p}$, with a set of rooted splits $\mathcal{R}(\taxonomy_{p})$ such that
-each edge in $\taxonomy_{p}$ is encoded in one split in $\mathcal{R}(\taxonomy_{p})$. If the order of trees is set
-to be: $\mathcal{R}(\taxonomy_{p})=\mathcal{Z}(\taxonomy_{p}) + 
-\mathcal{Y}(\taxonomy_{p})$, then using $\mathcal{R}(\taxonomy_{p})$ in place of $\taxonomy_{p}$ in the supertree
-algorithm would produce the same output.
-
-\paragraph{Definition of rooted splits}
-As is typically done, we define rooted splits for each edge by noting that
-removing the edge would produce two trees. The leaf label sets of these trees
-the partition the full leaf set $\mathcal{L}$ into two groups. The include set $\mathcal{I}(e)$ which
-does not contain the root, and the exclude set $\mathcal{E}(e)$ which does contain the root.
-Such a split may be written
-\[ \mathcal{I}(e)|\bullet\mathcal{E}(e).\]
-where the $\bullet$ indicates the root. The include set is the set of leaves that are
-descendants of the daughter node of the edge. The exclude set is the set of
-leaves that are not descendants of that node. If no taxa are \emph{incertae sedis},
-then the exclude set for a node is just the total tip set minus the include set
-for the node: $\mathcal{E}(e)  =\mathcal{L}-\mathcal{I}(e)$. For a node $n$ on the tip-ward side of an edge $e$, we
-may also write $\mathcal{I}(n)$ for $\mathcal{I}(e)$, and $\mathcal{E}(n)$ for $\mathcal{E}(e)$. For any two rooted splits $A=A_{1}|\bullet A_{2}$
-and $B=B_{1}|\bullet B_{2}$ , we say that $A$ \emph{displays }$B$ if $B_{1}\subseteq A_{1}$ and $B_{2}\subseteq A_{2}$.
-
-Note that when a taxon rooted at node $n$ is not marked as \emph{incertae sedis},
-then the leaf labels that are descendants of $n$ will be in the include set for
-$n$ and all its ancestors and will be in the exclude set for all other nodes/edges.
-
-In terms of phylogenetic nomenclature the supertree pipeline acts as if the
-taxonomy defines a branch-based name \citep[see][]{deQueiroz2013} for each
-higher taxon of the form. So for the name labeling node $n$ in the taxonomic
-tree, the operational definition would be: ``the clade defined by all taxa that
-are more closely related to $\mathcal{I}(n)$ than to $\mathcal{E}(n)$, if such a clade exists on the tree.''
-
 \subsection{Split-based semantics for \emph{incertae sedis} taxa}
@@
 An \emph{incertae sedis} taxon $n$ differs from a normal taxon because it can be
-moved into the descendants of any of its siblings.  Thus, annotating $n$ as
-\emph{incertae sedis} really modifies the meanings of its siblings so that they
-no longer exclude $n$.  In terms of rooted splits, we can represent this lack of
-exclusion by omitting taxa from the exclude set for a node.  Such a split does
+moved into the descendants of any of its siblings.
+Thus, annotating $n$ as
+\emph{incertae sedis} really modifies the meanings of the taxanomic statements
+that of the siblings of $n$ and all taxonomic nodes that descend from those siblings.
+ so that they
+no longer exclude $n$.
+If we denote this set of taxa (the set of taxa that are {\emph{potential}} 
+ancestors for \emph{incertae sedis} taxon $n$) as $\mathcal{P}_n$, 
+then we may simply remove the descendants of $n$ from the exclude
+set for each $p\in\mathcal{P}_n$.
+%In terms of rooted splits, we can represent this lack of
+%   exclusion by omitting taxa from the exclude set for a node.
+% We seek to define modified splits for each branch of the taxonomy tree.
+The include sets of these splits remain unchanged, but the exclude sets may be reduced
+\emph{incertae sedis} taxa to intrude into their sibling taxa.
+% Therefore, a semantics can be given simply in terms of the exclude set for each taxonomy node.
+Such a split does
 not mention the complete set of tip labels $\mathcal{L}$ and is therefore called a
@@
 the supertree problem.
-
-We seek to define modified splits for each branch of the taxonomy tree. The
-include sets of these splits remain unchanged, but the exclude sets may be reduced
-\emph{incertae sedis} taxa to intrude into their sibling taxa.  Therefore, a
-semantics can be given simply in terms of the exclude set for each taxonomy node.
 

### 2018-05-07 doc/incertae_sedis.tex
up to naming section

@@
 
-%\subsubsection{Terminology for rooted splits}
+\subsection{Terminology for rooted splits}
 
@@
 
+\paragraph{Equivalence of a taxonomy tree and a list of rooted splits}
+
+Note that, for the purposes of the supertree method, all of the information from
+the taxonomic tree is contained in the set of rooted splits. Thus, one could
+imagine replacing that taxonomy, $\taxonomy_{p}$, with a set of rooted splits $\mathcal{R}(\taxonomy_{p})$ such that
+each edge in $\taxonomy_{p}$ is encoded in one split in $\mathcal{R}(\taxonomy_{p})$. If the order of trees is set
+to be: $\mathcal{R}(\taxonomy_{p})=\mathcal{Z}(\taxonomy_{p}) + 
+\mathcal{Y}(\taxonomy_{p})$, then using $\mathcal{R}(\taxonomy_{p})$ in place of $\taxonomy_{p}$ in the supertree
+algorithm would produce the same output.
+
+\paragraph{Equivalence rooted split definitions and phylogenetic definitions of names}
 In terms of phylogenetic nomenclature the supertree pipeline acts as if the
@@
 
-\paragraph{Equivalence of a taxonomy tree and a list of rooted splits}
-
-Note that, for the purposes of the supertree method, all of the information from
-the taxonomic tree is contained in the set of rooted splits. Thus, one could
-imagine replacing that taxonomy, $\taxonomy_{p}$, with a set of rooted splits $\mathcal{R}(\taxonomy_{p})$ such that
-each edge in $\taxonomy_{p}$ is encoded in one split in $\mathcal{R}(\taxonomy_{p})$. If the order of trees is set
-to be: $\mathcal{R}(\taxonomy_{p})=\mathcal{Z}(\taxonomy_{p}) + 
-\mathcal{Y}(\taxonomy_{p})$, then using $\mathcal{R}(\taxonomy_{p})$ in place of $\taxonomy_{p}$ in the supertree
-algorithm would produce the same output.
-
-\subsubsection{Unrestricted range and ignoring additional information}
-The choice to allow each \emph{incertae sedis} taxon to attach anywhere within
-the descendants of its siblings requires some explanation, for two reasons.
-First, more precise information about where a taxon can attach is easy to
-incorporate. 
-If $\mathcal{P}$ is the set of taxa that are potential ancestors for \emph{incertae
-sedis} taxon $x$, then we may simply remove the descendants of $x$ from the exclude
-set for each $p\in\mathcal{P}$.  Such an approach would require that the path from $x$ to $p$
-go through only other potential ancestors.
-
-Second, as mentioned above, it may be expected that an \emph{incertae sedis} taxon
-attach only a specific points within the taxonomy. For example, a genus
-might be characterized as \emph{incertae sedis} at the level of the subfamily.
-Thus, it could be placed inside any of the hypothesized sub-families, or within
-its own subfamily, but it would be unexpected for it to be placed inside one of
-the other genera.
-
+\subsection{Split-based semantics for \emph{incertae sedis} taxa}
+
+An \emph{incertae sedis} taxon $n$ differs from a normal taxon because it can be
+placed as a descendant of some set of taxa without invalidating these taxa.
+We will use $\mathcal{P}_n$ to denote this set of taxa which are {\emph{potential}} 
+ancestors for \emph{incertae sedis} taxon $n$.
+Note that this distinct from the taxa that are ancestral to $n$ in the taxonomy;
+    those are taxa which must contain all of the descandants of $n$.
+The effect of the \emph{incertae sedis} annotation of taxon $n$ is
+    to simply remove $\mathcal{I}(n)$ from the exclude set for each $p\in\mathcal{P}_n$.
+The include sets of these splits remain unchanged, but the exclude sets may be reduced
+\emph{incertae sedis} taxa to intrude.
+Such a split does
+not mention the complete set of tip labels $\mathcal{L}$ and is therefore called a
+`partial split'', as opposed to a ``full split'' that contains the entirely of $\mathcal{L}$.
+
+When considering \emph{incertae sedis} annotations, the taxonomy is still comprehensive 
+if we include the degenerate grouping for the base
+    node $\rho$ which has $\mathcal{I}(\rho) = \mathcal{L}$ and $\mathcal{E}(\rho) = \emptyset$,
+    but many of the taxonomic nodes will correspond to partial splits.
+Thus, when interpreted correctly the \emph{incertae sedis} 
+annotations are reducing the amount of information that the taxonomy brings to
+the supertree problem.
+
+%  of any of its siblings.
+% Thus, annotating $n$ as
+% \emph{incertae sedis} really modifies the meanings of the taxanomic statements (rooted splits)
+% of other
+% that of the siblings of $n$ and all taxonomic nodes that descend from those siblings.
+%  so that they
+% no longer exclude $n$.
+% %In terms of rooted splits, we can represent this lack of
+% %   exclusion by omitting taxa from the exclude set for a node.
+% % We seek to define modified splits for each branch of the taxonomy tree.
+%  into their sibling taxa.
+% % Therefore, a semantics can be given simply in terms of the exclude set for each taxonomy node.
+
+\subsection{Unrestricted range and ignoring additional information}
+In principle, an \emph{incertae sedis} taxon $n$ could carry with it additional
+    information that indicates which specific points within the taxonomic tree
+    it could attach.
 Unfortunately, reliable information about the range of valid attachment points
@@
 taxonomies.
-Ideally, the \emph{incertae
-  sedis} annotation for some taxon $x$ would be accompanied by a specific statement
-about which other taxa might include $x$.  However, OTT does not provide this
-directly.  Constructing a range for each incertae sedis taxon relies on detailed
+Constructing a range for each incertae sedis taxon relies on detailed
 readings of character argumentation (which is lacking in OTT) or rank-based
@@
 
-\subsection{Split-based semantics for \emph{incertae sedis} taxa}
-
-An \emph{incertae sedis} taxon $n$ differs from a normal taxon because it can be
-moved into the descendants of any of its siblings.
-Thus, annotating $n$ as
-\emph{incertae sedis} really modifies the meanings of the taxanomic statements
-that of the siblings of $n$ and all taxonomic nodes that descend from those siblings.
- so that they
-no longer exclude $n$.
-If we denote this set of taxa (the set of taxa that are {\emph{potential}} 
-ancestors for \emph{incertae sedis} taxon $n$) as $\mathcal{P}_n$, 
-then we may simply remove the descendants of $n$ from the exclude
-set for each $p\in\mathcal{P}_n$.
-%In terms of rooted splits, we can represent this lack of
-%   exclusion by omitting taxa from the exclude set for a node.
-% We seek to define modified splits for each branch of the taxonomy tree.
-The include sets of these splits remain unchanged, but the exclude sets may be reduced
-\emph{incertae sedis} taxa to intrude into their sibling taxa.
-% Therefore, a semantics can be given simply in terms of the exclude set for each taxonomy node.
-Such a split does
-not mention the complete set of tip labels $\mathcal{L}$ and is therefore called a
-`partial split'', as opposed to a ``full split'' that contains the entirely of $\mathcal{L}$.
-Thus, when interpreted correctly the \emph{incertae sedis} 
-annotations are reducing the amount of information that the taxonomy brings to
-the supertree problem.
-
-In order to determine how the exclude sets should be modified to account for
-\emph{incertae sedis} taxa, we first attempt to express the unmodified exclude
-sets in a helpful way.  We note that a node $n$ ordinarily excludes the
-descendants of $\mathcal{I}(s)$ for all of its siblings, as well as the siblings
-of all of its ancestors.  This leads to the following equation:
+Specifically, the methods described here only consider cases in which 
+    $\mathcal{P}_n$ can be determined from the structure of the tree.
+Our method takes $\mathcal{P}_n$ to be the set of nodes that
+    are in the taxonomic subtrees rooted at the sibling taxa of \emph{incertae sedis} taxon $n$.
+This allows us to encode the exclude set of some node $x$ efficiently by sweeping
+    over the tree from root to tip.
+If $\mathcal{S}(x)$ is the set of siblings of $x$ which are not annotated as being 
+    \emph{incertae sedis} and $p(x)$ is the parent of $x$, then
 \begin{align}
-\mathcal{E}(n) & = \mathcal{L} - \mathcal{I}(n)\nonumber\\
- & =\left[\mathcal{I}(s)\big|a\in n+ancestors(n),s\in siblings(a)\right]\nonumber\\
- & =\left[\mathcal{I}(s)\big|s\in siblings(n)\right]\cup\left[\mathcal{I}(s)\big|a\in ancestors(n),s\in siblings(a)\right]\nonumber\\
- & =\left[\mathcal{I}(s)\big|s\in siblings(n)\right]\cup\mathcal{E}(parent(n)).\label{eq:exsetformonea-traditional}
+    \mathcal{E}(x) = \mathcal{E}(p(x)) \cup \bigcup_{s\in \mathcal{S}(x)} \mathcal{E}(s)
 \end{align}
-
-When modifying the exclude set for a node $n$, we seek to find all the \emph{incertae
-sedis} taxa that may intrude into it. It turns out that the incertae sedis taxa that
-might intrude are just the \emph{incertae sedis} siblings of $n$ and all of its ancestors.
-\begin{align}
-\mathcal{E}(n) & =
-    \left[\mathcal{I}(s)\big|a\in n+ancestors(n),s\in siblings(a),s\text{ not \emph{incertae sedis}}\right]\nonumber\\
-& =\left[\mathcal{I}(s)\big|s\in siblings(n),s\text{ not \emph{incertae sedis}}\right]\cup\nonumber\\
-& \,\,\,\,\,\,\,\,\,\,\,\,\left[\mathcal{I}(s)\big|a\in ancestors(n),s\in siblings(a),s\text{ not \emph{incertae sedis}}\right]\nonumber\\
-& =\left[\mathcal{I}(s)\big|s\in siblings(n),s\text{ not \emph{incertae sedis}}\right]\cup\mathcal{E}(parent(n)).\label{eq:exsetformonea}
-\end{align}
-This formula allows us to compute exclude sets via a pre-order traversal on the
-taxonomy tree with the exclude set for the root node defined to be $\emptyset$.
+with $\mathcal{E}(\rho) = \emptyset$ for the root of the taxonomic tree.
+
+%If $\mathcal{P}$ is the set of taxa that are potential ancestors for \emph{incertae
+%  sedis} taxon $x$, then we may simply remove the descendants of $x$ from the exclude
+%  set for each $p\in\mathcal{P}$.  Such an approach would require that the path from $x$ to $p$
+%  go through only other potential ancestors.
+
+% Supporting more precise information about $\mathcal{P}_n$  for each \emph{incertae sedis}
+%    taxon $n$ is an important future direction to be explored.
+% For example, a genus
+% might be characterized as \emph{incertae sedis} at the level of the subfamily.
+% Thus, it could be placed inside any of the hypothesized sub-families, or within
+% its own subfamily, but it would be unexpected for it to be placed inside one of
+% the other genera.
+
+% In order to determine how the exclude sets should be modified to account for
+% \emph{incertae sedis} taxa, we first attempt to express the unmodified exclude
+% sets in a helpful way.  We note that a node $n$ ordinarily excludes the
+% descendants of $\mathcal{I}(s)$ for all of its siblings, as well as the siblings
+% of all of its ancestors.  This leads to the following equation:
+% \begin{align}
+% \mathcal{E}(n) & = \mathcal{L} - \mathcal{I}(n)\nonumber\\
+%  & =\left[\mathcal{I}(s)\big|a\in n+ancestors(n),s\in siblings(a)\right]\nonumber\\
+%  & =\left[\mathcal{I}(s)\big|s\in siblings(n)\right]\cup\left[\mathcal{I}(s)\big|a\in ancestors(n),s\in siblings(a)\right]\nonumber\\
+%  & =\left[\mathcal{I}(s)\big|s\in siblings(n)\right]\cup\mathcal{E}(parent(n)).\label{eq:exsetformonea-traditional}
+% \end{align}
+
+% When modifying the exclude set for a node $n$, we seek to find all the \emph{incertae
+% sedis} taxa that may intrude into it. It turns out that the incertae sedis taxa that
+% might intrude are just the \emph{incertae sedis} siblings of $n$ and all of its ancestors.
+% \begin{align}
+% \mathcal{E}(n) & =
+%     \left[\mathcal{I}(s)\big|a\in n+ancestors(n),s\in siblings(a),s\text{ not \emph{incertae sedis}}\right]\nonumber\\
+% & =\left[\mathcal{I}(s)\big|s\in siblings(n),s\text{ not \emph{incertae sedis}}\right]\cup\nonumber\\
+% & \,\,\,\,\,\,\,\,\,\,\,\,\left[\mathcal{I}(s)\big|a\in ancestors(n),s\in siblings(a),s\text{ not \emph{incertae sedis}}\right]\nonumber\\
+% & =\left[\mathcal{I}(s)\big|s\in siblings(n),s\text{ not \emph{incertae sedis}}\right]\cup\mathcal{E}(parent(n)).\label{eq:exsetformonea}
+% \end{align}
+% This formula allows us to compute exclude sets via a pre-order traversal on the
+% taxonomy tree with the exclude set for the root node defined to be $\emptyset$.
 

### 2018-05-07 doc/incertae_sedis.tex
minor changes to avoid name repetition across figures 1 and 2

@@
 apply to multiple nodes. For example, in Figure \ref{fig:One-name-can}b, the
-name $A$ can apply to nodes $x$ and $y$. Here the name $A$ corresponds to the split
-$A1\,A2\,|\bullet B\,C$, leaving out $D$ since it is \emph{incertae sedis}. The two nodes $x$
-and $y$ display the splits $A1\,A2\,|\bullet B\,C\,D$ and $A1\,A2\,D|\bullet
-B\,C$ respectively, and both
-of these splits display the split $A1\,A2\,|\bullet B\,C$, so the name $A$ can apply to both
-$x$ and $y$. This cannot happen without\emph{ incertae sedis} taxa except at
-monotypic nodes. When faced with a choice about where to place a name, our
+name $E$ can apply to nodes $x$ and $y$. Here the name $E$ corresponds to the split
+$E1\,E2\,|\bullet F\,G$, leaving $H$ out of the definition since $H$ is \emph{incertae sedis}. 
+The two nodes $x$
+and $y$ display the splits $E1\,E2\,|\bullet F\,G\,H$ and $E1\,E2\,H|\bullet
+F\,G$ respectively, and both
+of these splits display the split $E1\,E2\,|\bullet F\,G$, so the name $E$ can apply to both
+$x$ and $y$. 
+This cannot happen when the rooted splits that correspond to a taxonomic name
+    are generated without\emph{ incertae sedis} taxa except at
+monotypic nodes.
+
+This situation is the result of the fact that it is not clear whether the names
+    in the taxonomic tree should be treated as node-based or branch-based
+    phylogenetic defintions.
+A node-based definition of $E$ would be ``the clade rooted at the MRCA of $E1$ and $E2$ as long as it is not
+    the ancestor of $F$ or $G$''; this would correspond to node $y$ in the solution
+    depicted in figure Figure \ref{fig:One-name-can}b.
+The branch-based definition would take the form ``The clade containing all lineages more closely related to 
+   $E1$ and $E2$ than to either $F$ or $G$, if such a clade exists''; this definition
+   would include node $x$ and the lineage connecting it to its parent.
+The ambiguity arises because both definitions refer to the same node in the taxonomy,
+    so it is unclear which is a preferable way to define the taxon concept of $E$.
+
+When faced with a choice about where to place a name, our
 solution is to find the most tip-ward node where the name can apply and attach
-the name to this node. This corresponds to a conservative choice about whether
-\emph{incertae sedis} taxa should be added to an existing taxon.
+the name to this node.
+This corresponds to using a node-based definition of the taxonomic names.
+The result of this naming choice is that the algorithm is conservative with respect
+    to when the \emph{incertae sedis} taxa should be placed within another named
+    taxon.
+
+Note that the intrusion of an \emph{incertae sedis} taxon within another taxon does not
+    always lead to ambiguity in which node should be named. 
+For example, $H$ is placed inside taxon $EF$ in Figure \ref{fig:One-name-can}, but the
+    the name $EF$ only applies to one node in the solution tree.
 

### 2018-05-07 doc/incertae_sedis.tex
figure 3

@@
 It is also possible for multiple names to apply to a single node. For example,
-in Figure \ref{fig:Multiple-names-can}a, the taxon $A$ corresponds to the split
-$B1\,B2\,C\,|\bullet\,Y$, and its child taxon $B$ corresponds to the split $B1\,B2\,|\bullet\,Y$. The
-edge leading to $A$ is consistent with the split for $B$, but the name $B$ is
-applied to the node with the smallest include group. However, in the
+in Figure \ref{fig:Multiple-names-can}a, the taxon $M$ corresponds to the split
+$L1\,L2\,K\,|\bullet\,J$, and its child taxon $L$ corresponds to the split $L1\,L2\,|\bullet\,Y$.
+The edge leading to $M$ is consistent with the split for $L$, but the name $L$ is
+applied to the node with the smallest include group.
+However, in the
 solution tree (Fig. \ref{fig:Multiple-names-can}b), there is is only node for
-both names $A$ and $B$ to apply to.  This kind of situation arises when a taxon
+both names $M$ and $L$ to apply to.
+This kind of situation arises when a taxon
 becomes monotypic after its \emph{incertae sedis} children are placed elsewhere.
-For such cases, we can solve this problem by (i) selecting a taxon at the
+For such cases, we can address this synonymy by (i) selecting a taxon at the
 node that is ancestral to all other taxa and (ii) moving this taxon to a
-newly-introduced out-degree-1 parent.  For example, $A$ is moved to an
+newly-introduced out-degree-1 parent.  For example, $M$ is moved to an
 out-degree-1 parent in Fig. \ref{fig:Multiple-names-can}c.  We repeatedly perform
@@
 However, there are cases where thus procedure does not result in one name per
-node.  This case can occur, for example, when two incertae sedis siblings are 
+node.  This case can occur, for example, when two \emph{incertae sedis} siblings are 
 ``inter-digitated'', or  combined into a single clade in such a way that their

### 2018-05-07 doc/incertae_sedis.tex
figure 4
figure 5

@@
 newly-introduced out-degree-1 parent.  For example, $M$ is moved to an
-out-degree-1 parent in Fig. \ref{fig:Multiple-names-can}c.  We repeatedly perform
+out-degree-1 parent in Fig. \ref{fig:Multiple-names-can}c.
+We repeatedly perform
 this procedure until no taxon at the node is the parent of all taxa.
@@
 apply to the same node.  However, in this case, neither name is ancestral to
-the other, and the previous solution does not apply.  In such cases, we choose
-one of the names arbitrarily.
+the other, and the previous solution does not apply.
+In such cases, we treat all applicable names as synonyms for the node.
+\mthcomment{Need to verify that the namer records this situation as synonyms!}
 
@@
 
-\subsection{Placement}\label{subsec:Placement}
+\subsection{Taxonomic revision}\label{subsec:Placement}
 
@@
 Figure \ref{fig:Broken-incertae-sedis}.
-Here taxon $C1$ is an descendant of $A$ in the
+Here taxon $T1$ is an descendant of $R$ in the
 synthesis tree, but not the taxonomy.
-Thus $C1$ is placed within $A$.
-We note that the most recent common ancestor (MRCA) of $A$ and $C1$ in
-the taxonomy is the root node, and the path to the MRCA from $A$ is
-$A\to AB\to root$, while the path to the MRCA from $C1$ is $C1\to C\to
+Thus $T1$ is placed within $R$.
+We note that the MRCA of $R$ and $T1$ in
+the taxonomy is the root node, and the path to the MRCA from $R$ is
+$R\to RS\to root$, while the path to the MRCA from $T1$ is $T1\to T\to
 root$.
-Here the additional node $AB$ on the path to the MRCA from $A$
-indicates that $C1$ is placed within $AB$ as well as $A$.
+Here the additional node $RS$ on the path to the MRCA from $R$
+indicates that $T1$ is placed within $RS$ as well as $R$.
 In contrast,
-the additional node $C$ on the path to the MRCA indicates that $C1$ is
+the additional node $T$ on the path to the MRCA indicates that $T1$ is
 a broken \emph{incertae sedis} taxon.
 Thus we may consider an
-algorithm that, for each taxon $B$ on the synthesis tree, finds the
+algorithm that, for each taxon $S$ on the synthesis tree, finds the
 closest ancestral taxon $A$, and checks if $A$ is an ancestor of $B$
@@
 MRCA(A,B)$ and $B\to B_{1}\to\ldots\to MRCA(A,B)$.
+
+\begin{figure}
+\hfill{}\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-clade/tree1-shaded}
+
+}\hfill{}\subfloat[Input tree \#2]{\includegraphics[scale=0.5]{Figures/is-clade/tree3-shaded}
+
+}\hfill{}
+
+\hfill{}\subfloat[Taxonomy]{\includegraphics[scale=0.5]{Figures/is-clade/tax-shaded}
+
+}\hfill{}\subfloat[Synthesis Tree \#1]{\includegraphics[scale=0.5]{Figures/is-clade/synth2-shaded}
+
+}\hfill{}
+
+\caption{\label{fig:Broken-incertae-sedis} \emph{Incertae sedis} clade $T$
+broken by conflicting placement.
+Input trees (a) and (b) conflict
+in the placement of taxa in $T$.
+The synthesis supertree (d) places
+places $T1$ and $T2$ separately within $R$, while unplaced taxa
+float upwards.
+}
+
+\end{figure}
 
@@
 members of the clade to be placed separately (Figure \ref{fig:Broken-incertae-sedis}).
-\begin{figure}
-\hfill{}\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-clade/tree1-shaded}
-
-}\hfill{}\subfloat[Input tree \#2]{\includegraphics[scale=0.5]{Figures/is-clade/tree3-shaded}
-
-}\hfill{}
-
-\hfill{}\subfloat[Taxonomy]{\includegraphics[scale=0.5]{Figures/is-clade/tax-shaded}
-
-}\hfill{}\subfloat[Synthesis Tree \#1]{\includegraphics[scale=0.5]{Figures/is-clade/synth2-shaded}
-
-}\hfill{}
-
-\caption{\label{fig:Broken-incertae-sedis} \emph{Incertae sedis} clade $C$
-broken by conflicting placement.
-Input trees (a) and (b) conflict
-in the placement of taxa in $C$.
-The synthesis supertree (d) places
-places $C1$ and $C2$ separately within $A$, while unplaced taxa
-float upwards.
-(\textbf{Note:} This is what the sub-problem solver
-does.
-The entire pipeline would attach $C3$ at $A$ actually, since
-it is the MRCA of $A1$ and $A2$.
-Should we change this?) }
-
-\end{figure}
 
@@
 
-In Figure \ref{fig:Broken-incertae-sedis}, tree $T_{1}$ places $C1$
-next to $A1$, while $T_{2}$ places $C2$ next to $A2$.
+In Figure \ref{fig:Broken-incertae-sedis}, tree $T_{1}$ places $T1$
+next to $R1$, while $T_{2}$ places $T2$ next to $R2$.
 Since different
-members of $C$ are placed, neither placement for $C$ is rejected.
-Instead the taxon $C$ is broken, the name $C$ disappears, and $C3$
-floats to the top level.
-Furthermore, since taxon $C$ is a broken
-\emph{incertae sedis }taxon, all of its children are effectively
+members of $C$ are placed, neither placement for $T$ is rejected.
+Instead the taxon $T$ is broken, the name $T$ disappears, and $T3$
+floats to the MRCA of the placed members of $T$.
+Furthermore, since taxon $T$ is a broken
+\emph{incertae sedis} taxon, all of its children are effectively
 incertae sedis independently, with the exception that they cannot be
 placed within each other.
-Therefore, the names $A$ and $AB$ are not
+Therefore, the names $R$ and $RS$ are not
 lost, since the taxa placed within them are \emph{incertae sedis}
@@
 
-The situation here would be different if the monophyly of $C$ was
+The situation here would be different if the monophyly of $T$ was
 supported by an high-ranked input phylogeny.
 In that case, instead of
-breaking $C$, we would choose the placement in the highest-ranked
+breaking $T$, we would choose the placement in the highest-ranked
 tree.

### 2018-05-07 doc/incertae_sedis.tex
restored fig 1 content

--- 2018-05-07 doc/incertae_sedis.tex
@@
 \begin{figure}
-\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-clade/tree1-shaded}
-
-}\hfill{}\subfloat[Taxonomy]{\includegraphics[scale=0.5]{Figures/is-clade/tax-shaded}
+\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/fig1/tree1-shaded}
+
+}\hfill{}\subfloat[Taxonomy]{\includegraphics[scale=0.5]{Figures/fig1/tax-shaded}
 
@@
 
-\subfloat[Synthesis Tree (no \emph{incertae sedis})]{\includegraphics[scale=0.5]{Figures/is-clade/synth1-no-is-shaded}
-
-}\hfill{}\subfloat[Synthesis Tree (\emph{incertae sedis} aware)]{\includegraphics[scale=0.5]{Figures/is-clade/synth1-shaded}
+\subfloat[Synthesis Tree (no \emph{incertae sedis})]{\includegraphics[scale=0.5]{Figures/fig1/synth1-no-is-shaded}
+
+}\hfill{}\subfloat[Synthesis Tree (\emph{incertae sedis} aware)]{\includegraphics[scale=0.5]{Figures/fig1/synth1-shaded}
 

### 2018-05-07 doc/incertae_sedis.tex
up to listed cases

@@
     over the tree from root to tip.
-If $\mathcal{S}(x)$ is the set of siblings of $x$ which are not annotated as being 
-    \emph{incertae sedis} and $p(x)$ is the parent of $x$, then
+If $\mathcal{S}(x)$ is the set of siblings of $x$and $p(x)$ is the parent of $x$, then
+the naive approach of ignoring the \emph{incertae sedis}  annotation amount to setting
+    the exclude set as:
 \begin{align}
-    \mathcal{E}(x) = \mathcal{E}(p(x)) \cup \bigcup_{s\in \mathcal{S}(x)} \mathcal{E}(s)
+    \mathcal{E}(x) = \mathcal{E}(p(x)) \cup \left\{ \bigcup_{s\in \mathcal{S}(x)} \mathcal{I}(s) \right\}
+    \label{eq:exsetformonea-traditional}
 \end{align}
 with $\mathcal{E}(\rho) = \emptyset$ for the root of the taxonomic tree.
+
+If we treat the  \emph{incertae sedis} annotation as an indication that the 
+    \emph{incertae sedis} taxon $x$ can be place in any of its sister taxa, then
+    we let $\mathcal{Q}(x)$ denote the set of siblings of $x$ that are not
+    flagged as being \emph{incertae sedis} and we define the exclude set as:
+\begin{align}
+    \mathcal{E}(x) = \mathcal{E}(p(x)) \cup \left\{ \bigcup_{s\in \mathcal{Q}(x)} \mathcal{I}(s)\right\}
+    \label{eq:exsetformonea}
+\end{align}
+We will refer to this interpretation of the \emph{incertae sedis} annotation as ``unrestricted
+    intrusion'' semantics.
 
@@
 The synthesis supertree (d) places
-places $T1$ and $T2$ separately within $R$, while unplaced taxa
-float upwards.
+places $T1$ and $T2$ separately within $R$.
+The BUILD algorithm would place the unsampled taxon $T3$ at the
+root of the tree, while our unprune operation implemented to 
+handle taxonomy-only taxa would place $T3$ as a child of the MRCA
+of the other members of $T$, as pictured in (d).
 }
@@
 \emph{incertae sedis}, as well as the outcomes for the rooted-split semantics.
-We think that they constitute one of the main contributions of the paper.  In
-each case, the synthesis trees in this section are constructed using the
+%We think that they constitute one of the main contributions of the paper.
+In each case, the synthesis trees in this section are constructed using the
 BUILD-based algorithm described in section \ref{sec:background}, without
@@
 
-With the naive semantics, intrusion of an incertae sedis taxon into a sibling is
+With the naive semantics, intrusion of an \emph{incertae sedis} taxon into a sibling is
 interpreted as conflict with that sibling.  This results in the loss of the edge
@@
 
-We see from Fig. \ref{fig:Handling-incertae-sedis}d that use of the incertae sedis
+We see from Fig. \ref{fig:Handling-incertae-sedis}d that use of the unrestricted intrusion
 semantics results the taxonomy edge $AB$ grouping $A$ and $B$.  It also restores the
@@
 
-\subsection{Conflict with incertae sedis taxa}
+\subsection{Conflict with \emph{incertae sedis} taxa}
 

### 2018-05-08 doc/incertae_sedis.tex
citing some of the early MRP stuff, and introducing the matrix rep that might help elsewhere.

@@
 \usepackage{setspace}
-\linespread{1.5}
+\linespread{1.0}
+\usepackage{multirow}
+
 \usepackage[authoryear]{natbib}
@@
 labels for leaves that are not descendants of that node.
+The ``include set'' and ``exclude set'' of a ``rooted split'' used here are equivalent
+    to the ``inside set'' and ``outside set'' of a ``$n$-taxon statement''
+    \citep[\emph{sensu}][]{Wilkinson1994}.
+
 If no taxa are \emph{incertae sedis},
@@
     supertree.
+
+\paragraph{Equivalence rooted split definitions with character encodings}
+Representing a split in a tree by a character is the basis of one of the oldest
+    supertree methods: matrix representation with parsimony \citep[MRP, see][]{Baum1992,Ragan1992}.
+In the rooted form of such an approach, character state 1 is used for tips that are
+    descendants of a internal node in an input tree and 0 is used for other tips.
+The partial splits caused by \emph{incertae sedis} annotations naturally map to
+    an encoding of missing data (typically denoted with the `?' symbol in a matrix).
+Thus the taxonomy shown in Figure \ref{fig:Handling-incertae-sedis}b could be represented as the matrix shown in 
+    Figure \ref{fig:Handling-incertae-sedis-mrp}.
+In the naive treatment of ignoring the \emph{incertae sedis} annotation, the `?' characters
+   in the row for taxa $C1$, $C2$, and $C3$ would be replaced with 0's.
+
+\begin{figure}
+\begin{center}
+\begin{tabular}{|r|p{2em}|p{2em}|p{2em}|p{2em}|}
+\hline
+& \multicolumn{4}{|c|}{Character/clade label} \\
+leaves & C & AB & B & A \\
+\hline
+$A1$, $A2$ & 0 & 1 & 0 & 1 \\
+$B1$, $B2$ & 0 & 1 & 1 & 0 \\
+$C1$, $C2$, $C3$ & 1 & ? & ? & ?  \\
+$D$  & 0 & 0 & 0 & 0 \\
+\hline
+\end{tabular}
+\end{center}
+\caption{\label{fig:Handling-incertae-sedis-mrp} The matrix encoding 
+of an \emph{incertae sedis} interpretation of the taxonomy shown in  Figure \ref{fig:Handling-incertae-sedis}b.
+Only the unique character encodings are show, thus the leftmost column
+lists all of the tip taxa that would be assigned a particular character encoding.
+}
+
+\end{figure}
 
@@
 In such cases, we treat all applicable names as synonyms for the node.
-\mthcomment{Need to verify that the namer records this situation as synonyms!}
+\mthcomment{Need to verify that the namer records this situation as synonyms, as it should!}
 

### 2018-05-09 doc/incertae_sedis.tex
char encoding of intermed

@@
 Thus, we could say that the synthetic tree places
-$C$ within $A$. Furthermore, the synthetic tree additionally places
-$C$ within $AB$.
-
-A slightly more general case is shown in 
-Figure \ref{fig:Broken-incertae-sedis}.
+$C$ within $A$. 
+%Furthermore, the synthetic tree additionally places
+% $C$ within $AB$.
+
+In general, after naming the clades in the supertree, the
+    taxonomic placement of an \emph{incertae sedis} taxon
+    can be detected if the most tip-ward named ancestor of
+    the clade in the supertree was not an ancestor in the
+    taxonomic tree.
+Note that if an \emph{incertae sedis} taxon is broken
+    during the supertree algorithm, it can be ambiguous whether
+    or not all of the members should be placed in a new
+    parent taxon.
+See, for example, Figure \ref{fig:Broken-incertae-sedis}.
 Here taxon $T1$ is an descendant of $R$ in the
 synthesis tree, but not the taxonomy.
-Thus $T1$ is placed within $R$.
-We note that the MRCA of $R$ and $T1$ in
-the taxonomy is the root node, and the path to the MRCA from $R$ is
-$R\to RS\to root$, while the path to the MRCA from $T1$ is $T1\to T\to
-root$.
-Here the additional node $RS$ on the path to the MRCA from $R$
-indicates that $T1$ is placed within $RS$ as well as $R$.
-In contrast,
-the additional node $T$ on the path to the MRCA indicates that $T1$ is
-a broken \emph{incertae sedis} taxon.
-Thus we may consider an
-algorithm that, for each taxon $S$ on the synthesis tree, finds the
-closest ancestral taxon $A$, and checks if $A$ is an ancestor of $B$
-on the taxonomy.
-If not, we compute the paths $A\to A_{1}\to\ldots\to
-MRCA(A,B)$ and $B\to B_{1}\to\ldots\to MRCA(A,B)$.
+Thus $T1$ is placed within $R$; taxon
+ $T2$ is similarly placed with in $R$.
+$T3$ is not mentioned in either input tree, and the 
+    only tree that speaks to its placement is the taxonomic input.
+However, the taxon $T$ cannot be monophyletic based on the
+     two input trees.
+Thus the correct placement of $T3$ is uncertain.
+
+The ambiguity of where to place $T3$ can also be compared to the
+    behavior of maximum compatibility tree estimation and parsimony.
+The matrix representation context the taxonomic tree's
+    character representation of taxon $T$ would include $T1$, $T2$ and $T3$
+    all being assigned state 1 and all of the other leaves being assigned 
+    state 0.
+This character is not compatible with the solution tree. 
+The compatability approach would be to conclude that the character encoding 
+    that taxon was incorrect and should be ignored.
+Since, no other character speaks to the position of $T3$, it would fall to being
+    a child of the root of the tree.
+In a similar result, the BUILD algorithm would place $T3$ as a child of the root.
+
+In a parsimony view, while the character cannot be mapped onto the solution in
+    a homoplasy-free manner, one might hope that the character data that the
+    taxonomy was based on was not completely unreliable.
+Placing of $T3$ as either sister to $T1$ or $T2$ could explain the character
+    representation of taxon $T$ with only two acquisitions of the state 1.
+Other placements of $T3$ would require at least three 0 to 1 changes (or a reversion
+    to state 0)l
+Placing $T3$ at the MRCA 
+    of the other members of $T$ corresponds
+    to handling of ``rogue'' taxa in an Adams consensus method of the two
+    most parsimonious placements of $T3$ under an irreversible parsimony model.
+
+% We note that the MRCA of $R$ and $T1$ in
+% the taxonomy is the root node, and the path to the ch from $R$ is
+% $R\to RS\to root$, while the path to the MRCA from $T1$ is $T1\to T\to
+% root$.
+% Here the additional node $RS$ on the path to the MRCA from $R$
+% indicates that $T1$ is placed within $RS$ as well as $R$.
+% In contrast,
+% the additional node $T$ on the path to the MRCA indicates that $T1$ is
+% a broken \emph{incertae sedis} taxon.
+% Thus we may consider an
+% algorithm that, for each taxon $S$ on the synthesis tree, finds the
+% closest ancestral taxon $A$, and checks if $A$ is an ancestor of $B$
+% on the taxonomy.
+% If not, we compute the paths $A\to A_{1}\to\ldots\to
+% MRCA(A,B)$ and $B\to B_{1}\to\ldots\to MRCA(A,B)$.
 
@@
 
+\end{figure}
+
+\subsection{Ambiguity of imputed taxonomic characters}
+It is tempting to view the matrix representation of the taxonomic tree
+    as an encoding of the presence, absence, and uncertainty about
+    suites of synapomorphies which allow us to diagnose a taxon.
+Clearly by simply representing the taxonomic tree, we will not be
+    able to recover the identity or even the number of characters that
+    a taxonomist might have used when classifying the organisms in question.
+Nevertheless, one might hope that a matrix representation of a taxonomy (such
+    as the matrix in Figure \ref{fig:Handling-incertae-sedis-mrp} as a 
+    representation of Figure \ref{fig:Handling-incertae-sedis}b) has a unique, uncontroversial 
+    association with the taxonomic tree.
+
+The rules for associated leaves in $\mathcal{I}(n)$ with state 1, leaves
+     in $\mathcal{E}(n)$, and other leaves with the missing data symbol 
+     do allow to encode the taxon $n$ in a taxonomic tree unambiguously.
+However, these are not the only set of rules for generating a character matrix
+    representation that would cause the taxonomic tree allows the characters to
+    be mapped without homoplasy.
+For example consider the case of the taxonomy shown in Figure \ref{fig:Placing-one-incertae}a.
+Either of the two matrices shown in \ref{fig:Placing-one-incertae-MRP} could encode this
+    taxonomy along with its \emph{incertae sedis} interpretation.
+
+\begin{figure}
+%\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-within-is/tree1\lyxdot tre}}
+\subfloat[Taxonomy]{\includegraphics[scale=0.5]{Figures/is-within-is/tax-shaded}
+}
+\subfloat[Synthesis tree]{\includegraphics[scale=0.5]{Figures/is-within-is/synth1-shaded}
+}
+\subfloat[Possible intermediate]{\includegraphics[scale=0.5]{Figures/is-within-is/intermed-shaded}
+}
+
+\caption{\label{fig:Placing-one-incertae}
+One \emph{incertae sedis} group ($W$) being placed within another ($V$), with possible
+loss of the taxon $V$.}
+\end{figure}
+
+\begin{figure}
+%\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-within-is/tree1\lyxdot tre}}
+\subfloat[maximally scored]{
+    \begin{tabular}{|r|p{2em}|p{2em}|}
+\hline
+& \multicolumn{2}{|c|}{Character} \\
+taxa & $V$ & $WX$  \\
+\hline
+$U$  & 0 & 0  \\
+$V1$ & 1 & ?  \\
+$V2$ & 1 & ?  \\
+$W$  & 0 & 1  \\
+$X$  & 0 & 1 \\
+\hline
+\end{tabular}
+}
+\subfloat[less informative]{
+    \begin{tabular}{|r|p{2em}|p{2em}|}
+\hline
+& \multicolumn{2}{|c|}{Character} \\
+taxa & $V$ & $WX$  \\
+\hline
+$U$  & 0 & 0 \\
+$V1$ & 1 & ? \\
+$V2$ & 1 & ? \\
+$W$  & ? & 1 \\
+$X$  & 0 & 1 \\
+\hline
+\end{tabular}
+}
+\caption{\label{fig:Placing-one-incertae-MRP}
+Two possible encodings that would support the $V$ and $WX$ branches and \emph{incertae sedis}
+annotation shown in the taxonomy from in Figure \ref{fig:Placing-one-incertae}a.
+The representations only vary in the scoring of leaf $W$ for character $V$.}
 \end{figure}
@@
 
-\begin{figure}
-\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-nested/tree3-shaded}
-
-}\hfill{}\subfloat[Taxonomy tree]{\includegraphics[scale=0.5]{Figures/is-nested/tax-shaded}
-
-}
-
-\subfloat[Synthesis tree (\emph{incertae sedis} ignored)]{\includegraphics[scale=0.5]{Figures/is-nested/synth3-no-is-shaded}
-
-}\hfill{}\subfloat[Synthesis tree (\emph{incertae sedis} considered)]{\includegraphics[scale=0.5]{Figures/is-nested/synth3-is-shaded}
-
-}
-
-\caption{\label{fig:Nested-incertae-sedis}Nested incertae sedis taxa.
-(a)
-Input tree \#1 places CI1 within CG, and C within $A$.
-(b) In the
-taxonomy, CI is \emph{incertae sedis }within C, while C is \emph{incertae
-sedis} under the root.
-(c) The synthesis tree when \emph{incertae
-sedis} taxa are not considered.
-The names AB, A, and CG are lost.
-(d) The synthesis tree when \emph{incertae sedis} taxa are considered.
-The names AB, A, and CG are regained, as well as the edges leading
-to AB and CG.}
-
-\end{figure}
+% \begin{figure}
+% \subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-nested/tree3-shaded}
+
+% }\hfill{}\subfloat[Taxonomy tree]{\includegraphics[scale=0.5]{Figures/is-nested/tax-shaded}
+
+% }
+
+% \subfloat[Synthesis tree (\emph{incertae sedis} ignored)]{\includegraphics[scale=0.5]{Figures/is-nested/synth3-no-is-shaded}
+
+% }\hfill{}\subfloat[Synthesis tree (\emph{incertae sedis} considered)]{\includegraphics[scale=0.5]{Figures/is-nested/synth3-is-shaded}
+
+% }
+
+% \caption{\label{fig:Nested-incertae-sedis}Nested incertae sedis taxa.
+% (a)
+% Input tree \#1 places CI1 within CG, and C within $A$.
+% (b) In the
+% taxonomy, CI is \emph{incertae sedis }within C, while C is \emph{incertae
+% sedis} under the root.
+% (c) The synthesis tree when \emph{incertae
+% sedis} taxa are not considered.
+% The names AB, A, and CG are lost.
+% (d) The synthesis tree when \emph{incertae sedis} taxa are considered.
+% The names AB, A, and CG are regained, as well as the edges leading
+% to AB and CG.}
+% \end{figure}
+
 When \emph{incertae sedis} clades are allowed, it is possible for an
@@
 $A$ (Figure \#).
-
-\begin{figure}
-\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-within-is/tree1\lyxdot tre}
-
-}\subfloat[Taxonomy]{\includegraphics[scale=0.5]{Figures/is-within-is/tax-shaded}
-
-}\subfloat[Synthesis tree]{\includegraphics[scale=0.5]{Figures/is-within-is/synth1-shaded}
-
-}
-
-\caption{\label{fig:Placing-one-incertae}Placing one incertae sedis group
-within another.
-Here I is placed within A, and then AI is placed within
-A.}
-\end{figure}
 

### 2018-05-09 doc/incertae_sedis.tex
order of operations point

@@
     taxonomy along with its \emph{incertae sedis} interpretation.
+The matrix in Figure \ref{fig:Placing-one-incertae-MRP}a would result in the 
+    taxon $V$ not being named in the supertree shown in Figure \ref{fig:Placing-one-incertae}b,
+    while the matrix shown in Figure \ref{fig:Placing-one-incertae-MRP}b would lead
+    to $V$ being named.
+Thus, accurately
+    honors the input taxonomic content we would need more taxonomic information
+    than simply a hierarchy and knowledge of which taxa can float more tipward.
+
+Note that this example also presents a challenge for iterative supertree building and refinement
+    of the taxonomy without retention of additional information.
+Consider the case, in which there are two phylogenetic inputs along with the taxonomy from
+    tree Figure \ref{fig:Placing-one-incertae}b with the first phylogeny, $T_1$, providing the single
+    rooted split: $V1\,X \mid \bullet\,U$ and the second phylogeny, $T_2$ providing the 
+    rooted split: $W\, V1 \mid \bullet\,V2$.
+If the supertree pipeline were run with just the taxonomy and and $T_1$ as inputs, then the supertree
+    would be the tree shown in Figure \ref{fig:Placing-one-incertae}c.
+One might hope that using that tree as the revised taxonomy and then adding the split from $T_2$ in
+    a subsequent analysis would result in the same tree as if $T_1$ and $T_2$ to the original taxonomy.
+While the topology for the final tree in either order of analyses would indeed be the tree shown in 
+    Figure \ref{fig:Placing-one-incertae}b}, the decision about whether or not to name $V$ would differ.
+Our handling of \emph{incertae sedis} in the original taxonomy via rooted partial splits
+    would encode the taxonomic definitions in the manner depicted in Figure \ref{fig:Placing-one-incertae-MRP}a,
+    but the encoding of the revised taxonomy in Figure \ref{fig:Placing-one-incertae}c would result
+    in an altered definition of taxon $V$, as shown in \ref{fig:Placing-one-incertae-MRP}c.
+As expected, in the encoding of the revised taxonomy, the definition of $WX$ has been altered to reflect the
+    fact $T_1$ placed $V$ inside $WX$.
+Perhaps, unexpected is the fact that storing the revised taxonomy with an \emph{incertae sedis} annotation
+    and then reprocessing the taxonomy using our rooted partial splits approach would also shift
+    the definition of $V$ from a taxon that must exclude $W$ to one that may include $W$.
 
@@
 }
-\subfloat[Synthesis tree]{\includegraphics[scale=0.5]{Figures/is-within-is/synth1-shaded}
+\subfloat[Final supertree]{\includegraphics[scale=0.5]{Figures/is-within-is/synth1-shaded}
 }
@@
 }
+\subfloat[from revised taxonomy]{
+    \begin{tabular}{|r|p{2em}|p{2em}|}
+\hline
+& \multicolumn{2}{|c|}{Character} \\
+taxa & $V$ & $WX$  \\
+\hline
+$U$  & 0 & 0 \\
+$V1$ & 1 & 1 \\
+$V2$ & 1 & 1 \\
+$W$  & ? & 1 \\
+$X$  & 0 & 1 \\
+\hline
+\end{tabular}
+}
 \caption{\label{fig:Placing-one-incertae-MRP}
-Two possible encodings that would support the $V$ and $WX$ branches and \emph{incertae sedis}
+Panels (a) and (b) show two possible encodings that would support the $V$ and $WX$ branches and \emph{incertae sedis}
 annotation shown in the taxonomy from in Figure \ref{fig:Placing-one-incertae}a.
-The representations only vary in the scoring of leaf $W$ for character $V$.}
+The representations only vary in the scoring of leaf $W$ for character $V$. Panel (c) shows the information encoded by the tree in \ref{fig:Placing-one-incertae}c. }
 \end{figure}

### 2018-05-09 doc/incertae_sedis.tex
blah

@@
 
-However, this assumes that \emph{incertae sedis} taxa are always
-placed within their siblings, or descendants of their siblings.
-As
-such, it does not handle cases such as Figure \ref{fig:Placing-one-incertae}, 
-in which $I$ is first placed as sibling to $AI$, and then
-$AI$ is placed within $I$.
-Note that $I$ is not a sibling of $AI$ or a
-descendant of a sibling.
-Furthermore, the MRCA of $I$ and $AI$ is the
-root, and the path from $AI$ to the root includes $A$, but $A$ is not
-broken.
-We suggest that this conundrum may be resolved by first
-processing the placement of $I$ within $A$ by modifying the taxonomy
-to reflect this placement\footnote{Does this actually work? Some
-placements break taxa.}.
-When we consider the placement of $AI$ within
-$I$ on this modified taxonomy, the MRCA of $I$ and $AI$ is now $A$
-instead of the root, and the correct interpretation results.
-We
-suggest that by walking the tree in a pre-order fashion and processing
-shallowing placements first, the previous approach of finding the MRCA
-and labelling intermediate nodes as either additional placements or
-broken incertae sedis (or incertae\_sedis\_inherited) taxa holds.
-Such
-an approach should also handle cases containing nested incertae sedis
-taxa, as in Figure \ref{fig:Placing-one-incertae}.
+% However, this assumes that \emph{incertae sedis} taxa are always
+% placed within their siblings, or descendants of their siblings.
+% As
+% such, it does not handle cases such as Figure \ref{fig:Placing-one-incertae}, 
+% in which $I$ is first placed as sibling to $AI$, and then
+% $AI$ is placed within $I$.
+% Note that $I$ is not a sibling of $AI$ or a
+% descendant of a sibling.
+% Furthermore, the MRCA of $I$ and $AI$ is the
+% root, and the path from $AI$ to the root includes $A$, but $A$ is not
+% broken.
+% We suggest that this conundrum may be resolved by first
+% processing the placement of $I$ within $A$ by modifying the taxonomy
+% to reflect this placement\footnote{Does this actually work? Some
+% placements break taxa.}.
+% When we consider the placement of $AI$ within
+% $I$ on this modified taxonomy, the MRCA of $I$ and $AI$ is now $A$
+% instead of the root, and the correct interpretation results.
+% We
+% suggest that by walking the tree in a pre-order fashion and processing
+% shallowing placements first, the previous approach of finding the MRCA
+% and labelling intermediate nodes as either additional placements or
+% broken incertae sedis (or incertae\_sedis\_inherited) taxa holds.
+% Such
+% an approach should also handle cases containing nested incertae sedis
+% taxa, as in Figure \ref{fig:Placing-one-incertae}.
 

### 2018-05-09 doc/incertae_sedis.tex
shorter...

--- 2018-05-09 doc/incertae_sedis.tex
@@
 
-However, this assumes that \emph{incertae sedis} taxa are always
-placed within their siblings, or descendants of their siblings.
-As
-such, it does not handle cases such as Figure \ref{fig:Placing-one-incertae}, 
-in which $I$ is first placed as sibling to $AI$, and then
-$AI$ is placed within $I$.
-Note that $I$ is not a sibling of $AI$ or a
-descendant of a sibling.
-Furthermore, the MRCA of $I$ and $AI$ is the
-root, and the path from $AI$ to the root includes $A$, but $A$ is not
-broken.
-We suggest that this conundrum may be resolved by first
-processing the placement of $I$ within $A$ by modifying the taxonomy
-to reflect this placement\footnote{Does this actually work? Some
-placements break taxa.}.
-When we consider the placement of $AI$ within
-$I$ on this modified taxonomy, the MRCA of $I$ and $AI$ is now $A$
-instead of the root, and the correct interpretation results.
-We
-suggest that by walking the tree in a pre-order fashion and processing
-shallowing placements first, the previous approach of finding the MRCA
-and labelling intermediate nodes as either additional placements or
-broken incertae sedis (or incertae\_sedis\_inherited) taxa holds.
-Such
-an approach should also handle cases containing nested incertae sedis
-taxa, as in Figure \ref{fig:Placing-one-incertae}.
-
-\begin{comment}
-One question is whether this approach is able to handle nested
-incertae sedis taxa.
-
-Placement of incertae sedis taxa by input trees is unfortunately not
-quite as simple as finding a single location where an I.S.
-taxon
-should attach.
-For example, when an incertae sedis taxon is broken,
-its children need to be ``placed'' separately.
-
-Each input tree can relate to an \emph{incertae sedis} taxon
-$(A,B,C)D$ in a number of ways
-\begin{itemize}
-    \item it could resolve
-$A$, $B$, $C$, or $D$.
-    \item it could place $D$ on a degree-2 (=out-
-degree-1) node that bisects a branch 
-    \item it could place a descendant
-taxon of $D$ 
-    \item it could place a descendant taxon of $D$ in a
-\emph{different place} than another input tree.
-    \item it could place
-children of $D$ in multiple places, thus conflicting with the branch.
-If the \emph{incertae sedis} taxon $(A,B,C)$ is broken, then $A$, $B$
-and $C$ become \emph{incertae sedis} clades in their own right, that
-may attach separately, except that they .
-This is because none of $A$,
-$B$, or $C$ is in the exclude set of the siblings of $D$.
-\end{itemize}
-\end{comment}
-
-\section{The rooted-split semantics in practice}
-
-In order to examine the results of the rooted splits-semantics above, we explore
-the outcome on a series of complex cases.  These cases are intended to
-illustrate some of the issues that arise when developing an operational semantics of
-\emph{incertae sedis}, as well as the outcomes for the rooted-split semantics.
-%We think that they constitute one of the main contributions of the paper.
-In each case, the synthesis trees in this section are constructed using the
-BUILD-based algorithm described in section \ref{sec:background}, without
-decomposing them into smaller subproblems or pruning and then unpruning
-taxonomy-only taxa.  Synthesis trees that are described as \emph{incertae
-sedis}-aware apply the BUILD algorithm to splits constructed using equation
-\ref{eq:exsetformonea}, whereas the synthesis trees that are not \emph{incertae
-sedis}-aware are constructed from splits using equation
-\ref{eq:exsetformonea-traditional}.
-
-Much of the complexity that we explore here comes from the fact that it is not
-only tip taxa that can be \emph{incertae sedis}.  It is also possible to have
-higher ranked  \emph{incertae sedis} taxa, such as genera that are marked
-\emph{incertae familia}.
-
-\subsection{The \emph{incertae sedis} semantics restores taxonomic groups and names}
-
-With the naive semantics, intrusion of an \emph{incertae sedis} taxon into a sibling is
-interpreted as conflict with that sibling.  This results in the loss of the edge
-corresponding to the sibling taxon as well as the taxon name.
-
-We see from Fig. \ref{fig:Handling-incertae-sedis}d that use of the unrestricted intrusion
-semantics results the taxonomy edge $AB$ grouping $A$ and $B$.  It also restores the
-names $AB$ and $A$.  We may thus determine that $C$ has been placed within $A$ (and
-its ancestor $AB$).
-
-\subsection{Conflict with \emph{incertae sedis} taxa}
-
-When only tip nodes on the taxonomy are \emph{incertae sedis}, we have
-a very simple form of the \emph{incertae sedis} supertree problem.
-However, in practice entire clades may be \emph{incertae sedis}, and
-so more complex issue arise.
-In this section we consider a number of
-cases that must be handled when attempting to place \emph{incertae
-sedis} taxa.
-
-\emph{Note that the synthesis of all input trees before the taxonomy
-is unaffected by incertae sedis information.}
-
-\subsection{Conflicting placement among input trees}
-
-{[}\textbf{BDR:} \emph{maybe we should merge this with Figure.
-\ref{fig:Broken-incertae-sedis}.
-Although the point of figure
-\ref{fig:placement-example1} is to talk about sub-problem
-decomposition as well as what the result looks like.}{]}
-
-The addition of \emph{incertae sedis} taxa allows new types of
-conflict between input trees.
-For example, different input trees might
-place an incertae sedis taxon in conflicting locations.
-This is
-illustrated in Figure \ref{fig:placement-example1}, where the IS
-taxon (ott7,ott6)ott10 is placed as sister to ott1 by phylogeny
-$\tau_{1}$ and as sister ott3 by phylogeny $\tau_{2}$.
-
-When this happens, the placement of the IS taxon is not influenced by
-its being marked IS on the taxonomy.
-Thus, in Example 1, the higher
-ranked tree $\tau_{1}$ will be reflected in the synthesis tree, ott10
-will be placed as sister to ott1.
-In contrast, the conflicting
-placement in $\tau_{2}$ will not be reflected in the synth tree.
-
-All this would occur in the previous version of propinquity.
-Where the
-updated version differs that (a) the names ott9 and ott8 are retained
-instead of being dropped.
-(b) as a result of not breaking ott8 and
-ott9, we do not move ott3 and ott2 up to ott11.
-{[}BDR: Extend more figures!{]}
-
-\subsection{Case 1: An \emph{incertae sedis} clade}
-
-When a clade is marked as \emph{incertae sedis}, we need to consider
-two cases.
-In the first case, the clade may be placed within a sister
-clade intact (Figure \ref{fig:Handling-incertae-sedis}).
-However, if
-the clade is not monophyletic in the synthesis tree, then we allow the
-members of the clade to be placed separately (Figure \ref{fig:Broken-incertae-sedis}).
-
-When the input phylogenies conflict on where members of an incertae
-sedis clade should be placed, neither placement is rejected.
-Instead,
-the taxon $C$ is broken, the name $C$ disappears, and $C3$ floats to
-the top level.
-Algorithmic, alternatively placements of children of an
-incertae sedis clade lead to contesting of all edges along with taxa
-might be placed, creating a large, merged sub-problem, as in Figure
-\ref{fig:placement-example1}.
-
-In Figure \ref{fig:Broken-incertae-sedis}, tree $T_{1}$ places $T1$
-next to $R1$, while $T_{2}$ places $T2$ next to $R2$.
-Since different
-members of $C$ are placed, neither placement for $T$ is rejected.
-Instead the taxon $T$ is broken, the name $T$ disappears, and $T3$
-floats to the MRCA of the placed members of $T$.
-Furthermore, since taxon $T$ is a broken
-\emph{incertae sedis} taxon, all of its children are effectively
-incertae sedis independently, with the exception that they cannot be
-placed within each other.
-Therefore, the names $R$ and $RS$ are not
-lost, since the taxa placed within them are \emph{incertae sedis}
-names.
-
-The situation here would be different if the monophyly of $T$ was
-supported by an high-ranked input phylogeny.
-In that case, instead of
-breaking $T$, we would choose the placement in the highest-ranked
-tree.
-
-\subsection{Case 2: A nested \emph{incertae sedis} clade}
-
-% \begin{figure}
-% \subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-nested/tree3-shaded}
-
-% }\hfill{}\subfloat[Taxonomy tree]{\includegraphics[scale=0.5]{Figures/is-nested/tax-shaded}
-
-% }
-
-% \subfloat[Synthesis tree (\emph{incertae sedis} ignored)]{\includegraphics[scale=0.5]{Figures/is-nested/synth3-no-is-shaded}
-
-% }\hfill{}\subfloat[Synthesis tree (\emph{incertae sedis} considered)]{\includegraphics[scale=0.5]{Figures/is-nested/synth3-is-shaded}
-
-% }
-
-% \caption{\label{fig:Nested-incertae-sedis}Nested incertae sedis taxa.
-% (a)
-% Input tree \#1 places CI1 within CG, and C within $A$.
-% (b) In the
-% taxonomy, CI is \emph{incertae sedis }within C, while C is \emph{incertae
-% sedis} under the root.
-% (c) The synthesis tree when \emph{incertae
-% sedis} taxa are not considered.
-% The names AB, A, and CG are lost.
-% (d) The synthesis tree when \emph{incertae sedis} taxa are considered.
-% The names AB, A, and CG are regained, as well as the edges leading
-% to AB and CG.}
-% \end{figure}
-
-When \emph{incertae sedis} clades are allowed, it is possible for an
-\emph{incertae sedis} taxon to be nested within another \emph{incertae
-sedis} taxon.
-In such a case we must allow the higher-ranked taxon to
-be placed within a sibling, while preserving the right of the lower
-ranked taxon to be placed within a lower-level sibling (Figure
-\ref{fig:Nested-incertae-sedis}).
-
-\subsection{Case 3: Placement of one \emph{incertae sedis} taxon within
-another.}
-\subsection{Surprising results of the split-based semantics}
-
-\subsubsection{Inter-digitation}
-1. Inter-digitation
-
-\bdrcomment{Ben: discuss figure \ref{fig:interdigitation}.}
-
-\subsubsection{Placing one \emph{incertae sedis} group within another}
-It would be nice if our semantics of incertae sedis taxa would allow us to place
-one \emph{incertae sedis} sibling inside another.
-A special case of
-this is when a higher-rank incertae sedis taxon $A$ is placed as
-sister to an incertae sedis taxon $B$, and then $B$ is placed within
-$A$ (Figure \#).
-
-2. Placing A next to B, then B within A.
-
-\bdrcomment{Ben: Write discussion of Figure \ref{fig:Placing-one-incertae}}
-
-\subsection{Splits-based semantics versus edit-based semantics}
-
-Here we focus on a split-based semantics instead of a semantics based on applying
-of a series of edit operations to the tree. The split-based semantics allows us
-to seek trees that are simultaneously consistent with a set of split-constraints.
-The use of a split-based semantics does impose some limitations. These come
-primarily because in the edit-based semantics, after we place an
-\emph{incertae sedis} taxon, it is no loner \emph{incertae sedis}, whereas with
-the split-based semantics taxa are either always \emph{incertae sedis} or never
-\emph{incertae sedis}.
-
-This has an effect, for example, when we consider two incertae sedis siblings $A$
-and $B$. If we can place $A$ within $B$, and $B$ within $A$, then a split- based
-semantics allows us to freely inter-digitate $A$ and $B$. If we do not want to
-allow $A$ and $B$ to be inter-digitated, then we cannot both $A$ within $B$ and $B$
-within $A$. In addition, we can imagine that a family $A$ is incertae sedis within a
-kingdom. In an edit-based semantics, we could place $A$ as a sibling to a genus $B$,
-mark $A$ as non-\emph{incertae-sedis}, and then place $B$ within $A$. In a split-based
-semantics, we must disallow either $B$ within $A$ or $A$ within $B$ if we want to disallow
-interdigitation.
-
-As a result of this, we cannot satisfy all five of these criteria. Thus, we
-describe two alternative split-based semantics. The non-interdigitation
-semantics gets \# (1,2,4,5), and the interdigitation semantics gets (1,3,4,5).
-This makes the point that each semantics results from different choices about
-which criteria to satisfy.
+\begin{figure}
+\subfloat[Taxonomy $\protect\taxonomy$]{\includegraphics[scale=0.5]{Figures/placement1/tax\lyxdot tre}
+
+}\subfloat[$\tau_{1}$]{\includegraphics[scale=0.5]{Figures/placement1/input1\lyxdot tre}
+
+}\subfloat[$\tau_{2}$]{\includegraphics[scale=0.5]{Figures/placement1/input2\lyxdot tre}
+
+}
+
+\caption{\label{fig:placement-example1}Example.
+    An incertae sedis clade (ott6,ott7)
+    is placed in different subtrees by input trees $\tau_{1}$ and $\tau_{2}$.
+    In $\tau_{1}$, two nodes that correspond to the taxonomy their ingroup
+    extended to include (ott6,ott7), and the branches leading to these
+    nodes have been colored blue.
+    The dashed blue edge leads to a node
+    that is a newly-introduced degree-2 node which does not correspond
+    to any taxonomy node.
+    In $\tau_{2}$, only one node that corresponds
+    to a taxonomy node needs to have its ingroup extended.
+    The placement
+    of (ott6,ott7) into ott8 toward ott1 by $\tau_{1}$ conflicts with
+    the placement of (ott6,ott7) into ott9 toward ott3 by $\tau_{2},$}
+\end{figure}
 
@@
 
-In order to handle \emph{incertae sedis} taxa within propinquity, we
+In order to handle \emph{incertae sedis} taxa within propinquity \citep{redelings2017supertree}, we
 must modify some of the stages of the propinquity pipeline.
@@
 subproblem.
-Subproblem files must indicate which taxa are incertae
-sedis.
+Subproblem files must indicate which taxa are \emph{incertae sedis}.
 The subproblem solver must read this information, account for
@@
 taxa so that it does not consider taxa broken when they have an
-incertae sedis taxon placed inside them.
-
-\subsection{Exemplifying taxa}
-
-One current problem is that well-known taxa like Fungi or Mammalia
-tend to have a very large number of incertae sedis children, making
-browsing in the tree viewer difficult.
-This can happen when, for
-example, fossils or other hard-to- place taxa get classified only to
-the level of these well-known nodes and no further.
-This leads to a
-situation where well-known taxa serve as a dumping ground for unplaced
-taxa.
-
-Our current approach to this problem is to perform a second round of
-pruning, or ``cleaning'', during the exemplification step.
-Incertae
-sedis taxa are pruned at this stage if they do not occur in any input
-trees.
-We thus generate a second ``cleaned taxonomy'' that has
-undergone this further round of cleaning.
-This approach improves on
-the previous approach in that \emph{incertae sedis} taxa in input
-trees are no longer pruned.
-This approach also removes tons of
-\emph{incertae sedis} children from nodes like ``Fungi'', where a lot
-of unplaced fossils with few observable characters have been dumped.
-
-However, this approach has the negative effect of pruning some
-incertae sedis taxa that need not be pruned.
-For example, suppose
-\emph{incertae sedis} taxon $A$ contains 5 children, of which only 1
-child $A_{1}$ occurs in an input tree.
-If the taxon $A$ is not broken,
-then it should be possible to attach the other 4 members of $A$ next
-to $A_{1}$, without cluttering up the synthesis tree.
-Such taxa have
-been successfully placed even though they are not in any input tree.
-This can only be discovered after synthesis is complete, though.
-
-We should be able to to figure out which incertae sedis taxa are
-placed (see section \ref{subsec:Placement}), and then mark the other
-ones incertae sedis in the \emph{synthesis tree}.
-We could then filter
-out the taxa that \emph{remain }incertae sedis (or incertae\_sedis
-inherited) in the tree viewer instead of the synthesis pipeline.
+\emph{incertae sedis} taxon placed inside them.
 
@@
 
-\begin{figure}
-\subfloat[Taxonomy $\protect\taxonomy$]{\includegraphics[scale=0.5]{Figures/placement1/tax\lyxdot tre}
-
-}\subfloat[$\tau_{1}$]{\includegraphics[scale=0.5]{Figures/placement1/input1\lyxdot tre}
-
-}\subfloat[$\tau_{2}$]{\includegraphics[scale=0.5]{Figures/placement1/input2\lyxdot tre}
-
-}
-
-\caption{\label{fig:placement-example1}Example.
-An incertae sedis clade (ott6,ott7)
-is placed in different subtrees by input trees $\tau_{1}$ and $\tau_{2}$.
-In $\tau_{1}$, two nodes that correspond to the taxonomy their ingroup
-extended to include (ott6,ott7), and the branches leading to these
-nodes have been colored blue.
-The dashed blue edge leads to a node
-that is a newly-introduced degree-2 node which does not correspond
-to any taxonomy node.
-In $\tau_{2}$, only one node that corresponds
-to a taxonomy node needs to have its ingroup extended.
-The placement
-of (ott6,ott7) into ott8 toward ott1 by $\tau_{1}$ conflicts with
-the placement of (ott6,ott7) into ott9 toward ott3 by $\tau_{2},$}
-\end{figure}
  For example, in Figure

### 2018-05-09 doc/incertae_sedis.tex
shortened for the time being

@@
 \begin{figure}
-\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/fig1/tree1-shaded}
-
-}\hfill{}\subfloat[Taxonomy]{\includegraphics[scale=0.5]{Figures/fig1/tax-shaded}
-
-}
-
-\subfloat[Synthesis Tree (no \emph{incertae sedis})]{\includegraphics[scale=0.5]{Figures/fig1/synth1-no-is-shaded}
-
-}\hfill{}\subfloat[Synthesis Tree (\emph{incertae sedis} aware)]{\includegraphics[scale=0.5]{Figures/fig1/synth1-shaded}
-
-}
+\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/fig1/tree1-shaded}}
+\hfill{}\subfloat[Taxonomy]{\includegraphics[scale=0.5]{Figures/fig1/tax-shaded}}
+
+\subfloat[Synthesis Tree (no \emph{incertae sedis})]{\includegraphics[scale=0.5]{Figures/fig1/synth1-no-is-shaded}}
+\hfill{}\subfloat[Synthesis Tree (\emph{incertae sedis} aware)]{\includegraphics[scale=0.5]{Figures/fig1/synth1-shaded}}
 
@@
 that groups $A$ and $B$. Taxon names and edges that are conditional on handling
-\emph{incertae sedis}
-are shaded grey.
+\emph{incertae sedis} are shaded grey. \mthcomment{We should probably remove $C3$ and replace $B1$ and $B2$ with $B$
+to simplify this. We could use triangle to indicate that $D$ and $B$ could be leaves or clades.}
 }
@@
 
-\subsubsection{Caveats}
-
-When multiple I.S.
-taxa have been moved to the root node of a
-subproblem, they may be I.S.
-over the entire subproblem, and some may
-be I.S.
-over others in an asymmetric manner.
-Therefore, we might need
-to specify additional information about the original attachment
-location of the I.S.
-taxa, such as their depth.
-This only affects
-problems that have been decomposed.
-
-\textbf{BDR:} \emph{currently we don't actually move taxa to get them
-into a subproblem.
-So, is this even an issue?}
-
-\subsection{Grafted supertree}
-
-\emph{Question:} Does the synthesis tree contain any \emph{incertae
-sedis} groups?\\ \emph{Answer:} The grafted supertree will not contain
-any \emph{incertae sedis} groups.
-However, when we attach pruned nodes
-to a parent in the grafted supertree, we could mark such nodes
-\emph{incertae sedis} if we want.
-
-\subsection{Unpruning}
-
-Currently the unpruner \emph{does not} require that the OTT ids are
-named in the grafted solution before unpruning starts.
-According to
-Mark's document, he wasn't sure if such names were generated for nodes
-that had an IS taxon placed inside of them, so otc-unprune-solution-
-and-name-unnamed nodes throws away all the names and generates them
-itself.
-
-\textbf{BDR}: See document texttt{otcetera/doc/unprune-solution-and-name-unnamed-nodes.pdf}
-
-The unpruner should record when unpruned nodes are \emph{incertae
-sedis}.
-Such nodes are unaffected by phylogenies, and so
-\emph{incertae sedis} annotations for them make good sense.
-
 \subsection{Annotation}
@@
 to the conflict-analysis code in the annotation step.
-\begin{comment}
-
-\subsection{Conflict service}
-
-The current conflict service considers a group $A$ to conflict with
-the taxonomy if group $A$ has an incertae sedis group $B$ placed
-within it.
-This doesn't affect the annotations, since taxon names are
-added by the unpruner.
-But it could make perfectly fine input trees
-incorrectly look like they are the cause of broken taxa, if they
-contain IS taxa.
-Thus, it would be nice to have a modified conflict
-algorithm.
-
-\subsubsection{Current conflict algorithm}
-
-The current conflict algorithm is pretty fast, but it works by
-classifying tips into either (i) the include group or (ii) the exclude
-group.
-To avoid counting the exclude group for every split, we instead
-count the total number of children for each node, and assume that any
-children not in the include set are in the exclude set.
-This is no
-longer true when we have incertae sedis taxa.
-I suspect that if we
-want to handle incertae sedies, we'd need a third category (iii) for
-``neither include group nor exclude group''.
-\begin{enumerate}
- \item
-Get induced trees on intersection of leaf sets \item Compute depth for
-each node (nd->depth)
- \item Compute number of tips at or below each
-node (nd->n\_tips)
- \item for each input tree node -> $nd$
- \begin{enumerate}
-   \item skip the root \item skip monotypic \item if
-its a tip then find corresponding (``terminal'') edges in synth tree
-and continue \item leaves1 <- get the list of leaves in the include
-group of $nd$ (in input) \item L2 <- find the total number of tips (L2
-= sum {[}nd->n\_tips| nd <- leaves1{]})
-   \item leaves2 <- get list of
-corresponding synth leaf nodes (in synth)
-   \item nodes <- find all
-nodes between leaves2 and the MRCA (in synth)
-   \item MRCA <- mrca of
-leaves2 (in synth.
-this uses the nd->depth annotation)
-   \item Compute
-number of tips in the include set (nd->include\_tips) below each node
-in $nodes$ (in synth)
-   \item if n\_include\_tips(MRCA) == n\_tips(MRCA)
-then the MRCA displays $nd$
-   \item if n\_include\_tips(MRCA) <
-n\_tips(MRCA) then 
-   \begin{itemize} 
-     \item foreach node in nodes
-      \begin{itemize}
-         \item if (n\_include\_tips(nd) < n\_tips(nd) and
-n\_include\_tips(nd) < l2)
-          \begin{itemize}
-            \item this is a conflict!
-\end{itemize} 
-           \end{itemize}
-         \item if there are no conflicts, then this
-is a resolved_by.
-
-      \end{itemize}
-\end{enumerate}
-\end{enumerate}
-
-\subsubsection{Modified conflict algorithm?}
-
-This probably is outside the scope of the paper, but if we could come
-up with a modified conflict algorithm, that would be nice/useful, and
-probably novel.
-It would also probably be slower...
-
-\end{comment}
-
-\section{Results}
-
-Should we do this? We could say:  \begin{itemize} \item there are 331
-incertae sedis taxa that are mentioned in input trees.
-\begin{itemize}
-\item what is the kingom / phylum / class / order / genus / species
-\end{itemize} \item we placed 14 incertae sedis taxa inside a sister
-taxon.
-(``incertae sedis'', ``versus ``unplaced''?) \item we placed
-$o$ incertae sedis taxa \emph{outside} all sister taxa.
-\item we
-confirmed $n$ incertae sedis taxa as being separate from all sister
-(sampled) taxa?? \item we avoiding breaking 8 taxa that had IS taxa
-placed inside them.
-\item we allowed $z_{1}$ new taxa into the
-synthesis tree that were incertae sedis.
-\item we allowed $z_{2}$ new
-taxa into the synthesis tree that are marked as extinct.
-\item some
-nodes have as many as $w$\emph{ incertae sedis} children, making them
-unbrowseable when incertae sedis children are not excluded.
-\item
-$v_{1}$ input trees were previously excluded \emph{entirely} because
-they are nested within in an incertae sedis taxon.
-\item $v_{2}$ input
-trees were previously excluded \emph{partially} because they are
-nested within in an incertae sedis taxon.
-\end{itemize} Currently the
-numbers $z_{1}$ and $z_{2}$.
-
-\section{Discussion}
-
-One thing we could do (perhaps) that we are not currently doing, is to
-have nodes marked as incertae sedis on the synth tree.
-This would be
-easy enough if such nodes are not affected in any way by the input
-trees.
-Thus, when unpruning nodes we could mark any nodes
-\emph{incertae sedis} if they were marked \emph{incertae sedis} on the
-taxonomy.
-
-Secondly, I think we need to distinguish \emph{incertae sedis} taxa
-that are ``unplaced'' from \emph{incertae sedis} taxa that do not
-occur in any input tree.
-I think that if $A$ contains child $A_{1}$
-that is an input tree, and the taxon $A$ is not broken, then $A$ will
-be placed, and thus any other children $A_{2},A_{3},\ldots,A_{n}$ will
-also be placed, since they will be added as children of the (placed)
-node $A$ by the unpruner.
-This could be considered when deciding which
-nodes to suppress in the tree viewer.
 
@@
 
-\subsubsection{Comparison to operational definition}
-
-The splits-based semantics for incertae sedis taxa has many desirable
-properties.
-In allows the use of the BUILD algorithm to assess
-compatibility of split sets.
-However, it also has some properties that
-may not be expected.
-For example, two incertae sedis siblings can be
-freely interdigitated, which might not be the expected outcome.
-The
-current definition, also ignore ranks.
-Thus if we had two incertae
-sedis siblings that were genera, but other siblings were families,
-then one might expect that the genus-level incertae sedis taxa could
-be placed within families, but not intermixed.
-Since the current
-approach ignores ranks, it cannot do that.
-
-The above definition of incertae sedis taxa does allow two incertae
-sedis sister taxa to be interdigitated.
-It also
-
-Also, the splits-based semantics might not completely recover the
-edit-operation semantics.
-The corner cases are ...
-.
-
-\subsubsection{Intrusion into incertae sedis taxa}
-
-In theory it could be possible for an \emph{incertae sedis} taxon $A$
-to be placed within an \emph{incertae sedis} sibling $B$.
-We initially
-implemented this semantics.
-It requires updating the recursion to
-define $I(m,n)$ as the descendants of $m$ that can be access from $n$
-without traversing an \emph{incertae sedis} node.
-
-With a splits-based semantics, we cannot allow both (i) placing $A$ in
-$B$ and (ii) placing $B$ in $A$ and also forbid interdigitating $A$
-and $B$.
-
-However, if we allow this behavior by reducing the exclude sets for
-$A$ and $B$, then this has the unfortunate consequence that the
-members of $A$ and $B$ may be inter-digitated.
-In many cases,
-\emph{incertae sedis} genera have siblings that are families, and the
-expected semantics is that genera may be placed within their sibling
-families, or may remain outside existing families.
-In such cases,
-inter-digitating the genera would not be expected behavior.
-Thus, we
-choose the semantics that \emph{incertae sedis} taxa exclude their
-\emph{incertae sedis} siblings.
-This is not because the other
-semantics is impossible.
-However, we note that use of the
-interdigitating semantics allows a situation where the names for $A$
-and $B$ both apply to the same node.
-
-This semantics also complicates placement, since it is possible to
-place $A$ as a sibling of $B$, and then place $B$ within $A$.
-This
-violates the current invariant that taxa are only ever placed within
-more tipward taxa.
-This makes the description of placement more
-complicated, since we can no longer simply record all cases where a
-parent node is not ancestor on the taxonomy.
-
-\subsubsection{Usage in the literature}
-
-The term \emph{incertae sedis} can be used in a number of different
-ways.
-For example,\emph{ }one author might describe \emph{incertae
-sedis} genera that are asserted to be monophyletic, whereas another
-author might describe \emph{incertae sedis} genera that may be
-interdigitated.
-
-\appendix
-\section{Alternative semantics}
-\subsubsection{semantics \#1b: modified to disallow inter-digitating
-\emph{incertae sedis} siblings}
-
-We additionally stipulate that it is not possible for \emph{incertae sedis} taxa
-to intrude into their \emph{incertae sedis} siblings.  We thus define $\mathcal{I}(m,n)$ to
-indicate descendants of $m$ that are excluded from $n$. Thus:
-\begin{align*}
-\mathcal{I}(m,n) & =
-    \begin{cases}
-        \emptyset & \text{if }n\text{ is \emph{incertae sedis}\,and }m\text{
-is not}\\
-        \mathcal{I}(m) & \text{otherwise}
-    \end{cases}
-\end{align*}
-
-We can write the exclude set of a node $n$ in terms of
-the exclude set of its parent:
-\begin{eqnarray*}
-    \mathcal{E}(n) & = \mathcal{E}(parent(n))\cup\left[\mathcal{I}(m,n)\big|m\in
-           siblings(n)\right].
-           \label{eq:exsetformone}
-\end{eqnarray*}
-\subsubsection{An alternative semantics that allows placing an incertae sedis group within an incertae-sedis great-uncle}
-
-\bdrcomment{This semantics is actually designed to allow the situation in Figure \ref{fig:Placing-one-incertae}.
-
-  The problem is (might be) as follows: if we have the tree ``((?A1,A2)A,((D)C)B)MRCA;'', where A1 is incertae sedis, then we can place A1 within D.
-This semantics does not exclude A1 from B, C, or D.
-So, if A is broken, then nothing prohibits A1 from intruding into B, C, D, etc.
-
-  The purpose of this was to allow placing B as a siblign to A1 and then placing A1 within B -- \emph{if} B were incertae sedis.  But B is not incertae sedis.
-
-  We could try to fix up this situation by adding cases to the exclude rules to say that B excludes A1 if B is not incertae sedis.
-But adding cases makes things messier.
-I think it also does not actually solve the problem: suppose we mark B incertae sedis, but do NOT place it as sibling to A1.
-Then again if A is broken, A1 can intrude into B.
-We really want to allow A1 into B only if B is placed as sibling to A1, and I'm not sure how to do that with the partial-split-based semantics.
-If you see a way, then great!
-}
-
-If we want to allow placing two incertae sedis siblings within each
-other, then incertae sedis siblings are never excluded.
-
-Furthermore, consider placing $A$ as a sibling of $B$, and then
-placing $B$ within $A$.
-In order for this to work, $A$ must not
-exclude the incertae sedis children of $B$.
-
-In order to accomplish both of these goals, we define the descendants
-$\mathcal{I}^{\prime}(n)$ that are accessible without passing through
-an incertae sedis node.
-
-\begin{align*}
-    \mathcal{I^{\prime}}(n) & =
-    \begin{cases}
-        \emptyset & \text{if }n\in\mathcal{U}\\
-        \{n\} & \text{if }n\text{ is a leaf}\\
-        [\mathcal{I}^{\prime}(c)|c\in children(n)] & \text{otherwise.}
-\end{cases}
-\end{align*} We may then define the taxa excluded from $n$
-as:
-
-\begin{eqnarray*} 
-\mathcal{E}(n) & =\mathcal{E}(parent(n))\cup\left[\mathcal{I}(m,n)\big|m\in siblings(n)\right].\label{eq:exsetformoneb}
-\end{eqnarray*}
-
-This semantics satisfies criteria 1,3,4, and 5 above.
-
-\subsubsection{A third semantics}
-
-\bdrcomment{I didn't really finish working this one out.}
-
-OK, the idea here is that we want to avoid inter-digitation.
-\emph{Incertae sedis} taxa thus exclude\emph{ incertae sedis}
-siblings.
-However, where the first semantics allows \emph{incertae
-sedis }taxa to intrude into \emph{incertae sedis} descendants of
-siblings, but not allow\emph{ incertae sedis} descendants of siblings
-to intrude into \emph{incertae sedis} taxa, here we do the reverse.
-
-OK, so the idea here is that an\emph{ incertae sedis} group excludes
-\begin{itemize} \item \emph{incertae sedis} siblings \item
-\emph{incertae sedis }descendants of ancestors \item but not\emph{
-incertae sedis} descendants of non-incertae-sedis siblings
-\end{itemize} It is also possible to allow placing $A$ as a sibling of
-$B$ and then placing $B$ within $A$ if we disallow placing $A$ into
-incertae sedis descendants of its siblings.
-We may then allow placing
-those descendants within $A$.
-
-\begin{align*} \mathcal{I^{\prime}}(n) & =\begin{cases} \emptyset &
-\text{if }n\in\mathcal{U}\\ \{n\} & \text{if }n\text{ is a leaf}\\{}
-[\mathcal{I}^{\prime}(c)|c\in children(n)] & \text{otherwise.}
-\end{cases} \end{align*}
-
-We use $\mathcal{I}(m,n)$ to indicate the descendants of $m$ that are
-excluded from $n$.
-\begin{align*} \mathcal{I}(m,n) & =\begin{cases}
-\emptyset & \text{if }n\text{ is \emph{incertae sedis}\,and }m\text{
-is not}\\ \mathcal{I}(m) & \text{if }n\text{ and }m\text{ are both
-}incertae\,sedis\\ \mathcal{I}^{\prime}(m) & \text{otherwise.}
-\end{cases} \end{align*}
-
-We may then define the taxa excluded from $n$ as:
-\begin{align*}
-\mathcal{E}(n) & =\begin{cases}
-\left[\mathcal{L}-\mathcal{I}(parent(n))\right]\cup[\mathcal{I}(s,n)|s\in siblings(n)] & \text{if }n\in\mathcal{U}\\
-\mathcal{E}(parent(n))\cup[\mathcal{I}(s,n)|s\in siblings(n)] & \text{otherwise.}
-\end{cases}
-\end{align*}
-I think this works... but need to double-check.
-
-Also: can we genericize $\mathcal{E}(parent(n))$?
-
-\section{Details \& Methods}
-
-\subsection{Cases of incertae sedis taxa}
-1.1 Some examples in OTT This paragraph pulled from intro to
-otcetera/doc /handling-incertae-sedis.pdf
-
-The problem is becoming more acute because NCBI is putting a larger
-number of taxa into groups that are marked as ``classified''. For
-example, when OTT 2.9 was created NCBI's clasification of the bird
-family Sylvidae included a group five genera that were placed within
-\url{http://www.ncbi.nlm.nih.gov/Taxonomy/Browser/wwwtax.cgi?mode=Undef&id=36270&lvl=3&lin=f&keep=1&srchmode=1&unlock||unclassified Sylviidae}.
-This includes the genus Regulus. Thus in OTT 2.9 Regulus (ott ID = 3599326) is placed inside the family Silvidae (OTT ID =
-259942).
-The ``unclassified Sylviidae'' does not appear in OTT; instead
-Regulus is flagged as ``unclassified,sibling higher'' and all of the
-species within Regulus are flagged as ``unclassified inherited.''
-
-The taxonomy (as of version 2.9) also contains 685 cases of taxa that
-are flagged as both ``unclassified inherited'' and ``unclassified'' based
-on {\tt grep unclass.*unclassified\_inherited taxonomy.tsv}.
-
-We would like to stop suppressing (pruning) incertae sedis taxa, so
-that groups such as Regulus can appear appear in the synthetic tree.
-
-\bdrcomment{We get Regulus back, bu we do not end up placing it in Sylviidae
-because Sylviidae is broken. It conflicts with about 20 different trees.}
-
-\subsection{Taxonomy details}
-
-\paragraph{Flags.}
-The Open Tree's reference taxonomy is produced by a tool called
-smasher that notices hints that a taxon is incertae sedis and labels
-that taxon with one of five flags (see https://github.com/OpenTreeOfLife
-/reference-taxonomy/wiki /Taxon-flags). This document is not concerned
-with the flagging system per se, so ``incertae sedis'' will be used here
-to refer to all of the flags that denote taxa with uncertain
-placement.
-
-\paragraph{Limited attachment range}
-While a taxonomist may intend to use incertae sedis to
-indicate a limited number of possible positions for a taxon to go in
-the taxonomy, we do not retain any such details.
-
-\paragraph{Container nodes.}
-Some taxonomies have nodes with names like ``Incertae sedis (Bacteria)''.
-Such nodes are called containers. They indicate that each child node is an
-incertae sedis child of the container node's parent. The OTT retains container
-nodes with the was\_container flag, but seems to have already moved all of the
-container's children to the container's parent.
-
-\section{Placements}
-\textbf{Whale:}~ ?ott6145835~{[}6145835~Brandtocetus/genus{]}~<-~ott698424~{[}698424~Cetacea/orde r{]}~placed~under~ott4942423~{[}4942423~Cetotherium~/~genus{]}~<-~ott698424~{[}6 98424~Cetacea/order{]}
-
-\textbf{Whale:}~?ott6145873~{[}6145873{]}~<-~ott698424~{[}698424~Cetacea~/~order{]}~placed~under~ott4942423~{[}4942423{]}~<-~ott698424~{[}698424{]}~
-
--~Cetacea~broken.
-
-\textbf{Insect:}~?ott18090~{[}18090{]}~<-~ott672501~{[}672501{]}~placed~under~ot t275501~{[}275501{]}~<-~ott672501~{[}672501{]}
-
-\textbf{Insect:}~?ott190269~{[}190269{]}~<-~ott672501~{[}672501{]}~placed~under~ ott275501~{[}275501{]}~<-~ott672501~{[}672501{]}~
-
-\textbf{Insect:}~?ott349638~{[}349638{]}~<-~ott672501~{[}672501{]}~placed~under~ ott275501~{[}275501{]}~<-~ott672501~{[}672501{]}~
-
-\textbf{Insect:}~?ott1009313~{[}1009313{]}~<-~ott672501~{[}672501~Reduviidae~/~f amily{]}~placed~under~ott275501~{[}275501{]}~<-~ott672501~{[}672501{]}~
-
--~Reduviidae~broken.
-
-\textbf{Insect:~}?ott564110~{[}564110{]}~<-~ott939130~{[}939130~Thyrididae~/~family{]}~placed~under~ott757344~{[}757344{]}~<-~ott939130~{[}939130{]}~
-
-\textbf{Plant:}~?ott679485~{[}679485{]}~<-~ott1074933~{[}1074933~Lamioidea~/~sub family{]}~placed~under~ott534796~{[}534796{]}~<-~
-
-\textbf{Plant:}~?ott729184~{[}729184{]}~<-~ott1074933~{[}1074933~Lamioidea~/~sub family~{]}~placed~under~ott534796~{[}534796{]}~<-~ott1074933~{[}1074933{]}~
-
-ott1074933~{[}1074933{]}~
-
--~Lamioidea~\emph{not}~broken
-
-\textbf{Plant:}~?ott5144555~{[}5144555{]}~<-~ott23373~{[}23373{]}~placed~under~o tt447802~{[}447802{]}~<-~ott23373~{[}23373{]}
-
--~Orobanchaceae~broken
-
-\textbf{Fly:}~?ott441413~{[}441413~Rhodesiella~lungliensis/species{]}~<-~ott1014 022~{[}1014022~Rubiaceae{]}~placed~under~ott571842~{[}571842~Pavetteae~/~tribe{ ]}~<-~ott997381~{[}997381~Ixoroidea/subfamily{]}~<-~ott1014022~{[}1014022~Rubia ceae/family{]}
-
-\textbf{Plant:}~?ott31033~{[}31033{]}~<-~ott627035~{[}627035{]}~placed~under~ott 6019314~{[}6019314{]}~<-~ott627035~{[}627035{]}~
-
-\textbf{Bacteria:}~?ott4018918~{[}4018918{]}~<-~ott5248084~{[}5248084{]}~placed~ under~ott167767~{[}167767{]}~<-~ott5248084~{[}5248084{]}~
-
-\textbf{Bacteria:}?ott4792759~{[}4792759{]}~<-~ott5248084~{[}5248084{]}~placed~u nder~ott167767~{[}167767{]}~<-~ott5248084~{[}5248084{]}~
-
--~Erysipelotrichaceae~broken
-
 \bibliographystyle{upmplainnat}

### 2018-05-10 doc/incertae_sedis.tex
pruned
up to last couple of pages

@@
 While the topology for the final tree in either order of analyses would indeed be the tree shown in 
-    Figure \ref{fig:Placing-one-incertae}b}, the decision about whether or not to name $V$ would differ.
+    Figure \ref{fig:Placing-one-incertae}b, the decision about whether or not to name $V$ would differ.
 Our handling of \emph{incertae sedis} in the original taxonomy via rooted partial splits
@@
 The representations only vary in the scoring of leaf $W$ for character $V$. Panel (c) shows the information encoded by the tree in \ref{fig:Placing-one-incertae}c. }
-\end{figure}
-
-\begin{figure}
-\subfloat[Taxonomy $\protect\taxonomy$]{\includegraphics[scale=0.5]{Figures/placement1/tax\lyxdot tre}
-
-}\subfloat[$\tau_{1}$]{\includegraphics[scale=0.5]{Figures/placement1/input1\lyxdot tre}
-
-}\subfloat[$\tau_{2}$]{\includegraphics[scale=0.5]{Figures/placement1/input2\lyxdot tre}
-
-}
-
-\caption{\label{fig:placement-example1}Example.
-    An incertae sedis clade (ott6,ott7)
-    is placed in different subtrees by input trees $\tau_{1}$ and $\tau_{2}$.
-    In $\tau_{1}$, two nodes that correspond to the taxonomy their ingroup
-    extended to include (ott6,ott7), and the branches leading to these
-    nodes have been colored blue.
-    The dashed blue edge leads to a node
-    that is a newly-introduced degree-2 node which does not correspond
-    to any taxonomy node.
-    In $\tau_{2}$, only one node that corresponds
-    to a taxonomy node needs to have its ingroup extended.
-    The placement
-    of (ott6,ott7) into ott8 toward ott1 by $\tau_{1}$ conflicts with
-    the placement of (ott6,ott7) into ott9 toward ott3 by $\tau_{2},$}
 \end{figure}
@@
 
+\newpage %TEMP
+
+\begin{figure}
+\subfloat[$\tau_{1}$]{\includegraphics[scale=0.5]{Figures/decomp/tree1}
+}
+\subfloat[$\tau_{2}$]{\includegraphics[scale=0.5]{Figures/decomp/tree2}
+}
+\subfloat[$\tau_{3}$]{\includegraphics[scale=0.5]{Figures/decomp/tree3}
+}
+\subfloat[$\tau_{4}$]{\includegraphics[scale=0.5]{Figures/decomp/tree4}
+}
+
+\subfloat[Taxonomy $\protect\taxonomy$]{\includegraphics[scale=0.5]{Figures/decomp/tax}
+}
+\subfloat[$Y$ and $Z$ decomposition]{\includegraphics[scale=0.5]{Figures/decomp/decompRep}
+}
+\subfloat[Pruned $Y$ and $Z$ decomposition]{\includegraphics[scale=0.5]{Figures/decomp/decompPruned}
+}
+
+\caption{\label{fig:decomp} Four phylogenetic trees (a-d) an
+example taxonomy (e) which are possible inputs to the propinquity supertree pipeline.
+(f) shows the taxonomic component of the subproblem decompositions for taxa $Y$ and $Z$ subproblems 
+that would result
+from an unmodified pipeline with used $\tau_1$ and either $\tau_2$ or $\tau_3$ as the phylogenetic
+inputs.
+(g) shows a possible decomposition from a decomposition tool that avoids duplication of taxa.
+}
+\end{figure}
+
 \subsection{Sub-problem decomposition}
@@
 
-We must also handle conflicting placements of \emph{incertae sedis}
-taxa by different input trees.
-Thus, if one input tree places the
-\emph{incertae sedis} taxon $X$ in $((X)B)A$ and another places $X$ in
-$((X)C)A$ then we must mark both edges $B$ and $C$ as contested edges,
-even if these edges would \emph{not} be contested were taxon $X$ to be
-removed.
-This results in a new way to contest edges that involves the
-interaction of two input trees, and not just the interaction of each
-input tree with the taxonomy.
+Simply allowing the subproblem decomposer to recognize that intrusion
+    of an \emph{incertae sedis} taxon does not contest the existence
+    of a clade would be straightforward, but problems arise if the
+    decomposer is not further altered.
+Consider, for example, the inputs of the $\tau_1$, $\tau_2$, $\tau_3$, and the taxonomy
+    shown in Figure \ref{fig:decomp}.
+Because the taxon $\alpha$ is annotated as being \emph{incertae sedis}, 
+    neither  $\tau_1$, $\tau_2$, nor $\tau_3$ contest the monophyly of 
+    the taxa $Y$, $Z$, or $\alpha$.
+This would lead to the taxonomic portion of both the $Y$ and $Z$ subproblems
+    containing the taxonomy for $\alpha$, leading to the members of
+    $\alpha$ occurring in two different subproblems (see Figure \ref{fig:decomp}f).
+Solving and merging these problems would result in the duplication of $\alpha$ 
+    in the supertree.
+
+One could easily
+    imagine a ``pruning decomposer'' that does not allow the say taxon to
+    occur in more than one subproblem.
+This approach would work in a straightforward way if the phylogenetic inputs were only $\tau_1$ and $\tau_3$,
+    yielding the decomposition shown in Figure \ref{fig:decomp}g, but if
+    both $\tau_1$ and $\tau_2$ were inputs the decomposer would have to
+    resolve conflict about the placement of $\alpha_1$.
+Implementing the behavior in the subproblem decomposition step would
+    break the separation between the ``divide'' and ``conquer'' steps
+    which is a key aspect of propinquity's current efficiency.
+
+Even considering ranked phylogenetic inputs of $\tau_4$ then $\tau_1$ from
+    Figure \ref{fig:decomp} reveal some difficulties.
+In this case the inclusion of $\alpha$ within the taxon $Y$ in $\tau_1$ would
+    cause $\alpha$ to be solved in the subproblem for $Y$.
+If $\tau_4$ were ranked higher than $\tau_1$, then the grouping of $Y_1+Y_2+\alpha$ would
+    not be permitted because they conflict with the split $Y_1\,Y_3 \mid \bullet\,Y_2$ from
+    $\tau_4$.
+Thus, $\alpha$ would be placed inside of $Y$ on the basis of a rooted split
+    that is not displayed in the final tree.
+While this is placement is permissible based on the \emph{incertae sedis} annotation of the
+    taxonomy, a major goal of the propinquity pipeline is to make it
+    easy for biologists to understand the cause of different groupings in the 
+    supertree.
+Allowing splits which are not displayed to determine the placement of groups runs counter to that goal.
 
@@
 subproblem decomposer.
-One downside of this approach is that, if we
-have $((X)B)A)$ in one input tree, and $X$ is mentioned nowhere else,
-then by marking $B$ as contested, we are creating unnecessarily large
-subproblems.
-We could instead placed $X$ in $B$ and avoid contesting
-the edge $B$.
-However, this approach is more complex and does not seem
-necessary in practice.
-
- For example, in Figure
-\ref{fig:placement-example1}, input tree $\tau_{1}$ contests ott9 and
-input tree $\tau_{2}$ contests ott9 and ott8.
-Thus ott1, ott2, ott3,
-ott6, and ott7 end up in the same sub-problem.
+One downside of this approach is that, an \emph{incertae sedis} taxon
+    place in only one input tree, the clade it is placed in will be
+    marked as contested even though none of the counterintuitive
+    behaviors mentioned above would arise if we did not
+    mark it as contested.
 
@@
 uncertainty.
-
-\subsubsection{Reading incertae sedis information}
-
-Currently, we read the \emph{incertae sedis} information as a list of
-OTT ids for \emph{incertae sedis} taxa.
-This does not require adding
-further annotations to the node names.
-Only taxonomy nodes can be
-\emph{incertae sedis} at the moment, and only the taxonomy tree for
-the subproblem contains OTT ids for internal nodes.
-Therefore we
-handle \emph{incertae sedis} information by constructing modified
-split sets for the lowest-ranked tree when the list of \emph{incertae
-sedis} nodes is not empty.
-
-\subsubsection{Exclude sets modified by \emph{incertae sedis} marks}
-
-Equation (\ref{eq:exsetformone}) leads to the following
-algorithm to compute the exclude set for all nodes in a tree.
-\begin{enumerate} \item Set the exclude set of the root node to be
-empty \item For each \emph{node} (except the root) in preorder
-\begin{itemize} \item combine the \emph{exclude} set of the parent
-node with the \emph{include} set of non-\emph{incertae-sedis}
-siblings.
-\item store this set in a hash, with key \emph{node}
-\end{itemize} \end{enumerate} This algorithm is currently implemented
-in \emph{otc-solve- subproblem}.
-We store the sets as \emph{std::set}.
 

### 2018-05-21 doc/incertae_sedis.tex
Small edits.

@@
     over the tree from root to tip.
-If $\mathcal{S}(x)$ is the set of siblings of $x$and $p(x)$ is the parent of $x$, then
-the naive approach of ignoring the \emph{incertae sedis}  annotation amount to setting
+If $\mathcal{S}(x)$ is the set of siblings of $x$ and $p(x)$ is the parent of $x$, then
+the naive approach of ignoring the \emph{incertae sedis}  annotation amounts to setting
     the exclude set as:
@@
 If we treat the  \emph{incertae sedis} annotation as an indication that the 
-    \emph{incertae sedis} taxon $x$ can be place in any of its sister taxa, then
+    \emph{incertae sedis} taxon $x$ can be placed in any of its sister taxa, then
     we let $\mathcal{Q}(x)$ denote the set of siblings of $x$ that are not
@@
 \caption{\label{fig:interdigitation}Interdigitation of \emph{incertae sedis} siblings.
-Here the names $A$ and $B$ apply to the same node on the solution tree, but neither is ancestral to the other on the taxonomy.}
+Here the names $P$ and $Q$ apply to the same node on the solution tree, but neither is ancestral to the other on the taxonomy.}
 \end{figure}

### 2018-05-29 doc/incertae_sedis.tex
Spacing.
Added some drafty discussion text.

@@
 \end{tabular}
-}
+}\hfill{}
 \subfloat[less informative]{
@@
 \end{tabular}
-}
+}\hfill()
 \subfloat[from revised taxonomy]{
@@
 \subfloat[$\tau_{1}$]{\includegraphics[scale=0.5]{Figures/decomp/tree1}
-}
+}\hfill{}
 \subfloat[$\tau_{2}$]{\includegraphics[scale=0.5]{Figures/decomp/tree2}
-}
+}\hfill{}
 \subfloat[$\tau_{3}$]{\includegraphics[scale=0.5]{Figures/decomp/tree3}
-}
+}\hfill{}
 \subfloat[$\tau_{4}$]{\includegraphics[scale=0.5]{Figures/decomp/tree4}
@@
 \subfloat[Taxonomy $\protect\taxonomy$]{\includegraphics[scale=0.5]{Figures/decomp/tax}
-}
+}\hfill{}
 \subfloat[$Y$ and $Z$ decomposition]{\includegraphics[scale=0.5]{Figures/decomp/decompRep}
-}
+}\hfill{}
 \subfloat[Pruned $Y$ and $Z$ decomposition]{\includegraphics[scale=0.5]{Figures/decomp/decompPruned}
@@
 
-\textbf{BDR:}\textbf{\emph{ }}\emph{we might get this behavior in a
+\bdrcomment{we might get this behavior in a
 nice an automatic way if we create a single fake leaf for each
@@
 to the conflict-analysis code in the annotation step.
+
+\section{Discussion}
+
+In this paper we've described a method for handling \emph{incertae sedis} taxa
+when constructing taxonomic supertrees.  Our method incorporates the uncertain
+placement information from such taxa when constructing the supertree.  It also
+handles higher taxon names and allows input trees to revise the taxonomy by
+determining where \emph{incertae sedis} taxa should be placed.
+
+However, our method leaves a number of problems for future work.  One of these
+problems is that we consider taxa with an \emph{incertae sedis} label to 
+about which siblings in attach within.  However, in practice,
+the taxonomist may only consider it plausible for the taxon to intrude only within specific
+siblings.  Our method also assumes that an \emph{incertae sedis} label on a taxon
+indicates that a taxon may attach at any depth within its siblings.  In practice, an
+\emph{incertae sedis} genus with sister families (for example) is unlikely to attach
+so deeply within a sister family that it becomes nested within another genus.
+
+In order to consider more precise attachment ranges for \emph{incertae sedis} taxa, we
+would have to obtain specifications for attachment ranges that are machine readable and
+robust.  Obtaining these specifications would be difficult or impossible in many cases.
+However, in cases that result from taxonomy merging (below) it would be possible to obtain
+them.  This necessitates the creation of a format \emph{incertae sedis} taxa that is more
+expressive that a simple label.
 

### 2018-06-12 doc/incertae_sedis.tex
Update.

@@
 that groups $A$ and $B$. Taxon names and edges that are conditional on handling
-\emph{incertae sedis} are shaded grey. \mthcomment{We should probably remove $C3$ and replace $B1$ and $B2$ with $B$
-to simplify this. We could use triangle to indicate that $D$ and $B$ could be leaves or clades.}
+\emph{incertae sedis} are shaded grey. \mthcomment{We could use triangle to indicate that $D$ and $B$ could be leaves or clades.}
 }
@@
 The rooted split associated with taxonomy node $n$ serves to determine both when $n$ is in conflict
-with an input tree, and to where to place the taxon name for clade $n$ on the synthesis tree.
+with an input tree, and where to place the taxon name for clade $n$ on the synthesis tree.
 % Removed mention of incertae sedis possibly affecting subproblem decomposition.  It
@@
 removing the edge connecting the node to its parent would produce two trees.
-The leaf label sets of these trees
-the partition the full leaf set $\mathcal{L}$ into two groups.
-The include set $\mathcal{I}(n)$ which
-does not contain the root, and the exclude set $\mathcal{E}(n)$ which does contain the root.
-Such a split may be written
+The leaf label sets of these trees partition the full leaf set $\mathcal{L}$ into two groups: the include
+set $\mathcal{I}(n)$ which does not contain the root, and the exclude set $\mathcal{E}(n)$
+which does contain the root. Such a split may be written
 \[ \mathcal{I}(n)|\bullet\mathcal{E}(n).\]
-where the $\bullet$ indicates the root. 
-The include set is the set of labels for leaves that are
-descendants of the node.
-The exclude set is the set of
+where the $\bullet$ indicates the root. The include set is the set of labels for leaves that are
+descendants of the node. The exclude set is the set of 
 labels for leaves that are not descendants of that node.
@@
 
-\paragraph{Equivalence rooted split definitions and phylogenetic definitions of names}
+\paragraph{Equivalence of rooted split definitions and phylogenetic definitions of names}
 In terms of phylogenetic nomenclature the supertree pipeline acts as if the
@@
 
-\paragraph{Equivalence rooted split definitions with character encodings}
+\paragraph{Equivalence of rooted split definitions with character encodings}
 Representing a split in a tree by a character is the basis of one of the oldest
@@
 In the rooted form of such an approach, character state 1 is used for tips that are
-    descendants of a internal node in an input tree and 0 is used for other tips.
+descendants of a internal node in an input tree and 0 is used for other tips and for the root.
 The partial splits caused by \emph{incertae sedis} annotations naturally map to
@@
 \hline
-& \multicolumn{4}{|c|}{Character/clade label} \\
-leaves & C & AB & B & A \\
+& \multicolumn{3}{|c|}{Character/clade label} \\
+leaves & C & AB & A \\
 \hline
-$A1$, $A2$ & 0 & 1 & 0 & 1 \\
-$B1$, $B2$ & 0 & 1 & 1 & 0 \\
-$C1$, $C2$, $C3$ & 1 & ? & ? & ?  \\
-$D$  & 0 & 0 & 0 & 0 \\
+$A1$       & 0 & 1 & 1 \\
+$A2$       & 0 & 1 & 1 \\
+$B$        & 0 & 1 & 0 \\
+$C1$       & 1 & ? & ? \\
+$C2$       & 1 & ? & ? \\
+  $D$      & 0 & 0 & 0 \\
+  root     & 0 & 0 & 0 \\
 \hline
@@
 of an \emph{incertae sedis} interpretation of the taxonomy shown in  Figure \ref{fig:Handling-incertae-sedis}b.
-Only the unique character encodings are show, thus the leftmost column
-lists all of the tip taxa that would be assigned a particular character encoding.
+The root is coded as a special leaf.
 }
@@
 \mthcomment{Need to verify that the namer records this situation as synonyms, as it should!}
+\bdrcomment{I assume this is not linked to publication}
 
@@
 One could easily
-    imagine a ``pruning decomposer'' that does not allow the say taxon to
+    imagine a ``pruning decomposer'' that does not allow the same taxon to
     occur in more than one subproblem.
@@
 solution tree.
-The solution tree is considered to a fixed tree and not
+The solution tree is considered to be a fixed tree and not
 to have any \emph{incertae sedis} nodes, or any other forms of

### 2018-06-13 doc/incertae_sedis.tex
Update.

@@
 
-\newcommand{\mthcomment}[1]{{\color{red} #1}\xspace}
-
-\newcommand{\bdrcomment}[1]{{\color{cyan} #1}\xspace}
+%\newcommand{\mthcomment}[1]{{\color{red} #1}\xspace}
+\newcommand{\mthcomment}[1]{}
+
+%\newcommand{\bdrcomment}[1]{{\color{cyan} #1}\xspace}
+\newcommand{\bdrcomment}[1]{}
 
@@
 descendants of a internal node in an input tree and 0 is used for other tips and for the root.
-The partial splits caused by \emph{incertae sedis} annotations naturally map to
-an encoding of missing data (typically denoted with the `?' symbol in a matrix).
-Thus the taxonomy shown in Figure \ref{fig:Handling-incertae-sedis}b could be represented as the matrix shown in 
-Figure \ref{fig:Handling-incertae-sedis-mrp}.
-In the naive treatment of ignoring the \emph{incertae sedis} annotation, the `?' characters
-   in the row for taxa $C1$, $C2$, and $C3$ would be replaced with 0's.
+Partial splits, such as those caused by \emph{incertae sedis} annotations, are naturally encoded by
+encoding missing labels in a split as missing data in the corresponding character
+(typically denoted with the `?' symbol in a matrix).
+
+A rooted split is equivalent to the corresponding character in a matrix representation in the
+sense that each can be treated as a condition that each tree may or may not satisfy.  A tree satisfies
+a split if there a branch which divides include and exclude group when cut.  A tree satisfies the
+corresponding character in a matrix representation if the character has a parsimony score of $1$ on the tree.
+This equivalence is illustrated in Figure \ref{fig:Handling-incertae-sedis-mrp}.
+Thus the taxonomy shown in Figure \ref{fig:Handling-incertae-sedis}b could be represented in terms of the splits
+in Figure \ref{fig:Handling-incertae-sedis-mrp}a or the characters in Figure \ref{fig:Handling-incertae-sedis-mrp}b.
+In the naive treatment of ignoring the \emph{incertae sedis} annotation, the `?' characters in the row for taxa $C1$ and $C2$ would be replaced with 0's.
+Equivalently, the naive interpretation would add $C1$ and $C2$ to the exclude set for the $A$ and $AB$ splits.
 
    \begin{figure}
-\begin{center}
+     \subfloat[Split representation]{
+\noindent\begin{minipage}[t]{0.45\columnwidth}%
+\begin{align*}
+C & =C1\,C2\,\phantom{B\,}|\bullet\,A1\,A2\,B\,D\\
+A & =A1\,A1\,\phantom{B\,}|\bullet\,B\,D\\
+AB & =A1\,A2\,B\,|\bullet\,D
+\end{align*}
+%
+\end{minipage}
+
+}
+     \subfloat[Matrix representation]{
          \begin{tabular}{|r|p{2em}|p{2em}|p{2em}|p{2em}|}
@@
            $D$      & 0 & 0 & 0 \\
-  root     & 0 & 0 & 0 \\
+           (root)     & 0 & 0 & 0 \\
            \hline
          \end{tabular}
-\end{center}
-\caption{\label{fig:Handling-incertae-sedis-mrp} The matrix encoding 
-of an \emph{incertae sedis} interpretation of the taxonomy shown in  Figure \ref{fig:Handling-incertae-sedis}b.
+}
+       \caption{\label{fig:Handling-incertae-sedis-mrp} Rooted splits have an equivalent matrix encoding.
+        (a) Rooted splits for an \emph{incertae sedis} interpretation of the taxonomy shown in  Figure \ref{fig:Handling-incertae-sedis}b.
+(b) The equivalent matrix representation.
 The root is coded as a special leaf.
+Split C is a full split, because it mentions all labels.
+Splits A and AB are partial splits, and therefore have missing labels in the left panel, and '?' in the right panel.
        }
@@
 ancestors for \emph{incertae sedis} taxon $n$.
-Note that this distinct from the taxa that are ancestral to $n$ in the taxonomy;
+Note that this is distinct from the taxa that are ancestral to $n$ in the taxonomy;
     those are taxa which must contain all of the descandants of $n$.
@@
 
-\newpage %TEMP
+% \newpage %TEMP
 

### 2018-06-15 doc/incertae_sedis.tex
very minor

@@
 for example,][]{HorandlS2010}, the priniciples of phylogenetic classification
-are used by such a large majority of practicing taxonomists, that we have chosen
+are used by such a large majority of practicing taxonomists.
+Thus we have chosen
 to treat the taxonomic tree that mirrors the hierarchy of the taxonomy as an
@@
 }taxa recovers additional edges and taxon names.
-Supertree construction on input tree (a) and taxonomy tree (b) with \emph{incertae sedis}
-clade $C$ leads to synthesis tree (c) if \emph{incertae sedis} annotations are
-ignored, but to synthesis tree (d) if \emph{incertae sedis} annotations are respected.
+Panel (a) shows the only phylogenetic to the supertree algorithm;
+(b) shows taxonomy tree the edge subtending the \emph{incertae sedis}
+clade $C$ shown using a dashed line.
+The tree in (c) would be the output of the supertree algorithm
+if \emph{incertae sedis} annotations are simply ignored.
+The tree shown in (d) is the supertree produced 
+if \emph{incertae sedis} annotations are respected.
 The tree in (d) recovers taxon names $AB$ and $A$, as well as the edge to clade $AB$
@@
 This approach incorrectly treats the intrusion of an \emph{incertae sedis} taxon
-into a sibling taxon as a conflicting with monophyly of the sibling taxon. For
-example, in Fig. \ref{fig:Handling-incertae-sedis}, the placement of the
+into a sibling taxon as a conflicting with monophyly of the sibling taxon.
+Consider, for example, the trees shown in Fig. \ref{fig:Handling-incertae-sedis}.
+In this figure (and throughout this chapter) dashed lines are used for edges that
+connect an \emph{incertae sedis} taxon to its parent.
+Note that if we were to use the naive semantics,  the placement of the
 \emph{incertae sedis} taxon $C$ within $A$ and $AB$ is considered

### 2018-06-15 doc/incertae_sedis.tex
Update Discussion.

--- 2018-06-13 doc/incertae_sedis.tex
@@
 
-%\newcommand{\mthcomment}[1]{{\color{red} #1}\xspace}
-\newcommand{\mthcomment}[1]{}
-
-%\newcommand{\bdrcomment}[1]{{\color{cyan} #1}\xspace}
-\newcommand{\bdrcomment}[1]{}
+\newcommand{\mthcomment}[1]{{\color{red} #1}\xspace}
+%\newcommand{\mthcomment}[1]{}
+
+\newcommand{\bdrcomment}[1]{{\color{cyan} #1}\xspace}
+%\newcommand{\bdrcomment}[1]{}
 
@@
 \section{Discussion}
-
-In this paper we've described a method for handling \emph{incertae sedis} taxa
-when constructing taxonomic supertrees.  Our method incorporates the uncertain
-placement information from such taxa when constructing the supertree.  It also
-handles higher taxon names and allows input trees to revise the taxonomy by
-determining where \emph{incertae sedis} taxa should be placed.
-
-However, our method leaves a number of problems for future work.  One of these
-problems is that we consider taxa with an \emph{incertae sedis} label to 
-about which siblings in attach within.  However, in practice,
-the taxonomist may only consider it plausible for the taxon to intrude only within specific
-siblings.  Our method also assumes that an \emph{incertae sedis} label on a taxon
+\bdrcomment{This is slightly exaggerated in the sense that there might be other
+  frameworks to incertae sedis taxa that do not require these exact moves.  Nonetheless,
+  I think this is not just a ``review'', in the sense that any well-founded approach
+  would have to work through all of these problems.  By working through them, we have
+  navigated the paths that subsequent approaches will need to walk.
+
+  I would argue that
+  the fact that our solution seems relatively simple should be seen as a success, and not
+  a sign that we haven't done enough work.  It often takes a lot of work to get a complex
+  solution down to a simple solution.
+
+  Additionally, the fact that we do not have
+  a lot of ``results'', in the sense of (say) large number of higher-taxon names that
+  are rescued by avoiding the naive approach, is a result of the fact that we didn't want
+  to do the work to add more input phylogenies with incertae sedis tips that will cause
+  problems.  We could always go back and try to get these ``results'', but the percieved value
+  wasn't very high.  Also, the algorithm that we came up with is itself a ``result''.
+}
+
+This paper introduces the necessary concepts and methods for handling \emph{incertae sedis}
+taxa in taxonomic supertrees.  We describe the goals of preserving internal edges,
+preserving higher taxon names, and preserving the opportunity to revise the taxonomy
+based on placement of \emph{incertae sedis} taxa.  In addition to explaining why
+the naive approach to taxonomic supertrees that ignores incertae sedis taxa sacrifices
+these three goals, we introduce a simple and satisfying way to achieve these goals
+through defining rooted splits for taxonomy edges with reduced exclude sets.  We also
+provide a simple and efficient recursion for computing these splits.
+
+Despite the simplicity of our approach, some complexities and conceptual difficulties
+remain.  While the correspondence of nodes and names is a simple one-to-one and onto
+relationship without \emph{incertae sedis} taxa, when \emph{incertae sedis} taxa are
+introduced, the relationship might be neither one-to-one (Figures \ref{fig:Multiple-names-can},\ref{fig:interdigitation}) nor onto (Figure \ref{fig:One-name-can}).
+This necessitates putting more thought into the semantics of taxon names, including
+whether names might become synonyms.  This may be considered a ``complexity''.
+
+In addition, our attempt to define reduced exclude sets for taxonomy splits yields
+a semantics that is not the only concievable semantics for \emph{incertae sedis}
+taxa (Figure \ref{fig:Placing-one-incertae}).  It allows \emph{incertae sedis} siblings to inter-digitate
+(Figure \ref{fig:interdigitation}), which might be a surprising result to the taxonomist.  Furthermore,
+in what might seem a lack of consistency, it does \emph{not} allow \emph{incertae sedis}
+taxa to intrude into each other if they become siblings through the result of
+placement, as is the case for taxa W and V in Figure \ref{fig:Placing-one-incertae}c.
+While each of these problems can be solved separately by adjusting the definition of the
+reduces exclude sets, no solution seems to be best in all respects.  This may be a result
+of the fact that split-based approaches to supertrees seek a simultaneous solution to a set
+of constraints.  This simultaneity can be contrasted with a solution that is phrased
+as a set of sequential operations to the taxonomy, as would be the case when (for example)
+(i) placing V within WX so that it is sister to W, (ii) marking V as no longer incertae sedis, and then (iii)
+placing W within V (which is no longer incertae sedis).  This maybe considered a ``conceptual difficulty''.
+
+Another issue that we have postponed for future work is how to more narrowly delimit
+the ranges in which \emph{incertae sedis} may attach.  The current paper allows
+\emph{incertae sedis} to attach anywhere within their sisters, but the taxonomist
+may only consider it plausible for the taxon to intrude only within specific
+siblings.  Furthermore, our method also assumes that an \emph{incertae sedis} label on a taxon
 indicates that a taxon may attach at any depth within its siblings.  In practice, an

### 2018-06-15 doc/incertae_sedis.tex
need some conclusion

--- 2018-06-15 doc/incertae_sedis.tex
@@
 C & =C1\,C2\,\phantom{B\,}|\bullet\,A1\,A2\,B\,D\\
-A & =A1\,A1\,\phantom{B\,}|\bullet\,B\,D\\
+A & =A1\,A2\,\phantom{B\,}|\bullet\,B\,D\\
 AB & =A1\,A2\,B\,|\bullet\,D
@@
 The effect of the \emph{incertae sedis} annotation of taxon $n$ is
-    to simply remove $\mathcal{I}(n)$ from the exclude set for each $p\in\mathcal{P}_n$.
+    to simply remove $\mathcal{I}(n)$ from the exclude set for each taxon in $\mathcal{P}_n$.
 The include sets of these splits remain unchanged, but the exclude sets may be reduced
@@
 Other placements of $T3$ would require at least three 0 to 1 changes (or a reversion
-    to state 0)l
+    to state 0).
 Placing $T3$ at the MRCA 
     of the other members of $T$ corresponds
-    to handling of ``rogue'' taxa in an Adams consensus method of the two
+    to handling of ``rogue'' taxa in an Adam's consensus tree \citep{Adams1972} method of the two
     most parsimonious placements of $T3$ under an irreversible parsimony model.
@@
     to $V$ being named.
-Thus, accurately
-    honors the input taxonomic content we would need more taxonomic information
+Thus, to honor the input taxonomic content accurately,
+   we need more taxonomic information
     than simply a hierarchy and knowledge of which taxa can float more tipward.
@@
     in an altered definition of taxon $V$, as shown in \ref{fig:Placing-one-incertae-MRP}c.
-As expected, in the encoding of the revised taxonomy, the definition of $WX$ has been altered to reflect the
-    fact $T_1$ placed $V$ inside $WX$.
+As expected,  the definition of $WX$ has been altered to reflect the
+    fact $T_1$ placed $V$ inside $WX$
+    when we encode the revised taxonomy.
 Perhaps, unexpected is the fact that storing the revised taxonomy with an \emph{incertae sedis} annotation
@@
 \end{tabular}
-}\hfill()
+}\hfill
 \subfloat[from revised taxonomy]{
@@
 trees from partial splits through its use of the BUILD algorithm.
-Handling \emph{incertae sedis} taxa thus requires loading incertae
-sedis information and computing partial splits for \emph{incertae
-sedis} taxa before solving a sub-problem.
-After solving a sub-problem,
+Handling \emph{incertae sedis} taxa thus requires loading \emph{incertae
+sedis} information and computing partial splits for \emph{incertae
+sedis} taxa before solving a subproblem.
+After solving a subproblem,
 we must apply taxon names from the taxonomy tree to the sub-problem
@@
 
+\mthcomment{We might be able to skip mentioning the annotation step
+for this chapter$\ldots 
 \subsection{Annotation}
-
-Annotation primarily involves running a conflict analysis between the
+The Open Tree of Life's supertree construction involves an 
+    ``annotation'' step which notes whether or not
+    different groupings in the input trees are displayed
+    by the supertree.
+Annotation essentially involves running a conflict analysis between the
 synthesis tree and each input tree.
@@
 to the conflict-analysis code in the annotation step.
+}
 
@@
 
-In this paper we've described a method for handling \emph{incertae sedis} taxa
-when constructing taxonomic supertrees.  Our method incorporates the uncertain
-placement information from such taxa when constructing the supertree.  It also
+In this chapter we haveve described a method for handling \emph{incertae sedis} taxa
+when constructing taxonomic supertrees.
+Our method incorporates the uncertain
+placement information from such taxa when constructing the supertree.
+It also
 handles higher taxon names and allows input trees to revise the taxonomy by
@@
 
-However, our method leaves a number of problems for future work.  One of these
+However, our method leaves a number of problems for future work.
+One of these
 problems is that we consider taxa with an \emph{incertae sedis} label to 
-about which siblings in attach within.  However, in practice,
-the taxonomist may only consider it plausible for the taxon to intrude only within specific
-siblings.  Our method also assumes that an \emph{incertae sedis} label on a taxon
+``invade'' a subtrees rooted by any of the taxon's siblings.  
+However,
+the taxonomist who made the \emph{incertae sedis} designation,
+may only consider it plausible for the taxon to intrude only within specific
+siblings.
+Our method also assumes that an \emph{incertae sedis} label on a taxon
 indicates that a taxon may attach at any depth within its siblings.  In practice, an
@@
 
-In order to consider more precise attachment ranges for \emph{incertae sedis} taxa, we
-would have to obtain specifications for attachment ranges that are machine readable and
-robust.  Obtaining these specifications would be difficult or impossible in many cases.
-However, in cases that result from taxonomy merging (below) it would be possible to obtain
-them.  This necessitates the creation of a format \emph{incertae sedis} taxa that is more
-expressive that a simple label.
-
-\paragraph{Taxonomy merging}
-
-Not all \emph{incertae sedis} taxa in our taxonomy are directly
+As we demonstrated above (in the discussion of figure \ref{fig:Placing-one-incertae-MRP}),
+if the \emph{incertae sedis} status of a taxon consists solely of
+    a flag on that taxon, we can encounter 
+    multiple equally plausible ways of encoding the information
+    into a character matrix.
+Resolving such ambiguities would require some datastore of intended
+    meanings of each restricted \emph{incertae sedis} annotation.
+In order to consider more precise attachment ranges for \emph{incertae sedis} taxa,
+    we would have to obtain machine readable specifications for attachment ranges
+    for groupings that a taxonomist has labeled as \emph{incertae sedis}
+Obtaining these specifications would be difficult or impossible in many cases.
+However,  not all \emph{incertae sedis} taxa in our taxonomy are directly
 labeled \emph{incertae sedis} by taxonomists.
@@
 uncertain placement.
-The reason is primarily that if taxonomy
-$\taxonomy_{1}$ contains more levels of hierarchy than taxonomy
-$\taxonomy_{2}$, then we must add internal nodes to $\taxonomy_{2}$ to
-align it to $\taxonomy_{1}$.
-However, if taxonomy $\taxonomy_{2}$
-contains more leaves than $\taxonomy_{1}$, then it is unclear if these
-extra leaves should be nested inside the additional internal nodes, or
-not.
-Thus, the extra leaves are marked \emph{incertae sedis}.
-
-For example, if $\taxonomy_{1}=((a,b)x,(c,d)y)z$ and
-$\taxonomy_{2}=(a,b,c,d,e)z$ then $a$ and $b$ in $\taxonomy_{2}$
-should be nested within $x$, but we do not know if $e$ should be
-nested within $x$ or not.
-Thus we obtain $((a,b)x,(c,d)y,?e)z$, where
-$?$ indicates that taxon $e$ is marked \emph{incertae sedis}.
-
-If incertae sedis taxa are cannot be placed within other incertae
-sedis taxa, then such an approach does not work if $y$ is already
-incertae sedis.
-However, we take the approach that \emph{incertae
-sedis} notations are not a general solution to expression of uncertain
-placement.
+
+Taxonomy merging generates a automated \emph{incertae sedis} annotation
+    in a number of ways.
+Perhaps, the easiest cause to contemplate is if
+one input taxonomy,
+$\taxonomy_{1}$, contains more levels of hierarchy than a second
+input taxonomy,
+$\taxonomy_{2}$.
+When the merger adds must add internal nodes that are unique to $\taxonomy_{2}$,
+it is unclear where these
+extra leaves should be nested be placed in the richer hierarchy of $\taxonomy_1$.
+In such a case, it is feasible to imagine the taxonomy merging procedure outputting
+    a list of plausible attachment points for each generated \emph{incertae sedis}
+    label.
+Case in which the \emph{incertae sedis} label is present in an input taxonomy,
+    restricting the range of attachment points would need to be treated as
+    some manually curated statements.
 

### 2018-06-17 doc/incertae_sedis.tex

@@
 
-%\newcommand{\mthcomment}[1]{{\color{red} #1}\xspace}
-\newcommand{\mthcomment}[1]{}
-
-%\newcommand{\bdrcomment}[1]{{\color{cyan} #1}\xspace}
-\newcommand{\bdrcomment}[1]{}
+\newcommand{\mthcomment}[1]{{\color{red} #1}\xspace}
+\renewcommand{\mthcomment}[1]{}
+
+\newcommand{\bdrcomment}[1]{{\color{cyan} #1}\xspace}
+\renewcommand{\bdrcomment}[1]{}
 
@@
 
-In this chapter we haveve described a method for handling \emph{incertae sedis} taxa
-when constructing taxonomic supertrees.
-Our method incorporates the uncertain
-placement information from such taxa when constructing the supertree.
-It also
-handles higher taxon names and allows input trees to revise the taxonomy by
-determining where \emph{incertae sedis} taxa should be placed.
-
-However, our method leaves a number of problems for future work.
-One of these
-problems is that we consider taxa with an \emph{incertae sedis} label to 
-``invade'' a subtrees rooted by any of the taxon's siblings.  
-However,
-the taxonomist who made the \emph{incertae sedis} designation,
+This paper introduces the necessary concepts and methods for handling \emph{incertae sedis}
+taxa in taxonomic supertrees.
+We describe the goals of preserving internal edges,
+preserving higher taxon names, and preserving the opportunity to revise the taxonomy
+based on placement of \emph{incertae sedis} taxa.
+In addition to explaining why
+the naive approach to taxonomic supertrees that ignores \emph{incertae sedis}
+ taxa sacrifices these three goals, we introduce a simple and satisfying way to achieve these goals
+through defining rooted splits for taxonomy edges with reduced exclude sets.
+We also
+provide a simple and efficient recursion for computing these splits.
+
+Despite the simplicity of our approach, some complexities and conceptual difficulties
+remain.
+While the correspondence of nodes and names is a simple one-to-one and onto
+relationship without \emph{incertae sedis} taxa, when \emph{incertae sedis} taxa are
+introduced, the relationship might be neither one-to-one (Figures \ref{fig:Multiple-names-can},\ref{fig:interdigitation}) nor onto (Figure \ref{fig:One-name-can}).
+This necessitates putting more thought into the semantics of taxon names, including
+whether names might become synonyms.  This may be considered a ``complexity''.
+
+In addition, our attempt to define reduced exclude sets for taxonomy splits yields
+a semantics that is not the only concievable semantics for \emph{incertae sedis}
+taxa (Figure \ref{fig:Placing-one-incertae}).  It allows \emph{incertae sedis} siblings to inter-digitate
+(Figure \ref{fig:interdigitation}), which might be a surprising result to the taxonomist.  Furthermore,
+in what might seem a lack of consistency, it does \emph{not} allow \emph{incertae sedis}
+taxa to intrude into each other if they become siblings through the result of
+placement, as is the case for taxa W and V in Figure \ref{fig:Placing-one-incertae}c.
+While each of these problems can be solved separately by adjusting the definition of the
+reduces exclude sets, no solution seems to be best in all respects.  This may be a result
+of the fact that split-based approaches to supertrees seek a simultaneous solution to a set
+of constraints.  This simultaneity can be contrasted with a solution that is phrased
+as a set of sequential operations to the taxonomy, as would be the case when (for example)
+(i) placing V within WX so that it is sister to W, (ii) marking V as no longer \emph{incertae sedis}, and then (iii)
+placing W within V (which is no longer \emph{incertae sedis}).  This maybe considered a ``conceptual difficulty''.
+
+Another issue that we have postponed for future work is how to more narrowly delimit
+the ranges in which \emph{incertae sedis} may attach.  The current paper allows
+\emph{incertae sedis} to attach anywhere within their sisters, but the taxonomist
 may only consider it plausible for the taxon to intrude only within specific
-siblings.
-Our method also assumes that an \emph{incertae sedis} label on a taxon
+siblings.  Furthermore, our method also assumes that an \emph{incertae sedis} label on a taxon
 indicates that a taxon may attach at any depth within its siblings.  In practice, an

### 2018-06-18 doc/incertae_sedis.tex
Add missing a \emph{ } for incertae sedis.

@@
 taxonomies.
-Constructing a range for each incertae sedis taxon relies on detailed
+Constructing a range for each \emph{incertae sedis} taxon relies on detailed
 readings of character argumentation (which is lacking in OTT) or rank-based

### 2018-06-18 doc/incertae_sedis.tex
Make section on naming a bit clearer by adding subsubsection headings.

@@
 
+\subsubsection{One name, multiple nodes}
 However, in the \emph{incertae sedis} framework, it is possible for one name to
@@
 
+\subsubsection{One node, multiple names}
 It is also possible for multiple names to apply to a single node. For example,

### 2018-06-18 doc/incertae_sedis.tex
Move "intermediate" before "final".
Add missing root taxon to character matrices.
Typo fixes.

@@
 See, for example, Figure \ref{fig:Broken-incertae-sedis}.
-Here taxon $T1$ is an descendant of $R$ in the
+Here taxon $T1$ is a descendant of $R$ in the
 synthesis tree, but not the taxonomy.
@@
 }
+\subfloat[Possible intermediate]{\includegraphics[scale=0.5]{Figures/is-within-is/intermed-shaded}
+}
 \subfloat[Final supertree]{\includegraphics[scale=0.5]{Figures/is-within-is/synth1-shaded}
-}
-\subfloat[Possible intermediate]{\includegraphics[scale=0.5]{Figures/is-within-is/intermed-shaded}
 }
@@
 $X$  & 0 & 1 \\
+(root) & 0 & 0 \\
 \hline
@@
 subproblem decomposer.
-One downside of this approach is that, an \emph{incertae sedis} taxon
-    place in only one input tree, the clade it is placed in will be
+One downside of this approach is that, when an \emph{incertae sedis} taxon
+    occurs in only one input tree, the clade it is placed in will be
     marked as contested even though none of the counterintuitive

### 2018-06-18 doc/incertae_sedis.tex
Updating section 2.6

@@
 
+\subsubsection{Revision of unbroken \emph{incertae sedis} taxa}
 After we have attached taxon names to the synthesis tree, we would
@@
 $C$ within $A$.
-%Furthermore, the synthetic tree additionally places
-% $C$ within $AB$.
 
@@
     taxonomic placement of an \emph{incertae sedis} taxon
-    can be detected if the most tip-ward named ancestor of
-    the clade in the supertree was not an ancestor in the
-    taxonomic tree.
-Note that if an \emph{incertae sedis} taxon is broken
-    during the supertree algorithm, it can be ambiguous whether
-    or not all of the members should be placed in a new
-    parent taxon.
-See, for example, Figure \ref{fig:Broken-incertae-sedis}.
-Here taxon $T1$ is a descendant of $R$ in the
-synthesis tree, but not the taxonomy.
-Thus $T1$ is placed within $R$; taxon
- $T2$ is similarly placed with in $R$.
-$T3$ is not mentioned in either input tree, and the 
-    only tree that speaks to its placement is the taxonomic input.
-However, the taxon $T$ cannot be monophyletic based on the
-     two input trees.
-Thus the correct placement of $T3$ is uncertain.
-
-The ambiguity of where to place $T3$ can also be compared to the
-    behavior of maximum compatibility tree estimation and parsimony.
-The matrix representation context the taxonomic tree's
-    character representation of taxon $T$ would include $T1$, $T2$ and $T3$
+    can be detected if the name of the \emph{incertae sedis} taxon
+    survives in the solution tree, and the most tip-ward named
+    ancestor of the clade in the supertree was not an ancestor in the
+    taxonomic tree.  In complex scenarios such as Figure
+    \ref{fig:Placing-one-incertae}c, it may be necessary to
+    walk the solution tree in pre-order and revise the taxonomy
+    incrementally, in order to (for example) place V within WX
+    before placing W within V.
+
+\subsubsection{Revision of broken \emph{incertae sedis} taxa}
+When an \emph{incertae sedis} taxon is broken and its name does not
+occur on the solution tree, taxonomic revision becomes considerably
+more complex.  For example, it can be ambiguous whether or not all of
+the members of a broken taxon should be placed in a new parent
+taxon. See, for example, Figure \ref{fig:Broken-incertae-sedis}. Here
+taxon $T1$ is a descendant of $R$ in the synthesis tree, but not the
+taxonomy. Thus $T1$ is placed within $R$; taxon  $T2$ is similarly
+placed with in $R$. $T3$ is not mentioned in either input tree, and
+the  only tree that speaks to its placement is the taxonomic
+input. However, the taxon $T$ cannot be monophyletic based on the two
+input trees.  Thus the correct placement of $T3$ is unclear.
+Furthermore, the taxon name for $T$ does not occur on the solution tree.
+
+The ambiguity of where to place $T3$ can explored by comparing the
+behavior of maximum-compatibility tree estimation and parsimony-based
+scoring schemes such as MRP. The taxonomic tree's character
+representation of taxon $T$ would include $T1$, $T2$ and $T3$ 
 all being assigned state 1 and all of the other leaves being assigned 
-    state 0.
-This character is not compatible with the solution tree. 
-The compatability approach would be to conclude that the character encoding 
-    that taxon was incorrect and should be ignored.
-Since, no other character speaks to the position of $T3$, it would fall to being
-    a child of the root of the tree.
-In a similar result, the BUILD algorithm would place $T3$ as a child of the root.
-
-In a parsimony view, while the character cannot be mapped onto the solution in
-    a homoplasy-free manner, one might hope that the character data that the
-    taxonomy was based on was not completely unreliable.
-Placing of $T3$ as either sister to $T1$ or $T2$ could explain the character
-    representation of taxon $T$ with only two acquisitions of the state 1.
-Other placements of $T3$ would require at least three 0 to 1 changes (or a reversion
-    to state 0).
-Placing $T3$ at the MRCA 
-    of the other members of $T$ corresponds
-    to handling of ``rogue'' taxa in an Adam's consensus tree \citep{Adams1972} method of the two
-    most parsimonious placements of $T3$ under an irreversible parsimony model.
+state 0.  This character is not compatible with the solution tree. 
+The compatibility approach would be to conclude that the character
+(or, equivalently, the corresponding rooted split) encoding that taxon was incorrect and should be completely ignored.
+Since, no other character speaks to the position of $T3$, it would
+fall to being  a child of the root of the tree.  This seems to
+contradict the intuition that $T3$ should be placed at the MRCA of
+$T1$ and $T2$.
+\bdrcomment{This is not really a feature of BUILD, but of our choice to
+completely throw all all triplets from incompatible splits.}
+
+Using a parsimony-based scoring scheme, the information from the
+character that encodes taxon $T$ is not completely ignored, even though it
+is contradicted by an input tree.  Placing $T3$ as either sister to
+$T1$ or $T2$ could explain the character representation of taxon $T$
+with only two acquisitions of the state 1. Other placements of $T3$
+would require at least three 0 to 1 changes (or a reversion to state
+0). Placing $T3$ at the MRCA of the other members of $T$ corresponds
+to handling of ``rogue'' taxa in an Adam's consensus tree
+\citep{Adams1972} method of the two  most parsimonious placements of
+$T3$ under an irreversible parsimony model.
+
+Our current supertree algorithm places $T3$ at the MRCA of $T1$ and
+$T2$ if $T3$ is a taxonomy-only taxon, but places $T3$ at the root
+of the tree if it occurs in an input tree.  This inconsistent behavior
+is not restricted to cases that contain \emph{incertae sedis} taxa, but is a
+general problem with compatibility-based algorithms that do not retain
+partial information from incompatible splits.  Such information could
+include compatible triplets that comprise parts of the incompatible split.
+The inconsistency results from the fact that our sub-problem solver
+implements compatibility-based consensus and completely discards
+incompatible splits, while the unpruner implements an \emph{ad hoc}
+rule to place pruned taxonomy-only taxa at the MRCA of taxa that occur
+in input trees.  It would be desirable to update the sub-problem
+solver to be consistent with the unpruner, so that we could treat the
+pruning and unpruning steps as an optimization that has no effect on
+semantics.
+
+It is possible to update our compatability-based supertree algorithm
+to achieve the intuitive result of placing $T3$ at the MRCA of $T1$
+and $T2$ without resorting to a parsimony-based scoring scheme. We
+note that the split $T1\,T2\,T3\,|\bullet\,U\,S\,R1\,R2$ implies
+many rooted triplets, and that we need not discard all of these
+triplets because some of them are contradicted by input trees.  A
+simple method of retaining partial information from incompatible
+rooted splits would be to note that the MRCA of $T1$, $T2$ and must
+also have $R1$ and $R2$ as descendants.  We would then remove these
+additional taxa from the exclude group, obtaining the partial split
+$T1\,T2\,T3\,|\bullet\,U\,S$.  Adding this split to the solution
+tree places $T3$ at the MRCA of $T1$ and $T2$ as desired.  While this
+approach places all members of $T$ inside $R$, it does not
+place a monophyletic $T$ within $R$, since $T$ is not monophyletic.
+Placing the name for $T$ at the MRCA of its members would make $T$ a
+synonym for $R$.
 

### 2018-06-18 doc/incertae_sedis.tex
Typos and clean up discussion.

@@
 While each of these problems can be solved separately by adjusting the definition of the
-reduces exclude sets, no solution seems to be best in all respects.  This may be a result
+reduced exclude sets, no solution seems to be best in all respects.  This may be a result
 of the fact that split-based approaches to supertrees seek a simultaneous solution to a set
@@
 the ranges in which \emph{incertae sedis} may attach.  The current paper allows
-\emph{incertae sedis} to attach anywhere within their sisters, but the taxonomist
+\emph{incertae sedis} to attach within any of their siblings, but the taxonomist
 may only consider it plausible for the taxon to intrude only within specific
-siblings.  Furthermore, our method also assumes that an \emph{incertae sedis} label on a taxon
-indicates that a taxon may attach at any depth within its siblings.  In practice, an
-\emph{incertae sedis} genus with sister families (for example) is unlikely to attach
-so deeply within a sister family that it becomes nested within another genus.
-
-As we demonstrated above (in the discussion of figure \ref{fig:Placing-one-incertae-MRP}),
-if the \emph{incertae sedis} status of a taxon consists solely of
-    a flag on that taxon, we can encounter 
-    multiple equally plausible ways of encoding the information
-    into a character matrix.
-Resolving such ambiguities would require some datastore of intended
-    meanings of each restricted \emph{incertae sedis} annotation.
-In order to consider more precise attachment ranges for \emph{incertae sedis} taxa,
-    we would have to obtain machine readable specifications for attachment ranges
-    for groupings that a taxonomist has labeled as \emph{incertae sedis}
-Obtaining these specifications would be difficult or impossible in many cases.
-However,  not all \emph{incertae sedis} taxa in our taxonomy are directly
-labeled \emph{incertae sedis} by taxonomists.
-\emph{Incertae sedis}
-taxa can also result from automatic merging of taxonomies to create
-the OpenTree taxonomy.
-For example, in Figure 3 of
-\citet{rees2017automated}, cases \#4 and \#6 illustrate examples where
-merging of two taxonomies leads to a taxonomy with a taxon of
-uncertain placement.
-
-Taxonomy merging generates a automated \emph{incertae sedis} annotation
-    in a number of ways.
-Perhaps, the easiest cause to contemplate is if
-one input taxonomy,
-$\taxonomy_{1}$, contains more levels of hierarchy than a second
-input taxonomy,
-$\taxonomy_{2}$.
-When the merger adds must add internal nodes that are unique to $\taxonomy_{2}$,
-it is unclear where these
-extra leaves should be nested be placed in the richer hierarchy of $\taxonomy_1$.
-In such a case, it is feasible to imagine the taxonomy merging procedure outputting
+siblings.  We also assume that a taxon labeled as \emph{incertae sedis} may
+attach at any depth within its siblings. In practice, an \emph{incertae sedis}
+genus with sister families (for example) is unlikely to attach so deeply within
+a sister family that it becomes nested within another genus.
+
+Our method also assumes that the \emph{incertae sedis} status of a taxon
+consists solely of a flag on that taxon.  In order to consider more precise
+attachment ranges for \emph{incertae sedis} taxa, we would have to obtain
+machine readable specifications for attachment ranges for groupings that a
+taxonomist has labeled as \emph{incertae sedis}.  Obtaining these specifications
+would be difficult or impossible in many cases. However, not all \emph{incertae
+sedis} taxa in our taxonomy are directly labeled \emph{incertae sedis} by
+taxonomists. \emph{Incertae sedis} taxa can also result from automatic merging
+of taxonomies to create the OpenTree taxonomy. For example, in Figure 3 of
+\citet{rees2017automated}, cases \#4 and \#6 illustrate examples where merging
+of two taxonomies leads to a taxonomy with a taxon of uncertain placement.
+
+Taxonomy merging generates an automated \emph{incertae sedis} annotation
+in a number of ways. Perhaps, the easiest cause to contemplate is if
+one input taxonomy, $\taxonomy_{1}$, contains more levels of hierarchy than a second
+input taxonomy, $\taxonomy_{2}$.
+When the merger adds internal nodes that are unique to $\taxonomy_{2}$,
+it is unclear how these
+extra leaves should be nested or placed in the richer hierarchy of $\taxonomy_1$.
+In such a case, it is feasible to imagine the taxonomy-merging procedure outputting
     a list of plausible attachment points for each generated \emph{incertae sedis}
     label.
-Case in which the \emph{incertae sedis} label is present in an input taxonomy,
+For the case in which the \emph{incertae sedis} label is present in an input taxonomy,
     restricting the range of attachment points would need to be treated as
-    some manually curated statements.
+    manually curated statements.
 

### 2018-06-19 doc/incertae_sedis.tex
minor change

@@
 pruning and unpruning steps as an optimization that has no effect on
-semantics.
+which tree is returned.
 

### 2018-06-20 doc/incertae_sedis.tex
Revise first part of 2.7

--- 2018-06-18 doc/incertae_sedis.tex
@@
 
-\subsection{Ambiguity of imputed taxonomic characters}
-It is tempting to view the matrix representation of the taxonomic tree
-    as an encoding of the presence, absence, and uncertainty about
-    suites of synapomorphies which allow us to diagnose a taxon.
-Clearly by simply representing the taxonomic tree, we will not be
-    able to recover the identity or even the number of characters that
-    a taxonomist might have used when classifying the organisms in question.
-Nevertheless, one might hope that a matrix representation of a taxonomy (such
-    as the matrix in Figure \ref{fig:Handling-incertae-sedis-mrp} as a 
-    representation of Figure \ref{fig:Handling-incertae-sedis}b) has a unique, uncontroversial 
-    association with the taxonomic tree.
-
-The rules for associated leaves in $\mathcal{I}(n)$ with state 1, leaves
-     in $\mathcal{E}(n)$, and other leaves with the missing data symbol 
-     do allow to encode the taxon $n$ in a taxonomic tree unambiguously.
-However, these are not the only set of rules for generating a character matrix
-    representation that would cause the taxonomic tree allows the characters to
-    be mapped without homoplasy.
-For example consider the case of the taxonomy shown in Figure \ref{fig:Placing-one-incertae}a.
-Either of the two matrices shown in \ref{fig:Placing-one-incertae-MRP} could encode this
-    taxonomy along with its \emph{incertae sedis} interpretation.
-The matrix in Figure \ref{fig:Placing-one-incertae-MRP}a would result in the 
-    taxon $V$ not being named in the supertree shown in Figure \ref{fig:Placing-one-incertae}b,
-    while the matrix shown in Figure \ref{fig:Placing-one-incertae-MRP}b would lead
-    to $V$ being named.
-Thus, to honor the input taxonomic content accurately,
-   we need more taxonomic information
-    than simply a hierarchy and knowledge of which taxa can float more tipward.
-
-Note that this example also presents a challenge for iterative supertree building and refinement
-    of the taxonomy without retention of additional information.
+\subsection{Ambiguity of split representation for \emph{incertae sedis} taxa}
+Equation (\ref{eq:exsetformonea}) introduces a method for encoding a
+taxonomy with \emph{incertae sedis} taxa into a set of partial splits.
+These partial splits have a unique represention as a matrix of
+0-1 characters with missing data indicated by '?'.  Clearly, we will
+not be able to recover the identity, or even the number, of characters
+that a taxonomist might have used when classifying a set of organisms,
+simply by looking at the character matrix that represents a taxonomy
+(as Figure \ref{fig:Handling-incertae-sedis-mrp} represents Figure
+\ref{fig:Handling-incertae-sedis}b). Nevertheless, one might hope that 
+equation (\ref{eq:exsetformonea}) is a unique an uncontroversial
+way to encode any taxonomic tree as a set of splits.
+
+However, equation (\ref{eq:exsetformonea}) is not the only set of rules 
+for generating a split representation for a taxonomy. For example
+consider the case of the taxonomy shown in Figure
+\ref{fig:Placing-one-incertae}a. Either of the two matrices shown in
+\ref{fig:Placing-one-incertae-MRP}a or
+\ref{fig:Placing-one-incertae-MRP}b could represent this taxonomy 
+along with its \emph{incertae sedis} interpretation.  The matrix in
+Figure \ref{fig:Placing-one-incertae-MRP}a, which corresponds to
+equation (\ref{eq:exsetformonea}), assumes that \emph{incertae sedis}
+taxa may only intrude into taxa that are siblings in the taxonomy.
+The placement of $W$ within $V$ leads to taxon $V$ not being named in
+the supertree shown in Figure \ref{fig:Placing-one-incertae}c under
+the encoding in \ref{fig:Placing-one-incertae-MRP}a because
+$V$ is not a sibling of $W$ in Figure \ref{fig:Placing-one-incertae}a.
+
+The matrix shown in Figure \ref{fig:Placing-one-incertae-MRP}b allows
+$V$ to be named in the supertree shown in Figure
+\ref{fig:Placing-one-incertae}c because it removes the assumption that
+\emph{incertae sedis} taxa can only be placed within siblings.
+However, as a side-effect, if taxon $WX$ is found to be
+non-monophyletic, the matrix from Figure \ref{fig:Placing-one-incertae-MRP}b
+allows $W$ to be placed within $V$.  In general, this approach would
+allow incertae sedis children of any taxon $\Gamma_1$ to be placed
+arbitrary deep into group $\Gamma_2$ if group $\Gamma_1$ is broken. This
+behavior can lead to surprising and non-intuitive results. Additionally, our unpruner currently assumes
+that \emph{incertae sedis} taxa 
+can only be placed within siblings, and therefore does not know what
+to do in this situation.  Thus, it is not clear that either Figure
+\ref{fig:Placing-one-incertae-MRP}a or Figure
+\ref{fig:Placing-one-incertae-MRP}b comprise a unique, best
+encoding of the taxonomy.  Thus, to honor the input taxonomic content
+accurately, we may need more taxonomic information than simply a
+hierarchy and knowledge of which taxa can float more tipward.
+
+Note that this example also exposes a lack of consistency in
+Equation (\ref{eq:exsetformonea}) that rears its head when attempting to
+iteratively revise placement of \emph{incertae sedis} taxa in the taxonomy
+on successive supertree builds.
+% without retention of additional information.
+
 Consider the case, in which there are two phylogenetic inputs along with the taxonomy from
@@
 \caption{\label{fig:Placing-one-incertae}
-One \emph{incertae sedis} group ($W$) being placed within another ($V$), with possible
-loss of the taxon $V$.}
+  Placement of incertae sedis taxa within non-siblings could occur if
+  taxa become siblings by placement.  Given the taxonomy (a),
+  placement of $V$ within $WX$ is shown in (b), followed by placement
+  of $W$ within $V$ as shown in (c).  Since $W$ is not a sibling of
+  $V$ in (a), the method proposed in this paper and illustrated in Figure
+  \ref{fig:Placing-one-incertae-MRP}a does not preserve the taxon name
+  for $V$.  However, with an alternative method for computing reduced
+  exclude sets that is shown in \ref{fig:Placing-one-incertae-MRP}b,
+  the name $V$ would be preserved.
+  }
 \end{figure}

### 2018-06-20 doc/incertae_sedis.tex
Finish revising section 2.7

@@
 
-Note that this example also exposes a lack of consistency in
-Equation (\ref{eq:exsetformonea}) that rears its head when attempting to
-iteratively revise placement of \emph{incertae sedis} taxa in the taxonomy
-on successive supertree builds.
-% without retention of additional information.
-
-Consider the case, in which there are two phylogenetic inputs along with the taxonomy from
-    tree Figure \ref{fig:Placing-one-incertae}b with the first phylogeny, $T_1$, providing the single
-    rooted split: $V1\,X \mid \bullet\,U$ and the second phylogeny, $T_2$ providing the 
-    rooted split: $W\, V1 \mid \bullet\,V2$.
-If the supertree pipeline were run with just the taxonomy and and $T_1$ as inputs, then the supertree
-    would be the tree shown in Figure \ref{fig:Placing-one-incertae}c.
-One might hope that using that tree as the revised taxonomy and then adding the split from $T_2$ in
-    a subsequent analysis would result in the same tree as if $T_1$ and $T_2$ to the original taxonomy.
-While the topology for the final tree in either order of analyses would indeed be the tree shown in 
-    Figure \ref{fig:Placing-one-incertae}b, the decision about whether or not to name $V$ would differ.
-Our handling of \emph{incertae sedis} in the original taxonomy via rooted partial splits
-    would encode the taxonomic definitions in the manner depicted in Figure \ref{fig:Placing-one-incertae-MRP}a,
-    but the encoding of the revised taxonomy in Figure \ref{fig:Placing-one-incertae}c would result
-    in an altered definition of taxon $V$, as shown in \ref{fig:Placing-one-incertae-MRP}c.
-As expected,  the definition of $WX$ has been altered to reflect the
-    fact $T_1$ placed $V$ inside $WX$
-    when we encode the revised taxonomy.
-Perhaps, unexpected is the fact that storing the revised taxonomy with an \emph{incertae sedis} annotation
-    and then reprocessing the taxonomy using our rooted partial splits approach would also shift
-    the definition of $V$ from a taxon that must exclude $W$ to one that may include $W$.
+Note that equation (\ref{eq:exsetformonea}) allows \emph{incertae sedis}
+taxa to intrude only into siblings in the original taxonomy, and not
+into taxa that become siblings through solution to the supertree
+problem. This can lead to cases where computing a supertree from two
+input trees $T_1$ and $T_2$ and an original taxonomy $\taxonomy$ can
+lead to a different outcome than if we compute a supertree from $T_1$
+and $\taxonomy$, and then compute a supertree from $T_2$ and a revised
+taxonomy $\taxonomy^\prime$. Consider a case in which the taxonomy
+$\taxonomy$ is given in Figure \ref{fig:Placing-one-incertae}a (with
+matrix representation given by \ref{fig:Placing-one-incertae-MRP}a)
+and $T_1$ and $T_2$ each provide the single rooted splits $V1\,X \mid \bullet\,U$ and $W\, V1 \mid
+\bullet\,V2$ respectively. 
+Running the supertree pipeline on $T_1$  and $\taxonomy$ yields the tree in
+Figure \ref{fig:Placing-one-incertae}b.  We use this tree as the revised
+taxonomy $\taxonomy^\prime$, and its matrix representation appears in
+Figure \ref{fig:Placing-one-incertae-MRP}c.  Running the supertree
+pipeline on $T_2$  and $\taxonomy^\prime$ then
+results in the tree in Figure \ref{fig:Placing-one-incertae}c.
+
+One might therefore hope that running the supertree pipeline on $T_1$,
+$T_2$, and $\taxonomy$ simultaneously would yield the same result.
+While the topology for the final tree would indeed be the tree shown in 
+    Figure \ref{fig:Placing-one-incertae}c, the decision about whether
+    or not to name $V$ differs.  As expected,  the definition of $WX$
+    in $\taxonomy^\prime$ has been altered to reflect the fact $T_1$
+    placed $V$ inside $WX$.  Perhaps unexpected is the fact that
+    storing the revised taxonomy with an \emph{incertae sedis}
+    annotation and then reprocessing the taxonomy using our rooted
+    partial splits approach would also shift the definition of $V$
+    from a taxon that must exclude $W$ to one that may include $W$. 
+    This results from the fact that equation (\ref{eq:exsetformonea})
+    only allows \emph{incertae sedis} taxa to intrude into taxa that
+    are their
+    sisters in the taxonomy.  Outcomes differ because $W$ is a sister of $V$ in
+    $\taxonomy^\prime$ but not $\taxonomy$.  Use of the alternate
+    representation of $\taxonomy$ in Figure \ref{fig:Placing-one-incertae-MRP}b
+    allows $V$ to be named in Figure \ref{fig:Placing-one-incertae}c,
+    which solves the seeming inconsistency.  However, this has some
+    surprising side-effects as mentioned above, and is therefore not
+    the approach that we present in this paper.
 
@@
 annotation shown in the taxonomy from in Figure \ref{fig:Placing-one-incertae}a.
-The representations only vary in the scoring of leaf $W$ for character $V$. Panel (c) shows the information encoded by the tree in \ref{fig:Placing-one-incertae}c. }
+The representations only vary in the scoring of leaf $W$ for character $V$. Panel (c) shows the information encoded by the tree in \ref{fig:Placing-one-incertae}b. }
 \end{figure}

### 2018-06-20 doc/incertae_sedis.tex
Interchange 8b and 8c.

@@
 \ref{fig:Placing-one-incertae-MRP}a or
-\ref{fig:Placing-one-incertae-MRP}b could represent this taxonomy 
+\ref{fig:Placing-one-incertae-MRP}c could represent this taxonomy 
 along with its \emph{incertae sedis} interpretation.  The matrix in
@@
 
-The matrix shown in Figure \ref{fig:Placing-one-incertae-MRP}b allows
+The matrix shown in Figure \ref{fig:Placing-one-incertae-MRP}c allows
 $V$ to be named in the supertree shown in Figure
@@
 However, as a side-effect, if taxon $WX$ is found to be
-non-monophyletic, the matrix from Figure \ref{fig:Placing-one-incertae-MRP}b
+non-monophyletic, the matrix from Figure \ref{fig:Placing-one-incertae-MRP}c
 allows $W$ to be placed within $V$.  In general, this approach would
@@
 \ref{fig:Placing-one-incertae-MRP}a or Figure
-\ref{fig:Placing-one-incertae-MRP}b comprise a unique, best
+\ref{fig:Placing-one-incertae-MRP}c comprise a unique, best
 encoding of the taxonomy.  Thus, to honor the input taxonomic content
@@
 taxonomy $\taxonomy^\prime$, and its matrix representation appears in
-Figure \ref{fig:Placing-one-incertae-MRP}c.  Running the supertree
+Figure \ref{fig:Placing-one-incertae-MRP}b.  Running the supertree
 pipeline on $T_2$  and $\taxonomy^\prime$ then
@@
     $\taxonomy^\prime$ but not $\taxonomy$.  Use of the alternate
-    representation of $\taxonomy$ in Figure \ref{fig:Placing-one-incertae-MRP}b
+    representation of $\taxonomy$ in Figure \ref{fig:Placing-one-incertae-MRP}c
     allows $V$ to be named in Figure \ref{fig:Placing-one-incertae}c,
@@
   for $V$.  However, with an alternative method for computing reduced
-  exclude sets that is shown in \ref{fig:Placing-one-incertae-MRP}b,
+  exclude sets that is shown in \ref{fig:Placing-one-incertae-MRP}c,
   the name $V$ would be preserved.
@@
 %\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-within-is/tree1\lyxdot tre}}
-\subfloat[maximally scored]{
+\subfloat[from original taxonomy]{
     \begin{tabular}{|r|p{2em}|p{2em}|}
@@
 }\hfill{}
-\subfloat[less informative]{
-    \begin{tabular}{|r|p{2em}|p{2em}|}
-\hline
-& \multicolumn{2}{|c|}{Character} \\
-taxa & $V$ & $WX$  \\
-\hline
-$U$  & 0 & 0 \\
-$V1$ & 1 & ? \\
-$V2$ & 1 & ? \\
-$W$  & ? & 1 \\
-$X$  & 0 & 1 \\
-(root) & 0 & 0 \\
-\hline
-\end{tabular}
-}\hfill
 \subfloat[from revised taxonomy]{
@@
 \end{tabular}
+}\hfill
+\subfloat[alternative encoding]{
+    \begin{tabular}{|r|p{2em}|p{2em}|}
+\hline
+& \multicolumn{2}{|c|}{Character} \\
+taxa & $V$ & $WX$  \\
+\hline
+$U$  & 0 & 0 \\
+$V1$ & 1 & ? \\
+$V2$ & 1 & ? \\
+$W$  & ? & 1 \\
+$X$  & 0 & 1 \\
+(root) & 0 & 0 \\
+\hline
+\end{tabular}
 }
 \caption{\label{fig:Placing-one-incertae-MRP}
-Panels (a) and (b) show two possible encodings that would support the $V$ and $WX$ branches and \emph{incertae sedis}
+Panels (a) and (c) show two possible encodings that would support the $V$ and $WX$ branches and \emph{incertae sedis}
 annotation shown in the taxonomy from in Figure \ref{fig:Placing-one-incertae}a.
-The representations only vary in the scoring of leaf $W$ for character $V$. Panel (c) shows the information encoded by the tree in \ref{fig:Placing-one-incertae}b. }
+The representations only vary in the scoring of leaf $W$ for character $V$. Panel (b) shows the information encoded by the tree in \ref{fig:Placing-one-incertae}b. }
 \end{figure}

### 2018-06-20 doc/incertae_sedis.tex
Update.

@@
 \ref{fig:Handling-incertae-sedis}b). Nevertheless, one might hope that 
-equation (\ref{eq:exsetformonea}) is a unique an uncontroversial
+equation (\ref{eq:exsetformonea}) is a unique and uncontroversial
 way to encode any taxonomic tree as a set of splits.
@@
 allows $W$ to be placed within $V$.  In general, this approach would
-allow incertae sedis children of any taxon $\Gamma_1$ to be placed
+allow \emph{incertae sedis} children of any taxon $\Gamma_1$ to be placed
 arbitrary deep into group $\Gamma_2$ if group $\Gamma_1$ is broken. This
@@
     surprising side-effects as mentioned above, and is therefore not
-    the approach that we present in this paper.
+    the approach that we focus on here.
 
@@
 remain.
-While the correspondence of nodes and names is a simple one-to-one and onto
-relationship without \emph{incertae sedis} taxa, when \emph{incertae sedis} taxa are
-introduced, the relationship might be neither one-to-one (Figures \ref{fig:Multiple-names-can},\ref{fig:interdigitation}) nor onto (Figure \ref{fig:One-name-can}).
+While the correspondence of nodes and names is a simple one-to-one function
+without \emph{incertae sedis} taxa, when \emph{incertae sedis} taxa are
+introduced, the relationship might be neither one-to-one (Figures
+\ref{fig:Multiple-names-can},\ref{fig:interdigitation}) nor a function (Figure \ref{fig:One-name-can}).
 This necessitates putting more thought into the semantics of taxon names, including

### 2020-12-05 stashed/doc/paper1-submission/paper5-adjusted.tex
moving to a committed stash to avoid clashes while restoring functionality

--- 2017-01-30 doc/paper1-submission/paper5-adjusted.tex
@@
 
-\caption{Exemplified input trees (a-b) and taxonomy tree (c) from Fig. \ref{fig:cleaned-phylo-example}.
-Taxon E in (a) is exemplified by E1. Pruned taxa are E2, F2, and D. The
-taxa E and F are retained as monotypic taxa in the pruned taxonomy
-$\protect\prunedTaxonomy$. The red
+\caption{Exemplified input trees (a-b) and pruned taxonomy tree (c) from Fig. \ref{fig:cleaned-phylo-example}.
+Taxon E in the first input tree is exemplified by E1 in (a). Pruned taxa are E2, F2, and D. The
+taxa E and F are retained as monotypic taxa in the pruned taxonomy. The red
 edge in the pruned taxonomy tree is an uncontested higher taxon in
@@
 
-\caption{Subproblems \emph{ABCD} and \emph{root} generated from the exemplified trees shown in Fig. \ref{fig:exemplify-example}.
+\caption{Subproblems (a) \emph{ABCD} and (b) \emph{root} generated from the exemplified trees shown in Fig. \ref{fig:exemplify-example}.
 A trivial statement from the first tree that a taxon labelled $ABCD$
 is sister to $E$ has been omitted, because trees with only 2 leaves
-do not contain phylogenetic information.
-Similarly, the trivial problems that list E1 and F1 as the only daughters
-of taxa E and F have been omitted.}
+do not contain phylogenetic information.}
 \label{fig:subproblems-example}
@@
 
-\caption{Solutions to subproblem \emph{ABCD} and subproblem \emph{root} depicted in Fig. \ref{fig:subproblems-example}.}
+\caption{Solutions to (a) subproblem \emph{ABCD} and (b) subproblem \emph{root} depicted in Fig. \ref{fig:subproblems-example}.}
 \label{fig:subproblem-solutions.}
@@
 backbone onto which taxa that are not included in any phylogeny
-will be placed.
-In a real analysis, the tips shown as E and F here
-would be monotypic taxa (with children E1 and F1, respectively)
-from the trivial subproblems that were omitted from from Fig. \ref{fig:subproblems-example}.}
+will be placed.}
 \label{fig:grafted-solution}
@@
 \caption{\label{fig:displayed-annotations}The relationship of edges in summary
-tree $\protect\summaryTree$ (b) to edges in the input tree $T_{1}$
+tree $\protect\summaryTree$ in panel (b) to edges in the input tree $T_{1}$ named ``tree1'' in panel
 (a). Only edges of $\protect\summaryTree$ that are present in the
-induced tree $\protect\summaryTree(1)$ are represented by JSON annotations
-(d). Taxon names are here suppressed in favor of OTT IDs, and edges
+induced tree $\protect\summaryTree(1)$ in panel (c) are represented by JSON annotations
+in panel (d). Taxon names are here suppressed in favor of OTT IDs, and edges
 are referenced via their tipward nodes. Edges in $\protect\summaryTree(1)$
